### 2024-11-24 README.qmd
Initial commit

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
# Please put your title here to include it in the file below.
Title <- "Title of your paper goes here"
```

# article3

[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh///master?urlpath=rstudio)

This repository contains the data and code for our paper:

> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>

Our pre-print is online here:

> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to cite

Please cite this compendium as:

> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

## Contents

The **analysis** directory contains:

  - [:file\_folder: paper](/analysis/paper): Quarto source document
    for manuscript. Includes code to reproduce the figures and tables
    generated by the analysis. It also has a rendered version,
    `paper.docx`, suitable for reading (the code is replaced by figures
    and tables in this file)
  - [:file\_folder: data](/analysis/data): Data used in the analysis.
  - [:file\_folder: figures](/analysis/figures): Plots and other
    illustrations
  - [:file\_folder:
    supplementary-materials](/analysis/supplementary-materials):
    Supplementary materials including notes and other documents
    prepared and collected during the analysis.

## How to run in your browser or download and run locally

This research compendium has been developed using the statistical programming
language R. To work with the compendium, you will need
installed on your computer the [R software](https://cloud.r-project.org/)
itself and optionally [RStudio Desktop](https://rstudio.com/products/rstudio/download/).

You can download the compendium as a zip from from this URL:
[master.zip](/archive/master.zip). After unzipping:
- open the `.Rproj` file in RStudio
- run `devtools::install()` to ensure you have the packages this analysis depends on (also listed in the
[DESCRIPTION](/DESCRIPTION) file).
- finally, open `analysis/paper/paper.Rmd` and knit to produce the `paper.docx`, or run `rmarkdown::render("analysis/paper/paper.qmd")` in the R console

### Licenses

**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse

### Contributions

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

### 2024-11-24 analysis/paper/paper.qmd
Git test

---
title: "Placeholder for article 3"
author:
  - Jane Doe:
      correspondence: "yes"
      email: janedoe@fosg.org
      orcid: 0000-0003-1689-0557
      institute:
        - fosg
        - fop
  - John Q. Doe:
      institute: fosg
      orcid: 0000-0003-1689-0558
  - Peder Ås:
      institute: fosg
      orcid: 0000-0003-1689-0559
  - Juan Pérez:
      orcid: 0000-0003-1689-0551
      institute:
        - name: Acme Corporation
  - Max Mustermann:
      orcid: 0000-0003-1689-0552
institute:
  - fosg:
      name: Formatting Open Science Group
      address: 23 Science Street, Eureka, Mississippi, USA
  - fop: Federation of Planets
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The actual document text starts here: -->

# Introduction

Here is a citation [@Marwick2017]

# Background

# Methods

# Results

<!-- Here's some example analysis code: -->

```{r}
#| label: get-data
#| eval: false
# Note the path that we need to use to access our data files when rendering this document
my_data <- read.csv(here::here('analysis/data/raw_data/my_csv_file.csv'))
```

```{r}
#| label: fig-demo-plot
#| fig-cap: "A plot of random numbers"
plot(rnorm(10))
```

@fig-demo-plot shows how we can have a caption and cross-reference for a plot. Note that figure label and cross-references must both be prefixed with `fig-`

```{r}
#| label: demo-inline-code
#| echo: false
x <- round(pi, 2)
```

Here is an example of inline code `r x` in the middle of a sentence.

# Discussion

# Conclusion

# Acknowledgements

<!-- The following line inserts a page break  -->

\newpage

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

::: {#refs}
:::

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r}
#| label: colophon
#| cache: false

# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```

### 2024-12-21 analysis/paper/paper.qmd
Long commit, includes: database cleanup and setup for analysis

@@
 execute:
-  echo: true
+  echo: false
   warning: false
@@
 ```{r}
-#| label: get-data
-#| eval: false
-# Note the path that we need to use to access our data files when rendering this document
-my_data <- read.csv(here::here('analysis/data/raw_data/my_csv_file.csv'))
+#| label: library
+
+library(readr)
+library(dplyr)
+library(tidyr)
+library(stringr)
+
 ```
@@
 ```{r}
-#| label: fig-demo-plot
-#| fig-cap: "A plot of random numbers"
-plot(rnorm(10))
+#| label: get-data
+#| eval: false
+
+# Technological data
+gravettian_tech <- read.csv(here::here('analysis/data/raw_data/VBT_6_7_JB.csv'))
+proto_tech <- read.csv(here::here('analysis/data/raw_data/VBT_4E_5_JB.csv'))
+solutrean_JC <- read.csv(here::here('analysis/data/raw_data/VBR_ABC_JC.csv'))
+solutrean_points <- read.csv(here::here('analysis/data/raw_data/shelter_points_JC.csv'))
+
+# RM data
+shelter_RM <- read_csv(here::here('analysis/data/raw_data/MACRO_IND_SHELTER.csv')) # Shelter RM analysis
+terrace_RM <- read_csv(here::here('analysis/data/raw_data/MACRO_IND_TERRACE.csv')) # Terrace RM analysis
+
+# Context data
+context <- read_csv(here::here('analysis/data/raw_data/VB_Terrace_context.csv')) # Terrace Field DB - lithic artefact info
+xyz <- read_csv(here::here('analysis/data/raw_data/VB_Terrace_xyz.csv')) # Terrace Field DB - just coordinates
+
 ```
 
-@fig-demo-plot shows how we can have a caption and cross-reference for a plot. Note that figure label and cross-references must both be prefixed with `fig-`
+```{r}
+#| label: terrace-databases-setup
+
+cols_to_concat_terrace <- c("UNIT", "ID")
+context$ID <- str_squish(context$ID) # Remove empty spaces from ID field
+xyz$ID <- str_squish(xyz$ID) # Remove empty spaces from UNIT field
+
+proto_tech_WIP <- proto_tech %>%
+  tidyr::unite(col='ID', all_of(cols_to_concat_terrace), sep="-", remove=TRUE) %>% 
+  rename(CLASS = Class, COREPREPPROD = CorePreparProd, PIECECOMPLETENESS = PieceCompleteness, CORTEX = Cortex, CORTEXLOCATION = CortexLocation, PLATFORMTYPE = PlatformType, PLATFORMCORTEX = PlatformCortex, SCARCOUNT = ScarCount, THICKNESS = Thickness,
+        MAXWIDTH = MaxWidth, LENGTH = Length, COREPLATFORM = CorePlatform, MAINFACECORTEX = MainFaceCortex, 
+        MAINFACESCARCOUNT = MainFaceScarCount, COREABANDON = CoreAbandon, RETOUCHEDPIECETYOPOLOGY = RetouchedPieceTypology) %>% 
+  mutate(RETOUCHEDPIECETYOPOLOGY = case_when(
+    RETOUCHEDPIECETYOPOLOGY == "Right truncation on retouched piece" ~ "Truncation",
+    RETOUCHEDPIECETYOPOLOGY == "Double backed bladelet" | RETOUCHEDPIECETYOPOLOGY == "Backed bladelet parcial" ~ "Backed bladelet",
+    RETOUCHEDPIECETYOPOLOGY == "Dentilhado" ~ "Denticulate",
+    RETOUCHEDPIECETYOPOLOGY == "Burin on truncation" ~ "Burin",
+    TRUE ~ RETOUCHEDPIECETYOPOLOGY)) %>% 
+  dplyr::select(ID, CLASS, COREPREPPROD, PIECECOMPLETENESS, CORTEX, CORTEXLOCATION, PLATFORMTYPE, PLATFORMCORTEX,
+                SCARCOUNT, THICKNESS, MAXWIDTH, LENGTH, COREPLATFORM, MAINFACECORTEX, MAINFACESCARCOUNT, COREABANDON,
+                RETOUCHEDPIECETYOPOLOGY)
+gravettian_tech <- gravettian_tech %>% 
+  mutate(RETOUCHEDPIECETYPOLOGY = case_when(
+    RETOUCHEDPIECETYPOLOGY == "Splintered piece with notch" | RETOUCHEDPIECETYPOLOGY == "Splintered piece with lateral retouch" ~ "Splintered piece",
+    RETOUCHEDPIECETYPOLOGY == "Retouched flake siret fragment" | RETOUCHEDPIECETYPOLOGY == "Retouched flake frag" ~ "Retouched flake fragment",
+    RETOUCHEDPIECETYPOLOGY == "Retouched bladelet frag" ~ "Retouched bladelet fragment",
+    RETOUCHEDPIECETYPOLOGY == "Multiple burin" | RETOUCHEDPIECETYPOLOGY == "Burin splintered piece" | RETOUCHEDPIECETYPOLOGY == "Burin on truncation" ~ "Burin",
+    RETOUCHEDPIECETYPOLOGY == "Double backed bladelet" ~ "Backed bladelet",
+    TRUE ~ RETOUCHEDPIECETYPOLOGY))
+
+terrace_tech <- bind_rows(proto_tech_WIP, gravettian_tech)
+
+terrace_RM_WIP <- terrace_RM %>%
+  mutate(RMTYPE = case_when(
+    RMTYPE == "Type 2" ~ "TYPE 2",
+    RMTYPE == "Chalcedony" ~ "TYPE 1", # Altering first group names to final group names. Subtypes are maintained as before.
+    RMTYPE == "TYPE 15" ~ "TYPE 3",
+    RMTYPE == "TYPE 7" ~ "TYPE 4",
+    RMTYPE == "TYPE 12" ~ "TYPE 5",
+    RMTYPE == "TYPE 11" ~ "TYPE 6",
+    RMTYPE == "TYPE 13" ~ "TYPE 7",
+    RMTYPE == "TYPE 11F" ~ "TYPE 8",
+    RMTYPE == "TYPE 13B" ~ "TYPE 9",
+    RMTYPE == "Oolitic" ~ "TYPE 10",
+    RMTYPE == "TYPE GB" ~ "TYPE 11",
+    .default = RMTYPE)) %>% 
+  select(AREA, ID, RMTYPE, SUBTYPE, WEIGHT, CORTEXTHICKNESS, CORTEXAPPEARANCE) %>%
+  distinct(ID, .keep_all = TRUE)
+
+context <- context %>%
+  dplyr::rename("UNIT" = "Unit") %>% # Renaming first column from Unit to UNIT to use cols_to_concat
+  unite(col='ID', cols_to_concat_terrace, sep="-", remove=FALSE) %>% # Unite UNIT with ID using cols_to_concat
+  select(ID, Level, Spit, Code) %>% # Selecting important variables: ID refers to artifact ID; Level refers to archaeologycal layer; Spit refers to artificial levels within an archaeological layer; Code refers to type of artifact (e.g., LITHIC or BONE)
+  distinct(ID, .keep_all = TRUE)
+
+xyz <- xyz %>%
+  unite(col='ID', cols_to_concat_terrace, sep="-", remove=FALSE) %>% # Unite UNIT with ID using cols_to_concat
+  filter(Suffix == 0) %>% # To avoid several entries related to several points of the same artefact
+  select(ID, X, Y, Z) %>% # Selecting important variables
+  distinct(ID, .keep_all = TRUE)
+
+field_data <- full_join(xyz, context, by = "ID") # Join field tables
+
+terrace_tech_WIP <- left_join(terrace_tech, field_data, by = "ID") # Join field and RM type tables
+terrace_dataset <- full_join(terrace_tech_WIP, terrace_RM_WIP, by = "ID") # Base 12690
+terrace_dataset <- terrace_dataset %>%
+  filter(!Level %in% c("3", "4", "4C", "4D")) %>%  # Dropping NAs on the technological class drops dataset to 1651
+  drop_na(RMTYPE) %>%  # Dropping NAs on the RM database drops dataset to 3635
+  drop_na(CLASS) # Dropping NAs on the technological class drops dataset to 1647
+
+# Mutate level 5 to add phases: top of 5 (4E/5) and bottom of 5 (5), following Belmiro et al. (2020).
+phase_dataset <- terrace_dataset %>% 
+  mutate(Level = case_when(
+    Level == 5 & Z > 24.1 ~ "4E/5",
+    Level == 5 & Z < 24.1 ~ "5",
+    Level == "7C" | Level == 8 ~ "7", # Collapsing samples from Level 7C and Level 8 into 7 to homogenise dataset
+    TRUE ~ Level)) %>% # Adding technocomplex name based on the levels
+  mutate(Technocomplex = case_when(
+    Level == "4E" | Level == "4E/5" | Level == 5 ~ "Proto-Solutrean",
+    Level == 6 | Level == 7 ~ "Gravettian"))
+
+```
 
 ```{r}
-#| label: demo-inline-code
-#| echo: false
-x <- round(pi, 2)
+#| label: shelter-solutrean-setup
+
+cols_to_concat_shelter <- c("YEAR", "ID") # Concatenate YEAR and ID variables in one single variable (e.g. 05-500)
+
+solutrean_WIP <- solutrean_JC %>% 
+  filter(Layer %in% c("A", "B", "C")) %>% # Filtering Solutrean layers
+  filter(Raw.material %in% c("Chert", "Chalcedony")) %>% # Filtering raw materials
+  filter(Screen != "Crivo") %>%  # Filtering out all bucket lithics since they were not included in the RM analysis
+  mutate_at("Date", str_replace, "20", "") %>% # Removing initial part of the year to match IDs
+  dplyr::rename("YEAR" = "Date") %>% 
+  tidyr::unite(col='ID', all_of(cols_to_concat_shelter), sep="-", remove=FALSE) # Concatenate YEAR and ID variables in one single variable (e.g. 05-500)
+
+# In Cascalheira 2010 Retouched piece was not a category but a variable. To match the other DBs, this category was added
+solutrean_WIP <- solutrean_WIP %>% 
+  mutate(Class = case_when( # This alteration is being done on the original class column
+    Retouched == "Yes" ~ "Retouched piece", # Whenever retouch is present, class is now Retouched piece
+    TRUE ~ Class)) %>% 
+  mutate(CLASS = case_when( # Creating a column for clustered classes, to match the other DBs
+    Class == "Blade" | Class == "Bladelet" ~ "ElongBlank", # Cluster blades with bladelets. The distinction can come later through a mutate and the measurements
+    Class == "Core front" | Class == "Core tablet" | Class == "Core trimming elem." | Class == "Crested piece" ~ "CorePrepProd", # Cluster all preparation products
+    Class == "Fragment" ~ "Shatter",
+    Class == "Flake distal frag." | Class == "Flake mesial frag." | Class == "Flake prox. frag." ~ "FlakeFrag", # Cluster flake frags into a single category
+    Class == "Blade distal frag." | Class == "Blade mesial frag." | Class == "Blade prox. frag." 
+    | Class == "Bladelet distal frag." | Class == "Bladelet mesial frag." | Class == "Bladelet prox. frag." ~ "ElongBlankFrag", # Cluster elong frags into a single category
+    TRUE ~ Class)) %>% 
+  mutate(COREPREPPROD = case_when( # Creating a column for the types of core preparation and maintenance products (COREPREPPROD)
+    Class == "Core front" ~ "Core front",
+    Class == "Core tablet" ~ "Core tablet",
+    Class == "Core trimming elem." ~ "Core trim",
+    Class == "Crested piece" ~ "Crested piece")) %>% 
+  mutate(PIECECOMPLETENESS = case_when( # Creating a column for the portions present in flake or elongblank fragments to match other DBs
+    Class == "Flake distal frag." | Class == "Blade distal frag." | Class == "Bladelet distal frag." ~ "Distal",
+    Class == "Flake mesial frag." | Class == "Blade mesial frag." | Class == "Bladelet mesial frag." ~ "Mesial",
+    Class == "Flake prox. frag." | Class == "Blade prox. frag." | Class == "Bladelet prox. frag." ~ "Proximal")) %>% 
+  rename(PLATFORMTYPE = Platform_type) %>% # Renaming column
+  mutate(PLATFORMTYPE = case_when( # Renaming variables to match correspondent variables in the other DBs
+    PLATFORMTYPE == "Multifacetted" ~ "Faceted",
+    PLATFORMTYPE == "Unfacetted" ~ "Plain",
+    TRUE ~ PLATFORMTYPE)) %>%
+  mutate(Type = case_when( # Translating retouched types from Zilhão 1997 to simplified typologies
+    Type == "92b" ~ "Retouched piece fragment",
+    Type == "92a" ~ "Retouched blank",
+    Type == "76" ~ "Splintered piece",
+    Type == "75" ~ "Denticulate",
+    Type == "74" ~ "Notch",
+    Type == "66" ~ "Retouched blade",
+    Type == "61" ~ "Truncation",
+    Type == "28" ~ "Burin",
+    Type == "7" | Type == "5b" | Type == "4"  ~ "Endscrapper",
+    TRUE ~ Type)) %>%
+  rename(LEVEL = Layer, UNIT = Unit, CORTEX = Percent_of_cortex, CORTEXLOCATION = Loc_of_cortex, THICKNESS = Thickness,
+         MAXWIDTH = Width, LENGTH = Length, COREPLATFORM = Core_platfform, SCARCOUNT = N_of_Scars, COREABANDON = Core_abandon,
+         RETOUCHEDPIECETYPOLOGY = Type)  # Renaming column
+
+solutrean_tech <- solutrean_WIP %>%
+  dplyr::select(ID, CLASS, LEVEL, UNIT, COREPREPPROD, PIECECOMPLETENESS, CORTEX, CORTEXLOCATION, PLATFORMTYPE,
+                SCARCOUNT, THICKNESS, MAXWIDTH, LENGTH, COREPLATFORM, COREABANDON, RETOUCHEDPIECETYPOLOGY)
+
+solutrean_points <- solutrean_points %>% # Creating ID from 
+  mutate_at("YEAR", str_replace, "20", "") %>% # Removing initial part of the year to match IDs
+  tidyr::unite(col='ID', all_of(cols_to_concat_shelter), sep="-", remove=TRUE)
+
+solutrean_tech <- bind_rows(solutrean_tech, solutrean_points) # Uniting technology dataset with points (2167 to 2228)
+
+solutrean_tech <- solutrean_tech %>% # Originally 2228
+  distinct(ID, .keep_all = TRUE) # Keeping individual IDs to avoid buckets (down to 1993)
+
+shelter_RM_WIP <- shelter_RM %>%
+  mutate_at("ID", str_replace, "VB", "") %>% # Remove "VB" from ID
+  mutate(RMTYPE = case_when(
+    RMTYPE == "Type 2" ~ "TYPE 2",
+    RMTYPE == "Chalcedony" ~ "TYPE 1", # Altering first group names to final group names. Subtypes are maintained as before.
+    RMTYPE == "TYPE 15" ~ "TYPE 3",
+    RMTYPE == "TYPE 7" ~ "TYPE 4",
+    RMTYPE == "TYPE 12" ~ "TYPE 5",
+    RMTYPE == "TYPE 11" ~ "TYPE 6",
+    RMTYPE == "TYPE 13" ~ "TYPE 7",
+    RMTYPE == "TYPE 11F" ~ "TYPE 8",
+    RMTYPE == "TYPE 13B" ~ "TYPE 9",
+    RMTYPE == "Oolitic" ~ "TYPE 10",
+    RMTYPE == "TYPE GB" ~ "TYPE 11",
+    RMTYPE == "ND_A" ~ "INDET",
+    .default = RMTYPE)) %>%
+  select(AREA, ID, RMTYPE, SUBTYPE, WEIGHT, CORTEXTHICKNESS, CORTEXAPPEARANCE) %>%  # Select variables
+  distinct(ID, .keep_all = TRUE) # Keeping individual IDs to avoid buckets
+
+solutrean_dataset <- full_join(solutrean_tech, shelter_RM_WIP, by = "ID") # The complete dataset has 2818 (44 points) entries
+solutrean_dataset <- solutrean_dataset %>%
+  drop_na(RMTYPE) %>%   # Dropping NAs on the RM database drops dataset to 1514
+  drop_na(CLASS) # Dropping NAs on the technological class drops dataset to 914
+
+```
+
+```{r}
+
+library(ggplot2)
+library(scales)
+
+solutrean_dataset %>%
+  filter(RMTYPE != "INDET") %>% 
+  dplyr::group_by(LEVEL, RMTYPE) %>% 
+  dplyr::summarise(n = n()) %>%
+  mutate(percent = n/sum(n)) %>% 
+  ggplot(aes(x = LEVEL, y = percent, fill = RMTYPE)) +
+  geom_bar(stat = "identity", width = 0.5) +
+  scale_y_continuous(labels = scales::percent_format()) +
+  #theme_minimal() + #type of theme
+  labs(y = "Count %") +# change labels
+  guides(fill=guide_legend(title=NULL)) + # remove legend title
+  geom_text(aes(label=ifelse(percent >= 0.05, paste0(sprintf("%.1f", percent*100),"%"),"")),
+                position=position_stack(vjust=0.5), colour="white", size=3) +
+  theme(axis.title.x = element_blank()) + 
+  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
+                              "#FACE7F", "#8B6969", "#CDAF95", "#2B4F60",
+                              "#996666", "#BDD2B6", "#798777", "#BDC7C9"))
+
 ```

### 2024-12-28 analysis/paper/paper.qmd
Goals added to main text. Code for Blank/Core ratios for all technocomplexes.

@@
 
+## Aims
+
+Our goal is to understand the provisioning, use and management strategies of chert lithic resources of the hunter-gatherer groups of Vale Boi during the Upper Paleolithic.
+We focus on the technological characteristics of the two broader groups of cherts identified at the site: local and non-local, based on the notion that local raw materials will show technological patterns different from those which are non-local.
+Expanding upon our previous works on chert types present throughout the UP sequence in the Terrace and Shelter area of Vale Boi, and their possible interpretations, and following the previously mentioned studies and models about raw material management and lithic technology organisation focusing mainly on the different management of local and non-local raw materials, associated with mobility and occupation length, we derive two main expectations (fig.
+1):
+
+1.	Local raw materials are expected to present a) low use-lives, with knapping strategies characterised by manufacture, use and discard; b) a decreased reduction intensity and low tools to debitage ratio; c) larger cores and lower blank-to-core ratio; d) less extensive reduction and reworking of tools; e) higher tool typology diversity.
+This may be related to the abundance of local raw materials which leads to less intensive life-use extension, but also to the residential characteristics of the site and low mobility, which increases the reliance on local raw materials.
+
+2.	Non-local raw materials are expected to present a) a majority of formal tools, in the case of individual provisioning or higher use-lives, with knapping strategies characterised by the manufacture, use and maintenance; b) an increased reduction intensity and high tools to debitage ratio; c) smaller cores and higher blank-to-core ratio; d) more extensive reduction and high presence of retouch and/or tool maintenance; e) lower tool diversity, characterised by less specialised tools, such as the presence of retouched blanks as multipurpose tools.
+This may be related to scarcer or poorer quality of raw materials which leads to the prepared transportation of lithic resources to a site, but also related to more mobile settlement patterns as well as short-term site occupations.
+
+Despite these expectations, it is important to notice that, especially in occupations with a high percentage of non-local cherts at the site (e.g., Gravettian occupations of level 7 and 6 of the Terrace at Vale Boi), these technological may not reflect a preoccupation with intensive management of a scarce and limited non-local resource, and be more a reflection of the type of provisioning applied following the suggested model by Kuhn (2004).
+In this sense, distance from source may have had little intrinsic meaning in the past, especially regarding non-local raw materials.
+There are several ways in which raw materials may be transported to a site, and the relationship between distance and cost may vary depending on how this takes place (Kuhn, 2004).
+When non-local raw materials are as abundant as local ones, the difference in the technological organization between these materials may be less obvious (Andrefsky, 2004).
+
 # Methods
@@
                 RETOUCHEDPIECETYOPOLOGY)
+
 gravettian_tech <- gravettian_tech %>% 
@@
     RETOUCHEDPIECETYPOLOGY == "Double backed bladelet" ~ "Backed bladelet",
-    TRUE ~ RETOUCHEDPIECETYPOLOGY))
+    TRUE ~ RETOUCHEDPIECETYPOLOGY)) %>% 
+  mutate(CLASS = case_when(
+    CLASS == "Flake" | CLASS == "ElongBlank" ~ "Blank",
+    CLASS == "FlakeFrag" | CLASS == "ElongBlankFrag" ~ "BlankFrag",
+    TRUE ~ CLASS))
 
@@
 
+```{r}
+#| label: ratios
+
+proto_ratio <- phase_dataset %>% 
+  filter(Level %in% c("4E", "4E/5")) %>% 
+  filter(RMTYPE != "INDET") %>% 
+  mutate(CLASS = case_when(
+    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
+    TRUE ~ CLASS)) %>% 
+  filter(CLASS %in% c("Blank", "Core")) %>% 
+  count(RMTYPE, CLASS, sort = TRUE) %>% 
+  group_by(RMTYPE, CLASS) %>%
+  summarise(n=sum(n)) %>% 
+  spread(CLASS, n, fill=0) %>% 
+  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
+  filter(!sum <= 5) %>% 
+  mutate("Blank/Core" = Blank / Core)
+
+grav_ratio <- phase_dataset %>% 
+  filter(Level %in% c(6, 7)) %>% 
+  filter(RMTYPE != "INDET") %>% 
+  mutate(CLASS = case_when(
+    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
+    TRUE ~ CLASS)) %>% 
+  filter(CLASS %in% c("Blank", "Core")) %>% 
+  count(RMTYPE, CLASS, sort = TRUE) %>% 
+  group_by(RMTYPE, CLASS) %>%
+  summarise(n=sum(n)) %>% 
+  spread(CLASS, n, fill=0) %>% 
+  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
+  filter(!sum <= 5) %>% 
+  mutate("Blank/Core" = Blank / Core)
+
+solu_ratio <- solutrean_dataset %>% 
+  mutate(CLASS = case_when(
+    CLASS == "Flake" | CLASS == "ElongBlank" | CLASS == "Thinning flake" ~ "Blank",
+    CLASS == "FlakeFrag" | CLASS == "ElongBlankFrag" ~ "BlankFrag",
+    TRUE ~ CLASS)) %>% 
+  mutate(CLASS = case_when(
+    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
+    TRUE ~ CLASS)) %>% 
+  filter(RMTYPE != "INDET") %>% 
+  filter(CLASS %in% c("Blank", "Core")) %>% 
+  count(RMTYPE, CLASS, sort = TRUE) %>% 
+  group_by(RMTYPE, CLASS) %>%
+  summarise(n=sum(n)) %>% 
+  spread(CLASS, n, fill=0) %>% 
+  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
+  filter(!sum <= 5) %>% 
+  mutate("Blank/Core" = Blank / Core)
+
+```
+
 Here is an example of inline code `r x` in the middle of a sentence.

### 2024-12-28 analysis/paper/paper.qmd
Tool to debitage ratio. Added codes for Level 5 (bottom).

@@
 ```{r}
-#| label: ratios
+#| label: ratio-blankcore
 
-proto_ratio <- phase_dataset %>% 
+proto_ratiobc <- phase_dataset %>% 
   filter(Level %in% c("4E", "4E/5")) %>% 
@@
 
-grav_ratio <- phase_dataset %>% 
+proto5_ratiobc <- phase_dataset %>% 
+  filter(Level == 5) %>% 
+  filter(RMTYPE != "INDET") %>% 
+  mutate(CLASS = case_when(
+    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
+    TRUE ~ CLASS)) %>% 
+  filter(CLASS %in% c("Blank", "Core")) %>% 
+  count(RMTYPE, CLASS, sort = TRUE) %>% 
+  group_by(RMTYPE, CLASS) %>%
+  summarise(n=sum(n)) %>% 
+  spread(CLASS, n, fill=0) %>% 
+  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
+  filter(!sum <= 5) %>% 
+  mutate("Blank/Core" = Blank / Core)
+
+grav_ratiobc <- phase_dataset %>% 
   filter(Level %in% c(6, 7)) %>% 
@@
 
-solu_ratio <- solutrean_dataset %>% 
+solu_ratiobc <- solutrean_dataset %>% 
   mutate(CLASS = case_when(
@@
   mutate("Blank/Core" = Blank / Core)
+
+```
+
+```{r}
+#| label: ratios-retouchdebitage
+
+proto_ratiotd <- phase_dataset %>% 
+  filter(Level %in% c("4E", "4E/5")) %>% 
+  filter(RMTYPE != "INDET") %>% 
+  mutate(CLASS = case_when(
+    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
+    CLASS == "RetouchedPieceFrag" & PIECECOMPLETENESS == "Proximal" ~ "RetouchedPiece",
+    TRUE ~ CLASS)) %>% 
+  filter(CLASS %in% c("Blank", "RetouchedPiece")) %>% 
+  count(RMTYPE, CLASS, sort = TRUE) %>% 
+  group_by(RMTYPE, CLASS) %>%
+  summarise(n=sum(n)) %>% 
+  spread(CLASS, n, fill=0) %>% 
+  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
+  filter(!sum <= 5) %>% 
+  mutate("Tool/Blank" = RetouchedPiece / Blank)
+
+proto5_ratiotd <- phase_dataset %>% 
+  filter(Level == 5) %>% 
+  filter(RMTYPE != "INDET") %>% 
+  mutate(CLASS = case_when(
+    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
+    CLASS == "RetouchedPieceFrag" & PIECECOMPLETENESS == "Proximal" ~ "RetouchedPiece",
+    TRUE ~ CLASS)) %>% 
+  filter(CLASS %in% c("Blank", "RetouchedPiece")) %>% 
+  count(RMTYPE, CLASS, sort = TRUE) %>% 
+  group_by(RMTYPE, CLASS) %>%
+  summarise(n=sum(n)) %>% 
+  spread(CLASS, n, fill=0) %>% 
+  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
+  filter(!sum <= 5) %>% 
+  mutate("Tool/Blank" = RetouchedPiece / Blank)
+
+grav_ratiotd <- phase_dataset %>% 
+  filter(Level %in% c(6, 7)) %>% 
+  filter(RMTYPE != "INDET") %>% 
+  mutate(CLASS = case_when(
+    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
+    CLASS == "RetouchedPieceFrag" & PIECECOMPLETENESS == "Proximal" ~ "RetouchedPiece",
+    TRUE ~ CLASS)) %>% 
+  filter(CLASS %in% c("Blank", "RetouchedPiece")) %>% 
+  count(RMTYPE, CLASS, sort = TRUE) %>% 
+  group_by(RMTYPE, CLASS) %>%
+  summarise(n=sum(n)) %>% 
+  spread(CLASS, n, fill=0) %>% 
+  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
+  filter(!sum <= 5) %>% 
+  mutate("Tool/Blank" = RetouchedPiece / Blank)
+
+solu_ratiotd <- solutrean_dataset %>% 
+  filter(RMTYPE != "INDET") %>% 
+  mutate(CLASS = case_when(
+    CLASS == "Flake" | CLASS == "ElongBlank" | CLASS == "Thinning flake" ~ "Blank",
+    CLASS == "FlakeFrag" | CLASS == "ElongBlankFrag" ~ "BlankFrag",
+    CLASS == "Retouched piece" ~ "RetouchedPiece",
+    CLASS == "Retouched piece fragment" ~ "RetouchedPieceFrag",
+    TRUE ~ CLASS)) %>% 
+  mutate(CLASS = case_when(
+    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
+    CLASS == "RetouchedPieceFrag" & PIECECOMPLETENESS == "Proximal" ~ "RetouchedPiece",
+    TRUE ~ CLASS)) %>% 
+  filter(CLASS %in% c("Blank", "RetouchedPiece")) %>% 
+  count(RMTYPE, CLASS, sort = TRUE) %>% 
+  group_by(RMTYPE, CLASS) %>%
+  summarise(n=sum(n)) %>% 
+  spread(CLASS, n, fill=0) %>% 
+  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
+  filter(!sum <= 5) %>% 
+  mutate("Tool/Blank" = RetouchedPiece / Blank)
 

### 2024-12-28 analysis/paper/paper.qmd
Retouched table start.

@@
         MAXWIDTH = MaxWidth, LENGTH = Length, COREPLATFORM = CorePlatform, MAINFACECORTEX = MainFaceCortex, 
-        MAINFACESCARCOUNT = MainFaceScarCount, COREABANDON = CoreAbandon, RETOUCHEDPIECETYOPOLOGY = RetouchedPieceTypology) %>% 
-  mutate(RETOUCHEDPIECETYOPOLOGY = case_when(
-    RETOUCHEDPIECETYOPOLOGY == "Right truncation on retouched piece" ~ "Truncation",
-    RETOUCHEDPIECETYOPOLOGY == "Double backed bladelet" | RETOUCHEDPIECETYOPOLOGY == "Backed bladelet parcial" ~ "Backed bladelet",
-    RETOUCHEDPIECETYOPOLOGY == "Dentilhado" ~ "Denticulate",
-    RETOUCHEDPIECETYOPOLOGY == "Burin on truncation" ~ "Burin",
-    TRUE ~ RETOUCHEDPIECETYOPOLOGY)) %>% 
+        MAINFACESCARCOUNT = MainFaceScarCount, COREABANDON = CoreAbandon, RETOUCHEDPIECETYPOLOGY = RetouchedPieceTypology) %>% 
+  mutate(RETOUCHEDPIECETYPOLOGY = case_when(
+    RETOUCHEDPIECETYPOLOGY == "Right truncation on retouched piece" ~ "Truncation",
+    RETOUCHEDPIECETYPOLOGY == "Double backed bladelet" | RETOUCHEDPIECETYPOLOGY == "Backed bladelet parcial" ~ "Backed bladelet",
+    RETOUCHEDPIECETYPOLOGY == "Dentilhado" ~ "Denticulate",
+    RETOUCHEDPIECETYPOLOGY == "Burin on truncation" ~ "Burin",
+    TRUE ~ RETOUCHEDPIECETYPOLOGY)) %>% 
   dplyr::select(ID, CLASS, COREPREPPROD, PIECECOMPLETENESS, CORTEX, CORTEXLOCATION, PLATFORMTYPE, PLATFORMCORTEX,
-                SCARCOUNT, THICKNESS, MAXWIDTH, LENGTH, COREPLATFORM, MAINFACECORTEX, MAINFACESCARCOUNT, COREABANDON,
-                RETOUCHEDPIECETYOPOLOGY)
+                SCARCOUNT, THICKNESS, MAXWIDTH, LENGTH, COREPLATFORM, MAINFACECORTEX, MAINFACESCARCOUNT, COREABANDON, RETOUCHEDPIECETYPOLOGY)
 
@@
 
+```{r}
+#| label: tool-diversity
+
+proto_retouch <- phase_dataset %>% 
+  filter(CLASS %in% c("RetouchedPiece", "RetouchedPieceFrag")) %>%
+  filter(Level %in% c("4E", "4E/5")) %>% 
+  select(RMTYPE, RETOUCHEDPIECETYPOLOGY) %>% 
+  dplyr::group_by(RMTYPE)
+
+proto_retouch_tab <- table(proto_retouch$RETOUCHEDPIECETYPOLOGY, proto_retouch$RMTYPE)
+proto_retouch_tab <- as.data.frame.matrix(proto_retouch_tab)
+
+```
+
 Here is an example of inline code `r x` in the middle of a sentence.

### 2024-12-30 analysis/paper/paper.qmd
Graphs

@@
 library(stringr)
+library(ggplot2)
 
@@
 ```{r}
+#| label: general-tables
+
+library(janitor)
+
+proto_general <- phase_dataset %>% 
+  filter(Level %in% c("4E", "4E/5"))
+  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
+                                        "TYPE 5", "TYPE 6","TYPE 7", 
+                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL", "INDET"))) %>%
+  tabyl(RMTYPE, CLASS) %>%
+  adorn_totals(c("row", "col")) %>%
+  adorn_percentages("col") %>%
+  adorn_pct_formatting(2) %>%
+  adorn_ns("front")
+
+proto_graph <- proto_general %>% 
+  dplyr::group_by(RMTYPE, CLASS) %>%
+  dplyr::summarise(n = n())
+  #mutate(percent = n/sum(n))
+
+proto_graph <- ggplot(proto_graph, aes(x=RMTYPE,y=n, fill = CLASS)) +
+  geom_bar(stat = "identity", width = 0.5) +
+  #scale_y_continuous(labels = scales::percent_format()) +
+  #theme_minimal() + #type of theme
+  labs(y = "Count %") +# change labels
+  guides(fill=guide_legend(title=NULL)) + # remove legend title
+  geom_text() +
+  #geom_text(aes(label=ifelse(percent >= 0.05, paste0(sprintf("%.1f", percent*100),"%"),"")),
+  #          position=position_stack(vjust=0.5), colour="white", size=3) +
+  theme(axis.title.x = element_blank()) +
+  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
+                               "#FACE7F", "#845460", "#EAD3CB", "#BDC7C9",
+                               "#2B4F60", "#BDD2B6", "#798777", "#FCF0C8"))
+
+```
+
+```{r}
 
@@
                               "#996666", "#BDD2B6", "#798777", "#BDC7C9"))
+
+```
+
+```{r}
+#| label: stats
+
+proto_stats_blanks <- phase_dataset %>% 
+  filter(Level %in% c("4E", "4E/5")) %>% 
+  filter(RMTYPE != "INDET") %>% 
+  mutate(CLASS = case_when(
+    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "BlankFragProx", # To add proximal fragments to ratio
+    TRUE ~ CLASS)) %>% 
+  filter(CLASS %in% c("Blank", "Core", "BlankFragProx", "RetouchedPiece")) %>%
+  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
+  group_by(RMTYPE, CLASS) %>% 
+  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
+
+proto_stats_blanks <- phase_dataset %>% 
+  filter(Level %in% c("4E", "4E/5")) %>% 
+  filter(RMTYPE != "INDET") %>% 
+  mutate(CLASS = case_when(
+    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "BlankFragProx", # To add proximal fragments to ratio
+    TRUE ~ CLASS)) %>% 
+  filter(CLASS %in% c("Blank", "Core", "BlankFragProx", "RetouchedPiece")) %>%
+  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
+  group_by(RMTYPE, CLASS) %>% 
+  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
 

### 2025-01-04 analysis/paper/paper.qmd
Retouch diversity tables

@@
 
+# Solutrean table
+solu_retouch <- solutrean_dataset %>%
+    mutate(CLASS = case_when(
+    CLASS == "Retouched piece" ~ "RetouchedPiece",
+    CLASS == "Retouched piece fragment" ~ "RetouchedPieceFrag",
+    TRUE ~ CLASS)) %>% 
+  filter(CLASS %in% c("RetouchedPiece", "RetouchedPieceFrag")) %>% 
+  select(RMTYPE, RETOUCHEDPIECETYPOLOGY) %>% 
+  group_by(RMTYPE)
+
+solu_retouch_tab <- solu_retouch %>% 
+  tabyl(RETOUCHEDPIECETYPOLOGY, RMTYPE) %>% 
+  adorn_totals(c("row", "col")) %>% 
+  adorn_percentages("col") %>% 
+  adorn_pct_formatting(2) %>% 
+  adorn_ns("front")
+
+# Proto-solutrean table
 proto_retouch <- phase_dataset %>% 
@@
 
-proto_retouch_tab <- table(proto_retouch$RETOUCHEDPIECETYPOLOGY, proto_retouch$RMTYPE)
-proto_retouch_tab <- as.data.frame.matrix(proto_retouch_tab)
+proto_retouch_tab <- proto_retouch %>%
+  tabyl(RETOUCHEDPIECETYPOLOGY, RMTYPE) %>%
+  adorn_totals(c("row", "col")) %>% 
+  adorn_percentages("col") %>% 
+  adorn_pct_formatting(2) %>% 
+  adorn_ns("front")
+
+# Gravettian table
+grav_retouch <- phase_dataset %>% 
+  filter(CLASS %in% c("RetouchedPiece", "RetouchedPieceFrag")) %>%
+  filter(Level %in% c(6, 7)) %>% 
+  select(RMTYPE, RETOUCHEDPIECETYPOLOGY) %>% 
+  dplyr::group_by(RMTYPE)
+
+grav_retouched_tab <- grav_retouch %>% 
+  tabyl(RETOUCHEDPIECETYPOLOGY, RMTYPE) %>%
+  adorn_totals(c("row", "col")) %>% 
+  adorn_percentages("col") %>% 
+  adorn_pct_formatting(2) %>% 
+  adorn_ns("front")
 

### 2025-01-04 analysis/paper/paper.qmd
Finishing up results. Retouched tool variability analysis of tables.

@@
 
+Local raw materials are expected to present a) low use-lives, with knapping strategies characterised by manufacture, use and discard; b) a decreased reduction intensity and low tools to debitage ratio; c) larger cores and lower blank-to-core ratio; d) less extensive reduction and reworking of tools; e) higher tool typology diversity.
+This may be related to the abundance of local raw materials which leads to less intensive life-use extension, but also to the residential characteristics of the site and low mobility, which increases the reliance on local raw materials.
+
 <!-- Here's some example analysis code: -->
@@
 library(ggplot2)
+library(janitor)
 
@@
 
-library(janitor)
-
 proto_general <- phase_dataset %>% 
@@
 
+# Solutrean statistics
+solu_stats <- solutrean_dataset %>% 
+  filter(RMTYPE != "INDET") %>%
+   mutate(CLASS = case_when(
+    CLASS == "Thinning flake" ~ "ThinningFlake",
+    CLASS == "ElongBlank" ~ "ElongatedProd",
+    CLASS == "FlakeFrag" | CLASS == "ElongBlankFrag" ~ "BlankFrag",
+    TRUE ~ CLASS)) %>% 
+  mutate(CLASS = case_when(
+    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "BlankFragProx", # To add proximal fragments to ratio
+    TRUE ~ CLASS)) %>%
+  filter(CLASS %in% c("Flake", "ElongatedProd", "ThinningFlake", "Core", "BlankFragProx", "RetouchedPiece")) %>%
+  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
+  group_by(RMTYPE, CLASS) %>% 
+  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
+
+# Proto-solutrean statistics
 proto_stats_blanks <- phase_dataset %>% 
@@
     TRUE ~ CLASS)) %>%
-  filter(CLASS %in% c("Blank", "Core", "BlankFragProx", "RetouchedPiece")) %>%
+  mutate(elong = MAXWIDTH*2-LENGTH) %>%
+  mutate(CLASS = case_when(
+    CLASS == "Blank" & elong > 0 ~ "Flake",
+    CLASS == "Blank" & elong <= 0 ~ "ElongatedProd",
+    TRUE ~ CLASS)) %>% 
+  filter(CLASS %in% c("Flake", "ElongatedProd", "Core", "BlankFragProx", "RetouchedPiece")) %>%
   select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
@@
 
-proto_stats_blanks <- phase_dataset %>% 
-  filter(Level %in% c("4E", "4E/5")) %>% 
+# Gravettian statistics
+grav_stats_blanks <- phase_dataset %>%
+  filter(Level %in% c(6, 7)) %>%
   filter(RMTYPE != "INDET") %>%
@@
 
+## Tool typology diversity
+
+Solutrean has a total of 43 retouched tools (complete and fragments), with a high variety of retouched tools (11), ranging from 5 (11.6%) to 4 (9.3%) per tool type.
+The typologies which are most frequently present in the studied sample are bifacial blanks and bifacial fragments, amounting to a total of 23.2% of products related to bifacial tool/point production.
+Two types of points were identified: shouldered points (2 complete and 1 fragment, amounting to a total of 6.9%) and tanged and 4 winged points (9.3%).
+Other frequent retouched typologies are end scrappers (9.3%), notches (11.6%), retouched blanks (~14%) and splintered pieces (11.6%).
+Regarding chert type, a substantial portion of retouched tools are of unidentified chert types (n=13), mostly due to alterations (fire and post depositional).
+Most retouched tools with identified chert are local (n=13, accounting to `r 13*100/43`%), followed by Type 6 cherts, previously suggested to be from Central Portugal (n=7, accounting to `r 7*100/43`%).
+Local types show a higher variability retouched typologies (9 types), although this may be related to the higher amount of tools.
+In comparison, Type 6 shows only 5 different typologies.
+
+For the Proto-Solutrean, 65 chert retouched tools were identified, with `r 15*100/65`% (n=15) having an unidentified chert type.
+A total of `r 1+10+2+8+5` retouched tools correspond to local cherts, while `r 10+9+3+1` correspond to non-local cherts.
+Regarding the typology, 14 different types of retouched tools were identified.
+Endscrappers (26.2%), splintered pieces (17%), burins (12.3%) and retouched flakes (7.7%) were the most prevalent.
+In comparison to the high frequency of point-related technology observed during the Solutrean, Vale Comprido points are barely present in chalcedony and chert (n=1), as they are mostly present in dolerite and chert was mainly used for other tools as seen in previous studies (**).
+
 ```{r}
@@
   select(RMTYPE, RETOUCHEDPIECETYPOLOGY) %>%
+  filter(RETOUCHEDPIECETYPOLOGY != "") %>% 
   group_by(RMTYPE)

### 2025-01-07 analysis/paper/paper.qmd
Results description and alterations, all sections except general table.

@@
 
+## Measurement and weight descriptive statistics
+
+In the Gravettian assemblages, blanks and retouched tools show little differences in the measurements of artefacts of local and non-local chert types.
+The only exception are T6 and T7 blanks, which show slightly lower weight means and lower maximum  length (~35 mm) than the other types.
+For cores, T2 shows smaller (lighter) values than T6 (means of 21.3 and 12.6, and maximum weight of 37.2 gr and 22.5 gr, respectively), even if they seem to vary in morphology, with little patterns regarding thickness, width and length.
+
+In the Proto-Solutrean assemblage, blanks show lower values for length and weight in non-local cherts (T6-T8), varying between a mean of ~19-20 mm and maximum of <30 mm for length, and mean of <3 gr and maximum of <7 gr for weight.
+In retouched tools, weight also shows smaller means for non-local cherts (~3-2 gr), although with higher standard deviation (~1.6).
+Regarding cores, the sample size for non-local cherts is small and only T7 statistics could be obtained.
+Despite this, when compared to local cherts (T2-T5), T7 cores show smaller mean and maximum values (7.3 gr and 8 gr, respectively).
+
+For the Solutrean assemblage, there seems to be no clear difference in the measurements and weights of artefacts between local and non-local cherts.
+This is true especially for the most represented classes (flakes and elongated blanks).
+Regarding cores, many of the local and non-local cherts are not available for the statistics due to missing data.
+However, the smallest weight mean seem to be for cores of T6 (16.5 gr), although with high standard deviation and maximum value (16.5 and 48.6 gr, respectively).
+
 ```{r}
@@
     CLASS == "FlakeFrag" | CLASS == "ElongBlankFrag" ~ "BlankFrag",
+    CLASS == "Retouched piece" ~ "RetouchedPiece",
     TRUE ~ CLASS)) %>% 
@@
     CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "BlankFragProx", # To add proximal fragments to ratio
-    TRUE ~ CLASS)) %>%
-  filter(CLASS %in% c("Flake", "ElongatedProd", "ThinningFlake", "Core", "BlankFragProx", "RetouchedPiece")) %>%
+    TRUE ~ CLASS))
+# For flakes (complete)
+solu_stats_blanks <- solu_stats %>% 
+  filter(CLASS %in% c("Flake", "ElongatedProd")) %>%
+  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
+  group_by(RMTYPE, CLASS) %>% 
+  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
+# For Thinning flakes
+solu_stats_thin <- solu_stats %>% 
+  filter(CLASS == "ThinningFlake") %>%
+  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
+  group_by(RMTYPE, CLASS) %>% 
+  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
+# For cores
+solu_stats_cores <- solu_stats %>% 
+  filter(CLASS == "Core") %>%
+  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
+  group_by(RMTYPE, CLASS) %>% 
+  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
+# For retouched
+solu_stats_retouched <- solu_stats %>% 
+  filter(CLASS == "RetouchedPiece") %>%
   select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
@@
 
-# Proto-solutrean statistics
-proto_stats_blanks <- phase_dataset %>% 
+# Proto-solutrean statistics ds prep
+proto_stats <- phase_dataset %>% 
   filter(Level %in% c("4E", "4E/5")) %>% 
@@
     CLASS == "Blank" & elong <= 0 ~ "ElongatedProd",
-    TRUE ~ CLASS)) %>% 
-  filter(CLASS %in% c("Flake", "ElongatedProd", "Core", "BlankFragProx", "RetouchedPiece")) %>%
+    TRUE ~ CLASS))
+
+# For blanks (complete)
+proto_stats_blanks <- proto_stats %>% 
+  filter(CLASS == "Flake") %>%
   select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
@@
 
-# Gravettian statistics
-grav_stats_blanks <- phase_dataset %>%
+# For cores
+proto_stats_cores <- proto_stats %>% 
+  filter(CLASS == "Core") %>%
+  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
+  group_by(RMTYPE, CLASS) %>% 
+  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
+
+# For retouched
+proto_stats_retouched <- proto_stats %>% 
+  filter(CLASS == "RetouchedPiece") %>%
+  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
+  group_by(RMTYPE, CLASS) %>% 
+  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
+
+# Gravettian statistics ds prep
+grav_stats <- phase_dataset %>%
   filter(Level %in% c(6, 7)) %>%
@@
     CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "BlankFragProx", # To add proximal fragments to ratio
-    TRUE ~ CLASS)) %>%
-  filter(CLASS %in% c("Blank", "Core", "BlankFragProx", "RetouchedPiece")) %>%
+    TRUE ~ CLASS))
+
+# For blanks (complete)
+grav_stats_blanks <- grav_stats %>% 
+  filter(CLASS == "Blank") %>%
+  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
+  group_by(RMTYPE, CLASS) %>% 
+  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
+
+# For cores
+grav_stats_cores <- grav_stats %>% 
+  filter(CLASS == "Core") %>%
+  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
+  group_by(RMTYPE, CLASS) %>% 
+  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
+
+# For retouched
+grav_stats_retouch <- grav_stats %>% 
+  filter(CLASS == "RetouchedPiece") %>%
   select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
@@
 ```
+
+## Blank to core ratios
+
+The analysis shows different patterns in the blank to core ratios not only between chert types, but also between assemblages.
+For the Gravettian, local types (T1 and T2) show lower ratios (<4) of blanks to cores, although other local types (T3-T5) have no identified cores.
+The non-local types, especially T6 and T7 show higher ratios (~8) of blanks to cores.
+In the Proto-Solutrean occupation, T2 and T4 show the lowest ratios (8.7 and 6.2 respectively) of blanks per cores, while the non-local T7 shows the highest ratios (11.0) of blanks to cores.
+Unlike the Gravettian, T1 (local chalcedony) and T6 (non-local chert) have no cores.
+Finally, the Solutrean assemblage shows similar patterns, with the local cherts showing lower blank to core ratios (<7), while the non-local cherts show higher ratios (>9).
+Similar to the Proto-Solutrean, chalcedony cores were also not identified in the studied sample.
 
@@
 ```
+
+## Retouched tools to debitage ratios
+
+Throughout the different assemblages, the difference in retouched tools to debitage ratios by chert types, especially between local and non-local types, seem to show similar patterns.
+In this sense, in general, non-local types seem to consistently show the higher ratios when compared to local cherts.
+In the Gravettian assemblages this pattern is the most accentuated since T1 to T5 show low ratios (0.25-0.5) of retouched to debitage, while TL and T6-T8 show higher ratios (0.9-1.0).
+The exception is T10 (non-local type), which shows a lower ratio (0.28), more akin to local cherts.
+In the Proto-Solutrean assemblage, the lowest ratio of retouched to debitage is seen in T1 and T4, with values below 0.1, while other local types show slightly higher values (0.1-0.2).
+Following the same trend as the Gravettian, the non-local cherts show slightly higher ratios (0.3-0.5) of retouched to debitage.
+In the Solutrean assemblage, all local cherts (T2-T5) show low retouch to debitage ratios (<0.1), with the exception of T1 (chalcedony), which shows a ratio of 1.8.
+Other non-local cherts (T6-T7 and T11) show slightly higher retouch to debitage ratios (0.1-0.3).
 
@@
 In comparison to the high frequency of point-related technology observed during the Solutrean, Vale Comprido points are barely present in chalcedony and chert (n=1), as they are mostly present in dolerite and chert was mainly used for other tools as seen in previous studies (**).
+Regarding the types of chalcedony and chert, local raw materials show 9 different typologies, with Vale Comprido points, Solutrean retouch and notches being present only in the local types, even if in small numbers.
+Non-local cherts show 10 different typologies of retouched tools.
+Perforator-endscrapper, truncation, backed bladelet and retouched blades typologies can only be found in non-local varieties, albeit also in small numbers.
+
+In the Gravettian assemblage, 170 retouched tools were identified, with 22.9% (n=39) having an unidentified chert type.
+From the identified chert types, 24.1% of retouched tools are from local types and 52.9% from non-local cherts, with 34.1% of these being from T6 and 14.1% of T7.
+Regarding typologies, 15 types of retouched tools were identified (excluding fragments).
+All these typologies were found in the non-local chert group, while only 7 retouched tool typologies were identified in local cherts/chalcedony.
+The most common typologies identified in the assemblage were burins (n=52), endscrappers (n=17), retouched flakes (n=30) and splintered pieces (n=26).
+From all these, the majority were produced in non-local chert, mainly T6 and in lesser numbers T7.
+Although in lesser numbers as well, backed bladelets were also identified (n=6), but only present in non-local chert types (T6 and T7).
 
@@
   adorn_totals(c("row", "col")) %>% 
-  adorn_percentages("col") %>% 
+  adorn_percentages("row") %>% 
   adorn_pct_formatting(2) %>% 

### 2025-01-07 analysis/paper/paper.qmd
General tables and general results.

@@
 
-proto_general <- phase_dataset %>% 
-  filter(Level %in% c("4E", "4E/5"))
+solutrean_general <- solutrean_dataset %>%
   mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
@@
                                         "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL", "INDET"))) %>% 
-  tabyl(RMTYPE, CLASS) %>%
+  tabyl(CLASS, RMTYPE) %>%
   adorn_totals(c("row", "col")) %>%
@@
 
-proto_graph <- proto_general %>% 
-  dplyr::group_by(RMTYPE, CLASS) %>%
-  dplyr::summarise(n = n())
-  #mutate(percent = n/sum(n))
+proto_general <- phase_dataset %>% 
+  filter(Level %in% c("4E", "4E/5")) %>% 
+  filter(CLASS != "Chip") %>% 
+  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
+                                        "TYPE 5", "TYPE 6","TYPE 7", 
+                                        "TYPE 8", "TYPE 9", "TYPE 10", "TL", "INDET"))) %>% 
+  tabyl(CLASS, RMTYPE) %>%
+  adorn_totals(c("row", "col")) %>%
+  adorn_percentages("col") %>%
+  adorn_pct_formatting(2) %>%
+  adorn_ns("front")
 
-proto_graph <- ggplot(proto_graph, aes(x=RMTYPE,y=n, fill = CLASS)) +
-  geom_bar(stat = "identity", width = 0.5) +
-  #scale_y_continuous(labels = scales::percent_format()) +
-  #theme_minimal() + #type of theme
-  labs(y = "Count %") +# change labels
-  guides(fill=guide_legend(title=NULL)) + # remove legend title
-  geom_text() +
-  #geom_text(aes(label=ifelse(percent >= 0.05, paste0(sprintf("%.1f", percent*100),"%"),"")),
-  #          position=position_stack(vjust=0.5), colour="white", size=3) +
-  theme(axis.title.x = element_blank()) +
-  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
-                               "#FACE7F", "#845460", "#EAD3CB", "#BDC7C9",
-                               "#2B4F60", "#BDD2B6", "#798777", "#FCF0C8"))
+gravettian_general <- phase_dataset %>% 
+  filter(Level %in% c(6, 7)) %>%
+  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
+                                        "TYPE 5", "TYPE 6","TYPE 7", 
+                                        "TYPE 8", "TYPE 9", "TYPE 10", "TL", "INDET"))) %>% 
+  tabyl(CLASS, RMTYPE) %>%
+  adorn_totals(c("row", "col")) %>%
+  adorn_percentages("col") %>%
+  adorn_pct_formatting(2) %>%
+  adorn_ns("front")
 
@@
 ```
+
+## Assemblage description
+
+For the Gravettian assemblage, local cherts show similar patterns, with T2 and T5 being the types with higher numbers of artefacts (n=110 and n=72, respectively).
+For all local cherts, blanks represent more than 50% of the artefacts.
+Cores and core maintenance products are only present in T2 and T5, although in small frequencies.
+For T2, cores represent 13.6% of the group and core preparation and maintenance products represent 1.8% (n=2).
+For T5 only 4 core fragments and 1 core preparation  and maintenance product are present.
+Retouched tools seem to be especially relevant in T2 (~18%), being the most frequent class after blanks.
+The presence of burin spalls (8.1%) in T2 also hints at the production of burins.
+In comparison, T5 shows similar percentages of retouched tools and shatter (~12% and ~14%, respectively).
+T1 (chalcedony) shows similar patterns, although with higher percentages of shatter (34.3%).
+
+In comparison, non-local chert types (especially T6 with n=169 and T7 with n=86) show similar percentages of blanks (~40-50%) but higher percentages of retouched tools (~30%).
+Cores, core fragments and core preparation and maintenance products are also present even if in small percentages (<4%), as well as shatter (~12%).
+In both types burin spalls are also present (7.1% for T6 and 4.6% for T7).
+The TL cherts show, as expected, blanks and blank fragments (n=3 and n=2, respectively), and retouched tools (n=4).
+The presence of core maintenance products in TL is possibly related to the limited knapping of TL01 (jasper).
+
+TL - blank and retouched, plus a shatter. The latter is unexpected since TL should be expected to be a finished product. Means it is either a byproduct of knapping or not TL?
+T1 (chalcedony) is composed mostly of blanks and a retouched tool (Vale Comprido point).
+The absence of cores but presence of shatter may hint at the knapping of this raw material at the site, but the absence of formal cores perhaps due to the quality of nodules/lenses, which are characterised by fractures and natural flaws.
+The local types show high frequency of blanks (>50%, including blank fragments), followed by shatter (~11-27%), cores (~9-12%) and retouched tools (~5-15% ).
+A single burin spall was identified in T4.
+There is also the absence of core preparation and maintenance products across all local cherts.
+For non-local types, especially those with higher number of artefacts (T6-T8), blanks and blank fragments are the most frequent class (`r 38.3+10`%, `r 40.4+12.7`% and `r 46.1+23.1`%, respectively).
+For T6 the absence of cores but the presence of shatter and a single core preparation and maintenance product suggests knapping on site.
+Similarly, T7 shows the presence of cores and core fragments  (`r 4.2+4.2`%), shatter (17.0%) and a single core preparation and maintenance product.
+In comparison, T8, despite more limited in number (n=13), shows only one core, with the rest of the artefacts being blanks (as previously noted) or retouched tools (23.1% and n=3).
+Aside from the TL artefacts, all non-local cores show at least 1 core and frequently the presence of shatter.
+
+The Solutrean assemblages show similar patterns to those observed in previous assemblages.
+For local cherts, especially T2 (n=506) which accounts for the majority of local chert present in the studied samples, the majority of the debitage is composed of flakes and elongated blanks (`r 42+13.4+17.8+10.7`%).
+Cores are also present (10.5%), alongside a small percentage of core preparation and maintenance products (1.3%).
+In comparison to debitage products, retouched tools are scarce in the sample (~3%).
+The non-local cherts most present in the Solutrean of the Shelter area are T6 (n=86) and T8 (n=30).
+These show similar patterns to the local cherts, with high percentages of blanks (>50%).
+Cores and core preparation and maintenance are also present in small percentages (<10%), as well as retouched tools (<10%).
+However, these specific results are truncated by the lack of shatter, which were not present in the sample used for the study.
+It is important to notice, however, that alike previous results from the Solutrean assemblages of the Terrace area, a large portion of the studied sample was altered and did not allow for a chert type attribution (`r 201*100/914`%).
 

### 2025-01-07 analysis/paper/paper.qmd
Results first draft. Interpretation draft and discussion initial organization.

@@
 # Results
-
-Local raw materials are expected to present a) low use-lives, with knapping strategies characterised by manufacture, use and discard; b) a decreased reduction intensity and low tools to debitage ratio; c) larger cores and lower blank-to-core ratio; d) less extensive reduction and reworking of tools; e) higher tool typology diversity.
-This may be related to the abundance of local raw materials which leads to less intensive life-use extension, but also to the residential characteristics of the site and low mobility, which increases the reliance on local raw materials.
-
-<!-- Here's some example analysis code: -->
 
@@
 T1 (chalcedony) shows similar patterns, although with higher percentages of shatter (34.3%).
-
 In comparison, non-local chert types (especially T6 with n=169 and T7 with n=86) show similar percentages of blanks (~40-50%) but higher percentages of retouched tools (~30%).
@@
 
-TL - blank and retouched, plus a shatter. The latter is unexpected since TL should be expected to be a finished product. Means it is either a byproduct of knapping or not TL?
-T1 (chalcedony) is composed mostly of blanks and a retouched tool (Vale Comprido point).
-The absence of cores but presence of shatter may hint at the knapping of this raw material at the site, but the absence of formal cores perhaps due to the quality of nodules/lenses, which are characterised by fractures and natural flaws.
-The local types show high frequency of blanks (>50%, including blank fragments), followed by shatter (~11-27%), cores (~9-12%) and retouched tools (~5-15% ).
+Regarding the Proto-Solutrean assemblage, the local types show high frequency of blanks (>50%, including blank fragments), followed by shatter (~11-27%), cores (~9-12%) and retouched tools (~5-15% ).
 A single burin spall was identified in T4.
 There is also the absence of core preparation and maintenance products across all local cherts.
+T1 (chalcedony) is composed mostly of blanks and a retouched tool (Vale Comprido point).
+The absence of cores but presence of shatter may hint at the knapping of this raw material at the site, but the absence of formal cores perhaps due to the quality of nodules/lenses, which are characterised by fractures.
 For non-local types, especially those with higher number of artefacts (T6-T8), blanks and blank fragments are the most frequent class (`r 38.3+10`%, `r 40.4+12.7`% and `r 46.1+23.1`%, respectively).
@@
 Aside from the TL artefacts, all non-local cores show at least 1 core and frequently the presence of shatter.
+As expected, the TL artifacts include a blank and a retouched.
+However, a shatter was also identified.
 
@@
 The only exception are T6 and T7 blanks, which show slightly lower weight means and lower maximum length (~35 mm) than the other types.
-For cores, T2 shows smaller (lighter) values than T6 (means of 21.3 and 12.6, and maximum weight of 37.2 gr and 22.5 gr, respectively), even if they seem to vary in morphology, with little patterns regarding thickness, width and length.
+For cores, T2 shows bigger (heavier) values than T6 (means of 21.3 and 12.6, and maximum weight of 37.2 gr and 22.5 gr, respectively), even if they seem to vary in morphology, with little patterns regarding thickness, width and length.
 
@@
   summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
+
+grav_stats %>%
+  filter(CLASS %in% c("Blank", "Core", "RetouchedPiece")) %>% 
+  filter(WEIGHT !="0" & !RMTYPE %in% c("INDET","GB","TL")) %>%
+  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
+                                        "TYPE 5", "TYPE 6","TYPE 7", 
+                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
+  mutate(CLASS = factor(CLASS, levels = c("Blank", "Core", "RetouchedPiece"))) %>%
+  ggplot(aes(RMTYPE, WEIGHT, fill = RMTYPE)) +
+  geom_boxplot(aes(fill=factor(RMTYPE)), outlier.shape=21) +
+  facet_wrap(~CLASS) +
+  stat_summary(fun.y=mean, geom="point", shape=23) +
+  labs(y="Weight (gr)") +
+  coord_cartesian(ylim =  c(0, 60)) +
+  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
+                              "#FACE7F", "#8B6969", "#CDAF95", "#2B4F60",
+                              "#996666", "#BDD2B6", "#798777", "#BDC7C9")) +
+  theme(axis.text.x=element_blank(),
+        axis.ticks.x=element_blank(), axis.title.x = element_blank())
+
+grav_stats %>%
+  filter(CLASS %in% c("Blank", "Core", "RetouchedPiece")) %>% 
+  filter(LENGTH !="0" & !RMTYPE %in% c("INDET","GB","TL")) %>%
+  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
+                                        "TYPE 5", "TYPE 6","TYPE 7", 
+                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
+  mutate(CLASS = factor(CLASS, levels = c("Blank", "Core", "RetouchedPiece"))) %>%
+  ggplot(aes(RMTYPE, LENGTH, fill = RMTYPE)) +
+  geom_boxplot(aes(fill=factor(RMTYPE)), outlier.shape=21) +
+  facet_wrap(~CLASS) +
+  stat_summary(fun.y=mean, geom="point", shape=23) +
+  labs(y="Length (mm)") +
+  #coord_cartesian(ylim =  c(0, 60)) +
+  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
+                              "#FACE7F", "#8B6969", "#CDAF95", "#2B4F60",
+                              "#996666", "#BDD2B6", "#798777", "#BDC7C9")) +
+  theme(axis.text.x=element_blank(),
+        axis.ticks.x=element_blank(), axis.title.x = element_blank())
 
@@
 
-Here is an example of inline code `r x` in the middle of a sentence.
+# Discussion
 
-# Discussion
+Hypotheses to test: a) use lives derived from knapping strategies; b) reduction intensity and tools-to-debitage ratios; c) core size and blank-to-core ratios; d) reduction intensity and tool reworking; e) tool typology diversity and specialization.
+
+To test these hypotheses, the following parameters were evaluated for each assemblage and compared between chert types: P1) frequency of retouched tools; P2) frequency of maintenance products; P3) debitage sizes; P4) tools-to-debitage ratios ; P5) core sizes; P6) blank-to-core ratios; P7) retouched tool sizes; P8) tool diversity; P9) tool specialization.
+The results may be summarised as follows:
+
+Retouched tools are present in higher percentages in the Gravettian assemblages.
+Within the assemblage, retouched tools (P1) are present in higher frequencies in the non-local cherts, as well as in core preparation and maintenance products (P2).
+Debitage sizes (P3) show no differences between local and non-local cherts.
+Tools-to-debitage ratios (P4) are significantly higher in non-local cherts.
+Core sizes (P5) are slightly smaller in T6, a non-local chert, and the blank-to-core ratios (P6) are higher in non-local cherts.
+Alike blanks, retouched sizes (P7) show no relevant differences between local and non-local cherts; however, tool diversity (P8) is higher in non-local cherts, although with high frequency of less specialised tools such as notches and retouched tools, alongside hunting implements such as backed bladelets (P9).
+The high frequency of burins seen in this assemblage may be related to their use as cores, representing a knapping strategy more than the presence of retouch.
+
+The Proto-Solutrean assemblages show a smaller presence of retouched tools than the Gravettian, and within the assemblage, retouched tools (P1) are slightly more frequent in non-local cherts.
+Core preparation and maintenance products (P2) are only present in non-local cherts, albeit in very small numbers (n=2).
+Debitage sizes (P3), core sizes (P5), and retouched tool sizes (P7) are slightly smaller in non-local cherts.
+Regarding the ratios, tool-to-debitage ratios (P4) are higher in non-local cherts, albeit smaller than those seen in the Gravettian assemblage, and blank-to-core ratios (P6) are also higher in non-local cherts.
+Little differences were observed in tool diversity (P8) and specialisation (P9) between local and non-local cherts.
+
+Finally, the Solutrean shows no relevant differences in the frequencies of retouched tools (P1) and core preparation and maintenance products (P2) between local and non-local cherts.
+In general, the retouch frequency is lower than that observed in the previous assemblages.
+Debitage (P3), core (P5), and retouched tool (P7) sizes show no differences between local and non-local cherts.
+Similarly to the Proto-Solutrean, tools-to-debitage ratios (P4) and blank-to-core ratios (P6) are slightly higher in non-local cherts.
+Both tool diversity (P8) and tool specialization (P9) are higher in local cherts.
+
+How the results fit or do not fit within the hypothesis, and possible reasons why.
+
+Focus on the high presence of retouch and burins during the Gravettian.
+
+Fitting the technological data with the suggested interpretations from article 2.
 

### 2025-01-08 analysis/paper/paper.qmd
Width, length and weight boxplots + ggarrange

@@
 library(janitor)
+library(ggpubr)
 
@@
 ```{r}
-#| label: stats
+#| label: stats-setup
 
 # Solutrean statistics
-solu_stats <- solutrean_dataset %>% 
+# DS preparation
+solutrean_stats <- solutrean_dataset %>% 
   filter(RMTYPE != "INDET") %>%
@@
     TRUE ~ CLASS))
+
+# Boxplot for individual measurements
+# Length
+solutrean_stats_length <- solutrean_stats %>%
+  filter(CLASS %in% c("Flake", "ElongatedProd", "Core", "RetouchedPiece")) %>% 
+  filter(LENGTH !="0" & !RMTYPE %in% c("INDET","GB","TL")) %>%
+  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
+                                        "TYPE 5", "TYPE 6","TYPE 7", 
+                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
+  mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) %>%
+  ggplot(aes(RMTYPE, LENGTH, fill = RMTYPE)) +
+  geom_boxplot(aes(fill=factor(RMTYPE)), outlier.shape=21) +
+  facet_wrap(~CLASS) +
+  stat_summary(fun.y=mean, geom="point", shape=23) +
+  labs(y="Length (mm)") +
+  #coord_cartesian(ylim =  c(0, 60)) +
+  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
+                              "#FACE7F", "#8B6969", "#CDAF95", "#2B4F60",
+                              "#996666", "#BDD2B6", "#798777", "#BDC7C9")) +
+  theme(axis.text.x=element_blank(),
+        axis.ticks.x=element_blank(), axis.title.x = element_blank())
+# Weight
+solutrean_stats_weight <- solutrean_stats %>%
+  filter(CLASS %in% c("Flake", "ElongatedProd", "Core", "RetouchedPiece")) %>% 
+  filter(WEIGHT !="0" & !RMTYPE %in% c("INDET","GB","TL")) %>%
+  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
+                                        "TYPE 5", "TYPE 6","TYPE 7", 
+                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
+  mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) %>%
+  ggplot(aes(RMTYPE, WEIGHT, fill = RMTYPE)) +
+  geom_boxplot(aes(fill=factor(RMTYPE)), outlier.shape=21) +
+  facet_wrap(~CLASS) +
+  stat_summary(fun.y=mean, geom="point", shape=23) +
+  labs(y="Weight (gr)") +
+  coord_cartesian(ylim =  c(0, 50)) +
+  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
+                              "#FACE7F", "#8B6969", "#CDAF95", "#2B4F60",
+                              "#996666", "#BDD2B6", "#798777", "#BDC7C9")) +
+  theme(axis.text.x=element_blank(),
+        axis.ticks.x=element_blank(), axis.title.x = element_blank())
+# Width
+solutrean_stats_width <- solutrean_stats %>%
+  filter(CLASS %in% c("Flake", "ElongatedProd", "Core", "RetouchedPiece")) %>% 
+  filter(MAXWIDTH !="0" & !RMTYPE %in% c("INDET","GB","TL")) %>%
+  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
+                                        "TYPE 5", "TYPE 6","TYPE 7", 
+                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
+  mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) %>%
+  ggplot(aes(RMTYPE, MAXWIDTH, fill = RMTYPE)) +
+  geom_boxplot(aes(fill=factor(RMTYPE)), outlier.shape=21) +
+  facet_wrap(~CLASS) +
+  stat_summary(fun.y=mean, geom="point", shape=23) +
+  labs(y="Width (mm)") +
+  #coord_cartesian(ylim =  c(0, 60)) +
+  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
+                              "#FACE7F", "#8B6969", "#CDAF95", "#2B4F60",
+                              "#996666", "#BDD2B6", "#798777", "#BDC7C9")) +
+  theme(axis.text.x=element_blank(),
+        axis.ticks.x=element_blank(), axis.title.x = element_blank())
+
+# Tables with individual class. For supplementary information if necessary
 # For flakes (complete)
-solu_stats_blanks <- solu_stats %>% 
+solutrean_stats_blanks <- solutrean_stats %>% 
   filter(CLASS %in% c("Flake", "ElongatedProd")) %>%
@@
   summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
-# For Thinning flakes
-solu_stats_thin <- solu_stats %>% 
-  filter(CLASS == "ThinningFlake") %>%
-  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
-  group_by(RMTYPE, CLASS) %>% 
-  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
 # For cores
-solu_stats_cores <- solu_stats %>% 
+solutrean_stats_cores <- solutrean_stats %>% 
   filter(CLASS == "Core") %>%
@@
 # For retouched
-solu_stats_retouched <- solu_stats %>% 
+solutrean_stats_retouched <- solutrean_stats %>% 
   filter(CLASS == "RetouchedPiece") %>%
@@
 
-# Proto-solutrean statistics ds prep
+# Proto-Solutrean statistics
+# DS prep
 proto_stats <- phase_dataset %>% 
@@
   mutate(CLASS = case_when(
-    CLASS == "Blank" & elong > 0 ~ "Flake",
+    CLASS == "Blank" & elong > 0 ~ "Flake", # QUESTION: do I keep flakes and elong separate or join into blanks?
     CLASS == "Blank" & elong <= 0 ~ "ElongatedProd",
@@
 
-# For blanks (complete)
-proto_stats_blanks <- proto_stats %>% 
+# Boxplots
+# Width
+proto_stats_width <- proto_stats %>%
+  filter(CLASS %in% c("Flake","ElongatedProd", "Core", "RetouchedPiece")) %>% 
+  filter(MAXWIDTH !="0" & !RMTYPE %in% c("INDET","GB","TL")) %>%
+  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
+                                        "TYPE 5", "TYPE 6","TYPE 7", 
+                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
+  mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) %>%
+  ggplot(aes(RMTYPE, MAXWIDTH, fill = RMTYPE)) +
+  geom_boxplot(aes(fill=factor(RMTYPE)), outlier.shape=21) +
+  facet_wrap(~CLASS) +
+  stat_summary(fun.y=mean, geom="point", shape=23) +
+  labs(y="Width (mm)") +
+  coord_cartesian(ylim =  c(0, 50)) +
+  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
+                              "#FACE7F", "#8B6969", "#CDAF95", "#2B4F60",
+                              "#996666", "#BDD2B6", "#798777", "#BDC7C9")) +
+  theme(axis.text.x=element_blank(),
+        axis.ticks.x=element_blank(), axis.title.x = element_blank())
+# Length
+proto_stats_length <- proto_stats %>%
+  filter(CLASS %in% c("Flake","ElongatedProd", "Core", "RetouchedPiece")) %>% 
+  filter(LENGTH !="0" & !RMTYPE %in% c("INDET","GB","TL")) %>%
+  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
+                                        "TYPE 5", "TYPE 6","TYPE 7", 
+                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
+  mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) %>%
+  ggplot(aes(RMTYPE, LENGTH, fill = RMTYPE)) +
+  geom_boxplot(aes(fill=factor(RMTYPE)), outlier.shape=21) +
+  facet_wrap(~CLASS) +
+  stat_summary(fun.y=mean, geom="point", shape=23) +
+  labs(y="Length (mm)") +
+  coord_cartesian(ylim =  c(0, 60)) +
+  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
+                              "#FACE7F", "#8B6969", "#CDAF95", "#2B4F60",
+                              "#996666", "#BDD2B6", "#798777", "#BDC7C9")) +
+  theme(axis.text.x=element_blank(),
+        axis.ticks.x=element_blank(), axis.title.x = element_blank())
+# Weight
+proto_stats_weight <- proto_stats %>%
+  filter(CLASS %in% c("Flake","ElongatedProd", "Core", "RetouchedPiece")) %>% 
+  filter(WEIGHT !="0" & !RMTYPE %in% c("INDET","GB","TL")) %>%
+  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
+                                        "TYPE 5", "TYPE 6","TYPE 7", 
+                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
+  mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) %>%
+  ggplot(aes(RMTYPE, WEIGHT, fill = RMTYPE)) +
+  geom_boxplot(aes(fill=factor(RMTYPE)), outlier.shape=21) +
+  facet_wrap(~CLASS) +
+  stat_summary(fun.y=mean, geom="point", shape=23) +
+  labs(y="Weight (gr)") +
+  coord_cartesian(ylim =  c(0, 50)) +
+  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
+                              "#FACE7F", "#8B6969", "#CDAF95", "#2B4F60",
+                              "#996666", "#BDD2B6", "#798777", "#BDC7C9")) +
+  theme(axis.text.x=element_blank(),
+        axis.ticks.x=element_blank(), axis.title.x = element_blank())
+
+# Tables with individual class. For supplementary information if necessary
+# For flakes (complete)
+proto_stats_flakes <- proto_stats %>% 
   filter(CLASS == "Flake") %>%
@@
   summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
-
 # For cores
@@
   summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
-
 # For retouched
@@
 
-# Gravettian statistics ds prep
-grav_stats <- phase_dataset %>%
+# Gravettian statistics
+# DS prep
+gravettian_stats <- phase_dataset %>%
   filter(Level %in% c(6, 7)) %>%
@@
     CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "BlankFragProx", # To add proximal fragments to ratio
+    TRUE ~ CLASS)) %>% 
+  mutate(elong = MAXWIDTH*2-LENGTH) %>%
+  mutate(CLASS = case_when(
+    CLASS == "Blank" & elong > 0 ~ "Flake", # QUESTION: do I keep flakes and elong separate or join into blanks?
+    CLASS == "Blank" & elong <= 0 ~ "ElongatedProd",
     TRUE ~ CLASS))
 
-# For blanks (complete)
-grav_stats_blanks <- grav_stats %>% 
-  filter(CLASS == "Blank") %>%
-  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
-  group_by(RMTYPE, CLASS) %>% 
-  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
-
-# For cores
-grav_stats_cores <- grav_stats %>% 
-  filter(CLASS == "Core") %>%
-  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
-  group_by(RMTYPE, CLASS) %>% 
-  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
-
-# For retouched
-grav_stats_retouch <- grav_stats %>% 
-  filter(CLASS == "RetouchedPiece") %>%
-  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
-  group_by(RMTYPE, CLASS) %>% 
-  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
-
-grav_stats %>%
-  filter(CLASS %in% c("Blank", "Core", "RetouchedPiece")) %>% 
+# Boxplots
+# Width
+gravettian_stats_width <- gravettian_stats %>%
+  filter(CLASS %in% c("Flake","ElongatedProd", "Core", "RetouchedPiece")) %>% 
+  filter(MAXWIDTH !="0" & !RMTYPE %in% c("INDET","GB","TL")) %>%
+  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
+                                        "TYPE 5", "TYPE 6","TYPE 7", 
+                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
+  mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) %>%
+  ggplot(aes(RMTYPE, MAXWIDTH, fill = RMTYPE)) +
+  geom_boxplot(aes(fill=factor(RMTYPE)), outlier.shape=21) +
+  facet_wrap(~CLASS) +
+  stat_summary(fun.y=mean, geom="point", shape=23) +
+  labs(y="Width (mm)") +
+  coord_cartesian(ylim =  c(0, 50)) +
+  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
+                              "#FACE7F", "#8B6969", "#CDAF95", "#2B4F60",
+                              "#996666", "#BDD2B6", "#798777", "#BDC7C9")) +
+  theme(axis.text.x=element_blank(),
+        axis.ticks.x=element_blank(), axis.title.x = element_blank())
+# Length
+gravettian_stats_length <- gravettian_stats %>%
+  filter(CLASS %in% c("Flake","ElongatedProd", "Core", "RetouchedPiece")) %>% 
+  filter(LENGTH !="0" & !RMTYPE %in% c("INDET","GB","TL")) %>%
+  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
+                                        "TYPE 5", "TYPE 6","TYPE 7", 
+                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
+  mutate(CLASS = factor(CLASS, levels = c("Flake","ElongatedProd", "Core", "RetouchedPiece"))) %>%
+  ggplot(aes(RMTYPE, LENGTH, fill = RMTYPE)) +
+  geom_boxplot(aes(fill=factor(RMTYPE)), outlier.shape=21) +
+  facet_wrap(~CLASS) +
+  stat_summary(fun.y=mean, geom="point", shape=23) +
+  labs(y="Length (mm)") +
+  coord_cartesian(ylim =  c(10, 60)) +
+  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
+                              "#FACE7F", "#8B6969", "#CDAF95", "#2B4F60",
+                              "#996666", "#BDD2B6", "#798777", "#BDC7C9")) +
+  theme(axis.text.x=element_blank(),
+        axis.ticks.x=element_blank(), axis.title.x = element_blank())
+# Weight
+gravettian_stats_weight <- gravettian_stats %>%
+  filter(CLASS %in% c("Flake","ElongatedProd", "Core", "RetouchedPiece")) %>% 
   filter(WEIGHT !="0" & !RMTYPE %in% c("INDET","GB","TL")) %>%
@@
                                         "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
-  mutate(CLASS = factor(CLASS, levels = c("Blank", "Core", "RetouchedPiece"))) %>%
+  mutate(CLASS = factor(CLASS, levels = c("Flake","ElongatedProd", "Core", "RetouchedPiece"))) %>%
   ggplot(aes(RMTYPE, WEIGHT, fill = RMTYPE)) +
@@
 
-grav_stats %>%
-  filter(CLASS %in% c("Blank", "Core", "RetouchedPiece")) %>% 
-  filter(LENGTH !="0" & !RMTYPE %in% c("INDET","GB","TL")) %>%
-  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
-                                        "TYPE 5", "TYPE 6","TYPE 7", 
-                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
-  mutate(CLASS = factor(CLASS, levels = c("Blank", "Core", "RetouchedPiece"))) %>%
-  ggplot(aes(RMTYPE, LENGTH, fill = RMTYPE)) +
-  geom_boxplot(aes(fill=factor(RMTYPE)), outlier.shape=21) +
-  facet_wrap(~CLASS) +
-  stat_summary(fun.y=mean, geom="point", shape=23) +
-  labs(y="Length (mm)") +
-  #coord_cartesian(ylim =  c(0, 60)) +
-  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
-                              "#FACE7F", "#8B6969", "#CDAF95", "#2B4F60",
-                              "#996666", "#BDD2B6", "#798777", "#BDC7C9")) +
-  theme(axis.text.x=element_blank(),
-        axis.ticks.x=element_blank(), axis.title.x = element_blank())
+# Tables with individual class. For supplementary information if necessary
+# For flakes (complete)
+gravettian_stats_flakes <- gravettian_stats %>% 
+  filter(CLASS == "Blank") %>%
+  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
+  group_by(RMTYPE, CLASS) %>% 
+  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
+# For cores
+gravettian_stats_cores <- gravettian_stats %>% 
+  filter(CLASS == "Core") %>%
+  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
+  group_by(RMTYPE, CLASS) %>% 
+  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
+# For retouched
+gravettian_stats_retouch <- gravettian_stats %>% 
+  filter(CLASS == "RetouchedPiece") %>%
+  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
+  group_by(RMTYPE, CLASS) %>% 
+  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
+
+```
+
+```{r}
+#| label: fig-solutrean-boxplots
+
+ggpubr::ggarrange(solutrean_stats_width, solutrean_stats_length, solutrean_stats_weight, labels = c("a", "b", "c"), heights = c(5, 5, 5), ncol = 1, common.legend = TRUE, legend = "bottom")
+
+```
+
+```{r}
+#| label:  fig-proto-boxplots
+
+ggpubr::ggarrange(proto_stats_width, proto_stats_length, proto_stats_weight, labels = c("a", "b", "c"), heights = c(5, 5, 5), ncol = 1, common.legend = TRUE, legend = "bottom")
+
+```
+
+```{r}
+#| label: fig-gravettian-boxplots
+
+ggpubr::ggarrange(gravettian_stats_width, gravettian_stats_length, gravettian_stats_weight, labels = c("a", "b", "c"), heights = c(5, 5, 5), ncol = 1, common.legend = TRUE, legend = "bottom")
 

### 2025-01-08 analysis/paper/paper.qmd
Table kables for render.

@@
 library(ggpubr)
+library(knitr)
 
@@
 #| label: get-data
-#| eval: false
 
@@
 ```{r}
+#| label: tbl-general-gravettian
+#| tbl-cap: Number and percentage of technotypological classes by raw material type from the Gravettian assemblage (levels 6 and 7) of the Terrace area of Vale Boi.
+
+kable(gravettian_general)
+
+```
+
+```{r}
+#| label: tbl-general-proto
+#| tbl-cap: Number and percentage of technotypological classes by raw material type from the Proto-Solutrean assemblage (levels 4E and top of level 5) of the Terrace area of Vale Boi.
+
+kable(proto_general)
+
+```
+
+```{r}
+#| label: tbl-general-solutrean
+#| tbl-cap: Number and percentage of technotypological classes by raw material type from the Solutrean assemblage (levels A to C) of the Shelter area of Vale Boi.
+
+kable(solutrean_general)
+
+```
+
+```{r}
+#| label: raw-material-graph
+#| include: false
 
@@
 
-grav_ratiobc <- phase_dataset %>% 
+gravettian_ratiobc <- phase_dataset %>% 
   filter(Level %in% c(6, 7)) %>% 
@@
 
-solu_ratiobc <- solutrean_dataset %>% 
+solutrean_ratiobc <- solutrean_dataset %>% 
   mutate(CLASS = case_when(
@@
   mutate("Blank/Core" = Blank / Core)
+
+```
+
+```{r}
+#| label: tbl-bcratio-gravettian
+#| tbl-cap: Blank to core ratio by raw material type of the Gravettian assemblage (levels 6 and 7) of the Terrace area.
+
+kable(gravettian_ratiobc)
+
+```
+
+```{r}
+#| label: tbl-bcratio-proto
+#| tbl-cap: Blank to core ratio by raw material type of the Proto-Solutrean assemblage (levels 4E and top of level 5) of the Terrace area.
+
+kable(proto_ratiobc)
+
+```
+
+```{r}
+#| label: tbl-bcratio-solutrean
+#| tbl-cap: Blank to core ratio by raw material type of the Solutrean assemblage (levels A to C) of the Shelter area.
+
+kable(solutrean_ratiobc)
 
@@
 
-grav_ratiotd <- phase_dataset %>% 
+gravettian_ratiotd <- phase_dataset %>% 
   filter(Level %in% c(6, 7)) %>% 
@@
 
-solu_ratiotd <- solutrean_dataset %>% 
+solutrean_ratiotd <- solutrean_dataset %>% 
   filter(RMTYPE != "INDET") %>% 
@@
   mutate("Tool/Blank" = RetouchedPiece / Blank)
+
+```
+
+```{r}
+#| label: tbl-tdratio-gravettian
+#| tbl-cap: Tools to debitage ratio by raw material of the Gravettian assemblage (levels 6 and 7) of the Terrace area.
+
+kable(gravettian_ratiotd)
+
+```
+
+```{r}
+#| label: tbl-tdratio-proto
+#| tbl-cap: Tools to debitage ratio by raw material of the Proto-Solutrean assemblage (levels 4E and top of level 5) of the Terrace area.
+
+kable(proto_ratiotd)
+
+```
+
+```{r}
+#| label: tbl-tdratio-solutrean
+#| tbl-cap: Tools to debitage ratio by raw material of the Solutrean assemblage (levels A to C) of the Shelter area.
+
+kable(solutrean_ratiotd)
 
@@
 # Solutrean table
-solu_retouch <- solutrean_dataset %>%
+solutrean_retouch <- solutrean_dataset %>%
     mutate(CLASS = case_when(
@@
 
-solu_retouch_tab <- solu_retouch %>% 
+solutrean_retouch_tab <- solutrean_retouch %>% 
   tabyl(RETOUCHEDPIECETYPOLOGY, RMTYPE) %>% 
@@
 # Gravettian table
-grav_retouch <- phase_dataset %>% 
+gravettian_retouch <- phase_dataset %>% 
   filter(CLASS %in% c("RetouchedPiece", "RetouchedPieceFrag")) %>%
@@
 
-grav_retouched_tab <- grav_retouch %>% 
+gravettian_retouched_tab <- gravettian_retouch %>% 
   tabyl(RETOUCHEDPIECETYPOLOGY, RMTYPE) %>%
@@
   adorn_ns("front")
+
+```
+
+```{r}
+#| label: tbl-retouch-gravettian
+#| tbl-cap: Number and percentage of retouched tools by raw material of the Gravettian assemblage (levels 6 and 7) of the Terrace area.
+
+kable(gravettian_retouched_tab)
+
+```
+
+```{r}
+#| label: tbl-retouch-proto
+#| tbl-cap: Number and percentage of retouched tools by raw material of the Proto-Solutrean assemblage (levels 4E and top of level 5) of the Terrace area.
+
+kable(proto_retouch_tab)
+
+```
+
+```{r}
+#| label: tbl-retouch-solutrean
+#| tbl-cap: Number and percentage of retouched tools by raw material of the Solutrean assemblage (level A to C) of the Shelter area.
+
+kable(solutrean_retouch_tab)
 

### 2025-01-08 analysis/paper/paper.qmd
Fig-caps, updated word and Legion corrected commit with previous commit stashed.

@@
 #| label: fig-solutrean-boxplots
+#| fig-cap: Boxplots Gravettian.
 
@@
 #| label:  fig-proto-boxplots
+#| fig-cap: Boxplots Proto-Solutrean
 
@@
 #| label: fig-gravettian-boxplots
+#| fig-cap: Boxplots Solutrean
 
@@
 ```
-
-The current Git commit details are:
-
-```{r}
-# what commit is this file at? 
-if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
-```

### 2025-01-08 analysis/paper/paper.qmd
Testing zotero integration. All good.

@@
 
-1.	Local raw materials are expected to present a) low use-lives, with knapping strategies characterised by manufacture, use and discard; b) a decreased reduction intensity and low tools to debitage ratio; c) larger cores and lower blank-to-core ratio; d) less extensive reduction and reworking of tools; e) higher tool typology diversity.
+1.
+Local raw materials are expected to present a) low use-lives, with knapping strategies characterised by manufacture, use and discard; b) a decreased reduction intensity and low tools to debitage ratio; c) larger cores and lower blank-to-core ratio; d) less extensive reduction and reworking of tools; e) higher tool typology diversity.
 This may be related to the abundance of local raw materials which leads to less intensive life-use extension, but also to the residential characteristics of the site and low mobility, which increases the reliance on local raw materials.
 
-2.	Non-local raw materials are expected to present a) a majority of formal tools, in the case of individual provisioning or higher use-lives, with knapping strategies characterised by the manufacture, use and maintenance; b) an increased reduction intensity and high tools to debitage ratio; c) smaller cores and higher blank-to-core ratio; d) more extensive reduction and high presence of retouch and/or tool maintenance; e) lower tool diversity, characterised by less specialised tools, such as the presence of retouched blanks as multipurpose tools.
+2.
+Non-local raw materials are expected to present a) a majority of formal tools, in the case of individual provisioning or higher use-lives, with knapping strategies characterised by the manufacture, use and maintenance; b) an increased reduction intensity and high tools to debitage ratio; c) smaller cores and higher blank-to-core ratio; d) more extensive reduction and high presence of retouch and/or tool maintenance; e) lower tool diversity, characterised by less specialised tools, such as the presence of retouched blanks as multipurpose tools.
 This may be related to scarcer or poorer quality of raw materials which leads to the prepared transportation of lithic resources to a site, but also related to more mobile settlement patterns as well as short-term site occupations.
@@
 In this sense, distance from source may have had little intrinsic meaning in the past, especially regarding non-local raw materials.
-There are several ways in which raw materials may be transported to a site, and the relationship between distance and cost may vary depending on how this takes place (Kuhn, 2004).
+There are several ways in which raw materials may be transported to a site, and the relationship between distance and cost may vary depending on how this takes place [@kuhn2004].
 When non-local raw materials are as abundant as local ones, the difference in the technological organization between these materials may be less obvious (Andrefsky, 2004).
@@
 For T5 only 4 core fragments and 1 core preparation and maintenance product are present.
-Retouched tools seem to be especially relevant in T2 (~18%), being the most frequent class after blanks.
+Retouched tools seem to be especially relevant in T2 (\~18%), being the most frequent class after blanks.
 The presence of burin spalls (8.1%) in T2 also hints at the production of burins.
-In comparison, T5 shows similar percentages of retouched tools and shatter (~12% and ~14%, respectively).
+In comparison, T5 shows similar percentages of retouched tools and shatter (\~12% and \~14%, respectively).
 T1 (chalcedony) shows similar patterns, although with higher percentages of shatter (34.3%).
-In comparison, non-local chert types (especially T6 with n=169 and T7 with n=86) show similar percentages of blanks (~40-50%) but higher percentages of retouched tools (~30%).
-Cores, core fragments and core preparation and maintenance products are also present even if in small percentages (<4%), as well as shatter (~12%).
+In comparison, non-local chert types (especially T6 with n=169 and T7 with n=86) show similar percentages of blanks (\~40-50%) but higher percentages of retouched tools (\~30%).
+Cores, core fragments and core preparation and maintenance products are also present even if in small percentages (\<4%), as well as shatter (\~12%).
 In both types burin spalls are also present (7.1% for T6 and 4.6% for T7).
@@
 
-Regarding the Proto-Solutrean assemblage, the local types show high frequency of blanks (>50%, including blank fragments), followed by shatter (~11-27%), cores (~9-12%) and retouched tools (~5-15% ).
+Regarding the Proto-Solutrean assemblage, the local types show high frequency of blanks (\>50%, including blank fragments), followed by shatter (\~11-27%), cores (\~9-12%) and retouched tools (\~5-15% ).
 A single burin spall was identified in T4.
@@
 Cores are also present (10.5%), alongside a small percentage of core preparation and maintenance products (1.3%).
-In comparison to debitage products, retouched tools are scarce in the sample (~3%).
+In comparison to debitage products, retouched tools are scarce in the sample (\~3%).
 The non-local cherts most present in the Solutrean of the Shelter area are T6 (n=86) and T8 (n=30).
-These show similar patterns to the local cherts, with high percentages of blanks (>50%).
-Cores and core preparation and maintenance are also present in small percentages (<10%), as well as retouched tools (<10%).
+These show similar patterns to the local cherts, with high percentages of blanks (\>50%).
+Cores and core preparation and maintenance are also present in small percentages (\<10%), as well as retouched tools (\<10%).
 However, these specific results are truncated by the lack of shatter, which were not present in the sample used for the study.
@@
 In the Gravettian assemblages, blanks and retouched tools show little differences in the measurements of artefacts of local and non-local chert types.
-The only exception are T6 and T7 blanks, which show slightly lower weight means and lower maximum length (~35 mm) than the other types.
+The only exception are T6 and T7 blanks, which show slightly lower weight means and lower maximum length (\~35 mm) than the other types.
 For cores, T2 shows bigger (heavier) values than T6 (means of 21.3 and 12.6, and maximum weight of 37.2 gr and 22.5 gr, respectively), even if they seem to vary in morphology, with little patterns regarding thickness, width and length.
 
-In the Proto-Solutrean assemblage, blanks show lower values for length and weight in non-local cherts (T6-T8), varying between a mean of ~19-20 mm and maximum of <30 mm for length, and mean of <3 gr and maximum of <7 gr for weight.
-In retouched tools, weight also shows smaller means for non-local cherts (~3-2 gr), although with higher standard deviation (~1.6).
+In the Proto-Solutrean assemblage, blanks show lower values for length and weight in non-local cherts (T6-T8), varying between a mean of \~19-20 mm and maximum of \<30 mm for length, and mean of \<3 gr and maximum of \<7 gr for weight.
+In retouched tools, weight also shows smaller means for non-local cherts (\~3-2 gr), although with higher standard deviation (\~1.6).
 Regarding cores, the sample size for non-local cherts is small and only T7 statistics could be obtained.
@@
 The analysis shows different patterns in the blank to core ratios not only between chert types, but also between assemblages.
-For the Gravettian, local types (T1 and T2) show lower ratios (<4) of blanks to cores, although other local types (T3-T5) have no identified cores.
-The non-local types, especially T6 and T7 show higher ratios (~8) of blanks to cores.
+For the Gravettian, local types (T1 and T2) show lower ratios (\<4) of blanks to cores, although other local types (T3-T5) have no identified cores.
+The non-local types, especially T6 and T7 show higher ratios (\~8) of blanks to cores.
 In the Proto-Solutrean occupation, T2 and T4 show the lowest ratios (8.7 and 6.2 respectively) of blanks per cores, while the non-local T7 shows the highest ratios (11.0) of blanks to cores.
 Unlike the Gravettian, T1 (local chalcedony) and T6 (non-local chert) have no cores.
-Finally, the Solutrean assemblage shows similar patterns, with the local cherts showing lower blank to core ratios (<7), while the non-local cherts show higher ratios (>9).
+Finally, the Solutrean assemblage shows similar patterns, with the local cherts showing lower blank to core ratios (\<7), while the non-local cherts show higher ratios (\>9).
 Similar to the Proto-Solutrean, chalcedony cores were also not identified in the studied sample.
@@
 Following the same trend as the Gravettian, the non-local cherts show slightly higher ratios (0.3-0.5) of retouched to debitage.
-In the Solutrean assemblage, all local cherts (T2-T5) show low retouch to debitage ratios (<0.1), with the exception of T1 (chalcedony), which shows a ratio of 1.8.
+In the Solutrean assemblage, all local cherts (T2-T5) show low retouch to debitage ratios (\<0.1), with the exception of T1 (chalcedony), which shows a ratio of 1.8.
 Other non-local cherts (T6-T7 and T11) show slightly higher retouch to debitage ratios (0.1-0.3).
@@
 Two types of points were identified: shouldered points (2 complete and 1 fragment, amounting to a total of 6.9%) and tanged and 4 winged points (9.3%).
-Other frequent retouched typologies are end scrappers (9.3%), notches (11.6%), retouched blanks (~14%) and splintered pieces (11.6%).
+Other frequent retouched typologies are end scrappers (9.3%), notches (11.6%), retouched blanks (\~14%) and splintered pieces (11.6%).
 Regarding chert type, a substantial portion of retouched tools are of unidentified chert types (n=13), mostly due to alterations (fire and post depositional).
@@
 Endscrappers (26.2%), splintered pieces (17%), burins (12.3%) and retouched flakes (7.7%) were the most prevalent.
-In comparison to the high frequency of point-related technology observed during the Solutrean, Vale Comprido points are barely present in chalcedony and chert (n=1), as they are mostly present in dolerite and chert was mainly used for other tools as seen in previous studies (**).
+In comparison to the high frequency of point-related technology observed during the Solutrean, Vale Comprido points are barely present in chalcedony and chert (n=1), as they are mostly present in dolerite and chert was mainly used for other tools as seen in previous studies (\*\*).
 Regarding the types of chalcedony and chert, local raw materials show 9 different typologies, with Vale Comprido points, Solutrean retouch and notches being present only in the local types, even if in small numbers.

### 2025-01-10 analysis/paper/paper.qmd
Measurement comparisons between assemblages.

@@
 
-Here is a citation [@Marwick2017]
-
 # Background
@@
 However, the smallest weight mean seem to be for cores of T6 (16.5 gr), although with high standard deviation and maximum value (16.5 and 48.6 gr, respectively).
+
+When comparing between technocomplexes, there are little differences in the general size and weight of the artifacts.
+For example, for key classes where size and weight differences could be expected (e.g., cores), the length measurements are mostly all located between 20-40 mm and weight between 10-40 gr.
+However, a few key differences were noted.
+T4 cores (the chert type most frequent in the Proto-Solutrean assemblage) shows the largest weights of cores of all assemblages, both in the Solutrean and Proto-Solutrean assemblage, as well as T5 in the Solutrean assemblage.
+Similarly, retouched tools show weights between 0.5-10 gr in all assemblages.
+T6 and T7 have consistently the smaller weights, especially compared to T2 during the Proto-Solutrean and Gravettian which shows slightly higher weights.
 
@@
 Two types of points were identified: shouldered points (2 complete and 1 fragment, amounting to a total of 6.9%) and tanged and 4 winged points (9.3%).
-Other frequent retouched typologies are end scrappers (9.3%), notches (11.6%), retouched blanks (\~14%) and splintered pieces (11.6%).
+Other frequent retouched typologies are endscrappers (9.3%), notches (11.6%), retouched blanks (\~14%) and splintered pieces (11.6%).
 Regarding chert type, a substantial portion of retouched tools are of unidentified chert types (n=13), mostly due to alterations (fire and post depositional).

### 2025-01-12 analysis/paper/paper.qmd
Added measurement results for flakes and elongated blanks, replacing blanks. Summarised table for Parameter results, excel table for now.

@@
 
+proto_original_chert <- proto_tech %>% 
+  filter(RawMaterial %in% c("Chert", "Chalcedony"))
+
 gravettian_tech <- gravettian_tech %>% 
@@
   tidyr::unite(col='ID', all_of(cols_to_concat_shelter), sep="-", remove=FALSE) # Concatenate YEAR and ID variables in one single variable (e.g. 05-500)
+
+# Dataset filtering for chert/chalcedony and levels (A-C), 3298 + 61 solutrean points, n=3359
+solutrean_original_chert <- solutrean_JC %>% 
+  filter(Layer %in% c("A", "B", "C")) %>% # Filtering Solutrean layers
+  filter(Raw.material %in% c("Chert", "Chalcedony"))
 
@@
   summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
+# For elongated blanks (complete)
+proto_stats_elong <- proto_stats %>% 
+  filter(CLASS == "ElongatedProd") %>% 
+  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
+  group_by(RMTYPE, CLASS) %>% 
+  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
 # For cores
@@
 gravettian_stats_flakes <- gravettian_stats %>% 
-  filter(CLASS == "Blank") %>%
+  filter(CLASS == "Flake") %>%
+  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
+  group_by(RMTYPE, CLASS) %>% 
+  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
+# For elongated blanks (complete)
+gravettian_stats_elong <- gravettian_stats %>% 
+  filter(CLASS == "ElongatedProd") %>%
   select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 

### 2025-01-13 analysis/paper/paper.qmd
Tool-to-debitage ratio graph

@@
 ```{r}
+#| label: fig-ratiotd
+#| fig-cap: Plotted ratios of tools-to-debitage by chert types (x axis) and assemblage (colour). Distance to source groups (local and non-local) are marked by different shapes.
+
+gravettian_ratiotd_b <-gravettian_ratiotd %>% 
+  dplyr::mutate("Assemblage" = "Gravettian") %>% 
+  select(Assemblage, RMTYPE, `Tool/Blank`)
+proto_ratiotd_b <- proto_ratiotd %>% 
+  dplyr::mutate("Assemblage" = "Proto-Solutrean") %>% 
+  select(Assemblage, RMTYPE, `Tool/Blank`)
+solutrean_ratiotd_b <- solutrean_ratiotd %>% 
+  dplyr::mutate("Assemblage" = "Solutrean") %>% 
+  select(Assemblage, RMTYPE, `Tool/Blank`)
+ratio_tooldebitage <- bind_rows(gravettian_ratiotd_b, proto_ratiotd_b,solutrean_ratiotd_b)
+
+ratio_tooldebitage %>%
+  dplyr::mutate(Source = case_when(
+    RMTYPE == "TYPE 1" | RMTYPE == "TYPE 2" | RMTYPE == "TYPE 3" | RMTYPE == "TYPE 4" | RMTYPE == "TYPE 5" ~ "Local",
+    RMTYPE == "TYPE 6" | RMTYPE == "TYPE 7" | RMTYPE == "TYPE 8" | RMTYPE == "TYPE 9" | RMTYPE == "TYPE 10" | RMTYPE == "TYPE 11" | RMTYPE == "TL" ~ "Non-local")) %>% 
+  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
+                                        "TYPE 5", "TYPE 6","TYPE 7", 
+                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
+  ggplot(aes(x=RMTYPE, y=`Tool/Blank`, shape = Source, colour = Assemblage)) +
+  geom_point(size=2)
+
+```
+
+```{r}
 #| label: tbl-tdratio-gravettian

### 2025-01-13 analysis/paper/paper.qmd
Ratios tool-to-debitage graph changes

@@
                                         "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
-  ggplot(aes(x=RMTYPE, y=`Tool/Blank`, shape = Source, colour = Assemblage)) +
+  ggplot(aes(x=Assemblage, y=`Tool/Blank`, colour = Source)) +
   geom_point(size=2)
+  #geom_text(aes(label = RMTYPE),nudge_x = 0.2, check_overlap = TRUE)
 

### 2025-01-13 analysis/paper/paper.qmd
Improvements to tool-to-debitage ratio graph and fig-cap correction. Blank-to-core graph created.

--- 2025-01-13 analysis/paper/paper.qmd
@@
 
-ratio_tooldebitage %>%
+# Scatterplot only with local vs non-local
+#ratio_tooldebitage |>
+#  dplyr::mutate(Source = case_when(
+#    RMTYPE == "TYPE 1" | RMTYPE == "TYPE 2" | RMTYPE == "TYPE 3" | RMTYPE == "TYPE 4" | RMTYPE == "TYPE 5" ~ "Local",
+#    RMTYPE == "TYPE 6" | RMTYPE == "TYPE 7" | RMTYPE == "TYPE 8" | RMTYPE == "TYPE 9" | RMTYPE == "TYPE 10" | RMTYPE == "TYPE 11" | RMTYPE == "TL" ~ "Non-local")) |>
+#  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
+#                                        "TYPE 5", "TYPE 6","TYPE 7", 
+#                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) |>
+#  ggplot(aes(x=Assemblage, y=`Tool/Blank`, colour = Source)) +
+#  geom_point(size=2)
+
+# Scatterplot with local vs non-local RM Types
+ratio_tooldebitage |>
   dplyr::mutate(Source = case_when(
     RMTYPE == "TYPE 1" | RMTYPE == "TYPE 2" | RMTYPE == "TYPE 3" | RMTYPE == "TYPE 4" | RMTYPE == "TYPE 5" ~ "Local",
-    RMTYPE == "TYPE 6" | RMTYPE == "TYPE 7" | RMTYPE == "TYPE 8" | RMTYPE == "TYPE 9" | RMTYPE == "TYPE 10" | RMTYPE == "TYPE 11" | RMTYPE == "TL" ~ "Non-local")) %>% 
-  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
-                                        "TYPE 5", "TYPE 6","TYPE 7", 
-                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
-  ggplot(aes(x=RMTYPE, y=`Tool/Blank`, shape = Source, colour = Assemblage)) +
-  geom_point(size=2)
+    RMTYPE == "TYPE 6" ~ "T6",
+    RMTYPE == "TYPE 7" ~ "T7",
+    RMTYPE == "TYPE 8" ~ "T8",
+    RMTYPE == "TYPE 10" ~ "T10",
+    RMTYPE == "TYPE 11" ~ "T11",
+    TRUE ~ RMTYPE)) |>
+  mutate(Source = factor(Source, levels = c("Local", "T6","T7", 
+                                        "T8", "T9", "T10", "T11", "TL"))) |>
+  ggplot(aes(x=Assemblage, y=`Tool/Blank`, colour = Source)) +
+  geom_point(size=3) +
+  scale_colour_manual(values = c("#D49B54", "#8B6969", "#CDAF95", "#2B4F60",
+                              "#8B7D6B", "#BDD2B6", "#798777", "#BDC7C9"))
 

### 2025-01-13 analysis/paper/paper.qmd
Same as previous commit. Forgot to save :p

@@
 ```{r}
+#| label: fig-ratiobc
+#| fig-cap: Plotted ratios of blanks-to-core by assemblage and by chert type. All local cherts were grouped in a single category "Local", while non-local cherts maintained their individual IDs.
+
+gravettian_ratiobc_b <-gravettian_ratiobc %>%
+  filter(`Blank/Core` != "Inf") %>% 
+  dplyr::mutate("Assemblage" = "Gravettian") %>% 
+  select(Assemblage, RMTYPE, `Blank/Core`)
+proto_ratiobc_b <- proto_ratiobc %>% 
+  filter(`Blank/Core` != "Inf") %>% 
+  dplyr::mutate("Assemblage" = "Proto-Solutrean") %>% 
+  select(Assemblage, RMTYPE, `Blank/Core`)
+solutrean_ratiobc_b <- solutrean_ratiobc %>% 
+  filter(`Blank/Core` != "Inf") %>% 
+  dplyr::mutate("Assemblage" = "Solutrean") %>% 
+  select(Assemblage, RMTYPE, `Blank/Core`)
+ratio_blankcore <- bind_rows(gravettian_ratiobc_b, proto_ratiobc_b,solutrean_ratiobc_b)
+
+# Scatterplot with local vs non-local RM Types
+ratio_blankcore |>
+  dplyr::mutate(Source = case_when(
+    RMTYPE == "TYPE 1" | RMTYPE == "TYPE 2" | RMTYPE == "TYPE 3" | RMTYPE == "TYPE 4" | RMTYPE == "TYPE 5" ~ "Local",
+    RMTYPE == "TYPE 6" ~ "T6",
+    RMTYPE == "TYPE 7" ~ "T7",
+    RMTYPE == "TYPE 8" ~ "T8",
+    RMTYPE == "TYPE 9" ~ "T9",
+    RMTYPE == "TYPE 10" ~ "T10",
+    RMTYPE == "TYPE 11" ~ "T11",
+    TRUE ~ RMTYPE)) |>
+  mutate(Source = factor(Source, levels = c("Local", "T6","T7", 
+                                        "T8", "T9", "T10", "T11", "TL"))) |>
+  ggplot(aes(x=Assemblage, y=`Blank/Core`, colour = Source)) +
+  geom_point(size=3) +
+  coord_cartesian(clip = "off") +
+  scale_colour_manual(values = c("#D49B54", "#8B6969", "#CDAF95", "#2B4F60",
+                              "#8B7D6B", "#BDD2B6", "#798777", "#BDC7C9"))
+```
+
+```{r}
 #| label: tbl-bcratio-gravettian
@@
 #| label: fig-ratiotd
-#| fig-cap: Plotted ratios of tools-to-debitage by chert types (x axis) and assemblage (colour). Distance to source groups (local and non-local) are marked by different shapes.
+#| fig-cap: Plotted ratios of tools-to-debitage by assemblage and by chert type. All local cherts were grouped in a single category "Local", while non-local cherts maintained their individual IDs.
 

### 2025-01-14 analysis/paper/paper.qmd
Scatterplots for width and length of all assemblages. Plots and colours rearranged.

@@
 
+# Scatterplot for width and length
+solutrean_scatter <- solutrean_stats |>
+  filter(CLASS %in% c("Flake","ElongatedProd", "Core", "RetouchedPiece")) |>
+  filter(MAXWIDTH !="0" & LENGTH != "0" & !RMTYPE %in% c("INDET","GB","TL")) |>
+  dplyr::mutate(RMTYPE = case_when(
+    RMTYPE == "TYPE 1" ~ "T1",
+    RMTYPE == "TYPE 2" ~ "T2",
+    RMTYPE == "TYPE 3" ~ "T3",
+    RMTYPE == "TYPE 4" ~ "T4",
+    RMTYPE == "TYPE 5" ~ "T5",
+    RMTYPE == "TYPE 6" ~ "T6",
+    RMTYPE == "TYPE 7" ~ "T7",
+    RMTYPE == "TYPE 8" ~ "T8",
+    RMTYPE == "TYPE 9" ~ "T9",
+    RMTYPE == "TYPE 10" ~ "T10",
+    RMTYPE == "TYPE 11" ~ "T11",
+    TRUE ~ RMTYPE)) |>
+   mutate(RMTYPE = factor(RMTYPE, levels = c("T1", "T2", "T3", "T4", "T5", "T6","T7", 
+                                        "T8", "T9", "T10", "T11"))) |>
+  mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) |>
+  ggplot(aes(x = LENGTH, y = MAXWIDTH, colour = RMTYPE)) +
+  geom_point(alpha = 0.9, size = 3) +
+  facet_wrap(~CLASS) +
+  labs(x="Length (mm)", y="Max. Width (mm") +
+  scale_colour_manual(values = c("#8B7B8B", "#FFE1FF", "#EED2EE", "#D8BFD8",
+                              "#CDB5CD", "#2B4F60" , "#4A708B", "#798777",
+                            "#9FB6CD", "#BDD2B6", "#BDC7C9"))
+
 # Boxplot for individual measurements
@@
   stat_summary(fun.y=mean, geom="point", shape=23) +
-  labs(y="Weight (gr)") +
+  labs(x="Raw material types" ,y="Weight (gr)") +
   coord_cartesian(ylim =  c(0, 50)) +
-  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
-                              "#FACE7F", "#8B6969", "#CDAF95", "#2B4F60",
-                              "#996666", "#BDD2B6", "#798777", "#BDC7C9")) +
+  scale_fill_manual(values = c("#8B7B8B", "#FFE1FF", "#EED2EE", "#D8BFD8",
+                              "#CDB5CD", "#2B4F60" , "#4A708B", "#798777",
+                            "#9FB6CD", "#BDD2B6", "#BDC7C9")) +
   theme(axis.text.x=element_blank(),
-        axis.ticks.x=element_blank(), axis.title.x = element_blank())
+        axis.ticks.x=element_blank())
 # Width
@@
         axis.ticks.x=element_blank(), axis.title.x = element_blank())
+
+# Scatterplots for width and length
+proto_scatter <- proto_stats |>
+  filter(CLASS %in% c("Flake","ElongatedProd", "Core", "RetouchedPiece")) |>
+  filter(MAXWIDTH !="0" & LENGTH != "0" & !RMTYPE %in% c("INDET","GB","TL")) |>
+  dplyr::mutate(RMTYPE = case_when(
+    RMTYPE == "TYPE 1" ~ "T1",
+    RMTYPE == "TYPE 2" ~ "T2",
+    RMTYPE == "TYPE 3" ~ "T3",
+    RMTYPE == "TYPE 4" ~ "T4",
+    RMTYPE == "TYPE 5" ~ "T5",
+    RMTYPE == "TYPE 6" ~ "T6",
+    RMTYPE == "TYPE 7" ~ "T7",
+    RMTYPE == "TYPE 8" ~ "T8",
+    RMTYPE == "TYPE 9" ~ "T9",
+    RMTYPE == "TYPE 10" ~ "T10",
+    TRUE ~ RMTYPE)) |>
+   mutate(RMTYPE = factor(RMTYPE, levels = c("T1", "T2", "T3", "T4", "T5", "T6","T7", 
+                                        "T8", "T9", "T10"))) |>
+  mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) |>
+  ggplot(aes(x = LENGTH, y = MAXWIDTH, colour = RMTYPE)) +
+  geom_point(alpha = 0.9, size = 3) +
+  facet_wrap(~CLASS) +
+  labs(x="Length (mm)", y="Max. Width (mm") +
+  scale_colour_manual(values = c("#8B7B8B", "#FFE1FF", "#EED2EE", "#D8BFD8",
+                              "#CDB5CD", "#2B4F60" , "#4A708B", "#798777",
+                            "#9FB6CD", "#BDD2B6"))
+
 # Length
@@
   coord_cartesian(ylim =  c(0, 60)) +
-  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
-                              "#FACE7F", "#8B6969", "#CDAF95", "#2B4F60",
-                              "#996666", "#BDD2B6", "#798777", "#BDC7C9")) +
+  scale_fill_manual(values = c("#8B7B8B", "#FFE1FF", "#EED2EE", "#D8BFD8",
+                              "#CDB5CD", "#2B4F60" , "#4A708B", "#798777",
+                            "#9FB6CD", "#BDD2B6")) +
   theme(axis.text.x=element_blank(),
@@
                                         "TYPE 5", "TYPE 6","TYPE 7", 
-                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
+                                        "TYPE 8", "TYPE 9", "TYPE 10"))) %>%
   mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) %>%
@@
   stat_summary(fun.y=mean, geom="point", shape=23) +
-  labs(y="Weight (gr)") +
+  labs(x="Raw material types", y="Weight (gr)") +
   coord_cartesian(ylim =  c(0, 50)) +
-  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
-                              "#FACE7F", "#8B6969", "#CDAF95", "#2B4F60",
-                              "#996666", "#BDD2B6", "#798777", "#BDC7C9")) +
+  scale_fill_manual(values = c("#8B7B8B", "#FFE1FF", "#EED2EE", "#D8BFD8",
+                              "#CDB5CD", "#2B4F60" , "#4A708B", "#798777",
+                            "#9FB6CD", "#BDD2B6")) +
   theme(axis.text.x=element_blank(),
-        axis.ticks.x=element_blank(), axis.title.x = element_blank())
+        axis.ticks.x=element_blank())
 
@@
     TRUE ~ CLASS))
+
+# Scatterplots width and length
+# Scatterplots for width and length
+gravettian_scatter <- gravettian_stats |>
+  filter(CLASS %in% c("Flake","ElongatedProd", "Core", "RetouchedPiece")) |>
+  filter(MAXWIDTH !="0" & LENGTH != "0" & !RMTYPE %in% c("INDET","GB","TL")) |>
+  dplyr::mutate(RMTYPE = case_when(
+    RMTYPE == "TYPE 1" ~ "T1",
+    RMTYPE == "TYPE 2" ~ "T2",
+    RMTYPE == "TYPE 3" ~ "T3",
+    RMTYPE == "TYPE 4" ~ "T4",
+    RMTYPE == "TYPE 5" ~ "T5",
+    RMTYPE == "TYPE 6" ~ "T6",
+    RMTYPE == "TYPE 7" ~ "T7",
+    RMTYPE == "TYPE 8" ~ "T8",
+    RMTYPE == "TYPE 9" ~ "T9",
+    RMTYPE == "TYPE 10" ~ "T10",
+    TRUE ~ RMTYPE)) |>
+   mutate(RMTYPE = factor(RMTYPE, levels = c("T1", "T2", "T3", "T4", "T5", "T6","T7", 
+                                        "T8", "T9", "T10"))) |>
+  mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) |>
+  ggplot(aes(x = LENGTH, y = MAXWIDTH, colour = RMTYPE)) +
+  geom_point(alpha = 0.9, size = 3) +
+  facet_wrap(~CLASS) +
+  labs(x="Length (mm)", y="Max. Width (mm") +
+  scale_colour_manual(values = c("#8B7B8B", "#FFE1FF", "#EED2EE", "#D8BFD8",
+                              "#CDB5CD", "#2B4F60" , "#4A708B", "#798777",
+                            "#9FB6CD", "#BDD2B6"))
 
@@
   stat_summary(fun.y=mean, geom="point", shape=23) +
-  labs(y="Weight (gr)") +
+  labs(x= "Raw material types", y="Weight (gr)") +
   coord_cartesian(ylim =  c(0, 60)) +
-  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
-                              "#FACE7F", "#8B6969", "#CDAF95", "#2B4F60",
-                              "#996666", "#BDD2B6", "#798777", "#BDC7C9")) +
+  scale_fill_manual(values = c("#8B7B8B", "#FFE1FF", "#EED2EE", "#D8BFD8",
+                              "#CDB5CD", "#2B4F60" , "#4A708B", "#798777",
+                            "#9FB6CD", "#BDD2B6")) +
   theme(axis.text.x=element_blank(),
-        axis.ticks.x=element_blank(), axis.title.x = element_blank())
+        axis.ticks.x=element_blank())
 
@@
 
-ggpubr::ggarrange(solutrean_stats_width, solutrean_stats_length, solutrean_stats_weight, labels = c("a", "b", "c"), heights = c(5, 5, 5), ncol = 1, common.legend = TRUE, legend = "bottom")
+ggpubr::ggarrange(solutrean_scatter, solutrean_stats_weight, labels = c("a", "b"), ncol = 2, common.legend = TRUE, legend = "bottom")
 
@@
 
-ggpubr::ggarrange(proto_stats_width, proto_stats_length, proto_stats_weight, labels = c("a", "b", "c"), heights = c(5, 5, 5), ncol = 1, common.legend = TRUE, legend = "bottom")
+ggpubr::ggarrange(proto_scatter, proto_stats_weight, labels = c("a", "b"), ncol = 2, common.legend = TRUE, legend = "bottom")
 
@@
 
-ggpubr::ggarrange(gravettian_stats_width, gravettian_stats_length, gravettian_stats_weight, labels = c("a", "b", "c"), heights = c(5, 5, 5), ncol = 1, common.legend = TRUE, legend = "bottom")
+ggpubr::ggarrange(gravettian_scatter, gravettian_stats_weight, labels = c("a", "b"), ncol = 2, common.legend = TRUE, legend = "bottom")
 

### 2025-01-14 analysis/paper/paper.qmd
Improved retouched tool tables

@@
 # Solutrean table
-solutrean_retouch <- solutrean_dataset %>%
+## Table preparation
+solutrean_retouch <- solutrean_dataset |>
     mutate(CLASS = case_when(
@@
     CLASS == "Retouched piece fragment" ~ "RetouchedPieceFrag",
-    TRUE ~ CLASS)) %>% 
-  filter(CLASS %in% c("RetouchedPiece", "RetouchedPieceFrag")) %>% 
-  select(RMTYPE, RETOUCHEDPIECETYPOLOGY) %>%
-  filter(RETOUCHEDPIECETYPOLOGY != "") %>% 
-  group_by(RMTYPE)
-
-solutrean_retouch_tab <- solutrean_retouch %>% 
-  tabyl(RETOUCHEDPIECETYPOLOGY, RMTYPE) %>% 
-  adorn_totals(c("row", "col")) %>% 
-  adorn_percentages("col") %>% 
-  adorn_pct_formatting(2) %>% 
+    TRUE ~ CLASS)) |>
+  filter(CLASS %in% c("RetouchedPiece", "RetouchedPieceFrag")) |>
+  select(RMTYPE, RETOUCHEDPIECETYPOLOGY) |>
+  filter(RETOUCHEDPIECETYPOLOGY != "") |>
+  dplyr::mutate(RMTYPE = case_when(
+    RMTYPE == "TYPE 1" ~ "T1",
+    RMTYPE == "TYPE 2" ~ "T2",
+    RMTYPE == "TYPE 3" ~ "T3",
+    RMTYPE == "TYPE 4" ~ "T4",
+    RMTYPE == "TYPE 5" ~ "T5",
+    RMTYPE == "TYPE 6" ~ "T6",
+    RMTYPE == "TYPE 7" ~ "T7",
+    RMTYPE == "TYPE 8" ~ "T8",
+    RMTYPE == "TYPE 9" ~ "T9",
+    RMTYPE == "TYPE 10" ~ "T10",
+    RMTYPE == "TYPE 11" ~ "T11",
+    TRUE ~ RMTYPE))
+## Janitor tabyl for final table with n and %
+solutrean_retouch_tab <- solutrean_retouch |>
+  tabyl(RETOUCHEDPIECETYPOLOGY, RMTYPE) |>
+  adorn_totals(c("row", "col")) |>
+  adorn_percentages("col") |>
+  adorn_pct_formatting(2) |> 
   adorn_ns("front")
+## Cleaning final table
+solutrean_retouch_tab <- solutrean_retouch_tab |>
+    dplyr::mutate(T1 = case_when( # Replace 0% values with -
+    T1 == "0   (0.00%)" ~ "-",
+    TRUE ~ T1)) |>
+  dplyr::mutate(T2 = case_when(
+    T2 == "0   (0.00%)" ~ "-",
+    TRUE ~ T2)) |>
+  dplyr::mutate(T5 = case_when(
+    T5 == "0   (0.00%)" ~ "-",
+    TRUE ~ T5)) |>
+  dplyr::mutate(T6 = case_when(
+    T6 == "0   (0.00%)" ~ "-",
+    TRUE ~ T6)) |>
+  dplyr::mutate(T7 = case_when(
+    T7 == "0   (0.00%)" ~ "-",
+    TRUE ~ T7)) |>
+  dplyr::mutate(T8 = case_when(
+    T8 == "0   (0.00%)" ~ "-",
+    TRUE ~ T8)) |>
+  dplyr::mutate(T11 = case_when(
+    T11 == "0   (0.00%)" ~ "-",
+    TRUE ~ T11)) |>
+  dplyr::mutate(INDET = case_when(
+    INDET == "0   (0.00%)" ~ "-",
+    TRUE ~ INDET)) |>
+  select(RETOUCHEDPIECETYPOLOGY, T1, T2, T5, T6, T7, T8, T11, INDET, Total) # Ordering values
 
 # Proto-solutrean table
-proto_retouch <- phase_dataset %>% 
-  filter(CLASS %in% c("RetouchedPiece", "RetouchedPieceFrag")) %>%
-  filter(Level %in% c("4E", "4E/5")) %>% 
-  select(RMTYPE, RETOUCHEDPIECETYPOLOGY) %>% 
-  dplyr::group_by(RMTYPE)
-
+## Table preparation
+proto_retouch <- phase_dataset |>
+  filter(CLASS %in% c("RetouchedPiece", "RetouchedPieceFrag")) |>
+  filter(Level %in% c("4E", "4E/5")) |>
+  select(RMTYPE, RETOUCHEDPIECETYPOLOGY) |>
+  dplyr::mutate(RMTYPE = case_when(
+    RMTYPE == "TYPE 1" ~ "T1",
+    RMTYPE == "TYPE 2" ~ "T2",
+    RMTYPE == "TYPE 3" ~ "T3",
+    RMTYPE == "TYPE 4" ~ "T4",
+    RMTYPE == "TYPE 5" ~ "T5",
+    RMTYPE == "TYPE 6" ~ "T6",
+    RMTYPE == "TYPE 7" ~ "T7",
+    RMTYPE == "TYPE 8" ~ "T8",
+    RMTYPE == "TYPE 9" ~ "T9",
+    RMTYPE == "TYPE 10" ~ "T10",
+    RMTYPE == "TYPE 11" ~ "T11",
+    TRUE ~ RMTYPE))
+## Janitor tabyl for final table with n and %
 proto_retouch_tab <- proto_retouch %>%
@@
   adorn_ns("front")
+## Cleaning final table
+proto_retouch_tab <- proto_retouch_tab |>
+  dplyr::mutate(T1 = case_when( # Replace 0% values with -
+    T1 == "0   (0.00%)" ~ "-",
+    TRUE ~ T1)) |>
+  dplyr::mutate(T2 = case_when(
+    T2 == "0   (0.00%)" ~ "-",
+    TRUE ~ T2)) |>
+  dplyr::mutate(T3 = case_when(
+    T3 == "0   (0.00%)" ~ "-",
+    TRUE ~ T3)) |>
+  dplyr::mutate(T4 = case_when(
+    T4 == "0   (0.00%)" ~ "-",
+    TRUE ~ T4)) |>
+  dplyr::mutate(T5 = case_when(
+    T5 == "0   (0.00%)" ~ "-",
+    TRUE ~ T5)) |>
+  dplyr::mutate(T6 = case_when(
+    T6 == "0   (0.00%)" ~ "-",
+    TRUE ~ T6)) |>
+  dplyr::mutate(T7 = case_when(
+    T7 == "0   (0.00%)" ~ "-",
+    TRUE ~ T7)) |>
+  dplyr::mutate(T8 = case_when(
+    T8 == "0   (0.00%)" ~ "-",
+    TRUE ~ T8)) |>
+  dplyr::mutate(T9 = case_when(
+    T9 == "0   (0.00%)" ~ "-",
+    TRUE ~ T9)) |>
+  dplyr::mutate(TL = case_when(
+    TL == "0   (0.00%)" ~ "-",
+    TRUE ~ TL)) |>
+  dplyr::mutate(INDET = case_when(
+    INDET == "0   (0.00%)" ~ "-",
+    TRUE ~ INDET))
 
 # Gravettian table
+## Table preparation
 gravettian_retouch <- phase_dataset %>% 
@@
   select(RMTYPE, RETOUCHEDPIECETYPOLOGY) %>% 
-  dplyr::group_by(RMTYPE)
-
+  dplyr::group_by(RMTYPE) |>
+  dplyr::mutate(RMTYPE = case_when(
+    RMTYPE == "TYPE 1" ~ "T1",
+    RMTYPE == "TYPE 2" ~ "T2",
+    RMTYPE == "TYPE 3" ~ "T3",
+    RMTYPE == "TYPE 4" ~ "T4",
+    RMTYPE == "TYPE 5" ~ "T5",
+    RMTYPE == "TYPE 6" ~ "T6",
+    RMTYPE == "TYPE 7" ~ "T7",
+    RMTYPE == "TYPE 8" ~ "T8",
+    RMTYPE == "TYPE 9" ~ "T9",
+    RMTYPE == "TYPE 10" ~ "T10",
+    TRUE ~ RMTYPE))
+## Janitor tabyl for final table with n and %
 gravettian_retouched_tab <- gravettian_retouch %>% 
@@
   adorn_totals(c("row", "col")) %>% 
-  adorn_percentages("row") %>% 
+  adorn_percentages("col") %>% 
   adorn_pct_formatting(2) %>% 
   adorn_ns("front")
+## Cleaning final table
+gravettian_retouched_tab <- gravettian_retouched_tab |>
+  dplyr::mutate(T1 = case_when( # Replace 0% values with -
+    T1 == "0   (0.00%)" ~ "-",
+    TRUE ~ T1)) |>
+  dplyr::mutate(T2 = case_when(
+    T2 == "0   (0.00%)" ~ "-",
+    TRUE ~ T2)) |>
+  dplyr::mutate(T3 = case_when(
+    T3 == "0   (0.00%)" ~ "-",
+    TRUE ~ T3)) |>
+  dplyr::mutate(T4 = case_when(
+    T4 == "0   (0.00%)" ~ "-",
+    TRUE ~ T4)) |>
+  dplyr::mutate(T5 = case_when(
+    T5 == "0   (0.00%)" ~ "-",
+    TRUE ~ T5)) |>
+  dplyr::mutate(T6 = case_when(
+    T6 == "0   (0.00%)" ~ "-",
+    TRUE ~ T6)) |>
+  dplyr::mutate(T7 = case_when(
+    T7 == "0   (0.00%)" ~ "-",
+    TRUE ~ T7)) |>
+  dplyr::mutate(T8 = case_when(
+    T8 == "0   (0.00%)" ~ "-",
+    TRUE ~ T8)) |>
+  dplyr::mutate(T9 = case_when(
+    T9 == "0   (0.00%)" ~ "-",
+    TRUE ~ T9)) |>
+  dplyr::mutate(T10 = case_when(
+    T10 == "0   (0.00%)" ~ "-",
+    TRUE ~ T10)) |>
+  dplyr::mutate(TL = case_when(
+    TL == "0   (0.00%)" ~ "-",
+    TRUE ~ TL)) |>
+  dplyr::mutate(INDET = case_when(
+    INDET == "0   (0.00%)" ~ "-",
+    TRUE ~ INDET)) |>
+  select(RETOUCHEDPIECETYPOLOGY, T1, T2, T3, T4, T5, T6, T7, T8, T9, T10, TL, INDET, Total) # Ordering values
 

### 2025-01-14 analysis/Supplementary/supplementary_tables.qmd
Supplementary materials folder with supplementary tables. Removed supplementary tables from the main text, cleaned code

--- 2025-01-13 analysis/paper/paper.qmd
@@
 ---
-title: "Placeholder for article 3"
+title: "SOM result tables"
 author:
@@
 csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
-abstract: |
-  Text of abstract
-keywords: |
-  keyword 1; keyword 2; keyword 3
-highlights: |
-  These are the highlights. 
 ---
-
-<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->
-
-<!-- With the following code you can access and display values from the yml header above. -->
-
-Keywords: `r rmarkdown::metadata$keywords`
-
-Highlights: `r rmarkdown::metadata$highlights`
-
-<!-- The actual document text starts here: -->
-
-# Introduction
-
-# Background
-
-## Aims
-
-Our goal is to understand the provisioning, use and management strategies of chert lithic resources of the hunter-gatherer groups of Vale Boi during the Upper Paleolithic.
-We focus on the technological characteristics of the two broader groups of cherts identified at the site: local and non-local, based on the notion that local raw materials will show technological patterns different from those which are non-local.
-Expanding upon our previous works on chert types present throughout the UP sequence in the Terrace and Shelter area of Vale Boi, and their possible interpretations, and following the previously mentioned studies and models about raw material management and lithic technology organisation focusing mainly on the different management of local and non-local raw materials, associated with mobility and occupation length, we derive two main expectations (fig.
-1):
-
-1.
-Local raw materials are expected to present a) low use-lives, with knapping strategies characterised by manufacture, use and discard; b) a decreased reduction intensity and low tools to debitage ratio; c) larger cores and lower blank-to-core ratio; d) less extensive reduction and reworking of tools; e) higher tool typology diversity.
-This may be related to the abundance of local raw materials which leads to less intensive life-use extension, but also to the residential characteristics of the site and low mobility, which increases the reliance on local raw materials.
-
-2.
-Non-local raw materials are expected to present a) a majority of formal tools, in the case of individual provisioning or higher use-lives, with knapping strategies characterised by the manufacture, use and maintenance; b) an increased reduction intensity and high tools to debitage ratio; c) smaller cores and higher blank-to-core ratio; d) more extensive reduction and high presence of retouch and/or tool maintenance; e) lower tool diversity, characterised by less specialised tools, such as the presence of retouched blanks as multipurpose tools.
-This may be related to scarcer or poorer quality of raw materials which leads to the prepared transportation of lithic resources to a site, but also related to more mobile settlement patterns as well as short-term site occupations.
-
-Despite these expectations, it is important to notice that, especially in occupations with a high percentage of non-local cherts at the site (e.g., Gravettian occupations of level 7 and 6 of the Terrace at Vale Boi), these technological may not reflect a preoccupation with intensive management of a scarce and limited non-local resource, and be more a reflection of the type of provisioning applied following the suggested model by Kuhn (2004).
-In this sense, distance from source may have had little intrinsic meaning in the past, especially regarding non-local raw materials.
-There are several ways in which raw materials may be transported to a site, and the relationship between distance and cost may vary depending on how this takes place [@kuhn2004].
-When non-local raw materials are as abundant as local ones, the difference in the technological organization between these materials may be less obvious (Andrefsky, 2004).
-
-# Methods
-
-# Results
-
-```{r}
-#| label: library
-
-library(readr)
-library(dplyr)
-library(tidyr)
-library(stringr)
-library(ggplot2)
-library(janitor)
-library(ggpubr)
-library(knitr)
-
-```
 
@@
 ```{r}
-#| label: terrace-databases-setup
+#| label: databases-setup
 
+# Terrace dataset setup
 cols_to_concat_terrace <- c("UNIT", "ID")
@@
 
-```
-
-```{r}
-#| label: shelter-solutrean-setup
-
+# Shelter dataset setup
 cols_to_concat_shelter <- c("YEAR", "ID") # Concatenate YEAR and ID variables in one single variable (e.g. 05-500)
@@
 
+# Blank-to-core ratios
+
 ```{r}
-#| label: general-tables
+#| label: ratio-blankcore
 
-solutrean_general <- solutrean_dataset %>%
-  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
-                                        "TYPE 5", "TYPE 6","TYPE 7", 
-                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL", "INDET"))) %>% 
-  tabyl(CLASS, RMTYPE) %>%
-  adorn_totals(c("row", "col")) %>%
-  adorn_percentages("col") %>%
-  adorn_pct_formatting(2) %>%
-  adorn_ns("front")
+# Solutrean ratio
+solutrean_ratiobc <- solutrean_dataset %>% 
+  mutate(CLASS = case_when(
+    CLASS == "Flake" | CLASS == "ElongBlank" | CLASS == "Thinning flake" ~ "Blank",
+    CLASS == "FlakeFrag" | CLASS == "ElongBlankFrag" ~ "BlankFrag",
+    TRUE ~ CLASS)) %>% 
+  mutate(CLASS = case_when(
+    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
+    TRUE ~ CLASS)) %>% 
+  filter(RMTYPE != "INDET") %>% 
+  filter(CLASS %in% c("Blank", "Core")) %>% 
+  count(RMTYPE, CLASS, sort = TRUE) %>% 
+  group_by(RMTYPE, CLASS) %>%
+  summarise(n=sum(n)) %>% 
+  spread(CLASS, n, fill=0) %>% 
+  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
+  filter(!sum <= 5) %>% 
+  mutate("Blank/Core" = Blank / Core)
 
-proto_general <- phase_dataset %>% 
+# Proto-Solutrean ratio
+proto_ratiobc <- phase_dataset %>% 
   filter(Level %in% c("4E", "4E/5")) %>% 
-  filter(CLASS != "Chip") %>% 
-  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
-                                        "TYPE 5", "TYPE 6","TYPE 7", 
-                                        "TYPE 8", "TYPE 9", "TYPE 10", "TL", "INDET"))) %>% 
-  tabyl(CLASS, RMTYPE) %>%
-  adorn_totals(c("row", "col")) %>%
-  adorn_percentages("col") %>%
-  adorn_pct_formatting(2) %>%
-  adorn_ns("front")
+  filter(RMTYPE != "INDET") %>% 
+  mutate(CLASS = case_when(
+    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
+    TRUE ~ CLASS)) %>% 
+  filter(CLASS %in% c("Blank", "Core")) %>% 
+  count(RMTYPE, CLASS, sort = TRUE) %>% 
+  group_by(RMTYPE, CLASS) %>%
+  summarise(n=sum(n)) %>% 
+  spread(CLASS, n, fill=0) %>% 
+  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
+  filter(!sum <= 5) %>%
+  mutate("Blank/Core" = Blank / Core)
 
-gravettian_general <- phase_dataset %>% 
+# Gravettian ratio
+gravettian_ratiobc <- phase_dataset %>% 
   filter(Level %in% c(6, 7)) %>% 
-  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
-                                        "TYPE 5", "TYPE 6","TYPE 7", 
-                                        "TYPE 8", "TYPE 9", "TYPE 10", "TL", "INDET"))) %>% 
-  tabyl(CLASS, RMTYPE) %>%
-  adorn_totals(c("row", "col")) %>%
-  adorn_percentages("col") %>%
-  adorn_pct_formatting(2) %>%
-  adorn_ns("front")
+  filter(RMTYPE != "INDET") %>% 
+  mutate(CLASS = case_when(
+    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
+    TRUE ~ CLASS)) %>% 
+  filter(CLASS %in% c("Blank", "Core")) %>% 
+  count(RMTYPE, CLASS, sort = TRUE) %>% 
+  group_by(RMTYPE, CLASS) %>%
+  summarise(n=sum(n)) %>% 
+  spread(CLASS, n, fill=0) %>% 
+  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
+  filter(!sum <= 5) %>% 
+  mutate("Blank/Core" = Blank / Core)
 
@@
 ```{r}
-#| label: tbl-general-gravettian
-#| tbl-cap: Number and percentage of technotypological classes by raw material type from the Gravettian assemblage (levels 6 and 7) of the Terrace area of Vale Boi.
+#| label: tbl-bcratio-solutrean
+#| tbl-cap: Blank to core ratio by raw material type of the Solutrean assemblage (levels A to C) of the Shelter area.
 
-kable(gravettian_general)
+kable(solutrean_ratiobc)
 
@@
 ```{r}
-#| label: tbl-general-proto
-#| tbl-cap: Number and percentage of technotypological classes by raw material type from the Proto-Solutrean assemblage (levels 4E and top of level 5) of the Terrace area of Vale Boi.
+#| label: tbl-bcratio-proto
+#| tbl-cap: Blank to core ratio by raw material type of the Proto-Solutrean assemblage (levels 4E and top of level 5) of the Terrace area.
 
-kable(proto_general)
+kable(proto_ratiobc)
 
@@
 ```{r}
-#| label: tbl-general-solutrean
-#| tbl-cap: Number and percentage of technotypological classes by raw material type from the Solutrean assemblage (levels A to C) of the Shelter area of Vale Boi.
+#| label: tbl-bcratio-gravettian
+#| tbl-cap: Blank to core ratio by raw material type of the Gravettian assemblage (levels 6 and 7) of the Terrace area.
 
-kable(solutrean_general)
+kable(gravettian_ratiobc)
+
+```
+
+# Tool-to-debitage ratios
+
+```{r}
+#| label: ratios-retouchdebitage
+
+# Solutrean
+solutrean_ratiotd <- solutrean_dataset %>% 
+  filter(RMTYPE != "INDET") %>% 
+  mutate(CLASS = case_when(
+    CLASS == "Flake" | CLASS == "ElongBlank" | CLASS == "Thinning flake" ~ "Blank",
+    CLASS == "FlakeFrag" | CLASS == "ElongBlankFrag" ~ "BlankFrag",
+    CLASS == "Retouched piece" ~ "RetouchedPiece",
+    CLASS == "Retouched piece fragment" ~ "RetouchedPieceFrag",
+    TRUE ~ CLASS)) %>% 
+  mutate(CLASS = case_when(
+    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
+    CLASS == "RetouchedPieceFrag" & PIECECOMPLETENESS == "Proximal" ~ "RetouchedPiece",
+    TRUE ~ CLASS)) %>% 
+  filter(CLASS %in% c("Blank", "RetouchedPiece")) %>% 
+  count(RMTYPE, CLASS, sort = TRUE) %>% 
+  group_by(RMTYPE, CLASS) %>%
+  summarise(n=sum(n)) %>% 
+  spread(CLASS, n, fill=0) %>% 
+  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
+  filter(!sum <= 5) %>% 
+  mutate("Tool/Blank" = RetouchedPiece / Blank)
+
+# Proto-Solutrean
+proto_ratiotd <- phase_dataset %>% 
+  filter(Level %in% c("4E", "4E/5")) %>% 
+  filter(RMTYPE != "INDET") %>% 
+  mutate(CLASS = case_when(
+    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
+    CLASS == "RetouchedPieceFrag" & PIECECOMPLETENESS == "Proximal" ~ "RetouchedPiece",
+    TRUE ~ CLASS)) %>% 
+  filter(CLASS %in% c("Blank", "RetouchedPiece")) %>% 
+  count(RMTYPE, CLASS, sort = TRUE) %>% 
+  group_by(RMTYPE, CLASS) %>%
+  summarise(n=sum(n)) %>% 
+  spread(CLASS, n, fill=0) %>% 
+  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
+  filter(!sum <= 5) %>% 
+  mutate("Tool/Blank" = RetouchedPiece / Blank)
+
+proto5_ratiotd <- phase_dataset %>% 
+  filter(Level == 5) %>% 
+  filter(RMTYPE != "INDET") %>% 
+  mutate(CLASS = case_when(
+    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
+    CLASS == "RetouchedPieceFrag" & PIECECOMPLETENESS == "Proximal" ~ "RetouchedPiece",
+    TRUE ~ CLASS)) %>% 
+  filter(CLASS %in% c("Blank", "RetouchedPiece")) %>% 
+  count(RMTYPE, CLASS, sort = TRUE) %>% 
+  group_by(RMTYPE, CLASS) %>%
+  summarise(n=sum(n)) %>% 
+  spread(CLASS, n, fill=0) %>% 
+  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
+  filter(!sum <= 5) %>% 
+  mutate("Tool/Blank" = RetouchedPiece / Blank)
+
+# Gravettian
+gravettian_ratiotd <- phase_dataset %>% 
+  filter(Level %in% c(6, 7)) %>% 
+  filter(RMTYPE != "INDET") %>% 
+  mutate(CLASS = case_when(
+    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
+    CLASS == "RetouchedPieceFrag" & PIECECOMPLETENESS == "Proximal" ~ "RetouchedPiece",
+    TRUE ~ CLASS)) %>% 
+  filter(CLASS %in% c("Blank", "RetouchedPiece")) %>% 
+  count(RMTYPE, CLASS, sort = TRUE) %>% 
+  group_by(RMTYPE, CLASS) %>%
+  summarise(n=sum(n)) %>% 
+  spread(CLASS, n, fill=0) %>% 
+  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
+  filter(!sum <= 5) %>% 
+  mutate("Tool/Blank" = RetouchedPiece / Blank)
 
@@
 ```{r}
-#| label: raw-material-graph
-#| include: false
+#| label: tbl-tdratio-solutrean
+#| tbl-cap: Tools to debitage ratio by raw material of the Solutrean assemblage (levels A to C) of the Shelter area.
 
-library(ggplot2)
-library(scales)
-
-solutrean_dataset %>%
-  filter(RMTYPE != "INDET") %>% 
-  dplyr::group_by(LEVEL, RMTYPE) %>% 
-  dplyr::summarise(n = n()) %>%
-  mutate(percent = n/sum(n)) %>% 
-  ggplot(aes(x = LEVEL, y = percent, fill = RMTYPE)) +
-  geom_bar(stat = "identity", width = 0.5) +
-  scale_y_continuous(labels = scales::percent_format()) +
-  #theme_minimal() + #type of theme
-  labs(y = "Count %") +# change labels
-  guides(fill=guide_legend(title=NULL)) + # remove legend title
-  geom_text(aes(label=ifelse(percent >= 0.05, paste0(sprintf("%.1f", percent*100),"%"),"")),
-                position=position_stack(vjust=0.5), colour="white", size=3) +
-  theme(axis.title.x = element_blank()) + 
-  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
-                              "#FACE7F", "#8B6969", "#CDAF95", "#2B4F60",
-                              "#996666", "#BDD2B6", "#798777", "#BDC7C9"))
+kable(solutrean_ratiotd)
 
@@
 
-## Assemblage description
+```{r}
+#| label: tbl-tdratio-proto
+#| tbl-cap: Tools to debitage ratio by raw material of the Proto-Solutrean assemblage (levels 4E and top of level 5) of the Terrace area.
 
-For the Gravettian assemblage, local cherts show similar patterns, with T2 and T5 being the types with higher numbers of artefacts (n=110 and n=72, respectively).
-For all local cherts, blanks represent more than 50% of the artefacts.
-Cores and core maintenance products are only present in T2 and T5, although in small frequencies.
-For T2, cores represent 13.6% of the group and core preparation and maintenance products represent 1.8% (n=2).
-For T5 only 4 core fragments and 1 core preparation and maintenance product are present.
-Retouched tools seem to be especially relevant in T2 (\~18%), being the most frequent class after blanks.
-The presence of burin spalls (8.1%) in T2 also hints at the production of burins.
-In comparison, T5 shows similar percentages of retouched tools and shatter (\~12% and \~14%, respectively).
-T1 (chalcedony) shows similar patterns, although with higher percentages of shatter (34.3%).
-In comparison, non-local chert types (especially T6 with n=169 and T7 with n=86) show similar percentages of blanks (\~40-50%) but higher percentages of retouched tools (\~30%).
-Cores, core fragments and core preparation and maintenance products are also present even if in small percentages (\<4%), as well as shatter (\~12%).
-In both types burin spalls are also present (7.1% for T6 and 4.6% for T7).
-The TL cherts show, as expected, blanks and blank fragments (n=3 and n=2, respectively), and retouched tools (n=4).
-The presence of core maintenance products in TL is possibly related to the limited knapping of TL01 (jasper).
+kable(proto_ratiotd)
 
-Regarding the Proto-Solutrean assemblage, the local types show high frequency of blanks (\>50%, including blank fragments), followed by shatter (\~11-27%), cores (\~9-12%) and retouched tools (\~5-15% ).
-A single burin spall was identified in T4.
-There is also the absence of core preparation and maintenance products across all local cherts.
-T1 (chalcedony) is composed mostly of blanks and a retouched tool (Vale Comprido point).
-The absence of cores but presence of shatter may hint at the knapping of this raw material at the site, but the absence of formal cores perhaps due to the quality of nodules/lenses, which are characterised by fractures.
-For non-local types, especially those with higher number of artefacts (T6-T8), blanks and blank fragments are the most frequent class (`r 38.3+10`%, `r 40.4+12.7`% and `r 46.1+23.1`%, respectively).
-For T6 the absence of cores but the presence of shatter and a single core preparation and maintenance product suggests knapping on site.
-Similarly, T7 shows the presence of cores and core fragments (`r 4.2+4.2`%), shatter (17.0%) and a single core preparation and maintenance product.
-In comparison, T8, despite more limited in number (n=13), shows only one core, with the rest of the artefacts being blanks (as previously noted) or retouched tools (23.1% and n=3).
-Aside from the TL artefacts, all non-local cores show at least 1 core and frequently the presence of shatter.
-As expected, the TL artifacts include a blank and a retouched.
-However, a shatter was also identified.
+```
 
-The Solutrean assemblages show similar patterns to those observed in previous assemblages.
-For local cherts, especially T2 (n=506) which accounts for the majority of local chert present in the studied samples, the majority of the debitage is composed of flakes and elongated blanks (`r 42+13.4+17.8+10.7`%).
-Cores are also present (10.5%), alongside a small percentage of core preparation and maintenance products (1.3%).
-In comparison to debitage products, retouched tools are scarce in the sample (\~3%).
-The non-local cherts most present in the Solutrean of the Shelter area are T6 (n=86) and T8 (n=30).
-These show similar patterns to the local cherts, with high percentages of blanks (\>50%).
-Cores and core preparation and maintenance are also present in small percentages (\<10%), as well as retouched tools (\<10%).
-However, these specific results are truncated by the lack of shatter, which were not present in the sample used for the study.
-It is important to notice, however, that alike previous results from the Solutrean assemblages of the Terrace area, a large portion of the studied sample was altered and did not allow for a chert type attribution (`r 201*100/914`%).
+```{r}
+#| label: tbl-tdratio-gravettian
+#| tbl-cap: Tools to debitage ratio by raw material of the Gravettian assemblage (levels 6 and 7) of the Terrace area.
 
-## Measurement and weight descriptive statistics
+kable(gravettian_ratiotd)
 
-In the Gravettian assemblages, blanks and retouched tools show little differences in the measurements of artefacts of local and non-local chert types.
-The only exception are T6 and T7 blanks, which show slightly lower weight means and lower maximum length (\~35 mm) than the other types.
-For cores, T2 shows bigger (heavier) values than T6 (means of 21.3 and 12.6, and maximum weight of 37.2 gr and 22.5 gr, respectively), even if they seem to vary in morphology, with little patterns regarding thickness, width and length.
+```
 
-In the Proto-Solutrean assemblage, blanks show lower values for length and weight in non-local cherts (T6-T8), varying between a mean of \~19-20 mm and maximum of \<30 mm for length, and mean of \<3 gr and maximum of \<7 gr for weight.
-In retouched tools, weight also shows smaller means for non-local cherts (\~3-2 gr), although with higher standard deviation (\~1.6).
-Regarding cores, the sample size for non-local cherts is small and only T7 statistics could be obtained.
-Despite this, when compared to local cherts (T2-T5), T7 cores show smaller mean and maximum values (7.3 gr and 8 gr, respectively).
-
-For the Solutrean assemblage, there seems to be no clear difference in the measurements and weights of artefacts between local and non-local cherts.
-This is true especially for the most represented classes (flakes and elongated blanks).
-Regarding cores, many of the local and non-local cherts are not available for the statistics due to missing data.
-However, the smallest weight mean seem to be for cores of T6 (16.5 gr), although with high standard deviation and maximum value (16.5 and 48.6 gr, respectively).
-
-When comparing between technocomplexes, there are little differences in the general size and weight of the artifacts.
-For example, for key classes where size and weight differences could be expected (e.g., cores), the length measurements are mostly all located between 20-40 mm and weight between 10-40 gr.
-However, a few key differences were noted.
-T4 cores (the chert type most frequent in the Proto-Solutrean assemblage) shows the largest weights of cores of all assemblages, both in the Solutrean and Proto-Solutrean assemblage, as well as T5 in the Solutrean assemblage.
-Similarly, retouched tools show weights between 0.5-10 gr in all assemblages.
-T6 and T7 have consistently the smaller weights, especially compared to T2 during the Proto-Solutrean and Gravettian which shows slightly higher weights.
+# Measurements
 
@@
 
-# Boxplot for individual measurements
-# Length
-solutrean_stats_length <- solutrean_stats %>%
-  filter(CLASS %in% c("Flake", "ElongatedProd", "Core", "RetouchedPiece")) %>% 
-  filter(LENGTH !="0" & !RMTYPE %in% c("INDET","GB","TL")) %>%
-  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
-                                        "TYPE 5", "TYPE 6","TYPE 7", 
-                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
-  mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) %>%
-  ggplot(aes(RMTYPE, LENGTH, fill = RMTYPE)) +
-  geom_boxplot(aes(fill=factor(RMTYPE)), outlier.shape=21) +
-  facet_wrap(~CLASS) +
-  stat_summary(fun.y=mean, geom="point", shape=23) +
-  labs(y="Length (mm)") +
-  #coord_cartesian(ylim =  c(0, 60)) +
-  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
-                              "#FACE7F", "#8B6969", "#CDAF95", "#2B4F60",
-                              "#996666", "#BDD2B6", "#798777", "#BDC7C9")) +
-  theme(axis.text.x=element_blank(),
-        axis.ticks.x=element_blank(), axis.title.x = element_blank())
-# Weight
-solutrean_stats_weight <- solutrean_stats %>%
-  filter(CLASS %in% c("Flake", "ElongatedProd", "Core", "RetouchedPiece")) %>% 
-  filter(WEIGHT !="0" & !RMTYPE %in% c("INDET","GB","TL")) %>%
-  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
-                                        "TYPE 5", "TYPE 6","TYPE 7", 
-                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
-  mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) %>%
-  ggplot(aes(RMTYPE, WEIGHT, fill = RMTYPE)) +
-  geom_boxplot(aes(fill=factor(RMTYPE)), outlier.shape=21) +
-  facet_wrap(~CLASS) +
-  stat_summary(fun.y=mean, geom="point", shape=23) +
-  labs(y="Weight (gr)") +
-  coord_cartesian(ylim =  c(0, 50)) +
-  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
-                              "#FACE7F", "#8B6969", "#CDAF95", "#2B4F60",
-                              "#996666", "#BDD2B6", "#798777", "#BDC7C9")) +
-  theme(axis.text.x=element_blank(),
-        axis.ticks.x=element_blank(), axis.title.x = element_blank())
-# Width
-solutrean_stats_width <- solutrean_stats %>%
-  filter(CLASS %in% c("Flake", "ElongatedProd", "Core", "RetouchedPiece")) %>% 
-  filter(MAXWIDTH !="0" & !RMTYPE %in% c("INDET","GB","TL")) %>%
-  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
-                                        "TYPE 5", "TYPE 6","TYPE 7", 
-                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
-  mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) %>%
-  ggplot(aes(RMTYPE, MAXWIDTH, fill = RMTYPE)) +
-  geom_boxplot(aes(fill=factor(RMTYPE)), outlier.shape=21) +
-  facet_wrap(~CLASS) +
-  stat_summary(fun.y=mean, geom="point", shape=23) +
-  labs(y="Width (mm)") +
-  #coord_cartesian(ylim =  c(0, 60)) +
-  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
-                              "#FACE7F", "#8B6969", "#CDAF95", "#2B4F60",
-                              "#996666", "#BDD2B6", "#798777", "#BDC7C9")) +
-  theme(axis.text.x=element_blank(),
-        axis.ticks.x=element_blank(), axis.title.x = element_blank())
-
 # Tables with individual class. For supplementary information if necessary
 # For flakes (complete)
-solutrean_stats_blanks <- solutrean_stats %>% 
-  filter(CLASS %in% c("Flake", "ElongatedProd")) %>%
+solutrean_stats_flakes <- solutrean_stats %>% 
+  filter(CLASS == "Flake") %>%
+  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
+  group_by(RMTYPE, CLASS) %>% 
+  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
+# For elongated blanks (complete)
+solutrean_stats_elong <- solutrean_stats %>% 
+  filter(CLASS == "ElongatedProd") %>%
   select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
@@
     TRUE ~ CLASS))
-
-# Boxplots
-# Width
-proto_stats_width <- proto_stats %>%
-  filter(CLASS %in% c("Flake","ElongatedProd", "Core", "RetouchedPiece")) %>% 
-  filter(MAXWIDTH !="0" & !RMTYPE %in% c("INDET","GB","TL")) %>%
-  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
-                                        "TYPE 5", "TYPE 6","TYPE 7", 
-                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
-  mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) %>%
-  ggplot(aes(RMTYPE, MAXWIDTH, fill = RMTYPE)) +
-  geom_boxplot(aes(fill=factor(RMTYPE)), outlier.shape=21) +
-  facet_wrap(~CLASS) +
-  stat_summary(fun.y=mean, geom="point", shape=23) +
-  labs(y="Width (mm)") +
-  coord_cartesian(ylim =  c(0, 50)) +
-  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
-                              "#FACE7F", "#8B6969", "#CDAF95", "#2B4F60",
-                              "#996666", "#BDD2B6", "#798777", "#BDC7C9")) +
-  theme(axis.text.x=element_blank(),
-        axis.ticks.x=element_blank(), axis.title.x = element_blank())
-# Length
-proto_stats_length <- proto_stats %>%
-  filter(CLASS %in% c("Flake","ElongatedProd", "Core", "RetouchedPiece")) %>% 
-  filter(LENGTH !="0" & !RMTYPE %in% c("INDET","GB","TL")) %>%
-  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
-                                        "TYPE 5", "TYPE 6","TYPE 7", 
-                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
-  mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) %>%
-  ggplot(aes(RMTYPE, LENGTH, fill = RMTYPE)) +
-  geom_boxplot(aes(fill=factor(RMTYPE)), outlier.shape=21) +
-  facet_wrap(~CLASS) +
-  stat_summary(fun.y=mean, geom="point", shape=23) +
-  labs(y="Length (mm)") +
-  coord_cartesian(ylim =  c(0, 60)) +
-  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
-                              "#FACE7F", "#8B6969", "#CDAF95", "#2B4F60",
-                              "#996666", "#BDD2B6", "#798777", "#BDC7C9")) +
-  theme(axis.text.x=element_blank(),
-        axis.ticks.x=element_blank(), axis.title.x = element_blank())
-# Weight
-proto_stats_weight <- proto_stats %>%
-  filter(CLASS %in% c("Flake","ElongatedProd", "Core", "RetouchedPiece")) %>% 
-  filter(WEIGHT !="0" & !RMTYPE %in% c("INDET","GB","TL")) %>%
-  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
-                                        "TYPE 5", "TYPE 6","TYPE 7", 
-                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
-  mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) %>%
-  ggplot(aes(RMTYPE, WEIGHT, fill = RMTYPE)) +
-  geom_boxplot(aes(fill=factor(RMTYPE)), outlier.shape=21) +
-  facet_wrap(~CLASS) +
-  stat_summary(fun.y=mean, geom="point", shape=23) +
-  labs(y="Weight (gr)") +
-  coord_cartesian(ylim =  c(0, 50)) +
-  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
-                              "#FACE7F", "#8B6969", "#CDAF95", "#2B4F60",
-                              "#996666", "#BDD2B6", "#798777", "#BDC7C9")) +
-  theme(axis.text.x=element_blank(),
-        axis.ticks.x=element_blank(), axis.title.x = element_blank())
 
@@
 
-# Boxplots
-# Width
-gravettian_stats_width <- gravettian_stats %>%
-  filter(CLASS %in% c("Flake","ElongatedProd", "Core", "RetouchedPiece")) %>% 
-  filter(MAXWIDTH !="0" & !RMTYPE %in% c("INDET","GB","TL")) %>%
-  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
-                                        "TYPE 5", "TYPE 6","TYPE 7", 
-                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
-  mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) %>%
-  ggplot(aes(RMTYPE, MAXWIDTH, fill = RMTYPE)) +
-  geom_boxplot(aes(fill=factor(RMTYPE)), outlier.shape=21) +
-  facet_wrap(~CLASS) +
-  stat_summary(fun.y=mean, geom="point", shape=23) +
-  labs(y="Width (mm)") +
-  coord_cartesian(ylim =  c(0, 50)) +
-  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
-                              "#FACE7F", "#8B6969", "#CDAF95", "#2B4F60",
-                              "#996666", "#BDD2B6", "#798777", "#BDC7C9")) +
-  theme(axis.text.x=element_blank(),
-        axis.ticks.x=element_blank(), axis.title.x = element_blank())
-# Length
-gravettian_stats_length <- gravettian_stats %>%
-  filter(CLASS %in% c("Flake","ElongatedProd", "Core", "RetouchedPiece")) %>% 
-  filter(LENGTH !="0" & !RMTYPE %in% c("INDET","GB","TL")) %>%
-  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
-                                        "TYPE 5", "TYPE 6","TYPE 7", 
-                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
-  mutate(CLASS = factor(CLASS, levels = c("Flake","ElongatedProd", "Core", "RetouchedPiece"))) %>%
-  ggplot(aes(RMTYPE, LENGTH, fill = RMTYPE)) +
-  geom_boxplot(aes(fill=factor(RMTYPE)), outlier.shape=21) +
-  facet_wrap(~CLASS) +
-  stat_summary(fun.y=mean, geom="point", shape=23) +
-  labs(y="Length (mm)") +
-  coord_cartesian(ylim =  c(10, 60)) +
-  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
-                              "#FACE7F", "#8B6969", "#CDAF95", "#2B4F60",
-                              "#996666", "#BDD2B6", "#798777", "#BDC7C9")) +
-  theme(axis.text.x=element_blank(),
-        axis.ticks.x=element_blank(), axis.title.x = element_blank())
-# Weight
-gravettian_stats_weight <- gravettian_stats %>%
-  filter(CLASS %in% c("Flake","ElongatedProd", "Core", "RetouchedPiece")) %>% 
-  filter(WEIGHT !="0" & !RMTYPE %in% c("INDET","GB","TL")) %>%
-  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
-                                        "TYPE 5", "TYPE 6","TYPE 7", 
-                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
-  mutate(CLASS = factor(CLASS, levels = c("Flake","ElongatedProd", "Core", "RetouchedPiece"))) %>%
-  ggplot(aes(RMTYPE, WEIGHT, fill = RMTYPE)) +
-  geom_boxplot(aes(fill=factor(RMTYPE)), outlier.shape=21) +
-  facet_wrap(~CLASS) +
-  stat_summary(fun.y=mean, geom="point", shape=23) +
-  labs(y="Weight (gr)") +
-  coord_cartesian(ylim =  c(0, 60)) +
-  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
-                              "#FACE7F", "#8B6969", "#CDAF95", "#2B4F60",
-                              "#996666", "#BDD2B6", "#798777", "#BDC7C9")) +
-  theme(axis.text.x=element_blank(),
-        axis.ticks.x=element_blank(), axis.title.x = element_blank())
-
 # Tables with individual class. For supplementary information if necessary
@@
 ```
-
-```{r}
-#| label: fig-solutrean-boxplots
-#| fig-cap: Boxplots Gravettian.
-
-ggpubr::ggarrange(solutrean_stats_width, solutrean_stats_length, solutrean_stats_weight, labels = c("a", "b", "c"), heights = c(5, 5, 5), ncol = 1, common.legend = TRUE, legend = "bottom")
-
-```
-
-```{r}
-#| label:  fig-proto-boxplots
-#| fig-cap: Boxplots Proto-Solutrean
-
-ggpubr::ggarrange(proto_stats_width, proto_stats_length, proto_stats_weight, labels = c("a", "b", "c"), heights = c(5, 5, 5), ncol = 1, common.legend = TRUE, legend = "bottom")
-
-```
-
-```{r}
-#| label: fig-gravettian-boxplots
-#| fig-cap: Boxplots Solutrean
-
-ggpubr::ggarrange(gravettian_stats_width, gravettian_stats_length, gravettian_stats_weight, labels = c("a", "b", "c"), heights = c(5, 5, 5), ncol = 1, common.legend = TRUE, legend = "bottom")
-
-```
-
-## Blank to core ratios
-
-The analysis shows different patterns in the blank to core ratios not only between chert types, but also between assemblages.
-For the Gravettian, local types (T1 and T2) show lower ratios (\<4) of blanks to cores, although other local types (T3-T5) have no identified cores.
-The non-local types, especially T6 and T7 show higher ratios (\~8) of blanks to cores.
-In the Proto-Solutrean occupation, T2 and T4 show the lowest ratios (8.7 and 6.2 respectively) of blanks per cores, while the non-local T7 shows the highest ratios (11.0) of blanks to cores.
-Unlike the Gravettian, T1 (local chalcedony) and T6 (non-local chert) have no cores.
-Finally, the Solutrean assemblage shows similar patterns, with the local cherts showing lower blank to core ratios (\<7), while the non-local cherts show higher ratios (\>9).
-Similar to the Proto-Solutrean, chalcedony cores were also not identified in the studied sample.
-
-```{r}
-#| label: ratio-blankcore
-
-proto_ratiobc <- phase_dataset %>% 
-  filter(Level %in% c("4E", "4E/5")) %>% 
-  filter(RMTYPE != "INDET") %>% 
-  mutate(CLASS = case_when(
-    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
-    TRUE ~ CLASS)) %>% 
-  filter(CLASS %in% c("Blank", "Core")) %>% 
-  count(RMTYPE, CLASS, sort = TRUE) %>% 
-  group_by(RMTYPE, CLASS) %>%
-  summarise(n=sum(n)) %>% 
-  spread(CLASS, n, fill=0) %>% 
-  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
-  filter(!sum <= 5) %>%
-  mutate("Blank/Core" = Blank / Core)
-
-proto5_ratiobc <- phase_dataset %>% 
-  filter(Level == 5) %>% 
-  filter(RMTYPE != "INDET") %>% 
-  mutate(CLASS = case_when(
-    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
-    TRUE ~ CLASS)) %>% 
-  filter(CLASS %in% c("Blank", "Core")) %>% 
-  count(RMTYPE, CLASS, sort = TRUE) %>% 
-  group_by(RMTYPE, CLASS) %>%
-  summarise(n=sum(n)) %>% 
-  spread(CLASS, n, fill=0) %>% 
-  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
-  filter(!sum <= 5) %>% 
-  mutate("Blank/Core" = Blank / Core)
-
-gravettian_ratiobc <- phase_dataset %>% 
-  filter(Level %in% c(6, 7)) %>% 
-  filter(RMTYPE != "INDET") %>% 
-  mutate(CLASS = case_when(
-    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
-    TRUE ~ CLASS)) %>% 
-  filter(CLASS %in% c("Blank", "Core")) %>% 
-  count(RMTYPE, CLASS, sort = TRUE) %>% 
-  group_by(RMTYPE, CLASS) %>%
-  summarise(n=sum(n)) %>% 
-  spread(CLASS, n, fill=0) %>% 
-  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
-  filter(!sum <= 5) %>% 
-  mutate("Blank/Core" = Blank / Core)
-
-solutrean_ratiobc <- solutrean_dataset %>% 
-  mutate(CLASS = case_when(
-    CLASS == "Flake" | CLASS == "ElongBlank" | CLASS == "Thinning flake" ~ "Blank",
-    CLASS == "FlakeFrag" | CLASS == "ElongBlankFrag" ~ "BlankFrag",
-    TRUE ~ CLASS)) %>% 
-  mutate(CLASS = case_when(
-    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
-    TRUE ~ CLASS)) %>% 
-  filter(RMTYPE != "INDET") %>% 
-  filter(CLASS %in% c("Blank", "Core")) %>% 
-  count(RMTYPE, CLASS, sort = TRUE) %>% 
-  group_by(RMTYPE, CLASS) %>%
-  summarise(n=sum(n)) %>% 
-  spread(CLASS, n, fill=0) %>% 
-  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
-  filter(!sum <= 5) %>% 
-  mutate("Blank/Core" = Blank / Core)
-
-```
-
-```{r}
-#| label: fig-ratiobc
-#| fig-cap: Plotted ratios of blanks-to-core by assemblage and by chert type. All local cherts were grouped in a single category "Local", while non-local cherts maintained their individual IDs.
-
-gravettian_ratiobc_b <-gravettian_ratiobc %>%
-  filter(`Blank/Core` != "Inf") %>% 
-  dplyr::mutate("Assemblage" = "Gravettian") %>% 
-  select(Assemblage, RMTYPE, `Blank/Core`)
-proto_ratiobc_b <- proto_ratiobc %>% 
-  filter(`Blank/Core` != "Inf") %>% 
-  dplyr::mutate("Assemblage" = "Proto-Solutrean") %>% 
-  select(Assemblage, RMTYPE, `Blank/Core`)
-solutrean_ratiobc_b <- solutrean_ratiobc %>% 
-  filter(`Blank/Core` != "Inf") %>% 
-  dplyr::mutate("Assemblage" = "Solutrean") %>% 
-  select(Assemblage, RMTYPE, `Blank/Core`)
-ratio_blankcore <- bind_rows(gravettian_ratiobc_b, proto_ratiobc_b,solutrean_ratiobc_b)
-
-# Scatterplot with local vs non-local RM Types
-ratio_blankcore |>
-  dplyr::mutate(Source = case_when(
-    RMTYPE == "TYPE 1" | RMTYPE == "TYPE 2" | RMTYPE == "TYPE 3" | RMTYPE == "TYPE 4" | RMTYPE == "TYPE 5" ~ "Local",
-    RMTYPE == "TYPE 6" ~ "T6",
-    RMTYPE == "TYPE 7" ~ "T7",
-    RMTYPE == "TYPE 8" ~ "T8",
-    RMTYPE == "TYPE 9" ~ "T9",
-    RMTYPE == "TYPE 10" ~ "T10",
-    RMTYPE == "TYPE 11" ~ "T11",
-    TRUE ~ RMTYPE)) |>
-  mutate(Source = factor(Source, levels = c("Local", "T6","T7", 
-                                        "T8", "T9", "T10", "T11", "TL"))) |>
-  ggplot(aes(x=Assemblage, y=`Blank/Core`, colour = Source)) +
-  geom_point(size=3) +
-  coord_cartesian(clip = "off") +
-  scale_colour_manual(values = c("#D49B54", "#8B6969", "#CDAF95", "#2B4F60",
-                              "#8B7D6B", "#BDD2B6", "#798777", "#BDC7C9"))
-```
-
-```{r}
-#| label: tbl-bcratio-gravettian
-#| tbl-cap: Blank to core ratio by raw material type of the Gravettian assemblage (levels 6 and 7) of the Terrace area.
-
-kable(gravettian_ratiobc)
-
-```
-
-```{r}
-#| label: tbl-bcratio-proto
-#| tbl-cap: Blank to core ratio by raw material type of the Proto-Solutrean assemblage (levels 4E and top of level 5) of the Terrace area.
-
-kable(proto_ratiobc)
-
-```
-
-```{r}
-#| label: tbl-bcratio-solutrean
-#| tbl-cap: Blank to core ratio by raw material type of the Solutrean assemblage (levels A to C) of the Shelter area.
-
-kable(solutrean_ratiobc)
-
-```
-
-## Retouched tools to debitage ratios
-
-Throughout the different assemblages, the difference in retouched tools to debitage ratios by chert types, especially between local and non-local types, seem to show similar patterns.
-In this sense, in general, non-local types seem to consistently show the higher ratios when compared to local cherts.
-In the Gravettian assemblages this pattern is the most accentuated since T1 to T5 show low ratios (0.25-0.5) of retouched to debitage, while TL and T6-T8 show higher ratios (0.9-1.0).
-The exception is T10 (non-local type), which shows a lower ratio (0.28), more akin to local cherts.
-In the Proto-Solutrean assemblage, the lowest ratio of retouched to debitage is seen in T1 and T4, with values below 0.1, while other local types show slightly higher values (0.1-0.2).
-Following the same trend as the Gravettian, the non-local cherts show slightly higher ratios (0.3-0.5) of retouched to debitage.
-In the Solutrean assemblage, all local cherts (T2-T5) show low retouch to debitage ratios (\<0.1), with the exception of T1 (chalcedony), which shows a ratio of 1.8.
-Other non-local cherts (T6-T7 and T11) show slightly higher retouch to debitage ratios (0.1-0.3).
-
-```{r}
-#| label: ratios-retouchdebitage
-
-proto_ratiotd <- phase_dataset %>% 
-  filter(Level %in% c("4E", "4E/5")) %>% 
-  filter(RMTYPE != "INDET") %>% 
-  mutate(CLASS = case_when(
-    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
-    CLASS == "RetouchedPieceFrag" & PIECECOMPLETENESS == "Proximal" ~ "RetouchedPiece",
-    TRUE ~ CLASS)) %>% 
-  filter(CLASS %in% c("Blank", "RetouchedPiece")) %>% 
-  count(RMTYPE, CLASS, sort = TRUE) %>% 
-  group_by(RMTYPE, CLASS) %>%
-  summarise(n=sum(n)) %>% 
-  spread(CLASS, n, fill=0) %>% 
-  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
-  filter(!sum <= 5) %>% 
-  mutate("Tool/Blank" = RetouchedPiece / Blank)
-
-proto5_ratiotd <- phase_dataset %>% 
-  filter(Level == 5) %>% 
-  filter(RMTYPE != "INDET") %>% 
-  mutate(CLASS = case_when(
-    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
-    CLASS == "RetouchedPieceFrag" & PIECECOMPLETENESS == "Proximal" ~ "RetouchedPiece",
-    TRUE ~ CLASS)) %>% 
-  filter(CLASS %in% c("Blank", "RetouchedPiece")) %>% 
-  count(RMTYPE, CLASS, sort = TRUE) %>% 
-  group_by(RMTYPE, CLASS) %>%
-  summarise(n=sum(n)) %>% 
-  spread(CLASS, n, fill=0) %>% 
-  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
-  filter(!sum <= 5) %>% 
-  mutate("Tool/Blank" = RetouchedPiece / Blank)
-
-gravettian_ratiotd <- phase_dataset %>% 
-  filter(Level %in% c(6, 7)) %>% 
-  filter(RMTYPE != "INDET") %>% 
-  mutate(CLASS = case_when(
-    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
-    CLASS == "RetouchedPieceFrag" & PIECECOMPLETENESS == "Proximal" ~ "RetouchedPiece",
-    TRUE ~ CLASS)) %>% 
-  filter(CLASS %in% c("Blank", "RetouchedPiece")) %>% 
-  count(RMTYPE, CLASS, sort = TRUE) %>% 
-  group_by(RMTYPE, CLASS) %>%
-  summarise(n=sum(n)) %>% 
-  spread(CLASS, n, fill=0) %>% 
-  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
-  filter(!sum <= 5) %>% 
-  mutate("Tool/Blank" = RetouchedPiece / Blank)
-
-solutrean_ratiotd <- solutrean_dataset %>% 
-  filter(RMTYPE != "INDET") %>% 
-  mutate(CLASS = case_when(
-    CLASS == "Flake" | CLASS == "ElongBlank" | CLASS == "Thinning flake" ~ "Blank",
-    CLASS == "FlakeFrag" | CLASS == "ElongBlankFrag" ~ "BlankFrag",
-    CLASS == "Retouched piece" ~ "RetouchedPiece",
-    CLASS == "Retouched piece fragment" ~ "RetouchedPieceFrag",
-    TRUE ~ CLASS)) %>% 
-  mutate(CLASS = case_when(
-    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
-    CLASS == "RetouchedPieceFrag" & PIECECOMPLETENESS == "Proximal" ~ "RetouchedPiece",
-    TRUE ~ CLASS)) %>% 
-  filter(CLASS %in% c("Blank", "RetouchedPiece")) %>% 
-  count(RMTYPE, CLASS, sort = TRUE) %>% 
-  group_by(RMTYPE, CLASS) %>%
-  summarise(n=sum(n)) %>% 
-  spread(CLASS, n, fill=0) %>% 
-  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
-  filter(!sum <= 5) %>% 
-  mutate("Tool/Blank" = RetouchedPiece / Blank)
-
-```
-
-```{r}
-#| label: fig-ratiotd
-#| fig-cap: Plotted ratios of tools-to-debitage by assemblage and by chert type. All local cherts were grouped in a single category "Local", while non-local cherts maintained their individual IDs.
-
-gravettian_ratiotd_b <-gravettian_ratiotd %>% 
-  dplyr::mutate("Assemblage" = "Gravettian") %>% 
-  select(Assemblage, RMTYPE, `Tool/Blank`)
-proto_ratiotd_b <- proto_ratiotd %>% 
-  dplyr::mutate("Assemblage" = "Proto-Solutrean") %>% 
-  select(Assemblage, RMTYPE, `Tool/Blank`)
-solutrean_ratiotd_b <- solutrean_ratiotd %>% 
-  dplyr::mutate("Assemblage" = "Solutrean") %>% 
-  select(Assemblage, RMTYPE, `Tool/Blank`)
-ratio_tooldebitage <- bind_rows(gravettian_ratiotd_b, proto_ratiotd_b,solutrean_ratiotd_b)
-
-# Scatterplot only with local vs non-local
-#ratio_tooldebitage |>
-#  dplyr::mutate(Source = case_when(
-#    RMTYPE == "TYPE 1" | RMTYPE == "TYPE 2" | RMTYPE == "TYPE 3" | RMTYPE == "TYPE 4" | RMTYPE == "TYPE 5" ~ "Local",
-#    RMTYPE == "TYPE 6" | RMTYPE == "TYPE 7" | RMTYPE == "TYPE 8" | RMTYPE == "TYPE 9" | RMTYPE == "TYPE 10" | RMTYPE == "TYPE 11" | RMTYPE == "TL" ~ "Non-local")) |>
-#  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
-#                                        "TYPE 5", "TYPE 6","TYPE 7", 
-#                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) |>
-#  ggplot(aes(x=Assemblage, y=`Tool/Blank`, colour = Source)) +
-#  geom_point(size=2)
-
-# Scatterplot with local vs non-local RM Types
-ratio_tooldebitage |>
-  dplyr::mutate(Source = case_when(
-    RMTYPE == "TYPE 1" | RMTYPE == "TYPE 2" | RMTYPE == "TYPE 3" | RMTYPE == "TYPE 4" | RMTYPE == "TYPE 5" ~ "Local",
-    RMTYPE == "TYPE 6" ~ "T6",
-    RMTYPE == "TYPE 7" ~ "T7",
-    RMTYPE == "TYPE 8" ~ "T8",
-    RMTYPE == "TYPE 10" ~ "T10",
-    RMTYPE == "TYPE 11" ~ "T11",
-    TRUE ~ RMTYPE)) |>
-  mutate(Source = factor(Source, levels = c("Local", "T6","T7", 
-                                        "T8", "T9", "T10", "T11", "TL"))) |>
-  ggplot(aes(x=Assemblage, y=`Tool/Blank`, colour = Source)) +
-  geom_point(size=3) +
-  scale_colour_manual(values = c("#D49B54", "#8B6969", "#CDAF95", "#2B4F60",
-                              "#8B7D6B", "#BDD2B6", "#798777", "#BDC7C9"))
-
-```
-
-```{r}
-#| label: tbl-tdratio-gravettian
-#| tbl-cap: Tools to debitage ratio by raw material of the Gravettian assemblage (levels 6 and 7) of the Terrace area.
-
-kable(gravettian_ratiotd)
-
-```
-
-```{r}
-#| label: tbl-tdratio-proto
-#| tbl-cap: Tools to debitage ratio by raw material of the Proto-Solutrean assemblage (levels 4E and top of level 5) of the Terrace area.
-
-kable(proto_ratiotd)
-
-```
-
-```{r}
-#| label: tbl-tdratio-solutrean
-#| tbl-cap: Tools to debitage ratio by raw material of the Solutrean assemblage (levels A to C) of the Shelter area.
-
-kable(solutrean_ratiotd)
-
-```
-
-## Tool typology diversity
-
-Solutrean has a total of 43 retouched tools (complete and fragments), with a high variety of retouched tools (11), ranging from 5 (11.6%) to 4 (9.3%) per tool type.
-The typologies which are most frequently present in the studied sample are bifacial blanks and bifacial fragments, amounting to a total of 23.2% of products related to bifacial tool/point production.
-Two types of points were identified: shouldered points (2 complete and 1 fragment, amounting to a total of 6.9%) and tanged and 4 winged points (9.3%).
-Other frequent retouched typologies are endscrappers (9.3%), notches (11.6%), retouched blanks (\~14%) and splintered pieces (11.6%).
-Regarding chert type, a substantial portion of retouched tools are of unidentified chert types (n=13), mostly due to alterations (fire and post depositional).
-Most retouched tools with identified chert are local (n=13, accounting to `r 13*100/43`%), followed by Type 6 cherts, previously suggested to be from Central Portugal (n=7, accounting to `r 7*100/43`%).
-Local types show a higher variability retouched typologies (9 types), although this may be related to the higher amount of tools.
-In comparison, Type 6 shows only 5 different typologies.
-
-For the Proto-Solutrean, 65 chert retouched tools were identified, with `r 15*100/65`% (n=15) having an unidentified chert type.
-A total of `r 1+10+2+8+5` retouched tools correspond to local cherts, while `r 10+9+3+1` correspond to non-local cherts.
-Regarding the typology, 14 different types of retouched tools were identified.
-Endscrappers (26.2%), splintered pieces (17%), burins (12.3%) and retouched flakes (7.7%) were the most prevalent.
-In comparison to the high frequency of point-related technology observed during the Solutrean, Vale Comprido points are barely present in chalcedony and chert (n=1), as they are mostly present in dolerite and chert was mainly used for other tools as seen in previous studies (\*\*).
-Regarding the types of chalcedony and chert, local raw materials show 9 different typologies, with Vale Comprido points, Solutrean retouch and notches being present only in the local types, even if in small numbers.
-Non-local cherts show 10 different typologies of retouched tools.
-Perforator-endscrapper, truncation, backed bladelet and retouched blades typologies can only be found in non-local varieties, albeit also in small numbers.
-
-In the Gravettian assemblage, 170 retouched tools were identified, with 22.9% (n=39) having an unidentified chert type.
-From the identified chert types, 24.1% of retouched tools are from local types and 52.9% from non-local cherts, with 34.1% of these being from T6 and 14.1% of T7.
-Regarding typologies, 15 types of retouched tools were identified (excluding fragments).
-All these typologies were found in the non-local chert group, while only 7 retouched tool typologies were identified in local cherts/chalcedony.
-The most common typologies identified in the assemblage were burins (n=52), endscrappers (n=17), retouched flakes (n=30) and splintered pieces (n=26).
-From all these, the majority were produced in non-local chert, mainly T6 and in lesser numbers T7.
-Although in lesser numbers as well, backed bladelets were also identified (n=6), but only present in non-local chert types (T6 and T7).
-
-```{r}
-#| label: tool-diversity
-
-# Solutrean table
-solutrean_retouch <- solutrean_dataset %>%
-    mutate(CLASS = case_when(
-    CLASS == "Retouched piece" ~ "RetouchedPiece",
-    CLASS == "Retouched piece fragment" ~ "RetouchedPieceFrag",
-    TRUE ~ CLASS)) %>% 
-  filter(CLASS %in% c("RetouchedPiece", "RetouchedPieceFrag")) %>% 
-  select(RMTYPE, RETOUCHEDPIECETYPOLOGY) %>%
-  filter(RETOUCHEDPIECETYPOLOGY != "") %>% 
-  group_by(RMTYPE)
-
-solutrean_retouch_tab <- solutrean_retouch %>% 
-  tabyl(RETOUCHEDPIECETYPOLOGY, RMTYPE) %>% 
-  adorn_totals(c("row", "col")) %>% 
-  adorn_percentages("col") %>% 
-  adorn_pct_formatting(2) %>% 
-  adorn_ns("front")
-
-# Proto-solutrean table
-proto_retouch <- phase_dataset %>% 
-  filter(CLASS %in% c("RetouchedPiece", "RetouchedPieceFrag")) %>%
-  filter(Level %in% c("4E", "4E/5")) %>% 
-  select(RMTYPE, RETOUCHEDPIECETYPOLOGY) %>% 
-  dplyr::group_by(RMTYPE)
-
-proto_retouch_tab <- proto_retouch %>%
-  tabyl(RETOUCHEDPIECETYPOLOGY, RMTYPE) %>%
-  adorn_totals(c("row", "col")) %>% 
-  adorn_percentages("col") %>% 
-  adorn_pct_formatting(2) %>% 
-  adorn_ns("front")
-
-# Gravettian table
-gravettian_retouch <- phase_dataset %>% 
-  filter(CLASS %in% c("RetouchedPiece", "RetouchedPieceFrag")) %>%
-  filter(Level %in% c(6, 7)) %>% 
-  select(RMTYPE, RETOUCHEDPIECETYPOLOGY) %>% 
-  dplyr::group_by(RMTYPE)
-
-gravettian_retouched_tab <- gravettian_retouch %>% 
-  tabyl(RETOUCHEDPIECETYPOLOGY, RMTYPE) %>%
-  adorn_totals(c("row", "col")) %>% 
-  adorn_percentages("row") %>% 
-  adorn_pct_formatting(2) %>% 
-  adorn_ns("front")
-
-```
-
-```{r}
-#| label: tbl-retouch-gravettian
-#| tbl-cap: Number and percentage of retouched tools by raw material of the Gravettian assemblage (levels 6 and 7) of the Terrace area.
-
-kable(gravettian_retouched_tab)
-
-```
-
-```{r}
-#| label: tbl-retouch-proto
-#| tbl-cap: Number and percentage of retouched tools by raw material of the Proto-Solutrean assemblage (levels 4E and top of level 5) of the Terrace area.
-
-kable(proto_retouch_tab)
-
-```
-
-```{r}
-#| label: tbl-retouch-solutrean
-#| tbl-cap: Number and percentage of retouched tools by raw material of the Solutrean assemblage (level A to C) of the Shelter area.
-
-kable(solutrean_retouch_tab)
-
-```
-
-# Discussion
-
-Hypotheses to test: a) use lives derived from knapping strategies; b) reduction intensity and tools-to-debitage ratios; c) core size and blank-to-core ratios; d) reduction intensity and tool reworking; e) tool typology diversity and specialization.
-
-To test these hypotheses, the following parameters were evaluated for each assemblage and compared between chert types: P1) frequency of retouched tools; P2) frequency of maintenance products; P3) debitage sizes; P4) tools-to-debitage ratios ; P5) core sizes; P6) blank-to-core ratios; P7) retouched tool sizes; P8) tool diversity; P9) tool specialization.
-The results may be summarised as follows:
-
-Retouched tools are present in higher percentages in the Gravettian assemblages.
-Within the assemblage, retouched tools (P1) are present in higher frequencies in the non-local cherts, as well as in core preparation and maintenance products (P2).
-Debitage sizes (P3) show no differences between local and non-local cherts.
-Tools-to-debitage ratios (P4) are significantly higher in non-local cherts.
-Core sizes (P5) are slightly smaller in T6, a non-local chert, and the blank-to-core ratios (P6) are higher in non-local cherts.
-Alike blanks, retouched sizes (P7) show no relevant differences between local and non-local cherts; however, tool diversity (P8) is higher in non-local cherts, although with high frequency of less specialised tools such as notches and retouched tools, alongside hunting implements such as backed bladelets (P9).
-The high frequency of burins seen in this assemblage may be related to their use as cores, representing a knapping strategy more than the presence of retouch.
-
-The Proto-Solutrean assemblages show a smaller presence of retouched tools than the Gravettian, and within the assemblage, retouched tools (P1) are slightly more frequent in non-local cherts.
-Core preparation and maintenance products (P2) are only present in non-local cherts, albeit in very small numbers (n=2).
-Debitage sizes (P3), core sizes (P5), and retouched tool sizes (P7) are slightly smaller in non-local cherts.
-Regarding the ratios, tool-to-debitage ratios (P4) are higher in non-local cherts, albeit smaller than those seen in the Gravettian assemblage, and blank-to-core ratios (P6) are also higher in non-local cherts.
-Little differences were observed in tool diversity (P8) and specialisation (P9) between local and non-local cherts.
-
-Finally, the Solutrean shows no relevant differences in the frequencies of retouched tools (P1) and core preparation and maintenance products (P2) between local and non-local cherts.
-In general, the retouch frequency is lower than that observed in the previous assemblages.
-Debitage (P3), core (P5), and retouched tool (P7) sizes show no differences between local and non-local cherts.
-Similarly to the Proto-Solutrean, tools-to-debitage ratios (P4) and blank-to-core ratios (P6) are slightly higher in non-local cherts.
-Both tool diversity (P8) and tool specialization (P9) are higher in local cherts.
-
-How the results fit or do not fit within the hypothesis, and possible reasons why.
-
-Focus on the high presence of retouch and burins during the Gravettian.
-
-Fitting the technological data with the suggested interpretations from article 2.
-
-# Conclusion
-
-# Acknowledgements
-
-<!-- The following line inserts a page break  -->
-
-\newpage
-
-# References
-
-<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
-
-::: {#refs}
-:::
-
-\newpage
-
-### Colophon
-
-This report was generated on `r Sys.time()` using the following computational environment and dependencies:
-
-```{r}
-#| label: colophon
-#| cache: false
-
-# which R packages and versions?
-if ("devtools" %in% installed.packages()) devtools::session_info()
-```

### 2025-01-14 analysis/paper/paper.qmd
Supplementary materials folder with supplementary tables. Removed supplementary tables from the main text, cleaned code

--- 2025-01-14 analysis/paper/paper.qmd
@@
 # Boxplot for individual measurements
-# Length
-solutrean_stats_length <- solutrean_stats %>%
-  filter(CLASS %in% c("Flake", "ElongatedProd", "Core", "RetouchedPiece")) %>% 
-  filter(LENGTH !="0" & !RMTYPE %in% c("INDET","GB","TL")) %>%
-  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
-                                        "TYPE 5", "TYPE 6","TYPE 7", 
-                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
-  mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) %>%
-  ggplot(aes(RMTYPE, LENGTH, fill = RMTYPE)) +
-  geom_boxplot(aes(fill=factor(RMTYPE)), outlier.shape=21) +
-  facet_wrap(~CLASS) +
-  stat_summary(fun.y=mean, geom="point", shape=23) +
-  labs(y="Length (mm)") +
-  #coord_cartesian(ylim =  c(0, 60)) +
-  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
-                              "#FACE7F", "#8B6969", "#CDAF95", "#2B4F60",
-                              "#996666", "#BDD2B6", "#798777", "#BDC7C9")) +
-  theme(axis.text.x=element_blank(),
-        axis.ticks.x=element_blank(), axis.title.x = element_blank())
 # Weight
@@
         axis.ticks.x=element_blank())
-# Width
-solutrean_stats_width <- solutrean_stats %>%
-  filter(CLASS %in% c("Flake", "ElongatedProd", "Core", "RetouchedPiece")) %>% 
-  filter(MAXWIDTH !="0" & !RMTYPE %in% c("INDET","GB","TL")) %>%
-  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
-                                        "TYPE 5", "TYPE 6","TYPE 7", 
-                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
-  mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) %>%
-  ggplot(aes(RMTYPE, MAXWIDTH, fill = RMTYPE)) +
-  geom_boxplot(aes(fill=factor(RMTYPE)), outlier.shape=21) +
-  facet_wrap(~CLASS) +
-  stat_summary(fun.y=mean, geom="point", shape=23) +
-  labs(y="Width (mm)") +
-  #coord_cartesian(ylim =  c(0, 60)) +
-  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
-                              "#FACE7F", "#8B6969", "#CDAF95", "#2B4F60",
-                              "#996666", "#BDD2B6", "#798777", "#BDC7C9")) +
-  theme(axis.text.x=element_blank(),
-        axis.ticks.x=element_blank(), axis.title.x = element_blank())
-
-# Tables with individual class. For supplementary information if necessary
-# For flakes (complete)
-solutrean_stats_blanks <- solutrean_stats %>% 
-  filter(CLASS %in% c("Flake", "ElongatedProd")) %>%
-  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
-  group_by(RMTYPE, CLASS) %>% 
-  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
-# For cores
-solutrean_stats_cores <- solutrean_stats %>% 
-  filter(CLASS == "Core") %>%
-  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
-  group_by(RMTYPE, CLASS) %>% 
-  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
-# For retouched
-solutrean_stats_retouched <- solutrean_stats %>% 
-  filter(CLASS == "RetouchedPiece") %>%
-  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
-  group_by(RMTYPE, CLASS) %>% 
-  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
 
@@
     TRUE ~ CLASS))
-
-# Boxplots
-# Width
-proto_stats_width <- proto_stats %>%
-  filter(CLASS %in% c("Flake","ElongatedProd", "Core", "RetouchedPiece")) %>% 
-  filter(MAXWIDTH !="0" & !RMTYPE %in% c("INDET","GB","TL")) %>%
-  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
-                                        "TYPE 5", "TYPE 6","TYPE 7", 
-                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
-  mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) %>%
-  ggplot(aes(RMTYPE, MAXWIDTH, fill = RMTYPE)) +
-  geom_boxplot(aes(fill=factor(RMTYPE)), outlier.shape=21) +
-  facet_wrap(~CLASS) +
-  stat_summary(fun.y=mean, geom="point", shape=23) +
-  labs(y="Width (mm)") +
-  coord_cartesian(ylim =  c(0, 50)) +
-  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
-                              "#FACE7F", "#8B6969", "#CDAF95", "#2B4F60",
-                              "#996666", "#BDD2B6", "#798777", "#BDC7C9")) +
-  theme(axis.text.x=element_blank(),
-        axis.ticks.x=element_blank(), axis.title.x = element_blank())
 
@@
 
-# Length
-proto_stats_length <- proto_stats %>%
-  filter(CLASS %in% c("Flake","ElongatedProd", "Core", "RetouchedPiece")) %>% 
-  filter(LENGTH !="0" & !RMTYPE %in% c("INDET","GB","TL")) %>%
-  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
-                                        "TYPE 5", "TYPE 6","TYPE 7", 
-                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
-  mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) %>%
-  ggplot(aes(RMTYPE, LENGTH, fill = RMTYPE)) +
-  geom_boxplot(aes(fill=factor(RMTYPE)), outlier.shape=21) +
-  facet_wrap(~CLASS) +
-  stat_summary(fun.y=mean, geom="point", shape=23) +
-  labs(y="Length (mm)") +
-  coord_cartesian(ylim =  c(0, 60)) +
-  scale_fill_manual(values = c("#8B7B8B", "#FFE1FF", "#EED2EE", "#D8BFD8",
-                              "#CDB5CD", "#2B4F60" , "#4A708B", "#798777",
-                            "#9FB6CD", "#BDD2B6")) +
-  theme(axis.text.x=element_blank(),
-        axis.ticks.x=element_blank(), axis.title.x = element_blank())
-# Weight
+# Boxplots weight
 proto_stats_weight <- proto_stats %>%
@@
 
-# Tables with individual class. For supplementary information if necessary
-# For flakes (complete)
-proto_stats_flakes <- proto_stats %>% 
-  filter(CLASS == "Flake") %>%
-  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
-  group_by(RMTYPE, CLASS) %>% 
-  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
-# For elongated blanks (complete)
-proto_stats_elong <- proto_stats %>% 
-  filter(CLASS == "ElongatedProd") %>% 
-  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
-  group_by(RMTYPE, CLASS) %>% 
-  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
-# For cores
-proto_stats_cores <- proto_stats %>% 
-  filter(CLASS == "Core") %>%
-  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
-  group_by(RMTYPE, CLASS) %>% 
-  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
-# For retouched
-proto_stats_retouched <- proto_stats %>% 
-  filter(CLASS == "RetouchedPiece") %>%
-  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
-  group_by(RMTYPE, CLASS) %>% 
-  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
-
 # Gravettian statistics
@@
 
-# Scatterplots width and length
 # Scatterplots for width and length
@@
 # Boxplots
-# Width
-gravettian_stats_width <- gravettian_stats %>%
-  filter(CLASS %in% c("Flake","ElongatedProd", "Core", "RetouchedPiece")) %>% 
-  filter(MAXWIDTH !="0" & !RMTYPE %in% c("INDET","GB","TL")) %>%
-  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
-                                        "TYPE 5", "TYPE 6","TYPE 7", 
-                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
-  mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) %>%
-  ggplot(aes(RMTYPE, MAXWIDTH, fill = RMTYPE)) +
-  geom_boxplot(aes(fill=factor(RMTYPE)), outlier.shape=21) +
-  facet_wrap(~CLASS) +
-  stat_summary(fun.y=mean, geom="point", shape=23) +
-  labs(y="Width (mm)") +
-  coord_cartesian(ylim =  c(0, 50)) +
-  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
-                              "#FACE7F", "#8B6969", "#CDAF95", "#2B4F60",
-                              "#996666", "#BDD2B6", "#798777", "#BDC7C9")) +
-  theme(axis.text.x=element_blank(),
-        axis.ticks.x=element_blank(), axis.title.x = element_blank())
-# Length
-gravettian_stats_length <- gravettian_stats %>%
-  filter(CLASS %in% c("Flake","ElongatedProd", "Core", "RetouchedPiece")) %>% 
-  filter(LENGTH !="0" & !RMTYPE %in% c("INDET","GB","TL")) %>%
-  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
-                                        "TYPE 5", "TYPE 6","TYPE 7", 
-                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
-  mutate(CLASS = factor(CLASS, levels = c("Flake","ElongatedProd", "Core", "RetouchedPiece"))) %>%
-  ggplot(aes(RMTYPE, LENGTH, fill = RMTYPE)) +
-  geom_boxplot(aes(fill=factor(RMTYPE)), outlier.shape=21) +
-  facet_wrap(~CLASS) +
-  stat_summary(fun.y=mean, geom="point", shape=23) +
-  labs(y="Length (mm)") +
-  coord_cartesian(ylim =  c(10, 60)) +
-  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
-                              "#FACE7F", "#8B6969", "#CDAF95", "#2B4F60",
-                              "#996666", "#BDD2B6", "#798777", "#BDC7C9")) +
-  theme(axis.text.x=element_blank(),
-        axis.ticks.x=element_blank(), axis.title.x = element_blank())
 # Weight
@@
 
-# Tables with individual class. For supplementary information if necessary
-# For flakes (complete)
-gravettian_stats_flakes <- gravettian_stats %>% 
-  filter(CLASS == "Flake") %>%
-  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
-  group_by(RMTYPE, CLASS) %>% 
-  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
-# For elongated blanks (complete)
-gravettian_stats_elong <- gravettian_stats %>% 
-  filter(CLASS == "ElongatedProd") %>%
-  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
-  group_by(RMTYPE, CLASS) %>% 
-  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
-# For cores
-gravettian_stats_cores <- gravettian_stats %>% 
-  filter(CLASS == "Core") %>%
-  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
-  group_by(RMTYPE, CLASS) %>% 
-  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
-# For retouched
-gravettian_stats_retouch <- gravettian_stats %>% 
-  filter(CLASS == "RetouchedPiece") %>%
-  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
-  group_by(RMTYPE, CLASS) %>% 
-  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
-
 ```
@@
 
-ggpubr::ggarrange(solutrean_scatter, solutrean_stats_weight, labels = c("a", "b"), ncol = 2, common.legend = TRUE, legend = "bottom")
+solutrean_graphs <- ggpubr::ggarrange(solutrean_scatter, solutrean_stats_weight, labels = c("a1", "a2"), ncol = 2, common.legend = TRUE, legend = "bottom")
 
@@
 
-ggpubr::ggarrange(proto_scatter, proto_stats_weight, labels = c("a", "b"), ncol = 2, common.legend = TRUE, legend = "bottom")
+proto_graphs <- ggpubr::ggarrange(proto_scatter, proto_stats_weight, labels = c("b1", "b2"), ncol = 2, common.legend = TRUE, legend = "bottom")
 
@@
 
-ggpubr::ggarrange(gravettian_scatter, gravettian_stats_weight, labels = c("a", "b"), ncol = 2, common.legend = TRUE, legend = "bottom")
+gravettian_graphs <- ggpubr::ggarrange(gravettian_scatter, gravettian_stats_weight, labels = c("c1", "c2"), ncol = 2, common.legend = TRUE, legend = "bottom")
 
@@
 
-proto_ratiobc <- phase_dataset %>% 
-  filter(Level %in% c("4E", "4E/5")) %>% 
+# Solutrean ratio
+solutrean_ratiobc <- solutrean_dataset %>% 
+  mutate(CLASS = case_when(
+    CLASS == "Flake" | CLASS == "ElongBlank" | CLASS == "Thinning flake" ~ "Blank",
+    CLASS == "FlakeFrag" | CLASS == "ElongBlankFrag" ~ "BlankFrag",
+    TRUE ~ CLASS)) %>% 
+  mutate(CLASS = case_when(
+    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
+    TRUE ~ CLASS)) %>% 
   filter(RMTYPE != "INDET") %>% 
-  mutate(CLASS = case_when(
-    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
-    TRUE ~ CLASS)) %>% 
   filter(CLASS %in% c("Blank", "Core")) %>% 
@@
 
-proto5_ratiobc <- phase_dataset %>% 
-  filter(Level == 5) %>% 
+# Proto-Solutrean ratio
+proto_ratiobc <- phase_dataset %>% 
+  filter(Level %in% c("4E", "4E/5")) %>% 
   filter(RMTYPE != "INDET") %>% 
@@
 
-gravettian_ratiobc <- phase_dataset %>% 
-  filter(Level %in% c(6, 7)) %>% 
+proto5_ratiobc <- phase_dataset %>% 
+  filter(Level == 5) %>% 
   filter(RMTYPE != "INDET") %>% 
@@
 
-solutrean_ratiobc <- solutrean_dataset %>% 
-  mutate(CLASS = case_when(
-    CLASS == "Flake" | CLASS == "ElongBlank" | CLASS == "Thinning flake" ~ "Blank",
-    CLASS == "FlakeFrag" | CLASS == "ElongBlankFrag" ~ "BlankFrag",
-    TRUE ~ CLASS)) %>% 
+# Gravettian ratio
+gravettian_ratiobc <- phase_dataset %>% 
+  filter(Level %in% c(6, 7)) %>% 
+  filter(RMTYPE != "INDET") %>% 
   mutate(CLASS = case_when(
@@
     TRUE ~ CLASS)) %>% 
-  filter(RMTYPE != "INDET") %>% 
   filter(CLASS %in% c("Blank", "Core")) %>% 
@@
                               "#8B7D6B", "#BDD2B6", "#798777", "#BDC7C9"))
-```
-
-```{r}
-#| label: tbl-bcratio-gravettian
-#| tbl-cap: Blank to core ratio by raw material type of the Gravettian assemblage (levels 6 and 7) of the Terrace area.
-
-kable(gravettian_ratiobc)
-
-```
-
-```{r}
-#| label: tbl-bcratio-proto
-#| tbl-cap: Blank to core ratio by raw material type of the Proto-Solutrean assemblage (levels 4E and top of level 5) of the Terrace area.
-
-kable(proto_ratiobc)
-
-```
-
-```{r}
-#| label: tbl-bcratio-solutrean
-#| tbl-cap: Blank to core ratio by raw material type of the Solutrean assemblage (levels A to C) of the Shelter area.
-
-kable(solutrean_ratiobc)
 
@@
 ratio_tooldebitage <- bind_rows(gravettian_ratiotd_b, proto_ratiotd_b,solutrean_ratiotd_b)
-
-# Scatterplot only with local vs non-local
-#ratio_tooldebitage |>
-#  dplyr::mutate(Source = case_when(
-#    RMTYPE == "TYPE 1" | RMTYPE == "TYPE 2" | RMTYPE == "TYPE 3" | RMTYPE == "TYPE 4" | RMTYPE == "TYPE 5" ~ "Local",
-#    RMTYPE == "TYPE 6" | RMTYPE == "TYPE 7" | RMTYPE == "TYPE 8" | RMTYPE == "TYPE 9" | RMTYPE == "TYPE 10" | RMTYPE == "TYPE 11" | RMTYPE == "TL" ~ "Non-local")) |>
-#  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
-#                                        "TYPE 5", "TYPE 6","TYPE 7", 
-#                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) |>
-#  ggplot(aes(x=Assemblage, y=`Tool/Blank`, colour = Source)) +
-#  geom_point(size=2)
 
@@
 
-```{r}
-#| label: tbl-tdratio-gravettian
-#| tbl-cap: Tools to debitage ratio by raw material of the Gravettian assemblage (levels 6 and 7) of the Terrace area.
-
-kable(gravettian_ratiotd)
-
-```
-
-```{r}
-#| label: tbl-tdratio-proto
-#| tbl-cap: Tools to debitage ratio by raw material of the Proto-Solutrean assemblage (levels 4E and top of level 5) of the Terrace area.
-
-kable(proto_ratiotd)
-
-```
-
-```{r}
-#| label: tbl-tdratio-solutrean
-#| tbl-cap: Tools to debitage ratio by raw material of the Solutrean assemblage (levels A to C) of the Shelter area.
-
-kable(solutrean_ratiotd)
-
-```
-
 ## Tool typology diversity
@@
     CLASS == "Retouched piece fragment" ~ "RetouchedPieceFrag",
-    TRUE ~ CLASS)) %>% 
-  filter(CLASS %in% c("RetouchedPiece", "RetouchedPieceFrag")) %>% 
-  select(RMTYPE, RETOUCHEDPIECETYPOLOGY) %>%
-  filter(RETOUCHEDPIECETYPOLOGY != "") %>% 
-  group_by(RMTYPE)
-
-solutrean_retouch_tab <- solutrean_retouch %>% 
-  tabyl(RETOUCHEDPIECETYPOLOGY, RMTYPE) %>% 
-  adorn_totals(c("row", "col")) %>% 
-  adorn_percentages("col") %>% 
-  adorn_pct_formatting(2) %>% 
+    TRUE ~ CLASS)) |>
+  filter(CLASS %in% c("RetouchedPiece", "RetouchedPieceFrag")) |>
+  select(RMTYPE, RETOUCHEDPIECETYPOLOGY) |>
+  filter(RETOUCHEDPIECETYPOLOGY != "") |>
+  dplyr::mutate(RMTYPE = case_when(
+    RMTYPE == "TYPE 1" ~ "T1",
+    RMTYPE == "TYPE 2" ~ "T2",
+    RMTYPE == "TYPE 3" ~ "T3",
+    RMTYPE == "TYPE 4" ~ "T4",
+    RMTYPE == "TYPE 5" ~ "T5",
+    RMTYPE == "TYPE 6" ~ "T6",
+    RMTYPE == "TYPE 7" ~ "T7",
+    RMTYPE == "TYPE 8" ~ "T8",
+    RMTYPE == "TYPE 9" ~ "T9",
+    RMTYPE == "TYPE 10" ~ "T10",
+    RMTYPE == "TYPE 11" ~ "T11",
+    TRUE ~ RMTYPE))
+## Janitor tabyl for final table with n and %
+solutrean_retouch_tab <- solutrean_retouch |>
+  tabyl(RETOUCHEDPIECETYPOLOGY, RMTYPE) |>
+  adorn_totals(c("row", "col")) |>
+  adorn_percentages("col") |>
+  adorn_pct_formatting(2) |> 
   adorn_ns("front")
+## Cleaning final table
+solutrean_retouch_tab <- solutrean_retouch_tab |> # Replace 0% values with -
+  dplyr::mutate(T1 = case_when(T1 == "0   (0.00%)" ~ "-", TRUE ~ T1)) |>
+  dplyr::mutate(T2 = case_when(T2 == "0   (0.00%)" ~ "-", TRUE ~ T2)) |>
+  dplyr::mutate(T5 = case_when(T5 == "0   (0.00%)" ~ "-", TRUE ~ T5)) |>
+  dplyr::mutate(T6 = case_when(T6 == "0   (0.00%)" ~ "-", TRUE ~ T6)) |>
+  dplyr::mutate(T7 = case_when(T7 == "0   (0.00%)" ~ "-", TRUE ~ T7)) |>
+  dplyr::mutate(T8 = case_when(T8 == "0   (0.00%)" ~ "-", TRUE ~ T8)) |>
+  dplyr::mutate(T11 = case_when(T11 == "0   (0.00%)" ~ "-", TRUE ~ T11)) |>
+  dplyr::mutate(INDET = case_when(INDET == "0   (0.00%)" ~ "-", TRUE ~ INDET)) |>
+  select(RETOUCHEDPIECETYPOLOGY, T1, T2, T5, T6, T7, T8, T11, INDET, Total) # Ordering values
 
 # Proto-solutrean table
-proto_retouch <- phase_dataset %>% 
-  filter(CLASS %in% c("RetouchedPiece", "RetouchedPieceFrag")) %>%
-  filter(Level %in% c("4E", "4E/5")) %>% 
-  select(RMTYPE, RETOUCHEDPIECETYPOLOGY) %>% 
-  dplyr::group_by(RMTYPE)
-
+## Table preparation
+proto_retouch <- phase_dataset |>
+  filter(CLASS %in% c("RetouchedPiece", "RetouchedPieceFrag")) |>
+  filter(Level %in% c("4E", "4E/5")) |>
+  select(RMTYPE, RETOUCHEDPIECETYPOLOGY) |>
+  dplyr::mutate(RMTYPE = case_when(
+    RMTYPE == "TYPE 1" ~ "T1",
+    RMTYPE == "TYPE 2" ~ "T2",
+    RMTYPE == "TYPE 3" ~ "T3",
+    RMTYPE == "TYPE 4" ~ "T4",
+    RMTYPE == "TYPE 5" ~ "T5",
+    RMTYPE == "TYPE 6" ~ "T6",
+    RMTYPE == "TYPE 7" ~ "T7",
+    RMTYPE == "TYPE 8" ~ "T8",
+    RMTYPE == "TYPE 9" ~ "T9",
+    RMTYPE == "TYPE 10" ~ "T10",
+    RMTYPE == "TYPE 11" ~ "T11",
+    TRUE ~ RMTYPE))
+## Janitor tabyl for final table with n and %
 proto_retouch_tab <- proto_retouch %>%
@@
   adorn_ns("front")
+## Cleaning final table
+proto_retouch_tab <- proto_retouch_tab |> # Replace 0% values with -
+  dplyr::mutate(T1 = case_when(T1 == "0   (0.00%)" ~ "-", TRUE ~ T1)) |>
+  dplyr::mutate(T2 = case_when(T2 == "0   (0.00%)" ~ "-", TRUE ~ T2)) |>
+  dplyr::mutate(T3 = case_when(T3 == "0   (0.00%)" ~ "-", TRUE ~ T3)) |>
+  dplyr::mutate(T4 = case_when(T4 == "0   (0.00%)" ~ "-", TRUE ~ T4)) |>
+  dplyr::mutate(T5 = case_when(T5 == "0   (0.00%)" ~ "-", TRUE ~ T5)) |>
+  dplyr::mutate(T6 = case_when(T6 == "0   (0.00%)" ~ "-", TRUE ~ T6)) |>
+  dplyr::mutate(T7 = case_when(T7 == "0   (0.00%)" ~ "-", TRUE ~ T7)) |>
+  dplyr::mutate(T8 = case_when(T8 == "0   (0.00%)" ~ "-", TRUE ~ T8)) |>
+  dplyr::mutate(T9 = case_when(T9 == "0   (0.00%)" ~ "-", TRUE ~ T9)) |>
+  dplyr::mutate(TL = case_when(TL == "0   (0.00%)" ~ "-", TRUE ~ TL)) |>
+  dplyr::mutate(INDET = case_when(INDET == "0   (0.00%)" ~ "-", TRUE ~ INDET))
 
 # Gravettian table
+## Table preparation
 gravettian_retouch <- phase_dataset %>% 
@@
   adorn_totals(c("row", "col")) %>% 
-  adorn_percentages("row") %>% 
+  adorn_percentages("col") %>% 
   adorn_pct_formatting(2) %>% 
   adorn_ns("front")
+## Cleaning final table
+gravettian_retouched_tab <- gravettian_retouched_tab |> # Replace 0% values with -
+  dplyr::mutate(T1 = case_when(T1 == "0   (0.00%)" ~ "-", TRUE ~ T1)) |>
+  dplyr::mutate(T2 = case_when(T2 == "0   (0.00%)" ~ "-", TRUE ~ T2)) |>
+  dplyr::mutate(T3 = case_when(T3 == "0   (0.00%)" ~ "-", TRUE ~ T3)) |>
+  dplyr::mutate(T4 = case_when(T4 == "0   (0.00%)" ~ "-", TRUE ~ T4)) |>
+  dplyr::mutate(T5 = case_when(T5 == "0   (0.00%)" ~ "-", TRUE ~ T5)) |>
+  dplyr::mutate(T6 = case_when(T6 == "0   (0.00%)" ~ "-", TRUE ~ T6)) |>
+  dplyr::mutate(T7 = case_when(T7 == "0   (0.00%)" ~ "-", TRUE ~ T7)) |>
+  dplyr::mutate(T8 = case_when(T8 == "0   (0.00%)" ~ "-", TRUE ~ T8)) |>
+  dplyr::mutate(T9 = case_when(T9 == "0   (0.00%)" ~ "-", TRUE ~ T9)) |>
+  dplyr::mutate(T10 = case_when(T10 == "0   (0.00%)" ~ "-", TRUE ~ T10)) |>
+  dplyr::mutate(TL = case_when(TL == "0   (0.00%)" ~ "-", TRUE ~ TL)) |>
+  dplyr::mutate(INDET = case_when(INDET == "0   (0.00%)" ~ "-", TRUE ~ INDET)) |>
+  select(RETOUCHEDPIECETYPOLOGY, T1, T2, T3, T4, T5, T6, T7, T8, T9, T10, TL, INDET, Total) # Ordering values
 

### 2025-01-15 analysis/Supplementary/supplementary_tables.qmd
Supplementary tables improved and organised. WIP render.

@@
   - ../templates/pagebreak.lua
-bibliography: references.bib
+bibliography: ../paper/references.bib
 csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
 ---
+
+```{r}
+#| label: library
+
+library(readr)
+library(dplyr)
+library(tidyr)
+library(stringr)
+library(knitr)
+
+```
 
@@
   mutate(RMTYPE = case_when(
-    RMTYPE == "Type 2" ~ "TYPE 2",
-    RMTYPE == "Chalcedony" ~ "TYPE 1", # Altering first group names to final group names. Subtypes are maintained as before.
-    RMTYPE == "TYPE 15" ~ "TYPE 3",
-    RMTYPE == "TYPE 7" ~ "TYPE 4",
-    RMTYPE == "TYPE 12" ~ "TYPE 5",
-    RMTYPE == "TYPE 11" ~ "TYPE 6",
-    RMTYPE == "TYPE 13" ~ "TYPE 7",
-    RMTYPE == "TYPE 11F" ~ "TYPE 8",
-    RMTYPE == "TYPE 13B" ~ "TYPE 9",
-    RMTYPE == "Oolitic" ~ "TYPE 10",
-    RMTYPE == "TYPE GB" ~ "TYPE 11",
+    RMTYPE == "Type 2" ~ "T2",
+    RMTYPE == "Chalcedony" ~ "T1", # Altering first group names to final group names. Subtypes are maintained as before.
+    RMTYPE == "TYPE 15" ~ "T3",
+    RMTYPE == "TYPE 7" ~ "T4",
+    RMTYPE == "TYPE 12" ~ "T5",
+    RMTYPE == "TYPE 11" ~ "T6",
+    RMTYPE == "TYPE 13" ~ "T7",
+    RMTYPE == "TYPE 11F" ~ "T8",
+    RMTYPE == "TYPE 13B" ~ "T9",
+    RMTYPE == "Oolitic" ~ "T10",
+    RMTYPE == "TYPE GB" ~ "T11",
     .default = RMTYPE)) %>% 
@@
 # Mutate level 5 to add phases: top of 5 (4E/5) and bottom of 5 (5), following Belmiro et al. (2020).
-phase_dataset <- terrace_dataset %>% 
+phase_dataset <- terrace_dataset |>
   mutate(Level = case_when(
@@
     Level == "7C" | Level == 8 ~ "7", # Collapsing samples from Level 7C and Level 8 into 7 to homogenise dataset
-    TRUE ~ Level)) %>% # Adding technocomplex name based on the levels
+    TRUE ~ Level)) |> # Adding technocomplex name based on the levels
   mutate(Technocomplex = case_when(
@@
   mutate(RMTYPE = case_when(
-    RMTYPE == "Type 2" ~ "TYPE 2",
-    RMTYPE == "Chalcedony" ~ "TYPE 1", # Altering first group names to final group names. Subtypes are maintained as before.
-    RMTYPE == "TYPE 15" ~ "TYPE 3",
-    RMTYPE == "TYPE 7" ~ "TYPE 4",
-    RMTYPE == "TYPE 12" ~ "TYPE 5",
-    RMTYPE == "TYPE 11" ~ "TYPE 6",
-    RMTYPE == "TYPE 13" ~ "TYPE 7",
-    RMTYPE == "TYPE 11F" ~ "TYPE 8",
-    RMTYPE == "TYPE 13B" ~ "TYPE 9",
-    RMTYPE == "Oolitic" ~ "TYPE 10",
-    RMTYPE == "TYPE GB" ~ "TYPE 11",
+    RMTYPE == "Type 2" ~ "T2",
+    RMTYPE == "Chalcedony" ~ "T1", # Altering first group names to final group names. Subtypes are maintained as before.
+    RMTYPE == "TYPE 15" ~ "T3",
+    RMTYPE == "TYPE 7" ~ "T4",
+    RMTYPE == "TYPE 12" ~ "T5",
+    RMTYPE == "TYPE 11" ~ "T6",
+    RMTYPE == "TYPE 13" ~ "T7",
+    RMTYPE == "TYPE 11F" ~ "T8",
+    RMTYPE == "TYPE 13B" ~ "T9",
+    RMTYPE == "Oolitic" ~ "T10",
+    RMTYPE == "TYPE GB" ~ "T11",
     RMTYPE == "ND_A" ~ "INDET",
@@
 ```{r}
-#| label: tbl-bcratio-solutrean
-#| tbl-cap: Blank to core ratio by raw material type of the Solutrean assemblage (levels A to C) of the Shelter area.
+#| label: tbl-bcratio
+#| tbl-cap: Blank to core ratio by raw material of the Solutrean (a), Proto-Solutrean (b) and Gravettian (c) assemblages.
+#| tbl-subcap: 
+#|  - "Blank to core ratio by raw material of the Solutrean assemblage (levels A to C) of the Shelter area."
+#|  - "Blank to core ratio by raw material of the Proto-Solutrean assemblage (levels 4E and top of level 5) of the Terrace area."
+#|  - "Blank to core ratio by raw material of the Gravettian assemblage (levels 6 and 7) of the Terrace area."
+#| layout-nrow: 4
 
-kable(solutrean_ratiobc)
-
-```
-
-```{r}
-#| label: tbl-bcratio-proto
-#| tbl-cap: Blank to core ratio by raw material type of the Proto-Solutrean assemblage (levels 4E and top of level 5) of the Terrace area.
-
-kable(proto_ratiobc)
-
-```
-
-```{r}
-#| label: tbl-bcratio-gravettian
-#| tbl-cap: Blank to core ratio by raw material type of the Gravettian assemblage (levels 6 and 7) of the Terrace area.
-
-kable(gravettian_ratiobc)
+kable(head(solutrean_ratiobc))
+kable(head(proto_ratiobc))
+kable(head(gravettian_ratiobc))
 
@@
 #| label: tbl-tdratio-solutrean
-#| tbl-cap: Tools to debitage ratio by raw material of the Solutrean assemblage (levels A to C) of the Shelter area.
+#| tbl-cap: Tools to debitage ratio by raw material of the Solutrean (a), Proto-Solutrean (b) and Gravettian (c) assemblages.
+#| tbl-subcap: 
+#|  - "Tools to debitage ratio by raw material of the Solutrean assemblage (levels A to C) of the Shelter area."
+#|  - "Tools to debitage ratio by raw material of the Proto-Solutrean assemblage (levels 4E and top of level 5) of the Terrace area."
+#|  - "Tools to debitage ratio by raw material of the Gravettian assemblage (levels 6 and 7) of the Terrace area."
+#| layout-nrow: 4
 
-kable(solutrean_ratiotd)
-
-```
-
-```{r}
-#| label: tbl-tdratio-proto
-#| tbl-cap: Tools to debitage ratio by raw material of the Proto-Solutrean assemblage (levels 4E and top of level 5) of the Terrace area.
-
-kable(proto_ratiotd)
-
-```
-
-```{r}
-#| label: tbl-tdratio-gravettian
-#| tbl-cap: Tools to debitage ratio by raw material of the Gravettian assemblage (levels 6 and 7) of the Terrace area.
-
-kable(gravettian_ratiotd)
+kable(head(solutrean_ratiotd))
+kable(head(proto_ratiotd))
+kable(head(gravettian_ratiotd))
 
@@
 # For retouched
-solutrean_stats_retouched <- solutrean_stats %>% 
+solutrean_stats_retouch<- solutrean_stats %>% 
   filter(CLASS == "RetouchedPiece") %>%
@@
 # For retouched
-proto_stats_retouched <- proto_stats %>% 
+proto_stats_retouch <- proto_stats %>% 
   filter(CLASS == "RetouchedPiece") %>%
@@
 ```
+
+```{r}
+#| label: tbl-solutrean
+#| tbl-cap: "Stats for Solutrean occupation."
+#| tbl-subcap: 
+#|  - "Stats for flakes"
+#|  - "Stats for elongated blanks"
+#|  - "Stats for cores"
+#|  - "Stats for retouched"
+#| layout-nrow: 4
+
+kable(head(solutrean_stats_flakes))
+kable(head(solutrean_stats_elong))
+kable(head(solutrean_stats_cores))
+kable(head(solutrean_stats_retouch))
+
+```
+
+```{r}
+#| label: tbl-proto
+#| tbl-cap: "Stats for Proto-Solutrean occupation."
+#| tbl-subcap: 
+#|  - "Stats for flakes"
+#|  - "Stats for elongated blanks"
+#|  - "Stats for cores"
+#|  - "Stats for retouched"
+#| layout-nrow: 4
+
+kable(head(proto_stats_flakes))
+kable(head(proto_stats_elong))
+kable(head(proto_stats_cores))
+kable(head(proto_stats_retouch))
+
+```
+
+```{r}
+#| label: tbl-gravettian
+#| tbl-cap: "Stats for Gravettian occupation."
+#| tbl-subcap: 
+#|  - "Stats for flakes"
+#|  - "Stats for elongated blanks"
+#|  - "Stats for cores"
+#|  - "Stats for retouched"
+#| layout-nrow: 4
+
+kable(head(gravettian_stats_flakes))
+kable(head(gravettian_stats_elong))
+kable(head(gravettian_stats_cores))
+kable(head(gravettian_stats_retouch))
+
+```

### 2025-01-21 analysis/paper/paper.qmd
Introduction maintext draft

@@
 
-# Background
+Lithic technological organisation has long been a central focus in archaeology for the insights it offers into past human behaviour and lifeways.
+Nelson (1991) defines lithic technological organisation as the selection and integration of strategies for the production, use, transportation, discarding, and management of tools and raw materials.
+Kelly (1988) describes it as the interplay of space and time in lithic tool production within a cultural system.
+This includes activities such as raw material sourcing and distribution, tool manufacture, use, reuse, and discard, as well as broader behavioural patterns like mobility and settlement systems.
+Together, these definitions underscore the multifaceted nature of lithic technological organisation and its utility in examining key aspects of hunter-gatherer adaptations over time.
+
+Raw material procurement is a key aspect of lithic technological organisation, as it played a crucial role in the lives of past hunter-gatherer groups by determing the possibilities for tool production and use (Bamforth & Bleed, 1997).
+Approaches focused on the study of raw materials provide key knowledge on several topics of hunter-gatherer organisation such as procurement and mobility strategies (Ambrose, 1990; Binford, 1979, 1985; Gould, 1985; Kuhn, 1991; Mccall, 2007), occupation types and their duration (Kuhn, 2004; Surovell, 2009; Grove, 2023), and social networks and exchanges (Whallon, 2006; Gamble, 1999).
+These approaches focus on the characterisation of geological and archaeological raw materials to identify local and non-local raw materials within an assemblage, their possible sources, and examine their proportions within a site over different occupations and periods.
+
+Similar studies have also been applied to lithic assemblages from prehistoric archaeological sites in the Iberian Peninsula (e.g., Aubry et al.
+2016; Aubry and Igreja 2009; García-Rojas et al.
+2021; Gómez de Soler et al.
+2020; Herrero-Alonso et al.
+2020; Matias 2016; Nocete et al.
+2005; Ortega 2003; Pereira, Andrade, et al.
+2016; Pereira et al.
+2021, 2022; Ramacciotti et al.
+2022; Rodríguez et al.
+2011; Sánchez de la Torre et al.
+2023; Soto 2016), and contributed to understanding how different groups explored the available lithic resources, aided in identifying extensive movements and social networks throughout the Upper Palaeolithic (e.g., Aubry et al.
+2004, 2016;  Corchón Rodríguez et al.
+2016; Hermida et al.
+2016; Marreiros et al.
+2016; Sánchez de la Torre et al.
+2023).
+
+Until recently, the study of raw materials in southwestern Iberia was limited, and knowledge about mobility and social networks was limited to technological studies, derived from the archaeological site of Vale Boi.
+This site is crucial for understanding Upper Paleolithic (UP) adaptations, as it is currently the only site in southwestern Iberia with a long-term occupation spanning most of the UP, enabling the exploration of cultural and technological trends over time (Cascalheira et al.
+2017).
+For example, Vale Boi has provided valuable insights into hunter-gatherer subsistence strategies, technological organisation (Belmiro et al.
+2021; Nuno Bicho et al.
+2013; Cascalheira et al.
+2017; Cascalheira 2019; Horta et al.
+2019; Manne et al.
+2012; Marreiros et al.
+2016).
+This also contributed to understanding different patterns of territory exploitation and contact with other regions of Iberia throughout the UP.
+
+During the Solutrean period, Vale Boi served as a key connection point between central Portugal and southern Spain, facilitating social networks and cultural exchanges (Cascalheira, 2013; Cascalheira et al., 2017).
+Proto-Solutrean evidence also indicates technological similarities and close contact with central Portugal (Belmiro et al., 2021).
+In contrast, studies of Gravettian occupations reveal marked regional differences, suggesting more limited movement of people and information in southwestern Iberia (Marreiros and Bicho, 2013).
+
+Recent research has expanded our understanding of mobility and social networks through lithic raw material studies.
+Our team’s recent work (Belmiro et al., 2025) identified non-local cherts originating from central Portugal (>250 km from Vale Boi) and souuthern Spain (>200 km from Vale Boi).
+The proportions of these materials varied across the UP, with earlier occupations (Gravettian) showing high frequencies of non-local chert, while later ones (Proto-Solutrean and Solutrean) exhibited a marked decline.
+These findings suggest the establishment and maintenance of extensive social networks connecting southwestern Portugal to Southern spain and central Portugal as early as the beginning of the UP.
+
+These results, particularly those related to the Gravettian, were interpreted as outcomes of residential mobility rather than direct social network influences, though further investigation is required.
+While distance to source and raw material proportions offer valuable insights, they represent only one aspect of raw material procurement and management.
+A raw material-focused approach alone is insufficient for comprehensively understanding mobility patterns and technological organisation.
+Indeed, numerous studies and models from both ethnographic and archaeological perspectives have demonstrated the complex interplay between lithic technological variability, raw material provisioning, and mobility.
+
+For example, technological variability within a lithic assemblage may be explained by the abundance and quality of raw materials (Andrefsky, 1994; Brown, 1999; Oestmo, 2017).
+Whenever raw materials in proximity to a residential site are abundant and of good quality, a high percentage of formal tools are produced (Andrefsky, 1994; Perlès, 1992), and artefacts show less reduction intensity and decreased conversation through retouch (Thacker, 2002).
+Whenever local raw materials are scarce or of low quality, non-local resources are procured to produce mostly formal tools (Andrefsky, 1994; Perlès, 1992).
+Similarly, when groups move to areas where raw material is scarcer or of poorer quality, more effort is placed into the preparation of transportable cores and the production of tools (Kelly, 1988).
+
+However, the different patterns between local and non-local raw material proportion and use in lithic assemblages may be less obvious whenever prehistoric groups have easy access to non-local raw materials due to high mobility or exchanges between groups (Akerman et al., 2002; Torrence, 1996; Brown, 1999; Oestmo).
+Furthermore, Roth and Dibble (1998) demonstrate that even in areas where good quality raw material is abundant, non-local raw materials were still transported to the site, often in the form of large blanks or trimmed cores, and both local and non-local raw materials were retouched into tools.
+
+Tools, flakes and cores often functioned as portable sources of raw material within mobile toolkits or during non-local raw material procurement, enabling stone knapping during periods of limited time or in environments with scarce or unsuitable materials (Kelly, 1988; Morrow, 1996; Gould et al., 1971; Kuhn, 1994; Perlès, 1992; Shott, 1986).
+For example, the transport of formal cores (e.g., blade cores) and bifaces used as cores have been interpreted as lightweight and reconfigurable raw material storage strategies (Clark, 1987; Kelly, 1988).
+
+Formal tools and the degree of tool maintenance have also been associated with planned preparation, efficient use, and transportability, reflecting their connection to mobile settlement strategies, short-term site occupations (Torrence, 1983) and the regional distribution of raw materials (Bamforth, 1986).
+For example, non-local raw materials, transported as part of personal toolkits (frequently associated with residential mobility), would primarily arrive at sites as retouched pieces (Kuhn, 2004).
+These artefacts would show higher degrees of reduction and reworking (Kuhn, 2004) and a higher percentage of non-cortical flakes (Roth and Dibble 1998), as they would have been in use for an extended period before being discarded.
+When knapped on-site, non-local materials also tend to exhibit higher proportions of retouched tools compared to local materials.
+
+Tool variability and size are also impacted by mobility and uncertainty regarding raw material availability; increased group mobility may create a pattern of limited tool variability, and tools may become less specialised and easily reconfigurable, such as retouched blanks (Shott, 1986; Odell, 1981; Siegel, 1984).
+Increased mobility also often results in smaller tools (Shott, 1986), although other authors have noted that large tools may be well-suited to a mobile lifestyle due to their extended use-life, allowing them to remain functional in areas where raw materials are scarce or unsuitable for knapping (Morrow, 1996).
+
+The proportions of local vs non-local raw materials discussed thus far have focused greatly on group mobility, which is linked with residential mobility and occupation duration.
+Residential mobility (e.g., foraging range size and frequency of residential moves) is a defining characteristic of hunter-gatherer societies and plays a central role in their lifeways (Kelly, 1995).
+It is a key factor driving changes in lithic assemblages (Brown, 1999; Oestmo, 2017; Roth & Dibble, 1998) as it significantly impacts the cost and use of artefacts and raw materials since they affect the relation between the distance of raw material sources to sites (Kuhn, 2004).
+As such, different raw material and technological patterns have been observed between short-term and long-term occupations.
+
+Short-term occupations typically rely on transported (non-local) toolkits (Torrence, 1983) and raw materials (Surovell, 2009), show higher reduction intensity, and an increased tools-to-debitage ratio in the assemblages (Grove et al., 2023).
+Long-term occupations, typically rely on provisioning sites with locally available and suitable raw materials, while non-local materials are gradually used and discarded (Grove et al., 2023; Surovell, 2009).
+Non-local raw material cores are highly reduced and extensively exploited (Kuhn, 2004; Roth & Dibble, 1998) and exhibit higher blank-to-core ratios (Roth & Dibble, 1998).
+
+These studies highlight significant variability in lithic assemblages and raw material proportions, reflecting different facets of hunter-gatherer technological organisation.
+This variability appears between behaviourally and ethnically similar sites (Bamforth, 1986), across distinct chrono-cultural periods, and even within a single site among various raw materials.
+To fully understand past hunter-gatherer behaviours, organisation, and lifeways, it is crucial to examine all stages of lithic production—procurement, distribution, use, maintenance, and discard—through interconnected approaches that combine lithic technology and raw material analyses.
+
+Building on the studies mentioned above and existing (yet currently uncorrelated) data on UP lithic technology and raw material analyses at Vale Boi, we outline the expectations and objectives for the current study.
 
@@
 
-# Methods
+## Site descriptions
+
+Vale Boi is situated on the western coast of the Algarve region in southern Portugal, within a small valley that stretches southward to the Atlantic coast, approximately 2 km away.
+The site is bordered by limestone outcrops, which form rock shelters facing west and southwest.
+Covering more than 10,000 square meters along the valley slope, Vale Boi includes three main excavated areas: Slope, Terrace, and Shelter, which were excavated between 2000 and 2019 (Bicho et al., 2007, 2012, 2013; Cascalheira et al., 2017; Cascalheira & Bicho, 2013; Manne et al., 2012).
+Based on the available raw material and technological data, particularly the macroscopic and petrographic research (Belmiro et al., 2025), the Terrace and Shelter areas were selected for the current study.
+
+The Terrace spans occupations from the UP to the Early Neolithic.
+Although excavations began in 2000, a new 8-square-meter area was opened in 2012 to better understand the stratigraphic sequence and assess the potential presence of older cultural layers, which could be linked to an early Upper Paleolithic occupation.
+Between 2012 and 2016, eight primary litho-stratigraphic units were identified.
+The UP sequence includes multiple occupations, spanning ~10,000 years: Gravettian (levels 8 to 6) between approximately c.
+32 and 27 ka cal BP, Proto-Solutrean (levels 5 to 4E) between approximately c.
+26 and 24 ka cal BP, and Solutrean (levels 4D, 4C, 4, and Lower 3) between approximately c.
+24 and 20 ka cal BP (Belmiro et al., 2021; Cascalheira et al., 2017; Cascalheira & Bicho, 2013).
+
+The Shelter area contains four main lithostratigraphic units, with occupations from the Magdalenian, Solutrean, and Gravettian periods.
+The Magdalenian and Gravettian levels (Z and D, respectively) were partially eroded, while the Solutrean levels represent the most intensive occupation, with three well-preserved archaeological horizons (layers C to A), dated to approximately c.
+24-22 ka cal BP.
+These occupation levels were found beneath blocks of limestone that likely collapsed from the rockshelter ceiling, probably shortly after the Last Glacial Maximum (Cascalheira, 2010; Cascalheira et al., 2017).
+
+Most of these occupations resulted in large lithic assemblages, encompassing all lithic production stages and use of several raw materials, often for specific uses.
+Research on the lithic assemblages of the site have allowed the technological and cultural characterisation of human groups occupying the site, which show some differences through the different UP occupations.
+Given the chosen sample for this study (see section 2 “Materials and methods”), we summarise here the results for the technological analysis of the Solutrean assemblages (levels A-C) of the Shelter (Cascalheira, 2010; Cascalheira et al., 2013) and the Proto-Solutrean assemblages (levels 4E-5) of the Terrace (Belmiro, 2020; Belmiro et al., 2021).
+The Gravettian occupations (Levels 6-8) of the Terrace have been recently studied, but the results remain unpublished.
+
+The lithic analysis of the Shelter Solutrean levels (A-C) shows a high number of lithic artifacts, with over 22 000 artifacts analysed (including chips which correspond to ~60% of the assemblage and shatter ~17-19%).
+The volumes introduced into the site were mostly cortical, given the high frequencies of cortex in blanks and retouched tools.
+The knapping strategies are dominated by unidirectional and bidirectional prismatic, with frequent core maintenance products.
+Flakes dominate the debitage assemblage, and endscrapers, notches, splintered pieces and retouched blades/flakes are the most common retouched tools.
+Stemmed and winged points, laurel leaves, shouldered points and parpalló points were also identified (Cascalheira et al., 2013).
+
+The Proto-Solutrean horizons yielded more than 26,500 artifacts (including chips which correspond to ~65-70% of the assemblage and shatter ~20%), ~11,000 attributed to the lower bottom of level 5 and ~15,600 attributed to the top of level 5 and level 4E.
+Both horizons are characterised by knapping strategies dominated by unidirectional prismatic cores, with little platform preparation, focused on the obtention of bladelets (in bottom level 5) and flakes.
+Retouched tools are dominated by splintered pieces, endscrapers and retouched blanks, with Vale Comprido points (the Proto-Solutrean fossil director) being barely inexistent in chert and mostly produced in dolerite (Belmiro et al., 2021).
+
+Previous studies showed that the main raw materials used at the site are chert, quartz and greywacke, with the percentages of these changing between occupations.
+For example, while studies show a high degree of quartz in the lower Proto-Solutrean levels (Belmiro et al., 2021), in the Solutrean occupations of the Shelter a dominance of chert was observed (Cascalheira, 2010).
+Similarly, different raw materials have been observed to play specific roles, such as the case of chert for the production of a high percentage of blanks and retouched tools.
+While quartz and greywacke can be found locally, in the streams and adjacent to the site, chert sources are further away, although still within a ~20 km radius of the site.
+Our previous study showed, as expected, the existence of local chert in the assemblages, originating from the Lower Jurassic outcrops in proximity to the site (Belmiro et al., 2025).
+As previously mentioned, a portion of non-local cherts was also identified, with some identified sources, such as the Cretaceous outcrops of Central Portugal (Lisbon and Rio Maior areas) and the Upper Jurassic oolitic/peloidal cherts of the Betic Systems in southern Spain.
+This suggests contacts or long-range mobility patterns of the groups occupying Vale Boi all throughout the UP, although in different degrees, possibly related to changes in the settlement patterns and duration of site occupation.
+
+Both these areas have been previously interpreted as seasonal residential camps, repeatedly used for extended stays, due to the abundance of lithic debitage, stone tools, heat-cracked rocks related to grease rendering activities, large quantities of faunal remains (both marine and terrestrial) and the presence of ornaments and portable art [@manneIntensiveSubsistencePractices2012].
+This is true for most of the occupations (levels), with exception of the Early Gravettian occupations of the Terrace area and Magdalenian occupation of the Shelter.
+
+Using the Whole Assemblage Behavioral Index (WABI), previously used in other studies (Riel-Salvatore, 2010; Riel-Salvatore & Barton, 2004), Cascalheira et al., (2017) used the lithic results from the Shelter Solutrean levels (Cascalheira, 2010; Cascalheira et al., 2013) and Magdalenian levels (Mendonça, 2009), and the Gravettian levels of Terrace (Marreiros et al., 2015) to reconstruct and compare the mobility strategies of the several occupations, through the lithic volumetric density and the frequency of retouched tools.
+This method allowed the confirmation that, except from the Early Gravettian and Magdalenian occupations, all other occupations are composed of high-density assemblages with low degree of retouch, correspondent to a residential base-camp occupation, for extended periods of time.
+Regarding the Terrace area, however, the results were based on previous excavations (<2012), and the identification of new layers and reorganization of the occupations is yet to be tested through the use of lithic techno-typological methods.
+
+A similar conclusion was suggested based on the interpretation of the chert sourcing and proportions in the several UP occupations of the Terrace and Shelter of Vale Boi, and following other models relating local/non-local raw materials with mobility (Grove et al., 2023; Surovell, 2009).
+The high proportions of non-local chert (~50%) in the Gravettian occupations (levels 6 and 7), and gradual reduction in the following occupations, were interpreted as a possible result of settlement changes, where the Gravettian occupations corresponded to short-term occupations, and the other occupations corresponded to residential, long-term occupations with more infrequent moves (Belmiro et al., 2025).
+
+# Materials and methods
 

### 2025-01-21 analysis/paper/paper.qmd
Results restructured (by assemblage instead of by analysis), and code cleaned/organised

@@
 
+# Read libraries necessary for running the script
 library(readr)
@@
 
-# Technological data
+# Reading the datasets
+## Technological data
 gravettian_tech <- read.csv(here::here('analysis/data/raw_data/VBT_6_7_JB.csv'))
@@
 
-# RM data
+## RM data
 shelter_RM <- read_csv(here::here('analysis/data/raw_data/MACRO_IND_SHELTER.csv')) # Shelter RM analysis
@@
 
-# Context data
+## Context data
 context <- read_csv(here::here('analysis/data/raw_data/VB_Terrace_context.csv')) # Terrace Field DB - lithic artefact info
@@
 
-cols_to_concat_terrace <- c("UNIT", "ID")
+# Setting up the Terrace dataset
+cols_to_concat_terrace <- c("UNIT", "ID") # Vector to join UNIT and ID into a single variable
 context$ID <- str_squish(context$ID) # Remove empty spaces from ID field
@@
 
+# Cleaning up DS and uniformising variable names
 proto_tech_WIP <- proto_tech %>%
@@
 
-proto_original_chert <- proto_tech %>% 
-  filter(RawMaterial %in% c("Chert", "Chalcedony"))
-
 gravettian_tech <- gravettian_tech %>% 
@@
 
-terrace_tech <- bind_rows(proto_tech_WIP, gravettian_tech)
+terrace_tech <- bind_rows(proto_tech_WIP, gravettian_tech) # Creating single DS with Proto and Gravettian
 
@@
 
+# Setting up the Terrace dataset
 cols_to_concat_shelter <- c("YEAR", "ID") # Concatenate YEAR and ID variables in one single variable (e.g. 05-500)
@@
 
-solutrean_points <- solutrean_points %>% # Creating ID from 
+solutrean_points <- solutrean_points %>% 
   mutate_at("YEAR", str_replace, "20", "") %>% # Removing initial part of the year to match IDs
@@
 
-solutrean_general <- solutrean_dataset %>%
+# Tables with class by raw material type (n and %)
+## For Gravettian
+gravettian_general <- phase_dataset |>
+  filter(Level %in% c(6, 7)) |>
   mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
                                         "TYPE 5", "TYPE 6","TYPE 7", 
-                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL", "INDET"))) %>% 
-  tabyl(CLASS, RMTYPE) %>%
-  adorn_totals(c("row", "col")) %>%
-  adorn_percentages("col") %>%
-  adorn_pct_formatting(2) %>%
+                                        "TYPE 8", "TYPE 9", "TYPE 10", "TL", "INDET"))) |>
+  tabyl(CLASS, RMTYPE) |> 
+  adorn_totals(c("row", "col")) |> 
+  adorn_percentages("col") |> 
+  adorn_pct_formatting(2) |> 
   adorn_ns("front")
 
-proto_general <- phase_dataset %>% 
-  filter(Level %in% c("4E", "4E/5")) %>% 
-  filter(CLASS != "Chip") %>% 
+## For Proto-Solutrean (4E and top of 5)
+proto_general <- phase_dataset |> 
+  filter(Level %in% c("4E", "4E/5")) |> 
+  filter(CLASS != "Chip") |> 
   mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
                                         "TYPE 5", "TYPE 6","TYPE 7", 
-                                        "TYPE 8", "TYPE 9", "TYPE 10", "TL", "INDET"))) %>% 
-  tabyl(CLASS, RMTYPE) %>%
-  adorn_totals(c("row", "col")) %>%
-  adorn_percentages("col") %>%
-  adorn_pct_formatting(2) %>%
+                                        "TYPE 8", "TYPE 9", "TYPE 10", "TL", "INDET"))) |> 
+  tabyl(CLASS, RMTYPE) |> 
+  adorn_totals(c("row", "col")) |> 
+  adorn_percentages("col") |> 
+  adorn_pct_formatting(2) |> 
   adorn_ns("front")
 
-gravettian_general <- phase_dataset %>% 
-  filter(Level %in% c(6, 7)) %>%
+## For Solutrean
+solutrean_general <- solutrean_dataset |> 
   mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
                                         "TYPE 5", "TYPE 6","TYPE 7", 
-                                        "TYPE 8", "TYPE 9", "TYPE 10", "TL", "INDET"))) %>% 
-  tabyl(CLASS, RMTYPE) %>%
-  adorn_totals(c("row", "col")) %>%
-  adorn_percentages("col") %>%
-  adorn_pct_formatting(2) %>%
+                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL", "INDET"))) |> 
+  tabyl(CLASS, RMTYPE) |> 
+  adorn_totals(c("row", "col")) |> 
+  adorn_percentages("col") |> 
+  adorn_pct_formatting(2) |> 
   adorn_ns("front")
@@
 ```{r}
-#| label: tbl-general-gravettian
-#| tbl-cap: Number and percentage of technotypological classes by raw material type from the Gravettian assemblage (levels 6 and 7) of the Terrace area of Vale Boi.
-
-kable(gravettian_general)
-
-```
-
-```{r}
-#| label: tbl-general-proto
-#| tbl-cap: Number and percentage of technotypological classes by raw material type from the Proto-Solutrean assemblage (levels 4E and top of level 5) of the Terrace area of Vale Boi.
-
-kable(proto_general)
-
-```
-
-```{r}
-#| label: tbl-general-solutrean
-#| tbl-cap: Number and percentage of technotypological classes by raw material type from the Solutrean assemblage (levels A to C) of the Shelter area of Vale Boi.
-
-kable(solutrean_general)
-
-```
-
-```{r}
-#| label: raw-material-graph
-#| include: false
-
-library(ggplot2)
-library(scales)
-
-solutrean_dataset %>%
-  filter(RMTYPE != "INDET") %>% 
-  dplyr::group_by(LEVEL, RMTYPE) %>% 
-  dplyr::summarise(n = n()) %>%
-  mutate(percent = n/sum(n)) %>% 
-  ggplot(aes(x = LEVEL, y = percent, fill = RMTYPE)) +
-  geom_bar(stat = "identity", width = 0.5) +
-  scale_y_continuous(labels = scales::percent_format()) +
-  #theme_minimal() + #type of theme
-  labs(y = "Count %") +# change labels
-  guides(fill=guide_legend(title=NULL)) + # remove legend title
-  geom_text(aes(label=ifelse(percent >= 0.05, paste0(sprintf("%.1f", percent*100),"%"),"")),
-                position=position_stack(vjust=0.5), colour="white", size=3) +
-  theme(axis.title.x = element_blank()) + 
-  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
-                              "#FACE7F", "#8B6969", "#CDAF95", "#2B4F60",
-                              "#996666", "#BDD2B6", "#798777", "#BDC7C9"))
-
-```
-
-## Assemblage description
-
-For the Gravettian assemblage, local cherts show similar patterns, with T2 and T5 being the types with higher numbers of artefacts (n=110 and n=72, respectively).
-For all local cherts, blanks represent more than 50% of the artefacts.
-Cores and core maintenance products are only present in T2 and T5, although in small frequencies.
-For T2, cores represent 13.6% of the group and core preparation and maintenance products represent 1.8% (n=2).
-For T5 only 4 core fragments and 1 core preparation and maintenance product are present.
-Retouched tools seem to be especially relevant in T2 (\~18%), being the most frequent class after blanks.
-The presence of burin spalls (8.1%) in T2 also hints at the production of burins.
-In comparison, T5 shows similar percentages of retouched tools and shatter (\~12% and \~14%, respectively).
-T1 (chalcedony) shows similar patterns, although with higher percentages of shatter (34.3%).
-In comparison, non-local chert types (especially T6 with n=169 and T7 with n=86) show similar percentages of blanks (\~40-50%) but higher percentages of retouched tools (\~30%).
-Cores, core fragments and core preparation and maintenance products are also present even if in small percentages (\<4%), as well as shatter (\~12%).
-In both types burin spalls are also present (7.1% for T6 and 4.6% for T7).
-The TL cherts show, as expected, blanks and blank fragments (n=3 and n=2, respectively), and retouched tools (n=4).
-The presence of core maintenance products in TL is possibly related to the limited knapping of TL01 (jasper).
-
-Regarding the Proto-Solutrean assemblage, the local types show high frequency of blanks (\>50%, including blank fragments), followed by shatter (\~11-27%), cores (\~9-12%) and retouched tools (\~5-15% ).
-A single burin spall was identified in T4.
-There is also the absence of core preparation and maintenance products across all local cherts.
-T1 (chalcedony) is composed mostly of blanks and a retouched tool (Vale Comprido point).
-The absence of cores but presence of shatter may hint at the knapping of this raw material at the site, but the absence of formal cores perhaps due to the quality of nodules/lenses, which are characterised by fractures.
-For non-local types, especially those with higher number of artefacts (T6-T8), blanks and blank fragments are the most frequent class (`r 38.3+10`%, `r 40.4+12.7`% and `r 46.1+23.1`%, respectively).
-For T6 the absence of cores but the presence of shatter and a single core preparation and maintenance product suggests knapping on site.
-Similarly, T7 shows the presence of cores and core fragments (`r 4.2+4.2`%), shatter (17.0%) and a single core preparation and maintenance product.
-In comparison, T8, despite more limited in number (n=13), shows only one core, with the rest of the artefacts being blanks (as previously noted) or retouched tools (23.1% and n=3).
-Aside from the TL artefacts, all non-local cores show at least 1 core and frequently the presence of shatter.
-As expected, the TL artifacts include a blank and a retouched.
-However, a shatter was also identified.
-
-The Solutrean assemblages show similar patterns to those observed in previous assemblages.
-For local cherts, especially T2 (n=506) which accounts for the majority of local chert present in the studied samples, the majority of the debitage is composed of flakes and elongated blanks (`r 42+13.4+17.8+10.7`%).
-Cores are also present (10.5%), alongside a small percentage of core preparation and maintenance products (1.3%).
-In comparison to debitage products, retouched tools are scarce in the sample (\~3%).
-The non-local cherts most present in the Solutrean of the Shelter area are T6 (n=86) and T8 (n=30).
-These show similar patterns to the local cherts, with high percentages of blanks (\>50%).
-Cores and core preparation and maintenance are also present in small percentages (\<10%), as well as retouched tools (\<10%).
-However, these specific results are truncated by the lack of shatter, which were not present in the sample used for the study.
-It is important to notice, however, that alike previous results from the Solutrean assemblages of the Terrace area, a large portion of the studied sample was altered and did not allow for a chert type attribution (`r 201*100/914`%).
-
-## Measurement and weight descriptive statistics
-
-In the Gravettian assemblages, blanks and retouched tools show little differences in the measurements of artefacts of local and non-local chert types.
-The only exception are T6 and T7 blanks, which show slightly lower weight means and lower maximum length (\~35 mm) than the other types.
-For cores, T2 shows bigger (heavier) values than T6 (means of 21.3 and 12.6, and maximum weight of 37.2 gr and 22.5 gr, respectively), even if they seem to vary in morphology, with little patterns regarding thickness, width and length.
-
-In the Proto-Solutrean assemblage, blanks show lower values for length and weight in non-local cherts (T6-T8), varying between a mean of \~19-20 mm and maximum of \<30 mm for length, and mean of \<3 gr and maximum of \<7 gr for weight.
-In retouched tools, weight also shows smaller means for non-local cherts (\~3-2 gr), although with higher standard deviation (\~1.6).
-Regarding cores, the sample size for non-local cherts is small and only T7 statistics could be obtained.
-Despite this, when compared to local cherts (T2-T5), T7 cores show smaller mean and maximum values (7.3 gr and 8 gr, respectively).
-
-For the Solutrean assemblage, there seems to be no clear difference in the measurements and weights of artefacts between local and non-local cherts.
-This is true especially for the most represented classes (flakes and elongated blanks).
-Regarding cores, many of the local and non-local cherts are not available for the statistics due to missing data.
-However, the smallest weight mean seem to be for cores of T6 (16.5 gr), although with high standard deviation and maximum value (16.5 and 48.6 gr, respectively).
-
-When comparing between technocomplexes, there are little differences in the general size and weight of the artifacts.
-For example, for key classes where size and weight differences could be expected (e.g., cores), the length measurements are mostly all located between 20-40 mm and weight between 10-40 gr.
-However, a few key differences were noted.
-T4 cores (the chert type most frequent in the Proto-Solutrean assemblage) shows the largest weights of cores of all assemblages, both in the Solutrean and Proto-Solutrean assemblage, as well as T5 in the Solutrean assemblage.
-Similarly, retouched tools show weights between 0.5-10 gr in all assemblages.
-T6 and T7 have consistently the smaller weights, especially compared to T2 during the Proto-Solutrean and Gravettian which shows slightly higher weights.
-
-```{r}
 #| label: stats-setup
+
+# Scatterplots for width and length and Boxplots for weight by assemblage
 
@@
 ```
-
-```{r}
-#| label: fig-solutrean-boxplots
-#| fig-cap: Boxplots Gravettian.
-
-solutrean_graphs <- ggpubr::ggarrange(solutrean_scatter, solutrean_stats_weight, labels = c("a1", "a2"), ncol = 2, common.legend = TRUE, legend = "bottom")
-
-```
-
-```{r}
-#| label:  fig-proto-boxplots
-#| fig-cap: Boxplots Proto-Solutrean
-
-proto_graphs <- ggpubr::ggarrange(proto_scatter, proto_stats_weight, labels = c("b1", "b2"), ncol = 2, common.legend = TRUE, legend = "bottom")
-
-```
-
-```{r}
-#| label: fig-gravettian-boxplots
-#| fig-cap: Boxplots Solutrean
-
-gravettian_graphs <- ggpubr::ggarrange(gravettian_scatter, gravettian_stats_weight, labels = c("c1", "c2"), ncol = 2, common.legend = TRUE, legend = "bottom")
-
-```
-
-## Blank to core ratios
-
-The analysis shows different patterns in the blank to core ratios not only between chert types, but also between assemblages.
-For the Gravettian, local types (T1 and T2) show lower ratios (\<4) of blanks to cores, although other local types (T3-T5) have no identified cores.
-The non-local types, especially T6 and T7 show higher ratios (\~8) of blanks to cores.
-In the Proto-Solutrean occupation, T2 and T4 show the lowest ratios (8.7 and 6.2 respectively) of blanks per cores, while the non-local T7 shows the highest ratios (11.0) of blanks to cores.
-Unlike the Gravettian, T1 (local chalcedony) and T6 (non-local chert) have no cores.
-Finally, the Solutrean assemblage shows similar patterns, with the local cherts showing lower blank to core ratios (\<7), while the non-local cherts show higher ratios (\>9).
-Similar to the Proto-Solutrean, chalcedony cores were also not identified in the studied sample.
-
-```{r}
-#| label: ratio-blankcore
-
-# Solutrean ratio
-solutrean_ratiobc <- solutrean_dataset %>% 
-  mutate(CLASS = case_when(
-    CLASS == "Flake" | CLASS == "ElongBlank" | CLASS == "Thinning flake" ~ "Blank",
-    CLASS == "FlakeFrag" | CLASS == "ElongBlankFrag" ~ "BlankFrag",
-    TRUE ~ CLASS)) %>% 
-  mutate(CLASS = case_when(
-    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
-    TRUE ~ CLASS)) %>% 
-  filter(RMTYPE != "INDET") %>% 
-  filter(CLASS %in% c("Blank", "Core")) %>% 
-  count(RMTYPE, CLASS, sort = TRUE) %>% 
-  group_by(RMTYPE, CLASS) %>%
-  summarise(n=sum(n)) %>% 
-  spread(CLASS, n, fill=0) %>% 
-  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
-  filter(!sum <= 5) %>% 
-  mutate("Blank/Core" = Blank / Core)
-
-# Proto-Solutrean ratio
-proto_ratiobc <- phase_dataset %>% 
-  filter(Level %in% c("4E", "4E/5")) %>% 
-  filter(RMTYPE != "INDET") %>% 
-  mutate(CLASS = case_when(
-    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
-    TRUE ~ CLASS)) %>% 
-  filter(CLASS %in% c("Blank", "Core")) %>% 
-  count(RMTYPE, CLASS, sort = TRUE) %>% 
-  group_by(RMTYPE, CLASS) %>%
-  summarise(n=sum(n)) %>% 
-  spread(CLASS, n, fill=0) %>% 
-  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
-  filter(!sum <= 5) %>%
-  mutate("Blank/Core" = Blank / Core)
-
-proto5_ratiobc <- phase_dataset %>% 
-  filter(Level == 5) %>% 
-  filter(RMTYPE != "INDET") %>% 
-  mutate(CLASS = case_when(
-    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
-    TRUE ~ CLASS)) %>% 
-  filter(CLASS %in% c("Blank", "Core")) %>% 
-  count(RMTYPE, CLASS, sort = TRUE) %>% 
-  group_by(RMTYPE, CLASS) %>%
-  summarise(n=sum(n)) %>% 
-  spread(CLASS, n, fill=0) %>% 
-  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
-  filter(!sum <= 5) %>% 
-  mutate("Blank/Core" = Blank / Core)
-
-# Gravettian ratio
-gravettian_ratiobc <- phase_dataset %>% 
-  filter(Level %in% c(6, 7)) %>% 
-  filter(RMTYPE != "INDET") %>% 
-  mutate(CLASS = case_when(
-    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
-    TRUE ~ CLASS)) %>% 
-  filter(CLASS %in% c("Blank", "Core")) %>% 
-  count(RMTYPE, CLASS, sort = TRUE) %>% 
-  group_by(RMTYPE, CLASS) %>%
-  summarise(n=sum(n)) %>% 
-  spread(CLASS, n, fill=0) %>% 
-  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
-  filter(!sum <= 5) %>% 
-  mutate("Blank/Core" = Blank / Core)
-
-```
-
-```{r}
-#| label: fig-ratiobc
-#| fig-cap: Plotted ratios of blanks-to-core by assemblage and by chert type. All local cherts were grouped in a single category "Local", while non-local cherts maintained their individual IDs.
-
-gravettian_ratiobc_b <-gravettian_ratiobc %>%
-  filter(`Blank/Core` != "Inf") %>% 
-  dplyr::mutate("Assemblage" = "Gravettian") %>% 
-  select(Assemblage, RMTYPE, `Blank/Core`)
-proto_ratiobc_b <- proto_ratiobc %>% 
-  filter(`Blank/Core` != "Inf") %>% 
-  dplyr::mutate("Assemblage" = "Proto-Solutrean") %>% 
-  select(Assemblage, RMTYPE, `Blank/Core`)
-solutrean_ratiobc_b <- solutrean_ratiobc %>% 
-  filter(`Blank/Core` != "Inf") %>% 
-  dplyr::mutate("Assemblage" = "Solutrean") %>% 
-  select(Assemblage, RMTYPE, `Blank/Core`)
-ratio_blankcore <- bind_rows(gravettian_ratiobc_b, proto_ratiobc_b,solutrean_ratiobc_b)
-
-# Scatterplot with local vs non-local RM Types
-ratio_blankcore |>
-  dplyr::mutate(Source = case_when(
-    RMTYPE == "TYPE 1" | RMTYPE == "TYPE 2" | RMTYPE == "TYPE 3" | RMTYPE == "TYPE 4" | RMTYPE == "TYPE 5" ~ "Local",
-    RMTYPE == "TYPE 6" ~ "T6",
-    RMTYPE == "TYPE 7" ~ "T7",
-    RMTYPE == "TYPE 8" ~ "T8",
-    RMTYPE == "TYPE 9" ~ "T9",
-    RMTYPE == "TYPE 10" ~ "T10",
-    RMTYPE == "TYPE 11" ~ "T11",
-    TRUE ~ RMTYPE)) |>
-  mutate(Source = factor(Source, levels = c("Local", "T6","T7", 
-                                        "T8", "T9", "T10", "T11", "TL"))) |>
-  ggplot(aes(x=Assemblage, y=`Blank/Core`, colour = Source)) +
-  geom_point(size=3) +
-  coord_cartesian(clip = "off") +
-  scale_colour_manual(values = c("#D49B54", "#8B6969", "#CDAF95", "#2B4F60",
-                              "#8B7D6B", "#BDD2B6", "#798777", "#BDC7C9"))
-
-```
-
-## Retouched tools to debitage ratios
-
-Throughout the different assemblages, the difference in retouched tools to debitage ratios by chert types, especially between local and non-local types, seem to show similar patterns.
-In this sense, in general, non-local types seem to consistently show the higher ratios when compared to local cherts.
-In the Gravettian assemblages this pattern is the most accentuated since T1 to T5 show low ratios (0.25-0.5) of retouched to debitage, while TL and T6-T8 show higher ratios (0.9-1.0).
-The exception is T10 (non-local type), which shows a lower ratio (0.28), more akin to local cherts.
-In the Proto-Solutrean assemblage, the lowest ratio of retouched to debitage is seen in T1 and T4, with values below 0.1, while other local types show slightly higher values (0.1-0.2).
-Following the same trend as the Gravettian, the non-local cherts show slightly higher ratios (0.3-0.5) of retouched to debitage.
-In the Solutrean assemblage, all local cherts (T2-T5) show low retouch to debitage ratios (\<0.1), with the exception of T1 (chalcedony), which shows a ratio of 1.8.
-Other non-local cherts (T6-T7 and T11) show slightly higher retouch to debitage ratios (0.1-0.3).
-
-```{r}
-#| label: ratios-retouchdebitage
-
-proto_ratiotd <- phase_dataset %>% 
-  filter(Level %in% c("4E", "4E/5")) %>% 
-  filter(RMTYPE != "INDET") %>% 
-  mutate(CLASS = case_when(
-    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
-    CLASS == "RetouchedPieceFrag" & PIECECOMPLETENESS == "Proximal" ~ "RetouchedPiece",
-    TRUE ~ CLASS)) %>% 
-  filter(CLASS %in% c("Blank", "RetouchedPiece")) %>% 
-  count(RMTYPE, CLASS, sort = TRUE) %>% 
-  group_by(RMTYPE, CLASS) %>%
-  summarise(n=sum(n)) %>% 
-  spread(CLASS, n, fill=0) %>% 
-  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
-  filter(!sum <= 5) %>% 
-  mutate("Tool/Blank" = RetouchedPiece / Blank)
-
-proto5_ratiotd <- phase_dataset %>% 
-  filter(Level == 5) %>% 
-  filter(RMTYPE != "INDET") %>% 
-  mutate(CLASS = case_when(
-    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
-    CLASS == "RetouchedPieceFrag" & PIECECOMPLETENESS == "Proximal" ~ "RetouchedPiece",
-    TRUE ~ CLASS)) %>% 
-  filter(CLASS %in% c("Blank", "RetouchedPiece")) %>% 
-  count(RMTYPE, CLASS, sort = TRUE) %>% 
-  group_by(RMTYPE, CLASS) %>%
-  summarise(n=sum(n)) %>% 
-  spread(CLASS, n, fill=0) %>% 
-  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
-  filter(!sum <= 5) %>% 
-  mutate("Tool/Blank" = RetouchedPiece / Blank)
-
-gravettian_ratiotd <- phase_dataset %>% 
-  filter(Level %in% c(6, 7)) %>% 
-  filter(RMTYPE != "INDET") %>% 
-  mutate(CLASS = case_when(
-    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
-    CLASS == "RetouchedPieceFrag" & PIECECOMPLETENESS == "Proximal" ~ "RetouchedPiece",
-    TRUE ~ CLASS)) %>% 
-  filter(CLASS %in% c("Blank", "RetouchedPiece")) %>% 
-  count(RMTYPE, CLASS, sort = TRUE) %>% 
-  group_by(RMTYPE, CLASS) %>%
-  summarise(n=sum(n)) %>% 
-  spread(CLASS, n, fill=0) %>% 
-  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
-  filter(!sum <= 5) %>% 
-  mutate("Tool/Blank" = RetouchedPiece / Blank)
-
-solutrean_ratiotd <- solutrean_dataset %>% 
-  filter(RMTYPE != "INDET") %>% 
-  mutate(CLASS = case_when(
-    CLASS == "Flake" | CLASS == "ElongBlank" | CLASS == "Thinning flake" ~ "Blank",
-    CLASS == "FlakeFrag" | CLASS == "ElongBlankFrag" ~ "BlankFrag",
-    CLASS == "Retouched piece" ~ "RetouchedPiece",
-    CLASS == "Retouched piece fragment" ~ "RetouchedPieceFrag",
-    TRUE ~ CLASS)) %>% 
-  mutate(CLASS = case_when(
-    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
-    CLASS == "RetouchedPieceFrag" & PIECECOMPLETENESS == "Proximal" ~ "RetouchedPiece",
-    TRUE ~ CLASS)) %>% 
-  filter(CLASS %in% c("Blank", "RetouchedPiece")) %>% 
-  count(RMTYPE, CLASS, sort = TRUE) %>% 
-  group_by(RMTYPE, CLASS) %>%
-  summarise(n=sum(n)) %>% 
-  spread(CLASS, n, fill=0) %>% 
-  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
-  filter(!sum <= 5) %>% 
-  mutate("Tool/Blank" = RetouchedPiece / Blank)
-
-```
-
-```{r}
-#| label: fig-ratiotd
-#| fig-cap: Plotted ratios of tools-to-debitage by assemblage and by chert type. All local cherts were grouped in a single category "Local", while non-local cherts maintained their individual IDs.
-
-gravettian_ratiotd_b <-gravettian_ratiotd %>% 
-  dplyr::mutate("Assemblage" = "Gravettian") %>% 
-  select(Assemblage, RMTYPE, `Tool/Blank`)
-proto_ratiotd_b <- proto_ratiotd %>% 
-  dplyr::mutate("Assemblage" = "Proto-Solutrean") %>% 
-  select(Assemblage, RMTYPE, `Tool/Blank`)
-solutrean_ratiotd_b <- solutrean_ratiotd %>% 
-  dplyr::mutate("Assemblage" = "Solutrean") %>% 
-  select(Assemblage, RMTYPE, `Tool/Blank`)
-ratio_tooldebitage <- bind_rows(gravettian_ratiotd_b, proto_ratiotd_b,solutrean_ratiotd_b)
-
-# Scatterplot with local vs non-local RM Types
-ratio_tooldebitage |>
-  dplyr::mutate(Source = case_when(
-    RMTYPE == "TYPE 1" | RMTYPE == "TYPE 2" | RMTYPE == "TYPE 3" | RMTYPE == "TYPE 4" | RMTYPE == "TYPE 5" ~ "Local",
-    RMTYPE == "TYPE 6" ~ "T6",
-    RMTYPE == "TYPE 7" ~ "T7",
-    RMTYPE == "TYPE 8" ~ "T8",
-    RMTYPE == "TYPE 10" ~ "T10",
-    RMTYPE == "TYPE 11" ~ "T11",
-    TRUE ~ RMTYPE)) |>
-  mutate(Source = factor(Source, levels = c("Local", "T6","T7", 
-                                        "T8", "T9", "T10", "T11", "TL"))) |>
-  ggplot(aes(x=Assemblage, y=`Tool/Blank`, colour = Source)) +
-  geom_point(size=3) +
-  scale_colour_manual(values = c("#D49B54", "#8B6969", "#CDAF95", "#2B4F60",
-                              "#8B7D6B", "#BDD2B6", "#798777", "#BDC7C9"))
-
-```
-
-## Tool typology diversity
-
-Solutrean has a total of 43 retouched tools (complete and fragments), with a high variety of retouched tools (11), ranging from 5 (11.6%) to 4 (9.3%) per tool type.
-The typologies which are most frequently present in the studied sample are bifacial blanks and bifacial fragments, amounting to a total of 23.2% of products related to bifacial tool/point production.
-Two types of points were identified: shouldered points (2 complete and 1 fragment, amounting to a total of 6.9%) and tanged and 4 winged points (9.3%).
-Other frequent retouched typologies are endscrappers (9.3%), notches (11.6%), retouched blanks (\~14%) and splintered pieces (11.6%).
-Regarding chert type, a substantial portion of retouched tools are of unidentified chert types (n=13), mostly due to alterations (fire and post depositional).
-Most retouched tools with identified chert are local (n=13, accounting to `r 13*100/43`%), followed by Type 6 cherts, previously suggested to be from Central Portugal (n=7, accounting to `r 7*100/43`%).
-Local types show a higher variability retouched typologies (9 types), although this may be related to the higher amount of tools.
-In comparison, Type 6 shows only 5 different typologies.
-
-For the Proto-Solutrean, 65 chert retouched tools were identified, with `r 15*100/65`% (n=15) having an unidentified chert type.
-A total of `r 1+10+2+8+5` retouched tools correspond to local cherts, while `r 10+9+3+1` correspond to non-local cherts.
-Regarding the typology, 14 different types of retouched tools were identified.
-Endscrappers (26.2%), splintered pieces (17%), burins (12.3%) and retouched flakes (7.7%) were the most prevalent.
-In comparison to the high frequency of point-related technology observed during the Solutrean, Vale Comprido points are barely present in chalcedony and chert (n=1), as they are mostly present in dolerite and chert was mainly used for other tools as seen in previous studies (\*\*).
-Regarding the types of chalcedony and chert, local raw materials show 9 different typologies, with Vale Comprido points, Solutrean retouch and notches being present only in the local types, even if in small numbers.
-Non-local cherts show 10 different typologies of retouched tools.
-Perforator-endscrapper, truncation, backed bladelet and retouched blades typologies can only be found in non-local varieties, albeit also in small numbers.
-
-In the Gravettian assemblage, 170 retouched tools were identified, with 22.9% (n=39) having an unidentified chert type.
-From the identified chert types, 24.1% of retouched tools are from local types and 52.9% from non-local cherts, with 34.1% of these being from T6 and 14.1% of T7.
-Regarding typologies, 15 types of retouched tools were identified (excluding fragments).
-All these typologies were found in the non-local chert group, while only 7 retouched tool typologies were identified in local cherts/chalcedony.
-The most common typologies identified in the assemblage were burins (n=52), endscrappers (n=17), retouched flakes (n=30) and splintered pieces (n=26).
-From all these, the majority were produced in non-local chert, mainly T6 and in lesser numbers T7.
-Although in lesser numbers as well, backed bladelets were also identified (n=6), but only present in non-local chert types (T6 and T7).
 
@@
 
+## Gravettian (levels 6-7)
+
+The sample studied for the Gravettian occupation (>2012 excavations) of the Terrace area included 766 artifacts.
+When crossing the lithic class with chert type, we see that the local cherts show similar patterns, with T2 and T5 being the types with higher numbers of artefacts (n=110 and n=72, respectively).
+For all local cherts, blanks represent more than 50% of the artefacts.
+Cores and core maintenance products are only present in T2 and T5, although in small frequencies: for T2, cores represent 13.6% of the group and core preparation and maintenance products represent 1.8% (n=2).
+For T5 only 4 core fragments and 1 core preparation and maintenance product are present.
+Retouched tools seem to be especially relevant in T2 (~18%), being the most frequent class after blanks.
+The presence of burin spalls (8.1%) in T2 is expected due to the prevalence of burins (see below).
+In comparison, T5 shows similar percentages of retouched tools and shatter (~12% and ~14%, respectively).
+T1 (chalcedony) shows similar patterns, although with higher percentages of shatter (34.3%).
+In comparison, non-local chert types (especially T6 with n=169 and T7 with n=86) show similar percentages of blanks (~40-50%) but higher percentages of retouched tools (~30%).
+Cores, core fragments and core preparation and maintenance products are also present even if in small percentages (<4%), as well as shatter (~12%).
+In both chert types burin spalls are also present (7.1% for T6 and 4.6% for T7).
+The TL cherts show, as expected, blanks and blank fragments (n=3 and n=2, respectively), and retouched tools (n=4).
+The presence of core maintenance products in TL is possibly related to the limited knapping of TL01 (jasper).
+
+```{r}
+#| label: tbl-general-gravettian
+#| tbl-cap: Number and percentage of technotypological classes by raw material type from the Gravettian assemblage (levels 6 and 7) of the Terrace area of Vale Boi.
+
+kable(gravettian_general)
+
+```
+
+As seen in table **, 170 retouched tools were identified, with 22.9% (n=39) having an unidentified chert type.
+From the identified chert types, 24.1% of retouched tools are from local types and 52.9% from non-local cherts.
+Regarding typologies, 15 types of retouched tools were identified (excluding retouched tool fragments).
+All these typologies were found in the non-local chert group, while only 7 retouched tool typologies were identified in local cherts/chalcedony.
+The most common typologies identified in the assemblage were burins (n=52), endscrappers (n=17), retouched flakes (n=30) and splintered pieces (n=26).
+From all these, the majority were produced in non-local chert, mainly T6 and in lesser numbers T7.
+Although in lesser numbers as well, backed bladelets were also identified (n=6), but only present in non-local chert types (T6 and T7).
+
 ```{r}
@@
 
+Regarding the sizes of the lithic artifacts (width, length and weight), blanks (flakes and elongated products) and retouched tools show little differences in the measurements of artefacts of local and non-local chert types.
+The only exception are T6 and T7 flakes, which show slightly lower weight means (~2.6-3.0 gr) and lower maximum length (~20-22 mm) than the other types.
+For cores, T2 shows bigger (heavier) values than T6 (means of 21.3 and 12.6, and maximum weight of 37.2 gr and 22.5 gr, respectively), even if they seem to vary in morphology, with little patterns regarding thickness, width and length.
+
+```{r}
+#| label: fig-gravettian-boxplots
+#| fig-cap: Boxplots Solutrean
+
+gravettian_graphs <- ggpubr::ggarrange(gravettian_scatter, gravettian_stats_weight, labels = c("c1", "c2"), ncol = 2, common.legend = TRUE, legend = "bottom")
+
+```
+
+As mentioned in the methodology, two different ratios were calculated: blank-to-core and tools-to-debitage.
+In general, non-local chert types show higher values for these ratios.
+Local types (T1 and T2) show lower ratios (<4) of blanks to cores, although other local types (T3-T5) have no identified cores.
+The non-local types, especially T6 and T7 show higher ratios (~8) of blanks to cores.
+T1 to T5 show low ratios (0.25-0.5) of retouched to debitage, while TL and T6-T8 show higher ratios (0.9-1.0).
+The exception is T10 (non-local type), which shows a lower ratio (0.28), more akin to local cherts.
+
+```{r}
+#| label: fig-ratiobc
+#| fig-cap: Plotted ratios of blanks-to-core by assemblage and by chert type. All local cherts were grouped in a single category "Local", while non-local cherts maintained their individual IDs.
+
+# DS preparation and tool-to-debitage ratio tables
+## For Gravettian
+gravettian_ratiobc <- phase_dataset %>% 
+  filter(Level %in% c(6, 7)) %>% 
+  filter(RMTYPE != "INDET") %>% 
+  mutate(CLASS = case_when(
+    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
+    TRUE ~ CLASS)) %>% 
+  filter(CLASS %in% c("Blank", "Core")) %>% 
+  count(RMTYPE, CLASS, sort = TRUE) %>% 
+  group_by(RMTYPE, CLASS) %>%
+  summarise(n=sum(n)) %>% 
+  spread(CLASS, n, fill=0) %>% 
+  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
+  filter(!sum <= 5) %>% 
+  mutate("Blank/Core" = Blank / Core)
+
+# For Proto-Solutrean
+proto_ratiobc <- phase_dataset %>% 
+  filter(Level %in% c("4E", "4E/5")) %>% 
+  filter(RMTYPE != "INDET") %>% 
+  mutate(CLASS = case_when(
+    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
+    TRUE ~ CLASS)) %>% 
+  filter(CLASS %in% c("Blank", "Core")) %>% 
+  count(RMTYPE, CLASS, sort = TRUE) %>% 
+  group_by(RMTYPE, CLASS) %>%
+  summarise(n=sum(n)) %>% 
+  spread(CLASS, n, fill=0) %>% 
+  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
+  filter(!sum <= 5) %>%
+  mutate("Blank/Core" = Blank / Core)
+
+# For Solutrean
+solutrean_ratiobc <- solutrean_dataset %>% 
+  mutate(CLASS = case_when(
+    CLASS == "Flake" | CLASS == "ElongBlank" | CLASS == "Thinning flake" ~ "Blank",
+    CLASS == "FlakeFrag" | CLASS == "ElongBlankFrag" ~ "BlankFrag",
+    TRUE ~ CLASS)) %>% 
+  mutate(CLASS = case_when(
+    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
+    TRUE ~ CLASS)) %>% 
+  filter(RMTYPE != "INDET") %>% 
+  filter(CLASS %in% c("Blank", "Core")) %>% 
+  count(RMTYPE, CLASS, sort = TRUE) %>% 
+  group_by(RMTYPE, CLASS) %>%
+  summarise(n=sum(n)) %>% 
+  spread(CLASS, n, fill=0) %>% 
+  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
+  filter(!sum <= 5) %>% 
+  mutate("Blank/Core" = Blank / Core)
+
+# Cleaning up to create a single table for scatterplot
+gravettian_ratiobc_b <-gravettian_ratiobc %>%
+  filter(`Blank/Core` != "Inf") %>% 
+  dplyr::mutate("Assemblage" = "Gravettian") %>% 
+  select(Assemblage, RMTYPE, `Blank/Core`)
+proto_ratiobc_b <- proto_ratiobc %>% 
+  filter(`Blank/Core` != "Inf") %>% 
+  dplyr::mutate("Assemblage" = "Proto-Solutrean") %>% 
+  select(Assemblage, RMTYPE, `Blank/Core`)
+solutrean_ratiobc_b <- solutrean_ratiobc %>% 
+  filter(`Blank/Core` != "Inf") %>% 
+  dplyr::mutate("Assemblage" = "Solutrean") %>% 
+  select(Assemblage, RMTYPE, `Blank/Core`)
+ratio_blankcore <- bind_rows(gravettian_ratiobc_b, proto_ratiobc_b,solutrean_ratiobc_b)
+
+# Scatterplot with local vs non-local RM Types
+ratio_blankcore |>
+  dplyr::mutate(Source = case_when(
+    RMTYPE == "TYPE 1" | RMTYPE == "TYPE 2" | RMTYPE == "TYPE 3" | RMTYPE == "TYPE 4" | RMTYPE == "TYPE 5" ~ "Local",
+    RMTYPE == "TYPE 6" ~ "T6",
+    RMTYPE == "TYPE 7" ~ "T7",
+    RMTYPE == "TYPE 8" ~ "T8",
+    RMTYPE == "TYPE 9" ~ "T9",
+    RMTYPE == "TYPE 10" ~ "T10",
+    RMTYPE == "TYPE 11" ~ "T11",
+    TRUE ~ RMTYPE)) |>
+  mutate(Source = factor(Source, levels = c("Local", "T6","T7", 
+                                        "T8", "T9", "T10", "T11", "TL"))) |>
+  ggplot(aes(x=Assemblage, y=`Blank/Core`, colour = Source)) +
+  geom_point(size=3) +
+  coord_cartesian(clip = "off") +
+  scale_colour_manual(values = c("#D49B54", "#8B6969", "#CDAF95", "#2B4F60",
+                              "#8B7D6B", "#BDD2B6", "#798777", "#BDC7C9"))
+
+```
+
+```{r}
+#| label: fig-ratiotd
+#| fig-cap: Plotted ratios of tools-to-debitage by assemblage and by chert type. All local cherts were grouped in a single category "Local", while non-local cherts maintained their individual IDs.
+
+# DS preparation and tool-to-debitage ratio tables
+## For Gravettian
+gravettian_ratiotd <- phase_dataset %>% 
+  filter(Level %in% c(6, 7)) %>% 
+  filter(RMTYPE != "INDET") %>% 
+  mutate(CLASS = case_when(
+    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
+    CLASS == "RetouchedPieceFrag" & PIECECOMPLETENESS == "Proximal" ~ "RetouchedPiece",
+    TRUE ~ CLASS)) %>% 
+  filter(CLASS %in% c("Blank", "RetouchedPiece")) %>% 
+  count(RMTYPE, CLASS, sort = TRUE) %>% 
+  group_by(RMTYPE, CLASS) %>%
+  summarise(n=sum(n)) %>% 
+  spread(CLASS, n, fill=0) %>% 
+  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
+  filter(!sum <= 5) %>% 
+  mutate("Tool/Blank" = RetouchedPiece / Blank)
+
+## For Proto-Solutrean
+proto_ratiotd <- phase_dataset %>% 
+  filter(Level %in% c("4E", "4E/5")) %>% 
+  filter(RMTYPE != "INDET") %>% 
+  mutate(CLASS = case_when(
+    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
+    CLASS == "RetouchedPieceFrag" & PIECECOMPLETENESS == "Proximal" ~ "RetouchedPiece",
+    TRUE ~ CLASS)) %>% 
+  filter(CLASS %in% c("Blank", "RetouchedPiece")) %>% 
+  count(RMTYPE, CLASS, sort = TRUE) %>% 
+  group_by(RMTYPE, CLASS) %>%
+  summarise(n=sum(n)) %>% 
+  spread(CLASS, n, fill=0) %>% 
+  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
+  filter(!sum <= 5) %>% 
+  mutate("Tool/Blank" = RetouchedPiece / Blank)
+
+## For Solutrean
+solutrean_ratiotd <- solutrean_dataset %>% 
+  filter(RMTYPE != "INDET") %>% 
+  mutate(CLASS = case_when(
+    CLASS == "Flake" | CLASS == "ElongBlank" | CLASS == "Thinning flake" ~ "Blank",
+    CLASS == "FlakeFrag" | CLASS == "ElongBlankFrag" ~ "BlankFrag",
+    CLASS == "Retouched piece" ~ "RetouchedPiece",
+    CLASS == "Retouched piece fragment" ~ "RetouchedPieceFrag",
+    TRUE ~ CLASS)) %>% 
+  mutate(CLASS = case_when(
+    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
+    CLASS == "RetouchedPieceFrag" & PIECECOMPLETENESS == "Proximal" ~ "RetouchedPiece",
+    TRUE ~ CLASS)) %>% 
+  filter(CLASS %in% c("Blank", "RetouchedPiece")) %>% 
+  count(RMTYPE, CLASS, sort = TRUE) %>% 
+  group_by(RMTYPE, CLASS) %>%
+  summarise(n=sum(n)) %>% 
+  spread(CLASS, n, fill=0) %>% 
+  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
+  filter(!sum <= 5) %>% 
+  mutate("Tool/Blank" = RetouchedPiece / Blank)
+
+# Creating a single table with all ratios by assemblage for the scatterplot
+gravettian_ratiotd_b <-gravettian_ratiotd %>% 
+  dplyr::mutate("Assemblage" = "Gravettian") %>% 
+  select(Assemblage, RMTYPE, `Tool/Blank`)
+proto_ratiotd_b <- proto_ratiotd %>% 
+  dplyr::mutate("Assemblage" = "Proto-Solutrean") %>% 
+  select(Assemblage, RMTYPE, `Tool/Blank`)
+solutrean_ratiotd_b <- solutrean_ratiotd %>% 
+  dplyr::mutate("Assemblage" = "Solutrean") %>% 
+  select(Assemblage, RMTYPE, `Tool/Blank`)
+ratio_tooldebitage <- bind_rows(gravettian_ratiotd_b, proto_ratiotd_b,solutrean_ratiotd_b)
+
+# Scatterplot with local vs non-local RM Types
+ratio_tooldebitage |>
+  dplyr::mutate(Source = case_when(
+    RMTYPE == "TYPE 1" | RMTYPE == "TYPE 2" | RMTYPE == "TYPE 3" | RMTYPE == "TYPE 4" | RMTYPE == "TYPE 5" ~ "Local",
+    RMTYPE == "TYPE 6" ~ "T6",
+    RMTYPE == "TYPE 7" ~ "T7",
+    RMTYPE == "TYPE 8" ~ "T8",
+    RMTYPE == "TYPE 10" ~ "T10",
+    RMTYPE == "TYPE 11" ~ "T11",
+    TRUE ~ RMTYPE)) |>
+  mutate(Source = factor(Source, levels = c("Local", "T6","T7", 
+                                        "T8", "T9", "T10", "T11", "TL"))) |>
+  ggplot(aes(x=Assemblage, y=`Tool/Blank`, colour = Source)) +
+  geom_point(size=3) +
+  scale_colour_manual(values = c("#D49B54", "#8B6969", "#CDAF95", "#2B4F60",
+                              "#8B7D6B", "#BDD2B6", "#798777", "#BDC7C9"))
+
+```
+
+## Proto-Solutrean (levels 4E and top of level 5)
+
+The final lithic sample from the Proto-Solutrean assemblage, consisting of the levels 4E and top of level 5, consists of 635 artifacts.
+When crossing the lithic class with chert type, we see that the local types show high frequency of blanks (>50%, including blank fragments), followed by shatter (~11-27%), cores (~9-12%) and retouched tools (~5-15%).
+A single burin spall was identified in T4.
+There is also the absence of core preparation and maintenance products across all local cherts.
+T1 (chalcedony) is composed mostly of blanks and a retouched tool (Vale Comprido point).
+The absence of cores but presence of shatter may hint at the knapping of this raw material at the site, but the absence of formal cores perhaps due to the quality of nodules/lenses, which are characterised by fractures.
+For non-local types, especially those with higher number of artefacts (T6-T8), blanks and blank fragments are the most frequent class (48.3%, 53.1% and 69.2%, respectively).
+Retouched tools are present in smaller frequencies for the previous mentioned chert types (~17-23%).
+For T6 the absence of cores but the presence of shatter and a single core preparation and maintenance product suggests knapping on site.
+Similarly, T7 shows the presence of cores and core fragments (8.4%), shatter (17.0%) and a single core preparation and maintenance product.
+In comparison, T8, despite more limited in number (n=13), shows only one core, with the rest of the artefacts being blanks (as previously noted) or retouched tools (23.1% and n=3).
+Aside from the TL artefacts, all non-local cores show at least 1 core and frequently the presence of shatter.
+As expected, the TL artifacts include a blank and a retouched.
+However, a shatter was also identified.
+In total, 65 chert retouched tools were identified, with 23.1% (n=15) having an unidentified chert type.
+A total of 26 retouched tools correspond to local cherts, while 23 correspond to non-local cherts.
+
+```{r}
+#| label: tbl-general-proto
+#| tbl-cap: Number and percentage of technotypological classes by raw material type from the Proto-Solutrean assemblage (levels 4E and top of level 5) of the Terrace area of Vale Boi.
+
+kable(proto_general)
+
+```
+
+Regarding the typology, 14 different types of retouched tools were identified.
+Endscrappers (26.2%), splintered pieces (17%), burins (12.3%) and retouched flakes (7.7%) were the most prevalent.
+Vale Comprido points, the Proto-Solutrean fossil director, are barely present in chalcedony and chert (n=1), as they are mostly present in dolerite, while chert was mainly used for other tools as seen in previous studies (Belmiro et al., 2021).
+Regarding the types of chalcedony and chert, local raw materials show 9 different typologies, with Vale Comprido points, Solutrean retouch and notches being present only in the local types, even if in small numbers.
+Non-local cherts show 10 different typologies of retouched tools.
+Perforator-endscrapper, truncation, backed bladelet and retouched blades typologies can only be found in non-local varieties, albeit also in small numbers.
+
 ```{r}
@@
 
+Regarding the debitage, cores and tool sizes, flakes show slightly higher means in local cherts (~21-22 mm for length and 3.5-4.5 gr for weight), in comparison to non-local cherts, especially T6-T8 (~19-20 mm for length and 1.3-2.6 gr in weight), with exception of chalcedony which shows consistently low sizes (21.8 mm for length and 2.8 gr for weight).
+Elongated blank however, show no differences between local and non-local chert types.
+In retouched tools, weight also shows smaller means for non-local cherts (~3-2 gr), although with higher standard deviation (~1.6).
+Regarding cores, the sample size for non-local cherts is small and only T7 statistics could be obtained.
+Despite this, when compared to local cherts (T2-T5), T7 cores show smaller mean and maximum values (7.3 gr and 8 gr, respectively).
+
+```{r}
+#| label:  fig-proto-boxplots
+#| fig-cap: Boxplots Proto-Solutrean
+
+proto_graphs <- ggpubr::ggarrange(proto_scatter, proto_stats_weight, labels = c("b1", "b2"), ncol = 2, common.legend = TRUE, legend = "bottom")
+
+```
+
+Similar to the Gravettian, non-local cherts seem to show the highest values in the calculated blank-to-core and tool-to-debitage ratios.
+T2 and T4 show the lowest ratios (8.7 and 6.2 respectively) of blanks per cores, while the non-local T7 shows the highest ratios (11.0) of blanks to cores.
+Unlike the Gravettian, T1 (local chalcedony) and T6 (non-local chert) have no cores.
+The lowest ratio of retouched to debitage is seen in T1 and T4, with values below 0.1, while other local types show slightly higher values (0.1-0.2).
+Following the same trend as the Gravettian, the non-local cherts show slightly higher ratios (0.3-0.5) of retouched to debitage.
+
+## Solutrean (levels A to C)
+
+The Solutrean assemblages show similar patterns to those observed in previous assemblages.
+The final sample from levels A, B and C from the Solutrean occupations of the Shelter area consists of 914 artifacts.
+When crossing the class data with type of raw material, we see that local chert, especially T2 (n=506) which accounts for the majority of local chert present in the studied samples, the majority of the debitage is composed of flakes and elongated blanks (83.9%).
+Cores are also present (10.5%), alongside a small percentage of core preparation and maintenance products (1.3%).
+In comparison to debitage products, retouched tools are scarce in the sample (~3%).
+The non-local cherts most present in the Solutrean of the Shelter area are T6 (n=86) and T8 (n=30).
+These show similar patterns to the local cherts, with high percentages of blanks (>50%).
+Cores and core preparation and maintenance are also present in small percentages (~3-6% and 3.5% only for T6, respectively), as well as retouched tools (<10%).
+However, these specific results are truncated by the lack of shatter, which were not present in the sample used for the study (see Materials and Methods section).
+It is important to notice, however, that alike previous results from the Solutrean assemblages of the Terrace area, a large portion of the studied sample was altered and did not allow for a chert type attribution (21.9912473%).
+
+```{r}
+#| label: tbl-general-solutrean
+#| tbl-cap: Number and percentage of technotypological classes by raw material type from the Solutrean assemblage (levels A to C) of the Shelter area of Vale Boi.
+
+kable(solutrean_general)
+
+```
+
+The analysed Solutrean sample has a total of 43 retouched tools (complete and fragments), with a high variety of retouched tools (11), ranging from 5 (11.6%) to 4 (9.3%) per tool type.
+These results however, are significantly reduced from the original dataset (Cascalheira et al., 2013), possibly due to no bucket coordinates being used.
+The typologies which are most frequently present in the studied sample are bifacial blanks and bifacial fragments, amounting to a total of 23.2% of products related to bifacial tool/point production.
+Two types of points were identified: shouldered points (2 complete and 1 fragment, amounting to a total of 6.9%) and tanged and 4 winged points (9.3%).
+Other frequent retouched typologies are end scrappers (9.3%), notches (11.6%), retouched blanks (~14%) and splintered pieces (11.6%).
+Regarding chert type, a substantial portion of retouched tools are of unidentified chert types (n=13), mostly due to alterations (fire and post depositional).
+Most retouched tools with identified chert are local (n=13, accounting to 30.2325581%), followed by Type 6 cherts, previously suggested to be from Central Portugal (n=7, accounting to 16.2790698%).
+Local types show a higher variability retouched typologies (9 types), although this may be related to the higher number of tools.
+In comparison, Type 6 shows only 5 different typologies.
+
 ```{r}
@@
 ```
+
+For the Solutrean assemblage, there seems to be no clear difference in the measurements and weights of artefacts between local and non-local cherts.
+This is true especially for the most represented classes (flakes and elongated blanks).
+Regarding cores, many of the local and non-local cherts are not available for the statistics due to missing data.
+However, the smallest weight mean seem to be for cores of T6 (16.5 gr), although with high standard deviation and maximum value (16.5 and 48.6 gr, respectively).
+
+```{r}
+#| label: fig-solutrean-boxplots
+#| fig-cap: Boxplots Gravettian.
+
+solutrean_graphs <- ggpubr::ggarrange(solutrean_scatter, solutrean_stats_weight, labels = c("a1", "a2"), ncol = 2, common.legend = TRUE, legend = "bottom")
+
+```
+
+Regarding the blank-to-core and tools-to-debitage ratios, the Solutrean assemblage shows similar patterns as the previous assemblages, with the local cherts showing lower blank to core ratios (<7), while the non-local cherts show higher ratios (>9).
+Similar to the Proto-Solutrean, chalcedony cores were also not identified in the studied sample.
+Similarly, local cherts (T2-T5) show low retouch to debitage ratios (<0.1), with the exception of T1 (chalcedony), which shows a ratio of 1.8.
+Other non-local cherts (T6-T7 and T11) show slightly higher retouch to debitage ratios (0.1-0.3).
 

### 2025-01-21 analysis/paper/paper.qmd
Pipe %>% replaced by |>

@@
 # Cleaning up DS and uniformising variable names
-proto_tech_WIP <- proto_tech %>%
-  tidyr::unite(col='ID', all_of(cols_to_concat_terrace), sep="-", remove=TRUE) %>% 
+proto_tech_WIP <- proto_tech |>
+  tidyr::unite(col='ID', all_of(cols_to_concat_terrace), sep="-", remove=TRUE) |> 
   rename(CLASS = Class, COREPREPPROD = CorePreparProd, PIECECOMPLETENESS = PieceCompleteness, CORTEX = Cortex, CORTEXLOCATION = CortexLocation, PLATFORMTYPE = PlatformType, PLATFORMCORTEX = PlatformCortex, SCARCOUNT = ScarCount, THICKNESS = Thickness,
         MAXWIDTH = MaxWidth, LENGTH = Length, COREPLATFORM = CorePlatform, MAINFACECORTEX = MainFaceCortex, 
-        MAINFACESCARCOUNT = MainFaceScarCount, COREABANDON = CoreAbandon, RETOUCHEDPIECETYPOLOGY = RetouchedPieceTypology) %>% 
+        MAINFACESCARCOUNT = MainFaceScarCount, COREABANDON = CoreAbandon, RETOUCHEDPIECETYPOLOGY = RetouchedPieceTypology) |> 
   mutate(RETOUCHEDPIECETYPOLOGY = case_when(
@@
     RETOUCHEDPIECETYPOLOGY == "Burin on truncation" ~ "Burin",
-    TRUE ~ RETOUCHEDPIECETYPOLOGY)) %>% 
+    TRUE ~ RETOUCHEDPIECETYPOLOGY)) |> 
   dplyr::select(ID, CLASS, COREPREPPROD, PIECECOMPLETENESS, CORTEX, CORTEXLOCATION, PLATFORMTYPE, PLATFORMCORTEX,
@@
 
-gravettian_tech <- gravettian_tech %>% 
+gravettian_tech <- gravettian_tech |> 
   mutate(RETOUCHEDPIECETYPOLOGY = case_when(
@@
     RETOUCHEDPIECETYPOLOGY == "Double backed bladelet" ~ "Backed bladelet",
-    TRUE ~ RETOUCHEDPIECETYPOLOGY)) %>% 
+    TRUE ~ RETOUCHEDPIECETYPOLOGY)) |> 
   mutate(CLASS = case_when(
@@
 
-terrace_RM_WIP <- terrace_RM %>%
+terrace_RM_WIP <- terrace_RM |>
   mutate(RMTYPE = case_when(
@@
     RMTYPE == "TYPE GB" ~ "TYPE 11",
-    .default = RMTYPE)) %>% 
-  select(AREA, ID, RMTYPE, SUBTYPE, WEIGHT, CORTEXTHICKNESS, CORTEXAPPEARANCE) %>%
+    .default = RMTYPE)) |> 
+  select(AREA, ID, RMTYPE, SUBTYPE, WEIGHT, CORTEXTHICKNESS, CORTEXAPPEARANCE) |>
   distinct(ID, .keep_all = TRUE)
 
-context <- context %>%
-  dplyr::rename("UNIT" = "Unit") %>% # Renaming first column from Unit to UNIT to use cols_to_concat
-  unite(col='ID', cols_to_concat_terrace, sep="-", remove=FALSE) %>% # Unite UNIT with ID using cols_to_concat
-  select(ID, Level, Spit, Code) %>% # Selecting important variables: ID refers to artifact ID; Level refers to archaeologycal layer; Spit refers to artificial levels within an archaeological layer; Code refers to type of artifact (e.g., LITHIC or BONE)
+context <- context |>
+  dplyr::rename("UNIT" = "Unit") |> # Renaming first column from Unit to UNIT to use cols_to_concat
+  unite(col='ID', cols_to_concat_terrace, sep="-", remove=FALSE) |> # Unite UNIT with ID using cols_to_concat
+  select(ID, Level, Spit, Code) |> # Selecting important variables: ID refers to artifact ID; Level refers to archaeologycal layer; Spit refers to artificial levels within an archaeological layer; Code refers to type of artifact (e.g., LITHIC or BONE)
   distinct(ID, .keep_all = TRUE)
 
-xyz <- xyz %>%
-  unite(col='ID', cols_to_concat_terrace, sep="-", remove=FALSE) %>% # Unite UNIT with ID using cols_to_concat
-  filter(Suffix == 0) %>% # To avoid several entries related to several points of the same artefact
-  select(ID, X, Y, Z) %>% # Selecting important variables
+xyz <- xyz |>
+  unite(col='ID', cols_to_concat_terrace, sep="-", remove=FALSE) |> # Unite UNIT with ID using cols_to_concat
+  filter(Suffix == 0) |> # To avoid several entries related to several points of the same artefact
+  select(ID, X, Y, Z) |> # Selecting important variables
   distinct(ID, .keep_all = TRUE)
@@
 terrace_dataset <- full_join(terrace_tech_WIP, terrace_RM_WIP, by = "ID") # Base 12690
-terrace_dataset <- terrace_dataset %>%
-  filter(!Level %in% c("3", "4", "4C", "4D")) %>%  # Dropping NAs on the technological class drops dataset to 1651
-  drop_na(RMTYPE) %>%  # Dropping NAs on the RM database drops dataset to 3635
+terrace_dataset <- terrace_dataset |>
+  filter(!Level %in% c("3", "4", "4C", "4D")) |>  # Dropping NAs on the technological class drops dataset to 1651
+  drop_na(RMTYPE) |>  # Dropping NAs on the RM database drops dataset to 3635
   drop_na(CLASS) # Dropping NAs on the technological class drops dataset to 1647
@@
 # Mutate level 5 to add phases: top of 5 (4E/5) and bottom of 5 (5), following Belmiro et al. (2020).
-phase_dataset <- terrace_dataset %>% 
+phase_dataset <- terrace_dataset |> 
   mutate(Level = case_when(
@@
 
-solutrean_WIP <- solutrean_JC %>% 
-  filter(Layer %in% c("A", "B", "C")) %>% # Filtering Solutrean layers
-  filter(Raw.material %in% c("Chert", "Chalcedony")) %>% # Filtering raw materials
-  filter(Screen != "Crivo") %>%  # Filtering out all bucket lithics since they were not included in the RM analysis
-  mutate_at("Date", str_replace, "20", "") %>% # Removing initial part of the year to match IDs
-  dplyr::rename("YEAR" = "Date") %>% 
+solutrean_WIP <- solutrean_JC |> 
+  filter(Layer %in% c("A", "B", "C")) |> # Filtering Solutrean layers
+  filter(Raw.material %in% c("Chert", "Chalcedony")) |> # Filtering raw materials
+  filter(Screen != "Crivo") |>  # Filtering out all bucket lithics since they were not included in the RM analysis
+  mutate_at("Date", str_replace, "20", "") |> # Removing initial part of the year to match IDs
+  dplyr::rename("YEAR" = "Date") |> 
   tidyr::unite(col='ID', all_of(cols_to_concat_shelter), sep="-", remove=FALSE) # Concatenate YEAR and ID variables in one single variable (e.g. 05-500)
@@
 # Dataset filtering for chert/chalcedony and levels (A-C), 3298 + 61 solutrean points, n=3359
-solutrean_original_chert <- solutrean_JC %>% 
-  filter(Layer %in% c("A", "B", "C")) %>% # Filtering Solutrean layers
+solutrean_original_chert <- solutrean_JC |> 
+  filter(Layer %in% c("A", "B", "C")) |> # Filtering Solutrean layers
   filter(Raw.material %in% c("Chert", "Chalcedony"))
@@
 # In Cascalheira 2010 Retouched piece was not a category but a variable. To match the other DBs, this category was added
-solutrean_WIP <- solutrean_WIP %>% 
+solutrean_WIP <- solutrean_WIP |> 
   mutate(Class = case_when( # This alteration is being done on the original class column
     Retouched == "Yes" ~ "Retouched piece", # Whenever retouch is present, class is now Retouched piece
-    TRUE ~ Class)) %>% 
+    TRUE ~ Class)) |> 
   mutate(CLASS = case_when( # Creating a column for clustered classes, to match the other DBs
@@
     | Class == "Bladelet distal frag." | Class == "Bladelet mesial frag." | Class == "Bladelet prox. frag." ~ "ElongBlankFrag", # Cluster elong frags into a single category
-    TRUE ~ Class)) %>% 
+    TRUE ~ Class)) |> 
   mutate(COREPREPPROD = case_when( # Creating a column for the types of core preparation and maintenance products (COREPREPPROD)
@@
     Class == "Core trimming elem." ~ "Core trim",
-    Class == "Crested piece" ~ "Crested piece")) %>% 
+    Class == "Crested piece" ~ "Crested piece")) |> 
   mutate(PIECECOMPLETENESS = case_when( # Creating a column for the portions present in flake or elongblank fragments to match other DBs
@@
     Class == "Flake mesial frag." | Class == "Blade mesial frag." | Class == "Bladelet mesial frag." ~ "Mesial",
-    Class == "Flake prox. frag." | Class == "Blade prox. frag." | Class == "Bladelet prox. frag." ~ "Proximal")) %>% 
-  rename(PLATFORMTYPE = Platform_type) %>% # Renaming column
+    Class == "Flake prox. frag." | Class == "Blade prox. frag." | Class == "Bladelet prox. frag." ~ "Proximal")) |> 
+  rename(PLATFORMTYPE = Platform_type) |> # Renaming column
   mutate(PLATFORMTYPE = case_when( # Renaming variables to match correspondent variables in the other DBs
@@
     PLATFORMTYPE == "Unfacetted" ~ "Plain",
-    TRUE ~ PLATFORMTYPE)) %>%
+    TRUE ~ PLATFORMTYPE)) |>
   mutate(Type = case_when( # Translating retouched types from Zilhão 1997 to simplified typologies
@@
     Type == "7" | Type == "5b" | Type == "4"  ~ "Endscrapper",
-    TRUE ~ Type)) %>%
+    TRUE ~ Type)) |>
   rename(LEVEL = Layer, UNIT = Unit, CORTEX = Percent_of_cortex, CORTEXLOCATION = Loc_of_cortex, THICKNESS = Thickness,
@@
 
-solutrean_tech <- solutrean_WIP %>%
+solutrean_tech <- solutrean_WIP |>
   dplyr::select(ID, CLASS, LEVEL, UNIT, COREPREPPROD, PIECECOMPLETENESS, CORTEX, CORTEXLOCATION, PLATFORMTYPE,
@@
 
-solutrean_points <- solutrean_points %>% 
-  mutate_at("YEAR", str_replace, "20", "") %>% # Removing initial part of the year to match IDs
+solutrean_points <- solutrean_points |> 
+  mutate_at("YEAR", str_replace, "20", "") |> # Removing initial part of the year to match IDs
   tidyr::unite(col='ID', all_of(cols_to_concat_shelter), sep="-", remove=TRUE)
@@
 
-solutrean_tech <- solutrean_tech %>% # Originally 2228
+solutrean_tech <- solutrean_tech |> # Originally 2228
   distinct(ID, .keep_all = TRUE) # Keeping individual IDs to avoid buckets (down to 1993)
 
-shelter_RM_WIP <- shelter_RM %>%
-  mutate_at("ID", str_replace, "VB", "") %>% # Remove "VB" from ID
+shelter_RM_WIP <- shelter_RM |>
+  mutate_at("ID", str_replace, "VB", "") |> # Remove "VB" from ID
   mutate(RMTYPE = case_when(
@@
     RMTYPE == "ND_A" ~ "INDET",
-    .default = RMTYPE)) %>%
-  select(AREA, ID, RMTYPE, SUBTYPE, WEIGHT, CORTEXTHICKNESS, CORTEXAPPEARANCE) %>%  # Select variables
+    .default = RMTYPE)) |>
+  select(AREA, ID, RMTYPE, SUBTYPE, WEIGHT, CORTEXTHICKNESS, CORTEXAPPEARANCE) |>  # Select variables
   distinct(ID, .keep_all = TRUE) # Keeping individual IDs to avoid buckets
@@
 solutrean_dataset <- full_join(solutrean_tech, shelter_RM_WIP, by = "ID") # The complete dataset has 2818 (44 points) entries
-solutrean_dataset <- solutrean_dataset %>%
-  drop_na(RMTYPE) %>%   # Dropping NAs on the RM database drops dataset to 1514
+solutrean_dataset <- solutrean_dataset |>
+  drop_na(RMTYPE) |>   # Dropping NAs on the RM database drops dataset to 1514
   drop_na(CLASS) # Dropping NAs on the technological class drops dataset to 914
@@
 # DS preparation
-solutrean_stats <- solutrean_dataset %>% 
-  filter(RMTYPE != "INDET") %>%
+solutrean_stats <- solutrean_dataset |> 
+  filter(RMTYPE != "INDET") |>
    mutate(CLASS = case_when(
@@
     CLASS == "Retouched piece" ~ "RetouchedPiece",
-    TRUE ~ CLASS)) %>% 
+    TRUE ~ CLASS)) |> 
   mutate(CLASS = case_when(
@@
 # Weight
-solutrean_stats_weight <- solutrean_stats %>%
-  filter(CLASS %in% c("Flake", "ElongatedProd", "Core", "RetouchedPiece")) %>% 
-  filter(WEIGHT !="0" & !RMTYPE %in% c("INDET","GB","TL")) %>%
+solutrean_stats_weight <- solutrean_stats |>
+  filter(CLASS %in% c("Flake", "ElongatedProd", "Core", "RetouchedPiece")) |> 
+  filter(WEIGHT !="0" & !RMTYPE %in% c("INDET","GB","TL")) |>
   mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
                                         "TYPE 5", "TYPE 6","TYPE 7", 
-                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
-  mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) %>%
+                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) |>
+  mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) |>
   ggplot(aes(RMTYPE, WEIGHT, fill = RMTYPE)) +
@@
 # DS prep
-proto_stats <- phase_dataset %>% 
-  filter(Level %in% c("4E", "4E/5")) %>% 
-  filter(RMTYPE != "INDET") %>% 
+proto_stats <- phase_dataset |> 
+  filter(Level %in% c("4E", "4E/5")) |> 
+  filter(RMTYPE != "INDET") |> 
   mutate(CLASS = case_when(
     CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "BlankFragProx", # To add proximal fragments to ratio
-    TRUE ~ CLASS)) %>%
-  mutate(elong = MAXWIDTH*2-LENGTH) %>%
+    TRUE ~ CLASS)) |>
+  mutate(elong = MAXWIDTH*2-LENGTH) |>
   mutate(CLASS = case_when(
@@
 # Boxplots weight
-proto_stats_weight <- proto_stats %>%
-  filter(CLASS %in% c("Flake","ElongatedProd", "Core", "RetouchedPiece")) %>% 
-  filter(WEIGHT !="0" & !RMTYPE %in% c("INDET","GB","TL")) %>%
+proto_stats_weight <- proto_stats |>
+  filter(CLASS %in% c("Flake","ElongatedProd", "Core", "RetouchedPiece")) |> 
+  filter(WEIGHT !="0" & !RMTYPE %in% c("INDET","GB","TL")) |>
   mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
                                         "TYPE 5", "TYPE 6","TYPE 7", 
-                                        "TYPE 8", "TYPE 9", "TYPE 10"))) %>%
-  mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) %>%
+                                        "TYPE 8", "TYPE 9", "TYPE 10"))) |>
+  mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) |>
   ggplot(aes(RMTYPE, WEIGHT, fill = RMTYPE)) +
@@
 # DS prep
-gravettian_stats <- phase_dataset %>%
-  filter(Level %in% c(6, 7)) %>%
-  filter(RMTYPE != "INDET") %>%
+gravettian_stats <- phase_dataset |>
+  filter(Level %in% c(6, 7)) |>
+  filter(RMTYPE != "INDET") |>
   mutate(CLASS = case_when(
     CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "BlankFragProx", # To add proximal fragments to ratio
-    TRUE ~ CLASS)) %>% 
-  mutate(elong = MAXWIDTH*2-LENGTH) %>%
+    TRUE ~ CLASS)) |> 
+  mutate(elong = MAXWIDTH*2-LENGTH) |>
   mutate(CLASS = case_when(
@@
 # Weight
-gravettian_stats_weight <- gravettian_stats %>%
-  filter(CLASS %in% c("Flake","ElongatedProd", "Core", "RetouchedPiece")) %>% 
-  filter(WEIGHT !="0" & !RMTYPE %in% c("INDET","GB","TL")) %>%
+gravettian_stats_weight <- gravettian_stats |>
+  filter(CLASS %in% c("Flake","ElongatedProd", "Core", "RetouchedPiece")) |> 
+  filter(WEIGHT !="0" & !RMTYPE %in% c("INDET","GB","TL")) |>
   mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
                                         "TYPE 5", "TYPE 6","TYPE 7", 
-                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
-  mutate(CLASS = factor(CLASS, levels = c("Flake","ElongatedProd", "Core", "RetouchedPiece"))) %>%
+                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) |>
+  mutate(CLASS = factor(CLASS, levels = c("Flake","ElongatedProd", "Core", "RetouchedPiece"))) |>
   ggplot(aes(RMTYPE, WEIGHT, fill = RMTYPE)) +
@@
 ## Janitor tabyl for final table with n and %
-proto_retouch_tab <- proto_retouch %>%
-  tabyl(RETOUCHEDPIECETYPOLOGY, RMTYPE) %>%
-  adorn_totals(c("row", "col")) %>% 
-  adorn_percentages("col") %>% 
-  adorn_pct_formatting(2) %>% 
+proto_retouch_tab <- proto_retouch |>
+  tabyl(RETOUCHEDPIECETYPOLOGY, RMTYPE) |>
+  adorn_totals(c("row", "col")) |> 
+  adorn_percentages("col") |> 
+  adorn_pct_formatting(2) |> 
   adorn_ns("front")
@@
 ## Table preparation
-gravettian_retouch <- phase_dataset %>% 
-  filter(CLASS %in% c("RetouchedPiece", "RetouchedPieceFrag")) %>%
-  filter(Level %in% c(6, 7)) %>% 
-  select(RMTYPE, RETOUCHEDPIECETYPOLOGY) %>% 
+gravettian_retouch <- phase_dataset |> 
+  filter(CLASS %in% c("RetouchedPiece", "RetouchedPieceFrag")) |>
+  filter(Level %in% c(6, 7)) |> 
+  select(RMTYPE, RETOUCHEDPIECETYPOLOGY) |> 
   dplyr::group_by(RMTYPE) |>
@@
 ## Janitor tabyl for final table with n and %
-gravettian_retouched_tab <- gravettian_retouch %>% 
-  tabyl(RETOUCHEDPIECETYPOLOGY, RMTYPE) %>%
-  adorn_totals(c("row", "col")) %>% 
-  adorn_percentages("col") %>% 
-  adorn_pct_formatting(2) %>% 
+gravettian_retouched_tab <- gravettian_retouch |> 
+  tabyl(RETOUCHEDPIECETYPOLOGY, RMTYPE) |>
+  adorn_totals(c("row", "col")) |> 
+  adorn_percentages("col") |> 
+  adorn_pct_formatting(2) |> 
   adorn_ns("front")
@@
 ## For Gravettian
-gravettian_ratiobc <- phase_dataset %>% 
-  filter(Level %in% c(6, 7)) %>% 
-  filter(RMTYPE != "INDET") %>% 
+gravettian_ratiobc <- phase_dataset |> 
+  filter(Level %in% c(6, 7)) |> 
+  filter(RMTYPE != "INDET") |> 
   mutate(CLASS = case_when(
     CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
-    TRUE ~ CLASS)) %>% 
-  filter(CLASS %in% c("Blank", "Core")) %>% 
-  count(RMTYPE, CLASS, sort = TRUE) %>% 
-  group_by(RMTYPE, CLASS) %>%
-  summarise(n=sum(n)) %>% 
-  spread(CLASS, n, fill=0) %>% 
-  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
-  filter(!sum <= 5) %>% 
+    TRUE ~ CLASS)) |> 
+  filter(CLASS %in% c("Blank", "Core")) |> 
+  count(RMTYPE, CLASS, sort = TRUE) |> 
+  group_by(RMTYPE, CLASS) |>
+  summarise(n=sum(n)) |> 
+  spread(CLASS, n, fill=0) |> 
+  mutate(sum = rowSums(across(where(is.numeric)))) |> 
+  filter(!sum <= 5) |> 
   mutate("Blank/Core" = Blank / Core)
@@
 # For Proto-Solutrean
-proto_ratiobc <- phase_dataset %>% 
-  filter(Level %in% c("4E", "4E/5")) %>% 
-  filter(RMTYPE != "INDET") %>% 
+proto_ratiobc <- phase_dataset |> 
+  filter(Level %in% c("4E", "4E/5")) |> 
+  filter(RMTYPE != "INDET") |> 
   mutate(CLASS = case_when(
     CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
-    TRUE ~ CLASS)) %>% 
-  filter(CLASS %in% c("Blank", "Core")) %>% 
-  count(RMTYPE, CLASS, sort = TRUE) %>% 
-  group_by(RMTYPE, CLASS) %>%
-  summarise(n=sum(n)) %>% 
-  spread(CLASS, n, fill=0) %>% 
-  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
-  filter(!sum <= 5) %>%
+    TRUE ~ CLASS)) |> 
+  filter(CLASS %in% c("Blank", "Core")) |> 
+  count(RMTYPE, CLASS, sort = TRUE) |> 
+  group_by(RMTYPE, CLASS) |>
+  summarise(n=sum(n)) |> 
+  spread(CLASS, n, fill=0) |> 
+  mutate(sum = rowSums(across(where(is.numeric)))) |> 
+  filter(!sum <= 5) |>
   mutate("Blank/Core" = Blank / Core)
@@
 # For Solutrean
-solutrean_ratiobc <- solutrean_dataset %>% 
+solutrean_ratiobc <- solutrean_dataset |> 
   mutate(CLASS = case_when(
@@
     CLASS == "FlakeFrag" | CLASS == "ElongBlankFrag" ~ "BlankFrag",
-    TRUE ~ CLASS)) %>% 
+    TRUE ~ CLASS)) |> 
   mutate(CLASS = case_when(
     CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
-    TRUE ~ CLASS)) %>% 
-  filter(RMTYPE != "INDET") %>% 
-  filter(CLASS %in% c("Blank", "Core")) %>% 
-  count(RMTYPE, CLASS, sort = TRUE) %>% 
-  group_by(RMTYPE, CLASS) %>%
-  summarise(n=sum(n)) %>% 
-  spread(CLASS, n, fill=0) %>% 
-  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
-  filter(!sum <= 5) %>% 
+    TRUE ~ CLASS)) |> 
+  filter(RMTYPE != "INDET") |> 
+  filter(CLASS %in% c("Blank", "Core")) |> 
+  count(RMTYPE, CLASS, sort = TRUE) |> 
+  group_by(RMTYPE, CLASS) |>
+  summarise(n=sum(n)) |> 
+  spread(CLASS, n, fill=0) |> 
+  mutate(sum = rowSums(across(where(is.numeric)))) |> 
+  filter(!sum <= 5) |> 
   mutate("Blank/Core" = Blank / Core)
@@
 # Cleaning up to create a single table for scatterplot
-gravettian_ratiobc_b <-gravettian_ratiobc %>%
-  filter(`Blank/Core` != "Inf") %>% 
-  dplyr::mutate("Assemblage" = "Gravettian") %>% 
+gravettian_ratiobc_b <-gravettian_ratiobc |>
+  filter(`Blank/Core` != "Inf") |> 
+  dplyr::mutate("Assemblage" = "Gravettian") |> 
   select(Assemblage, RMTYPE, `Blank/Core`)
-proto_ratiobc_b <- proto_ratiobc %>% 
-  filter(`Blank/Core` != "Inf") %>% 
-  dplyr::mutate("Assemblage" = "Proto-Solutrean") %>% 
+proto_ratiobc_b <- proto_ratiobc |> 
+  filter(`Blank/Core` != "Inf") |> 
+  dplyr::mutate("Assemblage" = "Proto-Solutrean") |> 
   select(Assemblage, RMTYPE, `Blank/Core`)
-solutrean_ratiobc_b <- solutrean_ratiobc %>% 
-  filter(`Blank/Core` != "Inf") %>% 
-  dplyr::mutate("Assemblage" = "Solutrean") %>% 
+solutrean_ratiobc_b <- solutrean_ratiobc |> 
+  filter(`Blank/Core` != "Inf") |> 
+  dplyr::mutate("Assemblage" = "Solutrean") |> 
   select(Assemblage, RMTYPE, `Blank/Core`)
@@
 ## For Gravettian
-gravettian_ratiotd <- phase_dataset %>% 
-  filter(Level %in% c(6, 7)) %>% 
-  filter(RMTYPE != "INDET") %>% 
+gravettian_ratiotd <- phase_dataset |> 
+  filter(Level %in% c(6, 7)) |> 
+  filter(RMTYPE != "INDET") |> 
   mutate(CLASS = case_when(
@@
     CLASS == "RetouchedPieceFrag" & PIECECOMPLETENESS == "Proximal" ~ "RetouchedPiece",
-    TRUE ~ CLASS)) %>% 
-  filter(CLASS %in% c("Blank", "RetouchedPiece")) %>% 
-  count(RMTYPE, CLASS, sort = TRUE) %>% 
-  group_by(RMTYPE, CLASS) %>%
-  summarise(n=sum(n)) %>% 
-  spread(CLASS, n, fill=0) %>% 
-  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
-  filter(!sum <= 5) %>% 
+    TRUE ~ CLASS)) |> 
+  filter(CLASS %in% c("Blank", "RetouchedPiece")) |> 
+  count(RMTYPE, CLASS, sort = TRUE) |> 
+  group_by(RMTYPE, CLASS) |>
+  summarise(n=sum(n)) |> 
+  spread(CLASS, n, fill=0) |> 
+  mutate(sum = rowSums(across(where(is.numeric)))) |> 
+  filter(!sum <= 5) |> 
   mutate("Tool/Blank" = RetouchedPiece / Blank)
@@
 ## For Proto-Solutrean
-proto_ratiotd <- phase_dataset %>% 
-  filter(Level %in% c("4E", "4E/5")) %>% 
-  filter(RMTYPE != "INDET") %>% 
+proto_ratiotd <- phase_dataset |> 
+  filter(Level %in% c("4E", "4E/5")) |> 
+  filter(RMTYPE != "INDET") |> 
   mutate(CLASS = case_when(
@@
 ## For Solutrean
-solutrean_ratiotd <- solutrean_dataset %>% 
-  filter(RMTYPE != "INDET") %>% 
+solutrean_ratiotd <- solutrean_dataset |> 
+  filter(RMTYPE != "INDET") |> 
   mutate(CLASS = case_when(
@@
     CLASS == "Retouched piece fragment" ~ "RetouchedPieceFrag",
-    TRUE ~ CLASS)) %>% 
+    TRUE ~ CLASS)) |> 
   mutate(CLASS = case_when(
@@
 # Creating a single table with all ratios by assemblage for the scatterplot
-gravettian_ratiotd_b <-gravettian_ratiotd %>% 
-  dplyr::mutate("Assemblage" = "Gravettian") %>% 
+gravettian_ratiotd_b <-gravettian_ratiotd |> 
+  dplyr::mutate("Assemblage" = "Gravettian") |> 
   select(Assemblage, RMTYPE, `Tool/Blank`)
-proto_ratiotd_b <- proto_ratiotd %>% 
-  dplyr::mutate("Assemblage" = "Proto-Solutrean") %>% 
+proto_ratiotd_b <- proto_ratiotd |> 
+  dplyr::mutate("Assemblage" = "Proto-Solutrean") |> 
   select(Assemblage, RMTYPE, `Tool/Blank`)
-solutrean_ratiotd_b <- solutrean_ratiotd %>% 
-  dplyr::mutate("Assemblage" = "Solutrean") %>% 
+solutrean_ratiotd_b <- solutrean_ratiotd |> 
+  dplyr::mutate("Assemblage" = "Solutrean") |> 
   select(Assemblage, RMTYPE, `Tool/Blank`)

### 2025-01-21 analysis/paper/paper.qmd
Added methodology. Paper rendered.

@@
 
+The current study includes the statistical correlation between the techno-typological analysis and the geoarchaeological study of the chert lithic assemblages of the UP of Vale Boi.
+As mentioned above, previous studies have individually analysed the lithic assemblages from Vale Boi focusing on different subjects and methods, obtaining different kinds of data.
+Specifically, our previous study focused on the macroscopic and petrographic characterisation of cherts, and the identification of distance to sources and proportions through the several UP technocomplexes.
+This study provided the basis for understanding which cherts were being used at the site, where they were coming from and how/why their use became more or less intense.
+In that same study, we used previously known data related to settlement type and seasonality to contextualise the changes in chert sources and proportions through time, to better understand the possible strategies of raw material procurement (Hahn, 2002).
+However, and as mentioned in Belmiro et al.
+2025, correlating the obtained data with technological analysis data provides a key step to better understand how these raw materials were being introduced into the site and used, through the correlation with data such as type of lithic class, since it may point towards the specific mobility of artifacts (Hahn, 2002).
+
+This is especially relevant in a site like Vale Boi, where several technological studies have been key to understand the cultural characteristics of the groups inhabiting throughout the UP, and their possible relations within the UP in Iberia.
+By using previously acquired data, it becomes possible to integrate different datasets to provide a wider picture of UP occupations at Vale Boi, and using the potential of lithics to inform us of hunter-gatherer lifeways.
+
+As such, this study includes datasets which have been previously interpreted and published, most of which can also be found in online repositories and freely consulted.
+As previously mentioned, the macroscopic and petrographic data of cherts was previously collected and published by our team (Belmiro et al., 2025), and can also be found in an online repository (OSF**).
+This study focused on the UP chalcedony and chert assemblages of Vale Boi, especially the Gravettian (levels 6 and 7), Proto-Solutrean (levels 4E and 5) and Solutrean (levels 4 to 4D) from the Terrace area and the Solutrean (levels A to C) from the Shelter area.
+The methodology used in the study, and supporting documents, including the chert type characterisation and analysis data dictionaries can be found in the article and the online repository.
+
+The technological analysis datasets have been collected by our team and published in previous works.
+The Solutrean assemblage technological analysis was published by Cascalheira (-Cascalheira, 2010), as part of a Master’s thesis and following publications, with the study of levels A to C of the Shelter area.
+The dataset used for this study was not originally published, but can be consulted in our online repository (++).
+Given the inexistence of technological studies for the Solutrean levels of the Terrace with a representative amount of lithics (level 4), the Solutrean analysis in this study is limited to the Shelter area.
+However, and as seen in Belmiro et al.
+(2025), the raw material patterns seem to show little differences between the Solutrean in both areas, thus coupled with the similar radiocarbon dates between area (c.
+24-22 and 24-20 ka cal BP), we assume that they are representative of similar, long-term, residential occupations which aptly characterise the Solutrean groups of Vale Boi.
+The Proto-Solutrean technological analysis was the result of a Master’s thesis (Belmiro, 2020) and following publication (Belmiro et al., 2021), with the study of levels 4E and 5 of the Terrace area.
+The dataset used for this study can be consulted in the original online repository (++), as well as in our RStudio reproducible project, available in our online repository (**).
+Finally, the Gravettian technological analysis was the result of a recent and unpublished study, focusing on the study of levels 6 and 7 of the Terrace area.
+This dataset can also be found in our online repository (++).
+Given the use of different datasets, collected through a wide span of time and by different researchers, a data dictionary with the variables used in this study can also be found in the Supplementary materials and in our online repository.
+
+An inherent caveat of using datasets collected by different researchers is related to the use of different methodologies, variables and measurements.
+This caveat is avoided in this study since the literature, approach and analysis method used in all data collection processes was similar.
+The lithic analysis adopted an attribute-based methodology, which aims to describe morphological and metrical attributes of technological classes.
+The attributes analysed followed those present in specialized literature, such as Brézillon (1968), and Tixier and Inizan (1980), paired with other Paleolithic lithic attribute analyses (see e.g., Bicho, 1992; Zilhão, 1997; Cascalheira, 2019; Almeida, 2000; Scerri et al., 2014; Tostevin, 2013).
+
+After data collection, the databases were imported into R Programming Environment, where the information was processed through the creation of descriptive statistical analysis and writing of this article.
+Some differences in the collection process, especially between the study by Cascalheira (2010) and the following studies, required homogenization.
+For example, where in Cascalheira (2010) retouch was an independent variable, with a Yes or No option followed by a variable to classify the type of retouch, in Belmiro (2020) and Belmiro et al.
+(2021), retouch was identified in the Class variable as RetouchedPiece, followed by a variable to choose the retouch typology.
+In this situation, the Solutrean dataset was corrected, so that all artifacts with retouch were classified as RetouchedPiece in the Class variable, instead of the original classification.
+The original datasets without alterations can be found in the online repository, as all the alterations to the original datasets were done through R Studio and can be found in the original R Script.
+Finally, technological and raw material data had to be merged in a single dataset.
+This was done through the merging of the several datasets through the artifact identifier “ID”, the sequential identifier which is attributed to each total station plot during excavation, whether relating to individual artifacts plotted in their exact location, or bucket coordinates, relating to the center of the excavated area, and attributed to all artifacts found in the sieved sediment in the bucket.
+Since the raw material study only focused piece plotted artifacts with individual coordinates (leaving out bucket artifacts), as these correspond to artifacts with dimensions superior to 2 cm or complete bladelets (thus excluding small fragments and chips), the final dataset used in this study does not directly correspond to the full datasets analysed in the original studies.
+Similarly, the raw material study used only currently accessible artifacts from the assemblages.
+This is especially relevant for the Shelter Solutrean chert assemblage, where shatter is absent from the sample due to the availability for study.
+This stands as a caveat for the current study, since it is using only portion of the assemblages, leaving out possibly important debris.
+Despite this, the current study still offers a novel and key approach to the study of lithic assemblages at the site, with a methodology that includes all the knapping sequence, of a large number of artifacts from the UP sequence of a single site, as seen in table **, which shows the number of artifacts used in the final analysis of this paper, after corrections, homogenization and merging.
+
+After the homogenization and merging of the datasets, descriptive statistical analysis was applied to describe and interpret the relation between the raw material and technological data.
+As mentioned above, the goal of the study was to test two hypotheses related to the technological organisation of local and non-local cherts, and between different UP occupations.
+To test these hypotheses, we evaluated several parameters for each dataset and assemblage and compared the results between chert types and assemblages.
+The parameters evaluated in the current study were: P1) frequency of retouched tools; P2) frequency of maintenance products; P3) debitage sizes; P4) tools-to-debitage ratios; P5) core sizes; P6) blank-to-core ratios; P7) retouched tool sizes; P8) tool diversity; P9) tool specialization.
+Debitage, core and retouched tool sizes were analysed through descriptive statistics (mean, sd, min and max, tables available in the SOM**) and plotted through boxplots.
+The ratios were obtained through the division of each group (e.g., tools by debitage and blanks by core), by chert type and assemblage.
+All other parameters were observed through number and percentages tables.
+All the analyses were done through code in R Studio.
+
+The entirety of the R code used for the analysis, datasets, and visual representations contained in this paper can be accessed through our online research compendium.
+We used the rrtools package by Marwick et al.
+[-@marwickPackagingDataAnalytical2018] to create a research compendium and write a reproducible journal article.
+The provided files include the complete set of raw data used in the analysis, along with a custom R project [@wickhamPackagesOrganizeTest2015] containing the code required to generate all tables and figures.
+To enable maximum reuse, the code is made available under the MIT license, data under CC‐0, and figures under CC‐BY [additional details can be found in @marwickComputationalReproducibilityArchaeological2017].
+
 # Results

### 2025-01-21 analysis/paper/paper.qmd
Corrected missing graphs and render.

@@
 
-gravettian_graphs <- ggpubr::ggarrange(gravettian_scatter, gravettian_stats_weight, labels = c("c1", "c2"), ncol = 2, common.legend = TRUE, legend = "bottom")
+ggpubr::ggarrange(gravettian_scatter, gravettian_stats_weight, labels = c("a", "b"), ncol = 2, common.legend = TRUE, legend = "bottom")
 
@@
 
-proto_graphs <- ggpubr::ggarrange(proto_scatter, proto_stats_weight, labels = c("b1", "b2"), ncol = 2, common.legend = TRUE, legend = "bottom")
+ggpubr::ggarrange(proto_scatter, proto_stats_weight, labels = c("b1", "b2"), ncol = 2, common.legend = TRUE, legend = "bottom")
 
@@
 
-solutrean_graphs <- ggpubr::ggarrange(solutrean_scatter, solutrean_stats_weight, labels = c("a1", "a2"), ncol = 2, common.legend = TRUE, legend = "bottom")
+ggpubr::ggarrange(solutrean_scatter, solutrean_stats_weight, labels = c("a", "b"), ncol = 2, common.legend = TRUE, legend = "bottom")
 

### 2025-01-21 analysis/paper/paper.qmd
Removed % from typology table to make them cleaner + render

@@
   adorn_totals(c("row", "col")) |>
-  adorn_percentages("col") |>
-  adorn_pct_formatting(2) |> 
-  adorn_ns("front")
-## Cleaning final table
-solutrean_retouch_tab <- solutrean_retouch_tab |> # Replace 0% values with -
-  dplyr::mutate(T1 = case_when(T1 == "0   (0.00%)" ~ "-", TRUE ~ T1)) |>
-  dplyr::mutate(T2 = case_when(T2 == "0   (0.00%)" ~ "-", TRUE ~ T2)) |>
-  dplyr::mutate(T5 = case_when(T5 == "0   (0.00%)" ~ "-", TRUE ~ T5)) |>
-  dplyr::mutate(T6 = case_when(T6 == "0   (0.00%)" ~ "-", TRUE ~ T6)) |>
-  dplyr::mutate(T7 = case_when(T7 == "0   (0.00%)" ~ "-", TRUE ~ T7)) |>
-  dplyr::mutate(T8 = case_when(T8 == "0   (0.00%)" ~ "-", TRUE ~ T8)) |>
-  dplyr::mutate(T11 = case_when(T11 == "0   (0.00%)" ~ "-", TRUE ~ T11)) |>
-  dplyr::mutate(INDET = case_when(INDET == "0   (0.00%)" ~ "-", TRUE ~ INDET)) |>
-  select(RETOUCHEDPIECETYPOLOGY, T1, T2, T5, T6, T7, T8, T11, INDET, Total) # Ordering values
+  mutate(across(where(is.numeric), ~ ifelse(. == 0, " ", .))) |>
+  select(RETOUCHEDPIECETYPOLOGY, T1, T2, T5, T6, T7, T8, T11, INDET, Total) |> # Ordering values
+  rename(Typology = RETOUCHEDPIECETYPOLOGY)
 
@@
   adorn_totals(c("row", "col")) |> 
-  adorn_percentages("col") |> 
-  adorn_pct_formatting(2) |> 
-  adorn_ns("front")
-## Cleaning final table
-proto_retouch_tab <- proto_retouch_tab |> # Replace 0% values with -
-  dplyr::mutate(T1 = case_when(T1 == "0   (0.00%)" ~ "-", TRUE ~ T1)) |>
-  dplyr::mutate(T2 = case_when(T2 == "0   (0.00%)" ~ "-", TRUE ~ T2)) |>
-  dplyr::mutate(T3 = case_when(T3 == "0   (0.00%)" ~ "-", TRUE ~ T3)) |>
-  dplyr::mutate(T4 = case_when(T4 == "0   (0.00%)" ~ "-", TRUE ~ T4)) |>
-  dplyr::mutate(T5 = case_when(T5 == "0   (0.00%)" ~ "-", TRUE ~ T5)) |>
-  dplyr::mutate(T6 = case_when(T6 == "0   (0.00%)" ~ "-", TRUE ~ T6)) |>
-  dplyr::mutate(T7 = case_when(T7 == "0   (0.00%)" ~ "-", TRUE ~ T7)) |>
-  dplyr::mutate(T8 = case_when(T8 == "0   (0.00%)" ~ "-", TRUE ~ T8)) |>
-  dplyr::mutate(T9 = case_when(T9 == "0   (0.00%)" ~ "-", TRUE ~ T9)) |>
-  dplyr::mutate(TL = case_when(TL == "0   (0.00%)" ~ "-", TRUE ~ TL)) |>
-  dplyr::mutate(INDET = case_when(INDET == "0   (0.00%)" ~ "-", TRUE ~ INDET))
+  mutate(across(where(is.numeric), ~ ifelse(. == 0, " ", .))) |>
+  select(RETOUCHEDPIECETYPOLOGY, T1, T2, T3, T4, T5, T6, T7, T8, T9, TL, INDET, Total) |> # Ordering values
+  rename(Typology = RETOUCHEDPIECETYPOLOGY)
 
@@
   adorn_totals(c("row", "col")) |>
-  adorn_percentages("col") |> 
-  adorn_pct_formatting(2) |> 
-  adorn_ns("front")
-## Cleaning final table
-gravettian_retouched_tab <- gravettian_retouched_tab |> # Replace 0% values with -
-  dplyr::mutate(T1 = case_when(T1 == "0   (0.00%)" ~ "-", TRUE ~ T1)) |>
-  dplyr::mutate(T2 = case_when(T2 == "0   (0.00%)" ~ "-", TRUE ~ T2)) |>
-  dplyr::mutate(T3 = case_when(T3 == "0   (0.00%)" ~ "-", TRUE ~ T3)) |>
-  dplyr::mutate(T4 = case_when(T4 == "0   (0.00%)" ~ "-", TRUE ~ T4)) |>
-  dplyr::mutate(T5 = case_when(T5 == "0   (0.00%)" ~ "-", TRUE ~ T5)) |>
-  dplyr::mutate(T6 = case_when(T6 == "0   (0.00%)" ~ "-", TRUE ~ T6)) |>
-  dplyr::mutate(T7 = case_when(T7 == "0   (0.00%)" ~ "-", TRUE ~ T7)) |>
-  dplyr::mutate(T8 = case_when(T8 == "0   (0.00%)" ~ "-", TRUE ~ T8)) |>
-  dplyr::mutate(T9 = case_when(T9 == "0   (0.00%)" ~ "-", TRUE ~ T9)) |>
-  dplyr::mutate(T10 = case_when(T10 == "0   (0.00%)" ~ "-", TRUE ~ T10)) |>
-  dplyr::mutate(TL = case_when(TL == "0   (0.00%)" ~ "-", TRUE ~ TL)) |>
-  dplyr::mutate(INDET = case_when(INDET == "0   (0.00%)" ~ "-", TRUE ~ INDET)) |>
-  select(RETOUCHEDPIECETYPOLOGY, T1, T2, T3, T4, T5, T6, T7, T8, T9, T10, TL, INDET, Total) # Ordering values
+  mutate(across(where(is.numeric), ~ ifelse(. == 0, " ", .))) |>
+  select(RETOUCHEDPIECETYPOLOGY, T1, T2, T3, T4, T5, T6, T7, T8, T9, T10, TL, INDET, Total) |> # Ordering values
+  rename(Typology = RETOUCHEDPIECETYPOLOGY)
 

### 2025-01-21 analysis/paper/paper.qmd
Small changes do main doc, template + render
TL table with classes to understand how this specific group of cherts relates to the technological variables.

@@
 
-ggpubr::ggarrange(proto_scatter, proto_stats_weight, labels = c("b1", "b2"), ncol = 2, common.legend = TRUE, legend = "bottom")
+ggpubr::ggarrange(proto_scatter, proto_stats_weight, labels = c("a", "b"), ncol = 2, common.legend = TRUE, legend = "bottom")
 
@@
 Following the same trend as the Gravettian, the non-local cherts show slightly higher ratios (0.3-0.5) of retouched to debitage.
+
+```{r}
+#| label: tbl-TL-classes
+#| tbl-cap: Trace lithologies cherts from the Gravettian and Proto-Solutrean assemblages of the Terrace and Solutrean assemblage from the Shelter and corresponding technological class.
+
+TL_classes_terrace <- phase_dataset |> 
+  filter(RMTYPE == "TL") |> 
+  select(Technocomplex, SUBTYPE, ID, CLASS, RETOUCHEDPIECETYPOLOGY)
+
+TL_classes_shelter <- solutrean_dataset |> 
+  mutate(Technocomplex = case_when(
+    LEVEL == "A" | LEVEL == "B" | LEVEL == "C" ~ "Solutrean")) |> 
+  filter(RMTYPE == "TL") |> 
+  select(Technocomplex, SUBTYPE, ID, CLASS, RETOUCHEDPIECETYPOLOGY)
+
+TL_classes <- rbind(TL_classes_terrace, TL_classes_shelter)
+TL_classes <- TL_classes |> 
+  rename(TL = SUBTYPE) |> 
+  arrange(TL)
+
+kable(head(TL_classes))
+
+```
 

### 2025-01-22 analysis/paper/paper.qmd
Final daily commit. Graph and table improvements. Results description improved and restructured (on-going). Document rendered.

@@
 Recent research has expanded our understanding of mobility and social networks through lithic raw material studies.
-Our team’s recent work (Belmiro et al., 2025) identified non-local cherts originating from central Portugal (>250 km from Vale Boi) and souuthern Spain (>200 km from Vale Boi).
+Our team’s recent work (Belmiro et al., 2025) identified non-local cherts originating from central Portugal (\>250 km from Vale Boi) and souuthern Spain (\>200 km from Vale Boi).
 The proportions of these materials varied across the UP, with earlier occupations (Gravettian) showing high frequencies of non-local chert, while later ones (Proto-Solutrean and Solutrean) exhibited a marked decline.
@@
 Between 2012 and 2016, eight primary litho-stratigraphic units were identified.
-The UP sequence includes multiple occupations, spanning ~10,000 years: Gravettian (levels 8 to 6) between approximately c.
+The UP sequence includes multiple occupations, spanning \~10,000 years: Gravettian (levels 8 to 6) between approximately c.
 32 and 27 ka cal BP, Proto-Solutrean (levels 5 to 4E) between approximately c.
@@
 
-The lithic analysis of the Shelter Solutrean levels (A-C) shows a high number of lithic artifacts, with over 22 000 artifacts analysed (including chips which correspond to ~60% of the assemblage and shatter ~17-19%).
+The lithic analysis of the Shelter Solutrean levels (A-C) shows a high number of lithic artifacts, with over 22 000 artifacts analysed (including chips which correspond to \~60% of the assemblage and shatter \~17-19%).
 The volumes introduced into the site were mostly cortical, given the high frequencies of cortex in blanks and retouched tools.
@@
 
-The Proto-Solutrean horizons yielded more than 26,500 artifacts (including chips which correspond to ~65-70% of the assemblage and shatter ~20%), ~11,000 attributed to the lower bottom of level 5 and ~15,600 attributed to the top of level 5 and level 4E.
+The Proto-Solutrean horizons yielded more than 26,500 artifacts (including chips which correspond to \~65-70% of the assemblage and shatter \~20%), \~11,000 attributed to the lower bottom of level 5 and \~15,600 attributed to the top of level 5 and level 4E.
 Both horizons are characterised by knapping strategies dominated by unidirectional prismatic cores, with little platform preparation, focused on the obtention of bladelets (in bottom level 5) and flakes.
@@
 Similarly, different raw materials have been observed to play specific roles, such as the case of chert for the production of a high percentage of blanks and retouched tools.
-While quartz and greywacke can be found locally, in the streams and adjacent to the site, chert sources are further away, although still within a ~20 km radius of the site.
+While quartz and greywacke can be found locally, in the streams and adjacent to the site, chert sources are further away, although still within a \~20 km radius of the site.
 Our previous study showed, as expected, the existence of local chert in the assemblages, originating from the Lower Jurassic outcrops in proximity to the site (Belmiro et al., 2025).
@@
 This method allowed the confirmation that, except from the Early Gravettian and Magdalenian occupations, all other occupations are composed of high-density assemblages with low degree of retouch, correspondent to a residential base-camp occupation, for extended periods of time.
-Regarding the Terrace area, however, the results were based on previous excavations (<2012), and the identification of new layers and reorganization of the occupations is yet to be tested through the use of lithic techno-typological methods.
+Regarding the Terrace area, however, the results were based on previous excavations (\<2012), and the identification of new layers and reorganization of the occupations is yet to be tested through the use of lithic techno-typological methods.
 
 A similar conclusion was suggested based on the interpretation of the chert sourcing and proportions in the several UP occupations of the Terrace and Shelter of Vale Boi, and following other models relating local/non-local raw materials with mobility (Grove et al., 2023; Surovell, 2009).
-The high proportions of non-local chert (~50%) in the Gravettian occupations (levels 6 and 7), and gradual reduction in the following occupations, were interpreted as a possible result of settlement changes, where the Gravettian occupations corresponded to short-term occupations, and the other occupations corresponded to residential, long-term occupations with more infrequent moves (Belmiro et al., 2025).
+The high proportions of non-local chert (\~50%) in the Gravettian occupations (levels 6 and 7), and gradual reduction in the following occupations, were interpreted as a possible result of settlement changes, where the Gravettian occupations corresponded to short-term occupations, and the other occupations corresponded to residential, long-term occupations with more infrequent moves (Belmiro et al., 2025).
 
@@
 As such, this study includes datasets which have been previously interpreted and published, most of which can also be found in online repositories and freely consulted.
-As previously mentioned, the macroscopic and petrographic data of cherts was previously collected and published by our team (Belmiro et al., 2025), and can also be found in an online repository (OSF**).
+As previously mentioned, the macroscopic and petrographic data of cherts was previously collected and published by our team (Belmiro et al., 2025), and can also be found in an online repository (OSF\*\*).
 This study focused on the UP chalcedony and chert assemblages of Vale Boi, especially the Gravettian (levels 6 and 7), Proto-Solutrean (levels 4E and 5) and Solutrean (levels 4 to 4D) from the Terrace area and the Solutrean (levels A to C) from the Shelter area.
@@
 The Proto-Solutrean technological analysis was the result of a Master’s thesis (Belmiro, 2020) and following publication (Belmiro et al., 2021), with the study of levels 4E and 5 of the Terrace area.
-The dataset used for this study can be consulted in the original online repository (++), as well as in our RStudio reproducible project, available in our online repository (**).
+The dataset used for this study can be consulted in the original online repository (++), as well as in our RStudio reproducible project, available in our online repository (\*\*).
 Finally, the Gravettian technological analysis was the result of a recent and unpublished study, focusing on the study of levels 6 and 7 of the Terrace area.
@@
 This stands as a caveat for the current study, since it is using only portion of the assemblages, leaving out possibly important debris.
-Despite this, the current study still offers a novel and key approach to the study of lithic assemblages at the site, with a methodology that includes all the knapping sequence, of a large number of artifacts from the UP sequence of a single site, as seen in table **, which shows the number of artifacts used in the final analysis of this paper, after corrections, homogenization and merging.
+Despite this, the current study still offers a novel and key approach to the study of lithic assemblages at the site, with a methodology that includes all the knapping sequence, of a large number of artifacts from the UP sequence of a single site, as seen in table \*\*, which shows the number of artifacts used in the final analysis of this paper, after corrections, homogenization and merging.
 
@@
 The parameters evaluated in the current study were: P1) frequency of retouched tools; P2) frequency of maintenance products; P3) debitage sizes; P4) tools-to-debitage ratios; P5) core sizes; P6) blank-to-core ratios; P7) retouched tool sizes; P8) tool diversity; P9) tool specialization.
-Debitage, core and retouched tool sizes were analysed through descriptive statistics (mean, sd, min and max, tables available in the SOM**) and plotted through boxplots.
+Debitage, core and retouched tool sizes were analysed through descriptive statistics (mean, sd, min and max, tables available in the SOM\*\*) and plotted through boxplots.
 The ratios were obtained through the division of each group (e.g., tools by debitage and blanks by core), by chert type and assemblage.
@@
   adorn_pct_formatting(2) |> 
-  adorn_ns("front")
+  adorn_ns("front") |> 
+  mutate(across(where(is.character), ~ gsub("^0   \\(0\\.00%\\)$", " ", .)))
 
@@
   mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) |>
-  ggplot(aes(x = LENGTH, y = MAXWIDTH, colour = RMTYPE)) +
-  geom_point(alpha = 0.9, size = 3) +
+  ggplot(aes(x = LENGTH, y = MAXWIDTH, fill = RMTYPE)) +
+  geom_point(alpha = 0.9, size = 3, shape = 21, colour = "grey") +
   facet_wrap(~CLASS) +
   labs(x="Length (mm)", y="Max. Width (mm") +
-  scale_colour_manual(values = c("#8B7B8B", "#FFE1FF", "#EED2EE", "#D8BFD8",
+  scale_fill_manual(values = c("#8B7B8B", "#FFE1FF", "#EED2EE", "#D8BFD8",
                               "#CDB5CD", "#2B4F60" , "#4A708B", "#798777",
@@
   geom_boxplot(aes(fill=factor(RMTYPE)), outlier.shape=21) +
-  facet_wrap(~CLASS) +
+  facet_wrap(~CLASS, scales = "free_y") +
   stat_summary(fun.y=mean, geom="point", shape=23) +
   labs(x="Raw material types" ,y="Weight (gr)") +
-  coord_cartesian(ylim =  c(0, 50)) +
+  #coord_cartesian(ylim =  c(0, 50)) +
   scale_fill_manual(values = c("#8B7B8B", "#FFE1FF", "#EED2EE", "#D8BFD8",
@@
   geom_boxplot(aes(fill=factor(RMTYPE)), outlier.shape=21) +
-  facet_wrap(~CLASS) +
+  facet_wrap(~CLASS, scales = "free_y") +
   stat_summary(fun.y=mean, geom="point", shape=23) +
   labs(x="Raw material types", y="Weight (gr)") +
-  coord_cartesian(ylim =  c(0, 50)) +
+  #coord_cartesian(ylim =  c(0, 50)) +
   scale_fill_manual(values = c("#8B7B8B", "#FFE1FF", "#EED2EE", "#D8BFD8",
@@
   filter(WEIGHT !="0" & !RMTYPE %in% c("INDET","GB","TL")) |>
+  filter(!(RMTYPE == "TYPE 4" & WEIGHT > 40)) |> # Removed a single outlier at 60 weight while ensuring that the data remains consistent. Only mean changed significantly and was normalised
   mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
@@
   geom_boxplot(aes(fill=factor(RMTYPE)), outlier.shape=21) +
-  facet_wrap(~CLASS) +
+  facet_wrap(~CLASS, scales = "free_y") +
   stat_summary(fun.y=mean, geom="point", shape=23) +
   labs(x= "Raw material types", y="Weight (gr)") +
-  coord_cartesian(ylim =  c(0, 60)) +
+  #coord_cartesian(ylim =  c(0, 60)) +
   scale_fill_manual(values = c("#8B7B8B", "#FFE1FF", "#EED2EE", "#D8BFD8",
@@
 
-The sample studied for the Gravettian occupation (>2012 excavations) of the Terrace area included 766 artifacts.
-When crossing the lithic class with chert type, we see that the local cherts show similar patterns, with T2 and T5 being the types with higher numbers of artefacts (n=110 and n=72, respectively).
-For all local cherts, blanks represent more than 50% of the artefacts.
-Cores and core maintenance products are only present in T2 and T5, although in small frequencies: for T2, cores represent 13.6% of the group and core preparation and maintenance products represent 1.8% (n=2).
-For T5 only 4 core fragments and 1 core preparation and maintenance product are present.
-Retouched tools seem to be especially relevant in T2 (~18%), being the most frequent class after blanks.
-The presence of burin spalls (8.1%) in T2 is expected due to the prevalence of burins (see below).
-In comparison, T5 shows similar percentages of retouched tools and shatter (~12% and ~14%, respectively).
-T1 (chalcedony) shows similar patterns, although with higher percentages of shatter (34.3%).
-In comparison, non-local chert types (especially T6 with n=169 and T7 with n=86) show similar percentages of blanks (~40-50%) but higher percentages of retouched tools (~30%).
-Cores, core fragments and core preparation and maintenance products are also present even if in small percentages (<4%), as well as shatter (~12%).
-In both chert types burin spalls are also present (7.1% for T6 and 4.6% for T7).
-The TL cherts show, as expected, blanks and blank fragments (n=3 and n=2, respectively), and retouched tools (n=4).
-The presence of core maintenance products in TL is possibly related to the limited knapping of TL01 (jasper).
+The sample studied for the Gravettian occupation (\>2012 excavations) of the Terrace area included 766 artifacts.
+Of these, 209 (`r round(209*100/766, 2)`%) have an indeterminate chert type (@tbl-general-gravettian).
+Chalcedony (T1) and local cherts (T2-5) show a total of `r 35+110+17+13+72` artefacts (`r round((35+110+17+13+72)*100/766, 2)`%).
+Non-local cherts show different numbers of artefacts depending on the type, with T6 and T7 showing the highest values (169 or `r round(169*100/766, 2)`% of the total assemblage, and 86 or `r round(86*100/766, 2)`%, respectively); the other non-local cherts have artefact counts lower than 20 per type.
+The TL cherts also show smaller numbers, with 12 artefacts.
+
+@tbl-general-gravettian shows the number and percentages of technological classes per chert type.
+Local cherts show similar patterns between them, with T2 and T5 being the types with higher numbers of artefacts (n=110 and n=72, respectively).
+For all local cherts, blanks (including blank fragments) represent more than 50% of the artefacts.
+Cores and core maintenance products are only present in T2 and T5, although in small frequencies: for T2, cores represent 13.6% (n=15 including core fragments) of the group and core preparation and maintenance products represent 1.8% (n=2).
+For T5 only 4 core fragments and 1 core preparation and maintenance product are present, all frequencies between \~5-1%.
+Retouched tools seem to be especially relevant in T2 (n= 20, \~18% including retouched piece fragments), making them the most frequent class after blanks.
+T5 shows similar percentages of retouched tools and shatter (\~12% and \~14%, respectively).
+T1 (chalcedony) shows similar patterns to other local cherts, although with higher percentages of shatter (34.3%).
+
+When compared to local cherts, non-local chert types (especially T6 and T7) show similar percentages of blanks (n=`r 44+24` or 40% and `r 21+21`\`or \~50%, respectively) but higher percentages of retouched tools (\~30%).
+Cores, core fragments and core preparation and maintenance products are also present in all non-local cherts (excluding TL) even if in percentages below 4% (n\<4), as well as shatter (\~12%).
+The exception is T6 that shows the highest numbers of cores and core fragments (n=8), although with low frequencies (4.6%).
+The TL cherts show mostly blanks and blank fragments (n=3 and n=2, respectively), and retouched tools (n=4).
 
@@
 
-As seen in table **, 170 retouched tools were identified, with 22.9% (n=39) having an unidentified chert type.
-From the identified chert types, 24.1% of retouched tools are from local types and 52.9% from non-local cherts.
+As seen in table @tbl-retouch-gravettian, 170 retouched tools were identified, with 39 (`r round(39*100/170, 2)`%) having an unidentified chert type.
+From the identified chert types, `r 3+20+4+2+9` (`r round((3+20+4+2+9)*100/170, 2)`%) of retouched tools are from local types, `r 58+24+3+1+3` (`r round((58+24+3+1+3)*100/170, 2)`%) from non-local cherts and 4 are from the TL group.
+From all chert types, T6 shows the highest numbers of retouched pieces (n=58, `r round(58*100/170, 2)`%).
 Regarding typologies, 15 types of retouched tools were identified (excluding retouched tool fragments).
 All these typologies were found in the non-local chert group, while only 7 retouched tool typologies were identified in local cherts/chalcedony.
-The most common typologies identified in the assemblage were burins (n=52), endscrappers (n=17), retouched flakes (n=30) and splintered pieces (n=26).
+The most common typologies identified in the assemblage (and for all chert types) were burins (n=52), endscrappers (n=17), retouched flakes (n=30) and splintered pieces (n=26).
 From all these, the majority were produced in non-local chert, mainly T6 and in lesser numbers T7.
 Although in lesser numbers as well, backed bladelets were also identified (n=6), but only present in non-local chert types (T6 and T7).
+The presence of burin spalls in local chert, especially T2 (n=9) and non-local cherts (T6 with n=12 and T7 with n=4) is expected due to the prevalence of burins (@tbl-general-gravettian).
 
@@
 
-Regarding the sizes of the lithic artifacts (width, length and weight), blanks (flakes and elongated products) and retouched tools show little differences in the measurements of artefacts of local and non-local chert types.
-The only exception are T6 and T7 flakes, which show slightly lower weight means (~2.6-3.0 gr) and lower maximum length (~20-22 mm) than the other types.
-For cores, T2 shows bigger (heavier) values than T6 (means of 21.3 and 12.6, and maximum weight of 37.2 gr and 22.5 gr, respectively), even if they seem to vary in morphology, with little patterns regarding thickness, width and length.
-
-```{r}
-#| label: fig-gravettian-boxplots
-#| fig-cap: Boxplots Solutrean
-
-ggpubr::ggarrange(gravettian_scatter, gravettian_stats_weight, labels = c("a", "b"), ncol = 2, common.legend = TRUE, legend = "bottom")
-
-```
-
-As mentioned in the methodology, two different ratios were calculated: blank-to-core and tools-to-debitage.
-In general, non-local chert types show higher values for these ratios.
-Local types (T1 and T2) show lower ratios (<4) of blanks to cores, although other local types (T3-T5) have no identified cores.
-The non-local types, especially T6 and T7 show higher ratios (~8) of blanks to cores.
-T1 to T5 show low ratios (0.25-0.5) of retouched to debitage, while TL and T6-T8 show higher ratios (0.9-1.0).
-The exception is T10 (non-local type), which shows a lower ratio (0.28), more akin to local cherts.
+Retouched tools were further explored through the tools-to-debitage ratio (@fig-ratiotd and table s++).
+As seen in @fig-ratiotd, local cherts (T1-5) show low ratios (\~2.5-0.5) of retouched tools to blanks, while TL and T6-T8 show higher ratios (\~0.9-1.0).
+The exception is T10 (non-local type), which shows a lower ratio (\~0.3), more akin to local cherts.
+The relation between blank and core numbers was also explored through a blank-to-core ratio (@fig-ratiobc and table s++).
+The results show that local types T1 and T2 show lower ratios (\<4) of blanks to cores, although other local types (T3-T5) have no identified cores @tbl-general-gravettian.
+The non-local types, especially T6 and T7 show higher ratios (\~8) of blanks to cores, with T10 once again showing values similar to local cherts (\~4).
 
@@
 
+Regarding the sizes of the lithic artifacts (width, length and weight), blanks (flakes and elongated products) and retouched tools show little differences in the measurements of artefacts of local and non-local chert types.
+The only exception are T6 and T7 flakes, which show slightly lower weight means (\~2.6-3.0 gr) and lower maximum length (\~20-22 mm) than the other types.
+For cores, T2 shows bigger (heavier) values than T6 (means of 21.3 and 12.6, and maximum weight of 37.2 gr and 22.5 gr, respectively), even if they seem to vary in morphology, with little patterns regarding thickness, width and length.
+
+Regarding the weight (@fig-gravettian-boxplots b), although there is some variability, non-local types T6 and T7 seem to have the lowest weights in all classes (flakes, elongated blanks, cores and retouched tools).
+The lowest means for flakes are for T6 and T7 (\~2.5 gr).
+However, in general, the lightest flakes seem to be from T2 and T6, although T2 also shows several heavy outliers.
+Elongated products show even smaller means for T6 and T7 (0.5-1 gr), especially when compared to local cherts (\~2.5 gr).
+Regarding cores, @tbl-general-gravettian shows that type T2 and T6 have the highest number of cores; between these two types, T6 has the lightest cores, with a maximum of \~21 gr and with a mean of \~12 gr, in comparison to T2 with a maximum of \~37 gr and a mean of \~21 gr.
+Retouched tools shows similar median values among all chert types (2.5-4 gr), although T6 and T7 have the lowest means (\~4 gr); in general, and given the high number of retouched tools for T6 (@tbl-general-gravettian), this non-local chert seems to have been consistently reduced with a third quartile ending in the 5 gr mark.
+
+```{r}
+#| label: fig-gravettian-boxplots
+#| fig-cap: Boxplots Solutrean
+
+ggpubr::ggarrange(gravettian_scatter, gravettian_stats_weight, labels = c("a", "b"), ncol = 2, common.legend = TRUE, legend = "bottom")
+
+```
+
 ## Proto-Solutrean (levels 4E and top of level 5)
@@
 The final lithic sample from the Proto-Solutrean assemblage, consisting of the levels 4E and top of level 5, consists of 635 artifacts.
-When crossing the lithic class with chert type, we see that the local types show high frequency of blanks (>50%, including blank fragments), followed by shatter (~11-27%), cores (~9-12%) and retouched tools (~5-15%).
+When crossing the lithic class with chert type, we see that the local types show high frequency of blanks (\>50%, including blank fragments), followed by shatter (\~11-27%), cores (\~9-12%) and retouched tools (\~5-15%).
 A single burin spall was identified in T4.
@@
 For non-local types, especially those with higher number of artefacts (T6-T8), blanks and blank fragments are the most frequent class (48.3%, 53.1% and 69.2%, respectively).
-Retouched tools are present in smaller frequencies for the previous mentioned chert types (~17-23%).
+Retouched tools are present in smaller frequencies for the previous mentioned chert types (\~17-23%).
 For T6 the absence of cores but the presence of shatter and a single core preparation and maintenance product suggests knapping on site.
@@
 
-Regarding the debitage, cores and tool sizes, flakes show slightly higher means in local cherts (~21-22 mm for length and 3.5-4.5 gr for weight), in comparison to non-local cherts, especially T6-T8 (~19-20 mm for length and 1.3-2.6 gr in weight), with exception of chalcedony which shows consistently low sizes (21.8 mm for length and 2.8 gr for weight).
+Regarding the debitage, cores and tool sizes, flakes show slightly higher means in local cherts (\~21-22 mm for length and 3.5-4.5 gr for weight), in comparison to non-local cherts, especially T6-T8 (\~19-20 mm for length and 1.3-2.6 gr in weight), with exception of chalcedony which shows consistently low sizes (21.8 mm for length and 2.8 gr for weight).
 Elongated blank however, show no differences between local and non-local chert types.
-In retouched tools, weight also shows smaller means for non-local cherts (~3-2 gr), although with higher standard deviation (~1.6).
+In retouched tools, weight also shows smaller means for non-local cherts (\~3-2 gr), although with higher standard deviation (\~1.6).
 Regarding cores, the sample size for non-local cherts is small and only T7 statistics could be obtained.
@@
 Cores are also present (10.5%), alongside a small percentage of core preparation and maintenance products (1.3%).
-In comparison to debitage products, retouched tools are scarce in the sample (~3%).
+In comparison to debitage products, retouched tools are scarce in the sample (\~3%).
 The non-local cherts most present in the Solutrean of the Shelter area are T6 (n=86) and T8 (n=30).
-These show similar patterns to the local cherts, with high percentages of blanks (>50%).
-Cores and core preparation and maintenance are also present in small percentages (~3-6% and 3.5% only for T6, respectively), as well as retouched tools (<10%).
+These show similar patterns to the local cherts, with high percentages of blanks (\>50%).
+Cores and core preparation and maintenance are also present in small percentages (\~3-6% and 3.5% only for T6, respectively), as well as retouched tools (\<10%).
 However, these specific results are truncated by the lack of shatter, which were not present in the sample used for the study (see Materials and Methods section).
@@
 
-Regarding the blank-to-core and tools-to-debitage ratios, the Solutrean assemblage shows similar patterns as the previous assemblages, with the local cherts showing lower blank to core ratios (<7), while the non-local cherts show higher ratios (>9).
+Regarding the blank-to-core and tools-to-debitage ratios, the Solutrean assemblage shows similar patterns as the previous assemblages, with the local cherts showing lower blank to core ratios (\<7), while the non-local cherts show higher ratios (\>9).
 Similar to the Proto-Solutrean, chalcedony cores were also not identified in the studied sample.
-Similarly, local cherts (T2-T5) show low retouch to debitage ratios (<0.1), with the exception of T1 (chalcedony), which shows a ratio of 1.8.
+Similarly, local cherts (T2-T5) show low retouch to debitage ratios (\<0.1), with the exception of T1 (chalcedony), which shows a ratio of 1.8.
 Other non-local cherts (T6-T7 and T11) show slightly higher retouch to debitage ratios (0.1-0.3).

### 2025-01-22 analysis/paper/paper.qmd
Improved results main text for class/type and retouched tool tables.

@@
 The final lithic sample from the Proto-Solutrean assemblage, consisting of the levels 4E and top of level 5, consists of 635 artifacts.
-When crossing the lithic class with chert type, we see that the local types show high frequency of blanks (\>50%, including blank fragments), followed by shatter (\~11-27%), cores (\~9-12%) and retouched tools (\~5-15%).
-A single burin spall was identified in T4.
-There is also the absence of core preparation and maintenance products across all local cherts.
-T1 (chalcedony) is composed mostly of blanks and a retouched tool (Vale Comprido point).
-The absence of cores but presence of shatter may hint at the knapping of this raw material at the site, but the absence of formal cores perhaps due to the quality of nodules/lenses, which are characterised by fractures.
+Of these, 167 (`r round(167*100/635, 2)`%) have an indeterminate chert type.
+Local cherts correspond to `r 32+62+32+158+51` artefacts (`r round((32+62+32+158+51)*100/635, 2`%), with T4 being the most present local chert (n=158 or `r round(158*100/(32+62+32+158+51), 2)`% of the local types).
+Non-local types are present in smaller numbers (n=`r 60+47+13+9+1` or `r round((60+47+13+9+1)*100/635, 2)`%), with T6 and T7 being the most relevant (n=60 and n=47, respectively), and the TL group includes 3 artefacts.
+
+When crossing the lithic class with chert type (@tbl-general-proto), we see that the local types show high frequency of blanks (\>50%, including blank fragments), followed by shatter (\~11-27%).
+Cores are present in small numbers (n=`r 4+5+13+6`), corresponding to ~9-12% of the classes in each local type, with the exception of T1 (chalcedony), with no cores.
+Core preparation and maintenance products are absent in all local cherts.
 For non-local types, especially those with higher number of artefacts (T6-T8), blanks and blank fragments are the most frequent class (48.3%, 53.1% and 69.2%, respectively).
 Retouched tools are present in smaller frequencies for the previous mentioned chert types (\~17-23%).
-For T6 the absence of cores but the presence of shatter and a single core preparation and maintenance product suggests knapping on site.
-Similarly, T7 shows the presence of cores and core fragments (8.4%), shatter (17.0%) and a single core preparation and maintenance product.
-In comparison, T8, despite more limited in number (n=13), shows only one core, with the rest of the artefacts being blanks (as previously noted) or retouched tools (23.1% and n=3).
-Aside from the TL artefacts, all non-local cores show at least 1 core and frequently the presence of shatter.
+For T6 the absence of cores but the presence of shatter and a single core preparation and maintenance product suggests knapping on site; all other non-local types show at least 1 core and frequently the presence of shatter.
 As expected, the TL artifacts include a blank and a retouched.
 However, a shatter was also identified.
-In total, 65 chert retouched tools were identified, with 23.1% (n=15) having an unidentified chert type.
-A total of 26 retouched tools correspond to local cherts, while 23 correspond to non-local cherts.
 
@@
 
-Regarding the typology, 14 different types of retouched tools were identified.
+In total, 65 chert retouched tools were identified, with 23.1% (n=15) having an unidentified chert type (@tbl-retouch-proto).
+A total of `r 1+10+2+8+5` (`r round((1+10+2+8+5)*100/65, 2)`%) retouched tools correspond to local cherts, while `r 10+9+3+1+1` (`r round((10+9+3+1+1)*100/65, 2)`%) correspond to non-local and TL cherts.
+As seen in @tbl-retouch-proto, 14 different types of retouched tools were identified.
 Endscrappers (26.2%), splintered pieces (17%), burins (12.3%) and retouched flakes (7.7%) were the most prevalent.
-Vale Comprido points, the Proto-Solutrean fossil director, are barely present in chalcedony and chert (n=1), as they are mostly present in dolerite, while chert was mainly used for other tools as seen in previous studies (Belmiro et al., 2021).
-Regarding the types of chalcedony and chert, local raw materials show 9 different typologies, with Vale Comprido points, Solutrean retouch and notches being present only in the local types, even if in small numbers.
+Local raw materials show 9 different typologies, with Vale Comprido points, Solutrean retouch and notches being present only in the local types, even if in small numbers.
 Non-local cherts show 10 different typologies of retouched tools.
@@
 The final sample from levels A, B and C from the Solutrean occupations of the Shelter area consists of 914 artifacts.
-When crossing the class data with type of raw material, we see that local chert, especially T2 (n=506) which accounts for the majority of local chert present in the studied samples, the majority of the debitage is composed of flakes and elongated blanks (83.9%).
+Of these, 201 (`r round(201*100/914, 2)`%) correspond to the indeterminate chert group.
+Local cherts correspond to `r 16+506+4+10+30` (`r round((16+506+4+10+30)*100/914, 2)`%) artefacts, mostly from T2 (n=506).
+Non-local cherts (including the TL group) correspond to `r 86+9+30+3+1+17+1` (`r round((86+9+30+3+1+17+1)*100/914, 2)`%) artefacts, mostly from T6 (n=86).
+
+When crossing the class data with type of raw material (@tbl-general-solutrean), we see that for local chert, especially T2, the majority of the debitage is composed of flakes and elongated blanks (83.9%).
 Cores are also present (10.5%), alongside a small percentage of core preparation and maintenance products (1.3%).
 In comparison to debitage products, retouched tools are scarce in the sample (\~3%).
-The non-local cherts most present in the Solutrean of the Shelter area are T6 (n=86) and T8 (n=30).
-These show similar patterns to the local cherts, with high percentages of blanks (\>50%).
+Non-local cherts, especially T6 and T8, show similar patterns to the local cherts, with high percentages of blanks (\>50%).
 Cores and core preparation and maintenance are also present in small percentages (\~3-6% and 3.5% only for T6, respectively), as well as retouched tools (\<10%).
 However, these specific results are truncated by the lack of shatter, which were not present in the sample used for the study (see Materials and Methods section).
-It is important to notice, however, that alike previous results from the Solutrean assemblages of the Terrace area, a large portion of the studied sample was altered and did not allow for a chert type attribution (21.9912473%).
 
@@
 
-The analysed Solutrean sample has a total of 43 retouched tools (complete and fragments), with a high variety of retouched tools (11), ranging from 5 (11.6%) to 4 (9.3%) per tool type.
-These results however, are significantly reduced from the original dataset (Cascalheira et al., 2013), possibly due to no bucket coordinates being used.
-The typologies which are most frequently present in the studied sample are bifacial blanks and bifacial fragments, amounting to a total of 23.2% of products related to bifacial tool/point production.
-Two types of points were identified: shouldered points (2 complete and 1 fragment, amounting to a total of 6.9%) and tanged and 4 winged points (9.3%).
-Other frequent retouched typologies are end scrappers (9.3%), notches (11.6%), retouched blanks (\~14%) and splintered pieces (11.6%).
-Regarding chert type, a substantial portion of retouched tools are of unidentified chert types (n=13), mostly due to alterations (fire and post depositional).
-Most retouched tools with identified chert are local (n=13, accounting to 30.2325581%), followed by Type 6 cherts, previously suggested to be from Central Portugal (n=7, accounting to 16.2790698%).
+The analysed Solutrean sample has a total of 43 retouched tools (complete and fragments), with `r 3+13+1` corresponding to local cherts and `r 7+1+2+3` to non-local cherts, with the remaining being cherts of unidentified type.
+These results are significantly reduced from the original dataset (Cascalheira et al., 2013), possibly due to no bucket coordinates being used.
+From the 43 retouched tools, 11 different typologies were identified (without accounting for fragments). 
+The typologies which are most frequently present in the studied sample are bifacial blanks and bifacial fragments (n=10).
+Two types of points were identified: shouldered points (n=3, including fragments) and tanged and winged points (n=4).
+Other frequent retouched typologies are retouched blanks (n=6), notches (n=5), splintered pieces (n=5) and endscrappers (n=4).
 Local types show a higher variability retouched typologies (9 types), although this may be related to the higher number of tools.

### 2025-01-22 analysis/Supplementary/supplementary_tables.qmd
Results improved, structured and updated + render. Added TL table to SOM qmd.

@@
 ```
+
+# TL cherts
+
+```{r}
+#| label: tbl-TL-classes
+#| tbl-cap: Trace lithologies cherts from the Gravettian and Proto-Solutrean assemblages of the Terrace and Solutrean assemblage from the Shelter and corresponding technological class.
+
+TL_classes_terrace <- phase_dataset |> 
+  filter(RMTYPE == "TL") |> 
+  select(Technocomplex, SUBTYPE, ID, CLASS, RETOUCHEDPIECETYPOLOGY)
+
+TL_classes_shelter <- solutrean_dataset |> 
+  mutate(Technocomplex = case_when(
+    LEVEL == "A" | LEVEL == "B" | LEVEL == "C" ~ "Solutrean")) |> 
+  filter(RMTYPE == "TL") |> 
+  select(Technocomplex, SUBTYPE, ID, CLASS, RETOUCHEDPIECETYPOLOGY)
+
+TL_classes <- rbind(TL_classes_terrace, TL_classes_shelter)
+TL_classes <- TL_classes |> 
+  rename(TL = SUBTYPE) |> 
+  arrange(TL)
+
+kable(head(TL_classes))
+
+```

### 2025-01-22 analysis/paper/paper.qmd
Results improved, structured and updated + render. Added TL table to SOM qmd.

@@
 Non-local cherts show different numbers of artefacts depending on the type, with T6 and T7 showing the highest values (169 or `r round(169*100/766, 2)`% of the total assemblage, and 86 or `r round(86*100/766, 2)`%, respectively); the other non-local cherts have artefact counts lower than 20 per type.
-The TL cherts also show smaller numbers, with 12 artefacts.
+The TL cherts also show smaller numbers, with 12 artefacts, corresponding to 7 different types of TL cherts (table s++ in SOM ++).
 
@@
 
-Regarding the sizes of the lithic artifacts (width, length and weight), blanks (flakes and elongated products) and retouched tools show little differences in the measurements of artefacts of local and non-local chert types.
-The only exception are T6 and T7 flakes, which show slightly lower weight means (\~2.6-3.0 gr) and lower maximum length (\~20-22 mm) than the other types.
-For cores, T2 shows bigger (heavier) values than T6 (means of 21.3 and 12.6, and maximum weight of 37.2 gr and 22.5 gr, respectively), even if they seem to vary in morphology, with little patterns regarding thickness, width and length.
-
+The sizes of lithic artefacts (flakes, elongated blanks, cores and retouched tools) are explored through scatterplots of width and length (@fig-gravettian-boxplots a) and boxplots for weight (@fig-gravettian-boxplots b); complete tables with descriptive statistics can also be found in SOM ++, table s++.
+The scatterplots show different patterns in the several analysed classes.
+Flakes are longer and wider in local cherts and generally smaller in non-local types (\~20-22 mm maximum lengths for T6 and T7).
+This is similar for elongated blanks, with non-local cherts presenting thinner and short bladelets than local types.
+Core width is variable between both local and non-local cherts; however, non-local cherts show a concentration of cores in the ~20 mm length mark, while local cherts show a wider variation of longer cores.
+Retouched tools show no differences in width and length between local and non-local cherts.
 Regarding the weight (@fig-gravettian-boxplots b), although there is some variability, non-local types T6 and T7 seem to have the lowest weights in all classes (flakes, elongated blanks, cores and retouched tools).
@@
 Of these, 167 (`r round(167*100/635, 2)`%) have an indeterminate chert type.
-Local cherts correspond to `r 32+62+32+158+51` artefacts (`r round((32+62+32+158+51)*100/635, 2`%), with T4 being the most present local chert (n=158 or `r round(158*100/(32+62+32+158+51), 2)`% of the local types).
-Non-local types are present in smaller numbers (n=`r 60+47+13+9+1` or `r round((60+47+13+9+1)*100/635, 2)`%), with T6 and T7 being the most relevant (n=60 and n=47, respectively), and the TL group includes 3 artefacts.
+Local cherts correspond to `r 32+62+32+158+51` artefacts (`r round((32+62+32+158+51)*100/635, 2)`%), with T4 being the most present local chert (n=158 or `r round(158*100/(32+62+32+158+51), 2)`% of the local types).
+Non-local types are present in smaller numbers (n=`r 60+47+13+9+1` or `r round((60+47+13+9+1)*100/635, 2)`%), with T6 and T7 being the most relevant (n=60 and n=47, respectively), and the TL group includes 3 artefacts, corresponding to 2 types of TL cherts (Table s++ in SOM ++).
 
@@
 
-Regarding the debitage, cores and tool sizes, flakes show slightly higher means in local cherts (\~21-22 mm for length and 3.5-4.5 gr for weight), in comparison to non-local cherts, especially T6-T8 (\~19-20 mm for length and 1.3-2.6 gr in weight), with exception of chalcedony which shows consistently low sizes (21.8 mm for length and 2.8 gr for weight).
-Elongated blank however, show no differences between local and non-local chert types.
-In retouched tools, weight also shows smaller means for non-local cherts (\~3-2 gr), although with higher standard deviation (\~1.6).
-Regarding cores, the sample size for non-local cherts is small and only T7 statistics could be obtained.
-Despite this, when compared to local cherts (T2-T5), T7 cores show smaller mean and maximum values (7.3 gr and 8 gr, respectively).
+As seen in @fig-ratiotd (see also table s++ in the SOM ++), the lowest ratio of retouched to debitage is seen in T1 and T4, with values below 0.1, while other local types show slightly higher values (0.1-0.2).
+Following the same trend as the Gravettian, the non-local cherts show slightly higher ratios (0.3-0.5) of retouched to debitage.
+Regarding the blank to core ratio (@fig-ratiobc and table s++ in the SOM ++), T2 and T4 show the lowest ratios (8.7 and 6.2 respectively) of blanks per cores, while the non-local T7 shows the highest ratios (11.0) of blanks to cores.
+Unlike the Gravettian, T1 (local chalcedony) and T6 (non-local chert) have no cores (@tbl-general-proto).
+
+Regarding the size (width, length and weight) of lithic artefacts (flakes, elongated blanks, cores and retouched tools), several patterns can be seen between local and non-local chert types (@fig-proto-boxplots, table s++ and s++ in the SOM ++).
+The distribution of width and length (@fig-proto-boxplots a) for flakes shows a concentration of non-local flakes below the 25 mm width and length (with means around 20 mm), while local cherts are more variable and show larger widths and lengths.
+Similarly, the larger elongated blanks are from local cherts, although there is variability in the sizes.
+Although there are clear differences in the number of cores between local and non-local cherts (@tbl-general-proto) making patterns harder to identify, non-local cherts seem to show smaller cores.
+Finally, regarding retouched tools, non-local cherts also seem to be smaller, with widths concentrated below the 20 mm mark, while local cherts are concentrated above the 20 mm mark; no differences were identified regarding length.
+As for weight (@fig-proto-boxplots b and table s++ in SOM ++), flakes seem to be heavier than and non-local cherts (means of 3.5-4.5 gr and 1.3-2.6 gr, respectively); especially T4 shows heavier flakes and outliers, than T6 and T7.
+In comparison, elongated blanks show no differences between local and non-local cherts.
+Although non-local chert cores are barely represented in the assemblage (@tbl-general-proto), they show the lighter weights than local cherts.
+Retouched tools show the clearest differences, with non-local cherts presenting lighter weights than local cherts.
 
@@
 ggpubr::ggarrange(proto_scatter, proto_stats_weight, labels = c("a", "b"), ncol = 2, common.legend = TRUE, legend = "bottom")
-
-```
-
-Similar to the Gravettian, non-local cherts seem to show the highest values in the calculated blank-to-core and tool-to-debitage ratios.
-T2 and T4 show the lowest ratios (8.7 and 6.2 respectively) of blanks per cores, while the non-local T7 shows the highest ratios (11.0) of blanks to cores.
-Unlike the Gravettian, T1 (local chalcedony) and T6 (non-local chert) have no cores.
-The lowest ratio of retouched to debitage is seen in T1 and T4, with values below 0.1, while other local types show slightly higher values (0.1-0.2).
-Following the same trend as the Gravettian, the non-local cherts show slightly higher ratios (0.3-0.5) of retouched to debitage.
-
-```{r}
-#| label: tbl-TL-classes
-#| tbl-cap: Trace lithologies cherts from the Gravettian and Proto-Solutrean assemblages of the Terrace and Solutrean assemblage from the Shelter and corresponding technological class.
-
-TL_classes_terrace <- phase_dataset |> 
-  filter(RMTYPE == "TL") |> 
-  select(Technocomplex, SUBTYPE, ID, CLASS, RETOUCHEDPIECETYPOLOGY)
-
-TL_classes_shelter <- solutrean_dataset |> 
-  mutate(Technocomplex = case_when(
-    LEVEL == "A" | LEVEL == "B" | LEVEL == "C" ~ "Solutrean")) |> 
-  filter(RMTYPE == "TL") |> 
-  select(Technocomplex, SUBTYPE, ID, CLASS, RETOUCHEDPIECETYPOLOGY)
-
-TL_classes <- rbind(TL_classes_terrace, TL_classes_shelter)
-TL_classes <- TL_classes |> 
-  rename(TL = SUBTYPE) |> 
-  arrange(TL)
-
-kable(head(TL_classes))
 
@@
 
-For the Solutrean assemblage, there seems to be no clear difference in the measurements and weights of artefacts between local and non-local cherts.
-This is true especially for the most represented classes (flakes and elongated blanks).
-Regarding cores, many of the local and non-local cherts are not available for the statistics due to missing data.
-However, the smallest weight mean seem to be for cores of T6 (16.5 gr), although with high standard deviation and maximum value (16.5 and 48.6 gr, respectively).
+Regarding the tools-to-debitage and blank-to-core ratios, the Solutrean assemblage shows similar patterns as the previous assemblages (@fig-ratiotd and @fig-ratiobc, respectively).
+Local cherts (T2-T5) show low retouch to debitage ratios (\<0.1), with the exception of T1 (chalcedony), which shows a ratio of 1.8.
+Other non-local cherts (T6-T7 and T11) show slightly higher retouch to debitage ratios (0.1-0.3).
+Similarly, local cherts show lower blank to core ratios (\<7), while the non-local cherts show higher ratios (\>9).
+Similar to the Proto-Solutrean, chalcedony cores were also not identified in the studied sample (@tbl-general-solutrean).
+
+Unlike the other assemblages, there seems to be no clear difference in the measurements and weights of artefacts between local and non-local cherts (@fig-solutrean-boxplots and table s++ in SOM ++).
+The exception are elongated blanks, with show differences between T2 and T6, with T6 showing smaller width and lengths (<35 mm and 15 mm, respectively), than the local type.
+Similarly, elongated blank weights are smaller in T6 (ranging between 0-5-2 gr).
+Core weights are also slightly lower in T6 (means of 16.5 gr) than in T2, although the sample is comparatively small (@tbl-general-solutrean).
 
@@
 ```
-
-Regarding the blank-to-core and tools-to-debitage ratios, the Solutrean assemblage shows similar patterns as the previous assemblages, with the local cherts showing lower blank to core ratios (\<7), while the non-local cherts show higher ratios (\>9).
-Similar to the Proto-Solutrean, chalcedony cores were also not identified in the studied sample.
-Similarly, local cherts (T2-T5) show low retouch to debitage ratios (\<0.1), with the exception of T1 (chalcedony), which shows a ratio of 1.8.
-Other non-local cherts (T6-T7 and T11) show slightly higher retouch to debitage ratios (0.1-0.3).
 

### 2025-01-23 analysis/paper/paper.qmd
Discussion (on-going) added. Extra tables added as csv or written with tibble. Latest version rendered.

@@
 
+```{r}
+#| label: library
+
+# Read libraries necessary for running the script
+library(readr)
+library(dplyr)
+library(tidyr)
+library(stringr)
+library(ggplot2)
+library(janitor)
+library(ggpubr)
+library(knitr)
+library(tibble)
+
+```
+
 # Introduction
@@
 
+```{r}
+#| label: tbl-samples
+#| tbl-cap: "Sample absolute number comparison between original datasets (chert and chalcedony only) for the Solutrean occupation of the Shelter area (levels A-C) published in Cascalheira (2010), the Proto-Solutrean occupation of the Terrace (levels 4E and 5) published in Belmiro et.
+al., (2021) and the Solutrean occupation of the Terrace (levels 6-7) and the final datasets used in the study after filtering individually plotted IDs (excluding bucket points) and merging with the geoarchaeological dataset published in Belmiro et al., (2025)."
+
+table_samples <- tibble::tribble(
+  ~Assemblage, ~`Original dataset (n)`, ~`Merged dataset (n)`,
+  "Shelter level A-C", 3359, 910,
+  "Terrace level 4E-5", 1821, 635,
+  "Terrace level 6-7", 781, 766)
+
+knitr::kable(head(table_samples))
+
+```
+
 After the homogenization and merging of the datasets, descriptive statistical analysis was applied to describe and interpret the relation between the raw material and technological data.
@@
 ```{r}
-#| label: library
-
-# Read libraries necessary for running the script
-library(readr)
-library(dplyr)
-library(tidyr)
-library(stringr)
-library(ggplot2)
-library(janitor)
-library(ggpubr)
-library(knitr)
-
-```
-
-```{r}
 #| label: get-data
@@
 ## Technological data
-gravettian_tech <- read.csv(here::here('analysis/data/raw_data/VBT_6_7_JB.csv'))
-proto_tech <- read.csv(here::here('analysis/data/raw_data/VBT_4E_5_JB.csv'))
-solutrean_JC <- read.csv(here::here('analysis/data/raw_data/VBR_ABC_JC.csv'))
-solutrean_points <- read.csv(here::here('analysis/data/raw_data/shelter_points_JC.csv'))
+gravettian_tech <- read_csv(here::here('analysis/data/raw_data/VBT_6_7_JB.csv'))
+proto_tech <- read_csv(here::here('analysis/data/raw_data/VBT_4E_5_JB.csv'))
+solutrean_JC <- read_csv(here::here('analysis/data/raw_data/VBR_ABC_JC.csv'))
+solutrean_points <- read_csv(here::here('analysis/data/raw_data/shelter_points_JC.csv'))
 
@@
   filter(Layer %in% c("A", "B", "C")) |> # Filtering Solutrean layers
-  filter(Raw.material %in% c("Chert", "Chalcedony")) |> # Filtering raw materials
-  filter(Screen != "Crivo") |>  # Filtering out all bucket lithics since they were not included in the RM analysis
+  filter(`Raw material` %in% c("Chert", "Chalcedony")) |>  # Filtering raw materials
+  filter(!Screen %in% c("Crivo", "crivo")) |>  # Filtering out all bucket lithics since they were not included in the RM analysis
   mutate_at("Date", str_replace, "20", "") |> # Removing initial part of the year to match IDs
@@
   filter(Layer %in% c("A", "B", "C")) |> # Filtering Solutrean layers
-  filter(Raw.material %in% c("Chert", "Chalcedony"))
+  filter(`Raw material` %in% c("Chert", "Chalcedony"))
 
@@
 Finally, regarding retouched tools, non-local cherts also seem to be smaller, with widths concentrated below the 20 mm mark, while local cherts are concentrated above the 20 mm mark; no differences were identified regarding length.
-As for weight (@fig-proto-boxplots b and table s++ in SOM ++), flakes seem to be heavier than and non-local cherts (means of 3.5-4.5 gr and 1.3-2.6 gr, respectively); especially T4 shows heavier flakes and outliers, than T6 and T7.
+As for weight (@fig-proto-boxplots b and table s++ in SOM ++), local flakes seem to be heavier than  non-local cherts (means of 3.5-4.5 gr and 1.3-2.6 gr, respectively); especially T4 shows heavier flakes and outliers, than T6 and T7.
 In comparison, elongated blanks show no differences between local and non-local cherts.
@@
 
-Hypotheses to test: a) use lives derived from knapping strategies; b) reduction intensity and tools-to-debitage ratios; c) core size and blank-to-core ratios; d) reduction intensity and tool reworking; e) tool typology diversity and specialization.
+As mentioned before, this study aimed to test two expectations related to the use and knapping of local and non-local raw materials, in specific, chert and chalcedony, and their relation to settlement and mobility strategies, across several UP occupations.
+As such, our method derived the following hypotheses to test: a) use lives derived from knapping strategies; b) reduction intensity and tools-to-debitage ratios; c) core size and blank-to-core ratios; d) reduction intensity and tool reworking; e) tool typology diversity and specialization (@tbl-hypotheses).
+
+```{r}
+#| label: tbl-hypotheses
+#| tbl-cap: A table.
+
+hypotheses <- readr::read_csv(here::here('analysis/data/tables/hypotheses_short.csv'))
+
+knitr::kable(head(hypotheses))
+
+```
 
 To test these hypotheses, the following parameters were evaluated for each assemblage and compared between chert types: P1) frequency of retouched tools; P2) frequency of maintenance products; P3) debitage sizes; P4) tools-to-debitage ratios ; P5) core sizes; P6) blank-to-core ratios; P7) retouched tool sizes; P8) tool diversity; P9) tool specialization.
-The results may be summarised as follows:
-
-Retouched tools are present in higher percentages in the Gravettian assemblages.
-Within the assemblage, retouched tools (P1) are present in higher frequencies in the non-local cherts, as well as in core preparation and maintenance products (P2).
-Debitage sizes (P3) show no differences between local and non-local cherts.
+These parameters and their main results can be found in table @tbl-results-sum, and can be summarised as follows.
+
+In the Gravettian assemblage, retouched tools (P1) are present in higher frequencies in the non-local cherts, as well as in core preparation and maintenance products (P2).
+Debitage sizes (P3) show smaller sizes and weights for non-local cherts.
 Tools-to-debitage ratios (P4) are significantly higher in non-local cherts.
 Core sizes (P5) are slightly smaller in T6, a non-local chert, and the blank-to-core ratios (P6) are higher in non-local cherts.
-Alike blanks, retouched sizes (P7) show no relevant differences between local and non-local cherts; however, tool diversity (P8) is higher in non-local cherts, although with high frequency of less specialised tools such as notches and retouched tools, alongside hunting implements such as backed bladelets (P9).
+Retouched sizes (P7) show no relevant differences between local and non-local cherts; however, tool diversity (P8) is higher in non-local cherts, although with high frequency of less specialised tools such as notches and retouched tools, alongside hunting implements such as backed bladelets (P9).
 The high frequency of burins seen in this assemblage may be related to their use as cores, representing a knapping strategy more than the presence of retouch.
+
+```{r}
+#| label: tbl-results-sum
+#| tbl-cap: "Summarised results for analysis parameters. Gr: Gravettian; Pr: Proto-Solutrean; So: Solutrean; Exp: Expectations; L: Local; NL: Non-local."
+
+p_results <- readr::read_csv(here::here('analysis/data/tables/results_sum.csv'))
+
+knitr::kable(head(p_results))
+
+```
 
@@
 Core preparation and maintenance products (P2) are only present in non-local cherts, albeit in very small numbers (n=2).
-Debitage sizes (P3), core sizes (P5), and retouched tool sizes (P7) are slightly smaller in non-local cherts.
+Blank sizes (P3), core sizes (P5), and retouched tool sizes (P7) are slightly smaller in non-local cherts.
 Regarding the ratios, tool-to-debitage ratios (P4) are higher in non-local cherts, albeit smaller than those seen in the Gravettian assemblage, and blank-to-core ratios (P6) are also higher in non-local cherts.
@@
 
-Finally, the Solutrean shows no relevant differences in the frequencies of retouched tools (P1) and core preparation and maintenance products (P2) between local and non-local cherts.
+Finally, the Solutrean shows no relevant differences in the frequencies of retouched tools (P1) and core preparation and maintenance products (P2) between local and non-local cherts, with only slightly higher frequencies in the non-local cherts.
 In general, the retouch frequency is lower than that observed in the previous assemblages.
-Debitage (P3), core (P5), and retouched tool (P7) sizes show no differences between local and non-local cherts.
+Debitage (P3), core (P5), and retouched tool (P7) sizes show no differences between local and non-local cherts, except for elongated blanks which are smaller in T6 and bigger in T2, and for slightly smaller cores seen in T6, although with high standard deviations.
 Similarly to the Proto-Solutrean, tools-to-debitage ratios (P4) and blank-to-core ratios (P6) are slightly higher in non-local cherts.
@@
 
-How the results fit or do not fit within the hypothesis, and possible reasons why.
-
-Focus on the high presence of retouch and burins during the Gravettian.
-
-Fitting the technological data with the suggested interpretations from article 2.
+In general, the obtained results seem to fit the study expectations.
+The Gravettian assemblage seems to show agreement with most of the expectations (table **).
+Although ~50% of the assemblage is composed of non-local raw materials, making them as frequent as the local cherts, non-local cherts, especially T6 and T7 show differential techno-typological patterns from the local raw materials.
+Only in expectation e) there seems to be a difference from the expected outcome.
+
+In this sense, it would be expected a lower typological diversity from non-local cherts, since the distance to source and inferred mobility would assume that less specialized tools would be more adequate for several use-case scenarios; instead, our results show higher tool diversity in non-local cherts.
+However, this scenario also implies that tools are being transported as either less specialized tools or even blanks which are then maintained.
+Instead, our results also suggest that the long-distance raw materials, especially T6 and T7, are being knapped at the site.
+As such, the high diversity of tool typology may simply be a reflection of the transportation of volumes and their knapping on site, a possible preference for these raw materials for the production of retouched tools, and thus their high typological variability.
+Despite this high variability however, less specialized tools such as retouched blanks are still predominant in non-local cherts, as well as specialized tools such as backed bladelets, giving weight to the interpretation that given the possible ease of access to these raw materials, either through high-mobility or exchanged between groups, these were knapped on site, but with longer use-lives, extensive reduction and high presence of retouch and formal tools.
+
+Similarly, the Proto-Solutrean assemblage shows only disagreement with the expected outcomes within the variability of retouched tools (expectation e and P8-9) and possibly the focus on maintenance and prolonging use-lives of tools (expectation a), especially concerning P2.
+Unlike expected, there are little differences in the amount and type of retouched tool types between local and non-local raw materials.
+Unlike the Gravettian occupation, however, retouched tools are overall not well-represented in the assemblage.
+It is key to notice, however, that the Proto-Solutrean occupation of Vale Boi seemed to introduce other raw materials, such as quartz or dolerite, to the production of retouched tools.
+Although knapping strategies seemed to be overall simpler (e.g., simple platform cores in quartz), splintered pieces are highly represented in quartz.
+In opposition, most endscrapers are made of chert (Belmiro et al., 2021), while the majority of Vale Comprido points are found in dolerite (Belmiro et al., 2021; Marreiros, 2009).
+These patterns, then, may correspond to a specialisation of tool production to the type of raw material itself (whether quartz, chert or dolerite), thus making chert, retouched tool variability patterns similar and unrelated to distance of source.
+Regarding P2, although there is a slightly higher presence of core preparation and maintenance products in non-local cherts, there is an overall lack of this class of artefacts.
+This means that for the Proto-Solutrean assemblage of Vale Boi, core maintenance variables are not adequate to discern differences between local and non-local raw materials, since these knapping strategies were not consistently used during this chronology.
+
+Finally, the Solutrean assemblages also show only two parameters that disagree with the expectations: debitage sizes (P3) and retouched tool sizes (P7).
+Flakes show no differences in dimensions and weight between local and non-local raw materials.
+This may be related to the production process of Solutrean points, which may start with larger flakes or elongated products that are then reconfigured to the desired product, thus producing similar discarded flakes in all chert types.
+However, elongated blanks do follow the expected patterns, with non-local cherts having smaller products and local cherts.
+This may mean that non-local cherts, especially T6, are being further reduced in specific strategies for the production of elongated products, thus leading to different patterns between these two classes of blanks.
+The sizes of retouched tools, alike flakes, also show no relevant differences between local and non-local.
+However, this variable may not be adequate for the present study, especially regarding the size measurements; both width and length are used for scatterplots, however, the technological database used in this study is missing specific measurements (especially length) for typologies of retouched tools.
+As such, the scatterplots are further truncated by the difference in analysis methodology, making this pattern possibly invalid.
+
+Aside from the specific expectations for the treatment of local and non-local cherts within each assemblage, when comparing between technocomplexes, a few key patterns also stand out.
+Chert retouched tools are present in higher percentages in the Gravettian assemblages (~22%) than in the Proto-Solutrean and Solutrean assemblages (<10%).
+This is also clear through the tool-to-debitage ratios, which are higher in the Gravettian assemblage (@fig-ratiotd).
+
+There are little differences in the general size and weight of the artifacts between assemblages.
+For example, for key classes where size and weight differences could be expected (e.g., cores), the length measurements are mostly all located between 20-40 mm and weight between 10-40 gr.
+The major differences seem to be explained better by bigger cores in local chert types and smaller cores for non-local chert types.
+However, a few key differences were noted.
+T4 cores (the chert type most frequent in the Proto-Solutrean assemblage) shows the largest weights of cores of all assemblages, both in the Solutrean and Proto-Solutrean assemblage, as well as T5 in the Solutrean assemblage.
+T6 and T7 have consistently the smaller weights, especially compared to T2 during the Proto-Solutrean and Gravettian, which shows slightly higher weights.
+This consistency in artefact size throughout the assemblages may be explained by a combination of local lithic resource availability and transport and knapping strategies focusing on intensive reduction and maintenance of non-local raw materials.
+Belmiro et al.
+(2023) highlight the small size of chert nodules (usually between ~5-10 cm with a maximum of 20 cm of diameter) in the Algarve region.
+Although abundant and possibly available throughout all of the UP, given their location in accessible cliffs, these lithic resources are often small and fracture easily when obtained directly from the source.
+Given the possibly initial small size of the local chert nodules, it may be possible that the small artefact sizes are related not to intensive knapping of the nodules, but to the initial size of the knapped volumes.
+This is also supported by the lower blank-to-core ratios seen in local raw materials.
+
+In comparison, at least for T6 cherts from central Portugal (for which we have a comparable reference collection of unknapped volumes collected from secondary sources), the initial volumes are frequently larger (>20 cm), although smaller nodules can also be found.
+This suggests that the small artefact sizes of this non-local resource are not necessarily a product of resource availability but an intentional choice of either transportability of smaller volumes/nodules, knapping strategies focused on intensive reduction and maintenance, or both.
+This is also supported by the higher blank-to-core and tool-to-debitage ratios seen in non-local raw materials, especially T6.
+
+A key interpretation can be derived.
+Cores, shatter and debitage are present even in non-local chert types (excluding the TL category), meaning although possibly coming from long-distance (e.g., T6 and T7 with sources over ~200 km, as the crow flies), these raw materials were being introduced to the site as knappable volumes and not only as finished products.
+This has also been previously suggested, based on the presence of cortex attributed to small transport distances from source to procurement area, in non-local cherts in all assemblages (Belmiro et al., 2025).
+
+## Some extra notes and structure
+
+- Goals of the study
+- Parameters studied
+- Summary of the results based on expectations
+- How they fit within the expectations + how they differ by assemblage
+- General differences between expectations
+- Interpretation of chert procurement and management, and changes through time + settlement changes
+
+When groups move to areas where raw materials are scarcer or of poorer quality (understanding that poorer quality may not mean they are bad quality, simply not as good quality as the raw materials group have current access), more effort is placed in the preparation of transportable cores and tools (Kelly, 1988).
+This may be the case for the Gravettian assemblage; not only is the percentage of tools high, large enough volumes of stone were transported to the site and knapped, but also high frequency of burins, which may have also been a technological choice to obtain elongated products in an efficient way.
+There seems to be an effort in transporting raw materials of known good quality in transportable yet adaptable shapes to later shape and use.
+
+It would be also expected that whenever groups have easy access to good quality, non-local raw materials, either due to high mobility or exchanges between groups, these resources become more abundant in the assemblage, but also be less extensively used (Andrefsky, 1994).
+The idea is that the patterns of non-local raw material use are based on their conservation due to cost of procurement; when that cost is reduced, so is the necessity to conserve the resources, and thus the expected patterns may not be as clear.
+Despite the high frequency of non-local raw materials, which implies some sort of ease of access to non-local raw materials of good quality, probably due to high residential mobility, the assemblage still shows conservation of raw materials.
+This is seen in the high frequence of retouched tools and high ratios of tool to debitage when compared to local raw materials.
+Artefact sizes are also smaller in non-local raw materials then local raw materials and blank-to-core ratios are higher, showing that these non-local raw materials were extensively used.
+This shows that access to non-local raw materials, making them highly represented in the assemblage, especially the cherts from central Portugal (T6), did not impact the patterns expected for the treatment of non-local raw materials.
+
+A strategy used to raw material transportation is stockpilling lithic raw materials in various production stages, thus reducing the need for extensive lithic production and tool reworking (Kuhn, 2004).
+Longer-term residential occupations occupations are more likely to involve stockpilling strategies and reduced investment in tool maintenance (Bamforth, 1990).
+This would create assemblages with more blanks, less maintenance products and less retouched tools such as retouched flakes.
+However, even with stockpilling,
 

### 2025-02-18 analysis/Supplementary/supplementary_tables.qmd
Correction to TL supplementary tables
Macbook morning session upload: Complete assemblage description and tables moved to supplementary. Main paper tables simplified with categories collapsed for an easier read. tbl- chunks added tbl-subcap to present table groups (assemblage and retouched) in a single plot.

@@
   drop_na(CLASS) # Dropping NAs on the technological class drops dataset to 914
+
+```
+
+# Assemblage description
+
+```{r}
+#| label: general-tables
+
+# Tables with class by raw material type (n and %)
+## For Gravettian
+gravettian_general <- phase_dataset |>
+  filter(Level %in% c(6, 7)) |>
+  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
+                                        "TYPE 5", "TYPE 6","TYPE 7", 
+                                        "TYPE 8", "TYPE 9", "TYPE 10", "TL", "INDET"))) |>
+  tabyl(CLASS, RMTYPE) |> 
+  adorn_totals(c("row", "col")) |> 
+  adorn_percentages("col") |> 
+  adorn_pct_formatting(2) |> 
+  adorn_ns("front") |> 
+  mutate(across(where(is.character), ~ gsub("^0   \\(0\\.00%\\)$", " ", .)))
+
+## For Proto-Solutrean (4E and top of 5)
+proto_general <- phase_dataset |> 
+  filter(Level %in% c("4E", "4E/5")) |> 
+  filter(CLASS != "Chip") |> 
+  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
+                                        "TYPE 5", "TYPE 6","TYPE 7", 
+                                        "TYPE 8", "TYPE 9", "TYPE 10", "TL", "INDET"))) |> 
+  tabyl(CLASS, RMTYPE) |> 
+  adorn_totals(c("row", "col")) |> 
+  adorn_percentages("col") |> 
+  adorn_pct_formatting(2) |> 
+  adorn_ns("front") |> 
+  mutate(across(where(is.character), ~ gsub("^0   \\(0\\.00%\\)$", " ", .)))
+
+## For Solutrean
+solutrean_general <- solutrean_dataset |> 
+  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
+                                        "TYPE 5", "TYPE 6","TYPE 7", 
+                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL", "INDET"))) |> 
+  tabyl(CLASS, RMTYPE) |> 
+  adorn_totals(c("row", "col")) |> 
+  adorn_percentages("col") |> 
+  adorn_pct_formatting(2) |> 
+  adorn_ns("front") |> 
+  mutate(across(where(is.character), ~ gsub("^0   \\(0\\.00%\\)$", " ", .)))
+
+```
+
+```{r}
+#| label: tool-diversity
+
+# Solutrean table
+## Table preparation
+solutrean_retouch <- solutrean_dataset |>
+    mutate(CLASS = case_when(
+    CLASS == "Retouched piece" ~ "RetouchedPiece",
+    CLASS == "Retouched piece fragment" ~ "RetouchedPieceFrag",
+    TRUE ~ CLASS)) |>
+  filter(CLASS %in% c("RetouchedPiece", "RetouchedPieceFrag")) |>
+  select(RMTYPE, RETOUCHEDPIECETYPOLOGY) |>
+  filter(RETOUCHEDPIECETYPOLOGY != "") |>
+  dplyr::mutate(RMTYPE = case_when(
+    RMTYPE == "TYPE 1" ~ "T1",
+    RMTYPE == "TYPE 2" ~ "T2",
+    RMTYPE == "TYPE 3" ~ "T3",
+    RMTYPE == "TYPE 4" ~ "T4",
+    RMTYPE == "TYPE 5" ~ "T5",
+    RMTYPE == "TYPE 6" ~ "T6",
+    RMTYPE == "TYPE 7" ~ "T7",
+    RMTYPE == "TYPE 8" ~ "T8",
+    RMTYPE == "TYPE 9" ~ "T9",
+    RMTYPE == "TYPE 10" ~ "T10",
+    RMTYPE == "TYPE 11" ~ "T11",
+    TRUE ~ RMTYPE))
+## Janitor tabyl for final table with n and %
+solutrean_retouch_tab <- solutrean_retouch |>
+  tabyl(RETOUCHEDPIECETYPOLOGY, RMTYPE) |>
+  adorn_totals(c("row", "col")) |>
+  mutate(across(where(is.numeric), ~ ifelse(. == 0, " ", .))) |>
+  select(RETOUCHEDPIECETYPOLOGY, T1, T2, T5, T6, T7, T8, T11, INDET, Total) |> # Ordering values
+  rename(Typology = RETOUCHEDPIECETYPOLOGY)
+
+# Proto-solutrean table
+## Table preparation
+proto_retouch <- phase_dataset |>
+  filter(CLASS %in% c("RetouchedPiece", "RetouchedPieceFrag")) |>
+  filter(Level %in% c("4E", "4E/5")) |>
+  select(RMTYPE, RETOUCHEDPIECETYPOLOGY) |>
+  dplyr::mutate(RMTYPE = case_when(
+    RMTYPE == "TYPE 1" ~ "T1",
+    RMTYPE == "TYPE 2" ~ "T2",
+    RMTYPE == "TYPE 3" ~ "T3",
+    RMTYPE == "TYPE 4" ~ "T4",
+    RMTYPE == "TYPE 5" ~ "T5",
+    RMTYPE == "TYPE 6" ~ "T6",
+    RMTYPE == "TYPE 7" ~ "T7",
+    RMTYPE == "TYPE 8" ~ "T8",
+    RMTYPE == "TYPE 9" ~ "T9",
+    RMTYPE == "TYPE 10" ~ "T10",
+    RMTYPE == "TYPE 11" ~ "T11",
+    TRUE ~ RMTYPE))
+## Janitor tabyl for final table with n and %
+proto_retouch_tab <- proto_retouch |>
+  tabyl(RETOUCHEDPIECETYPOLOGY, RMTYPE) |>
+  adorn_totals(c("row", "col")) |> 
+  mutate(across(where(is.numeric), ~ ifelse(. == 0, " ", .))) |>
+  select(RETOUCHEDPIECETYPOLOGY, T1, T2, T3, T4, T5, T6, T7, T8, T9, TL, INDET, Total) |> # Ordering values
+  rename(Typology = RETOUCHEDPIECETYPOLOGY)
+
+# Gravettian table
+## Table preparation
+gravettian_retouch <- phase_dataset |> 
+  filter(CLASS %in% c("RetouchedPiece", "RetouchedPieceFrag")) |>
+  filter(Level %in% c(6, 7)) |> 
+  select(RMTYPE, RETOUCHEDPIECETYPOLOGY) |> 
+  dplyr::group_by(RMTYPE) |>
+  dplyr::mutate(RMTYPE = case_when(
+    RMTYPE == "TYPE 1" ~ "T1",
+    RMTYPE == "TYPE 2" ~ "T2",
+    RMTYPE == "TYPE 3" ~ "T3",
+    RMTYPE == "TYPE 4" ~ "T4",
+    RMTYPE == "TYPE 5" ~ "T5",
+    RMTYPE == "TYPE 6" ~ "T6",
+    RMTYPE == "TYPE 7" ~ "T7",
+    RMTYPE == "TYPE 8" ~ "T8",
+    RMTYPE == "TYPE 9" ~ "T9",
+    RMTYPE == "TYPE 10" ~ "T10",
+    TRUE ~ RMTYPE))
+## Janitor tabyl for final table with n and %
+gravettian_retouched_tab <- gravettian_retouch |> 
+  tabyl(RETOUCHEDPIECETYPOLOGY, RMTYPE) |>
+  adorn_totals(c("row", "col")) |>
+  mutate(across(where(is.numeric), ~ ifelse(. == 0, " ", .))) |>
+  select(RETOUCHEDPIECETYPOLOGY, T1, T2, T3, T4, T5, T6, T7, T8, T9, T10, TL, INDET, Total) |> # Ordering values
+  rename(Typology = RETOUCHEDPIECETYPOLOGY)
+
+```
+
+## Gravettian (levels 6-7)
+
+The sample studied for the Gravettian occupation (\>2012 excavations) of the Terrace area included 766 artifacts.
+Of these, 209 (`r round(209*100/766, 2)`%) have an indeterminate chert type (@tbl-general-gravettian).
+Chalcedony (T1) and local cherts (T2-5) show a total of `r 35+110+17+13+72` artefacts (`r round((35+110+17+13+72)*100/766, 2)`%).
+Non-local cherts show different numbers of artefacts depending on the type, with T6 and T7 showing the highest values (169 or `r round(169*100/766, 2)`% of the total assemblage, and 86 or `r round(86*100/766, 2)`%, respectively); the other non-local cherts have artefact counts lower than 20 per type.
+The TL cherts also show smaller numbers, with 12 artefacts, corresponding to 7 different types of TL cherts (table s++ in SOM ++).
+
+@tbl-general-gravettian shows the number and percentages of technological classes per chert type.
+Local cherts show similar patterns between them, with T2 and T5 being the types with higher numbers of artefacts (n=110 and n=72, respectively).
+For all local cherts, blanks (including blank fragments) represent more than 50% of the artefacts.
+Cores and core maintenance products are only present in T2 and T5, although in small frequencies: for T2, cores represent 13.6% (n=15 including core fragments) of the group and core preparation and maintenance products represent 1.8% (n=2).
+For T5 only 4 core fragments and 1 core preparation and maintenance product are present, all frequencies between \~5-1%.
+Retouched tools seem to be especially relevant in T2 (n= 20, \~18% including retouched piece fragments), making them the most frequent class after blanks.
+T5 shows similar percentages of retouched tools and shatter (\~12% and \~14%, respectively).
+T1 (chalcedony) shows similar patterns to other local cherts, although with higher percentages of shatter (34.3%).
+
+When compared to local cherts, non-local chert types (especially T6 and T7) show similar percentages of blanks (n=`r 44+24` or 40% and `r 21+21`\`or \~50%, respectively) but higher percentages of retouched tools (\~30%).
+Cores, core fragments and core preparation and maintenance products are also present in all non-local cherts (excluding TL) even if in percentages below 4% (n\<4), as well as shatter (\~12%).
+The exception is T6 that shows the highest numbers of cores and core fragments (n=8), although with low frequencies (4.6%).
+The TL cherts show mostly blanks and blank fragments (n=3 and n=2, respectively), and retouched tools (n=4).
+
+```{r}
+#| label: tbl-general-gravettian
+#| tbl-cap: Number and percentage of technotypological classes by raw material type from the Gravettian assemblage (levels 6 and 7) of the Terrace area of Vale Boi.
+
+kable(gravettian_general)
+
+```
+
+As seen in table @tbl-retouch-gravettian, 170 retouched tools were identified, with 39 (`r round(39*100/170, 2)`%) having an unidentified chert type.
+From the identified chert types, `r 3+20+4+2+9` (`r round((3+20+4+2+9)*100/170, 2)`%) of retouched tools are from local types, `r 58+24+3+1+3` (`r round((58+24+3+1+3)*100/170, 2)`%) from non-local cherts and 4 are from the TL group.
+From all chert types, T6 shows the highest numbers of retouched pieces (n=58, `r round(58*100/170, 2)`%).
+Regarding typologies, 15 types of retouched tools were identified (excluding retouched tool fragments).
+All these typologies were found in the non-local chert group, while only 7 retouched tool typologies were identified in local cherts/chalcedony.
+The most common typologies identified in the assemblage (and for all chert types) were burins (n=52), endscrappers (n=17), retouched flakes (n=30) and splintered pieces (n=26).
+From all these, the majority were produced in non-local chert, mainly T6 and in lesser numbers T7.
+Although in lesser numbers as well, backed bladelets were also identified (n=6), but only present in non-local chert types (T6 and T7).
+The presence of burin spalls in local chert, especially T2 (n=9) and non-local cherts (T6 with n=12 and T7 with n=4) is expected due to the prevalence of burins (@tbl-general-gravettian).
+
+```{r}
+#| label: tbl-retouch-gravettian
+#| tbl-cap: Number and percentage of retouched tools by raw material of the Gravettian assemblage (levels 6 and 7) of the Terrace area.
+
+kable(gravettian_retouched_tab)
+
+```
+
+## Proto-Solutrean (levels 4E and top of level 5)
+
+The final lithic sample from the Proto-Solutrean assemblage, consisting of the levels 4E and top of level 5, consists of 635 artifacts.
+Of these, 167 (`r round(167*100/635, 2)`%) have an indeterminate chert type.
+Local cherts correspond to `r 32+62+32+158+51` artefacts (`r round((32+62+32+158+51)*100/635, 2)`%), with T4 being the most present local chert (n=158 or `r round(158*100/(32+62+32+158+51), 2)`% of the local types).
+Non-local types are present in smaller numbers (n=`r 60+47+13+9+1` or `r round((60+47+13+9+1)*100/635, 2)`%), with T6 and T7 being the most relevant (n=60 and n=47, respectively), and the TL group includes 3 artefacts, corresponding to 2 types of TL cherts (Table s++ in SOM ++).
+
+When crossing the lithic class with chert type (@tbl-general-proto), we see that the local types show high frequency of blanks (\>50%, including blank fragments), followed by shatter (\~11-27%).
+Cores are present in small numbers (n=`r 4+5+13+6`), corresponding to ~9-12% of the classes in each local type, with the exception of T1 (chalcedony), with no cores.
+Core preparation and maintenance products are absent in all local cherts.
+For non-local types, especially those with higher number of artefacts (T6-T8), blanks and blank fragments are the most frequent class (48.3%, 53.1% and 69.2%, respectively).
+Retouched tools are present in smaller frequencies for the previous mentioned chert types (\~17-23%).
+For T6 the absence of cores but the presence of shatter and a single core preparation and maintenance product suggests knapping on site; all other non-local types show at least 1 core and frequently the presence of shatter.
+As expected, the TL artifacts include a blank and a retouched.
+However, a shatter was also identified.
+
+```{r}
+#| label: tbl-general-proto
+#| tbl-cap: Number and percentage of technotypological classes by raw material type from the Proto-Solutrean assemblage (levels 4E and top of level 5) of the Terrace area of Vale Boi.
+
+kable(proto_general)
+
+```
+
+In total, 65 chert retouched tools were identified, with 23.1% (n=15) having an unidentified chert type (@tbl-retouch-proto).
+A total of `r 1+10+2+8+5` (`r round((1+10+2+8+5)*100/65, 2)`%) retouched tools correspond to local cherts, while `r 10+9+3+1+1` (`r round((10+9+3+1+1)*100/65, 2)`%) correspond to non-local and TL cherts.
+As seen in @tbl-retouch-proto, 14 different types of retouched tools were identified.
+Endscrappers (26.2%), splintered pieces (17%), burins (12.3%) and retouched flakes (7.7%) were the most prevalent.
+Local raw materials show 9 different typologies, with Vale Comprido points, Solutrean retouch and notches being present only in the local types, even if in small numbers.
+Non-local cherts show 10 different typologies of retouched tools.
+Perforator-endscrapper, truncation, backed bladelet and retouched blades typologies can only be found in non-local varieties, albeit also in small numbers.
+
+```{r}
+#| label: tbl-retouch-proto
+#| tbl-cap: Number and percentage of retouched tools by raw material of the Proto-Solutrean assemblage (levels 4E and top of level 5) of the Terrace area.
+
+kable(proto_retouch_tab)
+
+```
+
+## Solutrean (levels A to C)
+
+The Solutrean assemblages show similar patterns to those observed in previous assemblages.
+The final sample from levels A, B and C from the Solutrean occupations of the Shelter area consists of 914 artifacts.
+Of these, 201 (`r round(201*100/914, 2)`%) correspond to the indeterminate chert group.
+Local cherts correspond to `r 16+506+4+10+30` (`r round((16+506+4+10+30)*100/914, 2)`%) artefacts, mostly from T2 (n=506).
+Non-local cherts (including the TL group) correspond to `r 86+9+30+3+1+17+1` (`r round((86+9+30+3+1+17+1)*100/914, 2)`%) artefacts, mostly from T6 (n=86).
+
+When crossing the class data with type of raw material (@tbl-general-solutrean), we see that for local chert, especially T2, the majority of the debitage is composed of flakes and elongated blanks (83.9%).
+Cores are also present (10.5%), alongside a small percentage of core preparation and maintenance products (1.3%).
+In comparison to debitage products, retouched tools are scarce in the sample (\~3%).
+Non-local cherts, especially T6 and T8, show similar patterns to the local cherts, with high percentages of blanks (\>50%).
+Cores and core preparation and maintenance are also present in small percentages (\~3-6% and 3.5% only for T6, respectively), as well as retouched tools (\<10%).
+However, these specific results are truncated by the lack of shatter, which were not present in the sample used for the study (see Materials and Methods section).
+
+```{r}
+#| label: tbl-general-solutrean
+#| tbl-cap: Number and percentage of technotypological classes by raw material type from the Solutrean assemblage (levels A to C) of the Shelter area of Vale Boi.
+
+kable(solutrean_general)
+
+```
+
+The analysed Solutrean sample has a total of 43 retouched tools (complete and fragments), with `r 3+13+1` corresponding to local cherts and `r 7+1+2+3` to non-local cherts, with the remaining being cherts of unidentified type.
+These results are significantly reduced from the original dataset (Cascalheira et al., 2013), possibly due to no bucket coordinates being used.
+From the 43 retouched tools, 11 different typologies were identified (without accounting for fragments). 
+The typologies which are most frequently present in the studied sample are bifacial blanks and bifacial fragments (n=10).
+Two types of points were identified: shouldered points (n=3, including fragments) and tanged and winged points (n=4).
+Other frequent retouched typologies are retouched blanks (n=6), notches (n=5), splintered pieces (n=5) and endscrappers (n=4).
+Local types show a higher variability retouched typologies (9 types), although this may be related to the higher number of tools.
+In comparison, Type 6 shows only 5 different typologies.
+
+```{r}
+#| label: tbl-retouch-solutrean
+#| tbl-cap: Number and percentage of retouched tools by raw material of the Solutrean assemblage (level A to C) of the Shelter area.
+
+kable(solutrean_retouch_tab)
 
@@
   filter(RMTYPE == "TL") |> 
+  filter(Level != "5") |>
   select(Technocomplex, SUBTYPE, ID, CLASS, RETOUCHEDPIECETYPOLOGY)

### 2025-02-18 analysis/paper/paper.qmd
Macbook morning session upload: Complete assemblage description and tables moved to supplementary. Main paper tables simplified with categories collapsed for an easier read. tbl- chunks added tbl-subcap to present table groups (assemblage and retouched) in a single plot.

@@
   mutate(Class = case_when( # This alteration is being done on the original class column
-    Retouched == "Yes" ~ "Retouched piece", # Whenever retouch is present, class is now Retouched piece
+    Retouched == "Yes" ~ "RetouchedPiece", # Whenever retouch is present, class is now Retouched piece
     TRUE ~ Class)) |> 
@@
 # Tables with class by raw material type (n and %)
+##
+general_table <- phase_dataset |> 
+    dplyr::mutate(RMTYPE = case_when(
+    RMTYPE == "TYPE 1" | RMTYPE == "TYPE 2" | RMTYPE == "TYPE 3" | RMTYPE == "TYPE 4" | RMTYPE == "TYPE 5" ~ "Local",
+    RMTYPE == "TYPE 6" ~ "T6",
+    RMTYPE == "TYPE 7" ~ "T7",
+    RMTYPE == "TYPE 8" ~ "T8",
+    RMTYPE == "TYPE 9" ~ "T9",
+    RMTYPE == "TYPE 10" ~ "T10",
+    TRUE ~ RMTYPE)) |> 
+  dplyr::mutate(CLASS = case_when(
+    CLASS == "BlankFrag" ~ "Blank fragment",
+    CLASS == "BurinSpall" ~ "Burin spall",
+    CLASS == "CoreFrag" ~ "Core fragment",
+    CLASS == "CorePreparProd" ~ "Core prep product",
+    CLASS == "RetouchedPiece" ~ "Retouched",
+    CLASS == "RetouchedPieceFrag" ~ "Retouched fragment",
+    TRUE ~ CLASS)) |> 
+  dplyr::mutate(RETOUCHEDPIECETYPOLOGY = case_when(
+    RETOUCHEDPIECETYPOLOGY == "Retouched blade" | RETOUCHEDPIECETYPOLOGY == "Retouched bladelet" | RETOUCHEDPIECETYPOLOGY == "Retouched bladelet fragment" | RETOUCHEDPIECETYPOLOGY == "Retouched flake" | RETOUCHEDPIECETYPOLOGY == "Retouched flake fragment" ~ "Retouched blank",
+    RETOUCHEDPIECETYPOLOGY == "Splintered piece fragment" ~ "Splintered piece",
+    TRUE ~ RETOUCHEDPIECETYPOLOGY))
+
+general_table_solutrean <- solutrean_dataset |> 
+   dplyr::mutate(RMTYPE = case_when(
+    RMTYPE == "TYPE 1" | RMTYPE == "TYPE 2" | RMTYPE == "TYPE 3" | RMTYPE == "TYPE 4" | RMTYPE == "TYPE 5" ~ "Local",
+    RMTYPE == "TYPE 6" ~ "T6",
+    RMTYPE == "TYPE 7" ~ "T7",
+    RMTYPE == "TYPE 8" ~ "T8",
+    RMTYPE == "TYPE 9" ~ "T9",
+    RMTYPE == "TYPE 10" ~ "T10",
+    RMTYPE == "TYPE 11" ~ "T11",
+    TRUE ~ RMTYPE)) |> 
+  dplyr::mutate(CLASS = case_when(
+    CLASS == "ElongBlank" | CLASS == "Flake" ~ "Blank",
+    CLASS == "ElongBlankFrag" | CLASS == "FlakeFrag" ~ "Blank fragment",
+    CLASS == "Thinning flake" ~ "Blank",
+    CLASS == "CorePrepProd" ~ "Core prep product",
+    CLASS == "RetouchedPiece" | CLASS == "Retouched piece" ~ "Retouched",
+    CLASS == "Retouched piece fragment" ~ "Retouched fragment",
+    TRUE ~ CLASS))
+
 ## For Gravettian
-gravettian_general <- phase_dataset |>
+gravettian_general <- general_table |>
   filter(Level %in% c(6, 7)) |>
-  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
-                                        "TYPE 5", "TYPE 6","TYPE 7", 
-                                        "TYPE 8", "TYPE 9", "TYPE 10", "TL", "INDET"))) |>
+  filter(RMTYPE != "INDET") |> 
+  mutate(RMTYPE = factor(RMTYPE, levels = c("Local", "T6","T7", 
+                                        "T8", "T9", "T10", "TL"))) |>
   tabyl(CLASS, RMTYPE) |> 
@@
 ## For Proto-Solutrean (4E and top of 5)
-proto_general <- phase_dataset |> 
+proto_general <- general_table |> 
   filter(Level %in% c("4E", "4E/5")) |> 
   filter(CLASS != "Chip") |> 
-  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
-                                        "TYPE 5", "TYPE 6","TYPE 7", 
-                                        "TYPE 8", "TYPE 9", "TYPE 10", "TL", "INDET"))) |> 
+  filter(RMTYPE != "INDET") |> 
+  mutate(RMTYPE = factor(RMTYPE, levels = c("Local", "T6","T7", 
+                                        "T8", "T9", "T10", "TL"))) |> 
   tabyl(CLASS, RMTYPE) |> 
@@
 ## For Solutrean
-solutrean_general <- solutrean_dataset |> 
-  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
-                                        "TYPE 5", "TYPE 6","TYPE 7", 
-                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL", "INDET"))) |> 
+solutrean_general <- general_table_solutrean |> 
+  filter(RMTYPE != "INDET") |> 
+  mutate(RMTYPE = factor(RMTYPE, levels = c("Local", "T6","T7", 
+                                        "T8", "T9", "T10", "T11", "TL"))) |> 
   tabyl(CLASS, RMTYPE) |> 
@@
 ## Table preparation
-solutrean_retouch <- solutrean_dataset |>
-    mutate(CLASS = case_when(
-    CLASS == "Retouched piece" ~ "RetouchedPiece",
-    CLASS == "Retouched piece fragment" ~ "RetouchedPieceFrag",
-    TRUE ~ CLASS)) |>
-  filter(CLASS %in% c("RetouchedPiece", "RetouchedPieceFrag")) |>
-  select(RMTYPE, RETOUCHEDPIECETYPOLOGY) |>
+solutrean_retouch <- general_table_solutrean |>
+  filter(CLASS %in% c("Retouched", "Retouched fragment")) |>
+  filter(RMTYPE != "INDET") |> 
   filter(RETOUCHEDPIECETYPOLOGY != "") |> 
-  dplyr::mutate(RMTYPE = case_when(
-    RMTYPE == "TYPE 1" ~ "T1",
-    RMTYPE == "TYPE 2" ~ "T2",
-    RMTYPE == "TYPE 3" ~ "T3",
-    RMTYPE == "TYPE 4" ~ "T4",
-    RMTYPE == "TYPE 5" ~ "T5",
-    RMTYPE == "TYPE 6" ~ "T6",
-    RMTYPE == "TYPE 7" ~ "T7",
-    RMTYPE == "TYPE 8" ~ "T8",
-    RMTYPE == "TYPE 9" ~ "T9",
-    RMTYPE == "TYPE 10" ~ "T10",
-    RMTYPE == "TYPE 11" ~ "T11",
-    TRUE ~ RMTYPE))
-## Janitor tabyl for final table with n and %
-solutrean_retouch_tab <- solutrean_retouch |>
   tabyl(RETOUCHEDPIECETYPOLOGY, RMTYPE) |>
@@
   mutate(across(where(is.numeric), ~ ifelse(. == 0, " ", .))) |>
-  select(RETOUCHEDPIECETYPOLOGY, T1, T2, T5, T6, T7, T8, T11, INDET, Total) |> # Ordering values
+  select(RETOUCHEDPIECETYPOLOGY, Local, T6, T7, T8, T11, Total) |> # Ordering values
   rename(Typology = RETOUCHEDPIECETYPOLOGY)
@@
 ## Table preparation
-proto_retouch <- phase_dataset |>
-  filter(CLASS %in% c("RetouchedPiece", "RetouchedPieceFrag")) |>
+proto_retouch <- general_table |>
+  filter(CLASS %in% c("Retouched", "Retouched fragment")) |>
   filter(Level %in% c("4E", "4E/5")) |>
-  select(RMTYPE, RETOUCHEDPIECETYPOLOGY) |>
-  dplyr::mutate(RMTYPE = case_when(
-    RMTYPE == "TYPE 1" ~ "T1",
-    RMTYPE == "TYPE 2" ~ "T2",
-    RMTYPE == "TYPE 3" ~ "T3",
-    RMTYPE == "TYPE 4" ~ "T4",
-    RMTYPE == "TYPE 5" ~ "T5",
-    RMTYPE == "TYPE 6" ~ "T6",
-    RMTYPE == "TYPE 7" ~ "T7",
-    RMTYPE == "TYPE 8" ~ "T8",
-    RMTYPE == "TYPE 9" ~ "T9",
-    RMTYPE == "TYPE 10" ~ "T10",
-    RMTYPE == "TYPE 11" ~ "T11",
-    TRUE ~ RMTYPE))
-## Janitor tabyl for final table with n and %
-proto_retouch_tab <- proto_retouch |>
+  filter(!RMTYPE == "INDET") |> 
   tabyl(RETOUCHEDPIECETYPOLOGY, RMTYPE) |>
@@
   mutate(across(where(is.numeric), ~ ifelse(. == 0, " ", .))) |>
-  select(RETOUCHEDPIECETYPOLOGY, T1, T2, T3, T4, T5, T6, T7, T8, T9, TL, INDET, Total) |> # Ordering values
+  select(RETOUCHEDPIECETYPOLOGY, Local, T6, T7, T8, T9, TL, Total) |> # Ordering values
   rename(Typology = RETOUCHEDPIECETYPOLOGY)
@@
 ## Table preparation
-gravettian_retouch <- phase_dataset |> 
-  filter(CLASS %in% c("RetouchedPiece", "RetouchedPieceFrag")) |>
+gravettian_retouch <- general_table |> 
+  filter(CLASS %in% c("Retouched", "Retouched fragment")) |>
   filter(Level %in% c(6, 7)) |> 
-  select(RMTYPE, RETOUCHEDPIECETYPOLOGY) |> 
-  dplyr::group_by(RMTYPE) |>
-  dplyr::mutate(RMTYPE = case_when(
-    RMTYPE == "TYPE 1" ~ "T1",
-    RMTYPE == "TYPE 2" ~ "T2",
-    RMTYPE == "TYPE 3" ~ "T3",
-    RMTYPE == "TYPE 4" ~ "T4",
-    RMTYPE == "TYPE 5" ~ "T5",
-    RMTYPE == "TYPE 6" ~ "T6",
-    RMTYPE == "TYPE 7" ~ "T7",
-    RMTYPE == "TYPE 8" ~ "T8",
-    RMTYPE == "TYPE 9" ~ "T9",
-    RMTYPE == "TYPE 10" ~ "T10",
-    TRUE ~ RMTYPE))
-## Janitor tabyl for final table with n and %
-gravettian_retouched_tab <- gravettian_retouch |> 
+  filter(!RMTYPE == "INDET") |> 
   tabyl(RETOUCHEDPIECETYPOLOGY, RMTYPE) |>
@@
   mutate(across(where(is.numeric), ~ ifelse(. == 0, " ", .))) |>
-  select(RETOUCHEDPIECETYPOLOGY, T1, T2, T3, T4, T5, T6, T7, T8, T9, T10, TL, INDET, Total) |> # Ordering values
+  select(RETOUCHEDPIECETYPOLOGY, Local, T6, T7, T8, T9, T10, TL, Total) |> # Ordering values
   rename(Typology = RETOUCHEDPIECETYPOLOGY)
@@
 ```{r}
-#| label: tbl-general-gravettian
-#| tbl-cap: Number and percentage of technotypological classes by raw material type from the Gravettian assemblage (levels 6 and 7) of the Terrace area of Vale Boi.
-
-kable(gravettian_general)
+#| label: tbl-general
+#| tbl-cap: Number and percentage of technotypological classes by raw material type of the Gravettian (a), Proto-Solutrean (b) and Solutrean (c) assemblages.
+#| tbl-subcap: 
+#| - "Number and percentage of technotypological classes by raw material from the Gravettian assemblage (levels 6 and 7) of the Terrace area."
+#| - "Number and percentage of technotypological classes by raw material from the Proto-Solutrean assemblage (levels 4E and 5) of the Terrace area."
+#| - "Number and percentage of technotypological classes by raw material from the Solutrean assemblage (levels A-C) of the Shelter area."
+#| layout-nrow: 4
+
+kable(head(gravettian_general))
+kable(head(proto_general))
+kable(head(solutrean_general))
 
@@
 ```{r}
-#| label: tbl-retouch-gravettian
-#| tbl-cap: Number and percentage of retouched tools by raw material of the Gravettian assemblage (levels 6 and 7) of the Terrace area.
-
-kable(gravettian_retouched_tab)
+#| label: tbl-retouch
+#| tbl-cap: Number and percentage of retouched tools by raw material of the Gravettian (a), Proto-Solutrean (b) and Solutrean (c) assemblages.
+Retouched blank typology includes flakes, blades and bladelets and respective fragments.
+#| tbl-subcap: 
+#| - "Number and percentage of retouched tools by raw material from the Gravettian assemblage (levels 6 and 7) of the Terrace area."
+#| - "Number and percentage of retouched tools by raw material from the Proto-Solutrean assemblage (levels 4E and 5) of the Terrace area."
+#| - "Number and percentage of retouched tools by raw material from the Solutrean assemblage (levels A-C) of the Shelter area."
+#| layout-nrow: 4
+
+kable(head(gravettian_retouched))
+kable(head(proto_retouch))
+kable(head(solutrean_retouch))
 

### 2025-02-18 analysis/paper/paper.qmd
Graphs followed similar alterations of tables. Colors homogenised by RM type through the complete document.

@@
     CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "BlankFragProx", # To add proximal fragments to ratio
-    TRUE ~ CLASS))
-
-# Scatterplot for width and length
-solutrean_scatter <- solutrean_stats |>
-  filter(CLASS %in% c("Flake","ElongatedProd", "Core", "RetouchedPiece")) |>
-  filter(MAXWIDTH !="0" & LENGTH != "0" & !RMTYPE %in% c("INDET","GB","TL")) |>
+    TRUE ~ CLASS)) |> 
   dplyr::mutate(RMTYPE = case_when(
-    RMTYPE == "TYPE 1" ~ "T1",
-    RMTYPE == "TYPE 2" ~ "T2",
-    RMTYPE == "TYPE 3" ~ "T3",
-    RMTYPE == "TYPE 4" ~ "T4",
-    RMTYPE == "TYPE 5" ~ "T5",
+    RMTYPE == "TYPE 1" | RMTYPE == "TYPE 2" | RMTYPE == "TYPE 3" | RMTYPE == "TYPE 4" | RMTYPE == "TYPE 5" ~ "Local",
     RMTYPE == "TYPE 6" ~ "T6",
@@
     RMTYPE == "TYPE 11" ~ "T11",
-    TRUE ~ RMTYPE)) |>
-   mutate(RMTYPE = factor(RMTYPE, levels = c("T1", "T2", "T3", "T4", "T5", "T6","T7", 
-                                        "T8", "T9", "T10", "T11"))) |>
+    TRUE ~ RMTYPE))
+
+# Scatterplot for width and length
+solutrean_scatter <- solutrean_stats |>
+  filter(CLASS %in% c("Flake","ElongatedProd", "Core", "RetouchedPiece")) |>
+  filter(MAXWIDTH !="0" & LENGTH != "0") |>
+  mutate(RMTYPE = factor(RMTYPE, levels = c("Local", "T6","T7", 
+                                        "T8", "T9", "T10", "T11", "TL"))) |>
   mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) |>
   ggplot(aes(x = LENGTH, y = MAXWIDTH, fill = RMTYPE)) +
-  geom_point(alpha = 0.9, size = 3, shape = 21, colour = "grey") +
+  geom_point(alpha = 0.9, size = 3, shape = 21, colour = "black") +
   facet_wrap(~CLASS) +
   labs(x="Length (mm)", y="Max. Width (mm") +
-  scale_fill_manual(values = c("#8B7B8B", "#FFE1FF", "#EED2EE", "#D8BFD8",
-                              "#CDB5CD", "#2B4F60" , "#4A708B", "#798777",
-                            "#9FB6CD", "#BDD2B6", "#BDC7C9"))
+  scale_fill_manual(values = c("#D49B54", "#8B6969", "#CDAF95", "#2B4F60",
+                              "tomato4", "#BDC7C9", "#798777"))
 
@@
   filter(CLASS %in% c("Flake", "ElongatedProd", "Core", "RetouchedPiece")) |> 
-  filter(WEIGHT !="0" & !RMTYPE %in% c("INDET","GB","TL")) |>
-  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
-                                        "TYPE 5", "TYPE 6","TYPE 7", 
-                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) |>
+  filter(WEIGHT !="0") |>
+  mutate(RMTYPE = factor(RMTYPE, levels = c("Local", "T6","T7", 
+                                        "T8", "T9", "T10", "T11", "TL"))) |>
   mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) |>
@@
   facet_wrap(~CLASS, scales = "free_y") +
-  stat_summary(fun.y=mean, geom="point", shape=23) +
+  stat_summary(fun=mean, geom="point", shape=23) +
   labs(x="Raw material types" ,y="Weight (gr)") +
   #coord_cartesian(ylim =  c(0, 50)) +
-  scale_fill_manual(values = c("#8B7B8B", "#FFE1FF", "#EED2EE", "#D8BFD8",
-                              "#CDB5CD", "#2B4F60" , "#4A708B", "#798777",
-                            "#9FB6CD", "#BDD2B6", "#BDC7C9")) +
+  scale_fill_manual(values = c("#D49B54", "#8B6969", "#CDAF95", "#2B4F60",
+                              "tomato4", "#BDC7C9", "#798777")) +
   theme(axis.text.x=element_blank(),
@@
   filter(Level %in% c("4E", "4E/5")) |> 
-  filter(RMTYPE != "INDET") |> 
+  filter(!RMTYPE == "INDET") |> 
   mutate(CLASS = case_when(
@@
     CLASS == "Blank" & elong <= 0 ~ "ElongatedProd",
-    TRUE ~ CLASS))
-
-# Scatterplots for width and length
-proto_scatter <- proto_stats |>
-  filter(CLASS %in% c("Flake","ElongatedProd", "Core", "RetouchedPiece")) |>
-  filter(MAXWIDTH !="0" & LENGTH != "0" & !RMTYPE %in% c("INDET","GB","TL")) |>
+    TRUE ~ CLASS)) |> 
   dplyr::mutate(RMTYPE = case_when(
-    RMTYPE == "TYPE 1" ~ "T1",
-    RMTYPE == "TYPE 2" ~ "T2",
-    RMTYPE == "TYPE 3" ~ "T3",
-    RMTYPE == "TYPE 4" ~ "T4",
-    RMTYPE == "TYPE 5" ~ "T5",
+    RMTYPE == "TYPE 1" | RMTYPE == "TYPE 2" | RMTYPE == "TYPE 3" | RMTYPE == "TYPE 4" | RMTYPE == "TYPE 5" ~ "Local",
     RMTYPE == "TYPE 6" ~ "T6",
@@
     RMTYPE == "TYPE 10" ~ "T10",
-    TRUE ~ RMTYPE)) |>
-   mutate(RMTYPE = factor(RMTYPE, levels = c("T1", "T2", "T3", "T4", "T5", "T6","T7", 
-                                        "T8", "T9", "T10"))) |>
+    TRUE ~ RMTYPE))
+
+# Scatterplots for width and length
+proto_scatter <- proto_stats |>
+  filter(CLASS %in% c("Flake","ElongatedProd", "Core", "RetouchedPiece")) |>
+  filter(MAXWIDTH !="0" & LENGTH != "0") |>
+   mutate(RMTYPE = factor(RMTYPE, levels = c("Local", "T6","T7", 
+                                        "T8", "T9", "T10", "TL"))) |>
   mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) |>
   ggplot(aes(x = LENGTH, y = MAXWIDTH, fill = RMTYPE)) +
-  geom_point(alpha = 0.9, size = 3, shape = 21, colour = "grey") +
+  geom_point(alpha = 0.9, size = 3, shape = 21, colour = "black") +
   facet_wrap(~CLASS) +
   labs(x="Length (mm)", y="Max. Width (mm") +
-  scale_fill_manual(values = c("#8B7B8B", "#FFE1FF", "#EED2EE", "#D8BFD8",
-                              "#CDB5CD", "#2B4F60" , "#4A708B", "#798777",
-                            "#9FB6CD", "#BDD2B6"))
+  scale_fill_manual(values = c("#D49B54", "#8B6969", "#CDAF95", "#2B4F60",
+                              "tomato4", "#BDD2B6", "#798777", "#BDC7C9"))
 
@@
   filter(CLASS %in% c("Flake","ElongatedProd", "Core", "RetouchedPiece")) |> 
-  filter(WEIGHT !="0" & !RMTYPE %in% c("INDET","GB","TL")) |>
-  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
-                                        "TYPE 5", "TYPE 6","TYPE 7", 
-                                        "TYPE 8", "TYPE 9", "TYPE 10"))) |>
+  filter(WEIGHT !="0") |>
+  mutate(RMTYPE = factor(RMTYPE, levels = c("Local", "T6", "T7", "T8", "T9", "T10", "TL"))) |>
   mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) |>
@@
   #coord_cartesian(ylim =  c(0, 50)) +
-  scale_fill_manual(values = c("#8B7B8B", "#FFE1FF", "#EED2EE", "#D8BFD8",
-                              "#CDB5CD", "#2B4F60" , "#4A708B", "#798777",
-                            "#9FB6CD", "#BDD2B6")) +
+  scale_fill_manual(values = c("#D49B54", "#8B6969", "#CDAF95", "#2B4F60",
+                              "tomato4", "#BDD2B6", "#798777", "#BDC7C9")) +
   theme(axis.text.x=element_blank(),
@@
     CLASS == "Blank" & elong <= 0 ~ "ElongatedProd",
-    TRUE ~ CLASS))
-
-# Scatterplots for width and length
-gravettian_scatter <- gravettian_stats |>
-  filter(CLASS %in% c("Flake","ElongatedProd", "Core", "RetouchedPiece")) |>
-  filter(MAXWIDTH !="0" & LENGTH != "0" & !RMTYPE %in% c("INDET","GB","TL")) |>
+    TRUE ~ CLASS)) |> 
   dplyr::mutate(RMTYPE = case_when(
-    RMTYPE == "TYPE 1" ~ "T1",
-    RMTYPE == "TYPE 2" ~ "T2",
-    RMTYPE == "TYPE 3" ~ "T3",
-    RMTYPE == "TYPE 4" ~ "T4",
-    RMTYPE == "TYPE 5" ~ "T5",
+    RMTYPE == "TYPE 1" | RMTYPE == "TYPE 2" | RMTYPE == "TYPE 3" | RMTYPE == "TYPE 4" | RMTYPE == "TYPE 5" ~ "Local",
     RMTYPE == "TYPE 6" ~ "T6",
@@
     RMTYPE == "TYPE 10" ~ "T10",
-    TRUE ~ RMTYPE)) |>
-   mutate(RMTYPE = factor(RMTYPE, levels = c("T1", "T2", "T3", "T4", "T5", "T6","T7", 
-                                        "T8", "T9", "T10"))) |>
+    TRUE ~ RMTYPE))
+
+# Scatterplots for width and length
+gravettian_scatter <- gravettian_stats |>
+  filter(CLASS %in% c("Flake","ElongatedProd", "Core", "RetouchedPiece")) |>
+  filter(MAXWIDTH !="0" & LENGTH != "0") |>
+   mutate(RMTYPE = factor(RMTYPE, levels = c("Local", "T6","T7", 
+                                        "T8", "T9", "T10", "TL"))) |>
   mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) |>
   ggplot(aes(x = LENGTH, y = MAXWIDTH, fill = RMTYPE)) +
-  geom_point(alpha = 0.9, size = 3, shape = 21, colour = "grey") +
+  geom_point(alpha = 0.9, size = 3, shape = 21, colour = "black") +
   facet_wrap(~CLASS) +
   labs(x="Length (mm)", y="Max. Width (mm") +
-  scale_fill_manual(values = c("#8B7B8B", "#FFE1FF", "#EED2EE", "#D8BFD8",
-                              "#CDB5CD", "#2B4F60" , "#4A708B", "#798777",
-                            "#9FB6CD", "#BDD2B6"))
+  scale_fill_manual(values = c("#D49B54", "#8B6969", "#CDAF95", "#2B4F60",
+                              "#8B7D6B", "#BDD2B6", "#798777", "#BDC7C9"))
 
@@
   filter(CLASS %in% c("Flake","ElongatedProd", "Core", "RetouchedPiece")) |> 
-  filter(WEIGHT !="0" & !RMTYPE %in% c("INDET","GB","TL")) |>
-  filter(!(RMTYPE == "TYPE 4" & WEIGHT > 40)) |> # Removed a single outlier at 60 weight while ensuring that the data remains consistent. Only mean changed significantly and was normalised
-  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
-                                        "TYPE 5", "TYPE 6","TYPE 7", 
-                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) |>
+  filter(WEIGHT !="0") |>
+  filter(!(RMTYPE == "Local" & WEIGHT > 40)) |> # Removed a single outlier at 60 weight while ensuring that the data remains consistent. Only mean changed significantly and was normalised
+  mutate(RMTYPE = factor(RMTYPE, levels = c("Local", "T6", "T7", "T8", "T9", "T10", "TL"))) |>
   mutate(CLASS = factor(CLASS, levels = c("Flake","ElongatedProd", "Core", "RetouchedPiece"))) |>
@@
   #coord_cartesian(ylim =  c(0, 60)) +
-  scale_fill_manual(values = c("#8B7B8B", "#FFE1FF", "#EED2EE", "#D8BFD8",
-                              "#CDB5CD", "#2B4F60" , "#4A708B", "#798777",
-                            "#9FB6CD", "#BDD2B6")) +
+  scale_fill_manual(values = c("#D49B54", "#8B6969", "#CDAF95", "#2B4F60",
+                              "#8B7D6B", "#BDD2B6", "#798777", "#BDC7C9")) +
   theme(axis.text.x=element_blank(),
@@
   scale_colour_manual(values = c("#D49B54", "#8B6969", "#CDAF95", "#2B4F60",
-                              "#8B7D6B", "#BDD2B6", "#798777", "#BDC7C9"))
+                              "#BDD2B6"))
 
@@
   scale_colour_manual(values = c("#D49B54", "#8B6969", "#CDAF95", "#2B4F60",
-                              "#8B7D6B", "#BDD2B6", "#798777", "#BDC7C9"))
+                              "#BDD2B6", "#BDC7C9", "#798777"))
 

### 2025-02-19 analysis/paper/paper.qmd
Macbook night work session.

@@
 
+```{r}
+#| label: tbl-expectations
+#| tbl-cap: A table.
+
+hypotheses <- readr::read_csv(here::here('analysis/data/tables/hypotheses_short.csv'))
+
+knitr::kable(head(hypotheses))
+
+```
+
 Despite these expectations, it is important to notice that, especially in occupations with a high percentage of non-local cherts at the site (e.g., Gravettian occupations of level 7 and 6 of the Terrace at Vale Boi), these technological may not reflect a preoccupation with intensive management of a scarce and limited non-local resource, and be more a reflection of the type of provisioning applied following the suggested model by Kuhn (2004).
@@
 
-## Gravettian (levels 6-7)
-
-The sample studied for the Gravettian occupation (\>2012 excavations) of the Terrace area included 766 artifacts.
-Of these, 209 (`r round(209*100/766, 2)`%) have an indeterminate chert type (@tbl-general-gravettian).
-Chalcedony (T1) and local cherts (T2-5) show a total of `r 35+110+17+13+72` artefacts (`r round((35+110+17+13+72)*100/766, 2)`%).
-Non-local cherts show different numbers of artefacts depending on the type, with T6 and T7 showing the highest values (169 or `r round(169*100/766, 2)`% of the total assemblage, and 86 or `r round(86*100/766, 2)`%, respectively); the other non-local cherts have artefact counts lower than 20 per type.
-The TL cherts also show smaller numbers, with 12 artefacts, corresponding to 7 different types of TL cherts (table s++ in SOM ++).
-
-@tbl-general-gravettian shows the number and percentages of technological classes per chert type.
-Local cherts show similar patterns between them, with T2 and T5 being the types with higher numbers of artefacts (n=110 and n=72, respectively).
-For all local cherts, blanks (including blank fragments) represent more than 50% of the artefacts.
-Cores and core maintenance products are only present in T2 and T5, although in small frequencies: for T2, cores represent 13.6% (n=15 including core fragments) of the group and core preparation and maintenance products represent 1.8% (n=2).
-For T5 only 4 core fragments and 1 core preparation and maintenance product are present, all frequencies between \~5-1%.
-Retouched tools seem to be especially relevant in T2 (n= 20, \~18% including retouched piece fragments), making them the most frequent class after blanks.
-T5 shows similar percentages of retouched tools and shatter (\~12% and \~14%, respectively).
-T1 (chalcedony) shows similar patterns to other local cherts, although with higher percentages of shatter (34.3%).
-
-When compared to local cherts, non-local chert types (especially T6 and T7) show similar percentages of blanks (n=`r 44+24` or 40% and `r 21+21`\`or \~50%, respectively) but higher percentages of retouched tools (\~30%).
-Cores, core fragments and core preparation and maintenance products are also present in all non-local cherts (excluding TL) even if in percentages below 4% (n\<4), as well as shatter (\~12%).
-The exception is T6 that shows the highest numbers of cores and core fragments (n=8), although with low frequencies (4.6%).
-The TL cherts show mostly blanks and blank fragments (n=3 and n=2, respectively), and retouched tools (n=4).
-
 ```{r}
@@
 
-kable(head(gravettian_general))
-kable(head(proto_general))
-kable(head(solutrean_general))
-
-```
-
-As seen in table @tbl-retouch-gravettian, 170 retouched tools were identified, with 39 (`r round(39*100/170, 2)`%) having an unidentified chert type.
-From the identified chert types, `r 3+20+4+2+9` (`r round((3+20+4+2+9)*100/170, 2)`%) of retouched tools are from local types, `r 58+24+3+1+3` (`r round((58+24+3+1+3)*100/170, 2)`%) from non-local cherts and 4 are from the TL group.
-From all chert types, T6 shows the highest numbers of retouched pieces (n=58, `r round(58*100/170, 2)`%).
-Regarding typologies, 15 types of retouched tools were identified (excluding retouched tool fragments).
-All these typologies were found in the non-local chert group, while only 7 retouched tool typologies were identified in local cherts/chalcedony.
-The most common typologies identified in the assemblage (and for all chert types) were burins (n=52), endscrappers (n=17), retouched flakes (n=30) and splintered pieces (n=26).
-From all these, the majority were produced in non-local chert, mainly T6 and in lesser numbers T7.
-Although in lesser numbers as well, backed bladelets were also identified (n=6), but only present in non-local chert types (T6 and T7).
-The presence of burin spalls in local chert, especially T2 (n=9) and non-local cherts (T6 with n=12 and T7 with n=4) is expected due to the prevalence of burins (@tbl-general-gravettian).
+kable(gravettian_general)
+kable(proto_general)
+kable(solutrean_general)
+
+```
 
@@
 
-kable(head(gravettian_retouched))
-kable(head(proto_retouch))
-kable(head(solutrean_retouch))
+kable(gravettian_retouch)
+kable(proto_retouch)
+kable(solutrean_retouch)
 
@@
 #| label: fig-gravettian-boxplots
-#| fig-cap: Boxplots Solutrean
+#| fig-cap: Boxplots Gravettian
 
 ggpubr::ggarrange(gravettian_scatter, gravettian_stats_weight, labels = c("a", "b"), ncol = 2, common.legend = TRUE, legend = "bottom")
-
-```
-
-## Proto-Solutrean (levels 4E and top of level 5)
-
-The final lithic sample from the Proto-Solutrean assemblage, consisting of the levels 4E and top of level 5, consists of 635 artifacts.
-Of these, 167 (`r round(167*100/635, 2)`%) have an indeterminate chert type.
-Local cherts correspond to `r 32+62+32+158+51` artefacts (`r round((32+62+32+158+51)*100/635, 2)`%), with T4 being the most present local chert (n=158 or `r round(158*100/(32+62+32+158+51), 2)`% of the local types).
-Non-local types are present in smaller numbers (n=`r 60+47+13+9+1` or `r round((60+47+13+9+1)*100/635, 2)`%), with T6 and T7 being the most relevant (n=60 and n=47, respectively), and the TL group includes 3 artefacts, corresponding to 2 types of TL cherts (Table s++ in SOM ++).
-
-When crossing the lithic class with chert type (@tbl-general-proto), we see that the local types show high frequency of blanks (\>50%, including blank fragments), followed by shatter (\~11-27%).
-Cores are present in small numbers (n=`r 4+5+13+6`), corresponding to ~9-12% of the classes in each local type, with the exception of T1 (chalcedony), with no cores.
-Core preparation and maintenance products are absent in all local cherts.
-For non-local types, especially those with higher number of artefacts (T6-T8), blanks and blank fragments are the most frequent class (48.3%, 53.1% and 69.2%, respectively).
-Retouched tools are present in smaller frequencies for the previous mentioned chert types (\~17-23%).
-For T6 the absence of cores but the presence of shatter and a single core preparation and maintenance product suggests knapping on site; all other non-local types show at least 1 core and frequently the presence of shatter.
-As expected, the TL artifacts include a blank and a retouched.
-However, a shatter was also identified.
-
-```{r}
-#| label: tbl-general-proto
-#| tbl-cap: Number and percentage of technotypological classes by raw material type from the Proto-Solutrean assemblage (levels 4E and top of level 5) of the Terrace area of Vale Boi.
-
-kable(proto_general)
-
-```
-
-In total, 65 chert retouched tools were identified, with 23.1% (n=15) having an unidentified chert type (@tbl-retouch-proto).
-A total of `r 1+10+2+8+5` (`r round((1+10+2+8+5)*100/65, 2)`%) retouched tools correspond to local cherts, while `r 10+9+3+1+1` (`r round((10+9+3+1+1)*100/65, 2)`%) correspond to non-local and TL cherts.
-As seen in @tbl-retouch-proto, 14 different types of retouched tools were identified.
-Endscrappers (26.2%), splintered pieces (17%), burins (12.3%) and retouched flakes (7.7%) were the most prevalent.
-Local raw materials show 9 different typologies, with Vale Comprido points, Solutrean retouch and notches being present only in the local types, even if in small numbers.
-Non-local cherts show 10 different typologies of retouched tools.
-Perforator-endscrapper, truncation, backed bladelet and retouched blades typologies can only be found in non-local varieties, albeit also in small numbers.
-
-```{r}
-#| label: tbl-retouch-proto
-#| tbl-cap: Number and percentage of retouched tools by raw material of the Proto-Solutrean assemblage (levels 4E and top of level 5) of the Terrace area.
-
-kable(proto_retouch_tab)
 
@@
 #| label: fig-proto-boxplots
-#| fig-cap: Boxplots Proto-Solutrean
+#| fig-cap: Boxplots Proto-solutrean.
 
 ggpubr::ggarrange(proto_scatter, proto_stats_weight, labels = c("a", "b"), ncol = 2, common.legend = TRUE, legend = "bottom")
-
-```
-
-## Solutrean (levels A to C)
-
-The Solutrean assemblages show similar patterns to those observed in previous assemblages.
-The final sample from levels A, B and C from the Solutrean occupations of the Shelter area consists of 914 artifacts.
-Of these, 201 (`r round(201*100/914, 2)`%) correspond to the indeterminate chert group.
-Local cherts correspond to `r 16+506+4+10+30` (`r round((16+506+4+10+30)*100/914, 2)`%) artefacts, mostly from T2 (n=506).
-Non-local cherts (including the TL group) correspond to `r 86+9+30+3+1+17+1` (`r round((86+9+30+3+1+17+1)*100/914, 2)`%) artefacts, mostly from T6 (n=86).
-
-When crossing the class data with type of raw material (@tbl-general-solutrean), we see that for local chert, especially T2, the majority of the debitage is composed of flakes and elongated blanks (83.9%).
-Cores are also present (10.5%), alongside a small percentage of core preparation and maintenance products (1.3%).
-In comparison to debitage products, retouched tools are scarce in the sample (\~3%).
-Non-local cherts, especially T6 and T8, show similar patterns to the local cherts, with high percentages of blanks (\>50%).
-Cores and core preparation and maintenance are also present in small percentages (\~3-6% and 3.5% only for T6, respectively), as well as retouched tools (\<10%).
-However, these specific results are truncated by the lack of shatter, which were not present in the sample used for the study (see Materials and Methods section).
-
-```{r}
-#| label: tbl-general-solutrean
-#| tbl-cap: Number and percentage of technotypological classes by raw material type from the Solutrean assemblage (levels A to C) of the Shelter area of Vale Boi.
-
-kable(solutrean_general)
-
-```
-
-The analysed Solutrean sample has a total of 43 retouched tools (complete and fragments), with `r 3+13+1` corresponding to local cherts and `r 7+1+2+3` to non-local cherts, with the remaining being cherts of unidentified type.
-These results are significantly reduced from the original dataset (Cascalheira et al., 2013), possibly due to no bucket coordinates being used.
-From the 43 retouched tools, 11 different typologies were identified (without accounting for fragments). 
-The typologies which are most frequently present in the studied sample are bifacial blanks and bifacial fragments (n=10).
-Two types of points were identified: shouldered points (n=3, including fragments) and tanged and winged points (n=4).
-Other frequent retouched typologies are retouched blanks (n=6), notches (n=5), splintered pieces (n=5) and endscrappers (n=4).
-Local types show a higher variability retouched typologies (9 types), although this may be related to the higher number of tools.
-In comparison, Type 6 shows only 5 different typologies.
-
-```{r}
-#| label: tbl-retouch-solutrean
-#| tbl-cap: Number and percentage of retouched tools by raw material of the Solutrean assemblage (level A to C) of the Shelter area.
-
-kable(solutrean_retouch_tab)
 
@@
 #| label: fig-solutrean-boxplots
-#| fig-cap: Boxplots Gravettian.
+#| fig-cap: Boxplots Solutrean.
 
@@
 As mentioned before, this study aimed to test two expectations related to the use and knapping of local and non-local raw materials, in specific, chert and chalcedony, and their relation to settlement and mobility strategies, across several UP occupations.
-As such, our method derived the following hypotheses to test: a) use lives derived from knapping strategies; b) reduction intensity and tools-to-debitage ratios; c) core size and blank-to-core ratios; d) reduction intensity and tool reworking; e) tool typology diversity and specialization (@tbl-hypotheses).
-
-```{r}
-#| label: tbl-hypotheses
-#| tbl-cap: A table.
-
-hypotheses <- readr::read_csv(here::here('analysis/data/tables/hypotheses_short.csv'))
-
-knitr::kable(head(hypotheses))
-
-```
+As such, our method derived the following hypotheses to test: a) use lives derived from knapping strategies; b) reduction intensity and tools-to-debitage ratios; c) core size and blank-to-core ratios; d) reduction intensity and tool reworking; e) tool typology diversity and specialization (@tbl-expectations).
 

### 2025-02-24 analysis/paper/paper.qmd
Corrections to measurement graphs. Joined elong and flake in a single blank category. Removed retouched from the graphs.

@@
    mutate(CLASS = case_when(
-    CLASS == "Thinning flake" ~ "ThinningFlake",
-    CLASS == "ElongBlank" ~ "ElongatedProd",
+    CLASS == "ElongBlank" ~ "Blank",
+    CLASS == "Flake" ~ "Blank",
     CLASS == "FlakeFrag" | CLASS == "ElongBlankFrag" ~ "BlankFrag",
-    CLASS == "Retouched piece" ~ "RetouchedPiece",
-    TRUE ~ CLASS)) |> 
-  mutate(CLASS = case_when(
-    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "BlankFragProx", # To add proximal fragments to ratio
     TRUE ~ CLASS)) |> 
@@
 solutrean_scatter <- solutrean_stats |>
-  filter(CLASS %in% c("Flake","ElongatedProd", "Core", "RetouchedPiece")) |>
+  filter(CLASS %in% c("Blank", "Core")) |>
   filter(MAXWIDTH !="0" & LENGTH != "0") |>
@@
                                         "T8", "T9", "T10", "T11", "TL"))) |>
-  mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) |>
+  mutate(CLASS = factor(CLASS, levels = c("Blank", "Core"))) |>
   ggplot(aes(x = LENGTH, y = MAXWIDTH, fill = RMTYPE)) +
@@
   facet_wrap(~CLASS) +
-  labs(x="Length (mm)", y="Max. Width (mm") +
+  labs(x="Length (mm)", y="Max. Width (mm)") +
   scale_fill_manual(values = c("#D49B54", "#8B6969", "#CDAF95", "#2B4F60",
@@
 solutrean_stats_weight <- solutrean_stats |>
-  filter(CLASS %in% c("Flake", "ElongatedProd", "Core", "RetouchedPiece")) |> 
+  filter(CLASS %in% c("Blank", "Core")) |> 
   filter(WEIGHT !="0") |>
@@
                                         "T8", "T9", "T10", "T11", "TL"))) |>
-  mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) |>
+  mutate(CLASS = factor(CLASS, levels = c("Blank", "Core"))) |>
   ggplot(aes(RMTYPE, WEIGHT, fill = RMTYPE)) +
@@
   filter(!RMTYPE == "INDET") |>
-  mutate(CLASS = case_when(
-    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "BlankFragProx", # To add proximal fragments to ratio
-    TRUE ~ CLASS)) |>
-  mutate(elong = MAXWIDTH*2-LENGTH) |>
-  mutate(CLASS = case_when(
-    CLASS == "Blank" & elong > 0 ~ "Flake", # QUESTION: do I keep flakes and elong separate or join into blanks?
-    CLASS == "Blank" & elong <= 0 ~ "ElongatedProd",
-    TRUE ~ CLASS)) |> 
   dplyr::mutate(RMTYPE = case_when(
@@
 proto_scatter <- proto_stats |>
-  filter(CLASS %in% c("Flake","ElongatedProd", "Core", "RetouchedPiece")) |>
+  filter(CLASS %in% c("Blank", "Core")) |>
   filter(MAXWIDTH !="0" & LENGTH != "0") |>
@@
                                         "T8", "T9", "T10", "TL"))) |>
-  mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) |>
+  mutate(CLASS = factor(CLASS, levels = c("Blank", "Core"))) |>
   ggplot(aes(x = LENGTH, y = MAXWIDTH, fill = RMTYPE)) +
@@
 proto_stats_weight <- proto_stats |>
-  filter(CLASS %in% c("Flake","ElongatedProd", "Core", "RetouchedPiece")) |> 
+  filter(CLASS %in% c("Blank", "Core")) |> 
   filter(WEIGHT !="0") |>
   mutate(RMTYPE = factor(RMTYPE, levels = c("Local", "T6", "T7", "T8", "T9", "T10", "TL"))) |>
-  mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) |>
+  mutate(CLASS = factor(CLASS, levels = c("Blank", "Core"))) |>
   ggplot(aes(RMTYPE, WEIGHT, fill = RMTYPE)) +
@@
   filter(RMTYPE != "INDET") |>
-  mutate(CLASS = case_when(
-    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "BlankFragProx", # To add proximal fragments to ratio
-    TRUE ~ CLASS)) |> 
-  mutate(elong = MAXWIDTH*2-LENGTH) |>
-  mutate(CLASS = case_when(
-    CLASS == "Blank" & elong > 0 ~ "Flake", # QUESTION: do I keep flakes and elong separate or join into blanks?
-    CLASS == "Blank" & elong <= 0 ~ "ElongatedProd",
-    TRUE ~ CLASS)) |> 
   dplyr::mutate(RMTYPE = case_when(
@@
 gravettian_scatter <- gravettian_stats |>
-  filter(CLASS %in% c("Flake","ElongatedProd", "Core", "RetouchedPiece")) |>
+  filter(CLASS %in% c("Blank", "Core")) |>
   filter(MAXWIDTH !="0" & LENGTH != "0") |>
@@
 gravettian_stats_weight <- gravettian_stats |>
-  filter(CLASS %in% c("Flake","ElongatedProd", "Core", "RetouchedPiece")) |> 
+  filter(CLASS %in% c("Blank", "Core")) |> 
   filter(WEIGHT !="0") |>
@@
   mutate(RMTYPE = factor(RMTYPE, levels = c("Local", "T6", "T7", "T8", "T9", "T10", "TL"))) |>
-  mutate(CLASS = factor(CLASS, levels = c("Flake","ElongatedProd", "Core", "RetouchedPiece"))) |>
+  mutate(CLASS = factor(CLASS, levels = c("Blank", "Core"))) |>
   ggplot(aes(RMTYPE, WEIGHT, fill = RMTYPE)) +

### 2025-02-24 analysis/paper/paper.qmd
Macbook night session commit.

@@
 
+```{r}
+#| label: stats-wilcox
+
+wilcox_t <- phase_dataset |> 
+  dplyr::mutate(Prov = case_when(
+    RMTYPE == "TYPE 1" | RMTYPE == "TYPE 2" | RMTYPE == "TYPE 3" | RMTYPE == "TYPE 4" | RMTYPE == "TYPE 5" ~ "Local",
+    RMTYPE == "TYPE 6" | RMTYPE == "TYPE 7" | RMTYPE == "TYPE 8" | RMTYPE == "TYPE 9" ~ "Non-local",
+    TRUE ~ RMTYPE)) |> 
+  filter(Prov %in% c("Local", "Non-local")) |> 
+  select(ID, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT, Prov, Level)
+
+wt_blank_proto <- wilcox_t |> 
+  filter(Level %in% c("4E", "4E/5")) |> 
+  filter(CLASS == "Blank")
+
+dplyr::group_by(wt_blank_proto, Prov) |> 
+  summarise(count = n(), median = median(WEIGHT, na.rm = TRUE))
+res_blank_proto <- wilcox.test(WEIGHT ~ Prov, data = wt_blank_proto, exact = FALSE)
+dplyr::group_by(wt_blank_proto, Prov) |> 
+  summarise(count = n(), median = median(LENGTH, na.rm = TRUE))
+res_bl_proto <- wilcox.test(LENGTH ~ Prov, data = wt_blank_proto, exact = FALSE)
+dplyr::group_by(wt_blank_proto, Prov) |> 
+  summarise(count = n(), median = median(MAXWIDTH, na.rm = TRUE))
+res_bw_proto <- wilcox.test(MAXWIDTH ~ Prov, data = wt_blank_proto, exact = FALSE)
+
+```
+
 The sizes of lithic artefacts (flakes, elongated blanks, cores and retouched tools) are explored through scatterplots of width and length (@fig-gravettian-boxplots a) and boxplots for weight (@fig-gravettian-boxplots b); complete tables with descriptive statistics can also be found in SOM ++, table s++.

### 2025-02-26 analysis/Supplementary/supplementary_tables.qmd
Macbook morning work session. Changes to tables in main paper and supplementary.

@@
 library(knitr)
+library(janitor)
 
@@
 
-# Terrace dataset setup
-cols_to_concat_terrace <- c("UNIT", "ID")
+# Setting up the Terrace dataset
+cols_to_concat_terrace <- c("UNIT", "ID") # Vector to join UNIT and ID into a single variable
 context$ID <- str_squish(context$ID) # Remove empty spaces from ID field
@@
 
-proto_tech_WIP <- proto_tech %>%
-  tidyr::unite(col='ID', all_of(cols_to_concat_terrace), sep="-", remove=TRUE) %>% 
+# Cleaning up DS and uniformising variable names
+proto_tech_WIP <- proto_tech |>
+  tidyr::unite(col='ID', all_of(cols_to_concat_terrace), sep="-", remove=TRUE) |> 
   rename(CLASS = Class, COREPREPPROD = CorePreparProd, PIECECOMPLETENESS = PieceCompleteness, CORTEX = Cortex, CORTEXLOCATION = CortexLocation, PLATFORMTYPE = PlatformType, PLATFORMCORTEX = PlatformCortex, SCARCOUNT = ScarCount, THICKNESS = Thickness,
         MAXWIDTH = MaxWidth, LENGTH = Length, COREPLATFORM = CorePlatform, MAINFACECORTEX = MainFaceCortex, 
-        MAINFACESCARCOUNT = MainFaceScarCount, COREABANDON = CoreAbandon, RETOUCHEDPIECETYPOLOGY = RetouchedPieceTypology) %>% 
+        MAINFACESCARCOUNT = MainFaceScarCount, COREABANDON = CoreAbandon, RETOUCHEDPIECETYPOLOGY = RetouchedPieceTypology) |> 
   mutate(RETOUCHEDPIECETYPOLOGY = case_when(
@@
 
-proto_original_chert <- proto_tech %>% 
-  filter(RawMaterial %in% c("Chert", "Chalcedony"))
-
-gravettian_tech <- gravettian_tech %>% 
+gravettian_tech <- gravettian_tech |> 
   mutate(RETOUCHEDPIECETYPOLOGY = case_when(
@@
 
-terrace_tech <- bind_rows(proto_tech_WIP, gravettian_tech)
+terrace_tech <- bind_rows(proto_tech_WIP, gravettian_tech) # Creating single DS with Proto and Gravettian
 
-terrace_RM_WIP <- terrace_RM %>%
+terrace_RM_WIP <- terrace_RM |> 
   mutate(RMTYPE = case_when(
@@
     RMTYPE == "TYPE GB" ~ "T11",
-    .default = RMTYPE)) %>% 
+    .default = RMTYPE)) |> 
   select(AREA, ID, RMTYPE, SUBTYPE, WEIGHT, CORTEXTHICKNESS, CORTEXAPPEARANCE) %>%
@@
 
-context <- context %>%
-  dplyr::rename("UNIT" = "Unit") %>% # Renaming first column from Unit to UNIT to use cols_to_concat
-  unite(col='ID', cols_to_concat_terrace, sep="-", remove=FALSE) %>% # Unite UNIT with ID using cols_to_concat
-  select(ID, Level, Spit, Code) %>% # Selecting important variables: ID refers to artifact ID; Level refers to archaeologycal layer; Spit refers to artificial levels within an archaeological layer; Code refers to type of artifact (e.g., LITHIC or BONE)
+context <- context |>
+  dplyr::rename("UNIT" = "Unit") |> # Renaming first column from Unit to UNIT to use cols_to_concat
+  unite(col='ID', cols_to_concat_terrace, sep="-", remove=FALSE) |> # Unite UNIT with ID using cols_to_concat
+  select(ID, Level, Spit, Code) |> # Selecting important variables: ID refers to artifact ID; Level refers to archaeologycal layer; Spit refers to artificial levels within an archaeological layer; Code refers to type of artifact (e.g., LITHIC or BONE)
   distinct(ID, .keep_all = TRUE)
 
-xyz <- xyz %>%
-  unite(col='ID', cols_to_concat_terrace, sep="-", remove=FALSE) %>% # Unite UNIT with ID using cols_to_concat
-  filter(Suffix == 0) %>% # To avoid several entries related to several points of the same artefact
-  select(ID, X, Y, Z) %>% # Selecting important variables
+xyz <- xyz |>
+  unite(col='ID', cols_to_concat_terrace, sep="-", remove=FALSE) |> # Unite UNIT with ID using cols_to_concat
+  filter(Suffix == 0) |> # To avoid several entries related to several points of the same artefact
+  select(ID, X, Y, Z) |> # Selecting important variables
   distinct(ID, .keep_all = TRUE)
@@
   filter(Level %in% c(6, 7)) |>
-  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
-                                        "TYPE 5", "TYPE 6","TYPE 7", 
-                                        "TYPE 8", "TYPE 9", "TYPE 10", "TL", "INDET"))) |>
+  dplyr::mutate(RMTYPE = case_when(
+      RMTYPE == "T2" | RMTYPE == "T3" | RMTYPE == "T4" | RMTYPE == "T5" ~ "T2-5",
+      TRUE ~ RMTYPE)) |> 
+  mutate(RMTYPE = factor(RMTYPE, levels = c("T1","T2-5", "T6","T7", 
+                                        "T8", "T9", "T10", "TL", "INDET"))) |>
   tabyl(CLASS, RMTYPE) |> 
@@
   filter(CLASS != "Chip") |> 
-  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
-                                        "TYPE 5", "TYPE 6","TYPE 7", 
-                                        "TYPE 8", "TYPE 9", "TYPE 10", "TL", "INDET"))) |> 
+  dplyr::mutate(RMTYPE = case_when(
+      RMTYPE == "T2" | RMTYPE == "T3" | RMTYPE == "T4" | RMTYPE == "T5" ~ "T2-5",
+      TRUE ~ RMTYPE)) |> 
+  mutate(RMTYPE = factor(RMTYPE, levels = c("T1","T2-5", "T6","T7", 
+                                        "T8", "T9", "T10", "TL", "INDET"))) |>
   tabyl(CLASS, RMTYPE) |> 
@@
 solutrean_general <- solutrean_dataset |> 
-  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
-                                        "TYPE 5", "TYPE 6","TYPE 7", 
-                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL", "INDET"))) |> 
+  dplyr::mutate(RMTYPE = case_when(
+      RMTYPE == "T2" | RMTYPE == "T3" | RMTYPE == "T4" | RMTYPE == "T5" ~ "T2-5",
+      TRUE ~ RMTYPE)) |> 
+  mutate(RMTYPE = factor(RMTYPE, levels = c("T1","T2-5", "T6","T7", 
+                                        "T8", "T9", "T10", "T11", "TL", "INDET"))) |>
   tabyl(CLASS, RMTYPE) |> 
@@
   filter(CLASS %in% c("RetouchedPiece", "RetouchedPieceFrag")) |>
+  dplyr::mutate(RMTYPE = case_when(
+      RMTYPE == "T2" | RMTYPE == "T3" | RMTYPE == "T4" | RMTYPE == "T5" ~ "T2-5",
+      TRUE ~ RMTYPE)) |> 
   select(RMTYPE, RETOUCHEDPIECETYPOLOGY) |>
-  filter(RETOUCHEDPIECETYPOLOGY != "") |>
-  dplyr::mutate(RMTYPE = case_when(
-    RMTYPE == "TYPE 1" ~ "T1",
-    RMTYPE == "TYPE 2" ~ "T2",
-    RMTYPE == "TYPE 3" ~ "T3",
-    RMTYPE == "TYPE 4" ~ "T4",
-    RMTYPE == "TYPE 5" ~ "T5",
-    RMTYPE == "TYPE 6" ~ "T6",
-    RMTYPE == "TYPE 7" ~ "T7",
-    RMTYPE == "TYPE 8" ~ "T8",
-    RMTYPE == "TYPE 9" ~ "T9",
-    RMTYPE == "TYPE 10" ~ "T10",
-    RMTYPE == "TYPE 11" ~ "T11",
-    TRUE ~ RMTYPE))
+  filter(RETOUCHEDPIECETYPOLOGY != "")
+
 ## Janitor tabyl for final table with n and %
@@
   mutate(across(where(is.numeric), ~ ifelse(. == 0, " ", .))) |>
-  select(RETOUCHEDPIECETYPOLOGY, T1, T2, T5, T6, T7, T8, T11, INDET, Total) |> # Ordering values
+  select(RETOUCHEDPIECETYPOLOGY, T1, "T2-5", T6, T7, T8, T11, INDET, Total) |> # Ordering values
   rename(Typology = RETOUCHEDPIECETYPOLOGY)
@@
   filter(Level %in% c("4E", "4E/5")) |>
-  select(RMTYPE, RETOUCHEDPIECETYPOLOGY) |>
   dplyr::mutate(RMTYPE = case_when(
-    RMTYPE == "TYPE 1" ~ "T1",
-    RMTYPE == "TYPE 2" ~ "T2",
-    RMTYPE == "TYPE 3" ~ "T3",
-    RMTYPE == "TYPE 4" ~ "T4",
-    RMTYPE == "TYPE 5" ~ "T5",
-    RMTYPE == "TYPE 6" ~ "T6",
-    RMTYPE == "TYPE 7" ~ "T7",
-    RMTYPE == "TYPE 8" ~ "T8",
-    RMTYPE == "TYPE 9" ~ "T9",
-    RMTYPE == "TYPE 10" ~ "T10",
-    RMTYPE == "TYPE 11" ~ "T11",
-    TRUE ~ RMTYPE))
+      RMTYPE == "T2" | RMTYPE == "T3" | RMTYPE == "T4" | RMTYPE == "T5" ~ "T2-5",
+      TRUE ~ RMTYPE)) |> 
+  select(RMTYPE, RETOUCHEDPIECETYPOLOGY)
+
 ## Janitor tabyl for final table with n and %
@@
   mutate(across(where(is.numeric), ~ ifelse(. == 0, " ", .))) |>
-  select(RETOUCHEDPIECETYPOLOGY, T1, T2, T3, T4, T5, T6, T7, T8, T9, TL, INDET, Total) |> # Ordering values
+  select(RETOUCHEDPIECETYPOLOGY, T1, "T2-5", T6, T7, T8, T9, TL, INDET, Total) |> # Ordering values
   rename(Typology = RETOUCHEDPIECETYPOLOGY)
@@
   filter(Level %in% c(6, 7)) |> 
+  dplyr::mutate(RMTYPE = case_when(
+      RMTYPE == "T2" | RMTYPE == "T3" | RMTYPE == "T4" | RMTYPE == "T5" ~ "T2-5",
+      TRUE ~ RMTYPE)) |> 
   select(RMTYPE, RETOUCHEDPIECETYPOLOGY) |> 
-  dplyr::group_by(RMTYPE) |>
-  dplyr::mutate(RMTYPE = case_when(
-    RMTYPE == "TYPE 1" ~ "T1",
-    RMTYPE == "TYPE 2" ~ "T2",
-    RMTYPE == "TYPE 3" ~ "T3",
-    RMTYPE == "TYPE 4" ~ "T4",
-    RMTYPE == "TYPE 5" ~ "T5",
-    RMTYPE == "TYPE 6" ~ "T6",
-    RMTYPE == "TYPE 7" ~ "T7",
-    RMTYPE == "TYPE 8" ~ "T8",
-    RMTYPE == "TYPE 9" ~ "T9",
-    RMTYPE == "TYPE 10" ~ "T10",
-    TRUE ~ RMTYPE))
+  dplyr::group_by(RMTYPE)
 ## Janitor tabyl for final table with n and %
@@
   mutate(across(where(is.numeric), ~ ifelse(. == 0, " ", .))) |>
-  select(RETOUCHEDPIECETYPOLOGY, T1, T2, T3, T4, T5, T6, T7, T8, T9, T10, TL, INDET, Total) |> # Ordering values
+  select(RETOUCHEDPIECETYPOLOGY, T1, "T2-5", T6, T7, T8, T9, T10, TL, INDET, Total) |> # Ordering values
   rename(Typology = RETOUCHEDPIECETYPOLOGY)
@@
 
-proto5_ratiotd <- phase_dataset %>% 
-  filter(Level == 5) %>% 
-  filter(RMTYPE != "INDET") %>% 
-  mutate(CLASS = case_when(
-    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
-    CLASS == "RetouchedPieceFrag" & PIECECOMPLETENESS == "Proximal" ~ "RetouchedPiece",
-    TRUE ~ CLASS)) %>% 
-  filter(CLASS %in% c("Blank", "RetouchedPiece")) %>% 
-  count(RMTYPE, CLASS, sort = TRUE) %>% 
-  group_by(RMTYPE, CLASS) %>%
-  summarise(n=sum(n)) %>% 
-  spread(CLASS, n, fill=0) %>% 
-  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
-  filter(!sum <= 5) %>% 
-  mutate("Tool/Blank" = RetouchedPiece / Blank)
-
 # Gravettian
@@
 # DS preparation
-solutrean_stats <- solutrean_dataset %>% 
-  filter(RMTYPE != "INDET") %>%
+solutrean_stats <- solutrean_dataset |> 
+  filter(RMTYPE != "INDET") |> 
    mutate(CLASS = case_when(
-    CLASS == "Thinning flake" ~ "ThinningFlake",
-    CLASS == "ElongBlank" ~ "ElongatedProd",
-    CLASS == "FlakeFrag" | CLASS == "ElongBlankFrag" ~ "BlankFrag",
-    CLASS == "Retouched piece" ~ "RetouchedPiece",
-    TRUE ~ CLASS)) %>% 
-  mutate(CLASS = case_when(
-    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "BlankFragProx", # To add proximal fragments to ratio
-    TRUE ~ CLASS))
+    CLASS == "Thinning flake" | CLASS == "ElongBlank" | CLASS == "Flake" ~ "Blank",
+    TRUE ~ CLASS)) |> 
+    dplyr::mutate(RMTYPE = case_when(
+      RMTYPE == "T2" | RMTYPE == "T3" | RMTYPE == "T4" | RMTYPE == "T5" ~ "T2-5",
+      TRUE ~ RMTYPE))
 
@@
 # For flakes (complete)
-solutrean_stats_flakes <- solutrean_stats %>% 
-  filter(CLASS == "Flake") %>%
-  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
-  group_by(RMTYPE, CLASS) %>% 
-  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
-# For elongated blanks (complete)
-solutrean_stats_elong <- solutrean_stats %>% 
-  filter(CLASS == "ElongatedProd") %>%
-  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
-  group_by(RMTYPE, CLASS) %>% 
-  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
+solutrean_stats_blank <- solutrean_stats |>  
+  filter(CLASS == "Blank") |> 
+  select(RMTYPE, CLASS, MAXWIDTH, LENGTH, WEIGHT) |> 
+  group_by(RMTYPE, CLASS) |> 
+  summarise(across(c(MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
 # For cores
-solutrean_stats_cores <- solutrean_stats %>% 
-  filter(CLASS == "Core") %>%
-  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
-  group_by(RMTYPE, CLASS) %>% 
-  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
-# For retouched
-solutrean_stats_retouch<- solutrean_stats %>% 
-  filter(CLASS == "RetouchedPiece") %>%
-  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
-  group_by(RMTYPE, CLASS) %>% 
-  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
+solutrean_stats_cores <- solutrean_stats |> 
+  filter(CLASS == "Core") |> 
+  select(RMTYPE, CLASS, MAXWIDTH, LENGTH, WEIGHT) |>  
+  group_by(RMTYPE, CLASS) |> 
+  summarise(across(c(MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
 
@@
 # DS prep
-proto_stats <- phase_dataset %>% 
-  filter(Level %in% c("4E", "4E/5")) %>% 
-  filter(RMTYPE != "INDET") %>% 
-  mutate(CLASS = case_when(
-    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "BlankFragProx", # To add proximal fragments to ratio
-    TRUE ~ CLASS)) %>%
-  mutate(elong = MAXWIDTH*2-LENGTH) %>%
-  mutate(CLASS = case_when(
-    CLASS == "Blank" & elong > 0 ~ "Flake", # QUESTION: do I keep flakes and elong separate or join into blanks?
-    CLASS == "Blank" & elong <= 0 ~ "ElongatedProd",
-    TRUE ~ CLASS))
+proto_stats <- phase_dataset |> 
+  filter(Level %in% c("4E", "4E/5")) |> 
+  filter(RMTYPE != "INDET") |> 
+  dplyr::mutate(RMTYPE = case_when(
+      RMTYPE == "T2" | RMTYPE == "T3" | RMTYPE == "T4" | RMTYPE == "T5" ~ "T2-5",
+      TRUE ~ RMTYPE))
 
@@
 # For flakes (complete)
-proto_stats_flakes <- proto_stats %>% 
-  filter(CLASS == "Flake") %>%
-  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
-  group_by(RMTYPE, CLASS) %>% 
-  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
-# For elongated blanks (complete)
-proto_stats_elong <- proto_stats %>% 
-  filter(CLASS == "ElongatedProd") %>% 
-  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
-  group_by(RMTYPE, CLASS) %>% 
-  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
+proto_stats_blank <- proto_stats |> 
+  filter(CLASS == "Blank") |> 
+  select(RMTYPE, CLASS, MAXWIDTH, LENGTH, WEIGHT) |> 
+  group_by(RMTYPE, CLASS) |> 
+  summarise(across(c(MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
 # For cores
-proto_stats_cores <- proto_stats %>% 
-  filter(CLASS == "Core") %>%
-  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
-  group_by(RMTYPE, CLASS) %>% 
-  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
-# For retouched
-proto_stats_retouch <- proto_stats %>% 
-  filter(CLASS == "RetouchedPiece") %>%
-  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
-  group_by(RMTYPE, CLASS) %>% 
-  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
+proto_stats_cores <- proto_stats |> 
+  filter(CLASS == "Core") |> 
+  select(RMTYPE, CLASS, MAXWIDTH, LENGTH, WEIGHT) |> 
+  group_by(RMTYPE, CLASS) |> 
+  summarise(across(c(MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
 
@@
 # DS prep
-gravettian_stats <- phase_dataset %>%
-  filter(Level %in% c(6, 7)) %>%
-  filter(RMTYPE != "INDET") %>%
-  mutate(CLASS = case_when(
-    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "BlankFragProx", # To add proximal fragments to ratio
-    TRUE ~ CLASS)) %>% 
-  mutate(elong = MAXWIDTH*2-LENGTH) %>%
-  mutate(CLASS = case_when(
-    CLASS == "Blank" & elong > 0 ~ "Flake", # QUESTION: do I keep flakes and elong separate or join into blanks?
-    CLASS == "Blank" & elong <= 0 ~ "ElongatedProd",
-    TRUE ~ CLASS))
+gravettian_stats <- phase_dataset |> 
+  filter(Level %in% c(6, 7)) |> 
+  filter(RMTYPE != "INDET")
 
@@
 # For flakes (complete)
-gravettian_stats_flakes <- gravettian_stats %>% 
-  filter(CLASS == "Flake") %>%
-  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
-  group_by(RMTYPE, CLASS) %>% 
-  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
-# For elongated blanks (complete)
-gravettian_stats_elong <- gravettian_stats %>% 
-  filter(CLASS == "ElongatedProd") %>%
-  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
-  group_by(RMTYPE, CLASS) %>% 
-  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
+gravettian_stats_blank <- gravettian_stats |> 
+  filter(CLASS == "Blank") |> 
+  select(RMTYPE, CLASS, MAXWIDTH, LENGTH, WEIGHT) |> 
+  group_by(RMTYPE, CLASS) |> 
+  summarise(across(c(MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
 # For cores
-gravettian_stats_cores <- gravettian_stats %>% 
-  filter(CLASS == "Core") %>%
-  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
-  group_by(RMTYPE, CLASS) %>% 
-  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
-# For retouched
-gravettian_stats_retouch <- gravettian_stats %>% 
-  filter(CLASS == "RetouchedPiece") %>%
-  select(RMTYPE, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT) %>% 
-  group_by(RMTYPE, CLASS) %>% 
-  summarise(across(c(THICKNESS, MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
+gravettian_stats_cores <- gravettian_stats |> 
+  filter(CLASS == "Core") |> 
+  select(RMTYPE, CLASS, MAXWIDTH, LENGTH, WEIGHT) |> 
+  group_by(RMTYPE, CLASS) |> 
+  summarise(across(c(MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
 
@@
 
-kable(head(solutrean_stats_flakes))
-kable(head(solutrean_stats_elong))
+kable(head(solutrean_stats_blank))
 kable(head(solutrean_stats_cores))
-kable(head(solutrean_stats_retouch))
 
@@
 
-kable(head(proto_stats_flakes))
-kable(head(proto_stats_elong))
+kable(head(proto_stats_blank))
 kable(head(proto_stats_cores))
-kable(head(proto_stats_retouch))
 
@@
 
-kable(head(gravettian_stats_flakes))
-kable(head(gravettian_stats_elong))
+kable(head(gravettian_stats_blank))
 kable(head(gravettian_stats_cores))
-kable(head(gravettian_stats_retouch))
 
 ```
+
+# Statistical tests
+
+A Mann-Whitney U was conducted to determine whether there are differences in the measurements (weight, length and width) of blanks and cores between local and non-local cherts.
+To do this, all local cherts (including chalcedony) were grouped in a Local category (L) while all non-local cherts were grouped in a non-local category (NL).
+Type 10 and TL cherts were excluded from the NL grouping to avoid adding cherts which are the result of probable different procurement strategies and clearly deviate from the other cherts in terms regarding the measurements.
+The Solutrean assemblage is not represented in the core statistics due to the small sample size of both local and non-local categories.
+
+The results indicate generally significant differences in blank measurements between local and non-local cherts in all technocomplexes; in almost all measurements and groups, local cherts show higher medians.
+Length seems to be consistently significant (p < 0.01 and < 0.001) between the two groups in all assemblages, with local cherts showing higher median lengths than non-local cherts.
+In the Gravettian and Proto-Solutrean assemblages, weight is also significantly different between local and non-local blanks (p < 0.05), with local cherts presenting higher medians.
+In comparison, width differences show no statistical significance, with exception of the Solutrean assemblage.
+In the Solutrean, blank width is significant (p < 0.05) with the non-local category showing higher medians compared local cherts.
+
+Regarding the cores, in almost all measurements and groups local cherts show higher means than non-local cherts.
+Despite these differences, the results are not significant  (p > 0.05).
+It is important to note, however, that the sample size for cores in the Gravettian assemblage and non-local cherts in the Solutrean assemblage are smaller (n < 15) than the blank sample size.
 

### 2025-02-26 analysis/paper/paper.qmd
Macbook morning work session. Changes to tables in main paper and supplementary.

--- 2025-02-24 analysis/paper/paper.qmd
@@
   dplyr::mutate(RETOUCHEDPIECETYPOLOGY = case_when(
-    RETOUCHEDPIECETYPOLOGY == "Retouched blade" | RETOUCHEDPIECETYPOLOGY == "Retouched bladelet" | RETOUCHEDPIECETYPOLOGY == "Retouched bladelet fragment" | RETOUCHEDPIECETYPOLOGY == "Retouched flake" | RETOUCHEDPIECETYPOLOGY == "Retouched flake fragment" ~ "Retouched blank",
+    RETOUCHEDPIECETYPOLOGY == "Retouched blade" | RETOUCHEDPIECETYPOLOGY == "Retouched bladelet" | RETOUCHEDPIECETYPOLOGY == "Retouched bladelet fragment" | RETOUCHEDPIECETYPOLOGY == "Retouched flake" | RETOUCHEDPIECETYPOLOGY == "Retouched flake fragment" | RETOUCHEDPIECETYPOLOGY == "Retouched burin spall" ~ "Retouched blank",
     RETOUCHEDPIECETYPOLOGY == "Splintered piece fragment" ~ "Splintered piece",
+    RETOUCHEDPIECETYPOLOGY == "Backed bladelet" | RETOUCHEDPIECETYPOLOGY == "Backed flake" ~ "Backed tool",
+    RETOUCHEDPIECETYPOLOGY == "Retouched burin" ~ "Burin",
     TRUE ~ RETOUCHEDPIECETYPOLOGY))
@@
   filter(CLASS %in% c("Retouched", "Retouched fragment")) |>
+  dplyr::mutate(RETOUCHEDPIECETYPOLOGY = case_when(
+    RETOUCHEDPIECETYPOLOGY == "Bifacial fragment" ~ "Bifacial blank",
+    RETOUCHEDPIECETYPOLOGY == "Shouldered point fragment" ~ "Shouldered point",
+    TRUE ~ RETOUCHEDPIECETYPOLOGY))
   filter(RMTYPE != "INDET") |> 
@@
 
+```{r}
+#| label: stats-wilcox
+
+# Terrace wilcox statistics
+## DB prep
+wilcox_t <- phase_dataset |> 
+  dplyr::mutate(Prov = case_when(
+    RMTYPE == "TYPE 1" | RMTYPE == "TYPE 2" | RMTYPE == "TYPE 3" | RMTYPE == "TYPE 4" | RMTYPE == "TYPE 5" ~ "Local",
+    RMTYPE == "TYPE 6" | RMTYPE == "TYPE 7" | RMTYPE == "TYPE 8" | RMTYPE == "TYPE 9" ~ "Non-local",
+    TRUE ~ RMTYPE)) |> 
+  filter(Prov %in% c("Local", "Non-local")) |> 
+  select(ID, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT, Prov, Level)
+
+# Proto-Solutrean blanks statistical test
+## NL cores were not represented enough (n) for statistics
+wt_blank_proto <- wilcox_t |> 
+  filter(Level %in% c("4E", "4E/5")) |> 
+  filter(CLASS == "Blank")
+
+## Blanks weight (bw)
+dplyr::group_by(wt_blank_proto, Prov) |> 
+  summarise(count = n(), median = median(WEIGHT, na.rm = TRUE))
+res_bw_proto <- wilcox.test(WEIGHT ~ Prov, data = wt_blank_proto, exact = FALSE)
+## Blanks length (bl)
+dplyr::group_by(wt_blank_proto, Prov) |> 
+  summarise(count = n(), median = median(LENGTH, na.rm = TRUE))
+res_bl_proto <- wilcox.test(LENGTH ~ Prov, data = wt_blank_proto, exact = FALSE)
+## Blanks width (bw)
+dplyr::group_by(wt_blank_proto, Prov) |> 
+  summarise(count = n(), median = median(MAXWIDTH, na.rm = TRUE))
+res_bwi_proto <- wilcox.test(MAXWIDTH ~ Prov, data = wt_blank_proto, exact = FALSE)
+
+# Gravettian blanks statistical test
+wt_blank_grav <- wilcox_t |> 
+  filter(Level %in% c("6", "7")) |> 
+  filter(CLASS == "Blank")
+
+## Blanks weight
+dplyr::group_by(wt_blank_grav, Prov) |> 
+  summarise(count = n(), median = median(WEIGHT, na.rm = TRUE))
+res_bw_grav <- wilcox.test(WEIGHT ~ Prov, data = wt_blank_grav, exact = FALSE)
+## Blanks length
+dplyr::group_by(wt_blank_grav, Prov) |> 
+  summarise(count = n(), median = median(LENGTH, na.rm = TRUE))
+res_bl_grav <- wilcox.test(LENGTH ~ Prov, data = wt_blank_grav, exact = FALSE)
+## Blanks width
+dplyr::group_by(wt_blank_grav, Prov) |> 
+  summarise(count = n(), median = median(MAXWIDTH, na.rm = TRUE))
+res_bwi_grav <- wilcox.test(MAXWIDTH ~ Prov, data = wt_blank_grav, exact = FALSE)
+
+## Gravettian cores statistical test
+wt_core_grav <- wilcox_t |> 
+  filter(Level %in% c("6", "7")) |> 
+  filter(CLASS == "Core")
+
+## Cores weight
+dplyr::group_by(wt_core_grav, Prov) |> 
+  summarise(count = n(), median = median(WEIGHT, na.rm = TRUE))
+res_cw_grav <- wilcox.test(WEIGHT ~ Prov, data = wt_core_grav, exact = FALSE)
+## Cores length
+dplyr::group_by(wt_core_grav, Prov) |> 
+  summarise(count = n(), median = median(LENGTH, na.rm = TRUE))
+res_cl_grav <- wilcox.test(LENGTH ~ Prov, data = wt_core_grav, exact = FALSE)
+## Cores width
+dplyr::group_by(wt_core_grav, Prov) |> 
+  summarise(count = n(), median = median(MAXWIDTH, na.rm = TRUE))
+res_cwi_grav <- wilcox.test(MAXWIDTH ~ Prov, data = wt_core_grav, exact = FALSE)
+
+# Shelter wilcox statistics
+## DB prep
+wilcox_t_shelter <- solutrean_dataset |> 
+  dplyr::mutate(Prov = case_when(
+    RMTYPE == "TYPE 1" | RMTYPE == "TYPE 2" | RMTYPE == "TYPE 3" | RMTYPE == "TYPE 4" | RMTYPE == "TYPE 5" ~ "Local",
+    RMTYPE == "TYPE 6" | RMTYPE == "TYPE 7" | RMTYPE == "TYPE 8" | RMTYPE == "TYPE 9" | RMTYPE == "TYPE 11" ~ "Non-local",
+    TRUE ~ RMTYPE)) |> 
+   mutate(CLASS = case_when(
+    CLASS == "Flake" | CLASS == "ElongBlank" | CLASS == "Thinning flake" ~ "Blank",
+    TRUE ~ CLASS)) |>
+  filter(Prov %in% c("Local", "Non-local")) |> 
+  select(ID, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT, Prov, LEVEL)
+
+# Solutrean blanks statistical test
+wt_blank_s <- wilcox_t_shelter |>
+  filter(CLASS == "Blank")
+
+## Blanks weight
+dplyr::group_by(wt_blank_s, Prov) |> 
+  summarise(count = n(), median = median(WEIGHT, na.rm = TRUE))
+res_bw_s <- wilcox.test(WEIGHT ~ Prov, data = wt_blank_s, exact = FALSE)
+## Blanks length
+dplyr::group_by(wt_blank_s, Prov) |> 
+  summarise(count = n(), median = median(LENGTH, na.rm = TRUE))
+res_bl_s <- wilcox.test(LENGTH ~ Prov, data = wt_blank_s, exact = FALSE)
+## Blanks width
+dplyr::group_by(wt_blank_s, Prov) |> 
+  summarise(count = n(), median = median(MAXWIDTH, na.rm = TRUE))
+res_bwi_s <- wilcox.test(MAXWIDTH ~ Prov, data = wt_blank_s, exact = FALSE)
+
+wt_core_s <- wilcox_t_shelter |>
+  filter(CLASS == "Core")
+
+## Cores weight
+dplyr::group_by(wt_core_s, Prov) |> 
+  summarise(count = n(), median = median(WEIGHT, na.rm = TRUE))
+res_cw_s <- wilcox.test(WEIGHT ~ Prov, data = wt_core_s, exact = FALSE)
+## Cores length
+dplyr::group_by(wt_core_s, Prov) |> 
+  summarise(count = n(), median = median(LENGTH, na.rm = TRUE))
+res_cl_s <- wilcox.test(LENGTH ~ Prov, data = wt_core_s, exact = FALSE)
+## Cores width
+dplyr::group_by(wt_core_s, Prov) |> 
+  summarise(count = n(), median = median(MAXWIDTH, na.rm = TRUE))
+res_cwi_s <- wilcox.test(MAXWIDTH ~ Prov, data = wt_core_s, exact = FALSE)
+
+p.value_blanks <- tibble(
+  Technocomplex = c("Gravettian", " "," ", " ", "Proto-Solutrean", " ", " ", " ", "Solutrean", " ", " ", " "),
+  Variable = c("Weight", "Length", "Width", "(n)","Weight", "Length", "Width","(n)", "Weight", "Length", "Width", "(n)"),
+  L_median = c("2.8", "25.8", "19.1", "82", "2.2", "23.4", "16.8", "154", "1.7", "25.1", "16.6", "342"),
+  NL_median = c("1.9", "21.2", "17.8", "69", "1.7", "19.4", "16.6", "49", "1.5", "21.9", "18.8", "86"),
+  P_Value = round(c(res_bw_grav$p.value, res_bl_grav$p.value, res_bwi_grav$p.value, 0, res_bw_proto$p.value, res_bl_proto$p.value, res_bwi_proto$p.value, 0, res_bw_s$p.value, res_bl_s$p.value, res_bwi_s$p.value, 0), 5)) |>   mutate(Significance = case_when(
+      P_Value <= 0.001 ~ "***",
+      P_Value <= 0.01  ~ "**",
+      P_Value <= 0.05  ~ "*",
+      TRUE             ~ ""),
+    P_value = paste0(format(P_Value, nsmall = 4), Significance)  # Keep decimal places
+  ) |> 
+  select(-P_Value, -Significance) |> 
+  dplyr::mutate(P_value = case_when(
+    P_value == "0.00000***" ~ "-",
+    TRUE ~ P_value))
+
+p.value_cores <- tibble(
+  Technocomplex = c("Gravettian", "", "", "", "Solutrean", "", "", ""),
+  Variable = c("Weight", "Length", "Width", "(n)", "Weight", "Length", "Width", "(n)"),
+  L_median = c("20", "29.6", "27.6", "14", "16.5", "28.3", "30.7", "59"),
+  NL_median = c("11.8", "21.9", "29.7", "10", "11.7", "25.6", "26.0", "7"),
+  P_Value = round(c(res_cw_grav$p.value, res_cl_grav$p.value, res_cwi_grav$p.value, 0, res_cw_s$p.value, res_cl_s$p.value, res_cwi_s$p.value, 0), 3)) |> 
+  mutate(Significance = case_when(
+      P_Value <= 0.001 ~ "***",
+      P_Value <= 0.01  ~ "**",
+      P_Value <= 0.05  ~ "*",
+      TRUE             ~ ""),
+    P_value = paste0(format(P_Value, nsmall = 3), Significance)  # Keep decimal places
+  ) |> 
+  select(-P_Value, -Significance) |> 
+  dplyr::mutate(P_value = case_when(
+    P_value == "0.000***" ~ "-",
+    TRUE ~ P_value))
+
+```
+
 The sizes of lithic artefacts (flakes, elongated blanks, cores and retouched tools) are explored through scatterplots of width and length (@fig-gravettian-boxplots a) and boxplots for weight (@fig-gravettian-boxplots b); complete tables with descriptive statistics can also be found in SOM ++, table s++.

### 2025-02-26 analysis/Supplementary/supplementary_tables.qmd
PC afternoon work session. Improved codes, trying graphs instead of tables for main text

@@
 
-# Technological data
-gravettian_tech <- read.csv(here::here('analysis/data/raw_data/VBT_6_7_JB.csv'))
-proto_tech <- read.csv(here::here('analysis/data/raw_data/VBT_4E_5_JB.csv'))
-solutrean_JC <- read.csv(here::here('analysis/data/raw_data/VBR_ABC_JC.csv'))
-solutrean_points <- read.csv(here::here('analysis/data/raw_data/shelter_points_JC.csv'))
+# Reading the datasets
+## Technological data
+gravettian_tech <- read_csv(here::here('analysis/data/raw_data/VBT_6_7_JB.csv'))
+proto_tech <- read_csv(here::here('analysis/data/raw_data/VBT_4E_5_JB.csv'))
+solutrean_JC <- read_csv(here::here('analysis/data/raw_data/VBR_ABC_JC.csv'))
+solutrean_points <- read_csv(here::here('analysis/data/raw_data/shelter_points_JC.csv'))
 
-# RM data
+## RM data
 shelter_RM <- read_csv(here::here('analysis/data/raw_data/MACRO_IND_SHELTER.csv')) # Shelter RM analysis
@@
         MAINFACESCARCOUNT = MainFaceScarCount, COREABANDON = CoreAbandon, RETOUCHEDPIECETYPOLOGY = RetouchedPieceTypology) |> 
-  mutate(RETOUCHEDPIECETYPOLOGY = case_when(
-    RETOUCHEDPIECETYPOLOGY == "Right truncation on retouched piece" ~ "Truncation",
-    RETOUCHEDPIECETYPOLOGY == "Double backed bladelet" | RETOUCHEDPIECETYPOLOGY == "Backed bladelet parcial" ~ "Backed bladelet",
-    RETOUCHEDPIECETYPOLOGY == "Dentilhado" ~ "Denticulate",
-    RETOUCHEDPIECETYPOLOGY == "Burin on truncation" ~ "Burin",
-    TRUE ~ RETOUCHEDPIECETYPOLOGY)) |> 
   dplyr::select(ID, CLASS, COREPREPPROD, PIECECOMPLETENESS, CORTEX, CORTEXLOCATION, PLATFORMTYPE, PLATFORMCORTEX,
@@
 gravettian_tech <- gravettian_tech |> 
-  mutate(RETOUCHEDPIECETYPOLOGY = case_when(
-    RETOUCHEDPIECETYPOLOGY == "Splintered piece with notch" | RETOUCHEDPIECETYPOLOGY == "Splintered piece with lateral retouch" ~ "Splintered piece",
-    RETOUCHEDPIECETYPOLOGY == "Retouched flake siret fragment" | RETOUCHEDPIECETYPOLOGY == "Retouched flake frag" ~ "Retouched flake fragment",
-    RETOUCHEDPIECETYPOLOGY == "Retouched bladelet frag" ~ "Retouched bladelet fragment",
-    RETOUCHEDPIECETYPOLOGY == "Multiple burin" | RETOUCHEDPIECETYPOLOGY == "Burin splintered piece" | RETOUCHEDPIECETYPOLOGY == "Burin on truncation" ~ "Burin",
-    RETOUCHEDPIECETYPOLOGY == "Double backed bladelet" ~ "Backed bladelet",
-    TRUE ~ RETOUCHEDPIECETYPOLOGY)) |> 
   mutate(CLASS = case_when(
@@
     CLASS == "FlakeFrag" | CLASS == "ElongBlankFrag" ~ "BlankFrag",
-    TRUE ~ CLASS))
+    TRUE ~ CLASS)) |> 
+  dplyr::mutate(RETOUCHEDPIECETYPOLOGY = case_when( # Solving typing issue
+    RETOUCHEDPIECETYPOLOGY == "Carenated endscrapper" ~ "Carenated endscraper",
+    RETOUCHEDPIECETYPOLOGY == "Endscrapper" ~ "Endscraper",
+    RETOUCHEDPIECETYPOLOGY == "Endscrapper-burin" ~ "Burin",
+    TRUE ~ RETOUCHEDPIECETYPOLOGY))
 
@@
   mutate(RMTYPE = case_when(
-    RMTYPE == "Type 2" ~ "T2",
+    RMTYPE == "TYPE 2" ~ "T2",
     RMTYPE == "Chalcedony" ~ "T1", # Altering first group names to final group names. Subtypes are maintained as before.
@@
     Level == "4E" | Level == "4E/5" | Level == 5 ~ "Proto-Solutrean",
-    Level == 6 | Level == 7 ~ "Gravettian"))
+    Level == 6 | Level == 7 ~ "Gravettian")) |> 
+  mutate(RETOUCHEDPIECETYPOLOGY = case_when(
+    RETOUCHEDPIECETYPOLOGY == "Splintered piece with notch" | RETOUCHEDPIECETYPOLOGY == "Splintered piece with lateral retouch" | RETOUCHEDPIECETYPOLOGY == "Splintered piece fragment" ~ "Splintered piece",
+    RETOUCHEDPIECETYPOLOGY == "Retouched blade" | RETOUCHEDPIECETYPOLOGY == "Retouched bladelet" | RETOUCHEDPIECETYPOLOGY == "Retouched flake" | RETOUCHEDPIECETYPOLOGY == "Retouched burin spall" ~ "Retouched blank",
+    RETOUCHEDPIECETYPOLOGY == "Retouched bladelet frag" | RETOUCHEDPIECETYPOLOGY == "Retouched flake frag" | RETOUCHEDPIECETYPOLOGY == "Retouched flake siret fragment"  ~ "Retouched blank fragment",
+    RETOUCHEDPIECETYPOLOGY == "Multiple burin" | RETOUCHEDPIECETYPOLOGY == "Burin splintered piece" | RETOUCHEDPIECETYPOLOGY == "Burin on truncation" | RETOUCHEDPIECETYPOLOGY == "Retouched burin" ~ "Burin",
+    RETOUCHEDPIECETYPOLOGY == "Dentilhado" ~ "Denticulate",
+    RETOUCHEDPIECETYPOLOGY == "Right truncation on retouched piece" ~ "Truncation",
+    RETOUCHEDPIECETYPOLOGY == "Backed bladelet" | RETOUCHEDPIECETYPOLOGY == "Backed flake" | RETOUCHEDPIECETYPOLOGY == "Double backed bladelet" | RETOUCHEDPIECETYPOLOGY == "Backed bladelet parcial" ~ "Backed tool",
+    RETOUCHEDPIECETYPOLOGY == "Carenated endscraper" ~ "Endscraper",
+    TRUE ~ RETOUCHEDPIECETYPOLOGY))
 
-# Shelter dataset setup
+# Setting up the Shelter dataset
 cols_to_concat_shelter <- c("YEAR", "ID") # Concatenate YEAR and ID variables in one single variable (e.g. 05-500)
 
-solutrean_WIP <- solutrean_JC %>% 
-  filter(Layer %in% c("A", "B", "C")) %>% # Filtering Solutrean layers
-  filter(Raw.material %in% c("Chert", "Chalcedony")) %>% # Filtering raw materials
-  filter(Screen != "Crivo") %>%  # Filtering out all bucket lithics since they were not included in the RM analysis
-  mutate_at("Date", str_replace, "20", "") %>% # Removing initial part of the year to match IDs
-  dplyr::rename("YEAR" = "Date") %>% 
+solutrean_WIP <- solutrean_JC |> 
+  filter(Layer %in% c("A", "B", "C")) |> # Filtering Solutrean layers
+  filter(`Raw material` %in% c("Chert", "Chalcedony")) |>  # Filtering raw materials
+  filter(!Screen %in% c("Crivo", "crivo")) |>  # Filtering out all bucket lithics since they were not included in the RM analysis
+  mutate_at("Date", str_replace, "20", "") |> # Removing initial part of the year to match IDs
+  dplyr::rename("YEAR" = "Date") |> 
   tidyr::unite(col='ID', all_of(cols_to_concat_shelter), sep="-", remove=FALSE) # Concatenate YEAR and ID variables in one single variable (e.g. 05-500)
@@
 # Dataset filtering for chert/chalcedony and levels (A-C), 3298 + 61 solutrean points, n=3359
-solutrean_original_chert <- solutrean_JC %>% 
-  filter(Layer %in% c("A", "B", "C")) %>% # Filtering Solutrean layers
-  filter(Raw.material %in% c("Chert", "Chalcedony"))
+solutrean_original_chert <- solutrean_JC |> 
+  filter(Layer %in% c("A", "B", "C")) |> # Filtering Solutrean layers
+  filter(`Raw material` %in% c("Chert", "Chalcedony"))
 
 # In Cascalheira 2010 Retouched piece was not a category but a variable. To match the other DBs, this category was added
-solutrean_WIP <- solutrean_WIP %>% 
+solutrean_WIP <- solutrean_WIP |> 
   mutate(Class = case_when( # This alteration is being done on the original class column
-    Retouched == "Yes" ~ "Retouched piece", # Whenever retouch is present, class is now Retouched piece
-    TRUE ~ Class)) %>% 
+    Retouched == "Yes" ~ "RetouchedPiece", # Whenever retouch is present, class is now Retouched piece
+    TRUE ~ Class)) |> 
   mutate(CLASS = case_when( # Creating a column for clustered classes, to match the other DBs
@@
 
-solutrean_points <- solutrean_points %>% # Creating ID from 
-  mutate_at("YEAR", str_replace, "20", "") %>% # Removing initial part of the year to match IDs
+solutrean_points <- solutrean_points |> 
+  mutate_at("YEAR", str_replace, "20", "") |> # Removing initial part of the year to match IDs
   tidyr::unite(col='ID', all_of(cols_to_concat_shelter), sep="-", remove=TRUE)
@@
 
-solutrean_tech <- solutrean_tech %>% # Originally 2228
+solutrean_tech <- solutrean_tech |> # Originally 2228
   distinct(ID, .keep_all = TRUE) # Keeping individual IDs to avoid buckets (down to 1993)
 
-shelter_RM_WIP <- shelter_RM %>%
-  mutate_at("ID", str_replace, "VB", "") %>% # Remove "VB" from ID
+shelter_RM_WIP <- shelter_RM |>
+  mutate_at("ID", str_replace, "VB", "") |> # Remove "VB" from ID
   mutate(RMTYPE = case_when(
-    RMTYPE == "Type 2" ~ "T2",
+    RMTYPE == "Type 2" | RMTYPE == "TYPE 2" ~ "T2",
     RMTYPE == "Chalcedony" ~ "T1", # Altering first group names to final group names. Subtypes are maintained as before.
@@
 solutrean_dataset <- full_join(solutrean_tech, shelter_RM_WIP, by = "ID") # The complete dataset has 2818 (44 points) entries
-solutrean_dataset <- solutrean_dataset %>%
-  drop_na(RMTYPE) %>%   # Dropping NAs on the RM database drops dataset to 1514
-  drop_na(CLASS) # Dropping NAs on the technological class drops dataset to 914
+solutrean_dataset <- solutrean_dataset |>
+  drop_na(RMTYPE) |>   # Dropping NAs on the RM database drops dataset to 1514
+  drop_na(CLASS) |>  # Dropping NAs on the technological class drops dataset to 914
+  dplyr::mutate(RETOUCHEDPIECETYPOLOGY = case_when(
+    RETOUCHEDPIECETYPOLOGY == "Bifacial fragment" ~ "Bifacial blank",
+    RETOUCHEDPIECETYPOLOGY == "Shouldered point fragment" ~ "Shouldered point",
+    TRUE ~ RETOUCHEDPIECETYPOLOGY))
 
@@
 # Tables with class by raw material type (n and %)
-## For Gravettian
-gravettian_general <- phase_dataset |>
-  filter(Level %in% c(6, 7)) |>
+general_table_som <- phase_dataset |> 
+  dplyr::mutate(CLASS = case_when(
+    CLASS == "BlankFrag" ~ "Blank fragment",
+    CLASS == "BurinSpall" ~ "Burin spall",
+    CLASS == "CoreFrag" ~ "Core fragment",
+    CLASS == "CorePreparProd" ~ "Core prep product",
+    CLASS == "RetouchedPiece" ~ "Retouched",
+    CLASS == "RetouchedPieceFrag" ~ "Retouched fragment",
+    TRUE ~ CLASS)) |> 
   dplyr::mutate(RMTYPE = case_when(
@@
       TRUE ~ RMTYPE)) |> 
+  filter(CLASS != "Chip")
+
+## For Gravettian
+gravettian_general_som <- general_table_som |>
+  filter(Level %in% c(6, 7)) |>
   mutate(RMTYPE = factor(RMTYPE, levels = c("T1","T2-5", "T6","T7", 
@@
 ## For Proto-Solutrean (4E and top of 5)
-proto_general <- phase_dataset |> 
+proto_general_som <- general_table_som |> 
   filter(Level %in% c("4E", "4E/5")) |> 
-  filter(CLASS != "Chip") |> 
-  dplyr::mutate(RMTYPE = case_when(
-      RMTYPE == "T2" | RMTYPE == "T3" | RMTYPE == "T4" | RMTYPE == "T5" ~ "T2-5",
-      TRUE ~ RMTYPE)) |> 
   mutate(RMTYPE = factor(RMTYPE, levels = c("T1","T2-5", "T6","T7", 
@@
 ## For Solutrean
-solutrean_general <- solutrean_dataset |> 
+solutrean_general_som <- solutrean_dataset |> 
   dplyr::mutate(RMTYPE = case_when(
@@
       TRUE ~ RMTYPE)) |>
+  dplyr::mutate(CLASS = case_when(
+    CLASS == "ElongBlank" | CLASS == "Flake" ~ "Blank",
+    CLASS == "ElongBlankFrag" | CLASS == "FlakeFrag" ~ "Blank fragment",
+    CLASS == "Thinning flake" ~ "Blank",
+    CLASS == "CorePrepProd" ~ "Core prep product",
+    CLASS == "RetouchedPiece" | CLASS == "Retouched piece" ~ "Retouched",
+    CLASS == "Retouched piece fragment" ~ "Retouched fragment",
+    TRUE ~ CLASS)) |> 
   mutate(RMTYPE = factor(RMTYPE, levels = c("T1","T2-5", "T6","T7", 
@@
 
-kable(gravettian_general)
+kable(gravettian_general_som)
 
@@
 
-kable(proto_general)
+kable(proto_general_som)
 
@@
 
-kable(solutrean_general)
+kable(solutrean_general_som)
 
@@
 #| tbl-subcap: 
-#|  - "Stats for flakes"
-#|  - "Stats for elongated blanks"
+#|  - "Stats for blanks"
 #|  - "Stats for cores"
-#|  - "Stats for retouched"
 #| layout-nrow: 4
@@
 #| tbl-subcap: 
-#|  - "Stats for flakes"
-#|  - "Stats for elongated blanks"
+#|  - "Stats for blanks"
 #|  - "Stats for cores"
-#|  - "Stats for retouched"
-#| layout-nrow: 4
+#| layout-nrow: 2
 
@@
 # Statistical tests
+
+```{r}
+#| label: stats-wilcox
+#| include: false
+
+# Terrace wilcox statistics
+## DB prep
+wilcox_t <- phase_dataset |> 
+  dplyr::mutate(Prov = case_when(
+    RMTYPE == "T1" | RMTYPE == "T2" | RMTYPE == "T3" | RMTYPE == "T4" | RMTYPE == "T5" ~ "Local",
+    RMTYPE == "T6" | RMTYPE == "T7" | RMTYPE == "T8" | RMTYPE == "T9" ~ "Non-local",
+    TRUE ~ RMTYPE)) |> 
+  filter(Prov %in% c("Local", "Non-local")) |> 
+  select(ID, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT, Prov, Level)
+
+# Proto-Solutrean blanks statistical test
+## NL cores were not represented enough (n) for statistics
+wt_blank_proto <- wilcox_t |> 
+  filter(Level %in% c("4E", "4E/5")) |> 
+  filter(CLASS == "Blank")
+
+## Blanks weight (bw)
+dplyr::group_by(wt_blank_proto, Prov) |> 
+  summarise(count = n(), median = median(WEIGHT, na.rm = TRUE))
+res_bw_proto <- wilcox.test(WEIGHT ~ Prov, data = wt_blank_proto, exact = FALSE)
+## Blanks length (bl)
+dplyr::group_by(wt_blank_proto, Prov) |> 
+  summarise(count = n(), median = median(LENGTH, na.rm = TRUE))
+res_bl_proto <- wilcox.test(LENGTH ~ Prov, data = wt_blank_proto, exact = FALSE)
+## Blanks width (bw)
+dplyr::group_by(wt_blank_proto, Prov) |> 
+  summarise(count = n(), median = median(MAXWIDTH, na.rm = TRUE))
+res_bwi_proto <- wilcox.test(MAXWIDTH ~ Prov, data = wt_blank_proto, exact = FALSE)
+
+# Gravettian blanks statistical test
+wt_blank_grav <- wilcox_t |> 
+  filter(Level %in% c("6", "7")) |> 
+  filter(CLASS == "Blank")
+
+## Blanks weight
+dplyr::group_by(wt_blank_grav, Prov) |> 
+  summarise(count = n(), median = median(WEIGHT, na.rm = TRUE))
+res_bw_grav <- wilcox.test(WEIGHT ~ Prov, data = wt_blank_grav, exact = FALSE)
+## Blanks length
+dplyr::group_by(wt_blank_grav, Prov) |> 
+  summarise(count = n(), median = median(LENGTH, na.rm = TRUE))
+res_bl_grav <- wilcox.test(LENGTH ~ Prov, data = wt_blank_grav, exact = FALSE)
+## Blanks width
+dplyr::group_by(wt_blank_grav, Prov) |> 
+  summarise(count = n(), median = median(MAXWIDTH, na.rm = TRUE))
+res_bwi_grav <- wilcox.test(MAXWIDTH ~ Prov, data = wt_blank_grav, exact = FALSE)
+
+## Gravettian cores statistical test
+wt_core_grav <- wilcox_t |> 
+  filter(Level %in% c("6", "7")) |> 
+  filter(CLASS == "Core")
+
+## Cores weight
+dplyr::group_by(wt_core_grav, Prov) |> 
+  summarise(count = n(), median = median(WEIGHT, na.rm = TRUE))
+res_cw_grav <- wilcox.test(WEIGHT ~ Prov, data = wt_core_grav, exact = FALSE)
+## Cores length
+dplyr::group_by(wt_core_grav, Prov) |> 
+  summarise(count = n(), median = median(LENGTH, na.rm = TRUE))
+res_cl_grav <- wilcox.test(LENGTH ~ Prov, data = wt_core_grav, exact = FALSE)
+## Cores width
+dplyr::group_by(wt_core_grav, Prov) |> 
+  summarise(count = n(), median = median(MAXWIDTH, na.rm = TRUE))
+res_cwi_grav <- wilcox.test(MAXWIDTH ~ Prov, data = wt_core_grav, exact = FALSE)
+
+# Shelter wilcox statistics
+## DB prep
+wilcox_t_shelter <- solutrean_dataset |> 
+  dplyr::mutate(Prov = case_when(
+    RMTYPE == "T1" | RMTYPE == "T2" | RMTYPE == "T3" | RMTYPE == "T4" | RMTYPE == "T5" ~ "Local",
+    RMTYPE == "T6" | RMTYPE == "T7" | RMTYPE == "T8" | RMTYPE == "T9" | RMTYPE == "T11" ~ "Non-local",
+    TRUE ~ RMTYPE)) |> 
+   mutate(CLASS = case_when(
+    CLASS == "Flake" | CLASS == "ElongBlank" | CLASS == "Thinning flake" ~ "Blank",
+    TRUE ~ CLASS)) |>
+  filter(Prov %in% c("Local", "Non-local")) |> 
+  select(ID, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT, Prov, LEVEL)
+
+# Solutrean blanks statistical test
+wt_blank_s <- wilcox_t_shelter |>
+  filter(CLASS == "Blank")
+
+## Blanks weight
+dplyr::group_by(wt_blank_s, Prov) |> 
+  summarise(count = n(), median = median(WEIGHT, na.rm = TRUE))
+res_bw_s <- wilcox.test(WEIGHT ~ Prov, data = wt_blank_s, exact = FALSE)
+## Blanks length
+dplyr::group_by(wt_blank_s, Prov) |> 
+  summarise(count = n(), median = median(LENGTH, na.rm = TRUE))
+res_bl_s <- wilcox.test(LENGTH ~ Prov, data = wt_blank_s, exact = FALSE)
+## Blanks width
+dplyr::group_by(wt_blank_s, Prov) |> 
+  summarise(count = n(), median = median(MAXWIDTH, na.rm = TRUE))
+res_bwi_s <- wilcox.test(MAXWIDTH ~ Prov, data = wt_blank_s, exact = FALSE)
+
+wt_core_s <- wilcox_t_shelter |>
+  filter(CLASS == "Core")
+
+## Cores weight
+dplyr::group_by(wt_core_s, Prov) |> 
+  summarise(count = n(), median = median(WEIGHT, na.rm = TRUE))
+res_cw_s <- wilcox.test(WEIGHT ~ Prov, data = wt_core_s, exact = FALSE)
+## Cores length
+dplyr::group_by(wt_core_s, Prov) |> 
+  summarise(count = n(), median = median(LENGTH, na.rm = TRUE))
+res_cl_s <- wilcox.test(LENGTH ~ Prov, data = wt_core_s, exact = FALSE)
+## Cores width
+dplyr::group_by(wt_core_s, Prov) |> 
+  summarise(count = n(), median = median(MAXWIDTH, na.rm = TRUE))
+res_cwi_s <- wilcox.test(MAXWIDTH ~ Prov, data = wt_core_s, exact = FALSE)
+
+p.value_blanks <- tibble(
+  Technocomplex = c("Gravettian", " "," ", " ", "Proto-Solutrean", " ", " ", " ", "Solutrean", " ", " ", " "),
+  Variable = c("Weight", "Length", "Width", "(n)","Weight", "Length", "Width","(n)", "Weight", "Length", "Width", "(n)"),
+  L_median = c("2.8", "25.8", "19.1", "82", "2.2", "23.4", "16.8", "154", "1.7", "25.1", "16.6", "342"),
+  NL_median = c("1.9", "21.2", "17.8", "69", "1.7", "19.4", "16.6", "49", "1.5", "21.9", "18.8", "86"),
+  P_Value = round(c(res_bw_grav$p.value, res_bl_grav$p.value, res_bwi_grav$p.value, 0, res_bw_proto$p.value, res_bl_proto$p.value, res_bwi_proto$p.value, 0, res_bw_s$p.value, res_bl_s$p.value, res_bwi_s$p.value, 0), 5)) |>   mutate(Significance = case_when(
+      P_Value <= 0.001 ~ "***",
+      P_Value <= 0.01  ~ "**",
+      P_Value <= 0.05  ~ "*",
+      TRUE             ~ ""),
+    P_value = paste0(format(P_Value, nsmall = 4), Significance)  # Keep decimal places
+  ) |> 
+  select(-P_Value, -Significance) |> 
+  dplyr::mutate(P_value = case_when(
+    P_value == "0.00000***" ~ "-",
+    TRUE ~ P_value))
+
+p.value_cores <- tibble(
+  Technocomplex = c("Gravettian", "", "", "", "Solutrean", "", "", ""),
+  Variable = c("Weight", "Length", "Width", "(n)", "Weight", "Length", "Width", "(n)"),
+  L_median = c("20", "29.6", "27.6", "14", "16.5", "28.3", "30.7", "59"),
+  NL_median = c("11.8", "21.9", "29.7", "10", "11.7", "25.6", "26.0", "7"),
+  P_Value = round(c(res_cw_grav$p.value, res_cl_grav$p.value, res_cwi_grav$p.value, 0, res_cw_s$p.value, res_cl_s$p.value, res_cwi_s$p.value, 0), 3)) |> 
+  mutate(Significance = case_when(
+      P_Value <= 0.001 ~ "***",
+      P_Value <= 0.01  ~ "**",
+      P_Value <= 0.05  ~ "*",
+      TRUE             ~ ""),
+    P_value = paste0(format(P_Value, nsmall = 3), Significance)  # Keep decimal places
+  ) |> 
+  select(-P_Value, -Significance) |> 
+  dplyr::mutate(P_value = case_when(
+    P_value == "0.000***" ~ "-",
+    TRUE ~ P_value))
+
+```
 
@@
 It is important to note, however, that the sample size for cores in the Gravettian assemblage and non-local cherts in the Solutrean assemblage are smaller (n < 15) than the blank sample size.
+
+```{r}
+#| label: tbl-pvalue
+#| tbl-cap: "Median values and p value results for the Mann-Whitney test applied to blank and core measurements (weight, length and width)."
+#| tbl-subcap: 
+#|  - "Median values and p value results for the Mann-Whitney test applied to blank measurements"
+#|  - "Median values and p value results for the Mann-Whitney test applied to core measurements"
+#| layout-nrow: 2
+
+kable(head(p.value_blanks))
+kable(head(p.value_cores))
+
+```
 

### 2025-02-26 analysis/paper/paper.qmd
PC afternoon work session. Improved codes, trying graphs instead of tables for main text

--- 2025-02-24 analysis/paper/paper.qmd
@@
   mutate(RMTYPE = case_when(
-    RMTYPE == "Type 2" ~ "TYPE 2",
-    RMTYPE == "Chalcedony" ~ "TYPE 1", # Altering first group names to final group names. Subtypes are maintained as before.
-    RMTYPE == "TYPE 15" ~ "TYPE 3",
-    RMTYPE == "TYPE 7" ~ "TYPE 4",
-    RMTYPE == "TYPE 12" ~ "TYPE 5",
-    RMTYPE == "TYPE 11" ~ "TYPE 6",
-    RMTYPE == "TYPE 13" ~ "TYPE 7",
-    RMTYPE == "TYPE 11F" ~ "TYPE 8",
-    RMTYPE == "TYPE 13B" ~ "TYPE 9",
-    RMTYPE == "Oolitic" ~ "TYPE 10",
-    RMTYPE == "TYPE GB" ~ "TYPE 11",
+    RMTYPE == "TYPE 2" | RMTYPE == "Type 2" ~ "T2",
+    RMTYPE == "Chalcedony" ~ "T1", # Altering first group names to final group names. Subtypes are maintained as before.
+    RMTYPE == "TYPE 15" ~ "T3",
+    RMTYPE == "TYPE 7" ~ "T4",
+    RMTYPE == "TYPE 12" ~ "T5",
+    RMTYPE == "TYPE 11" ~ "T6",
+    RMTYPE == "TYPE 13" ~ "T7",
+    RMTYPE == "TYPE 11F" ~ "T8",
+    RMTYPE == "TYPE 13B" ~ "T9",
+    RMTYPE == "Oolitic" ~ "T10",
+    RMTYPE == "TYPE GB" ~ "T11",
     .default = RMTYPE)) |> 
-  select(AREA, ID, RMTYPE, SUBTYPE, WEIGHT, CORTEXTHICKNESS, CORTEXAPPEARANCE) |>
+  select(AREA, ID, RMTYPE, SUBTYPE, WEIGHT, CORTEXTHICKNESS, CORTEXAPPEARANCE) %>%
   distinct(ID, .keep_all = TRUE)
@@
     Level == "4E" | Level == "4E/5" | Level == 5 ~ "Proto-Solutrean",
-    Level == 6 | Level == 7 ~ "Gravettian"))
+    Level == 6 | Level == 7 ~ "Gravettian")) |> 
+  mutate(RETOUCHEDPIECETYPOLOGY = case_when( # Simplifying the retouched typology
+    RETOUCHEDPIECETYPOLOGY == "Splintered piece with notch" | RETOUCHEDPIECETYPOLOGY == "Splintered piece with lateral retouch" | RETOUCHEDPIECETYPOLOGY == "Splintered piece fragment" ~ "Splintered piece",
+    RETOUCHEDPIECETYPOLOGY == "Retouched blade" | RETOUCHEDPIECETYPOLOGY == "Retouched bladelet" | RETOUCHEDPIECETYPOLOGY == "Retouched flake" | RETOUCHEDPIECETYPOLOGY == "Retouched burin spall" ~ "Retouched blank",
+    RETOUCHEDPIECETYPOLOGY == "Retouched bladelet frag" | RETOUCHEDPIECETYPOLOGY == "Retouched flake frag" | RETOUCHEDPIECETYPOLOGY == "Retouched flake siret fragment"  ~ "Retouched blank fragment",
+    RETOUCHEDPIECETYPOLOGY == "Multiple burin" | RETOUCHEDPIECETYPOLOGY == "Burin splintered piece" | RETOUCHEDPIECETYPOLOGY == "Burin on truncation" | RETOUCHEDPIECETYPOLOGY == "Retouched burin" ~ "Burin",
+    RETOUCHEDPIECETYPOLOGY == "Dentilhado" ~ "Denticulate",
+    RETOUCHEDPIECETYPOLOGY == "Right truncation on retouched piece" ~ "Truncation",
+    RETOUCHEDPIECETYPOLOGY == "Backed bladelet" | RETOUCHEDPIECETYPOLOGY == "Backed flake" | RETOUCHEDPIECETYPOLOGY == "Double backed bladelet" | RETOUCHEDPIECETYPOLOGY == "Backed bladelet parcial" ~ "Backed tool",
+    TRUE ~ RETOUCHEDPIECETYPOLOGY))
 
@@
 
-# Setting up the Terrace dataset
+# Setting up the Shelter dataset
 cols_to_concat_shelter <- c("YEAR", "ID") # Concatenate YEAR and ID variables in one single variable (e.g. 05-500)
@@
   mutate(RMTYPE = case_when(
-    RMTYPE == "Type 2" ~ "TYPE 2",
-    RMTYPE == "Chalcedony" ~ "TYPE 1", # Altering first group names to final group names. Subtypes are maintained as before.
-    RMTYPE == "TYPE 15" ~ "TYPE 3",
-    RMTYPE == "TYPE 7" ~ "TYPE 4",
-    RMTYPE == "TYPE 12" ~ "TYPE 5",
-    RMTYPE == "TYPE 11" ~ "TYPE 6",
-    RMTYPE == "TYPE 13" ~ "TYPE 7",
-    RMTYPE == "TYPE 11F" ~ "TYPE 8",
-    RMTYPE == "TYPE 13B" ~ "TYPE 9",
-    RMTYPE == "Oolitic" ~ "TYPE 10",
-    RMTYPE == "TYPE GB" ~ "TYPE 11",
+    RMTYPE == "Type 2" | RMTYPE == "TYPE 2" ~ "T2",
+    RMTYPE == "Chalcedony" ~ "T1", # Altering first group names to final group names. Subtypes are maintained as before.
+    RMTYPE == "TYPE 15" ~ "T3",
+    RMTYPE == "TYPE 7" ~ "T4",
+    RMTYPE == "TYPE 12" ~ "T5",
+    RMTYPE == "TYPE 11" ~ "T6",
+    RMTYPE == "TYPE 13" ~ "T7",
+    RMTYPE == "TYPE 11F" ~ "T8",
+    RMTYPE == "TYPE 13B" ~ "T9",
+    RMTYPE == "Oolitic" ~ "T10",
+    RMTYPE == "TYPE GB" ~ "T11",
     RMTYPE == "ND_A" ~ "INDET",
@@
   drop_na(RMTYPE) |>   # Dropping NAs on the RM database drops dataset to 1514
-  drop_na(CLASS) # Dropping NAs on the technological class drops dataset to 914
+  drop_na(CLASS) |>  # Dropping NAs on the technological class drops dataset to 914
+  dplyr::mutate(RETOUCHEDPIECETYPOLOGY = case_when(
+    RETOUCHEDPIECETYPOLOGY == "Bifacial fragment" ~ "Bifacial blank",
+    RETOUCHEDPIECETYPOLOGY == "Shouldered point fragment" ~ "Shouldered point",
+    TRUE ~ RETOUCHEDPIECETYPOLOGY))
 
@@
     dplyr::mutate(RMTYPE = case_when(
-    RMTYPE == "TYPE 1" | RMTYPE == "TYPE 2" | RMTYPE == "TYPE 3" | RMTYPE == "TYPE 4" | RMTYPE == "TYPE 5" ~ "Local",
-    RMTYPE == "TYPE 6" ~ "T6",
-    RMTYPE == "TYPE 7" ~ "T7",
-    RMTYPE == "TYPE 8" ~ "T8",
-    RMTYPE == "TYPE 9" ~ "T9",
-    RMTYPE == "TYPE 10" ~ "T10",
+    RMTYPE == "T1" | RMTYPE == "T2" | RMTYPE == "T3" | RMTYPE == "T4" | RMTYPE == "T5" ~ "Local",
     TRUE ~ RMTYPE)) |> 
@@
     CLASS == "RetouchedPieceFrag" ~ "Retouched fragment",
-    TRUE ~ CLASS)) |> 
-  dplyr::mutate(RETOUCHEDPIECETYPOLOGY = case_when(
-    RETOUCHEDPIECETYPOLOGY == "Retouched blade" | RETOUCHEDPIECETYPOLOGY == "Retouched bladelet" | RETOUCHEDPIECETYPOLOGY == "Retouched bladelet fragment" | RETOUCHEDPIECETYPOLOGY == "Retouched flake" | RETOUCHEDPIECETYPOLOGY == "Retouched flake fragment" ~ "Retouched blank",
-    RETOUCHEDPIECETYPOLOGY == "Splintered piece fragment" ~ "Splintered piece",
-    TRUE ~ RETOUCHEDPIECETYPOLOGY))
+    TRUE ~ CLASS))
 
@@
    dplyr::mutate(RMTYPE = case_when(
-    RMTYPE == "TYPE 1" | RMTYPE == "TYPE 2" | RMTYPE == "TYPE 3" | RMTYPE == "TYPE 4" | RMTYPE == "TYPE 5" ~ "Local",
-    RMTYPE == "TYPE 6" ~ "T6",
-    RMTYPE == "TYPE 7" ~ "T7",
-    RMTYPE == "TYPE 8" ~ "T8",
-    RMTYPE == "TYPE 9" ~ "T9",
-    RMTYPE == "TYPE 10" ~ "T10",
-    RMTYPE == "TYPE 11" ~ "T11",
+    RMTYPE == "T1" | RMTYPE == "T2" | RMTYPE == "T3" | RMTYPE == "T4" | RMTYPE == "T5" ~ "Local",
     TRUE ~ RMTYPE)) |> 
@@
 ```{r}
+#| label: general-graphs
+
+phase_dataset |> 
+    dplyr::mutate(RMTYPE = case_when(
+    RMTYPE == "T1" | RMTYPE == "T2" | RMTYPE == "T3" | RMTYPE == "T4" | RMTYPE == "T5" ~ "Local",
+    TRUE ~ RMTYPE)) |> 
+  dplyr::mutate(CLASS = case_when(
+    CLASS == "BlankFrag" ~ "Blank",
+    CLASS == "BurinSpall" ~ "Burin spall",
+    CLASS == "CoreFrag" ~ "Core",
+    CLASS == "CorePreparProd" ~ "Core prep product",
+    CLASS == "RetouchedPiece" | CLASS == "RetouchedPieceFrag" ~ "Retouched",
+    TRUE ~ CLASS)) |> 
+  filter(Level %in% c(6, 7)) |>
+  filter(RMTYPE != "INDET") |> 
+  mutate(RMTYPE = factor(RMTYPE, levels = c("Local", "T6","T7", 
+                                        "T8", "T9", "T10", "TL"))) |>
+  dplyr::group_by(RMTYPE, CLASS) |> 
+  dplyr::summarise(n = n()) |> 
+  mutate(percent = n/sum(n)) |>
+  ggplot(aes(x = RMTYPE, y = percent, fill = CLASS)) +
+  geom_bar(stat = "identity", width = 0.5) +
+  geom_text(aes(label=ifelse(n > 3, n, "")), position = position_stack(vjust = .5), size = 3, color = "white") +
+  scale_y_continuous(labels = scales::percent_format()) +
+  labs(x = "Gravettian", y = "Count %") +
+  theme_minimal() +
+  scale_fill_manual(values = c("#D49B54", "#8B6969", "#CDAF95", "#2B4F60",
+                              "tomato4", "#798777"))
+
+phase_dataset |> 
+    dplyr::mutate(RMTYPE = case_when(
+    RMTYPE == "T1" | RMTYPE == "T2" | RMTYPE == "T3" | RMTYPE == "T4" | RMTYPE == "T5" ~ "Local",
+    TRUE ~ RMTYPE)) |> 
+  dplyr::mutate(CLASS = case_when(
+    CLASS == "BlankFrag" ~ "Blank",
+    CLASS == "BurinSpall" ~ "Burin spall",
+    CLASS == "CoreFrag" ~ "Core",
+    CLASS == "CorePreparProd" ~ "Core prep product",
+    CLASS == "RetouchedPiece" | CLASS == "RetouchedPieceFrag" ~ "Retouched",
+    TRUE ~ CLASS)) |> 
+  filter(Level %in% c("4E", "4E/5")) |>
+  filter(RMTYPE != "INDET") |> 
+  filter(CLASS != "Chip") |> 
+  mutate(RMTYPE = factor(RMTYPE, levels = c("Local", "T6","T7", 
+                                        "T8", "T9", "T10", "TL"))) |>
+  dplyr::group_by(RMTYPE, CLASS) |> 
+  dplyr::summarise(n = n()) |> 
+  mutate(percent = n/sum(n)) |>
+  ggplot(aes(x = RMTYPE, y = percent, fill = CLASS)) +
+  geom_bar(stat = "identity", width = 0.5) +
+  geom_text(aes(label=ifelse(n > 3, n, "")), position = position_stack(vjust = .5), size = 3, color = "white") +
+  scale_y_continuous(labels = scales::percent_format()) +
+  labs(x = "Proto-Solutrean", y = "Count %") +
+  theme_minimal() +
+  scale_fill_manual(values = c("#D49B54", "#8B6969", "#CDAF95", "#2B4F60",
+                              "tomato4", "#798777"))
+
+solutrean_dataset |> 
+   dplyr::mutate(RMTYPE = case_when(
+    RMTYPE == "T1" | RMTYPE == "T2" | RMTYPE == "T3" | RMTYPE == "T4" | RMTYPE == "T5" ~ "Local",
+    TRUE ~ RMTYPE)) |> 
+  dplyr::mutate(CLASS = case_when(
+    CLASS == "ElongBlank" | CLASS == "Flake" ~ "Blank",
+    CLASS == "ElongBlankFrag" | CLASS == "FlakeFrag" ~ "Blank",
+    CLASS == "Thinning flake" ~ "Blank",
+    CLASS == "CorePrepProd" ~ "Core prep product",
+    CLASS == "RetouchedPiece" | CLASS == "Retouched piece" | CLASS == "Retouched piece fragment" ~ "Retouched",
+    CLASS == "Core fragment" ~ "Core",
+    TRUE ~ CLASS)) |> 
+  filter(RMTYPE != "INDET") |> 
+  mutate(RMTYPE = factor(RMTYPE, levels = c("Local", "T6","T7", 
+                                        "T8", "T9", "T10","T11", "TL"))) |>
+  dplyr::group_by(RMTYPE, CLASS) |> 
+  dplyr::summarise(n = n()) |> 
+  mutate(percent = n/sum(n)) |>
+  ggplot(aes(x = RMTYPE, y = percent, fill = CLASS)) +
+  geom_bar(stat = "identity", width = 0.5) +
+  geom_text(aes(label=ifelse(n > 3, n, "")), position = position_stack(vjust = .5), size = 3, color = "white") +
+  scale_y_continuous(labels = scales::percent_format()) +
+  labs(x = "Solutrean", y = "Count %") +
+  theme_minimal() +
+  scale_fill_manual(values = c("#D49B54", "#8B6969", "#CDAF95", "#2B4F60",
+                              "tomato4", "#798777"))
+
+```
+
+```{r}
 #| label: stats-setup
@@
   dplyr::mutate(RMTYPE = case_when(
-    RMTYPE == "TYPE 1" | RMTYPE == "TYPE 2" | RMTYPE == "TYPE 3" | RMTYPE == "TYPE 4" | RMTYPE == "TYPE 5" ~ "Local",
-    RMTYPE == "TYPE 6" ~ "T6",
-    RMTYPE == "TYPE 7" ~ "T7",
-    RMTYPE == "TYPE 8" ~ "T8",
-    RMTYPE == "TYPE 9" ~ "T9",
-    RMTYPE == "TYPE 10" ~ "T10",
-    RMTYPE == "TYPE 11" ~ "T11",
+    RMTYPE == "T1" | RMTYPE == "T2" | RMTYPE == "T3" | RMTYPE == "T4" | RMTYPE == "T5" ~ "Local",
     TRUE ~ RMTYPE))
@@
   dplyr::mutate(RMTYPE = case_when(
-    RMTYPE == "TYPE 1" | RMTYPE == "TYPE 2" | RMTYPE == "TYPE 3" | RMTYPE == "TYPE 4" | RMTYPE == "TYPE 5" ~ "Local",
-    RMTYPE == "TYPE 6" ~ "T6",
-    RMTYPE == "TYPE 7" ~ "T7",
-    RMTYPE == "TYPE 8" ~ "T8",
-    RMTYPE == "TYPE 9" ~ "T9",
-    RMTYPE == "TYPE 10" ~ "T10",
+    RMTYPE == "T1" | RMTYPE == "T2" | RMTYPE == "T3" | RMTYPE == "T4" | RMTYPE == "T5" ~ "Local",
     TRUE ~ RMTYPE))
@@
 ```{r}
-#| label: tool-diversity
-
-# Solutrean table
-## Table preparation
-solutrean_retouch <- general_table_solutrean |>
-  filter(CLASS %in% c("Retouched", "Retouched fragment")) |>
-  filter(RMTYPE != "INDET") |> 
-  filter(RETOUCHEDPIECETYPOLOGY != "") |> 
-  tabyl(RETOUCHEDPIECETYPOLOGY, RMTYPE) |>
-  adorn_totals(c("row", "col")) |>
-  mutate(across(where(is.numeric), ~ ifelse(. == 0, " ", .))) |>
-  select(RETOUCHEDPIECETYPOLOGY, Local, T6, T7, T8, T11, Total) |> # Ordering values
-  rename(Typology = RETOUCHEDPIECETYPOLOGY)
-
-# Proto-solutrean table
-## Table preparation
-proto_retouch <- general_table |>
-  filter(CLASS %in% c("Retouched", "Retouched fragment")) |>
-  filter(Level %in% c("4E", "4E/5")) |>
-  filter(!RMTYPE == "INDET") |> 
-  tabyl(RETOUCHEDPIECETYPOLOGY, RMTYPE) |>
-  adorn_totals(c("row", "col")) |> 
-  mutate(across(where(is.numeric), ~ ifelse(. == 0, " ", .))) |>
-  select(RETOUCHEDPIECETYPOLOGY, Local, T6, T7, T8, T9, TL, Total) |> # Ordering values
-  rename(Typology = RETOUCHEDPIECETYPOLOGY)
-
-# Gravettian table
-## Table preparation
-gravettian_retouch <- general_table |> 
-  filter(CLASS %in% c("Retouched", "Retouched fragment")) |>
-  filter(Level %in% c(6, 7)) |> 
-  filter(!RMTYPE == "INDET") |> 
-  tabyl(RETOUCHEDPIECETYPOLOGY, RMTYPE) |>
-  adorn_totals(c("row", "col")) |>
-  mutate(across(where(is.numeric), ~ ifelse(. == 0, " ", .))) |>
-  select(RETOUCHEDPIECETYPOLOGY, Local, T6, T7, T8, T9, T10, TL, Total) |> # Ordering values
-  rename(Typology = RETOUCHEDPIECETYPOLOGY)
-
-```
-
-```{r}
 #| label: tbl-general
@@
 kable(solutrean_general)
-
-```
-
-```{r}
-#| label: tbl-retouch
-#| tbl-cap: Number and percentage of retouched tools by raw material of the Gravettian (a), Proto-Solutrean (b) and Solutrean (c) assemblages.
-Retouched blank typology includes flakes, blades and bladelets and respective fragments.
-#| tbl-subcap: 
-#| - "Number and percentage of retouched tools by raw material from the Gravettian assemblage (levels 6 and 7) of the Terrace area."
-#| - "Number and percentage of retouched tools by raw material from the Proto-Solutrean assemblage (levels 4E and 5) of the Terrace area."
-#| - "Number and percentage of retouched tools by raw material from the Solutrean assemblage (levels A-C) of the Shelter area."
-#| layout-nrow: 4
-
-kable(gravettian_retouch)
-kable(proto_retouch)
-kable(solutrean_retouch)
 
@@
   dplyr::mutate(Source = case_when(
-    RMTYPE == "TYPE 1" | RMTYPE == "TYPE 2" | RMTYPE == "TYPE 3" | RMTYPE == "TYPE 4" | RMTYPE == "TYPE 5" ~ "Local",
-    RMTYPE == "TYPE 6" ~ "T6",
-    RMTYPE == "TYPE 7" ~ "T7",
-    RMTYPE == "TYPE 8" ~ "T8",
-    RMTYPE == "TYPE 9" ~ "T9",
-    RMTYPE == "TYPE 10" ~ "T10",
-    RMTYPE == "TYPE 11" ~ "T11",
+    RMTYPE == "T1" | RMTYPE == "T2" | RMTYPE == "T3" | RMTYPE == "T4" | RMTYPE == "T5" ~ "Local",
     TRUE ~ RMTYPE)) |>
@@
   dplyr::mutate(Source = case_when(
-    RMTYPE == "TYPE 1" | RMTYPE == "TYPE 2" | RMTYPE == "TYPE 3" | RMTYPE == "TYPE 4" | RMTYPE == "TYPE 5" ~ "Local",
-    RMTYPE == "TYPE 6" ~ "T6",
-    RMTYPE == "TYPE 7" ~ "T7",
-    RMTYPE == "TYPE 8" ~ "T8",
-    RMTYPE == "TYPE 10" ~ "T10",
-    RMTYPE == "TYPE 11" ~ "T11",
+    RMTYPE == "T1" | RMTYPE == "T2" | RMTYPE == "T3" | RMTYPE == "T4" | RMTYPE == "T5" ~ "Local",
     TRUE ~ RMTYPE)) |>

### 2025-02-28 analysis/Supplementary/supplementary_tables.qmd
Macbook daily work session. Results rewritten and added to main text, discussion altered. Main paper and SOM tables/graphs improved and reordered.

@@
 #| label: tbl-bcratio
-#| tbl-cap: Blank to core ratio by raw material of the Solutrean (a), Proto-Solutrean (b) and Gravettian (c) assemblages.
+#| tbl-cap: Blank to core ratio by raw material of the Gravettian (a), Proto-Solutrean (b) and Solutrean (c) assemblages.
 #| tbl-subcap: 
+#|  - "Blank to core ratio by raw material of the Gravettian assemblage (levels 6 and 7) of the Terrace area."
+#|  - "Blank to core ratio by raw material of the Proto-Solutrean assemblage (levels 4E and top of level 5) of the Terrace area."
 #|  - "Blank to core ratio by raw material of the Solutrean assemblage (levels A to C) of the Shelter area."
-#|  - "Blank to core ratio by raw material of the Proto-Solutrean assemblage (levels 4E and top of level 5) of the Terrace area."
-#|  - "Blank to core ratio by raw material of the Gravettian assemblage (levels 6 and 7) of the Terrace area."
-#| layout-nrow: 4
-
-kable(head(solutrean_ratiobc))
-kable(head(proto_ratiobc))
-kable(head(gravettian_ratiobc))
+#| layout-nrow: 3
+
+kable(gravettian_ratiobc)
+kable(proto_ratiobc)
+kable(solutrean_ratiobc)
 
@@
 #| label: tbl-tdratio-solutrean
-#| tbl-cap: Tools to debitage ratio by raw material of the Solutrean (a), Proto-Solutrean (b) and Gravettian (c) assemblages.
+#| tbl-cap: Tools to debitage ratio by raw material of the Gravettian (a), Proto-Solutrean (b) and Solutrean (c) assemblages.
 #| tbl-subcap: 
+#|  - "Tools to debitage ratio by raw material of the Gravettian assemblage (levels 6 and 7) of the Terrace area."
+#|  - "Tools to debitage ratio by raw material of the Proto-Solutrean assemblage (levels 4E and top of level 5) of the Terrace area."
 #|  - "Tools to debitage ratio by raw material of the Solutrean assemblage (levels A to C) of the Shelter area."
-#|  - "Tools to debitage ratio by raw material of the Proto-Solutrean assemblage (levels 4E and top of level 5) of the Terrace area."
-#|  - "Tools to debitage ratio by raw material of the Gravettian assemblage (levels 6 and 7) of the Terrace area."
-#| layout-nrow: 4
-
-kable(head(solutrean_ratiotd))
-kable(head(proto_ratiotd))
-kable(head(gravettian_ratiotd))
+#| layout-nrow: 3
+
+kable(gravettian_ratiotd)
+kable(proto_ratiotd)
+kable(solutrean_ratiotd)
 
@@
   select(RMTYPE, CLASS, MAXWIDTH, LENGTH, WEIGHT) |> 
+  drop_na() |> 
   group_by(RMTYPE, CLASS) |> 
-  summarise(across(c(MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
+  summarise(across(c(MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd)))
 # For cores
@@
   select(RMTYPE, CLASS, MAXWIDTH, LENGTH, WEIGHT) |> 
+  drop_na() |> 
   group_by(RMTYPE, CLASS) |> 
-  summarise(across(c(MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
+  summarise(across(c(MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd)))
 
@@
   group_by(RMTYPE, CLASS) |> 
-  summarise(across(c(MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
+  summarise(across(c(MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd)))
 # For cores
@@
   group_by(RMTYPE, CLASS) |> 
-  summarise(across(c(MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
+  summarise(across(c(MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd)))
 
@@
   filter(Level %in% c(6, 7)) |> 
-  filter(RMTYPE != "INDET")
+  filter(RMTYPE != "INDET") |> 
+  dplyr::mutate(RMTYPE = case_when(
+      RMTYPE == "T2" | RMTYPE == "T3" | RMTYPE == "T4" | RMTYPE == "T5" ~ "T2-5",
+      TRUE ~ RMTYPE))
 
@@
   group_by(RMTYPE, CLASS) |> 
-  summarise(across(c(MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd, max = max, min = min)))
+  summarise(across(c(MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd)))
+
+```
+
+```{r}
+#| label: tbl-gravettian
+#| tbl-cap: "Stats for Gravettian occupation."
+#| tbl-subcap: 
+#|  - "Stats for blanks"
+#|  - "Stats for cores"
+#| layout-nrow: 2
+
+kable(gravettian_stats_blank)
+kable(gravettian_stats_cores)
+
+```
+
+```{r}
+#| label: tbl-proto
+#| tbl-cap: "Stats for Proto-Solutrean occupation."
+#| tbl-subcap: 
+#|  - "Stats for blanks"
+#|  - "Stats for cores"
+#| layout-nrow: 2
+
+kable(proto_stats_blank)
+kable(proto_stats_cores)
 
@@
 
-kable(head(solutrean_stats_blank))
-kable(head(solutrean_stats_cores))
-
-```
-
-```{r}
-#| label: tbl-proto
-#| tbl-cap: "Stats for Proto-Solutrean occupation."
-#| tbl-subcap: 
-#|  - "Stats for blanks"
-#|  - "Stats for cores"
-#| layout-nrow: 2
-
-kable(head(proto_stats_blank))
-kable(head(proto_stats_cores))
-
-```
-
-```{r}
-#| label: tbl-gravettian
-#| tbl-cap: "Stats for Gravettian occupation."
-#| tbl-subcap: 
-#|  - "Stats for blanks"
-#|  - "Stats for cores"
-#| layout-nrow: 2
-
-kable(head(gravettian_stats_blank))
-kable(head(gravettian_stats_cores))
+kable(solutrean_stats_blank)
+kable(solutrean_stats_cores)
 
@@
 
-A Mann-Whitney U was conducted to determine whether there are differences in the measurements (weight, length and width) of blanks and cores between local and non-local cherts.
+A Mann-Whitney U statistical test was conducted to determine whether there are differences in the measurements (weight, length and width) of blanks and cores between local and non-local cherts.
 To do this, all local cherts (including chalcedony) were grouped in a Local category (L) while all non-local cherts were grouped in a non-local category (NL).
@@
 
-kable(head(p.value_blanks))
-kable(head(p.value_cores))
+kable(p.value_blanks)
+kable(p.value_cores)
 
@@
 
-kable(head(TL_classes))
-
-```
+kable(TL_classes)
+
+```

### 2025-02-28 analysis/paper/paper.qmd
Macbook daily work session. Results rewritten and added to main text, discussion altered. Main paper and SOM tables/graphs improved and reordered.

@@
 #| label: general-tables
+#| include: false
 
@@
 
-phase_dataset |> 
+barplot_gravettian <- phase_dataset |> 
     dplyr::mutate(RMTYPE = case_when(
@@
   scale_y_continuous(labels = scales::percent_format()) +
-  labs(x = "Gravettian", y = "Count %") +
   theme_minimal() +
+  theme(axis.title.x=element_blank()) +
+  labs(y = " ") +
   scale_fill_manual(values = c("#D49B54", "#8B6969", "#CDAF95", "#2B4F60",
@@
 
-phase_dataset |> 
+barplot_proto <- phase_dataset |> 
     dplyr::mutate(RMTYPE = case_when(
@@
   scale_y_continuous(labels = scales::percent_format()) +
-  labs(x = "Proto-Solutrean", y = "Count %") +
+  labs(y = " ") +
   theme_minimal() +
+  theme(axis.title.x=element_blank()) +
   scale_fill_manual(values = c("#D49B54", "#8B6969", "#CDAF95", "#2B4F60",
@@
 
-solutrean_dataset |> 
+barplot_solutrean <- solutrean_dataset |> 
    dplyr::mutate(RMTYPE = case_when(
@@
   scale_y_continuous(labels = scales::percent_format()) +
-  labs(x = "Solutrean", y = "Count %") +
+  labs(y = " ") +
   theme_minimal() +
+  theme(axis.title.x=element_blank()) +
   scale_fill_manual(values = c("#D49B54", "#8B6969", "#CDAF95", "#2B4F60",
@@
 
-```{r}
-#| label: tbl-general
-#| tbl-cap: Number and percentage of technotypological classes by raw material type of the Gravettian (a), Proto-Solutrean (b) and Solutrean (c) assemblages.
-#| tbl-subcap: 
-#| - "Number and percentage of technotypological classes by raw material from the Gravettian assemblage (levels 6 and 7) of the Terrace area."
-#| - "Number and percentage of technotypological classes by raw material from the Proto-Solutrean assemblage (levels 4E and 5) of the Terrace area."
-#| - "Number and percentage of technotypological classes by raw material from the Solutrean assemblage (levels A-C) of the Shelter area."
-#| layout-nrow: 4
-
-kable(gravettian_general)
-kable(proto_general)
-kable(solutrean_general)
-
-```
-
-Retouched tools were further explored through the tools-to-debitage ratio (@fig-ratiotd and table s++).
-As seen in @fig-ratiotd, local cherts (T1-5) show low ratios (\~2.5-0.5) of retouched tools to blanks, while TL and T6-T8 show higher ratios (\~0.9-1.0).
-The exception is T10 (non-local type), which shows a lower ratio (\~0.3), more akin to local cherts.
-The relation between blank and core numbers was also explored through a blank-to-core ratio (@fig-ratiobc and table s++).
-The results show that local types T1 and T2 show lower ratios (\<4) of blanks to cores, although other local types (T3-T5) have no identified cores @tbl-general-gravettian.
-The non-local types, especially T6 and T7 show higher ratios (\~8) of blanks to cores, with T10 once again showing values similar to local cherts (\~4).
+In all assemblages, blanks and blank fragments are the most frequent class of lithic artefacts and in all chert types (>40% of the assemblages).
+Cores are, as expected, found in smaller percentages in all assemblages (~3-8%).
+Cores are present in local cherts in larger numbers than in non-local types (P++); their presence is noted in almost all non-local cherts in all assemblages, except for the TL group.
+In the Proto-Solutrean assemblage however, non-local chert core presence is small and limited to T7-9 (n=4 complete cores and n=2 core fragments).
+Albeit in very small numbers in all assemblages (n<10), core preparation and maintenance products (P2) are also present in local and non-local cherts, especially in the non-local types which show the larger sample number such as T6-7.
+The presence of knapping by-products such as shatter is also seen in both local and non-local cherts in the Gravettian and Proto-Solutrean assemblages.
+Retouched tools (P1) were identified in all chert types from all studied assemblages; in general, all assemblages show a similar pattern where local cherts show smaller percentages of retouched tools than non-local cherts.
+However, the assemblages show clear differences in the number of retouched tools.
+
+```{r}
+#| label: fig-general-graphs
+#| fig-cap: Class by chert type in the Gravettian (a), Proto-Solutrean (b) and Solutrean (c) assemblages.
+Labels within the plots correspond to the number of artefacts within each class.
+Only labels of n>3 are shown.
+
+ggpubr::ggarrange(barplot_gravettian, barplot_proto, barplot_solutrean, labels = c("a", "b", "c"), nrow = 3, common.legend = TRUE, legend = "right")
+
+```
+
+The Gravettian assemblage shows higher percentages of retouched tools (~21%) than the other assemblages (<10%).
+Detailed tables with tool typology and counts by chert type and assemblage can be found in the SOM (tables +++).
+Tool diversity (P8) is higher in non-local cherts in the Gravettian assemblage, although with high frequency of less specialised tools such as notches and retouched tools, alongside hunting implements such as backed bladelets (P9).
+The high frequency of burins seen in this assemblage may be related to their use as cores, representing a knapping strategy more than the presence of retouch.
+In comparison to the Gravettian, other assemblages show smaller numbers of retouched tools.
+While in the Proto-Solutrean no pattern was identified between local and non-local retouched tools, in the Solutrean assemblage both tool diversity (P8) and tool specialization (P9) are higher in local cherts.
+
+The relation between blank and core numbers was explored through a blank-to-core ratio (P6; Figure 1 and table s++).
+In all assemblages, local cherts have generally smaller ratios than non-local cherts, although sample sizes for cores are variable (graph ++).
+For the Gravettian assemblages, the results show that non-local cherts show ratios above 7 (with the exception of T10), while local cherts have ratios located below 4.
+The Solutrean assemblage also shows higher ratios for non-local cherts (>12), with local cherts showing ratios below 7.
+The Proto-Solutrean assemblage has most blank-to-core ratios below 10, except for non-local T7.
+However, non-local cores are poorly represented within the assemblage (table ++).
 
@@
 ```
+
+Retouched tools were further explored through the tools-to-debitage ratio (P4; Figure 2 and table s++).
+In general, the result show that non-local chert types have higher retouch-to-debitage ratios than local types, albeit with differences between assemblages.
+The Gravettian assemblage shows the highest tool-to-debitage ratios compared to the remaining assemblages; non-local cherts show ratios above 0.9, while local cherts show values below 0.5.
+In the Proto-Solutrean non-local cherts show ratios between 0.3 to 0.5, while local cherts show ratios below 0.3.
+This pattern is less clear in the Solutrean assemblage since the ratios are the lowest of all assemblages.
+Non-local cherts show ratios above 0.1 (with exception of T8) while local cherts show ratios below 0.1 (with exception of the local chalcedony).
 
@@
 
-The sizes of lithic artefacts (flakes, elongated blanks, cores and retouched tools) are explored through scatterplots of width and length (@fig-gravettian-boxplots a) and boxplots for weight (@fig-gravettian-boxplots b); complete tables with descriptive statistics can also be found in SOM ++, table s++.
+The sizes of blanks (P3) and cores (P5) are explored through scatterplots of width and length (Figure 3 a) and boxplots for weight (Figure 3 b); complete tables with descriptive statistics and non-parametric statistical test results can also be found in SOM ++, table s++.
 The scatterplots show different patterns in the several analysed classes.
-Flakes are longer and wider in local cherts and generally smaller in non-local types (\~20-22 mm maximum lengths for T6 and T7).
-This is similar for elongated blanks, with non-local cherts presenting thinner and short bladelets than local types.
-Core width is variable between both local and non-local cherts; however, non-local cherts show a concentration of cores in the ~20 mm length mark, while local cherts show a wider variation of longer cores.
-Retouched tools show no differences in width and length between local and non-local cherts.
-Regarding the weight (@fig-gravettian-boxplots b), although there is some variability, non-local types T6 and T7 seem to have the lowest weights in all classes (flakes, elongated blanks, cores and retouched tools).
-The lowest means for flakes are for T6 and T7 (\~2.5 gr).
-However, in general, the lightest flakes seem to be from T2 and T6, although T2 also shows several heavy outliers.
-Elongated products show even smaller means for T6 and T7 (0.5-1 gr), especially when compared to local cherts (\~2.5 gr).
-Regarding cores, @tbl-general-gravettian shows that type T2 and T6 have the highest number of cores; between these two types, T6 has the lightest cores, with a maximum of \~21 gr and with a mean of \~12 gr, in comparison to T2 with a maximum of \~37 gr and a mean of \~21 gr.
-Retouched tools shows similar median values among all chert types (2.5-4 gr), although T6 and T7 have the lowest means (\~4 gr); in general, and given the high number of retouched tools for T6 (@tbl-general-gravettian), this non-local chert seems to have been consistently reduced with a third quartile ending in the 5 gr mark.
 
@@
 
-As seen in @fig-ratiotd (see also table s++ in the SOM ++), the lowest ratio of retouched to debitage is seen in T1 and T4, with values below 0.1, while other local types show slightly higher values (0.1-0.2).
-Following the same trend as the Gravettian, the non-local cherts show slightly higher ratios (0.3-0.5) of retouched to debitage.
-Regarding the blank to core ratio (@fig-ratiobc and table s++ in the SOM ++), T2 and T4 show the lowest ratios (8.7 and 6.2 respectively) of blanks per cores, while the non-local T7 shows the highest ratios (11.0) of blanks to cores.
-Unlike the Gravettian, T1 (local chalcedony) and T6 (non-local chert) have no cores (@tbl-general-proto).
-
-Regarding the size (width, length and weight) of lithic artefacts (flakes, elongated blanks, cores and retouched tools), several patterns can be seen between local and non-local chert types (@fig-proto-boxplots, table s++ and s++ in the SOM ++).
-The distribution of width and length (@fig-proto-boxplots a) for flakes shows a concentration of non-local flakes below the 25 mm width and length (with means around 20 mm), while local cherts are more variable and show larger widths and lengths.
-Similarly, the larger elongated blanks are from local cherts, although there is variability in the sizes.
-Although there are clear differences in the number of cores between local and non-local cherts (@tbl-general-proto) making patterns harder to identify, non-local cherts seem to show smaller cores.
-Finally, regarding retouched tools, non-local cherts also seem to be smaller, with widths concentrated below the 20 mm mark, while local cherts are concentrated above the 20 mm mark; no differences were identified regarding length.
-As for weight (@fig-proto-boxplots b and table s++ in SOM ++), local flakes seem to be heavier than  non-local cherts (means of 3.5-4.5 gr and 1.3-2.6 gr, respectively); especially T4 shows heavier flakes and outliers, than T6 and T7.
-In comparison, elongated blanks show no differences between local and non-local cherts.
-Although non-local chert cores are barely represented in the assemblage (@tbl-general-proto), they show the lighter weights than local cherts.
-Retouched tools show the clearest differences, with non-local cherts presenting lighter weights than local cherts.
-
 ```{r}
@@
 
-Regarding the tools-to-debitage and blank-to-core ratios, the Solutrean assemblage shows similar patterns as the previous assemblages (@fig-ratiotd and @fig-ratiobc, respectively).
-Local cherts (T2-T5) show low retouch to debitage ratios (\<0.1), with the exception of T1 (chalcedony), which shows a ratio of 1.8.
-Other non-local cherts (T6-T7 and T11) show slightly higher retouch to debitage ratios (0.1-0.3).
-Similarly, local cherts show lower blank to core ratios (\<7), while the non-local cherts show higher ratios (\>9).
-Similar to the Proto-Solutrean, chalcedony cores were also not identified in the studied sample (@tbl-general-solutrean).
-
-Unlike the other assemblages, there seems to be no clear difference in the measurements and weights of artefacts between local and non-local cherts (@fig-solutrean-boxplots and table s++ in SOM ++).
-The exception are elongated blanks, with show differences between T2 and T6, with T6 showing smaller width and lengths (<35 mm and 15 mm, respectively), than the local type.
-Similarly, elongated blank weights are smaller in T6 (ranging between 0-5-2 gr).
-Core weights are also slightly lower in T6 (means of 16.5 gr) than in T2, although the sample is comparatively small (@tbl-general-solutrean).
-
 ```{r}
@@
 
-# Discussion
-
-As mentioned before, this study aimed to test two expectations related to the use and knapping of local and non-local raw materials, in specific, chert and chalcedony, and their relation to settlement and mobility strategies, across several UP occupations.
-As such, our method derived the following hypotheses to test: a) use lives derived from knapping strategies; b) reduction intensity and tools-to-debitage ratios; c) core size and blank-to-core ratios; d) reduction intensity and tool reworking; e) tool typology diversity and specialization (@tbl-expectations).
-
-To test these hypotheses, the following parameters were evaluated for each assemblage and compared between chert types: P1) frequency of retouched tools; P2) frequency of maintenance products; P3) debitage sizes; P4) tools-to-debitage ratios ; P5) core sizes; P6) blank-to-core ratios; P7) retouched tool sizes; P8) tool diversity; P9) tool specialization.
-These parameters and their main results can be found in table @tbl-results-sum, and can be summarised as follows.
+In the Gravettian assemblage, non-local chert blanks (P3) are smaller than local chert blanks, especially regarding length (~ <25 mm) and weight (<5 gr), with statistically significant differences (p value of 0.00009*** and 0.012, respectively).
+The local chert blanks, although mostly concentrated below 25 mm width and length and with weights below 5 gr, have the largest number of outliers and variability in terms of size.
+The same patterns can be seen in the Proto-Solutrean assemblage, where length and weight also show statistically significant differences between chert types (0.046* and 0.004), with non-local types showing smaller values.
+In comparison, in the Solutrean assemblage, only non-local chert blank length seems to show statistically significant smaller values (p value of 0.0005**) than local chert blanks, being mostly concentrated below ~30 mm, while local chert blank lengths are more variable.
+While width seems to be more variable in both local and non-local chert blanks, the statistical analysis shows that non-local chert width is significantly higher than local chert blanks (p value of 0.02).
+
+Regarding cores (P5), the differences in sizes and weights seem between local and non-local cherts seem to be less obvious than those observed in blanks.
+In the Gravettian assemblage, there is a cluster of non-local chert cores below the 25 mm width and length, while in general, local chert cores seem to be more variable.
+Especially regarding weight, T6 cores (non-local) show lighter cores (~12 gr means) compared to local cherts.
+In the Solutrean assemblage, weight is the only measurement where differences can be seen between local and non-local chert cores; similarly to the Gravettian assemblage, T6 cores show slightly smaller cores (~18 gr means) than local cherts (~22 gr means).
+Despite these differences, the statistical tests applied to the dimensions between local and non-local chert cores show no statistically significant differences between the groups (p > 0.05).
+The Proto-Solutrean assemblage shows a small sample of non-local chert cores (n=4), not allowing a clear identification of patterns between the groups; alike the other assemblage however, local chert cores show a high range of sizes (~15-50 mm width and length) and weights (~3-85 gr).
+
+Following the analysis parameters previously described, the summarised results for each parameter can be found in @tbl-results-sum, and can be summarised as follows:
 
@@
 
-```{r}
-#| label: tbl-results-sum
-#| tbl-cap: "Summarised results for analysis parameters. Gr: Gravettian; Pr: Proto-Solutrean; So: Solutrean; Exp: Expectations; L: Local; NL: Non-local."
-
-p_results <- readr::read_csv(here::here('analysis/data/tables/results_sum.csv'))
-
-knitr::kable(head(p_results))
-
-```
-
 The Proto-Solutrean assemblages show a smaller presence of retouched tools than the Gravettian, and within the assemblage, retouched tools (P1) are slightly more frequent in non-local cherts.
@@
 Both tool diversity (P8) and tool specialization (P9) are higher in local cherts.
+
+```{r}
+#| label: tbl-results-sum
+#| tbl-cap: "Summarised results for analysis parameters. Gr: Gravettian; Pr: Proto-Solutrean; So: Solutrean; Exp: Expectations; L: Local; NL: Non-local."
+
+p_results <- readr::read_csv(here::here('analysis/data/tables/results_sum.csv'))
+
+knitr::kable(head(p_results))
+
+```
+
+# Discussion
 

### 2025-03-01 analysis/Supplementary/supplementary_tables.qmd
Macbook daily work session. Results in main document corrected and rewritten for emphasis on main patterns and comparison between groups. Supplementary document revised and corrected.

@@
     RETOUCHEDPIECETYPOLOGY == "Shouldered point fragment" ~ "Shouldered point",
+    RETOUCHEDPIECETYPOLOGY == "Retouched blade" ~ "Retouched blank",
     TRUE ~ RETOUCHEDPIECETYPOLOGY))
@@
 @tbl-general-gravettian shows the number and percentages of technological classes per chert type.
-Local cherts show similar patterns between them, with T2 and T5 being the types with higher numbers of artefacts (n=110 and n=72, respectively).
-For all local cherts, blanks (including blank fragments) represent more than 50% of the artefacts.
-Cores and core maintenance products are only present in T2 and T5, although in small frequencies: for T2, cores represent 13.6% (n=15 including core fragments) of the group and core preparation and maintenance products represent 1.8% (n=2).
-For T5 only 4 core fragments and 1 core preparation and maintenance product are present, all frequencies between \~5-1%.
-Retouched tools seem to be especially relevant in T2 (n= 20, \~18% including retouched piece fragments), making them the most frequent class after blanks.
-T5 shows similar percentages of retouched tools and shatter (\~12% and \~14%, respectively).
+For local cherts, blanks (including blank fragments) represent more than 50% of the artefacts.
+Cores and core maintenance products are only present in small frequencies: <10% and <2%, respectively.
+Retouched tools seem to be especially relevant in local T2-5 cherts (~16%), making them the most frequent class after blanks and blank fragments, followed by shatter (~10%).
 T1 (chalcedony) shows similar patterns to other local cherts, although with higher percentages of shatter (34.3%).
@@
 From all chert types, T6 shows the highest numbers of retouched pieces (n=58, `r round(58*100/170, 2)`%).
-Regarding typologies, 15 types of retouched tools were identified (excluding retouched tool fragments).
-All these typologies were found in the non-local chert group, while only 7 retouched tool typologies were identified in local cherts/chalcedony.
+Regarding typologies, 9 types of retouched tools were identified (excluding retouched tool fragments).
+All these typologies were found in the non-local chert group, while only 5 retouched tool typologies were identified in local cherts/chalcedony.
 The most common typologies identified in the assemblage (and for all chert types) were burins (n=52), endscrappers (n=17), retouched flakes (n=30) and splintered pieces (n=26).
@@
 
-The final lithic sample from the Proto-Solutrean assemblage, consisting of the levels 4E and top of level 5, consists of 635 artifacts.
+The final lithic sample from the Proto-Solutrean assemblage (levels 4E and top of level 5), consists of 635 artifacts.
 Of these, 167 (`r round(167*100/635, 2)`%) have an indeterminate chert type.
-Local cherts correspond to `r 32+62+32+158+51` artefacts (`r round((32+62+32+158+51)*100/635, 2)`%), with T4 being the most present local chert (n=158 or `r round(158*100/(32+62+32+158+51), 2)`% of the local types).
+Local cherts correspond to `r 32+62+32+158+51` artefacts (`r round((32+62+32+158+51)*100/635, 2)`%), with T4 being the most present local chert as seen in previous works (Belmiro et al., 2025).
 Non-local types are present in smaller numbers (n=`r 60+47+13+9+1` or `r round((60+47+13+9+1)*100/635, 2)`%), with T6 and T7 being the most relevant (n=60 and n=47, respectively), and the TL group includes 3 artefacts, corresponding to 2 types of TL cherts (Table s++ in SOM ++).
@@
 When crossing the lithic class with chert type (@tbl-general-proto), we see that the local types show high frequency of blanks (\>50%, including blank fragments), followed by shatter (\~11-27%).
-Cores are present in small numbers (n=`r 4+5+13+6`), corresponding to ~9-12% of the classes in each local type, with the exception of T1 (chalcedony), with no cores.
+Cores are present in small numbers (n=`r 4+5+13+6`), corresponding to ~10% of the classes in local types, with the exception of T1 (chalcedony), with no cores.
 Core preparation and maintenance products are absent in all local cherts.
@@
 A total of `r 1+10+2+8+5` (`r round((1+10+2+8+5)*100/65, 2)`%) retouched tools correspond to local cherts, while `r 10+9+3+1+1` (`r round((10+9+3+1+1)*100/65, 2)`%) correspond to non-local and TL cherts.
-As seen in @tbl-retouch-proto, 14 different types of retouched tools were identified.
-Endscrappers (26.2%), splintered pieces (17%), burins (12.3%) and retouched flakes (7.7%) were the most prevalent.
-Local raw materials show 9 different typologies, with Vale Comprido points, Solutrean retouch and notches being present only in the local types, even if in small numbers.
-Non-local cherts show 10 different typologies of retouched tools.
-Perforator-endscrapper, truncation, backed bladelet and retouched blades typologies can only be found in non-local varieties, albeit also in small numbers.
+As seen in @tbl-retouch-proto, 11 different types of retouched tools were identified.
+Endscrappers (n=18), retouched blanks (n=14), splintered pieces (n=11) and burins (n=8) were the most prevalent.
+Local raw materials show 8 different typologies, with Vale Comprido points, Solutrean retouch and notches being present only in the local types, even if in small numbers.
+Non-local cherts show 8 different typologies of retouched tools.
+Perforator-endscrapper, truncation and backed bladelet typologies can only be found in non-local varieties, albeit also in small numbers.
 
@@
 Of these, 201 (`r round(201*100/914, 2)`%) correspond to the indeterminate chert group.
-Local cherts correspond to `r 16+506+4+10+30` (`r round((16+506+4+10+30)*100/914, 2)`%) artefacts, mostly from T2 (n=506).
+Local cherts correspond to `r 16+506+4+10+30` (`r round((16+506+4+10+30)*100/914, 2)`%) artefacts, mostly from T2 as seen in previous works (Belmiro et al., 2025).
 Non-local cherts (including the TL group) correspond to `r 86+9+30+3+1+17+1` (`r round((86+9+30+3+1+17+1)*100/914, 2)`%) artefacts, mostly from T6 (n=86).
 
-When crossing the class data with type of raw material (@tbl-general-solutrean), we see that for local chert, especially T2, the majority of the debitage is composed of flakes and elongated blanks (83.9%).
+When crossing the class data with type of raw material (@tbl-general-solutrean), we see that for local chert, the majority of the debitage is composed of blanks (83.9%).
 Cores are also present (10.5%), alongside a small percentage of core preparation and maintenance products (1.3%).
@@
 # Blank-to-core ratios
-
 ```{r}
@@
   group_by(RMTYPE, CLASS) %>%
+  mutate(RMTYPE = factor(RMTYPE, levels = c("T1", "T2", "T3", "T4", "T5", "T6","T7", 
+                                        "T8", "T9", "T10", "T11", "TL"))) |>
   summarise(n=sum(n)) %>% 
@@
   group_by(RMTYPE, CLASS) %>%
+  mutate(RMTYPE = factor(RMTYPE, levels = c("T1", "T2", "T3", "T4", "T5", "T6","T7", 
+                                        "T8", "T9", "T10", "TL"))) |>
   summarise(n=sum(n)) %>% 
@@
 # Tool-to-debitage ratios
-
 ```{r}
@@
   group_by(RMTYPE, CLASS) %>%
+   mutate(RMTYPE = factor(RMTYPE, levels = c("T1", "T2", "T3", "T4", "T5", "T6","T7", 
+                                        "T8", "T9", "T10", "TL"))) |>
   summarise(n=sum(n)) %>% 
@@
 # Measurements
-
 ```{r}
@@
   group_by(RMTYPE, CLASS) |> 
+  mutate(RMTYPE = factor(RMTYPE, levels = c("T1", "T2-5", "T6","T7", 
+                                        "T8", "T9", "T10", "T11", "TL"))) |>
   summarise(across(c(MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd)))
@@
   group_by(RMTYPE, CLASS) |> 
+  mutate(RMTYPE = factor(RMTYPE, levels = c("T1", "T2-5", "T6","T7", 
+                                        "T8", "T9", "T10", "TL"))) |>
   summarise(across(c(MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd)))
@@
   group_by(RMTYPE, CLASS) |>
+  mutate(RMTYPE = factor(RMTYPE, levels = c("T1", "T2-5", "T6","T7", 
+                                        "T8", "T9", "T10", "TL"))) |>
   summarise(across(c(MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd)))
@@
 #|  - "Stats for cores"
-#| layout-nrow: 4
+#| layout-nrow: 2
 

### 2025-03-01 analysis/paper/paper.qmd
Macbook daily work session. Results in main document corrected and rewritten for emphasis on main patterns and comparison between groups. Supplementary document revised and corrected.

@@
 To test these hypotheses, we evaluated several parameters for each dataset and assemblage and compared the results between chert types and assemblages.
-The parameters evaluated in the current study were: P1) frequency of retouched tools; P2) frequency of maintenance products; P3) debitage sizes; P4) tools-to-debitage ratios; P5) core sizes; P6) blank-to-core ratios; P7) retouched tool sizes; P8) tool diversity; P9) tool specialization.
-Debitage, core and retouched tool sizes were analysed through descriptive statistics (mean, sd, min and max, tables available in the SOM\*\*) and plotted through boxplots.
+The parameters evaluated in the current study were: P1) frequency of retouched tools; P2) frequency of cores; P3) frequency of maintenance products; P4) debitage sizes; P5) tools-to-debitage ratios; P6) core sizes; P7) blank-to-core ratios; P8) tool diversity; P9) tool specialization.
+Debitage and core sizes were analysed through descriptive statistics (mean and sd tables available in the SOM\*\*) and plotted through boxplots and through non-parametric statistical analyses (Mann-Whitney U test).
 The ratios were obtained through the division of each group (e.g., tools by debitage and blanks by core), by chert type and assemblage.
@@
 
+The results are presented following the analysis parameters, focusing on the comparison between wider chert groups (local and non-local cherts) and between assemblages.
+A complete description of the assemblages and detailed tables for each analysis presented in this study can be found in the supplementary materials (SOM ++).
+
 In all assemblages, blanks and blank fragments are the most frequent class of lithic artefacts and in all chert types (>40% of the assemblages).
 Cores are, as expected, found in smaller percentages in all assemblages (~3-8%).
-Cores are present in local cherts in larger numbers than in non-local types (P++); their presence is noted in almost all non-local cherts in all assemblages, except for the TL group.
+Cores are present in local cherts in larger numbers than in non-local types (P2); their presence is noted in almost all non-local cherts in all assemblages, except for the TL group.
 In the Proto-Solutrean assemblage however, non-local chert core presence is small and limited to T7-9 (n=4 complete cores and n=2 core fragments).
-Albeit in very small numbers in all assemblages (n<10), core preparation and maintenance products (P2) are also present in local and non-local cherts, especially in the non-local types which show the larger sample number such as T6-7.
+Albeit in very small numbers in all assemblages (n<10), core preparation and maintenance products (P3) are also present in local and non-local cherts, especially in the non-local types which show the larger sample number such as T6-7.
 The presence of knapping by-products such as shatter is also seen in both local and non-local cherts in the Gravettian and Proto-Solutrean assemblages.
@@
 
-The relation between blank and core numbers was explored through a blank-to-core ratio (P6; Figure 1 and table s++).
+The relation between blank and core numbers was explored through a blank-to-core ratio (P7; Figure 1 and table s++).
 In all assemblages, local cherts have generally smaller ratios than non-local cherts, although sample sizes for cores are variable (graph ++).
@@
 
-Retouched tools were further explored through the tools-to-debitage ratio (P4; Figure 2 and table s++).
+Retouched tools were further explored through the tools-to-debitage ratio (P5; Figure 2 and table s++).
 In general, the result show that non-local chert types have higher retouch-to-debitage ratios than local types, albeit with differences between assemblages.
@@
 
-The sizes of blanks (P3) and cores (P5) are explored through scatterplots of width and length (Figure 3 a) and boxplots for weight (Figure 3 b); complete tables with descriptive statistics and non-parametric statistical test results can also be found in SOM ++, table s++.
+The sizes of blanks (P4) and cores (P6) are explored through scatterplots of width and length (Figure 3 a) and boxplots for weight (Figure 3 b); complete tables with descriptive statistics and non-parametric statistical test results can also be found in SOM ++, table s++.
 The scatterplots show different patterns in the several analysed classes.
@@
 
-ggpubr::ggarrange(gravettian_scatter, gravettian_stats_weight, labels = c("a", "b"), ncol = 2, common.legend = TRUE, legend = "bottom")
+ggpubr::ggarrange(gravettian_scatter, gravettian_stats_weight, labels = c("a", "b"), nrow = 2, common.legend = TRUE, legend = "bottom")
 
@@
 
-ggpubr::ggarrange(proto_scatter, proto_stats_weight, labels = c("a", "b"), ncol = 2, common.legend = TRUE, legend = "bottom")
+ggpubr::ggarrange(proto_scatter, proto_stats_weight, labels = c("a", "b"), nrow = 2, common.legend = TRUE, legend = "bottom")
 
@@
 
-ggpubr::ggarrange(solutrean_scatter, solutrean_stats_weight, labels = c("a", "b"), ncol = 2, common.legend = TRUE, legend = "bottom")
-
-```
-
-In the Gravettian assemblage, non-local chert blanks (P3) are smaller than local chert blanks, especially regarding length (~ <25 mm) and weight (<5 gr), with statistically significant differences (p value of 0.00009*** and 0.012, respectively).
+ggpubr::ggarrange(solutrean_scatter, solutrean_stats_weight, labels = c("a", "b"), nrow = 2, common.legend = TRUE, legend = "bottom")
+
+```
+
+In the Gravettian assemblage, non-local chert blanks (P4) are smaller than local chert blanks, especially regarding length (~ <25 mm) and weight (<5 gr), with statistically significant differences (p value of 0.00009*** and 0.012, respectively).
 The local chert blanks, although mostly concentrated below 25 mm width and length and with weights below 5 gr, have the largest number of outliers and variability in terms of size.
@@
 
-Regarding cores (P5), the differences in sizes and weights seem between local and non-local cherts seem to be less obvious than those observed in blanks.
+Regarding cores (P6), the differences in sizes and weights seem between local and non-local cherts seem to be less obvious than those observed in blanks.
 In the Gravettian assemblage, there is a cluster of non-local chert cores below the 25 mm width and length, while in general, local chert cores seem to be more variable.
@@
 
-In the Gravettian assemblage, retouched tools (P1) are present in higher frequencies in the non-local cherts, as well as in core preparation and maintenance products (P2).
-Debitage sizes (P3) show smaller sizes and weights for non-local cherts.
-Tools-to-debitage ratios (P4) are significantly higher in non-local cherts.
-Core sizes (P5) are slightly smaller in T6, a non-local chert, and the blank-to-core ratios (P6) are higher in non-local cherts.
-Retouched sizes (P7) show no relevant differences between local and non-local cherts; however, tool diversity (P8) is higher in non-local cherts, although with high frequency of less specialised tools such as notches and retouched tools, alongside hunting implements such as backed bladelets (P9).
+In the Gravettian assemblage, retouched tools (P1) are present in higher frequencies in the non-local cherts, as well as in core preparation and maintenance products (P3).
+Cores are present in even in non-local chert types (P2).
+Debitage sizes (P4) show smaller sizes and weights for non-local cherts.
+Tools-to-debitage ratios (P5) are significantly higher in non-local cherts.
+Core sizes (P6) are slightly smaller in T6, a non-local chert, and the blank-to-core ratios (P7) are higher in non-local cherts.
+Tool diversity (P8) is higher in non-local cherts, although with high frequency of less specialised tools such as notches and retouched tools, alongside hunting implements such as backed bladelets (P9).
 The high frequency of burins seen in this assemblage may be related to their use as cores, representing a knapping strategy more than the presence of retouch.
@@
 The Proto-Solutrean assemblages show a smaller presence of retouched tools than the Gravettian, and within the assemblage, retouched tools (P1) are slightly more frequent in non-local cherts.
-Core preparation and maintenance products (P2) are only present in non-local cherts, albeit in very small numbers (n=2).
-Blank sizes (P3), core sizes (P5), and retouched tool sizes (P7) are slightly smaller in non-local cherts.
-Regarding the ratios, tool-to-debitage ratios (P4) are higher in non-local cherts, albeit smaller than those seen in the Gravettian assemblage, and blank-to-core ratios (P6) are also higher in non-local cherts.
+Cores of non-local chert types are found in very small numbers (P2), although knapping is likely occurring on site.
+Core preparation and maintenance products (P3) are only present in non-local cherts, albeit in very small numbers (n=2).
+Blank sizes (P4) and core sizes (P6) are slightly smaller in non-local cherts.
+Regarding the ratios, tool-to-debitage ratios (P5) are higher in non-local cherts, albeit smaller than those seen in the Gravettian assemblage, and blank-to-core ratios (P7) are also higher in non-local cherts.
 Little differences were observed in tool diversity (P8) and specialisation (P9) between local and non-local cherts.
@@
 In general, the retouch frequency is lower than that observed in the previous assemblages.
-Debitage (P3), core (P5), and retouched tool (P7) sizes show no differences between local and non-local cherts, except for elongated blanks which are smaller in T6 and bigger in T2, and for slightly smaller cores seen in T6, although with high standard deviations.
-Similarly to the Proto-Solutrean, tools-to-debitage ratios (P4) and blank-to-core ratios (P6) are slightly higher in non-local cherts.
+Cores are present in relevant quantities in non-local chert types (P2).
+Debitage (P4) mass is slightly smaller in non-local cherts, although with larger widths.
+Core (P6) sizes show slightly smaller values for T6.
+Similarly to the Proto-Solutrean, tools-to-debitage ratios (P5) and blank-to-core ratios (P7) are slightly higher in non-local cherts.
 Both tool diversity (P8) and tool specialization (P9) are higher in local cherts.

### 2025-03-01 analysis/paper/paper.qmd
Revised and rewritten introduction

@@
 ---
-title: "Placeholder for article 3"
+title: "From Stone to Tool: How Raw Materials Influenced Upper Paleolithic Technology in Southwestern Iberia (Vale Boi)"
 author:
-  - Jane Doe:
+  - Joana Belmiro:
       correspondence: "yes"
-      email: janedoe@fosg.org
+      email: jfbelmiro@ualg.pt
       orcid: 0000-0003-1689-0557
       institute:
-        - fosg
-        - fop
-  - John Q. Doe:
-      institute: fosg
+        - ICArEHB
+  - Nuno Bicho:
+      institute: ICArEHB
       orcid: 0000-0003-1689-0558
-  - Peder Ås:
-      institute: fosg
+  - Xavier Terradas:
+      institute: CSIC
       orcid: 0000-0003-1689-0559
-  - Juan Pérez:
+  - João Cascalheira:
       orcid: 0000-0003-1689-0551
+      institute: ICArEHB
 institute:
-        - name: Acme Corporation
-  - Max Mustermann:
-      orcid: 0000-0003-1689-0552
-institute:
-  - fosg:
-      name: Formatting Open Science Group
-      address: 23 Science Street, Eureka, Mississippi, USA
-  - fop: Federation of Planets
+  - ICArEHB:
+      name: Interdisciplinary Center for Archaeology and Human Evolution
+      address: 
+  - CSIC: CSIC
 title-block-published: "Last updated"  
@@
 
+The Iberian Peninsula has been a key region for the study of human adaptations and cultural transitions in Archaeology.
+Located in the westernmost tip of Europe and with a generally stable climate even during the coldest periods punctuating the Late Pleistocene, this region has yielded some of the most significant glacial refugia in Europe [@hewitt2000; @gómez2007; @gonzález-sampériz2010; @carvalho2022].
+
+For this reason, the Iberian Peninsula has been the focal point of several discussions regarding Late Pleistocene human adaptations [@finlayson2006; @zilhão2017; @cascalheira2017].
+
+A particularly good example is the region's role in the discussion of Upper Paleolithic technocomplexes transitions.
+Previous studies have discussed the territory's potential as a refugium during cold and harsh climatic conditions [@gómez2007; @rodríguez-sánchez2010; @jennings2011].
+Other studies have expanded upon this notion of climatic refugia during harsh climatic events, to understand the Iberian Peninsula as a long-term eco-cultural refugia [@cascalheira2017].
+Using this framework, this territory would consist of several ecological niches, consistently used through time, possibly due to the stability in the richness and variety of resources.
+This continuous use would then create long-term regional adaptive structures, which when correlated with the ecological niches, @cascalheira2017 have referred to as eco-cultural niches.
+In fact, it seems that a large number of caves and rockshelters in Iberia are multi-layered, giving validity to the aforementioned framework [@schmidt2012].
+These eco-cultural niches provide an exceptional opportunity to understand long-term dynamics regarding biotic and abiotic resource exploitation since they can provide details on how human populations maintained or changed their adaptive systems when facing environmental changes, and cultural and social transformations or constraints.
+
+More specifically, southwestern Iberia has been shown to be a key eco-cultural niche to study cultural transitions and the role of such a peripheral coastal region within hunter-gatherer social networks [@cascalheira2017].
+This has been mostly due to the archaeological site of Vale Boi, located in southwesternmost Iberia and characterized by a nearly complete UP sequence (++).
+This site is crucial for understanding Upper Paleolithic (UP) adaptations, as it is currently the only site in southwestern Iberia with a long-term occupation spanning most of the UP, enabling the exploration of cultural and technological trends over time (Cascalheira et al.
+2017).
+
+For example, Vale Boi has provided valuable insights into hunter-gatherer subsistence strategies and technological organisation (Belmiro et al.
+2021; Bicho et al.
+2013; Cascalheira et al.
+2017; Cascalheira 2019; Horta et al.
+2019; Manne et al.
+2012; Marreiros et al.
+2016), contributing to a better understanding of territory exploitation and interactions with other regions of Iberia throughout the UP.
+Especially due to lithic technology studies from this site, the region has been shown to have played a shifting and complex role within the larger territory.
+
+During the Gravettian, the peripheral nature of the region studies of Gravettian occupations reveal marked regional differences, suggesting more limited movement of people and information in southwestern Iberia (Marreiros and Bicho, 2013).
+In comparison, Proto-Solutrean evidence indicates technological similarities and close contact with central Portugal (Belmiro et al., 2021).
+The Solutrean period stands out as a particularly dynamic phase at Vale Boi, during which the site served as a key connection point between central Portugal and southern Spain, facilitating social networks and cultural exchanges (Cascalheira, 2013; Cascalheira et al., 2017).
+
+Expanding on these findings, recent research has further refined our understanding of mobility and social networks through lithic raw material studies at the site.
+Our team’s recent work (Belmiro et al., 2025) identified non-local cherts originating from central Portugal (>250 km from Vale Boi) and southern Spain (>200 km from Vale Boi).
+The proportions of these materials varied across the UP, with earlier occupations (Gravettian) showing high frequencies of non-local chert, while later ones (Proto-Solutrean and Solutrean) exhibited a marked decline.
+These findings suggest the establishment and maintenance of extensive social networks connecting southwestern Portugal to Southern spain and central Portugal as early as the beginning of the UP.
+
+These results, particularly those related to the Gravettian, were interpreted as outcomes of residential mobility rather than direct social network influences, though further investigation is required.
+
+Despite the insights gained from previous studies, the relationship between lithic technology and raw material use at Vale Boi remains largely uncorrelated.
+While broader research highlights the complexity of lithic technological organisation, a more integrated approach is needed to fully understand how raw material provisioning influenced technological strategies at the site and what this can tell us about how hunter-gatherer groups organized, explored the territory and connected to others across Iberia.
+To address this, our goal is to combine lithic technological and chert raw material studies to better characterize hunter-gatherer technological and social adaptations throughout the UP in southwestern Iberia.
+To do so, we create and test expectations built upon previous archaeological and ethnographic studies of the technological organization of hunter-gatherers.
+
+## Theoretical framework and objectives
+
 Lithic technological organisation has long been a central focus in archaeology for the insights it offers into past human behaviour and lifeways.
@@
 
-Similar studies have also been applied to lithic assemblages from prehistoric archaeological sites in the Iberian Peninsula (e.g., Aubry et al.
-2016; Aubry and Igreja 2009; García-Rojas et al.
-2021; Gómez de Soler et al.
-2020; Herrero-Alonso et al.
-2020; Matias 2016; Nocete et al.
-2005; Ortega 2003; Pereira, Andrade, et al.
-2016; Pereira et al.
-2021, 2022; Ramacciotti et al.
-2022; Rodríguez et al.
-2011; Sánchez de la Torre et al.
-2023; Soto 2016), and contributed to understanding how different groups explored the available lithic resources, aided in identifying extensive movements and social networks throughout the Upper Palaeolithic (e.g., Aubry et al.
-2004, 2016; Corchón Rodríguez et al.
-2016; Hermida et al.
-2016; Marreiros et al.
-2016; Sánchez de la Torre et al.
-2023).
-
-Until recently, the study of raw materials in southwestern Iberia was limited, and knowledge about mobility and social networks was limited to technological studies, derived from the archaeological site of Vale Boi.
-This site is crucial for understanding Upper Paleolithic (UP) adaptations, as it is currently the only site in southwestern Iberia with a long-term occupation spanning most of the UP, enabling the exploration of cultural and technological trends over time (Cascalheira et al.
-2017).
-For example, Vale Boi has provided valuable insights into hunter-gatherer subsistence strategies, technological organisation (Belmiro et al.
-2021; Nuno Bicho et al.
-2013; Cascalheira et al.
-2017; Cascalheira 2019; Horta et al.
-2019; Manne et al.
-2012; Marreiros et al.
-2016).
-This also contributed to understanding different patterns of territory exploitation and contact with other regions of Iberia throughout the UP.
-
-During the Solutrean period, Vale Boi served as a key connection point between central Portugal and southern Spain, facilitating social networks and cultural exchanges (Cascalheira, 2013; Cascalheira et al., 2017).
-Proto-Solutrean evidence also indicates technological similarities and close contact with central Portugal (Belmiro et al., 2021).
-In contrast, studies of Gravettian occupations reveal marked regional differences, suggesting more limited movement of people and information in southwestern Iberia (Marreiros and Bicho, 2013).
-
-Recent research has expanded our understanding of mobility and social networks through lithic raw material studies.
-Our team’s recent work (Belmiro et al., 2025) identified non-local cherts originating from central Portugal (\>250 km from Vale Boi) and souuthern Spain (\>200 km from Vale Boi).
-The proportions of these materials varied across the UP, with earlier occupations (Gravettian) showing high frequencies of non-local chert, while later ones (Proto-Solutrean and Solutrean) exhibited a marked decline.
-These findings suggest the establishment and maintenance of extensive social networks connecting southwestern Portugal to Southern spain and central Portugal as early as the beginning of the UP.
-
-These results, particularly those related to the Gravettian, were interpreted as outcomes of residential mobility rather than direct social network influences, though further investigation is required.
 While distance to source and raw material proportions offer valuable insights, they represent only one aspect of raw material procurement and management.
@@
 For example, the transport of formal cores (e.g., blade cores) and bifaces used as cores have been interpreted as lightweight and reconfigurable raw material storage strategies (Clark, 1987; Kelly, 1988).
-
 Formal tools and the degree of tool maintenance have also been associated with planned preparation, efficient use, and transportability, reflecting their connection to mobile settlement strategies, short-term site occupations (Torrence, 1983) and the regional distribution of raw materials (Bamforth, 1986).
@@
 When knapped on-site, non-local materials also tend to exhibit higher proportions of retouched tools compared to local materials.
-
 Tool variability and size are also impacted by mobility and uncertainty regarding raw material availability; increased group mobility may create a pattern of limited tool variability, and tools may become less specialised and easily reconfigurable, such as retouched blanks (Shott, 1986; Odell, 1981; Siegel, 1984).
@@
 
-These studies highlight significant variability in lithic assemblages and raw material proportions, reflecting different facets of hunter-gatherer technological organisation.
+Altogether, different studies highlight significant variability in lithic assemblages and raw material proportions, reflecting different facets of hunter-gatherer technological organisation.
 This variability appears between behaviourally and ethnically similar sites (Bamforth, 1986), across distinct chrono-cultural periods, and even within a single site among various raw materials.
-To fully understand past hunter-gatherer behaviours, organisation, and lifeways, it is crucial to examine all stages of lithic production—procurement, distribution, use, maintenance, and discard—through interconnected approaches that combine lithic technology and raw material analyses.
-
-Building on the studies mentioned above and existing (yet currently uncorrelated) data on UP lithic technology and raw material analyses at Vale Boi, we outline the expectations and objectives for the current study.
-
-## Aims
+To fully understand past hunter-gatherer behaviours, organisation, and lifeways, it is crucial to examine all stages of lithic production - procurement, distribution, use, maintenance, and discard - through interconnected approaches that combine lithic technology and raw material analyses.
 
@@
 We focus on the technological characteristics of the two broader groups of cherts identified at the site: local and non-local, based on the notion that local raw materials will show technological patterns different from those which are non-local.
-Expanding upon our previous works on chert types present throughout the UP sequence in the Terrace and Shelter area of Vale Boi, and their possible interpretations, and following the previously mentioned studies and models about raw material management and lithic technology organisation focusing mainly on the different management of local and non-local raw materials, associated with mobility and occupation length, we derive two main expectations (fig.
-1):
-
-1.
-Local raw materials are expected to present a) low use-lives, with knapping strategies characterised by manufacture, use and discard; b) a decreased reduction intensity and low tools to debitage ratio; c) larger cores and lower blank-to-core ratio; d) less extensive reduction and reworking of tools; e) higher tool typology diversity.
-This may be related to the abundance of local raw materials which leads to less intensive life-use extension, but also to the residential characteristics of the site and low mobility, which increases the reliance on local raw materials.
-
-2.
-Non-local raw materials are expected to present a) a majority of formal tools, in the case of individual provisioning or higher use-lives, with knapping strategies characterised by the manufacture, use and maintenance; b) an increased reduction intensity and high tools to debitage ratio; c) smaller cores and higher blank-to-core ratio; d) more extensive reduction and high presence of retouch and/or tool maintenance; e) lower tool diversity, characterised by less specialised tools, such as the presence of retouched blanks as multipurpose tools.
-This may be related to scarcer or poorer quality of raw materials which leads to the prepared transportation of lithic resources to a site, but also related to more mobile settlement patterns as well as short-term site occupations.
+Expanding upon our previous works on chert types present throughout the UP sequence in the Terrace and Shelter area of Vale Boi, and their possible interpretations, and following the previously mentioned studies and models about raw material management and lithic technology organisation focusing mainly on the different management of local and non-local raw materials, associated with mobility and occupation length, we derive several expectations regarding local and non-local raw materials that can be found summarized in @tbl-expectations.
 
@@
 In this sense, distance from source may have had little intrinsic meaning in the past, especially regarding non-local raw materials.
-There are several ways in which raw materials may be transported to a site, and the relationship between distance and cost may vary depending on how this takes place [@kuhn2004].
+There are several ways in which raw materials may be transported to a site, and the relationship between distance and cost may vary depending on how this takes place (Kuhn, 2004).
 When non-local raw materials are as abundant as local ones, the difference in the technological organization between these materials may be less obvious (Andrefsky, 2004).
 
-## Site descriptions
+# Materials and methods
+
+## Site description
 
@@
 
-# Materials and methods
+## Lithic analysis
 

### 2025-03-05 analysis/paper/paper.qmd
Backlog of several workdays. Discussion written. Introduction revised. Alterations to supporting tables in the main paper. Updated render.

@@
 The Iberian Peninsula has been a key region for the study of human adaptations and cultural transitions in Archaeology.
-Located in the westernmost tip of Europe and with a generally stable climate even during the coldest periods punctuating the Late Pleistocene, this region has yielded some of the most significant glacial refugia in Europe [@hewitt2000; @gómez2007; @gonzález-sampériz2010; @carvalho2022].
+Located in the westernmost tip of Europe and with a generally stable climate even during the coldest periods punctuating the Late Pleistocene, this region has yielded some of the most significant glacial refugia in Europe [@hewitt2000; @gómez2007; @gonzález-sampériz2010; @carvalho2022], and has been considered a long-term eco-cultural refugia [@cascalheira2017].
 
@@
 
-A particularly good example is the region's role in the discussion of Upper Paleolithic technocomplexes transitions.
-Previous studies have discussed the territory's potential as a refugium during cold and harsh climatic conditions [@gómez2007; @rodríguez-sánchez2010; @jennings2011].
-Other studies have expanded upon this notion of climatic refugia during harsh climatic events, to understand the Iberian Peninsula as a long-term eco-cultural refugia [@cascalheira2017].
-Using this framework, this territory would consist of several ecological niches, consistently used through time, possibly due to the stability in the richness and variety of resources.
-This continuous use would then create long-term regional adaptive structures, which when correlated with the ecological niches, @cascalheira2017 have referred to as eco-cultural niches.
-In fact, it seems that a large number of caves and rockshelters in Iberia are multi-layered, giving validity to the aforementioned framework [@schmidt2012].
-These eco-cultural niches provide an exceptional opportunity to understand long-term dynamics regarding biotic and abiotic resource exploitation since they can provide details on how human populations maintained or changed their adaptive systems when facing environmental changes, and cultural and social transformations or constraints.
+A particularly good example is the region’s role in the discussion of Upper Paleolithic technocomplexes transitions.
+This has been possible due to the existence of several ecological niches, consistently used through time, possibly due to the stability in the richness and variety of resources in Iberia.
+This continuous use would then create long-term regional adaptive structures, which when correlated with the ecological niches, have been referred to as eco-cultural niches [@cascalheira2017].
+These eco-cultural niches—which yielded multi-layered sites—.
+provide an exceptional opportunity to understand long-term dynamics regarding biotic and abiotic resource exploitation since they can provide details on how human populations maintained or changed their adaptive systems when facing cultural and social transformations.
 
@@
 Expanding on these findings, recent research has further refined our understanding of mobility and social networks through lithic raw material studies at the site.
-Our team’s recent work (Belmiro et al., 2025) identified non-local cherts originating from central Portugal (>250 km from Vale Boi) and southern Spain (>200 km from Vale Boi).
+Our team’s recent work (Belmiro et al., 2025) identified non-local cherts originating from central Portugal (\>250 km from Vale Boi) and southern Spain (\>200 km from Vale Boi).
 The proportions of these materials varied across the UP, with earlier occupations (Gravettian) showing high frequencies of non-local chert, while later ones (Proto-Solutrean and Solutrean) exhibited a marked decline.
@@
   labs(y = " ") +
-  scale_fill_manual(values = c("#D49B54", "#8B6969", "#CDAF95", "#2B4F60",
+  scale_fill_manual(name = "Class", values = c("#D49B54", "#8B6969", "#CDAF95", "#2B4F60",
                               "tomato4", "#798777"))
@@
   labs(x="Length (mm)", y="Max. Width (mm)") +
-  scale_fill_manual(values = c("#D49B54", "#8B6969", "#CDAF95", "#2B4F60",
+  scale_fill_manual(name = "Chert type", values = c("#D49B54", "#8B6969", "#CDAF95", "#2B4F60",
                               "tomato4", "#BDC7C9", "#798777"))
@@
   stat_summary(fun=mean, geom="point", shape=23) +
-  labs(x="Raw material types" ,y="Weight (gr)") +
+  labs(y="Weight (gr)") +
   #coord_cartesian(ylim =  c(0, 50)) +
@@
   theme(axis.text.x=element_blank(),
-        axis.ticks.x=element_blank())
+        axis.ticks.x=element_blank(), axis.title.x = element_blank())
 
@@
   labs(x="Length (mm)", y="Max. Width (mm)") +
-  scale_fill_manual(values = c("#D49B54", "#8B6969", "#CDAF95", "#2B4F60",
+  scale_fill_manual(name = "Chert type", values = c("#D49B54", "#8B6969", "#CDAF95", "#2B4F60",
                               "tomato4", "#BDD2B6", "#798777", "#BDC7C9"))
@@
   stat_summary(fun.y=mean, geom="point", shape=23) +
-  labs(x="Raw material types", y="Weight (gr)") +
+  labs(y="Weight (gr)") +
   #coord_cartesian(ylim =  c(0, 50)) +
@@
   labs(x="Length (mm)", y="Max. Width (mm)") +
-  scale_fill_manual(values = c("#D49B54", "#8B6969", "#CDAF95", "#2B4F60",
+  scale_fill_manual(name = "Chert type", values = c("#D49B54", "#8B6969", "#CDAF95", "#2B4F60",
                               "#8B7D6B", "#BDD2B6", "#798777", "#BDC7C9"))
@@
   stat_summary(fun.y=mean, geom="point", shape=23) +
-  labs(x= "Raw material types", y="Weight (gr)") +
+  labs(y="Weight (gr)") +
   #coord_cartesian(ylim =  c(0, 60)) +
@@
   theme(axis.text.x=element_blank(),
-        axis.ticks.x=element_blank())
-
-```
-
-The results are presented following the analysis parameters, focusing on the comparison between wider chert groups (local and non-local cherts) and between assemblages.
-A complete description of the assemblages and detailed tables for each analysis presented in this study can be found in the supplementary materials (SOM ++).
-
-In all assemblages, blanks and blank fragments are the most frequent class of lithic artefacts and in all chert types (>40% of the assemblages).
-Cores are, as expected, found in smaller percentages in all assemblages (~3-8%).
-Cores are present in local cherts in larger numbers than in non-local types (P2); their presence is noted in almost all non-local cherts in all assemblages, except for the TL group.
-In the Proto-Solutrean assemblage however, non-local chert core presence is small and limited to T7-9 (n=4 complete cores and n=2 core fragments).
-Albeit in very small numbers in all assemblages (n<10), core preparation and maintenance products (P3) are also present in local and non-local cherts, especially in the non-local types which show the larger sample number such as T6-7.
-The presence of knapping by-products such as shatter is also seen in both local and non-local cherts in the Gravettian and Proto-Solutrean assemblages.
-Retouched tools (P1) were identified in all chert types from all studied assemblages; in general, all assemblages show a similar pattern where local cherts show smaller percentages of retouched tools than non-local cherts.
-However, the assemblages show clear differences in the number of retouched tools.
-
-```{r}
-#| label: fig-general-graphs
-#| fig-cap: Class by chert type in the Gravettian (a), Proto-Solutrean (b) and Solutrean (c) assemblages.
-Labels within the plots correspond to the number of artefacts within each class.
-Only labels of n>3 are shown.
-
-ggpubr::ggarrange(barplot_gravettian, barplot_proto, barplot_solutrean, labels = c("a", "b", "c"), nrow = 3, common.legend = TRUE, legend = "right")
-
-```
-
-The Gravettian assemblage shows higher percentages of retouched tools (~21%) than the other assemblages (<10%).
-Detailed tables with tool typology and counts by chert type and assemblage can be found in the SOM (tables +++).
-Tool diversity (P8) is higher in non-local cherts in the Gravettian assemblage, although with high frequency of less specialised tools such as notches and retouched tools, alongside hunting implements such as backed bladelets (P9).
-The high frequency of burins seen in this assemblage may be related to their use as cores, representing a knapping strategy more than the presence of retouch.
-In comparison to the Gravettian, other assemblages show smaller numbers of retouched tools.
-While in the Proto-Solutrean no pattern was identified between local and non-local retouched tools, in the Solutrean assemblage both tool diversity (P8) and tool specialization (P9) are higher in local cherts.
-
-The relation between blank and core numbers was explored through a blank-to-core ratio (P7; Figure 1 and table s++).
-In all assemblages, local cherts have generally smaller ratios than non-local cherts, although sample sizes for cores are variable (graph ++).
-For the Gravettian assemblages, the results show that non-local cherts show ratios above 7 (with the exception of T10), while local cherts have ratios located below 4.
-The Solutrean assemblage also shows higher ratios for non-local cherts (>12), with local cherts showing ratios below 7.
-The Proto-Solutrean assemblage has most blank-to-core ratios below 10, except for non-local T7.
-However, non-local cores are poorly represented within the assemblage (table ++).
-
-```{r}
-#| label: fig-ratiobc
-#| fig-cap: Plotted ratios of blanks-to-core by assemblage and by chert type. All local cherts were grouped in a single category "Local", while non-local cherts maintained their individual IDs.
+        axis.ticks.x=element_blank(), axis.title.x = element_blank())
+
+```
+
+```{r}
+#| label: ratiobc
 
@@
 # Scatterplot with local vs non-local RM Types
-ratio_blankcore |>
+bc_ratio <- ratio_blankcore |>
   dplyr::mutate(Source = case_when(
@@
   coord_cartesian(clip = "off") +
+  theme(legend.position = "none", axis.title.x = element_blank()) +
   scale_colour_manual(values = c("#D49B54", "#8B6969", "#CDAF95", "#2B4F60",
@@
 
-Retouched tools were further explored through the tools-to-debitage ratio (P5; Figure 2 and table s++).
-In general, the result show that non-local chert types have higher retouch-to-debitage ratios than local types, albeit with differences between assemblages.
-The Gravettian assemblage shows the highest tool-to-debitage ratios compared to the remaining assemblages; non-local cherts show ratios above 0.9, while local cherts show values below 0.5.
-In the Proto-Solutrean non-local cherts show ratios between 0.3 to 0.5, while local cherts show ratios below 0.3.
-This pattern is less clear in the Solutrean assemblage since the ratios are the lowest of all assemblages.
-Non-local cherts show ratios above 0.1 (with exception of T8) while local cherts show ratios below 0.1 (with exception of the local chalcedony).
-
-```{r}
-#| label: fig-ratiotd
-#| fig-cap: Plotted ratios of tools-to-debitage by assemblage and by chert type. All local cherts were grouped in a single category "Local", while non-local cherts maintained their individual IDs.
+```{r}
+#| label: ratiotd
 
@@
 # Scatterplot with local vs non-local RM Types
-ratio_tooldebitage |>
+td_ratio <- ratio_tooldebitage |>
   dplyr::mutate(Source = case_when(
@@
   geom_point(size=3) +
-  scale_colour_manual(values = c("#D49B54", "#8B6969", "#CDAF95", "#2B4F60",
+  theme(legend.position = "bottom", axis.title.x = element_blank()) +
+  scale_colour_manual(name = "Chert type", values = c("#D49B54", "#8B6969", "#CDAF95", "#2B4F60",
                               "#BDD2B6", "#BDC7C9", "#798777"))
@@
 ```
+
+The results are presented following the analysis parameters, focusing on the comparison between wider chert groups (local and non-local cherts) and between assemblages.
+A complete description of the assemblages and detailed tables for each analysis presented in this study can be found in the supplementary materials (SOM ++).
+
+In all assemblages, blanks and blank fragments are the most frequent class of lithic artefacts and in all chert types (\>40% of the assemblages).
+Cores are, as expected, found in smaller percentages in all assemblages (\~3-8%).
+Cores are present in local cherts in larger numbers than in non-local types (P2); their presence is noted in almost all non-local cherts in all assemblages, except for the TL group.
+In the Proto-Solutrean assemblage however, non-local chert core presence is small and limited to T7-9 (n=4 complete cores and n=2 core fragments).
+Albeit in very small numbers in all assemblages (n\<10), core preparation and maintenance products (P3) are also present in local and non-local cherts, especially in the non-local types which show the larger sample number such as T6-7.
+The presence of knapping by-products such as shatter is also seen in both local and non-local cherts in the Gravettian and Proto-Solutrean assemblages.
+Retouched tools (P1) were identified in all chert types from all studied assemblages; in general, all assemblages show a similar pattern where local cherts show smaller percentages of retouched tools than non-local cherts.
+However, the assemblages show clear differences in the number of retouched tools.
+
+```{r}
+#| label: fig-general-graphs
+#| fig-cap: Class by chert type in the Gravettian (a), Proto-Solutrean (b) and Solutrean (c) assemblages.
+Labels within the plots correspond to the number of artefacts within each class.
+Only labels of n>3 are shown.
+
+ggpubr::ggarrange(barplot_gravettian, barplot_proto, barplot_solutrean, labels = c("a", "b", "c"), nrow = 3, common.legend = TRUE, legend = "right")
+
+```
+
+The Gravettian assemblage shows higher percentages of retouched tools (\~21%) than the other assemblages (\<10%).
+Detailed tables with tool typology and counts by chert type and assemblage can be found in the SOM (tables +++).
+Tool diversity (P8) is higher in non-local cherts in the Gravettian assemblage, although with high frequency of less specialised tools such as notches and retouched tools, alongside hunting implements such as backed bladelets (P9).
+The high frequency of burins seen in this assemblage may be related to their use as cores, representing a knapping strategy more than the presence of retouch.
+In comparison to the Gravettian, other assemblages show smaller numbers of retouched tools.
+While in the Proto-Solutrean no pattern was identified between local and non-local retouched tools, in the Solutrean assemblage both tool diversity (P8) and tool specialization (P9) are higher in local cherts.
+
+The relation between blank and core numbers was explored through a blank-to-core ratio (P7; Figure 1 and table s++).
+In all assemblages, local cherts have generally smaller ratios than non-local cherts, although sample sizes for cores are variable (graph ++).
+For the Gravettian assemblages, the results show that non-local cherts show ratios above 7 (with the exception of T10), while local cherts have ratios located below 4.
+The Solutrean assemblage also shows higher ratios for non-local cherts (\>12), with local cherts showing ratios below 7.
+The Proto-Solutrean assemblage has most blank-to-core ratios below 10, except for non-local T7.
+However, non-local cores are poorly represented within the assemblage (table ++).
+
+```{r}
+#| label: fig-ratios
+#| fig-cap: Plotted ratios of blanks-to-core (a) and tool-to-debitage (b) by assemblage (Gravettian, Proto-Solutrean and Solutrean) and by chert type.
+All local cherts were grouped in a single category "Local", while non-local cherts maintained their individual types.
+
+# Extract legend from plot B using ggpubr function since plot B has more categories than plot A
+legend_b <- as_ggplot(get_legend(td_ratio))  # Convert legend to a ggpubr-compatible plot
+td_ratio <- td_ratio + theme(legend.position = "none")
+ggpubr::ggarrange(bc_ratio, td_ratio, legend_b, labels = c("a", "b", ""), ncol = 1, heights = c(3, 3, 1))
+
+```
+
+Retouched tools were further explored through the tools-to-debitage ratio (P5; Figure 2 and table s++).
+In general, the result show that non-local chert types have higher retouch-to-debitage ratios than local types, albeit with differences between assemblages.
+The Gravettian assemblage shows the highest tool-to-debitage ratios compared to the remaining assemblages; non-local cherts show ratios above 0.9, while local cherts show values below 0.5.
+In the Proto-Solutrean non-local cherts show ratios between 0.3 to 0.5, while local cherts show ratios below 0.3.
+This pattern is less clear in the Solutrean assemblage since the ratios are the lowest of all assemblages.
+Non-local cherts show ratios above 0.1 (with exception of T8) while local cherts show ratios below 0.1 (with exception of the local chalcedony).
 
@@
 
-In the Gravettian assemblage, non-local chert blanks (P4) are smaller than local chert blanks, especially regarding length (~ <25 mm) and weight (<5 gr), with statistically significant differences (p value of 0.00009*** and 0.012, respectively).
+In the Gravettian assemblage, non-local chert blanks (P4) are smaller than local chert blanks, especially regarding length (\~ \<25 mm) and weight (\<5 gr), with statistically significant differences (p value of 0.00009\*\*\* and 0.012, respectively).
 The local chert blanks, although mostly concentrated below 25 mm width and length and with weights below 5 gr, have the largest number of outliers and variability in terms of size.
-The same patterns can be seen in the Proto-Solutrean assemblage, where length and weight also show statistically significant differences between chert types (0.046* and 0.004), with non-local types showing smaller values.
-In comparison, in the Solutrean assemblage, only non-local chert blank length seems to show statistically significant smaller values (p value of 0.0005**) than local chert blanks, being mostly concentrated below ~30 mm, while local chert blank lengths are more variable.
+The same patterns can be seen in the Proto-Solutrean assemblage, where length and weight also show statistically significant differences between chert types (0.046\* and 0.004), with non-local types showing smaller values.
+In comparison, in the Solutrean assemblage, only non-local chert blank length seems to show statistically significant smaller values (p value of 0.0005\*\*) than local chert blanks, being mostly concentrated below \~30 mm, while local chert blank lengths are more variable.
 While width seems to be more variable in both local and non-local chert blanks, the statistical analysis shows that non-local chert width is significantly higher than local chert blanks (p value of 0.02).
@@
 In the Gravettian assemblage, there is a cluster of non-local chert cores below the 25 mm width and length, while in general, local chert cores seem to be more variable.
-Especially regarding weight, T6 cores (non-local) show lighter cores (~12 gr means) compared to local cherts.
-In the Solutrean assemblage, weight is the only measurement where differences can be seen between local and non-local chert cores; similarly to the Gravettian assemblage, T6 cores show slightly smaller cores (~18 gr means) than local cherts (~22 gr means).
-Despite these differences, the statistical tests applied to the dimensions between local and non-local chert cores show no statistically significant differences between the groups (p > 0.05).
-The Proto-Solutrean assemblage shows a small sample of non-local chert cores (n=4), not allowing a clear identification of patterns between the groups; alike the other assemblage however, local chert cores show a high range of sizes (~15-50 mm width and length) and weights (~3-85 gr).
+Especially regarding weight, T6 cores (non-local) show lighter cores (\~12 gr means) compared to local cherts.
+In the Solutrean assemblage, weight is the only measurement where differences can be seen between local and non-local chert cores; similarly to the Gravettian assemblage, T6 cores show slightly smaller cores (\~18 gr means) than local cherts (\~22 gr means).
+Despite these differences, the statistical tests applied to the dimensions between local and non-local chert cores show no statistically significant differences between the groups (p \> 0.05).
+The Proto-Solutrean assemblage shows a small sample of non-local chert cores (n=4), not allowing a clear identification of patterns between the groups; alike the other assemblage however, local chert cores show a high range of sizes (\~15-50 mm width and length) and weights (\~3-85 gr).
 
@@
 #| label: tbl-results-sum
-#| tbl-cap: "Summarised results for analysis parameters. Gr: Gravettian; Pr: Proto-Solutrean; So: Solutrean; Exp: Expectations; L: Local; NL: Non-local."
+#| tbl-cap: "Summarised results for analysis parameters between grouped chert types (local and non-local) by assemblage."
 
@@
 
-knitr::kable(head(p_results))
+knitr::kable(p_results)
 
@@
 
-In general, the obtained results seem to fit the study expectations.
-The Gravettian assemblage seems to show agreement with most of the expectations (table **).
-Although ~50% of the assemblage is composed of non-local raw materials, making them as frequent as the local cherts, non-local cherts, especially T6 and T7 show differential techno-typological patterns from the local raw materials.
-Only in expectation e) there seems to be a difference from the expected outcome.
-
-In this sense, it would be expected a lower typological diversity from non-local cherts, since the distance to source and inferred mobility would assume that less specialized tools would be more adequate for several use-case scenarios; instead, our results show higher tool diversity in non-local cherts.
-However, this scenario also implies that tools are being transported as either less specialized tools or even blanks which are then maintained.
-Instead, our results also suggest that the long-distance raw materials, especially T6 and T7, are being knapped at the site.
-As such, the high diversity of tool typology may simply be a reflection of the transportation of volumes and their knapping on site, a possible preference for these raw materials for the production of retouched tools, and thus their high typological variability.
-Despite this high variability however, less specialized tools such as retouched blanks are still predominant in non-local cherts, as well as specialized tools such as backed bladelets, giving weight to the interpretation that given the possible ease of access to these raw materials, either through high-mobility or exchanged between groups, these were knapped on site, but with longer use-lives, extensive reduction and high presence of retouch and formal tools.
-
-Similarly, the Proto-Solutrean assemblage shows only disagreement with the expected outcomes within the variability of retouched tools (expectation e and P8-9) and possibly the focus on maintenance and prolonging use-lives of tools (expectation a), especially concerning P2.
-Unlike expected, there are little differences in the amount and type of retouched tool types between local and non-local raw materials.
-Unlike the Gravettian occupation, however, retouched tools are overall not well-represented in the assemblage.
-It is key to notice, however, that the Proto-Solutrean occupation of Vale Boi seemed to introduce other raw materials, such as quartz or dolerite, to the production of retouched tools.
-Although knapping strategies seemed to be overall simpler (e.g., simple platform cores in quartz), splintered pieces are highly represented in quartz.
-In opposition, most endscrapers are made of chert (Belmiro et al., 2021), while the majority of Vale Comprido points are found in dolerite (Belmiro et al., 2021; Marreiros, 2009).
-These patterns, then, may correspond to a specialisation of tool production to the type of raw material itself (whether quartz, chert or dolerite), thus making chert, retouched tool variability patterns similar and unrelated to distance of source.
-Regarding P2, although there is a slightly higher presence of core preparation and maintenance products in non-local cherts, there is an overall lack of this class of artefacts.
-This means that for the Proto-Solutrean assemblage of Vale Boi, core maintenance variables are not adequate to discern differences between local and non-local raw materials, since these knapping strategies were not consistently used during this chronology.
-
-Finally, the Solutrean assemblages also show only two parameters that disagree with the expectations: debitage sizes (P3) and retouched tool sizes (P7).
-Flakes show no differences in dimensions and weight between local and non-local raw materials.
-This may be related to the production process of Solutrean points, which may start with larger flakes or elongated products that are then reconfigured to the desired product, thus producing similar discarded flakes in all chert types.
-However, elongated blanks do follow the expected patterns, with non-local cherts having smaller products and local cherts.
-This may mean that non-local cherts, especially T6, are being further reduced in specific strategies for the production of elongated products, thus leading to different patterns between these two classes of blanks.
-The sizes of retouched tools, alike flakes, also show no relevant differences between local and non-local.
-However, this variable may not be adequate for the present study, especially regarding the size measurements; both width and length are used for scatterplots, however, the technological database used in this study is missing specific measurements (especially length) for typologies of retouched tools.
-As such, the scatterplots are further truncated by the difference in analysis methodology, making this pattern possibly invalid.
-
-Aside from the specific expectations for the treatment of local and non-local cherts within each assemblage, when comparing between technocomplexes, a few key patterns also stand out.
-Chert retouched tools are present in higher percentages in the Gravettian assemblages (~22%) than in the Proto-Solutrean and Solutrean assemblages (<10%).
-This is also clear through the tool-to-debitage ratios, which are higher in the Gravettian assemblage (@fig-ratiotd).
-
-There are little differences in the general size and weight of the artifacts between assemblages.
-For example, for key classes where size and weight differences could be expected (e.g., cores), the length measurements are mostly all located between 20-40 mm and weight between 10-40 gr.
-The major differences seem to be explained better by bigger cores in local chert types and smaller cores for non-local chert types.
-However, a few key differences were noted.
-T4 cores (the chert type most frequent in the Proto-Solutrean assemblage) shows the largest weights of cores of all assemblages, both in the Solutrean and Proto-Solutrean assemblage, as well as T5 in the Solutrean assemblage.
-T6 and T7 have consistently the smaller weights, especially compared to T2 during the Proto-Solutrean and Gravettian, which shows slightly higher weights.
-This consistency in artefact size throughout the assemblages may be explained by a combination of local lithic resource availability and transport and knapping strategies focusing on intensive reduction and maintenance of non-local raw materials.
+The results of the study align with our initial expectations, providing insights into mobility patterns, raw material use, and technological strategies throughout the Upper Paleolithic (UP) in southwestern Iberia.
+We initially anticipated differences in core reduction between local and non-local resources (Table ++, expectations a-b).
+However, these differences are influenced by the transportation strategies used to bring non-local cherts to Vale Boi.
+The presence of non-local chert cores at the site depends on whether these materials were transported as finished tools or blanks—likely part of an individual provisioning strategy (Kuhn, 2004)—or as trimmed cores or larger stone volumes, as seen in site provisioning strategies (Kelly, 1988; Roth & Dibble, 1998; Clark, 1987; Kuhn, 2004).
+
+Our findings indicate that cores were present in non-local chert groups across all studied assemblages, albeit in small numbers (e.g., in the Proto-Solutrean assemblage).
+While the smaller fraction of the assemblages—such as chips, which can serve as proxies for on-site knapping—was not analyzed, other knapping by-products, including shatter in non-local chert types (except in the Solutrean assemblage, where this category was not studied; see Section ++ Materials and Methods), as well as core preparation and maintenance products, may also serve as indicators of core transportation and on-site knapping.
+
+These findings suggest a strategy of transporting cores or nodules to Vale Boi, followed by on-site knapping for most non-local chert types.
+However, an exception appears in the case of certain TL cherts (TL2-6, TL9-10, TL14; Table S13).
+These artifacts, whose source remains unknown but do not match local cherts, lack both cores and associated knapping debris.
+This pattern suggests that these TL cherts may represent an instance of long-distance individual provisioning (Kuhn, 2004), where materials were transported as finished tools or blanks.
+
+The presence of cores and evidence of on-site knapping enabled a comprehensive comparison of knapping strategies—specifically, the degree of preparation and maintenance through maintenance products—as well as knapping intensity between local and non-local cherts.
+
+Our analysis identified core preparation and maintenance products in the studied assemblages, suggesting a possible effort to preserve raw material volume.
+However, these products were found in low percentages across all assemblages and chert types (\<4% and n\<7).
+This indicates that, at Vale Boi, core preparation and maintenance products may not serve as reliable proxies for distinguishing between chert types.
+Instead, their presence may be more closely linked to specific knapping strategies and objectives, such as prioritizing the production of blades and bladelets.
+
+Core and blank sizes and ratios yielded better results.
+Our findings suggest that non-local cherts underwent more intensive reduction than local cherts, as evidenced by the higher blank-to-core ratio and the statistically significant smaller size and weight of blanks made from non-local cherts.
+These results align with previous studies, which propose that the intensive reduction of cores—manifested in smaller core sizes, smaller debitage products, and a higher amount of debitage per core—may be linked to lower mobility.
+In this context, lower mobility results in reduced access to non-local raw materials, which are therefore more intensively reduced and eventually replaced by local raw materials (Kuhn, 2004; Roth & Dibble, 1998; Grove et al., 2023; Surovel, 2009).
+However, our results show that these differences are not limited to long-term occupations and low mobility.
+High blank-to-core ratios and smaller blank sizes in non-local cherts are also observed in the Gravettian occupation, previously interpreted as a short-term occupation which frequent moves (Cascalheira ++; Belmiro et al., 2025).
+Due to this, we may interpret that non-local raw materials were more intensively reduced than local cherts throughout all the UP, independent of the duration of the occupation.
+
+In contrast to blanks, cores exhibited small, though statistically non-significant, differences in size and weight between local and non-local cherts.
+These results may be explained by a combination of factors, including the availability of local lithic resources, transport, and knapping strategies focused on the intensive reduction and maintenance of non-local raw materials.
 Belmiro et al.
-(2023) highlight the small size of chert nodules (usually between ~5-10 cm with a maximum of 20 cm of diameter) in the Algarve region.
-Although abundant and possibly available throughout all of the UP, given their location in accessible cliffs, these lithic resources are often small and fracture easily when obtained directly from the source.
-Given the possibly initial small size of the local chert nodules, it may be possible that the small artefact sizes are related not to intensive knapping of the nodules, but to the initial size of the knapped volumes.
-This is also supported by the lower blank-to-core ratios seen in local raw materials.
-
-In comparison, at least for T6 cherts from central Portugal (for which we have a comparable reference collection of unknapped volumes collected from secondary sources), the initial volumes are frequently larger (>20 cm), although smaller nodules can also be found.
-This suggests that the small artefact sizes of this non-local resource are not necessarily a product of resource availability but an intentional choice of either transportability of smaller volumes/nodules, knapping strategies focused on intensive reduction and maintenance, or both.
-This is also supported by the higher blank-to-core and tool-to-debitage ratios seen in non-local raw materials, especially T6.
-
-A key interpretation can be derived.
-Cores, shatter and debitage are present even in non-local chert types (excluding the TL category), meaning although possibly coming from long-distance (e.g., T6 and T7 with sources over ~200 km, as the crow flies), these raw materials were being introduced to the site as knappable volumes and not only as finished products.
-This has also been previously suggested, based on the presence of cortex attributed to small transport distances from source to procurement area, in non-local cherts in all assemblages (Belmiro et al., 2025).
-
-## Some extra notes and structure
-
-- Goals of the study
-- Parameters studied
-- Summary of the results based on expectations
-- How they fit within the expectations + how they differ by assemblage
-- General differences between expectations
-- Interpretation of chert procurement and management, and changes through time + settlement changes
-
-When groups move to areas where raw materials are scarcer or of poorer quality (understanding that poorer quality may not mean they are bad quality, simply not as good quality as the raw materials group have current access), more effort is placed in the preparation of transportable cores and tools (Kelly, 1988).
-This may be the case for the Gravettian assemblage; not only is the percentage of tools high, large enough volumes of stone were transported to the site and knapped, but also high frequency of burins, which may have also been a technological choice to obtain elongated products in an efficient way.
-There seems to be an effort in transporting raw materials of known good quality in transportable yet adaptable shapes to later shape and use.
-
-It would be also expected that whenever groups have easy access to good quality, non-local raw materials, either due to high mobility or exchanges between groups, these resources become more abundant in the assemblage, but also be less extensively used (Andrefsky, 1994).
-The idea is that the patterns of non-local raw material use are based on their conservation due to cost of procurement; when that cost is reduced, so is the necessity to conserve the resources, and thus the expected patterns may not be as clear.
-Despite the high frequency of non-local raw materials, which implies some sort of ease of access to non-local raw materials of good quality, probably due to high residential mobility, the assemblage still shows conservation of raw materials.
-This is seen in the high frequence of retouched tools and high ratios of tool to debitage when compared to local raw materials.
-Artefact sizes are also smaller in non-local raw materials then local raw materials and blank-to-core ratios are higher, showing that these non-local raw materials were extensively used.
-This shows that access to non-local raw materials, making them highly represented in the assemblage, especially the cherts from central Portugal (T6), did not impact the patterns expected for the treatment of non-local raw materials.
-
-A strategy used to raw material transportation is stockpilling lithic raw materials in various production stages, thus reducing the need for extensive lithic production and tool reworking (Kuhn, 2004).
-Longer-term residential occupations occupations are more likely to involve stockpilling strategies and reduced investment in tool maintenance (Bamforth, 1990).
-This would create assemblages with more blanks, less maintenance products and less retouched tools such as retouched flakes.
-However, even with stockpilling,
+(2023) highlight that chert nodules in the Algarve region vary between \~5-10 cm, with a maximum diameter of 20 cm.
+Although these nodules are abundant and likely accessible throughout the Upper Paleolithic, given their location in accessible cliffs, they tend to be small and prone to fracture when extracted directly from the source.
+Given the initial size of local chert nodules, it is possible that the small core sizes are not the result of intensive knapping, but rather reflect the initial size of the knapped volumes.
+
+In comparison, cherts from central Portugal—the probably source from T6 cherts from Vale Boi—can be found in larger volumes (\>20 cm), although smaller nodules are also present.
+This suggests that the small artifact sizes associated with this non-local resource are not necessarily due to limited resource availability, but rather an intentional choice.
+This decision may involve the transportability of smaller volumes or nodules, knapping strategies focused on intensive reduction and maintenance, or a combination of both.
+This is further supported by the higher blank-to-core ratio observed in non-local raw materials, particularly T6.
+It may also explain the more frequent occurrence of heavier and larger outliers in local chert cores, compared to the relatively rare occurrence of larger outliers in non-local types (T6-7).
+
+The intensity of reduction and maintenance discussed thus far can also be inferred from the frequency and diversity of retouched tools.
+In fact, retouched tools are often key elements used in the literature to distinguish between differential use of local and non-local raw materials.
+Our results align with expected outcomes, as non-local cherts exhibit higher frequencies of retouched tools and higher tool-to-blank ratios across all assemblages.
+This supports the idea that non-local raw materials were actively procured and that greater effort was invested in tool production (Andrefsky, 1994; Perlès, 1992; Kelly, 1988) and the maintenance of existing artifacts through retouch (Kuhn, 2004).
+
+While the intensity of retouch aligns with the expected outcomes (Table ++, expectations 1c and 2c), tool diversity and typologies (Table ++, expectations 1d and 2d) differ from what was anticipated.
+It was expected that non-local cherts would show lower typological diversity, as the distance to the source and inferred mobility suggest that less specialized tools (e.g., retouched blanks) would be more suitable for various use-case scenarios (Shott, 1986; Odell, 1981; Siegel, 1984).
+
+However, with the exception of the Solutrean assemblage, our results show higher tool diversity in non-local cherts in the Gravettian assemblages, and similar tool diversity between chert types in the Proto-Solutrean assemblages.
+This suggests that tools were transported as either less specialized tools or blanks, which were then maintained on-site.
+Our results indicate that long-distance raw materials, particularly T6 and T7, were likely knapped at the site.
+Therefore, the high diversity in tool typology may reflect the transportation of raw material volumes and their subsequent knapping on-site, coupled with a possible preference for these materials in the production of retouched tools, contributing to their high typological variability.
+
+Despite this high variability, less specialized tools, such as retouched blanks, remain predominant in non-local cherts across all assemblages.
+Specialized tools (e.g., backed tools during the Gravettian occupation) are also present, supporting the interpretation that, due to the likely ease of access to these raw materials—either through high mobility or exchange between groups—these materials were knapped on-site.
+However, they were characterized by longer use-lives, extensive reduction, and a high presence of retouch, with a predominance of non-specialized formal tools.
+
+While the patterns identified so far have established a clear distinction in the use and maintenance of local and non-local resources, it is also important to discuss the possible reasons for the varying abundance and technological patterns of non-local raw materials throughout the Upper Paleolithic at Vale Boi, as well as the factors behind the differences observed across the different technocomplexes.
+
+Previous research has often focused on the concepts of abundance and quality in the procurement and use of non-local lithic resources over local ones to explain differences in proportion, use and maintenance.
+In this context, the scarcity or low quality of local resources would drive the procurement of non-local materials and require more effort in preparing transportable cores and producing tools (Andrefsky, 1994; Perlès, 1992; Kelly, 1988).
+However, quality may become a less significant factor when non-local raw materials are easily accessible (Akerman et al., 2002; Torrence, 1996; Brown, 1999; Oestmo, \*\*\*).
+Even in the presence of high-quality local materials, effort was still invested in transporting non-local resources (Roth & Dibble, 1998).
+
+Regarding Vale Boi, quality does not appear to be a defining factor in raw material choice, as typologies typically associated with high-quality lithic resources are also present in other materials.
+For example, Solutrean bifacial tools and points were produced using local chert, which may be considered lower quality than Rio Maior Cretaceous chert due to its coarser texture and less homogeneous composition.
+This suggests that the choice of raw material was not solely dependent on quality.
+
+Similarly, Vale Comprido points, a fossil indicator of the Proto-Solutrean, were primarily made from other lithic resources (e.g., dolerite), with only two identified points made of chalcedony and chert (Belmiro et al., 2021).
+This further supports the idea that raw material choice was not strictly based on quality, as these points were made from a variety of materials, including those less commonly associated with formal tools at the site.
+
+In this case, the impact of raw material quality seems to be more related to the differences between lithological groups, which have varying textures and breakage patterns (e.g., between greywacke and chert), rather than within micro-cryptocrystalline quartz resources like chert.
+This distinction suggests that raw material characteristics, such as texture and fracture properties, may have played a more significant role in tool production than raw material quality alone.
+
+Finally, a visual inspection of a cache found in the Shelter area attributed to the Gravettian revealed a high frequency of backed bladelets made from both local and non-local cherts.
+This observation further emphasizes the use of diverse raw materials, regardless of their quality.
+
+Therefore, it is not necessarily the case that non-local cherts were procured for their quality or because local cherts were unsuitable.
+Instead, their presence may reflect mobility strategies, territorial exploitation, and social networks that facilitated the introduction of long-distance cherts into the technological practices of southwestern hunter-gatherer groups.
+These non-local materials were used and maintained differently, possibly due to their initial size or the need for conservation, given their more restricted or higher-cost acquisition compared to local cherts.
+
+These mobility strategies, territorial exploitation, and social networks also help explain the distinct patterns of non-local raw material use observed between the Gravettian and other assemblages.
+One key difference is the high frequency of non-local cherts in the Gravettian assemblage at Vale Boi, which make up approximately 50% of the chert assemblage (Belmiro et al., 2025).
+When combined with technological data, this also reveals significant differences in the frequency of retouched non-local chert tools and the tool-to-blank ratio (\~30% and \~1, respectively), compared to other technocomplexes (\~15-19% and 0.1-0.5).
+Several studies have shown that short-term occupations often rely heavily on non-local toolkits and raw materials (Torrence, 1983; Surovell, 2009), and are characterized by higher reduction intensity and increased tool-to-debitage ratios (Grove et al., 2023).
+These differences support the interpretation of the Gravettian at Vale Boi as a short-term occupation (Belmiro et al., 2025).
+
+This allows us to further expand our knowledge about chert provisioning during the UP at Vale Boi and suggest an overall interpretation for how groups were using and managing their lithic resources through time.
+In southwesternmost Iberia, the first identified UP occupations are from the Gravettian.
+At Vale Boi, these occupations appear to correspond to short-term settlements with frequent mobility, where non-local raw materials were transported as cores or nodules and knapped on-site.
+Alternatively, non-local raw materials were brought in as blanks, which were consistently retouched, although individual toolkits made from limited non-local raw materials—primarily blanks or tools—were also present.
+Raw material volumes were intensively exploited at the site, yielding both specialized and non-specialized tools.
+Local raw materials were also used to produce high frequencies of retouched tools, though this was done less intensively.
+Given the technological differences between the Gravettian Vicentine facies and other regions such as central Portugal and southern Spain (Marreiros et al., ++), we suggest that the approximately 250 km distance between Vale Boi and identified non-local sources reflects raw material exchanges between groups at the edges of their territories, rather than an indication of exploited territory.
+
+These exchanges and patterns of mobility persisted throughout the Upper Paleolithic, though with changes in the mobility and settlement patterns at Vale Boi.
+Over time, Vale Boi was occupied as a long-term residential site.
+Proto-Solutrean and later Solutrean groups, while still having access to non-local raw materials, moved less frequently, which limited the influx of these materials.
+Although non-local raw materials were still intensively exploited and maintained at the site, hunter-gatherer groups increasingly relied on local raw materials.
+This shift was likely due to the proximity of local sources, which facilitated more frequent use, while non-local materials continued to be exploited and maintained with greater intensity.
 

### 2025-04-11 analysis/paper/paper.qmd
Slight changes to the code

@@
     .default = RMTYPE)) |> 
-  select(AREA, ID, RMTYPE, SUBTYPE, WEIGHT, CORTEXTHICKNESS, CORTEXAPPEARANCE) %>%
+  select(AREA, ID, RMTYPE, SUBTYPE, WEIGHT, CORTEXTHICKNESS, CORTEXAPPEARANCE) |> 
   distinct(ID, .keep_all = TRUE)
@@
 
-# Dataset filtering for chert/chalcedony and levels (A-C), 3298 + 61 solutrean points, n=3359
+# Dataset filtering for chert/chalcedony and levels (A-C), 3229 + 61 solutrean points, n=2390
 solutrean_original_chert <- solutrean_JC |> 
   filter(Layer %in% c("A", "B", "C")) |> # Filtering Solutrean layers
-  filter(`Raw material` %in% c("Chert", "Chalcedony"))
+  filter(`Raw material` %in% c("Chert", "Chalcedony")) |> 
+  filter(Class != "Chip")
 
@@
                                         "T8", "T9", "T10", "T11", "TL"))) |>
-  ggplot(aes(x=Assemblage, y=`Blank/Core`, colour = Source)) +
-  geom_point(size=3) +
+  ggplot(aes(x=Assemblage, y=`Blank/Core`, fill = Source)) +
+  geom_point(size=3, shape = 21, colour = "black") +
   coord_cartesian(clip = "off") +
   theme(legend.position = "none", axis.title.x = element_blank()) +
-  scale_colour_manual(values = c("#D49B54", "#8B6969", "#CDAF95", "#2B4F60",
+  scale_fill_manual(values = c("#D49B54", "#8B6969", "#CDAF95", "#2B4F60",
                               "#BDD2B6"))
@@
                                         "T8", "T9", "T10", "T11", "TL"))) |>
-  ggplot(aes(x=Assemblage, y=`Tool/Blank`, colour = Source)) +
-  geom_point(size=3) +
+  ggplot(aes(x=Assemblage, y=`Tool/Blank`, fill = Source)) +
+  geom_point(size=3, shape = 21, colour = "black") +
   theme(legend.position = "bottom", axis.title.x = element_blank()) +
-  scale_colour_manual(name = "Chert type", values = c("#D49B54", "#8B6969", "#CDAF95", "#2B4F60",
+  scale_fill_manual(name = "Chert type", values = c("#D49B54", "#8B6969", "#CDAF95", "#2B4F60",
                               "#BDD2B6", "#BDC7C9", "#798777"))
@@
 
-ggpubr::ggarrange(barplot_gravettian, barplot_proto, barplot_solutrean, labels = c("a", "b", "c"), nrow = 3, common.legend = TRUE, legend = "right")
+ggpubr::ggarrange(barplot_gravettian, barplot_proto, barplot_solutrean, labels = c("a", "b", "c"), nrow = 3, common.legend = TRUE, legend = "bottom")
 

### 2025-04-11 analysis/paper/paper.qmd
Updating qmd with final paper.

@@
 abstract: |
-  Text of abstract
+  The Upper Palaeolithic (UP) of Westernmost Europe was marked by technological and cultural transformations shaped by Late Pleistocene climatic shifts.
+The Iberian Peninsula, particularly southwestern Iberia, served as a key eco-cultural niche, sustaining hunter-gatherer populations through environmental fluctuations.
+Vale Boi, the only site in southwestern Iberia with a nearly complete UP sequence, provides critical insights into technological, economical and mobility patterns over time.
+This study examines lithic raw material procurement and technological organization at Vale Boi, focusing on the role of non-local cherts in mobility strategies and social interactions.Using a combined techno-typological and geoarchaeological approach, we analyzed chert assemblages from Gravettian, Proto-Solutrean, and Solutrean levels, integrating previously published macroscopic and petrographic data.
+Our results indicate that non-local cherts were intensively reduced, suggesting strategic curation.
+The presence of diverse tool types challenges the assumption that non-local materials were transported solely as finished tools, instead highlighting their role as versatile raw material sources.
+Gravettian occupations exhibited greater reliance on non-local cherts, implying frequent mobility, while later phases saw increased dependence on local materials, aligning with more long-term occupations.These findings refine our understanding of UP technological adaptations in southwestern Iberia, emphasizing the evolving role of lithic resources in mobility, settlement, and social networks.
+By integrating lithic technology and raw material sourcing, this study contributes to broader discussions on eco-cultural niches and UP hunter-gatherer adaptations.
+
 keywords: |
-  keyword 1; keyword 2; keyword 3
-highlights: |
-  These are the highlights. 
+  Upper Palaeolithic; Lithic technology; Hunter-gatherer mobility; Chert procurement; Iberian Peninsula
 ---
@@
 
-The Iberian Peninsula has been a key region for the study of human adaptations and cultural transitions in Archaeology.
-Located in the westernmost tip of Europe and with a generally stable climate even during the coldest periods punctuating the Late Pleistocene, this region has yielded some of the most significant glacial refugia in Europe [@hewitt2000; @gómez2007; @gonzález-sampériz2010; @carvalho2022], and has been considered a long-term eco-cultural refugia [@cascalheira2017].
-
-For this reason, the Iberian Peninsula has been the focal point of several discussions regarding Late Pleistocene human adaptations [@finlayson2006; @zilhão2017; @cascalheira2017].
-
-A particularly good example is the region’s role in the discussion of Upper Paleolithic technocomplexes transitions.
-This has been possible due to the existence of several ecological niches, consistently used through time, possibly due to the stability in the richness and variety of resources in Iberia.
-This continuous use would then create long-term regional adaptive structures, which when correlated with the ecological niches, have been referred to as eco-cultural niches [@cascalheira2017].
-These eco-cultural niches—which yielded multi-layered sites—.
-provide an exceptional opportunity to understand long-term dynamics regarding biotic and abiotic resource exploitation since they can provide details on how human populations maintained or changed their adaptive systems when facing cultural and social transformations.
-
-More specifically, southwestern Iberia has been shown to be a key eco-cultural niche to study cultural transitions and the role of such a peripheral coastal region within hunter-gatherer social networks [@cascalheira2017].
-This has been mostly due to the archaeological site of Vale Boi, located in southwesternmost Iberia and characterized by a nearly complete UP sequence (++).
-This site is crucial for understanding Upper Paleolithic (UP) adaptations, as it is currently the only site in southwestern Iberia with a long-term occupation spanning most of the UP, enabling the exploration of cultural and technological trends over time (Cascalheira et al.
-2017).
-
-For example, Vale Boi has provided valuable insights into hunter-gatherer subsistence strategies and technological organisation (Belmiro et al.
-2021; Bicho et al.
-2013; Cascalheira et al.
-2017; Cascalheira 2019; Horta et al.
-2019; Manne et al.
-2012; Marreiros et al.
-2016), contributing to a better understanding of territory exploitation and interactions with other regions of Iberia throughout the UP.
-Especially due to lithic technology studies from this site, the region has been shown to have played a shifting and complex role within the larger territory.
-
-During the Gravettian, the peripheral nature of the region studies of Gravettian occupations reveal marked regional differences, suggesting more limited movement of people and information in southwestern Iberia (Marreiros and Bicho, 2013).
-In comparison, Proto-Solutrean evidence indicates technological similarities and close contact with central Portugal (Belmiro et al., 2021).
-The Solutrean period stands out as a particularly dynamic phase at Vale Boi, during which the site served as a key connection point between central Portugal and southern Spain, facilitating social networks and cultural exchanges (Cascalheira, 2013; Cascalheira et al., 2017).
+The Upper Palaeolithic (UP) in Europe, dating to approximately 40,000–10,000 years ago, was marked by significant technological innovations, shifts in subsistence strategies, and the emergence of diverse regional traditions [@gamble2004; @straus1995].
+
+In Westernmost Europe, five main technocomplexes have been identified within the UP  - Aurignacian, Gravettian, Proto-Solutrean, Solutrean, and Magdalenian - each contributing new evidence on resilient behaviors and cultural continuity.
+These technocomplexes are distinguished by regional adaptations as well as technological and subsistence innovations [@baker2024; @bicho2017; @bicho2011; @bicho2008; @cascalheira2017; @straus1991; @straus1993; @zilhão1997].
+
+These innovations are believed to have resulted from human adaptations and resilience to the shifting conditions of the Late Pleistocene, which was marked by abrupt climatic events such as the Heinrich events and the Last Glacial Maximum (LGM) [@sanchezgoñi2010].
+These climatic shifts had a profound impact on flora and fauna, with exception of spatially limited refugia, likely disrupting settlement patterns and the organization of hunter-gatherer groups, ultimately leading to a process of cultural reorganization [@bradtmöller2012; @morin2008; @schmidt2012], which has frequently been a focus of UP research.
+
+In particular, the Iberian Peninsula has been recognized as a key region for studying UP adaptations and cultural transitions [e.g., @aubry2016; @aubry2022; @bicho2011; @bicho2017a; @bradtmöller2012; @cascalheira2017a; @cascalheira2021; @cascalheira2013; @finlayson2006; @gamble2004; @jennings2011; @jiménez-espejo2007; @linscott2023; @rasilla2020; @schmidt2012; @zilhão1997a; @zilhão2021; @zilhão2017a].
+This significance is largely due to its location in southwestern Europe, which enabled it to maintain a relatively temperate climate even during the coldest periods, such as the LGM.
+As a result, the Iberian Peninsula served as one of Europe’s most important glacial refugia [@gómez2007; @gonzález-sampériz2010; @hewitt2000; @jennings2011; @schmidt2012].
+
+These refugia provided stable environments that were continuously occupied by hunter-gatherers, leading to the formation of eco-cultural niches - stable socio-ecological systems where human settlements were recurrent, and cultural adaptations were shaped by ecological constraints and opportunities [@cascalheira2017].
+These eco-cultural niches present a unique opportunity to examine long-term dynamics in resource exploitation, offering valuable insights into human adaptive strategies in response to cultural and social transformations.
+
+More specifically, southwesternmost Iberia has been identified as a key eco-cultural niche for studying cultural transitions and the role of this peripheral coastal region within hunter-gatherer social networks.
+This is mainly due to the archaeological site of Vale Boi, the only site in southwestern Iberia with a nearly complete UP sequence, making it a crucial reference for examining long-term technological and mobility patterns in the region [@cascalheira2017].
+Its well-preserved stratigraphic continuity allows for a detailed analysis of cultural and technological trends across different phases of the UP.
+
+Building on this significance, research at Vale Boi has provided valuable insights into hunter-gatherer subsistence strategies and technological organisation [@belmiro2021; @bicho2013; @bicho2017a; @cascalheira2019; @horta2019; @manne2012; @marreiros2015; @pereira2016], and lithic technology studies from this site have highlighted the region’s shifting and complex role within the broader territorial landscape.
+
+In this context, comparisons between Vale Boi and the Gravettian occupations in central Portugal and southern Spain, revealed marked regional differences, suggesting more limited movement of people and information in southwestern Iberia.
+These differences were suggested to be related to Vale Boi’s peripheral location and the distinct ecological characteristics of westernmost Algarve [@marreiros2013].
+In contrast, Proto-Solutrean evidence from the site indicates technological similarities and strong connections with central Portugal [@belmiro2021], while the Solutrean period stands out as a highly dynamic phase at Vale Boi, during which the site functioned as a key connection point between central Portugal and southern Spain, facilitating social networks and cultural exchanges [@bicho2017a; \@cascalheira2013a].
 
 Expanding on these findings, recent research has further refined our understanding of mobility and social networks through lithic raw material studies at the site.
-Our team’s recent work (Belmiro et al., 2025) identified non-local cherts originating from central Portugal (\>250 km from Vale Boi) and southern Spain (\>200 km from Vale Boi).
-The proportions of these materials varied across the UP, with earlier occupations (Gravettian) showing high frequencies of non-local chert, while later ones (Proto-Solutrean and Solutrean) exhibited a marked decline.
-These findings suggest the establishment and maintenance of extensive social networks connecting southwestern Portugal to Southern spain and central Portugal as early as the beginning of the UP.
+Our team’s recent work identified non-local cherts originating from central Portugal and southern Spain in the UP lithic assemblages of Vale Boi [@belmiro2025].
+The proportions of these materials varied throughout the UP with earlier occupations (Gravettian) showing high frequencies of non-local chert, while later phases (Proto-Solutrean and Solutrean) exhibited a marked decline.
+These findings suggest the establishment and maintenance of extensive social networks connecting southwestern Portugal to southern Spain and central Portugal as early as the beginning of the UP.
 
@@
 
-Despite the insights gained from previous studies, the relationship between lithic technology and raw material use at Vale Boi remains largely uncorrelated.
-While broader research highlights the complexity of lithic technological organisation, a more integrated approach is needed to fully understand how raw material provisioning influenced technological strategies at the site and what this can tell us about how hunter-gatherer groups organized, explored the territory and connected to others across Iberia.
-To address this, our goal is to combine lithic technological and chert raw material studies to better characterize hunter-gatherer technological and social adaptations throughout the UP in southwestern Iberia.
-To do so, we create and test expectations built upon previous archaeological and ethnographic studies of the technological organization of hunter-gatherers.
+While previous studies have explored Vale Boi’s technological organisation, a direct analysis of how raw material sourcing shaped lithic production strategies remains missing.
+This study bridges that gap by examining chert procurement in relation to tool production, use, and discard, providing critical insights into hunter-gatherer mobility, settlement strategies, and social interactions across Iberia.
+
+To address this gap, this paper presents a combined analysis of lithic technology and raw material sourcing, aiming to refine our understanding of technological and social adaptations throughout the UP in southwestern Iberia.
+By integrating archaeological and ethnographic perspectives, we test expectations regarding the organisation of hunter-gatherer technological systems and their implications for mobility, resource management, and social interactions.
+In doing so, we contribute to ongoing discussions about eco-cultural niches and the role of southwestern Iberia within broader UP social networks.
 
@@
 
-Lithic technological organisation has long been a central focus in archaeology for the insights it offers into past human behaviour and lifeways.
-Nelson (1991) defines lithic technological organisation as the selection and integration of strategies for the production, use, transportation, discarding, and management of tools and raw materials.
-Kelly (1988) describes it as the interplay of space and time in lithic tool production within a cultural system.
-This includes activities such as raw material sourcing and distribution, tool manufacture, use, reuse, and discard, as well as broader behavioural patterns like mobility and settlement systems.
-Together, these definitions underscore the multifaceted nature of lithic technological organisation and its utility in examining key aspects of hunter-gatherer adaptations over time.
-
-Raw material procurement is a key aspect of lithic technological organisation, as it played a crucial role in the lives of past hunter-gatherer groups by determing the possibilities for tool production and use (Bamforth & Bleed, 1997).
-Approaches focused on the study of raw materials provide key knowledge on several topics of hunter-gatherer organisation such as procurement and mobility strategies (Ambrose, 1990; Binford, 1979, 1985; Gould, 1985; Kuhn, 1991; Mccall, 2007), occupation types and their duration (Kuhn, 2004; Surovell, 2009; Grove, 2023), and social networks and exchanges (Whallon, 2006; Gamble, 1999).
-These approaches focus on the characterisation of geological and archaeological raw materials to identify local and non-local raw materials within an assemblage, their possible sources, and examine their proportions within a site over different occupations and periods.
+Lithic technological organization has long been a central focus in archaeology, offering critical insights into past human behaviour and adaptive strategies.
+It encompasses all stages of stone tool production and use, from raw material acquisition to discard.
+Nelson [-@nelson1991] defines it as the selection and integration of strategies for tool production, use, transport, and maintenance, while Kelly [-@kelly1988] emphasizes its spatial and temporal dimensions within cultural systems.
+These perspectives highlight the interrelationship between raw material sourcing, mobility, and settlement organization—key factors in understanding hunter-gatherer adaptations.
+
+Raw material procurement is a fundamental aspect of lithic technological organisation, as it played a crucial role in the lives of past hunter-gatherer groups by determining the possibilities for tool production and use [@bamforth1997].
+Approaches focused on the study of raw materials provide key knowledge on several topics of hunter-gatherer organisation such as procurement and mobility strategies [@ambrose1990; @binford1979; @binford1985; @gould1985; @kuhn1991; @mccall2007; @gramly1980], occupation types and their duration [@torrence1983; @grove2023; @kuhn2004a; @surovell2009], and social networks and exchanges [@gamble1999; @whallon2006].
+However, raw material studies often focus on the characterisation of geological and archaeological raw materials to identify local and non-local raw materials within an assemblage, their possible sources, and examine their proportions within a site over different occupations and periods.
 
@@
 A raw material-focused approach alone is insufficient for comprehensively understanding mobility patterns and technological organisation.
-Indeed, numerous studies and models from both ethnographic and archaeological perspectives have demonstrated the complex interplay between lithic technological variability, raw material provisioning, and mobility.
-
-For example, technological variability within a lithic assemblage may be explained by the abundance and quality of raw materials (Andrefsky, 1994; Brown, 1999; Oestmo, 2017).
-Whenever raw materials in proximity to a residential site are abundant and of good quality, a high percentage of formal tools are produced (Andrefsky, 1994; Perlès, 1992), and artefacts show less reduction intensity and decreased conversation through retouch (Thacker, 2002).
-Whenever local raw materials are scarce or of low quality, non-local resources are procured to produce mostly formal tools (Andrefsky, 1994; Perlès, 1992).
-Similarly, when groups move to areas where raw material is scarcer or of poorer quality, more effort is placed into the preparation of transportable cores and the production of tools (Kelly, 1988).
-
-However, the different patterns between local and non-local raw material proportion and use in lithic assemblages may be less obvious whenever prehistoric groups have easy access to non-local raw materials due to high mobility or exchanges between groups (Akerman et al., 2002; Torrence, 1996; Brown, 1999; Oestmo).
-Furthermore, Roth and Dibble (1998) demonstrate that even in areas where good quality raw material is abundant, non-local raw materials were still transported to the site, often in the form of large blanks or trimmed cores, and both local and non-local raw materials were retouched into tools.
-
-Tools, flakes and cores often functioned as portable sources of raw material within mobile toolkits or during non-local raw material procurement, enabling stone knapping during periods of limited time or in environments with scarce or unsuitable materials (Kelly, 1988; Morrow, 1996; Gould et al., 1971; Kuhn, 1994; Perlès, 1992; Shott, 1986).
-For example, the transport of formal cores (e.g., blade cores) and bifaces used as cores have been interpreted as lightweight and reconfigurable raw material storage strategies (Clark, 1987; Kelly, 1988).
-Formal tools and the degree of tool maintenance have also been associated with planned preparation, efficient use, and transportability, reflecting their connection to mobile settlement strategies, short-term site occupations (Torrence, 1983) and the regional distribution of raw materials (Bamforth, 1986).
-For example, non-local raw materials, transported as part of personal toolkits (frequently associated with residential mobility), would primarily arrive at sites as retouched pieces (Kuhn, 2004).
-These artefacts would show higher degrees of reduction and reworking (Kuhn, 2004) and a higher percentage of non-cortical flakes (Roth and Dibble 1998), as they would have been in use for an extended period before being discarded.
-When knapped on-site, non-local materials also tend to exhibit higher proportions of retouched tools compared to local materials.
-Tool variability and size are also impacted by mobility and uncertainty regarding raw material availability; increased group mobility may create a pattern of limited tool variability, and tools may become less specialised and easily reconfigurable, such as retouched blanks (Shott, 1986; Odell, 1981; Siegel, 1984).
-Increased mobility also often results in smaller tools (Shott, 1986), although other authors have noted that large tools may be well-suited to a mobile lifestyle due to their extended use-life, allowing them to remain functional in areas where raw materials are scarce or unsuitable for knapping (Morrow, 1996).
+Numerous studies and models from both ethnographic and archaeological perspectives have demonstrated the complex interplay between lithic technological variability, raw material provisioning, and mobility.
+
+For example, technological variability within a lithic assemblage may be explained by the abundance and quality of raw materials [@andrefsky1994; @brown1999; @oestmo2017].
+Whenever raw materials in proximity to a residential site are abundant and of good quality, a high percentage of formal tools are produced [@andrefsky1994; @perlès1992], and artefacts show less reduction intensity and decreased conservation through retouching [@thacker2002].
+Whenever local raw materials are scarce or of low quality, non-local resources are procured to produce mostly formal tools [@andrefsky1994; @perlès1992].
+Similarly, when groups move to areas where raw material is scarcer or of poorer quality, more effort is placed into the preparation of transportable cores and the production of tools [@kelly1988].
+
+However, the different patterns between local and non-local raw material proportion and use in lithic assemblages may be less obvious whenever prehistoric groups have easy access to non-local raw materials due to high mobility or exchanges between groups [@akerman2002; @contexts1982; @torrence1986; @brown1999; @oestmo2017].
+Furthermore, Roth and Dibble [-@roth1998] demonstrate that even in areas where good quality raw material is abundant, non-local raw materials were still transported to the site, often in the form of large blanks or trimmed cores, and both local and non-local raw materials were retouched into tools.
+
+Tools, flakes and cores often functioned as portable sources of raw material within mobile toolkits or during non-local raw material procurement, enabling stone knapping during periods of limited time or in environments with scarce or unsuitable materials [@gould1971; @kelly1988; @kuhn1994; @morrow1996; @perlès1992; @shott1986].
+In this context, the transport of formal cores (e.g., blade cores) and bifaces used as cores have been interpreted as lightweight and reconfigurable raw material storage strategies [@clark1987; @kelly1988].
+
+Formal tools and the degree of tool maintenance have also been associated with planned preparation, efficient use, and transportability, reflecting their connection to mobile settlement strategies, short-term site occupations [@torrence1983] and the regional distribution of raw materials [@bamforth1986].
+In some cases, non-local raw materials, transported as part of personal toolkits (frequently associated with residential mobility), would primarily arrive at sites as retouched pieces [@kuhn2004].
+These artefacts would show higher degrees of reduction and reworking [@kuhn2004] and a higher percentage of non-cortical flakes [@roth1998], as they would have been in use for an extended period before being discarded.
+When knapped on-site, non-local materials also tended to exhibit higher proportions of retouched tools compared to local materials.
+
+Tool variability and size are also impacted by mobility and uncertainty regarding raw material availability: increased group mobility may create a pattern of limited tool variability, and tools may become less specialised and easily reconfigurable, such as retouched blanks [@odell1981; @shott1986; @siegel1984].
+Increased mobility also often results in smaller tools [@shott1986], although other authors have noted that large tools may be well-suited to a mobile lifestyle due to their extended use-life, allowing them to remain functional in areas where raw materials are scarce or unsuitable for knapping [@morrow1996].
 
@@
 Short-term occupations typically rely on transported (non-local) toolkits (Torrence, 1983) and raw materials (Surovell, 2009), show higher reduction intensity, and an increased tools-to-debitage ratio in the assemblages (Grove et al., 2023).
-Long-term occupations, typically rely on provisioning sites with locally available and suitable raw materials, while non-local materials are gradually used and discarded (Grove et al., 2023; Surovell, 2009).
+Long-term occupations typically rely on provisioning sites with locally available and suitable raw materials, while non-local materials are gradually used and discarded (Grove et al., 2023; Surovell, 2009).
 Non-local raw material cores are highly reduced and extensively exploited (Kuhn, 2004; Roth & Dibble, 1998) and exhibit higher blank-to-core ratios (Roth & Dibble, 1998).
@@
 This variability appears between behaviourally and ethnically similar sites (Bamforth, 1986), across distinct chrono-cultural periods, and even within a single site among various raw materials.
-To fully understand past hunter-gatherer behaviours, organisation, and lifeways, it is crucial to examine all stages of lithic production - procurement, distribution, use, maintenance, and discard - through interconnected approaches that combine lithic technology and raw material analyses.
-
-Our goal is to understand the provisioning, use and management strategies of chert lithic resources of the hunter-gatherer groups of Vale Boi during the Upper Paleolithic.
-We focus on the technological characteristics of the two broader groups of cherts identified at the site: local and non-local, based on the notion that local raw materials will show technological patterns different from those which are non-local.
-Expanding upon our previous works on chert types present throughout the UP sequence in the Terrace and Shelter area of Vale Boi, and their possible interpretations, and following the previously mentioned studies and models about raw material management and lithic technology organisation focusing mainly on the different management of local and non-local raw materials, associated with mobility and occupation length, we derive several expectations regarding local and non-local raw materials that can be found summarized in @tbl-expectations.
+To fully understand past hunter-gatherer behaviours, organisation, and lifeways, it is crucial to examine all stages of lithic production— procurement, distribution, use, maintenance, and discard—through interconnected approaches that combine lithic technology and raw material analyses.
+
+This study aims to refine our understanding of lithic resource provisioning, use, and management strategies among the UP hunter-gatherer groups settled at Vale Boi.
+To examine the role of chert procurement and reduction strategies at Vale Boi we test several expectations derived from the previously mentioned studies and models about raw material use and lithic technology organisation, focusing mainly on the different management of local and non-local raw materials, associated with mobility and occupation length, which are summarised in @tbl-expectations.
+These expectations are tested through the comparison of local and non-local chert technological and metric characteristics.
 
@@
 #| label: tbl-expectations
-#| tbl-cap: A table.
+#| tbl-cap: "Expected results of the technological and metric comparison between local and non-local cherts."
 
@@
 
-knitr::kable(head(hypotheses))
-
-```
-
-Despite these expectations, it is important to notice that, especially in occupations with a high percentage of non-local cherts at the site (e.g., Gravettian occupations of level 7 and 6 of the Terrace at Vale Boi), these technological may not reflect a preoccupation with intensive management of a scarce and limited non-local resource, and be more a reflection of the type of provisioning applied following the suggested model by Kuhn (2004).
-In this sense, distance from source may have had little intrinsic meaning in the past, especially regarding non-local raw materials.
-There are several ways in which raw materials may be transported to a site, and the relationship between distance and cost may vary depending on how this takes place (Kuhn, 2004).
-When non-local raw materials are as abundant as local ones, the difference in the technological organization between these materials may be less obvious (Andrefsky, 2004).
+knitr::kable(hypotheses)
+
+```
 
@@
 
-Vale Boi is situated on the western coast of the Algarve region in southern Portugal, within a small valley that stretches southward to the Atlantic coast, approximately 2 km away.
-The site is bordered by limestone outcrops, which form rock shelters facing west and southwest.
-Covering more than 10,000 square meters along the valley slope, Vale Boi includes three main excavated areas: Slope, Terrace, and Shelter, which were excavated between 2000 and 2019 (Bicho et al., 2007, 2012, 2013; Cascalheira et al., 2017; Cascalheira & Bicho, 2013; Manne et al., 2012).
-Based on the available raw material and technological data, particularly the macroscopic and petrographic research (Belmiro et al., 2025), the Terrace and Shelter areas were selected for the current study.
-
-The Terrace spans occupations from the UP to the Early Neolithic.
-Although excavations began in 2000, a new 8-square-meter area was opened in 2012 to better understand the stratigraphic sequence and assess the potential presence of older cultural layers, which could be linked to an early Upper Paleolithic occupation.
+Vale Boi is located on the western coast of the Algarve region, southern Portugal, within a small valley extending \~2 km southward to the Atlantic Ocean (\@fig-map a).
+The site is nestled among limestone outcrops that create west- and southwest-facing rock shelters (\@fig-map b).
+Covering more than 10,000 square meters along the valley slope, Vale Boi includes three main excavated areas, Slope, Terrace, and Shelter (\@fig-map c), which were excavated between 2000 and 2019 (Bicho et al., 2007, 2012, 2013; Cascalheira et al., 2017; Cascalheira & Bicho, 2013; Manne et al., 2012).
+Based on the available raw material and technological evidence, particularly the macroscopic and petrographic data (Belmiro et al., 2025), the Terrace and Shelter areas were selected for the current study.
+
+```{r}
+#| label: fig-map
+#| fig-cap: "a) Location of the archaeological site of Vale Boi (orange box) within the Iberian Peninsula; b) Photo of the limestone hill with terrace and shelter locations; c) Schematic of the site area.
+Figures b-c adapted from Cascalheira et al., 2017."
+
+```
+
+The Terrace area preserves occupations ranging from the UP to the Early Neolithic.
+While excavations started in 2000, an additional 8-square-meter trench was opened in 2012 to refine the stratigraphic sequence and assess the presence of earlier UP deposits.
 Between 2012 and 2016, eight primary litho-stratigraphic units were identified.
-The UP sequence includes multiple occupations, spanning \~10,000 years: Gravettian (levels 8 to 6) between approximately c.
-32 and 27 ka cal BP, Proto-Solutrean (levels 5 to 4E) between approximately c.
-26 and 24 ka cal BP, and Solutrean (levels 4D, 4C, 4, and Lower 3) between approximately c.
-24 and 20 ka cal BP (Belmiro et al., 2021; Cascalheira et al., 2017; Cascalheira & Bicho, 2013).
+The UP sequence includes multiple occupations, spanning \~10,000 years: Gravettian (levels 8 to 6) dated between approximately c. 32 and 27 ka cal BP, Proto-Solutrean (levels 5 to 4E) between approximately c. 26 and 24 ka cal BP, and Solutrean (levels 4D, 4C, 4, and Lower 3) between approximately c. 24 and 20 ka cal BP (Belmiro et al., 2021; Cascalheira et al., 2017; Cascalheira & Bicho, 2013).
 
 The Shelter area contains four main lithostratigraphic units, with occupations from the Magdalenian, Solutrean, and Gravettian periods.
-The Magdalenian and Gravettian levels (Z and D, respectively) were partially eroded, while the Solutrean levels represent the most intensive occupation, with three well-preserved archaeological horizons (layers C to A), dated to approximately c.
-24-22 ka cal BP.
-These occupation levels were found beneath blocks of limestone that likely collapsed from the rockshelter ceiling, probably shortly after the Last Glacial Maximum (Cascalheira, 2010; Cascalheira et al., 2017).
-
-Most of these occupations resulted in large lithic assemblages, encompassing all lithic production stages and use of several raw materials, often for specific uses.
-Research on the lithic assemblages of the site have allowed the technological and cultural characterisation of human groups occupying the site, which show some differences through the different UP occupations.
-Given the chosen sample for this study (see section 2 “Materials and methods”), we summarise here the results for the technological analysis of the Solutrean assemblages (levels A-C) of the Shelter (Cascalheira, 2010; Cascalheira et al., 2013) and the Proto-Solutrean assemblages (levels 4E-5) of the Terrace (Belmiro, 2020; Belmiro et al., 2021).
-The Gravettian occupations (Levels 6-8) of the Terrace have been recently studied, but the results remain unpublished.
-
-The lithic analysis of the Shelter Solutrean levels (A-C) shows a high number of lithic artifacts, with over 22 000 artifacts analysed (including chips which correspond to \~60% of the assemblage and shatter \~17-19%).
+The Magdalenian and Gravettian levels (Z and D, respectively) were partially eroded, while the Solutrean levels represent the most intensive occupation, with three well-preserved archaeological horizons (layers A to C), dated to approximately c. 24-22 ka cal BP.
+These occupation levels were found beneath blocks of limestone that likely collapsed from the rock shelter ceiling, probably shortly after the LGM (Cascalheira, 2010; Cascalheira et al., 2017).
+
+The site’s lithic assemblages represent all stages of stone tool production, with raw material use varying by occupation.
+Technological analyses of these assemblages have provided insights into the adaptive strategies of UP hunter-gatherers at Vale Boi, revealing technological shifts across different cultural phases.
+Given the chosen sample for this study (see **Section 2.2.
+Datasets and analysis**), we summarise here the results for the technological analysis of the Solutrean assemblages (levels A-C) of the Shelter (Cascalheira, 2010; Cascalheira et al., 2013) and the Proto-Solutrean assemblages (levels 4E-5) of the Terrace (Belmiro, 2020; Belmiro et al., 2021).
+The Gravettian occupations (levels 6-8) of the Terrace have been recently studied, but the results remain unpublished.
+
+The lithic analysis of the Solutrean levels from the Shelter (levels A-C) shows a high number of lithic artifacts, with over 22,000 artifacts analysed (including chips which correspond to \~60% of the assemblage and shatter \~17-19%).
 The volumes introduced into the site were mostly cortical, given the high frequencies of cortex in blanks and retouched tools.
-The knapping strategies are dominated by unidirectional and bidirectional prismatic, with frequent core maintenance products.
-Flakes dominate the debitage assemblage, and endscrapers, notches, splintered pieces and retouched blades/flakes are the most common retouched tools.
-Stemmed and winged points, laurel leaves, shouldered points and parpalló points were also identified (Cascalheira et al., 2013).
-
-The Proto-Solutrean horizons yielded more than 26,500 artifacts (including chips which correspond to \~65-70% of the assemblage and shatter \~20%), \~11,000 attributed to the lower bottom of level 5 and \~15,600 attributed to the top of level 5 and level 4E.
+The knapping strategies are dominated by unidirectional and bidirectional prismatic cores, with frequent core maintenance products.
+Flakes dominate the debitage assemblage, and endscrapers, notches, splintered pieces and retouched blanks are the most common retouched tools.
+Stemmed and winged points, laurel leaves, and shouldered points were also identified (Cascalheira et al., 2013).
+
+The Proto-Solutrean horizons yielded more than 26,500 artifacts (including chips which correspond to \~65-70% of the assemblage and shatter \~20%), \~11,000 attributed to the bottom of Terrace’s level 5, and \~15,600 attributed to the top of level 5 and level 4E.
 Both horizons are characterised by knapping strategies dominated by unidirectional prismatic cores, with little platform preparation, focused on the obtention of bladelets (in bottom level 5) and flakes.
@@
 
-Previous studies showed that the main raw materials used at the site are chert, quartz and greywacke, with the percentages of these changing between occupations.
-For example, while studies show a high degree of quartz in the lower Proto-Solutrean levels (Belmiro et al., 2021), in the Solutrean occupations of the Shelter a dominance of chert was observed (Cascalheira, 2010).
-Similarly, different raw materials have been observed to play specific roles, such as the case of chert for the production of a high percentage of blanks and retouched tools.
-While quartz and greywacke can be found locally, in the streams and adjacent to the site, chert sources are further away, although still within a \~20 km radius of the site.
-Our previous study showed, as expected, the existence of local chert in the assemblages, originating from the Lower Jurassic outcrops in proximity to the site (Belmiro et al., 2025).
-As previously mentioned, a portion of non-local cherts was also identified, with some identified sources, such as the Cretaceous outcrops of Central Portugal (Lisbon and Rio Maior areas) and the Upper Jurassic oolitic/peloidal cherts of the Betic Systems in southern Spain.
-This suggests contacts or long-range mobility patterns of the groups occupying Vale Boi all throughout the UP, although in different degrees, possibly related to changes in the settlement patterns and duration of site occupation.
-
-Both these areas have been previously interpreted as seasonal residential camps, repeatedly used for extended stays, due to the abundance of lithic debitage, stone tools, heat-cracked rocks related to grease rendering activities, large quantities of faunal remains (both marine and terrestrial) and the presence of ornaments and portable art [@manneIntensiveSubsistencePractices2012].
-This is true for most of the occupations (levels), with exception of the Early Gravettian occupations of the Terrace area and Magdalenian occupation of the Shelter.
-
-Using the Whole Assemblage Behavioral Index (WABI), previously used in other studies (Riel-Salvatore, 2010; Riel-Salvatore & Barton, 2004), Cascalheira et al., (2017) used the lithic results from the Shelter Solutrean levels (Cascalheira, 2010; Cascalheira et al., 2013) and Magdalenian levels (Mendonça, 2009), and the Gravettian levels of Terrace (Marreiros et al., 2015) to reconstruct and compare the mobility strategies of the several occupations, through the lithic volumetric density and the frequency of retouched tools.
-This method allowed the confirmation that, except from the Early Gravettian and Magdalenian occupations, all other occupations are composed of high-density assemblages with low degree of retouch, correspondent to a residential base-camp occupation, for extended periods of time.
+Previous studies showed that the main raw materials used at the site are chert, quartz, greywacke and smaller amounts of chalcedony, with the percentages of these changing across occupations.
+Notably, the lower Proto-Solutrean levels show high frequencies of quartz (Belmiro et al., 2021), whereas chert becomes the dominant raw material in the Solutrean occupations of the Shelter (Cascalheira, 2010).
+Likewise, different raw materials appear to have been used for specific purposes,with chert being particularly associated  with the production of high percentage of blanks and retouched tools, and greywacke being extensively used for anvils (Horta et al., 2019)
+
+Quartz and greywacke are locally available from streambeds near the site, whereas regional chert and chalcedony sources lie further afield, within a \~20 km radius.
+Our previous studies showed, as expected, the existence of local cherts in the assemblages (\@fig-cherts a), originating from the Lower Jurassic (cherts) and Triassic (chalcedony) outcrops in proximity to the site (Belmiro et al., 2025).
+Despite petrographic differences between cherts and chalcedony, for this study any chert varieties (e.g., chalcedony or jasper) are referred to as chert (Luedtke, 1978).
+A significant  portion of non-local cherts was also identified (\@fig-cherts b-c), with some identified sources, such as the Cretaceous outcrops of central Portugal (Lisbon and Rio Maior areas) and the Upper Jurassic oolitic/peloidal cherts of the Betic Systems in southern Spain.
+This indicates varying degrees of contact or long-range mobility throughout the UP, likely influenced by shifts in settlement patterns and the duration of site occupation.
+
+```{r}
+#| label: fig-cherts
+#| fig-cap: "Simplified scheme of chert types identified in the UP sequence of the Terrace and Shelter areas of Vale Boi: a) local chert macroscopic variability (T1 to T5), including chalcedony (T1; top left); b) non-local chert types present in all occupations except for T11 which is only present in the Solutrean; c) TL09 (left) and TL06 (right) varieties within the trace lithologies group (TL) composed of non-local cherts with 1-3 samples per TL type and limited to specific archaeological levels and areas.
+Summarised from Belmiro et al., (2025)."
+
+```
+
+Previous research suggests that both the Terrace and Shelter functioned as seasonal residential camps, repeatedly occupied for extended periods.
+This interpretation is based on high-density lithic assemblages, evidence of tool manufacture, heat-cracked rocks associated with grease rendering, diverse faunal remains (marine and terrestrial), and personal ornaments and portable art (Manne et al., 2012).
+This is true for most of the UP occupations, although with some exceptions.
+
+Cascalheira et al.
+(2017) applied the Whole Assemblage Behavioral Index (WABI) (Riel-Salvatore, 2010; Riel-Salvatore & Barton, 2004) to reconstruct and compare mobility strategies at Vale Boi.
+By analysing lithic volumetric density and retouched tool frequencies from the Shelter’s Solutrean (Cascalheira, 2010; Cascalheira et al., 2013) and Magdalenian (Mendonça, 2009) levels, as well as the Terrace’s Gravettian levels (Marreiros et al., 2015), the authors confirmed that most occupations corresponded to long-term residential base camps, composed of high-density assemblages with low degree of retouch, except for the Early Gravettian and Magdalenian occupations.
 Regarding the Terrace area, however, the results were based on previous excavations (\<2012), and the identification of new layers and reorganization of the occupations is yet to be tested through the use of lithic techno-typological methods.
 
-A similar conclusion was suggested based on the interpretation of the chert sourcing and proportions in the several UP occupations of the Terrace and Shelter of Vale Boi, and following other models relating local/non-local raw materials with mobility (Grove et al., 2023; Surovell, 2009).
-The high proportions of non-local chert (\~50%) in the Gravettian occupations (levels 6 and 7), and gradual reduction in the following occupations, were interpreted as a possible result of settlement changes, where the Gravettian occupations corresponded to short-term occupations, and the other occupations corresponded to residential, long-term occupations with more infrequent moves (Belmiro et al., 2025).
-
-## Lithic analysis
-
-The current study includes the statistical correlation between the techno-typological analysis and the geoarchaeological study of the chert lithic assemblages of the UP of Vale Boi.
-As mentioned above, previous studies have individually analysed the lithic assemblages from Vale Boi focusing on different subjects and methods, obtaining different kinds of data.
-Specifically, our previous study focused on the macroscopic and petrographic characterisation of cherts, and the identification of distance to sources and proportions through the several UP technocomplexes.
-This study provided the basis for understanding which cherts were being used at the site, where they were coming from and how/why their use became more or less intense.
-In that same study, we used previously known data related to settlement type and seasonality to contextualise the changes in chert sources and proportions through time, to better understand the possible strategies of raw material procurement (Hahn, 2002).
-However, and as mentioned in Belmiro et al.
-2025, correlating the obtained data with technological analysis data provides a key step to better understand how these raw materials were being introduced into the site and used, through the correlation with data such as type of lithic class, since it may point towards the specific mobility of artifacts (Hahn, 2002).
-
-This is especially relevant in a site like Vale Boi, where several technological studies have been key to understand the cultural characteristics of the groups inhabiting throughout the UP, and their possible relations within the UP in Iberia.
-By using previously acquired data, it becomes possible to integrate different datasets to provide a wider picture of UP occupations at Vale Boi, and using the potential of lithics to inform us of hunter-gatherer lifeways.
-
-As such, this study includes datasets which have been previously interpreted and published, most of which can also be found in online repositories and freely consulted.
-As previously mentioned, the macroscopic and petrographic data of cherts was previously collected and published by our team (Belmiro et al., 2025), and can also be found in an online repository (OSF\*\*).
-This study focused on the UP chalcedony and chert assemblages of Vale Boi, especially the Gravettian (levels 6 and 7), Proto-Solutrean (levels 4E and 5) and Solutrean (levels 4 to 4D) from the Terrace area and the Solutrean (levels A to C) from the Shelter area.
-The methodology used in the study, and supporting documents, including the chert type characterisation and analysis data dictionaries can be found in the article and the online repository.
-
-The technological analysis datasets have been collected by our team and published in previous works.
-The Solutrean assemblage technological analysis was published by Cascalheira (-Cascalheira, 2010), as part of a Master’s thesis and following publications, with the study of levels A to C of the Shelter area.
-The dataset used for this study was not originally published, but can be consulted in our online repository (++).
-Given the inexistence of technological studies for the Solutrean levels of the Terrace with a representative amount of lithics (level 4), the Solutrean analysis in this study is limited to the Shelter area.
-However, and as seen in Belmiro et al.
-(2025), the raw material patterns seem to show little differences between the Solutrean in both areas, thus coupled with the similar radiocarbon dates between area (c.
-24-22 and 24-20 ka cal BP), we assume that they are representative of similar, long-term, residential occupations which aptly characterise the Solutrean groups of Vale Boi.
+A similar conclusion was suggested based on the interpretation of the chert sourcing and proportions in the several UP occupations of the Terrace and Shelter of Vale Boi and following other models relating local/non-local raw materials with mobility (Grove et al., 2023; Surovell, 2009).
+The high proportions of non-local chert (\~50%) in the Gravettian occupations (levels 6 and 7), and gradual reduction in the following occupations, were interpreted as a possible result of settlement changes, where the Gravettian occupations corresponded to short-term occupations, and the other occupations corresponded to residential, long-term occupations with more infrequent moves (Belmiro et al., 2025).\
+
+## Datasets and analysis
+
+The current study focuses on the correlation between the techno-typological analysis and the geoarchaeological study of the chert lithic assemblages of the UP of Vale Boi.
+It integrates previously collected datasets, which have been published or are openly accessible in online repositories, allowing for transparency and reproducibility.
+
+The macroscopic and petrographic data of cherts was previously collected and published by our team (Belmiro et al., 2025) and can also be found in an online repository (<https://www.doi.org/10.17605/OSF.IO/DBFT2>).
+This study focused on the UP chert assemblages of Vale Boi, especially the Gravettian (levels 6 and 7), Proto-Solutrean (levels 4E and 5) and Solutrean (levels 4 to 4D) from the Terrace area and the Solutrean (levels A to C) from the Shelter area.
+The methodology used in the study, dataset and supporting documents can be found in the article and the online repository.
+
+The Solutrean assemblage technological analysis was published by Cascalheira (2009) as part of a Master’s thesis and following publications (e.g., Cascalheira, 2010; Cascalheira et al., 2013), with the study of levels A to C of the Shelter area.
+The dataset used for this study was not originally published but can be consulted in our online repository (<http://www.doi.org/10.17605/OSF.IO/UWMPK>).
+
+The Solutrean analysis in this study is limited to the Shelter area due to the inexistence of technological studies for the better preserved Solutrean levels of the Terrace (e.g., level 4).
+However, and as seen in Belmiro et al. (2025), the raw material patterns seem to show little differences between the Solutrean in both areas; coupled with the similar radiocarbon dates between the areas (c. 24-22 and 24-20 ka cal BP), we assume that they are representative of similar, long-term, residential occupations which aptly characterise the Solutrean groups of Vale Boi.
+
 The Proto-Solutrean technological analysis was the result of a Master’s thesis (Belmiro, 2020) and following publication (Belmiro et al., 2021), with the study of levels 4E and 5 of the Terrace area.
-The dataset used for this study can be consulted in the original online repository (++), as well as in our RStudio reproducible project, available in our online repository (\*\*).
-Finally, the Gravettian technological analysis was the result of a recent and unpublished study, focusing on the study of levels 6 and 7 of the Terrace area.
-This dataset can also be found in our online repository (++).
-Given the use of different datasets, collected through a wide span of time and by different researchers, a data dictionary with the variables used in this study can also be found in the Supplementary materials and in our online repository.
-
-An inherent caveat of using datasets collected by different researchers is related to the use of different methodologies, variables and measurements.
-This caveat is avoided in this study since the literature, approach and analysis method used in all data collection processes was similar.
-The lithic analysis adopted an attribute-based methodology, which aims to describe morphological and metrical attributes of technological classes.
-The attributes analysed followed those present in specialized literature, such as Brézillon (1968), and Tixier and Inizan (1980), paired with other Paleolithic lithic attribute analyses (see e.g., Bicho, 1992; Zilhão, 1997; Cascalheira, 2019; Almeida, 2000; Scerri et al., 2014; Tostevin, 2013).
+The dataset and analysis can be consulted in the original online repository (<http://www.doi.org/10.17605/OSF.IO/456EG>).
+
+Finally, the Gravettian technological analysis was the result of a recent and unpublished study, focusing on the chert lithic artefacts from levels 6 and 7 of the Terrace area (\> 2012 excavations).
+This dataset and preliminary technological analysis results can be found in our online repository (<http://www.doi.org/10.17605/OSF.IO/UWMPK>) and in the Online Supplementary Materials (SOM 1).
+
+One challenge of integrating datasets collected by different researchers is variability in methodologies and recorded variables.
+However, this study minimizes such inconsistencies by ensuring methodological alignment across datasets, as all data collection followed standardized attribute-based lithic analysis protocols.
+The lithic analysis datasets used in this study adopted an attribute-based methodology, which aims to describe morphological and metrical attributes of technological classes.
+The attributes analysed followed those present in specialized literature, such as Brézillon (1968), and Tixier and Inizan (1980), paired with other Palaeolithic lithic attribute analyses (see e.g., Bicho, 1992; Zilhão, 1997; Cascalheira, 2019; Almeida, 2000; Scerri et al., 2014; Tostevin, 2013).
 
@@
 Some differences in the collection process, especially between the study by Cascalheira (2010) and the following studies, required homogenization.
-For example, where in Cascalheira (2010) retouch was an independent variable, with a Yes or No option followed by a variable to classify the type of retouch, in Belmiro (2020) and Belmiro et al.
-(2021), retouch was identified in the Class variable as RetouchedPiece, followed by a variable to choose the retouch typology.
-In this situation, the Solutrean dataset was corrected, so that all artifacts with retouch were classified as RetouchedPiece in the Class variable, instead of the original classification.
-The original datasets without alterations can be found in the online repository, as all the alterations to the original datasets were done through R Studio and can be found in the original R Script.
-Finally, technological and raw material data had to be merged in a single dataset.
-This was done through the merging of the several datasets through the artifact identifier “ID”, the sequential identifier which is attributed to each total station plot during excavation, whether relating to individual artifacts plotted in their exact location, or bucket coordinates, relating to the center of the excavated area, and attributed to all artifacts found in the sieved sediment in the bucket.
-Since the raw material study only focused piece plotted artifacts with individual coordinates (leaving out bucket artifacts), as these correspond to artifacts with dimensions superior to 2 cm or complete bladelets (thus excluding small fragments and chips), the final dataset used in this study does not directly correspond to the full datasets analysed in the original studies.
+For example, where in Cascalheira (2010) retouch was an independent variable (e.g., Class as flake, Retouch as Yes), in Belmiro et al. (2021), retouch was identified as Class variable.
+In this situation, the Solutrean dataset was modified, so that all artifacts with retouch were identified in the Class variable, instead of the original classification.
+All changes applied to the original data made for homogenisation between datasets can be consulted in the R Studio scripts in our online repository (<http://www.doi.org/10.17605/OSF.IO/UWMPK>).
+
+To integrate technological and raw material data, we merged multiple datasets using the artifact identifier (“ID”), which corresponds to total station plot records.
+This identifier links artifacts to either precise excavation coordinates (for individually plotted pieces) or generalized spatial data (for sieved assemblages).
+
+Given that the raw material analysis focused exclusively on individually piece-plotted artifacts (dimensions \>2 cm), the final dataset does not include all excavated material but retains the most diagnostically relevant pieces for analysis.
+As a result, the final dataset used in this study does not directly correspond to the full datasets analysed in the original studies (\@tbl-samples).
 Similarly, the raw material study used only currently accessible artifacts from the assemblages.
-This is especially relevant for the Shelter Solutrean chert assemblage, where shatter is absent from the sample due to the availability for study.
-This stands as a caveat for the current study, since it is using only portion of the assemblages, leaving out possibly important debris.
-Despite this, the current study still offers a novel and key approach to the study of lithic assemblages at the site, with a methodology that includes all the knapping sequence, of a large number of artifacts from the UP sequence of a single site, as seen in table \*\*, which shows the number of artifacts used in the final analysis of this paper, after corrections, homogenization and merging.
+This is especially relevant for the Shelter Solutrean chert assemblage, where shatter is absent from the sample due to the lack of availability for study.
+
+A limitation of this study is its reliance on a subset of the total lithic assemblage, which excludes small debris.
+While this may omit some aspects of on-site knapping activities, the analysed sample remains representative of the broader technological trends at Vale Boi throughout the UP sequence.
 
@@
 #| label: tbl-samples
-#| tbl-cap: "Sample absolute number comparison between original datasets (chert and chalcedony only) for the Solutrean occupation of the Shelter area (levels A-C) published in Cascalheira (2010), the Proto-Solutrean occupation of the Terrace (levels 4E and 5) published in Belmiro et.
-al., (2021) and the Solutrean occupation of the Terrace (levels 6-7) and the final datasets used in the study after filtering individually plotted IDs (excluding bucket points) and merging with the geoarchaeological dataset published in Belmiro et al., (2025)."
+#| tbl-cap: "Sample absolute number comparison between original technological analysis datasets (chert and chalcedony only, excluding chips) for the Solutrean occupation of the Shelter area (levels A-C) published in Cascalheira (2010), the Proto-Solutrean occupation of the Terrace (levels 4E and 5) published in Belmiro et.
+al., (2021) and the Gravettian occupation of the Terrace (levels 6-7) and the final datasets used in the study after filtering individually plotted IDs (excluding bucket points) and merging with the geoarchaeological dataset published in Belmiro et al., (2025)."
 
@@
   ~Assemblage, ~`Original dataset (n)`, ~`Merged dataset (n)`,
-  "Shelter level A-C", 3359, 910,
-  "Terrace level 4E-5", 1821, 635,
+  "Shelter level A-C", 3290, 910,
+  "Terrace level 4E-5", 1413, 635,
   "Terrace level 6-7", 781, 766)
 
-knitr::kable(head(table_samples))
-
-```
-
-After the homogenization and merging of the datasets, descriptive statistical analysis was applied to describe and interpret the relation between the raw material and technological data.
-As mentioned above, the goal of the study was to test two hypotheses related to the technological organisation of local and non-local cherts, and between different UP occupations.
-To test these hypotheses, we evaluated several parameters for each dataset and assemblage and compared the results between chert types and assemblages.
-The parameters evaluated in the current study were: P1) frequency of retouched tools; P2) frequency of cores; P3) frequency of maintenance products; P4) debitage sizes; P5) tools-to-debitage ratios; P6) core sizes; P7) blank-to-core ratios; P8) tool diversity; P9) tool specialization.
-Debitage and core sizes were analysed through descriptive statistics (mean and sd tables available in the SOM\*\*) and plotted through boxplots and through non-parametric statistical analyses (Mann-Whitney U test).
-The ratios were obtained through the division of each group (e.g., tools by debitage and blanks by core), by chert type and assemblage.
-All other parameters were observed through number and percentages tables.
-All the analyses were done through code in R Studio.
-
-The entirety of the R code used for the analysis, datasets, and visual representations contained in this paper can be accessed through our online research compendium.
+knitr::kable(table_samples)
+
+```
+
+Following data homogenization and integration, we conducted descriptive statistical analyses and non-parametric statistical analyses to examine correlations between raw material use and technological patterns and test expectations related to the technological organisation of local and non-local cherts (see @tbl-expectations), and between different UP occupations.
+To test these expectations, we evaluated 9 parameters for each dataset and assemblage; the parameters and their calculation/measurement methods are detailed in @tbl-parameters.
+
+```{r}
+#| label: tbl-parameters
+#| tbl-cap: "Analysis parameters used in the study, calculation/measurement used for each parameter, and unit of the parameter results after analysis."
+
+table_parameters <- tibble::tribble(
+  ~Parameter, ~Calculation/Measurements, ~Unit,
+  "P1 - Frequency of retouched tools", "N retouched tools / Total assemblage * 100", "%",
+  "P2 - Frequency of cores", "N cores / Total assemblage * 100", "%",
+  "P3 - Frequency of maintenance products", "N maintenance products / Total assemblage * 100", "%",
+  "P4 - Debitage sizes", "Length, Maximum Width, Weight", "mm",
+  "P5 - Tool to debitage ratio", "N tools / N debitage", "ratio",
+  "P6 - Core sizes", "Length, Maximum Width, Weight", "mm",
+  "P7 - Blank to core ratio", "N blanks / N cores", "ratio",
+  "P8 - Tool diversity", "N of retouched typologies", "N",
+  "P9 - Tool specialization", "N of retouched blanks vs other typologies", "N")
+
+```
+
+The entirety of the R code used for the analysis, datasets, and visual representations contained in this paper can be accessed through our online repository (<http://www.doi.org/10.17605/OSF.IO/UWMPK>).
 We used the rrtools package by Marwick et al.
-[-@marwickPackagingDataAnalytical2018] to create a research compendium and write a reproducible journal article.
-The provided files include the complete set of raw data used in the analysis, along with a custom R project [@wickhamPackagesOrganizeTest2015] containing the code required to generate all tables and figures.
-To enable maximum reuse, the code is made available under the MIT license, data under CC‐0, and figures under CC‐BY [additional details can be found in @marwickComputationalReproducibilityArchaeological2017].
+(2018) to create a research compendium and write a reproducible journal article.
+The provided files include the complete set of raw data used in the analysis, along with a custom R project (Wickham, 2015) containing the code required to generate all tables and figures.
+To enable maximum reuse, the code is made available under the MIT license, data under CC‐0, and figures under CC‐BY (additional details can be found in Marwick, 2017).
 
@@
 
-The results are presented following the analysis parameters, focusing on the comparison between wider chert groups (local and non-local cherts) and between assemblages.
-A complete description of the assemblages and detailed tables for each analysis presented in this study can be found in the supplementary materials (SOM ++).
-
-In all assemblages, blanks and blank fragments are the most frequent class of lithic artefacts and in all chert types (\>40% of the assemblages).
-Cores are, as expected, found in smaller percentages in all assemblages (\~3-8%).
-Cores are present in local cherts in larger numbers than in non-local types (P2); their presence is noted in almost all non-local cherts in all assemblages, except for the TL group.
-In the Proto-Solutrean assemblage however, non-local chert core presence is small and limited to T7-9 (n=4 complete cores and n=2 core fragments).
-Albeit in very small numbers in all assemblages (n\<10), core preparation and maintenance products (P3) are also present in local and non-local cherts, especially in the non-local types which show the larger sample number such as T6-7.
-The presence of knapping by-products such as shatter is also seen in both local and non-local cherts in the Gravettian and Proto-Solutrean assemblages.
-Retouched tools (P1) were identified in all chert types from all studied assemblages; in general, all assemblages show a similar pattern where local cherts show smaller percentages of retouched tools than non-local cherts.
-However, the assemblages show clear differences in the number of retouched tools.
+Results are structured according to the predefined analytical parameters, comparing local and non-local cherts (\@fig-cherts) across different assemblages.
+A comprehensive description of each assemblage and detailed tables for all analyses are provided in the supplementary materials (**SOM 1**).
+
+Blanks and blank fragments represent the most abundant lithic class in all assemblages, accounting for over 40% of artifacts across all chert types (\@fig-general-graphs).
+As expected, cores are present in lower frequencies (\~3-8%) but are more numerous in local cherts than in non-local types and absent from the TL group.
+In the Proto-Solutrean assemblage, non-local chert cores are rare, with only T7-9 yielding a small number (n=4 complete cores, n=2 core fragments).
+Albeit in very small numbers in all assemblages (n\<10), core preparation and maintenance products are also present in local and non-local cherts, especially in the non-local types which show the larger sample number such as T6-7.
+The presence of knapping by-products such as shatter is also seen in both local and non-local cherts in the Gravettian (\@fig-general-graphs a)  and Proto-Solutrean (\@fig-general-graphs b) assemblages.
+Retouched tools are present in all chert types across all assemblages.
+In general, local cherts exhibit lower retouch frequencies than non-local cherts, though the total number of retouched tools varies significantly between assemblages.
 
@@
 
-The Gravettian assemblage shows higher percentages of retouched tools (\~21%) than the other assemblages (\<10%).
-Detailed tables with tool typology and counts by chert type and assemblage can be found in the SOM (tables +++).
-Tool diversity (P8) is higher in non-local cherts in the Gravettian assemblage, although with high frequency of less specialised tools such as notches and retouched tools, alongside hunting implements such as backed bladelets (P9).
-The high frequency of burins seen in this assemblage may be related to their use as cores, representing a knapping strategy more than the presence of retouch.
+The Gravettian assemblage exhibits the highest proportion of retouched tools (\~21%; @fig-general-graphs **a**), compared to \<10% in other assemblages.
+Tool diversity is greater among non-local cherts, which contain a high frequency of generalized tools (e.g., notches, retouched pieces) alongside hunting implements such as backed bladelets.
+The abundance of burins and burin spalls suggests they may have functioned as cores, indicating a technological strategy rather than solely a focus on retouch (**SOM 1** **Table s2**).
 In comparison to the Gravettian, other assemblages show smaller numbers of retouched tools.
-While in the Proto-Solutrean no pattern was identified between local and non-local retouched tools, in the Solutrean assemblage both tool diversity (P8) and tool specialization (P9) are higher in local cherts.
-
-The relation between blank and core numbers was explored through a blank-to-core ratio (P7; Figure 1 and table s++).
-In all assemblages, local cherts have generally smaller ratios than non-local cherts, although sample sizes for cores are variable (graph ++).
-For the Gravettian assemblages, the results show that non-local cherts show ratios above 7 (with the exception of T10), while local cherts have ratios located below 4.
+While in the Proto-Solutrean (**SOM 1** **Table s4**) no significant difference was identified between local and non-local retouched tools, in the Solutrean assemblage (**SOM 1** **Table s6**) both tool diversity and tool specialization are higher in local cherts.
+
+Retouched tools were further explored through the tools-to-debitage ratio (\@fig-ratios a and **SOM 1 Table s8**).
+In general, the results show that non-local chert types have higher retouch-to-debitage ratios than local types, albeit with differences between assemblages.
+The Gravettian assemblage shows the highest tool-to-debitage ratios compared to the remaining assemblages; non-local cherts show ratios above 0.9, while local cherts show values below 0.5.
+In the Proto-Solutrean non-local cherts show ratios between 0.3 to 0.5, while local cherts show ratios below 0.3.
+This pattern is less clear in the Solutrean assemblage since the ratios are the lowest of all assemblages.
+Non-local cherts show ratios above 0.1 (with exception of T8) while local cherts show ratios below 0.1 (with exception of T1; **SOM 1 Table s8 c**).
+
+Blank-to-core ratios were analysed across assemblages (\@fig-ratios b, **SOM 1 Table s7**).
+In all cases, local cherts generally exhibit lower blank-to-core ratios than non-local cherts, though variability in core sample sizes may influence these patterns.
+For the Gravettian assemblages, the results reveal that non-local cherts show ratios above 7 (with the exception of T10), while local cherts have ratios located below 4.
 The Solutrean assemblage also shows higher ratios for non-local cherts (\>12), with local cherts showing ratios below 7.
 The Proto-Solutrean assemblage has most blank-to-core ratios below 10, except for non-local T7.
-However, non-local cores are poorly represented within the assemblage (table ++).
+However, non-local cores are poorly represented within the assemblage (\@tbl-ratios b; **SOM 1 Table s3**).
 
@@
 td_ratio <- td_ratio + theme(legend.position = "none")
-ggpubr::ggarrange(bc_ratio, td_ratio, legend_b, labels = c("a", "b", ""), ncol = 1, heights = c(3, 3, 1))
-
-```
-
-Retouched tools were further explored through the tools-to-debitage ratio (P5; Figure 2 and table s++).
-In general, the result show that non-local chert types have higher retouch-to-debitage ratios than local types, albeit with differences between assemblages.
-The Gravettian assemblage shows the highest tool-to-debitage ratios compared to the remaining assemblages; non-local cherts show ratios above 0.9, while local cherts show values below 0.5.
-In the Proto-Solutrean non-local cherts show ratios between 0.3 to 0.5, while local cherts show ratios below 0.3.
-This pattern is less clear in the Solutrean assemblage since the ratios are the lowest of all assemblages.
-Non-local cherts show ratios above 0.1 (with exception of T8) while local cherts show ratios below 0.1 (with exception of the local chalcedony).
-
-The sizes of blanks (P4) and cores (P6) are explored through scatterplots of width and length (Figure 3 a) and boxplots for weight (Figure 3 b); complete tables with descriptive statistics and non-parametric statistical test results can also be found in SOM ++, table s++.
-The scatterplots show different patterns in the several analysed classes.
-
-```{r}
-#| label: fig-gravettian-boxplots
-#| fig-cap: Boxplots Gravettian
+ggpubr::ggarrange(td_ratio, bc_ratio, legend_b, labels = c("a", "b", ""), ncol = 1, heights = c(3, 3, 1))
+
+```
+
+The sizes of blanks and cores were explored through scatterplots of width and length (\@fig-gravettian-plots**, @fig-proto-plots**, \@fig-solutrean-plots a) and boxplots for weight (\@fig-gravettian-plots, **@fig-proto-plots**, \@fig-solutrean-plots b) and through non-parametric statistical tests; complete tables with descriptive statistics and non-parametric statistical test results can be found in **SOM 1 Tables s9-11** and **s12**, respectively.
+
+The scatterplots and boxplots show different patterns in the several analysed classes.
+In the Gravettian assemblage (@fig-gravettian-plots), non-local chert blanks  are significantly smaller than local chert blanks in both length (\<25 mm) and weight (\<5 g), with p-values of 0.00009\*\* and 0.012, respectively.
+While local chert blanks also cluster below 25 mm in width and length, they display greater variability and a higher number of outliers.
+The same patterns can be seen in the Proto-Solutrean assemblage (@fig-proto-plots) , where length and weight also show statistically significant differences between chert types (0.046\* and 0.004), with non-local types showing smaller values.
+In comparison, in the Solutrean assemblage (@fig-solutrean-plots), only non-local chert blank length seems to show statistically significant smaller values (p value of 0.0005\*\*) than local chert blanks, being mostly concentrated below \~30 mm, while local chert blank lengths are more variable.
+While width seems to be more variable in both local and non-local chert blanks, the statistical analysis shows that non-local chert width is significantly higher than local chert blanks (p value of 0.02\*).
+
+```{r}
+#| label: fig-gravettian-plots
+#| fig-cap: "Gravettian assemblage blank and core measurements and weight descriptive statistics in local and non-local chert types.
+a) Scatterplots of width and length.
+b) Boxplots of weight; diamond symbol represents mean values within each group."
 
@@
 ```{r}
-#| label: fig-proto-boxplots
-#| fig-cap: Boxplots Proto-solutrean.
+#| label: fig-proto-plots
+#| fig-cap: "Proto-Solutrean assemblage blank and core measurements and weight descriptive statistics in local and non-local chert types.
+a) Scatterplots of width and length.
+b) Boxplots of weight; diamond symbol represents mean values within each group."
 
@@
 ```{r}
-#| label: fig-solutrean-boxplots
-#| fig-cap: Boxplots Solutrean.
+#| label: fig-solutrean-plots
+#| fig-cap: "Solutrean assemblage blank and core measurements and weight descriptive statistics in local and non-local chert types.
+a) Scatterplots of width and length.
+b) Boxplots of weight; diamond symbol represents mean values within each group."
 
@@
 
-In the Gravettian assemblage, non-local chert blanks (P4) are smaller than local chert blanks, especially regarding length (\~ \<25 mm) and weight (\<5 gr), with statistically significant differences (p value of 0.00009\*\*\* and 0.012, respectively).
-The local chert blanks, although mostly concentrated below 25 mm width and length and with weights below 5 gr, have the largest number of outliers and variability in terms of size.
-The same patterns can be seen in the Proto-Solutrean assemblage, where length and weight also show statistically significant differences between chert types (0.046\* and 0.004), with non-local types showing smaller values.
-In comparison, in the Solutrean assemblage, only non-local chert blank length seems to show statistically significant smaller values (p value of 0.0005\*\*) than local chert blanks, being mostly concentrated below \~30 mm, while local chert blank lengths are more variable.
-While width seems to be more variable in both local and non-local chert blanks, the statistical analysis shows that non-local chert width is significantly higher than local chert blanks (p value of 0.02).
-
-Regarding cores (P6), the differences in sizes and weights seem between local and non-local cherts seem to be less obvious than those observed in blanks.
-In the Gravettian assemblage, there is a cluster of non-local chert cores below the 25 mm width and length, while in general, local chert cores seem to be more variable.
+For cores, size and weight differences between local and non-local cherts are less pronounced than those observed in blanks.
+In the Gravettian assemblage (@fig-gravettian-plots), there is a cluster of non-local chert cores below the 25 mm width and length, while in general, local chert cores seem to be more variable.
 Especially regarding weight, T6 cores (non-local) show lighter cores (\~12 gr means) compared to local cherts.
-In the Solutrean assemblage, weight is the only measurement where differences can be seen between local and non-local chert cores; similarly to the Gravettian assemblage, T6 cores show slightly smaller cores (\~18 gr means) than local cherts (\~22 gr means).
+In the Solutrean assemblage (@fig-solutrean-plots), weight is the only measurement where differences can be seen between local and non-local chert cores; similarly to the Gravettian assemblage, T6 cores show slightly smaller cores (\~18 gr means) than local cherts (\~22 gr means).
 Despite these differences, the statistical tests applied to the dimensions between local and non-local chert cores show no statistically significant differences between the groups (p \> 0.05).
-The Proto-Solutrean assemblage shows a small sample of non-local chert cores (n=4), not allowing a clear identification of patterns between the groups; alike the other assemblage however, local chert cores show a high range of sizes (\~15-50 mm width and length) and weights (\~3-85 gr).
-
-Following the analysis parameters previously described, the summarised results for each parameter can be found in @tbl-results-sum, and can be summarised as follows:
+The Proto-Solutrean assemblage (@fig-proto-plots) contains a limited sample of non-local chert cores (n=4), preventing clear pattern identification.
+However, as in other assemblages, local chert cores display a broad size range (\~15-50 mm in width/length, \~3-85 g in weight).
+
+Following the analysis parameters, the results described above can be found in @tbl-results-sum, and can be summarised as follows:
 
 In the Gravettian assemblage, retouched tools (P1) are present in higher frequencies in the non-local cherts, as well as in core preparation and maintenance products (P3).
-Cores are present in even in non-local chert types (P2).
+Cores are present even in non-local chert types (P2).
 Debitage sizes (P4) show smaller sizes and weights for non-local cherts.
@@
 Core sizes (P6) are slightly smaller in T6, a non-local chert, and the blank-to-core ratios (P7) are higher in non-local cherts.
-Tool diversity (P8) is higher in non-local cherts, although with high frequency of less specialised tools such as notches and retouched tools, alongside hunting implements such as backed bladelets (P9).
+Tool diversity (P8) is higher in non-local cherts, although with high frequency of less specialised tools such as notches and retouched blanks, alongside hunting implements such as backed bladelets (P9).
 The high frequency of burins seen in this assemblage may be related to their use as cores, representing a knapping strategy more than the presence of retouch.
 
-The Proto-Solutrean assemblages show a smaller presence of retouched tools than the Gravettian, and within the assemblage, retouched tools (P1) are slightly more frequent in non-local cherts.
-Cores of non-local chert types are found in very small numbers (P2), although knapping is likely occurring on site.
+Compared to the Gravettian, the Proto-Solutrean assemblage contains fewer retouched tools (P1), with non-local cherts exhibiting slightly higher retouch frequencies.
+Non-local chert cores (P2) are rare, though evidence suggests knapping occurred on-site.
 Core preparation and maintenance products (P3) are only present in non-local cherts, albeit in very small numbers (n=2).
@@
 #| label: tbl-results-sum
-#| tbl-cap: "Summarised results for analysis parameters between grouped chert types (local and non-local) by assemblage."
+#| tbl-cap: "Summarised results for analysis parameters between grouped chert types (local and non-local) by assemblage.
+Core sizes (P6) for the local Proto-Solutrean cherts are presented in mean values instead."
 
@@
 
-The results of the study align with our initial expectations, providing insights into mobility patterns, raw material use, and technological strategies throughout the Upper Paleolithic (UP) in southwestern Iberia.
-We initially anticipated differences in core reduction between local and non-local resources (Table ++, expectations a-b).
-However, these differences are influenced by the transportation strategies used to bring non-local cherts to Vale Boi.
-The presence of non-local chert cores at the site depends on whether these materials were transported as finished tools or blanks—likely part of an individual provisioning strategy (Kuhn, 2004)—or as trimmed cores or larger stone volumes, as seen in site provisioning strategies (Kelly, 1988; Roth & Dibble, 1998; Clark, 1987; Kuhn, 2004).
-
-Our findings indicate that cores were present in non-local chert groups across all studied assemblages, albeit in small numbers (e.g., in the Proto-Solutrean assemblage).
-While the smaller fraction of the assemblages—such as chips, which can serve as proxies for on-site knapping—was not analyzed, other knapping by-products, including shatter in non-local chert types (except in the Solutrean assemblage, where this category was not studied; see Section ++ Materials and Methods), as well as core preparation and maintenance products, may also serve as indicators of core transportation and on-site knapping.
-
-These findings suggest a strategy of transporting cores or nodules to Vale Boi, followed by on-site knapping for most non-local chert types.
-However, an exception appears in the case of certain TL cherts (TL2-6, TL9-10, TL14; Table S13).
-These artifacts, whose source remains unknown but do not match local cherts, lack both cores and associated knapping debris.
-This pattern suggests that these TL cherts may represent an instance of long-distance individual provisioning (Kuhn, 2004), where materials were transported as finished tools or blanks.
+Our findings confirm initial expectations regarding mobility patterns, raw material use, and technological strategies in southwestern Iberia during the UP.
+While core reduction differences between local and non-local cherts were anticipated (\@tbl-expectations, expectations a-b), these differences were largely shaped by transportation strategies used to bring non-local cherts to Vale Boi.
+As such, the occurrence of non-local chert cores at Vale Boi depended on whether these materials arrived as finished tools or blanks—likely reflecting an individual provisioning strategy (Kuhn, 2004)—or as trimmed cores and larger stone volumes, aligning with site provisioning models (Kelly, 1988; Roth & Dibble, 1998; Clark, 1987; Kuhn, 2004).
+
+Non-local chert cores appear in all studied assemblages, though in limited numbers (e.g., in the Proto-Solutrean).
+While chips—often proxies for on-site knapping—were not analysed, the presence of knapping by-products, such as shatter (except in the Solutrean assemblage; see **Section 2.2.
+Datasets and analysis**), and core preparation and maintenance products in non-local cherts suggests active core reduction and on-site knapping.
+
+Our results indicate that most non-local cherts were introduced to Vale Boi as cores or nodules, subsequently knapped on-site.
+However, an exception is seen in TL cherts (TL2-6, TL9-10, TL14; **SOM 1 Table S13**), which lack associated cores and knapping debris.
+This absence suggests that TL cherts were introduced to the site primarily as finished tools or blanks, aligning with long-distance individual provisioning models (Kuhn, 2004).
 
@@
 
-Our analysis identified core preparation and maintenance products in the studied assemblages, suggesting a possible effort to preserve raw material volume.
+Our analysis identified core preparation and maintenance products in the studied assemblages, suggesting a possible effort to preserve raw material volumes.
 However, these products were found in low percentages across all assemblages and chert types (\<4% and n\<7).
@@
 
-Core and blank sizes and ratios yielded better results.
-Our findings suggest that non-local cherts underwent more intensive reduction than local cherts, as evidenced by the higher blank-to-core ratio and the statistically significant smaller size and weight of blanks made from non-local cherts.
+Core and blank size comparisons provide further insights.
+Non-local cherts underwent more intensive reduction than local cherts, as indicated by their higher blank-to-core ratios and the significantly smaller size and weight of non-local chert blanks.
 These results align with previous studies, which propose that the intensive reduction of cores—manifested in smaller core sizes, smaller debitage products, and a higher amount of debitage per core—may be linked to lower mobility.
@@
 However, our results show that these differences are not limited to long-term occupations and low mobility.
-High blank-to-core ratios and smaller blank sizes in non-local cherts are also observed in the Gravettian occupation, previously interpreted as a short-term occupation which frequent moves (Cascalheira ++; Belmiro et al., 2025).
-Due to this, we may interpret that non-local raw materials were more intensively reduced than local cherts throughout all the UP, independent of the duration of the occupation.
+High blank-to-core ratios and smaller blank sizes in non-local cherts are also observed in the Gravettian occupation, previously interpreted as a short-term occupation with frequent moves (Cascalheira, 2017; Belmiro et al., 2025).
+Due to this, we may interpret that non-local raw materials were generally more intensively reduced than local cherts throughout all the UP, independent of the duration of the occupation.
 
@@
 These results may be explained by a combination of factors, including the availability of local lithic resources, transport, and knapping strategies focused on the intensive reduction and maintenance of non-local raw materials.
-Belmiro et al.
-(2023) highlight that chert nodules in the Algarve region vary between \~5-10 cm, with a maximum diameter of 20 cm.
-Although these nodules are abundant and likely accessible throughout the Upper Paleolithic, given their location in accessible cliffs, they tend to be small and prone to fracture when extracted directly from the source.
-Given the initial size of local chert nodules, it is possible that the small core sizes are not the result of intensive knapping, but rather reflect the initial size of the knapped volumes.
-
-In comparison, cherts from central Portugal—the probably source from T6 cherts from Vale Boi—can be found in larger volumes (\>20 cm), although smaller nodules are also present.
-This suggests that the small artifact sizes associated with this non-local resource are not necessarily due to limited resource availability, but rather an intentional choice.
-This decision may involve the transportability of smaller volumes or nodules, knapping strategies focused on intensive reduction and maintenance, or a combination of both.
+Belmiro et al. (2023) highlight that chert nodules in the Algarve region vary between \~5-10 cm, with a maximum diameter of 20 cm.
+Although abundant and easily accessible in exposed cliffs, local chert nodules are generally small and prone to fracturing when extracted, and normally occur fractured in secondary deposits associated to primary outcrops.
+As a result, the small size of local chert cores may not necessarily indicate intensive reduction but could instead reflect the initial size of available raw material volumes.
+
+In comparison, cherts from central Portugal—the probable source of T6 cherts from Vale Boi—can be found in larger volumes (\>20 cm), although smaller nodules are also present.
+These findings suggest that the small size of non-local chert artifacts was not necessarily constrained by raw material availability but was likely an intentional choice.
+This could reflect preferences for transportable nodules, strategic reduction to maximize resource utility, or a combination of both.
 This is further supported by the higher blank-to-core ratio observed in non-local raw materials, particularly T6.
@@
 The intensity of reduction and maintenance discussed thus far can also be inferred from the frequency and diversity of retouched tools.
-In fact, retouched tools are often key elements used in the literature to distinguish between differential use of local and non-local raw materials.
-Our results align with expected outcomes, as non-local cherts exhibit higher frequencies of retouched tools and higher tool-to-blank ratios across all assemblages.
-This supports the idea that non-local raw materials were actively procured and that greater effort was invested in tool production (Andrefsky, 1994; Perlès, 1992; Kelly, 1988) and the maintenance of existing artifacts through retouch (Kuhn, 2004).
-
-While the intensity of retouch aligns with the expected outcomes (Table ++, expectations 1c and 2c), tool diversity and typologies (Table ++, expectations 1d and 2d) differ from what was anticipated.
-It was expected that non-local cherts would show lower typological diversity, as the distance to the source and inferred mobility suggest that less specialized tools (e.g., retouched blanks) would be more suitable for various use-case scenarios (Shott, 1986; Odell, 1981; Siegel, 1984).
-
-However, with the exception of the Solutrean assemblage, our results show higher tool diversity in non-local cherts in the Gravettian assemblages, and similar tool diversity between chert types in the Proto-Solutrean assemblages.
-This suggests that tools were transported as either less specialized tools or blanks, which were then maintained on-site.
-Our results indicate that long-distance raw materials, particularly T6 and T7, were likely knapped at the site.
-Therefore, the high diversity in tool typology may reflect the transportation of raw material volumes and their subsequent knapping on-site, coupled with a possible preference for these materials in the production of retouched tools, contributing to their high typological variability.
-
-Despite this high variability, less specialized tools, such as retouched blanks, remain predominant in non-local cherts across all assemblages.
-Specialized tools (e.g., backed tools during the Gravettian occupation) are also present, supporting the interpretation that, due to the likely ease of access to these raw materials—either through high mobility or exchange between groups—these materials were knapped on-site.
-However, they were characterized by longer use-lives, extensive reduction, and a high presence of retouch, with a predominance of non-specialized formal tools.
-
-While the patterns identified so far have established a clear distinction in the use and maintenance of local and non-local resources, it is also important to discuss the possible reasons for the varying abundance and technological patterns of non-local raw materials throughout the Upper Paleolithic at Vale Boi, as well as the factors behind the differences observed across the different technocomplexes.
-
-Previous research has often focused on the concepts of abundance and quality in the procurement and use of non-local lithic resources over local ones to explain differences in proportion, use and maintenance.
+In fact, retouched tools are often key elements used in the literature to distinguish between differential use of local and non-local raw materials (Andrefsky, 1994; Bamforth, 1986; Kelly, 1988; Kuhn, 2004; Perlès, 1992).  Our results align with expected outcomes, as non-local cherts exhibit higher frequencies of retouched tools and higher tool-to-blank ratios across all assemblages.
+This supports the idea that non-local raw materials were actively procured, and that greater effort was invested in tool production (Andrefsky, 1994; Perlès, 1992; Kelly, 1988) and the maintenance of existing artifacts through retouch (Kuhn, 2004).
+
+While the high intensity of retouch in non-local cherts aligns with expectations (\@tbl-expectations, expectations 1c and 2c), their unexpectedly high tool diversity contradicts common mobility models which expect mobility to be linked with limited tool variability and an increase of easily reconfigurable tools (Odell, 1981; Shott, 1986; Siegel, 1984).
+Non-local cherts at Vale Boi contain a diverse set of tool types, although less specialized tools, such as retouched blanks, remain predominant in non-local cherts across all assemblages.
+This suggests that these materials were not only transported as unspecialised tools but as volumes that supported on-site tool production and diversification.
+
+While the patterns identified so far have established a clear distinction in the use and maintenance of local and non-local resources, it is also important to discuss the possible reasons for the varying abundance and technological patterns of non-local raw materials throughout the UP at Vale Boi, as well as the factors behind the differences observed across the different technocomplexes.
+
+Previous studies emphasize raw material abundance and quality as key factors driving the procurement and use of non-local lithic resources over local ones, shaping their proportions, use, and maintenance.
 In this context, the scarcity or low quality of local resources would drive the procurement of non-local materials and require more effort in preparing transportable cores and producing tools (Andrefsky, 1994; Perlès, 1992; Kelly, 1988).
-However, quality may become a less significant factor when non-local raw materials are easily accessible (Akerman et al., 2002; Torrence, 1996; Brown, 1999; Oestmo, \*\*\*).
+However, quality may become a less significant factor when non-local raw materials are easily accessible (Akerman et al., 2002; Torrence, 1996; Brown, 1999; Oestmo, 2017).
 Even in the presence of high-quality local materials, effort was still invested in transporting non-local resources (Roth & Dibble, 1998).
@@
 Regarding Vale Boi, quality does not appear to be a defining factor in raw material choice, as typologies typically associated with high-quality lithic resources are also present in other materials.
-For example, Solutrean bifacial tools and points were produced using local chert, which may be considered lower quality than Rio Maior Cretaceous chert due to its coarser texture and less homogeneous composition.
-This suggests that the choice of raw material was not solely dependent on quality.
+For example, Solutrean bifacial tools and points were produced using Lower Jurassic and Triassic local chert, which may be considered lower quality than Rio Maior Cretaceous chert due to its coarser texture and less homogeneous composition.
+This suggests that the choice of raw material was not solely dependent on quality, especially when techniques such as heat treatment—the process of heating lithic raw material to improve knapping quality (Schmidt et al., 2016)—have been identified during the Solutrean in European UP contexts (Bachellerie & Schmidt, 2021; Bordes, 1969; Inizan et al., 1976; Tiffagom, 1998).
 
@@
 
-In this case, the impact of raw material quality seems to be more related to the differences between lithological groups, which have varying textures and breakage patterns (e.g., between greywacke and chert), rather than within micro-cryptocrystalline quartz resources like chert.
-This distinction suggests that raw material characteristics, such as texture and fracture properties, may have played a more significant role in tool production than raw material quality alone.
-
-Finally, a visual inspection of a cache found in the Shelter area attributed to the Gravettian revealed a high frequency of backed bladelets made from both local and non-local cherts.
-This observation further emphasizes the use of diverse raw materials, regardless of their quality.
-
-Therefore, it is not necessarily the case that non-local cherts were procured for their quality or because local cherts were unsuitable.
-Instead, their presence may reflect mobility strategies, territorial exploitation, and social networks that facilitated the introduction of long-distance cherts into the technological practices of southwestern hunter-gatherer groups.
+In this case, the impact of raw material quality seems to be more related to the differences between lithological groups, which have varying textures and breakage patterns (e.g., between greywacke and chert), rather than within micro-cryptocrystalline or fibrous quartz resources like chert.
+
+Finally, a visual inspection of a cache found in the Shelter area attributed to the Gravettian revealed a high frequency of backed bladelets made from both local and non-local cherts (Bicho et al., 2016).
+This observation further emphasizes the use of diverse chert types, regardless of their quality.
+
+Thus, non-local cherts were likely not procured solely for their superior quality or due to local resource limitations.
+Instead, their presence reflects mobility strategies, territorial exploitation, and social networks that facilitated the circulation of long-distance raw materials within southwestern Iberian hunter-gatherer groups.
 These non-local materials were used and maintained differently, possibly due to their initial size or the need for conservation, given their more restricted or higher-cost acquisition compared to local cherts.
@@
 These mobility strategies, territorial exploitation, and social networks also help explain the distinct patterns of non-local raw material use observed between the Gravettian and other assemblages.
-One key difference is the high frequency of non-local cherts in the Gravettian assemblage at Vale Boi, which make up approximately 50% of the chert assemblage (Belmiro et al., 2025).
+A key distinction is the elevated proportion of non-local cherts in the Gravettian assemblage at Vale Boi, comprising approximately 50% of the total chert assemblage (Belmiro et al., 2025).
 When combined with technological data, this also reveals significant differences in the frequency of retouched non-local chert tools and the tool-to-blank ratio (\~30% and \~1, respectively), compared to other technocomplexes (\~15-19% and 0.1-0.5).
-Several studies have shown that short-term occupations often rely heavily on non-local toolkits and raw materials (Torrence, 1983; Surovell, 2009), and are characterized by higher reduction intensity and increased tool-to-debitage ratios (Grove et al., 2023).
-These differences support the interpretation of the Gravettian at Vale Boi as a short-term occupation (Belmiro et al., 2025).
+Several studies have shown that short-term occupations often rely heavily on non-local toolkits and raw materials (Torrence, 1983; Surovell, 2009) and are characterized by higher reduction intensity and increased tool-to-debitage ratios (Grove et al., 2023).
+These differences support the interpretation of the Gravettian at Vale Boi as a set of short-term occupations (Belmiro et al., 2025).
 
@@
 Local raw materials were also used to produce high frequencies of retouched tools, though this was done less intensively.
-Given the technological differences between the Gravettian Vicentine facies and other regions such as central Portugal and southern Spain (Marreiros et al., ++), we suggest that the approximately 250 km distance between Vale Boi and identified non-local sources reflects raw material exchanges between groups at the edges of their territories, rather than an indication of exploited territory.
-
-These exchanges and patterns of mobility persisted throughout the Upper Paleolithic, though with changes in the mobility and settlement patterns at Vale Boi.
+Given the technological differences between the Gravettian Vicentine facies and other regions such as central Portugal and southern Spain (Marreiros et al., 2015), we suggest that the approximately 250 km distance between Vale Boi and identified non-local sources reflects raw material exchanges between groups at the edges of their territories, rather than an indication of exploited territory.
+
+These exchanges and patterns of mobility persisted throughout the UP, though with changes in the mobility and settlement patterns at Vale Boi.
 Over time, Vale Boi was occupied as a long-term residential site.
-Proto-Solutrean and later Solutrean groups, while still having access to non-local raw materials, moved less frequently, which limited the influx of these materials.
-Although non-local raw materials were still intensively exploited and maintained at the site, hunter-gatherer groups increasingly relied on local raw materials.
-This shift was likely due to the proximity of local sources, which facilitated more frequent use, while non-local materials continued to be exploited and maintained with greater intensity.
+At the beginning and during the Last Glacial Maximum, Proto-Solutrean and later Solutrean groups, while still having access to non-local raw materials, moved less frequently, which limited the influx of these materials.
+
+These changes in mobility frequency may be related to the increasingly harsher climatic conditions, which impacted all of Europe, while allowing regions such as southern Iberia to maintain temperate conditions where flora and faunal resources may have remained much the same (Gómez & Lunt, 2007; González-Sampériz et al., 2010; Hewitt, 2000; Jennings et al., 2011).
+The climatic stability of this ecological niche may have provided the necessary conditions for longer-term occupations of Vale Boi (Cascalheira et al., 2017), while still maintaining social interactions and contacts with other groups from Iberia seen both through technological patterns (Belmiro et al., 2021; Cascalheira et al., 2013; Cascalheira, 2017) and the presence of non-local raw materials (Belmiro et al., 2025).
+
+These non-local raw materials were still intensively exploited and maintained at the site, although hunter-gatherer groups increasingly relied on local raw materials.
+This shift was likely due to the proximity of local sources, which facilitated more frequent use and discard, or even due to the introduction of new lithic treatment techniques.
+
+\
+Our findings confirm initial expectations regarding mobility patterns, raw material use, and technological strategies in southwestern Iberia during the UP.
+While core reduction differences between local and non-local cherts were anticipated (\@tbl-expectations, expectations a-b), these differences were largely shaped by transportation strategies used to bring non-local cherts to Vale Boi.
+As such, the occurrence of non-local chert cores at Vale Boi depended on whether these materials arrived as finished tools or blanks—likely reflecting an individual provisioning strategy (Kuhn, 2004)—or as trimmed cores and larger stone volumes, aligning with site provisioning models (Kelly, 1988; Roth & Dibble, 1998; Clark, 1987; Kuhn, 2004).
+
+Non-local chert cores appear in all studied assemblages, though in limited numbers (e.g., in the Proto-Solutrean).
+While chips—often proxies for on-site knapping—were not analysed, the presence of knapping by-products, such as shatter (except in the Solutrean assemblage; see **Section 2.2.
+Datasets and analysis**), and core preparation and maintenance products in non-local cherts suggests active core reduction and on-site knapping.
+
+Our results indicate that most non-local cherts were introduced to Vale Boi as cores or nodules, subsequently knapped on-site.
+However, an exception is seen in TL cherts (TL2-6, TL9-10, TL14; **SOM 1 Table S13**), which lack associated cores and knapping debris.
+This absence suggests that TL cherts were introduced to the site primarily as finished tools or blanks, aligning with long-distance individual provisioning models (Kuhn, 2004).
+
+The presence of cores and evidence of on-site knapping enabled a comprehensive comparison of knapping strategies—specifically, the degree of preparation and maintenance through maintenance products—as well as knapping intensity between local and non-local cherts.
+
+Our analysis identified core preparation and maintenance products in the studied assemblages, suggesting a possible effort to preserve raw material volumes.
+However, these products were found in low percentages across all assemblages and chert types (\<4% and n\<7).
+This indicates that, at Vale Boi, core preparation and maintenance products may not serve as reliable proxies for distinguishing between chert types.
+Instead, their presence may be more closely linked to specific knapping strategies and objectives, such as prioritizing the production of blades and bladelets.
+
+Core and blank size comparisons provide further insights.
+Non-local cherts underwent more intensive reduction than local cherts, as indicated by their higher blank-to-core ratios and the significantly smaller size and weight of non-local chert blanks.
+These results align with previous studies, which propose that the intensive reduction of cores—manifested in smaller core sizes, smaller debitage products, and a higher amount of debitage per core—may be linked to lower mobility.
+In this context, lower mobility results in reduced access to non-local raw materials, which are therefore more intensively reduced and eventually replaced by local raw materials (Kuhn, 2004; Roth & Dibble, 1998; Grove et al., 2023; Surovel, 2009).
+However, our results show that these differences are not limited to long-term occupations and low mobility.
+High blank-to-core ratios and smaller blank sizes in non-local cherts are also observed in the Gravettian occupation, previously interpreted as a short-term occupation with frequent moves (Cascalheira, 2017; Belmiro et al., 2025).
+Due to this, we may interpret that non-local raw materials were generally more intensively reduced than local cherts throughout all the UP, independent of the duration of the occupation.
+
+In contrast to blanks, cores exhibited small, though statistically non-significant, differences in size and weight between local and non-local cherts.
+These results may be explained by a combination of factors, including the availability of local lithic resources, transport, and knapping strategies focused on the intensive reduction and maintenance of non-local raw materials.
+Belmiro et al. (2023) highlight that chert nodules in the Algarve region vary between \~5-10 cm, with a maximum diameter of 20 cm.
+Although abundant and easily accessible in exposed cliffs, local chert nodules are generally small and prone to fracturing when extracted, and normally occur fractured in secondary deposits associated to primary outcrops.
+As a result, the small size of local chert cores may not necessarily indicate intensive reduction but could instead reflect the initial size of available raw material volumes.
+
+In comparison, cherts from central Portugal—the probable source of T6 cherts from Vale Boi—can be found in larger volumes (\>20 cm), although smaller nodules are also present.
+These findings suggest that the small size of non-local chert artifacts was not necessarily constrained by raw material availability but was likely an intentional choice.
+This could reflect preferences for transportable nodules, strategic reduction to maximize resource utility, or a combination of both.
+This is further supported by the higher blank-to-core ratio observed in non-local raw materials, particularly T6.
+It may also explain the more frequent occurrence of heavier and larger outliers in local chert cores, compared to the relatively rare occurrence of larger outliers in non-local types (T6-7).
+
+The intensity of reduction and maintenance discussed thus far can also be inferred from the frequency and diversity of retouched tools.
+In fact, retouched tools are often key elements used in the literature to distinguish between differential use of local and non-local raw materials (Andrefsky, 1994; Bamforth, 1986; Kelly, 1988; Kuhn, 2004; Perlès, 1992).  Our results align with expected outcomes, as non-local cherts exhibit higher frequencies of retouched tools and higher tool-to-blank ratios across all assemblages.
+This supports the idea that non-local raw materials were actively procured, and that greater effort was invested in tool production (Andrefsky, 1994; Perlès, 1992; Kelly, 1988) and the maintenance of existing artifacts through retouch (Kuhn, 2004).
+
+While the high intensity of retouch in non-local cherts aligns with expectations (\@tbl-expectations, expectations 1c and 2c), their unexpectedly high tool diversity contradicts common mobility models which expect mobility to be linked with limited tool variability and an increase of easily reconfigurable tools (Odell, 1981; Shott, 1986; Siegel, 1984).
+Non-local cherts at Vale Boi contain a diverse set of tool types, although less specialized tools, such as retouched blanks, remain predominant in non-local cherts across all assemblages.
+This suggests that these materials were not only transported as unspecialised tools but as volumes that supported on-site tool production and diversification.
+
+While the patterns identified so far have established a clear distinction in the use and maintenance of local and non-local resources, it is also important to discuss the possible reasons for the varying abundance and technological patterns of non-local raw materials throughout the UP at Vale Boi, as well as the factors behind the differences observed across the different technocomplexes.
+
+Previous studies emphasize raw material abundance and quality as key factors driving the procurement and use of non-local lithic resources over local ones, shaping their proportions, use, and maintenance.
+In this context, the scarcity or low quality of local resources would drive the procurement of non-local materials and require more effort in preparing transportable cores and producing tools (Andrefsky, 1994; Perlès, 1992; Kelly, 1988).
+However, quality may become a less significant factor when non-local raw materials are easily accessible (Akerman et al., 2002; Torrence, 1996; Brown, 1999; Oestmo, 2017).
+Even in the presence of high-quality local materials, effort was still invested in transporting non-local resources (Roth & Dibble, 1998).
+
+Regarding Vale Boi, quality does not appear to be a defining factor in raw material choice, as typologies typically associated with high-quality lithic resources are also present in other materials.
+For example, Solutrean bifacial tools and points were produced using Lower Jurassic and Triassic local chert, which may be considered lower quality than Rio Maior Cretaceous chert due to its coarser texture and less homogeneous composition.
+This suggests that the choice of raw material was not solely dependent on quality, especially when techniques such as heat treatment—the process of heating lithic raw material to improve knapping quality (Schmidt et al., 2016)—have been identified during the Solutrean in European UP contexts (Bachellerie & Schmidt, 2021; Bordes, 1969; Inizan et al., 1976; Tiffagom, 1998).
+
+Similarly, Vale Comprido points, a fossil indicator of the Proto-Solutrean, were primarily made from other lithic resources (e.g., dolerite), with only two identified points made of chalcedony and chert (Belmiro et al., 2021).
+This further supports the idea that raw material choice was not strictly based on quality, as these points were made from a variety of materials, including those less commonly associated with formal tools at the site.
+
+In this case, the impact of raw material quality seems to be more related to the differences between lithological groups, which have varying textures and breakage patterns (e.g., between greywacke and chert), rather than within micro-cryptocrystalline or fibrous quartz resources like chert.
+
+Finally, a visual inspection of a cache found in the Shelter area attributed to the Gravettian revealed a high frequency of backed bladelets made from both local and non-local cherts (Bicho et al., 2016).
+This observation further emphasizes the use of diverse chert types, regardless of their quality.
+
+Thus, non-local cherts were likely not procured solely for their superior quality or due to local resource limitations.
+Instead, their presence reflects mobility strategies, territorial exploitation, and social networks that facilitated the circulation of long-distance raw materials within southwestern Iberian hunter-gatherer groups.
+These non-local materials were used and maintained differently, possibly due to their initial size or the need for conservation, given their more restricted or higher-cost acquisition compared to local cherts.
+
+These mobility strategies, territorial exploitation, and social networks also help explain the distinct patterns of non-local raw material use observed between the Gravettian and other assemblages.
+A key distinction is the elevated proportion of non-local cherts in the Gravettian assemblage at Vale Boi, comprising approximately 50% of the total chert assemblage (Belmiro et al., 2025).
+When combined with technological data, this also reveals significant differences in the frequency of retouched non-local chert tools and the tool-to-blank ratio (\~30% and \~1, respectively), compared to other technocomplexes (\~15-19% and 0.1-0.5).
+Several studies have shown that short-term occupations often rely heavily on non-local toolkits and raw materials (Torrence, 1983; Surovell, 2009) and are characterized by higher reduction intensity and increased tool-to-debitage ratios (Grove et al., 2023).
+These differences support the interpretation of the Gravettian at Vale Boi as a set of short-term occupations (Belmiro et al., 2025).
+
+This allows us to further expand our knowledge about chert provisioning during the UP at Vale Boi and suggest an overall interpretation for how groups were using and managing their lithic resources through time.
+In southwesternmost Iberia, the first identified UP occupations are from the Gravettian.
+At Vale Boi, these occupations appear to correspond to short-term settlements with frequent mobility, where non-local raw materials were transported as cores or nodules and knapped on-site.
+Alternatively, non-local raw materials were brought in as blanks, which were consistently retouched, although individual toolkits made from limited non-local raw materials—primarily blanks or tools—were also present.
+Raw material volumes were intensively exploited at the site, yielding both specialized and non-specialized tools.
+Local raw materials were also used to produce high frequencies of retouched tools, though this was done less intensively.
+Given the technological differences between the Gravettian Vicentine facies and other regions such as central Portugal and southern Spain (Marreiros et al., 2015), we suggest that the approximately 250 km distance between Vale Boi and identified non-local sources reflects raw material exchanges between groups at the edges of their territories, rather than an indication of exploited territory.
+
+These exchanges and patterns of mobility persisted throughout the UP, though with changes in the mobility and settlement patterns at Vale Boi.
+Over time, Vale Boi was occupied as a long-term residential site.
+At the beginning and during the Last Glacial Maximum, Proto-Solutrean and later Solutrean groups, while still having access to non-local raw materials, moved less frequently, which limited the influx of these materials.
+
+These changes in mobility frequency may be related to the increasingly harsher climatic conditions, which impacted all of Europe, while allowing regions such as southern Iberia to maintain temperate conditions where flora and faunal resources may have remained much the same (Gómez & Lunt, 2007; González-Sampériz et al., 2010; Hewitt, 2000; Jennings et al., 2011).
+The climatic stability of this ecological niche may have provided the necessary conditions for longer-term occupations of Vale Boi (Cascalheira et al., 2017), while still maintaining social interactions and contacts with other groups from Iberia seen both through technological patterns (Belmiro et al., 2021; Cascalheira et al., 2013; Cascalheira, 2017) and the presence of non-local raw materials (Belmiro et al., 2025).
+
+These non-local raw materials were still intensively exploited and maintained at the site, although hunter-gatherer groups increasingly relied on local raw materials.
+This shift was likely due to the proximity of local sources, which facilitated more frequent use and discard, or even due to the introduction of new lithic treatment techniques.
 
 # Conclusion
+
+This study demonstrates that lithic raw material procurement and technological organization at Vale Boi were shaped by a combination of raw material accessibility, mobility strategies, and social interactions:
+
+-   Intensive reduction of non-local cherts, evidenced by higher blank-to-core ratios and increased retouch frequencies, suggests that these materials were strategically curated and maintained throughout their use-lives.
+
+-   The unexpectedly high tool diversity in non-local cherts challenges assumptions that non-local materials were only transported as finished tools.
+Instead, our results indicate that they functioned as versatile raw material sources, knapped on-site to meet diverse technological needs as they arose
+
+-   Differences between Gravettian, Proto-Solutrean, and Solutrean assemblages indicate shifts in raw material use over time, with the Gravettian phase characterized by higher reliance on non-local cherts, suggesting more frequent movement or trade connections.
+In later phases, increased dependence on local resources reflects a shift toward more stable residential occupations.
+
+These findings provide new insights into hunter-gatherer mobility, social networks, and tool production strategies in southwestern Iberia.
+While non-local raw materials were crucial throughout the UP, their role evolved in response to changing settlement patterns and technological needs.
+
+Future work should explore whether Vale Boi’s non-local cherts were obtained through direct procurement, down-the-line exchange, or social trade networks.
+Additionally, applying use-wear analysis to retouched tools could help clarify whether non-local chert tools had specialized functions, further informing our understanding of raw material management and mobility strategies.
 

### 2025-04-13 analysis/paper/paper.qmd
References updated on main doc

@@
 The proportions of local vs non-local raw materials discussed thus far have focused greatly on group mobility, which is linked with residential mobility and occupation duration.
-Residential mobility (e.g., foraging range size and frequency of residential moves) is a defining characteristic of hunter-gatherer societies and plays a central role in their lifeways  (Kelly, 1995).
-It is a key factor driving changes in lithic assemblages (Brown, 1999; Oestmo, 2017; Roth & Dibble, 1998) as it significantly impacts the cost and use of artefacts and raw materials since they affect the relation between the distance of raw material sources to sites (Kuhn, 2004).
+Residential mobility (e.g., foraging range size and frequency of residential moves) is a defining characteristic of hunter-gatherer societies and plays a central role in their lifeways [@kelly1995].
+It is a key factor driving changes in lithic assemblages [@brown1999; @oestmo2017; @roth1998] as it significantly impacts the cost and use of artefacts and raw materials since they affect the relation between the distance of raw material sources to sites [@kuhn2004].
 As such, different raw material and technological patterns have been observed between short-term and long-term occupations.
 
-Short-term occupations typically rely on transported (non-local) toolkits (Torrence, 1983) and raw materials (Surovell, 2009), show higher reduction intensity, and an increased tools-to-debitage ratio in the assemblages (Grove et al., 2023).
-Long-term occupations typically rely on provisioning sites with locally available and suitable raw materials, while non-local materials are gradually used and discarded (Grove et al., 2023; Surovell, 2009).
-Non-local raw material cores are highly reduced and extensively exploited (Kuhn, 2004; Roth & Dibble, 1998) and exhibit higher blank-to-core ratios (Roth & Dibble, 1998).
+Short-term occupations typically rely on transported (non-local) toolkits [@torrence1983] and raw materials [@surovell2009], show higher reduction intensity, and an increased tools-to-debitage ratio in the assemblages [@grove2023].
+Long-term occupations typically rely on provisioning sites with locally available and suitable raw materials, while non-local materials are gradually used and discarded [@grove2023; @surovell2009].
+Non-local raw material cores are highly reduced and extensively exploited [@kuhn2004; @roth1998] and exhibit higher blank-to-core ratios [@roth1998].
 
 Altogether, different studies highlight significant variability in lithic assemblages and raw material proportions, reflecting different facets of hunter-gatherer technological organisation.
-This variability appears between behaviourally and ethnically similar sites (Bamforth, 1986), across distinct chrono-cultural periods, and even within a single site among various raw materials.
+This variability appears between behaviourally and ethnically similar sites [@bamforth1986], across distinct chrono-cultural periods, and even within a single site among various raw materials.
 To fully understand past hunter-gatherer behaviours, organisation, and lifeways, it is crucial to examine all stages of lithic production— procurement, distribution, use, maintenance, and discard—through interconnected approaches that combine lithic technology and raw material analyses.
@@
 The site is nestled among limestone outcrops that create west- and southwest-facing rock shelters (\@fig-map b).
-Covering more than 10,000 square meters along the valley slope, Vale Boi includes three main excavated areas, Slope, Terrace, and Shelter (\@fig-map c), which were excavated between 2000 and 2019 (Bicho et al., 2007, 2012, 2013; Cascalheira et al., 2017; Cascalheira & Bicho, 2013; Manne et al., 2012).
-Based on the available raw material and technological evidence, particularly the macroscopic and petrographic data (Belmiro et al., 2025), the Terrace and Shelter areas were selected for the current study.
+Covering more than 10,000 square meters along the valley slope, Vale Boi includes three main excavated areas, Slope, Terrace, and Shelter (\@fig-map c), which were excavated between 2000 and 2019 [@bicho2007; @bicho2012; @bicho2013; @cascalheira2017; @cascalheira2013; @manne2012].
+Based on the available raw material and technological evidence, particularly the macroscopic and petrographic data [@belmiro2025], the Terrace and Shelter areas were selected for the current study.
 
@@
 Between 2012 and 2016, eight primary litho-stratigraphic units were identified.
-The UP sequence includes multiple occupations, spanning \~10,000 years: Gravettian (levels 8 to 6) dated between approximately c. 32 and 27 ka cal BP, Proto-Solutrean (levels 5 to 4E) between approximately c. 26 and 24 ka cal BP, and Solutrean (levels 4D, 4C, 4, and Lower 3) between approximately c. 24 and 20 ka cal BP (Belmiro et al., 2021; Cascalheira et al., 2017; Cascalheira & Bicho, 2013).
+The UP sequence includes multiple occupations, spanning \~10,000 years: Gravettian (levels 8 to 6) dated between approximately c.
+32 and 27 ka cal BP, Proto-Solutrean (levels 5 to 4E) between approximately c.
+26 and 24 ka cal BP, and Solutrean (levels 4D, 4C, 4, and Lower 3) between approximately c. 24 and 20 ka cal BP [@belmiro2021; @cascalheira2017; @cascalheira2013b].
 
@@
 The Magdalenian and Gravettian levels (Z and D, respectively) were partially eroded, while the Solutrean levels represent the most intensive occupation, with three well-preserved archaeological horizons (layers A to C), dated to approximately c. 24-22 ka cal BP.
-These occupation levels were found beneath blocks of limestone that likely collapsed from the rock shelter ceiling, probably shortly after the LGM (Cascalheira, 2010; Cascalheira et al., 2017).
+These occupation levels were found beneath blocks of limestone that likely collapsed from the rock shelter ceiling, probably shortly after the LGM [@cascalheira2010; @cascalheira2017].
 
@@
 Given the chosen sample for this study (see **Section 2.2.
-Datasets and analysis**), we summarise here the results for the technological analysis of the Solutrean assemblages (levels A-C) of the Shelter (Cascalheira, 2010; Cascalheira et al., 2013) and the Proto-Solutrean assemblages (levels 4E-5) of the Terrace (Belmiro, 2020; Belmiro et al., 2021).
+Datasets and analysis**), we summarise here the results for the technological analysis of the Solutrean assemblages (levels A-C) of the Shelter [@cascalheira2010; @cascalheira2013a] and the Proto-Solutrean assemblages (levels 4E-5) of the Terrace [@belmiro2020; @belmiro2021].
 The Gravettian occupations (levels 6-8) of the Terrace have been recently studied, but the results remain unpublished.
@@
 Flakes dominate the debitage assemblage, and endscrapers, notches, splintered pieces and retouched blanks are the most common retouched tools.
-Stemmed and winged points, laurel leaves, and shouldered points were also identified (Cascalheira et al., 2013).
+Stemmed and winged points, laurel leaves, and shouldered points were also identified [@cascalheira2013a].
 
@@
 Both horizons are characterised by knapping strategies dominated by unidirectional prismatic cores, with little platform preparation, focused on the obtention of bladelets (in bottom level 5) and flakes.
-Retouched tools are dominated by splintered pieces, endscrapers and retouched blanks, with Vale Comprido points (the Proto-Solutrean fossil director) being barely inexistent in chert and mostly produced in dolerite (Belmiro et al., 2021).
+Retouched tools are dominated by splintered pieces, endscrapers and retouched blanks, with Vale Comprido points (the Proto-Solutrean fossil director) being barely inexistent in chert and mostly produced in dolerite [@belmiro2021].
 
 Previous studies showed that the main raw materials used at the site are chert, quartz, greywacke and smaller amounts of chalcedony, with the percentages of these changing across occupations.
-Notably, the lower Proto-Solutrean levels show high frequencies of quartz (Belmiro et al., 2021), whereas chert becomes the dominant raw material in the Solutrean occupations of the Shelter (Cascalheira, 2010).
-Likewise, different raw materials appear to have been used for specific purposes,with chert being particularly associated  with the production of high percentage of blanks and retouched tools, and greywacke being extensively used for anvils (Horta et al., 2019)
+Notably, the lower Proto-Solutrean levels show high frequencies of quartz [@belmiro2021], whereas chert becomes the dominant raw material in the Solutrean occupations of the Shelter [@cascalheira2010].
+Likewise, different raw materials appear to have been used for specific purposes,with chert being particularly associated  with the production of high percentage of blanks and retouched tools, and greywacke being extensively used for anvils [@horta2019].
 
 Quartz and greywacke are locally available from streambeds near the site, whereas regional chert and chalcedony sources lie further afield, within a \~20 km radius.
-Our previous studies showed, as expected, the existence of local cherts in the assemblages (\@fig-cherts a), originating from the Lower Jurassic (cherts) and Triassic (chalcedony) outcrops in proximity to the site (Belmiro et al., 2025).
-Despite petrographic differences between cherts and chalcedony, for this study any chert varieties (e.g., chalcedony or jasper) are referred to as chert (Luedtke, 1978).
+Our previous studies showed, as expected, the existence of local cherts in the assemblages (\@fig-cherts a), originating from the Lower Jurassic (cherts) and Triassic (chalcedony) outcrops in proximity to the site [@belmiro2025].
+Despite petrographic differences between cherts and chalcedony, for this study any chert varieties (e.g., chalcedony or jasper) are referred to as chert [@luedtke1978].
 A significant  portion of non-local cherts was also identified (\@fig-cherts b-c), with some identified sources, such as the Cretaceous outcrops of central Portugal (Lisbon and Rio Maior areas) and the Upper Jurassic oolitic/peloidal cherts of the Betic Systems in southern Spain.
@@
 Previous research suggests that both the Terrace and Shelter functioned as seasonal residential camps, repeatedly occupied for extended periods.
-This interpretation is based on high-density lithic assemblages, evidence of tool manufacture, heat-cracked rocks associated with grease rendering, diverse faunal remains (marine and terrestrial), and personal ornaments and portable art (Manne et al., 2012).
+This interpretation is based on high-density lithic assemblages, evidence of tool manufacture, heat-cracked rocks associated with grease rendering, diverse faunal remains (marine and terrestrial), and personal ornaments and portable art [@manne2012].
 This is true for most of the UP occupations, although with some exceptions.
@@
 Cascalheira et al.
-(2017) applied the Whole Assemblage Behavioral Index (WABI) (Riel-Salvatore, 2010; Riel-Salvatore & Barton, 2004) to reconstruct and compare mobility strategies at Vale Boi.
-By analysing lithic volumetric density and retouched tool frequencies from the Shelter’s Solutrean (Cascalheira, 2010; Cascalheira et al., 2013) and Magdalenian (Mendonça, 2009) levels, as well as the Terrace’s Gravettian levels (Marreiros et al., 2015), the authors confirmed that most occupations corresponded to long-term residential base camps, composed of high-density assemblages with low degree of retouch, except for the Early Gravettian and Magdalenian occupations.
+(2017) applied the Whole Assemblage Behavioral Index (WABI) [@riel-salvatore2004; @riel-salvatore2010] to reconstruct and compare mobility strategies at Vale Boi.
+By analysing lithic volumetric density and retouched tool frequencies from the Shelter’s Solutrean [@cascalheira2010; @cascalheira2013a] and Magdalenian [@mendonça2009] levels, as well as the Terrace’s Gravettian levels [@marreiros2015], the authors confirmed that most occupations corresponded to long-term residential base camps, composed of high-density assemblages with low degree of retouch, except for the Early Gravettian and Magdalenian occupations.
 Regarding the Terrace area, however, the results were based on previous excavations (\<2012), and the identification of new layers and reorganization of the occupations is yet to be tested through the use of lithic techno-typological methods.
 
-A similar conclusion was suggested based on the interpretation of the chert sourcing and proportions in the several UP occupations of the Terrace and Shelter of Vale Boi and following other models relating local/non-local raw materials with mobility (Grove et al., 2023; Surovell, 2009).
-The high proportions of non-local chert (\~50%) in the Gravettian occupations (levels 6 and 7), and gradual reduction in the following occupations, were interpreted as a possible result of settlement changes, where the Gravettian occupations corresponded to short-term occupations, and the other occupations corresponded to residential, long-term occupations with more infrequent moves (Belmiro et al., 2025).\
+A similar conclusion was suggested based on the interpretation of the chert sourcing and proportions in the several UP occupations of the Terrace and Shelter of Vale Boi and following other models relating local/non-local raw materials with mobility [@grove2023; @surovell2009].
+The high proportions of non-local chert (\~50%) in the Gravettian occupations (levels 6 and 7), and gradual reduction in the following occupations, were interpreted as a possible result of settlement changes, where the Gravettian occupations corresponded to short-term occupations, and the other occupations corresponded to residential, long-term occupations with more infrequent moves [@belmiro2025].\
 
@@
 
-The macroscopic and petrographic data of cherts was previously collected and published by our team (Belmiro et al., 2025) and can also be found in an online repository (<https://www.doi.org/10.17605/OSF.IO/DBFT2>).
+The macroscopic and petrographic data of cherts was previously collected and published by our team [@belmiro2025] and can also be found in an online repository (<https://www.doi.org/10.17605/OSF.IO/DBFT2>).
 This study focused on the UP chert assemblages of Vale Boi, especially the Gravettian (levels 6 and 7), Proto-Solutrean (levels 4E and 5) and Solutrean (levels 4 to 4D) from the Terrace area and the Solutrean (levels A to C) from the Shelter area.
@@
 
-The Solutrean assemblage technological analysis was published by Cascalheira (2009) as part of a Master’s thesis and following publications (e.g., Cascalheira, 2010; Cascalheira et al., 2013), with the study of levels A to C of the Shelter area.
+The Solutrean assemblage technological analysis was published by Cascalheira [-@cascalheira2009] as part of a Master’s thesis and following publications [e.g., @cascalheira2010; @cascalheira2013a], with the study of levels A to C of the Shelter area.
 The dataset used for this study was not originally published but can be consulted in our online repository (<http://www.doi.org/10.17605/OSF.IO/UWMPK>).
@@
 The Solutrean analysis in this study is limited to the Shelter area due to the inexistence of technological studies for the better preserved Solutrean levels of the Terrace (e.g., level 4).
-However, and as seen in Belmiro et al. (2025), the raw material patterns seem to show little differences between the Solutrean in both areas; coupled with the similar radiocarbon dates between the areas (c. 24-22 and 24-20 ka cal BP), we assume that they are representative of similar, long-term, residential occupations which aptly characterise the Solutrean groups of Vale Boi.
-
-The Proto-Solutrean technological analysis was the result of a Master’s thesis (Belmiro, 2020) and following publication (Belmiro et al., 2021), with the study of levels 4E and 5 of the Terrace area.
+However, and as seen in [@belmiro2025], the raw material patterns seem to show little differences between the Solutrean in both areas; coupled with the similar radiocarbon dates between the areas (c. 24-22 and 24-20 ka cal BP), we assume that they are representative of similar, long-term, residential occupations which aptly characterise the Solutrean groups of Vale Boi.
+
+The Proto-Solutrean technological analysis was the result of a Master’s thesis [@belmiro2020] and following publication [@belmiro2021], with the study of levels 4E and 5 of the Terrace area.
 The dataset and analysis can be consulted in the original online repository (<http://www.doi.org/10.17605/OSF.IO/456EG>).
@@
 The lithic analysis datasets used in this study adopted an attribute-based methodology, which aims to describe morphological and metrical attributes of technological classes.
-The attributes analysed followed those present in specialized literature, such as Brézillon (1968), and Tixier and Inizan (1980), paired with other Palaeolithic lithic attribute analyses (see e.g., Bicho, 1992; Zilhão, 1997; Cascalheira, 2019; Almeida, 2000; Scerri et al., 2014; Tostevin, 2013).
+The attributes analysed followed those present in specialized literature, such as Brézillon [-@brézillon1968], and [@tixier1980], paired with other Palaeolithic lithic attribute analyses [e.g., @bicho1992; @zilhão1997; @cascalheira2019; @tostevin2012].
 
 After data collection, the databases were imported into R Programming Environment, where the information was processed through the creation of descriptive statistical analysis and writing of this article.
-Some differences in the collection process, especially between the study by Cascalheira (2010) and the following studies, required homogenization.
-For example, where in Cascalheira (2010) retouch was an independent variable (e.g., Class as flake, Retouch as Yes), in Belmiro et al. (2021), retouch was identified as Class variable.
+Some differences in the collection process, especially between the study by Cascalheira [-@cascalheira2010] and the following studies, required homogenization.
+For example, where in Cascalheira (2010) retouch was an independent variable (e.g., Class as flake, Retouch as Yes), in Belmiro et al.
+[-@belmiro2021], retouch was identified as Class variable.
 In this situation, the Solutrean dataset was modified, so that all artifacts with retouch were identified in the Class variable, instead of the original classification.
@@
 We used the rrtools package by Marwick et al.
-(2018) to create a research compendium and write a reproducible journal article.
-The provided files include the complete set of raw data used in the analysis, along with a custom R project (Wickham, 2015) containing the code required to generate all tables and figures.
-To enable maximum reuse, the code is made available under the MIT license, data under CC‐0, and figures under CC‐BY (additional details can be found in Marwick, 2017).
+[-@marwick2018] to create a research compendium and write a reproducible journal article.
+The provided files include the complete set of raw data used in the analysis, along with a custom R project [@wickham2015] containing the code required to generate all tables and figures.
+To enable maximum reuse, the code is made available under the MIT license, data under CC‐0, and figures under CC‐BY (additional details can be found in @Marwick2017).
 
@@
 
-The sizes of blanks and cores were explored through scatterplots of width and length (\@fig-gravettian-plots**, @fig-proto-plots**, \@fig-solutrean-plots a) and boxplots for weight (\@fig-gravettian-plots, **@fig-proto-plots**, \@fig-solutrean-plots b) and through non-parametric statistical tests; complete tables with descriptive statistics and non-parametric statistical test results can be found in **SOM 1 Tables s9-11** and **s12**, respectively.
+The sizes of blanks and cores were explored through scatterplots of width and length (\@fig-gravettian-plots**, @fig-proto-plots**, \@fig-solutrean-plots a) and boxplots for weight (\@fig-gravettian-plots, @fig-proto-plots, \@fig-solutrean-plots b) and through non-parametric statistical tests; complete tables with descriptive statistics and non-parametric statistical test results can be found in **SOM 1 Tables s9-11** and **s12**, respectively.
 
@@
 While core reduction differences between local and non-local cherts were anticipated (\@tbl-expectations, expectations a-b), these differences were largely shaped by transportation strategies used to bring non-local cherts to Vale Boi.
-As such, the occurrence of non-local chert cores at Vale Boi depended on whether these materials arrived as finished tools or blanks—likely reflecting an individual provisioning strategy (Kuhn, 2004)—or as trimmed cores and larger stone volumes, aligning with site provisioning models (Kelly, 1988; Roth & Dibble, 1998; Clark, 1987; Kuhn, 2004).
+As such, the occurrence of non-local chert cores at Vale Boi depended on whether these materials arrived as finished tools or blanks—likely reflecting an individual provisioning strategy [@kuhn2004]—or as trimmed cores and larger stone volumes, aligning with site provisioning models [@kelly1988; @roth1998; @clark1987; @kuhn2004] .
 
@@
 However, an exception is seen in TL cherts (TL2-6, TL9-10, TL14; **SOM 1 Table S13**), which lack associated cores and knapping debris.
-This absence suggests that TL cherts were introduced to the site primarily as finished tools or blanks, aligning with long-distance individual provisioning models (Kuhn, 2004).
+This absence suggests that TL cherts were introduced to the site primarily as finished tools or blanks, aligning with long-distance individual provisioning models [@kuhn2004].
 
@@
 These results align with previous studies, which propose that the intensive reduction of cores—manifested in smaller core sizes, smaller debitage products, and a higher amount of debitage per core—may be linked to lower mobility.
-In this context, lower mobility results in reduced access to non-local raw materials, which are therefore more intensively reduced and eventually replaced by local raw materials (Kuhn, 2004; Roth & Dibble, 1998; Grove et al., 2023; Surovel, 2009).
+In this context, lower mobility results in reduced access to non-local raw materials, which are therefore more intensively reduced and eventually replaced by local raw materials [@kuhn2004; @roth1998; @grove2023; @surovell2009].
 However, our results show that these differences are not limited to long-term occupations and low mobility.
-High blank-to-core ratios and smaller blank sizes in non-local cherts are also observed in the Gravettian occupation, previously interpreted as a short-term occupation with frequent moves (Cascalheira, 2017; Belmiro et al., 2025).
+High blank-to-core ratios and smaller blank sizes in non-local cherts are also observed in the Gravettian occupation, previously interpreted as a short-term occupation with frequent moves [@cascalheira2017a; @belmiro2025].
 Due to this, we may interpret that non-local raw materials were generally more intensively reduced than local cherts throughout all the UP, independent of the duration of the occupation.
@@
 These results may be explained by a combination of factors, including the availability of local lithic resources, transport, and knapping strategies focused on the intensive reduction and maintenance of non-local raw materials.
-Belmiro et al. (2023) highlight that chert nodules in the Algarve region vary between \~5-10 cm, with a maximum diameter of 20 cm.
+Belmiro et al.
+[-@belmiro2023] highlight that chert nodules in the Algarve region vary between \~5-10 cm, with a maximum diameter of 20 cm.
 Although abundant and easily accessible in exposed cliffs, local chert nodules are generally small and prone to fracturing when extracted, and normally occur fractured in secondary deposits associated to primary outcrops.
@@
 The intensity of reduction and maintenance discussed thus far can also be inferred from the frequency and diversity of retouched tools.
-In fact, retouched tools are often key elements used in the literature to distinguish between differential use of local and non-local raw materials (Andrefsky, 1994; Bamforth, 1986; Kelly, 1988; Kuhn, 2004; Perlès, 1992).  Our results align with expected outcomes, as non-local cherts exhibit higher frequencies of retouched tools and higher tool-to-blank ratios across all assemblages.
-This supports the idea that non-local raw materials were actively procured, and that greater effort was invested in tool production (Andrefsky, 1994; Perlès, 1992; Kelly, 1988) and the maintenance of existing artifacts through retouch (Kuhn, 2004).
-
-While the high intensity of retouch in non-local cherts aligns with expectations (\@tbl-expectations, expectations 1c and 2c), their unexpectedly high tool diversity contradicts common mobility models which expect mobility to be linked with limited tool variability and an increase of easily reconfigurable tools (Odell, 1981; Shott, 1986; Siegel, 1984).
+In fact, retouched tools are often key elements used in the literature to distinguish between differential use of local and non-local raw materials [@andrefsky1994; @bamforth1986; @kelly1988; @kuhn2004; @perlès1992].
+Our results align with expected outcomes, as non-local cherts exhibit higher frequencies of retouched tools and higher tool-to-blank ratios across all assemblages.
+This supports the idea that non-local raw materials were actively procured, and that greater effort was invested in tool production [@andrefsky1994; @perlès1992; @kelly1988] and the maintenance of existing artifacts through retouch [@kuhn2004].
+
+While the high intensity of retouch in non-local cherts aligns with expectations (\@tbl-expectations, expectations 1c and 2c), their unexpectedly high tool diversity contradicts common mobility models which expect mobility to be linked with limited tool variability and an increase of easily reconfigurable tools [@odell1981; @shott1986; @siegel1984].
 Non-local cherts at Vale Boi contain a diverse set of tool types, although less specialized tools, such as retouched blanks, remain predominant in non-local cherts across all assemblages.
@@
 Previous studies emphasize raw material abundance and quality as key factors driving the procurement and use of non-local lithic resources over local ones, shaping their proportions, use, and maintenance.
-In this context, the scarcity or low quality of local resources would drive the procurement of non-local materials and require more effort in preparing transportable cores and producing tools (Andrefsky, 1994; Perlès, 1992; Kelly, 1988).
-However, quality may become a less significant factor when non-local raw materials are easily accessible (Akerman et al., 2002; Torrence, 1996; Brown, 1999; Oestmo, 2017).
-Even in the presence of high-quality local materials, effort was still invested in transporting non-local resources (Roth & Dibble, 1998).
+In this context, the scarcity or low quality of local resources would drive the procurement of non-local materials and require more effort in preparing transportable cores and producing tools [@andrefsky1994; @perlès1992; @kelly1988].
+However, quality may become a less significant factor when non-local raw materials are easily accessible [@akerman2002; @torrence1986; @brown1999; @oestmo2017].
+Even in the presence of high-quality local materials, effort was still invested in transporting non-local resources [@roth1998].
 
@@
 For example, Solutrean bifacial tools and points were produced using Lower Jurassic and Triassic local chert, which may be considered lower quality than Rio Maior Cretaceous chert due to its coarser texture and less homogeneous composition.
-This suggests that the choice of raw material was not solely dependent on quality, especially when techniques such as heat treatment—the process of heating lithic raw material to improve knapping quality (Schmidt et al., 2016)—have been identified during the Solutrean in European UP contexts (Bachellerie & Schmidt, 2021; Bordes, 1969; Inizan et al., 1976; Tiffagom, 1998).
-
-Similarly, Vale Comprido points, a fossil indicator of the Proto-Solutrean, were primarily made from other lithic resources (e.g., dolerite), with only two identified points made of chalcedony and chert (Belmiro et al., 2021).
+This suggests that the choice of raw material was not solely dependent on quality, especially when techniques such as heat treatment—the process of heating lithic raw material to improve knapping quality [@schmidt2016]—have been identified during the Solutrean in European UP contexts [@bachellerie2022; @bordes1969; @inizan1976; @tiffagom1998].
+
+Similarly, Vale Comprido points, a fossil indicator of the Proto-Solutrean, were primarily made from other lithic resources (e.g., dolerite), with only two identified points made of chalcedony and chert [@belmiro2021].
 This further supports the idea that raw material choice was not strictly based on quality, as these points were made from a variety of materials, including those less commonly associated with formal tools at the site.
@@
 
-Finally, a visual inspection of a cache found in the Shelter area attributed to the Gravettian revealed a high frequency of backed bladelets made from both local and non-local cherts (Bicho et al., 2016).
+Finally, a visual inspection of a cache found in the Shelter area attributed to the Gravettian revealed a high frequency of backed bladelets made from both local and non-local cherts [@bicho2016].
 This observation further emphasizes the use of diverse chert types, regardless of their quality.
@@
 These mobility strategies, territorial exploitation, and social networks also help explain the distinct patterns of non-local raw material use observed between the Gravettian and other assemblages.
-A key distinction is the elevated proportion of non-local cherts in the Gravettian assemblage at Vale Boi, comprising approximately 50% of the total chert assemblage (Belmiro et al., 2025).
+A key distinction is the elevated proportion of non-local cherts in the Gravettian assemblage at Vale Boi, comprising approximately 50% of the total chert assemblage [@belmiro2025].
 When combined with technological data, this also reveals significant differences in the frequency of retouched non-local chert tools and the tool-to-blank ratio (\~30% and \~1, respectively), compared to other technocomplexes (\~15-19% and 0.1-0.5).
-Several studies have shown that short-term occupations often rely heavily on non-local toolkits and raw materials (Torrence, 1983; Surovell, 2009) and are characterized by higher reduction intensity and increased tool-to-debitage ratios (Grove et al., 2023).
-These differences support the interpretation of the Gravettian at Vale Boi as a set of short-term occupations (Belmiro et al., 2025).
+Several studies have shown that short-term occupations often rely heavily on non-local toolkits and raw materials [@torrence1983; @surovell2009] and are characterized by higher reduction intensity and increased tool-to-debitage ratios [@grove2023].
+These differences support the interpretation of the Gravettian at Vale Boi as a set of short-term occupations [@belmiro2025].
 
@@
 Local raw materials were also used to produce high frequencies of retouched tools, though this was done less intensively.
-Given the technological differences between the Gravettian Vicentine facies and other regions such as central Portugal and southern Spain (Marreiros et al., 2015), we suggest that the approximately 250 km distance between Vale Boi and identified non-local sources reflects raw material exchanges between groups at the edges of their territories, rather than an indication of exploited territory.
+Given the technological differences between the Gravettian Vicentine facies and other regions such as central Portugal and southern Spain [@marreiros2015], we suggest that the approximately 250 km distance between Vale Boi and identified non-local sources reflects raw material exchanges between groups at the edges of their territories, rather than an indication of exploited territory.
 
@@
 
-These changes in mobility frequency may be related to the increasingly harsher climatic conditions, which impacted all of Europe, while allowing regions such as southern Iberia to maintain temperate conditions where flora and faunal resources may have remained much the same (Gómez & Lunt, 2007; González-Sampériz et al., 2010; Hewitt, 2000; Jennings et al., 2011).
-The climatic stability of this ecological niche may have provided the necessary conditions for longer-term occupations of Vale Boi (Cascalheira et al., 2017), while still maintaining social interactions and contacts with other groups from Iberia seen both through technological patterns (Belmiro et al., 2021; Cascalheira et al., 2013; Cascalheira, 2017) and the presence of non-local raw materials (Belmiro et al., 2025).
+These changes in mobility frequency may be related to the increasingly harsher climatic conditions, which impacted all of Europe, while allowing regions such as southern Iberia to maintain temperate conditions where flora and faunal resources may have remained much the same [@gómez2007; @gonzález-sampériz2010; @hewitt2000; @jennings2011].
+The climatic stability of this ecological niche may have provided the necessary conditions for longer-term occupations of Vale Boi [@cascalheira2017], while still maintaining social interactions and contacts with other groups from Iberia seen both through technological patterns [@belmiro2021; @cascalheira2013c; @cascalheira2017] and the presence of non-local raw materials [@belmiro2025].
 
@@
 
-\
-Our findings confirm initial expectations regarding mobility patterns, raw material use, and technological strategies in southwestern Iberia during the UP.
-While core reduction differences between local and non-local cherts were anticipated (\@tbl-expectations, expectations a-b), these differences were largely shaped by transportation strategies used to bring non-local cherts to Vale Boi.
-As such, the occurrence of non-local chert cores at Vale Boi depended on whether these materials arrived as finished tools or blanks—likely reflecting an individual provisioning strategy (Kuhn, 2004)—or as trimmed cores and larger stone volumes, aligning with site provisioning models (Kelly, 1988; Roth & Dibble, 1998; Clark, 1987; Kuhn, 2004).
-
-Non-local chert cores appear in all studied assemblages, though in limited numbers (e.g., in the Proto-Solutrean).
-While chips—often proxies for on-site knapping—were not analysed, the presence of knapping by-products, such as shatter (except in the Solutrean assemblage; see **Section 2.2.
-Datasets and analysis**), and core preparation and maintenance products in non-local cherts suggests active core reduction and on-site knapping.
-
-Our results indicate that most non-local cherts were introduced to Vale Boi as cores or nodules, subsequently knapped on-site.
-However, an exception is seen in TL cherts (TL2-6, TL9-10, TL14; **SOM 1 Table S13**), which lack associated cores and knapping debris.
-This absence suggests that TL cherts were introduced to the site primarily as finished tools or blanks, aligning with long-distance individual provisioning models (Kuhn, 2004).
-
-The presence of cores and evidence of on-site knapping enabled a comprehensive comparison of knapping strategies—specifically, the degree of preparation and maintenance through maintenance products—as well as knapping intensity between local and non-local cherts.
-
-Our analysis identified core preparation and maintenance products in the studied assemblages, suggesting a possible effort to preserve raw material volumes.
-However, these products were found in low percentages across all assemblages and chert types (\<4% and n\<7).
-This indicates that, at Vale Boi, core preparation and maintenance products may not serve as reliable proxies for distinguishing between chert types.
-Instead, their presence may be more closely linked to specific knapping strategies and objectives, such as prioritizing the production of blades and bladelets.
-
-Core and blank size comparisons provide further insights.
-Non-local cherts underwent more intensive reduction than local cherts, as indicated by their higher blank-to-core ratios and the significantly smaller size and weight of non-local chert blanks.
-These results align with previous studies, which propose that the intensive reduction of cores—manifested in smaller core sizes, smaller debitage products, and a higher amount of debitage per core—may be linked to lower mobility.
-In this context, lower mobility results in reduced access to non-local raw materials, which are therefore more intensively reduced and eventually replaced by local raw materials (Kuhn, 2004; Roth & Dibble, 1998; Grove et al., 2023; Surovel, 2009).
-However, our results show that these differences are not limited to long-term occupations and low mobility.
-High blank-to-core ratios and smaller blank sizes in non-local cherts are also observed in the Gravettian occupation, previously interpreted as a short-term occupation with frequent moves (Cascalheira, 2017; Belmiro et al., 2025).
-Due to this, we may interpret that non-local raw materials were generally more intensively reduced than local cherts throughout all the UP, independent of the duration of the occupation.
-
-In contrast to blanks, cores exhibited small, though statistically non-significant, differences in size and weight between local and non-local cherts.
-These results may be explained by a combination of factors, including the availability of local lithic resources, transport, and knapping strategies focused on the intensive reduction and maintenance of non-local raw materials.
-Belmiro et al. (2023) highlight that chert nodules in the Algarve region vary between \~5-10 cm, with a maximum diameter of 20 cm.
-Although abundant and easily accessible in exposed cliffs, local chert nodules are generally small and prone to fracturing when extracted, and normally occur fractured in secondary deposits associated to primary outcrops.
-As a result, the small size of local chert cores may not necessarily indicate intensive reduction but could instead reflect the initial size of available raw material volumes.
-
-In comparison, cherts from central Portugal—the probable source of T6 cherts from Vale Boi—can be found in larger volumes (\>20 cm), although smaller nodules are also present.
-These findings suggest that the small size of non-local chert artifacts was not necessarily constrained by raw material availability but was likely an intentional choice.
-This could reflect preferences for transportable nodules, strategic reduction to maximize resource utility, or a combination of both.
-This is further supported by the higher blank-to-core ratio observed in non-local raw materials, particularly T6.
-It may also explain the more frequent occurrence of heavier and larger outliers in local chert cores, compared to the relatively rare occurrence of larger outliers in non-local types (T6-7).
-
-The intensity of reduction and maintenance discussed thus far can also be inferred from the frequency and diversity of retouched tools.
-In fact, retouched tools are often key elements used in the literature to distinguish between differential use of local and non-local raw materials (Andrefsky, 1994; Bamforth, 1986; Kelly, 1988; Kuhn, 2004; Perlès, 1992).  Our results align with expected outcomes, as non-local cherts exhibit higher frequencies of retouched tools and higher tool-to-blank ratios across all assemblages.
-This supports the idea that non-local raw materials were actively procured, and that greater effort was invested in tool production (Andrefsky, 1994; Perlès, 1992; Kelly, 1988) and the maintenance of existing artifacts through retouch (Kuhn, 2004).
-
-While the high intensity of retouch in non-local cherts aligns with expectations (\@tbl-expectations, expectations 1c and 2c), their unexpectedly high tool diversity contradicts common mobility models which expect mobility to be linked with limited tool variability and an increase of easily reconfigurable tools (Odell, 1981; Shott, 1986; Siegel, 1984).
-Non-local cherts at Vale Boi contain a diverse set of tool types, although less specialized tools, such as retouched blanks, remain predominant in non-local cherts across all assemblages.
-This suggests that these materials were not only transported as unspecialised tools but as volumes that supported on-site tool production and diversification.
-
-While the patterns identified so far have established a clear distinction in the use and maintenance of local and non-local resources, it is also important to discuss the possible reasons for the varying abundance and technological patterns of non-local raw materials throughout the UP at Vale Boi, as well as the factors behind the differences observed across the different technocomplexes.
-
-Previous studies emphasize raw material abundance and quality as key factors driving the procurement and use of non-local lithic resources over local ones, shaping their proportions, use, and maintenance.
-In this context, the scarcity or low quality of local resources would drive the procurement of non-local materials and require more effort in preparing transportable cores and producing tools (Andrefsky, 1994; Perlès, 1992; Kelly, 1988).
-However, quality may become a less significant factor when non-local raw materials are easily accessible (Akerman et al., 2002; Torrence, 1996; Brown, 1999; Oestmo, 2017).
-Even in the presence of high-quality local materials, effort was still invested in transporting non-local resources (Roth & Dibble, 1998).
-
-Regarding Vale Boi, quality does not appear to be a defining factor in raw material choice, as typologies typically associated with high-quality lithic resources are also present in other materials.
-For example, Solutrean bifacial tools and points were produced using Lower Jurassic and Triassic local chert, which may be considered lower quality than Rio Maior Cretaceous chert due to its coarser texture and less homogeneous composition.
-This suggests that the choice of raw material was not solely dependent on quality, especially when techniques such as heat treatment—the process of heating lithic raw material to improve knapping quality (Schmidt et al., 2016)—have been identified during the Solutrean in European UP contexts (Bachellerie & Schmidt, 2021; Bordes, 1969; Inizan et al., 1976; Tiffagom, 1998).
-
-Similarly, Vale Comprido points, a fossil indicator of the Proto-Solutrean, were primarily made from other lithic resources (e.g., dolerite), with only two identified points made of chalcedony and chert (Belmiro et al., 2021).
-This further supports the idea that raw material choice was not strictly based on quality, as these points were made from a variety of materials, including those less commonly associated with formal tools at the site.
-
-In this case, the impact of raw material quality seems to be more related to the differences between lithological groups, which have varying textures and breakage patterns (e.g., between greywacke and chert), rather than within micro-cryptocrystalline or fibrous quartz resources like chert.
-
-Finally, a visual inspection of a cache found in the Shelter area attributed to the Gravettian revealed a high frequency of backed bladelets made from both local and non-local cherts (Bicho et al., 2016).
-This observation further emphasizes the use of diverse chert types, regardless of their quality.
-
-Thus, non-local cherts were likely not procured solely for their superior quality or due to local resource limitations.
-Instead, their presence reflects mobility strategies, territorial exploitation, and social networks that facilitated the circulation of long-distance raw materials within southwestern Iberian hunter-gatherer groups.
-These non-local materials were used and maintained differently, possibly due to their initial size or the need for conservation, given their more restricted or higher-cost acquisition compared to local cherts.
-
-These mobility strategies, territorial exploitation, and social networks also help explain the distinct patterns of non-local raw material use observed between the Gravettian and other assemblages.
-A key distinction is the elevated proportion of non-local cherts in the Gravettian assemblage at Vale Boi, comprising approximately 50% of the total chert assemblage (Belmiro et al., 2025).
-When combined with technological data, this also reveals significant differences in the frequency of retouched non-local chert tools and the tool-to-blank ratio (\~30% and \~1, respectively), compared to other technocomplexes (\~15-19% and 0.1-0.5).
-Several studies have shown that short-term occupations often rely heavily on non-local toolkits and raw materials (Torrence, 1983; Surovell, 2009) and are characterized by higher reduction intensity and increased tool-to-debitage ratios (Grove et al., 2023).
-These differences support the interpretation of the Gravettian at Vale Boi as a set of short-term occupations (Belmiro et al., 2025).
-
-This allows us to further expand our knowledge about chert provisioning during the UP at Vale Boi and suggest an overall interpretation for how groups were using and managing their lithic resources through time.
-In southwesternmost Iberia, the first identified UP occupations are from the Gravettian.
-At Vale Boi, these occupations appear to correspond to short-term settlements with frequent mobility, where non-local raw materials were transported as cores or nodules and knapped on-site.
-Alternatively, non-local raw materials were brought in as blanks, which were consistently retouched, although individual toolkits made from limited non-local raw materials—primarily blanks or tools—were also present.
-Raw material volumes were intensively exploited at the site, yielding both specialized and non-specialized tools.
-Local raw materials were also used to produce high frequencies of retouched tools, though this was done less intensively.
-Given the technological differences between the Gravettian Vicentine facies and other regions such as central Portugal and southern Spain (Marreiros et al., 2015), we suggest that the approximately 250 km distance between Vale Boi and identified non-local sources reflects raw material exchanges between groups at the edges of their territories, rather than an indication of exploited territory.
-
-These exchanges and patterns of mobility persisted throughout the UP, though with changes in the mobility and settlement patterns at Vale Boi.
-Over time, Vale Boi was occupied as a long-term residential site.
-At the beginning and during the Last Glacial Maximum, Proto-Solutrean and later Solutrean groups, while still having access to non-local raw materials, moved less frequently, which limited the influx of these materials.
-
-These changes in mobility frequency may be related to the increasingly harsher climatic conditions, which impacted all of Europe, while allowing regions such as southern Iberia to maintain temperate conditions where flora and faunal resources may have remained much the same (Gómez & Lunt, 2007; González-Sampériz et al., 2010; Hewitt, 2000; Jennings et al., 2011).
-The climatic stability of this ecological niche may have provided the necessary conditions for longer-term occupations of Vale Boi (Cascalheira et al., 2017), while still maintaining social interactions and contacts with other groups from Iberia seen both through technological patterns (Belmiro et al., 2021; Cascalheira et al., 2013; Cascalheira, 2017) and the presence of non-local raw materials (Belmiro et al., 2025).
-
-These non-local raw materials were still intensively exploited and maintained at the site, although hunter-gatherer groups increasingly relied on local raw materials.
-This shift was likely due to the proximity of local sources, which facilitated more frequent use and discard, or even due to the introduction of new lithic treatment techniques.
-
 # Conclusion
@@
 -   The unexpectedly high tool diversity in non-local cherts challenges assumptions that non-local materials were only transported as finished tools.
-Instead, our results indicate that they functioned as versatile raw material sources, knapped on-site to meet diverse technological needs as they arose
+Instead, our results indicate that they functioned as versatile raw material sources, knapped on-site to meet diverse technological needs as they arose.
 

### 2025-04-13 analysis/Supplementary/ESM_1.qmd
Submission preparation 01

--- 2025-03-01 analysis/Supplementary/supplementary_tables.qmd
@@
 ---
-title: "SOM result tables"
+title: "Online Resources 1: Supplementary results and tables"
+subtitle: "From Stone to Tool: How Raw Materials Influenced Upper Paleolithic Technology in Southwestern Iberia (Vale Boi). Archaeological and Anthroplogical Sciences"
 author:
-  - Jane Doe:
+  - Joana Belmiro:
       correspondence: "yes"
-      email: janedoe@fosg.org
+      email: jfbelmiro@ualg.pt
       orcid: 0000-0003-1689-0557
       institute:
-        - fosg
-        - fop
-  - John Q. Doe:
-      institute: fosg
+        - ICArEHB
+  - Nuno Bicho:
+      institute: ICArEHB
       orcid: 0000-0003-1689-0558
-  - Peder Ås:
-      institute: fosg
+  - Xavier Terradas:
+      institute: CSIC
       orcid: 0000-0003-1689-0559
-  - Juan Pérez:
+  - João Cascalheira:
       orcid: 0000-0003-1689-0551
+      institute: ICArEHB
 institute:
-        - name: Acme Corporation
-  - Max Mustermann:
-      orcid: 0000-0003-1689-0552
-institute:
-  - fosg:
-      name: Formatting Open Science Group
-      address: 23 Science Street, Eureka, Mississippi, USA
-  - fop: Federation of Planets
+  - ICArEHB:
+      name: Interdisciplinary Center for Archaeology and Human Evolution
+      address: 
+  - CSIC: CSIC
 title-block-published: "Last updated"  

### 2025-04-13 analysis/paper/paper.qmd
Submission preparation 01

@@
 abstract: |
-  The Upper Palaeolithic (UP) of Westernmost Europe was marked by technological and cultural transformations shaped by Late Pleistocene climatic shifts.
+  The Upper Palaeolithic of Westernmost Europe was marked by technological and cultural transformations shaped by Late Pleistocene climatic shifts.
 The Iberian Peninsula, particularly southwestern Iberia, served as a key eco-cultural niche, sustaining hunter-gatherer populations through environmental fluctuations.
 Vale Boi, the only site in southwestern Iberia with a nearly complete UP sequence, provides critical insights into technological, economical and mobility patterns over time.
-This study examines lithic raw material procurement and technological organization at Vale Boi, focusing on the role of non-local cherts in mobility strategies and social interactions.Using a combined techno-typological and geoarchaeological approach, we analyzed chert assemblages from Gravettian, Proto-Solutrean, and Solutrean levels, integrating previously published macroscopic and petrographic data.
+This study examines lithic raw material procurement and technological organization at Vale Boi, focusing on the role of non-local cherts in mobility strategies and social interactions.
+Using a combined techno-typological and geoarchaeological approach, we analyzed chert assemblages from Gravettian, Proto-Solutrean, and Solutrean levels, integrating previously published macroscopic and petrographic data.
 Our results indicate that non-local cherts were intensively reduced, suggesting strategic curation.
@@
 <!-- The actual document text starts here: -->
+
+# Acknowledgements
+
+**Author contributions:** Study conception and design were performed by Joana Belmiro and João Cascalheira.
+Material preparation, data collection, database collection and treatment and analysis were performed by Joana Belmiro.
+The first draft of the manuscript was written by Joana Belmiro and all authors commented on previous versions of the manuscript.
+All authors read and approved the final manuscript.
 
@@
 #| fig-cap: "a) Location of the archaeological site of Vale Boi (orange box) within the Iberian Peninsula; b) Photo of the limestone hill with terrace and shelter locations; c) Schematic of the site area.
-Figures b-c adapted from Cascalheira et al., 2017."
+Figures b-c adapted from Cascalheira et al., 2017"
 
@@
 #| fig-cap: "Simplified scheme of chert types identified in the UP sequence of the Terrace and Shelter areas of Vale Boi: a) local chert macroscopic variability (T1 to T5), including chalcedony (T1; top left); b) non-local chert types present in all occupations except for T11 which is only present in the Solutrean; c) TL09 (left) and TL06 (right) varieties within the trace lithologies group (TL) composed of non-local cherts with 1-3 samples per TL type and limited to specific archaeological levels and areas.
-Summarised from Belmiro et al., (2025)."
+Summarised from Belmiro et al., (2025)"
 
@@
 Finally, the Gravettian technological analysis was the result of a recent and unpublished study, focusing on the chert lithic artefacts from levels 6 and 7 of the Terrace area (\> 2012 excavations).
-This dataset and preliminary technological analysis results can be found in our online repository (<http://www.doi.org/10.17605/OSF.IO/UWMPK>) and in the Online Supplementary Materials (SOM 1).
+This dataset and preliminary technological analysis results can be found in our online repository (<http://www.doi.org/10.17605/OSF.IO/UWMPK>) and in Online Resource 1.
 
@@
 Results are structured according to the predefined analytical parameters, comparing local and non-local cherts (\@fig-cherts) across different assemblages.
-A comprehensive description of each assemblage and detailed tables for all analyses are provided in the supplementary materials (**SOM 1**).
+A comprehensive description of each assemblage and detailed tables for all analyses are provided in Online Resource1.
 
@@
 Labels within the plots correspond to the number of artefacts within each class.
-Only labels of n>3 are shown.
+Only labels of n>3 are shown
 
@@
 Tool diversity is greater among non-local cherts, which contain a high frequency of generalized tools (e.g., notches, retouched pieces) alongside hunting implements such as backed bladelets.
-The abundance of burins and burin spalls suggests they may have functioned as cores, indicating a technological strategy rather than solely a focus on retouch (**SOM 1** **Table s2**).
+The abundance of burins and burin spalls suggests they may have functioned as cores, indicating a technological strategy rather than solely a focus on retouch (Online Resource1 Table s2).
 In comparison to the Gravettian, other assemblages show smaller numbers of retouched tools.
-While in the Proto-Solutrean (**SOM 1** **Table s4**) no significant difference was identified between local and non-local retouched tools, in the Solutrean assemblage (**SOM 1** **Table s6**) both tool diversity and tool specialization are higher in local cherts.
-
-Retouched tools were further explored through the tools-to-debitage ratio (\@fig-ratios a and **SOM 1 Table s8**).
+While in the Proto-Solutrean (Online Resource 1 Table s4) no significant difference was identified between local and non-local retouched tools, in the Solutrean assemblage (Online Resource 1 Table s6) both tool diversity and tool specialization are higher in local cherts.
+
+Retouched tools were further explored through the tools-to-debitage ratio (\@fig-ratios a and Online Resource1 Table s8).
 In general, the results show that non-local chert types have higher retouch-to-debitage ratios than local types, albeit with differences between assemblages.
@@
 This pattern is less clear in the Solutrean assemblage since the ratios are the lowest of all assemblages.
-Non-local cherts show ratios above 0.1 (with exception of T8) while local cherts show ratios below 0.1 (with exception of T1; **SOM 1 Table s8 c**).
-
-Blank-to-core ratios were analysed across assemblages (\@fig-ratios b, **SOM 1 Table s7**).
+Non-local cherts show ratios above 0.1 (with exception of T8) while local cherts show ratios below 0.1 (with exception of T1; Online Resource1 Table s8 c).
+
+Blank-to-core ratios were analysed across assemblages (\@fig-ratios b, Online Resource1 Table s7).
 In all cases, local cherts generally exhibit lower blank-to-core ratios than non-local cherts, though variability in core sample sizes may influence these patterns.
@@
 The Proto-Solutrean assemblage has most blank-to-core ratios below 10, except for non-local T7.
-However, non-local cores are poorly represented within the assemblage (\@tbl-ratios b; **SOM 1 Table s3**).
+However, non-local cores are poorly represented within the assemblage (\@tbl-ratios b; Online Resource 1 Table s3).
 
@@
 #| fig-cap: Plotted ratios of blanks-to-core (a) and tool-to-debitage (b) by assemblage (Gravettian, Proto-Solutrean and Solutrean) and by chert type.
-All local cherts were grouped in a single category "Local", while non-local cherts maintained their individual types.
+All local cherts were grouped in a single category "Local", while non-local cherts maintained their individual types
 
@@
 
-The sizes of blanks and cores were explored through scatterplots of width and length (\@fig-gravettian-plots**, @fig-proto-plots**, \@fig-solutrean-plots a) and boxplots for weight (\@fig-gravettian-plots, @fig-proto-plots, \@fig-solutrean-plots b) and through non-parametric statistical tests; complete tables with descriptive statistics and non-parametric statistical test results can be found in **SOM 1 Tables s9-11** and **s12**, respectively.
+The sizes of blanks and cores were explored through scatterplots of width and length (\@fig-gravettian-plots**, @fig-proto-plots**, \@fig-solutrean-plots a) and boxplots for weight (\@fig-gravettian-plots, @fig-proto-plots, \@fig-solutrean-plots b) and through non-parametric statistical tests; complete tables with descriptive statistics and non-parametric statistical test results can be found in Online Resource1 Tables s9-11 and s12, respectively.
 
@@
 a) Scatterplots of width and length.
-b) Boxplots of weight; diamond symbol represents mean values within each group."
+b) Boxplots of weight; diamond symbol represents mean values within each group"
 
@@
 Our results indicate that most non-local cherts were introduced to Vale Boi as cores or nodules, subsequently knapped on-site.
-However, an exception is seen in TL cherts (TL2-6, TL9-10, TL14; **SOM 1 Table S13**), which lack associated cores and knapping debris.
+However, an exception is seen in TL cherts (TL2-6, TL9-10, TL14; Online Resource1 Table S13), which lack associated cores and knapping debris.
 This absence suggests that TL cherts were introduced to the site primarily as finished tools or blanks, aligning with long-distance individual provisioning models [@kuhn2004].
@@
 
-# Acknowledgements
+# Supplementary Information
+
+**Online Resource 1:** Supplementary results and tables, supporting information for the results presented in the main paper.
+Includes detailed assemblage technological descriptions by chert type, tool-to-debitage and blank-to-core ratios tables, descriptive statistics tables for measurements (mean and standard deviation) and results of the Mann-Whitney U statistical analysis.
+
+# Statements and Declarations
+
+## Funding
+
+This study was supported by
+
+## Competing Interests
+
+The authors have no competing interests to declare that are relevant to the content of this article.
+
+## Data and code availability
+
+All datasets and code used to write this article and supplementary materials, including analyses and visual representations, can be accessed and downloaded in an online repository: <http://www.doi.org/10.17605/OSF.IO/UWMPK>.
 

### 2025-04-15 analysis/Supplementary/ESM_1.qmd
Cleaning up paper and som for submission 02

@@
       email: jfbelmiro@ualg.pt
-      orcid: 0000-0003-1689-0557
+      orcid: 0000-0003-4951-2907
       institute:
@@
       institute: ICArEHB
-      orcid: 0000-0003-1689-0558
+      orcid: 0000-0001-9655-0549
   - Xavier Terradas:
       institute: CSIC
-      orcid: 0000-0003-1689-0559
+      orcid: 0000-0002-8000-5607
   - João Cascalheira:
-      orcid: 0000-0003-1689-0551
+      orcid: 0000-0003-0321-8892
       institute: ICArEHB
@@
   - ICArEHB:
-      name: Interdisciplinary Center for Archaeology and Human Evolution
-      address: 
-  - CSIC: CSIC
+      name: Interdisciplinary Center for Archaeology and the Evolution of Human Behaviour, University of Algarve, Faro, Portugal
+  - CSIC:
+      name: Spanish National Research Council - Institute Milá y Fontanals of research on Humanities, Barcelona, Spain
 title-block-published: "Last updated"  
@@
 bibliography: ../paper/references.bib
-csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
+csl: "../templates/archaeological-and-anthropological-sciences.csl" # Insert path for the bib-style
 ---
@@
   mutate(RMTYPE = case_when(
-    RMTYPE == "TYPE 2" ~ "T2",
+    RMTYPE == "TYPE 2" | RMTYPE == "Type 2" ~ "T2",
     RMTYPE == "Chalcedony" ~ "T1", # Altering first group names to final group names. Subtypes are maintained as before.
@@
   filter(Layer %in% c("A", "B", "C")) |> # Filtering Solutrean layers
-  filter(`Raw material` %in% c("Chert", "Chalcedony"))
+  filter(`Raw material` %in% c("Chert", "Chalcedony")) |> 
+  filter(Class != "Chip")
 
@@
 
+```{r}
+#| label: TL-cherts
+
+TL_classes_terrace <- phase_dataset |> 
+  filter(RMTYPE == "TL") |> 
+  filter(Level != "5") |>
+  select(Technocomplex, SUBTYPE, ID, CLASS, RETOUCHEDPIECETYPOLOGY)
+
+TL_classes_shelter <- solutrean_dataset |> 
+  mutate(Technocomplex = case_when(
+    LEVEL == "A" | LEVEL == "B" | LEVEL == "C" ~ "Solutrean")) |> 
+  filter(RMTYPE == "TL") |> 
+  select(Technocomplex, SUBTYPE, ID, CLASS, RETOUCHEDPIECETYPOLOGY)
+
+TL_classes <- rbind(TL_classes_terrace, TL_classes_shelter)
+TL_classes <- TL_classes |> 
+  rename(TL = SUBTYPE) |> 
+  arrange(TL)
+
+```
+
 ## Gravettian (levels 6-7)
@@
 Of these, 209 (`r round(209*100/766, 2)`%) have an indeterminate chert type (@tbl-general-gravettian).
-Chalcedony (T1) and local cherts (T2-5) show a total of `r 35+110+17+13+72` artefacts (`r round((35+110+17+13+72)*100/766, 2)`%).
+Chalcedony (T1) and local cherts (T2-5) show a total of `r 35+212` artefacts (`r round((35+212)*100/766, 2)`%).
 Non-local cherts show different numbers of artefacts depending on the type, with T6 and T7 showing the highest values (169 or `r round(169*100/766, 2)`% of the total assemblage, and 86 or `r round(86*100/766, 2)`%, respectively); the other non-local cherts have artefact counts lower than 20 per type.
-The TL cherts also show smaller numbers, with 12 artefacts, corresponding to 7 different types of TL cherts (table s++ in SOM ++).
+The TL cherts also show smaller numbers, with 12 artefacts, corresponding to 7 different types of TL cherts [@tbl-TL-classes].
+
+```{r}
+#| label: tbl-general-gravettian
+#| tbl-cap: Number and percentage of technotypological classes by raw material type from the Gravettian assemblage (levels 6 and 7) of the Terrace area of Vale Boi.
+
+kable(gravettian_general_som)
+
+```
+
+```{r}
+#| label: tbl-TL-classes
+#| tbl-cap: Trace lithologies cherts from the Gravettian and Proto-Solutrean assemblages of the Terrace and Solutrean assemblage from the Shelter and corresponding technological class.
+
+kable(TL_classes)
+
+```
 
@@
 For local cherts, blanks (including blank fragments) represent more than 50% of the artefacts.
-Cores and core maintenance products are only present in small frequencies: <10% and <2%, respectively.
-Retouched tools seem to be especially relevant in local T2-5 cherts (~16%), making them the most frequent class after blanks and blank fragments, followed by shatter (~10%).
+Cores and core maintenance products are only present in small frequencies: \<10% and \<2%, respectively.
+Retouched tools seem to be especially relevant in local T2-5 cherts (\~16%), making them the most frequent class after blanks and blank fragments, followed by shatter (\~10%).
 T1 (chalcedony) shows similar patterns to other local cherts, although with higher percentages of shatter (34.3%).
@@
 
-```{r}
-#| label: tbl-general-gravettian
-#| tbl-cap: Number and percentage of technotypological classes by raw material type from the Gravettian assemblage (levels 6 and 7) of the Terrace area of Vale Boi.
-
-kable(gravettian_general_som)
-
-```
-
-As seen in table @tbl-retouch-gravettian, 170 retouched tools were identified, with 39 (`r round(39*100/170, 2)`%) having an unidentified chert type.
+As seen in @tbl-retouch-gravettian, 170 retouched tools were identified, with 39 (`r round(39*100/170, 2)`%) having an unidentified chert type.
 From the identified chert types, `r 3+20+4+2+9` (`r round((3+20+4+2+9)*100/170, 2)`%) of retouched tools are from local types, `r 58+24+3+1+3` (`r round((58+24+3+1+3)*100/170, 2)`%) from non-local cherts and 4 are from the TL group.
@@
 All these typologies were found in the non-local chert group, while only 5 retouched tool typologies were identified in local cherts/chalcedony.
-The most common typologies identified in the assemblage (and for all chert types) were burins (n=52), endscrappers (n=17), retouched flakes (n=30) and splintered pieces (n=26).
+The most common typologies identified in the assemblage (and for all chert types) were burins (n=54), endscrappers (n=18), retouched blanks (n=38) and splintered pieces (n=26).
 From all these, the majority were produced in non-local chert, mainly T6 and in lesser numbers T7.
 Although in lesser numbers as well, backed bladelets were also identified (n=6), but only present in non-local chert types (T6 and T7).
-The presence of burin spalls in local chert, especially T2 (n=9) and non-local cherts (T6 with n=12 and T7 with n=4) is expected due to the prevalence of burins (@tbl-general-gravettian).
+The presence of burin spalls in local chert, especially T2 (n=11) and non-local cherts (T6 with n=12 and T7 with n=4) is expected due to the prevalence of burins (@tbl-general-gravettian).
 
@@
 Of these, 167 (`r round(167*100/635, 2)`%) have an indeterminate chert type.
-Local cherts correspond to `r 32+62+32+158+51` artefacts (`r round((32+62+32+158+51)*100/635, 2)`%), with T4 being the most present local chert as seen in previous works (Belmiro et al., 2025).
-Non-local types are present in smaller numbers (n=`r 60+47+13+9+1` or `r round((60+47+13+9+1)*100/635, 2)`%), with T6 and T7 being the most relevant (n=60 and n=47, respectively), and the TL group includes 3 artefacts, corresponding to 2 types of TL cherts (Table s++ in SOM ++).
-
-When crossing the lithic class with chert type (@tbl-general-proto), we see that the local types show high frequency of blanks (\>50%, including blank fragments), followed by shatter (\~11-27%).
-Cores are present in small numbers (n=`r 4+5+13+6`), corresponding to ~10% of the classes in local types, with the exception of T1 (chalcedony), with no cores.
+Local cherts correspond to `r 32+62+32+158+51` artefacts (`r round((32+62+32+158+51)*100/635, 2)`%), with T4 being the most present local chert.
+Non-local types are present in smaller numbers (n=`r 60+47+13+9+1` or `r round((60+47+13+9+1)*100/635, 2)`%), with T6 and T7 being the most relevant (n=60 and n=47, respectively), and the TL group includes 3 artefacts, corresponding to 2 types of TL cherts [@tbl-TL-classes].
+
+When crossing the lithic class with chert type (@tbl-general-proto), we see that the local types show high frequency of blanks (\>50%, including blank fragments), followed by shatter (\~20-30%).
+Cores are present in small numbers (n=`r 4+5+13+6`), corresponding to \~10% of the classes in local types, with the exception of T1 (chalcedony), with no cores.
 Core preparation and maintenance products are absent in all local cherts.
@@
 The Solutrean assemblages show similar patterns to those observed in previous assemblages.
-The final sample from levels A, B and C from the Solutrean occupations of the Shelter area consists of 914 artifacts.
+The final sample from levels A, B and C from the Solutrean occupations of the Shelter area consists of 910 artifacts.
 Of these, 201 (`r round(201*100/914, 2)`%) correspond to the indeterminate chert group.
-Local cherts correspond to `r 16+506+4+10+30` (`r round((16+506+4+10+30)*100/914, 2)`%) artefacts, mostly from T2 as seen in previous works (Belmiro et al., 2025).
-Non-local cherts (including the TL group) correspond to `r 86+9+30+3+1+17+1` (`r round((86+9+30+3+1+17+1)*100/914, 2)`%) artefacts, mostly from T6 (n=86).
-
-When crossing the class data with type of raw material (@tbl-general-solutrean), we see that for local chert, the majority of the debitage is composed of blanks (83.9%).
-Cores are also present (10.5%), alongside a small percentage of core preparation and maintenance products (1.3%).
+Local cherts correspond to `r 16+548` (`r round((16+548)*100/914, 2)`%) artefacts, mostly from T2.
+Non-local cherts (including the TL group) correspond to `r 85+9+30+3+1+16+1` (`r round((85+9+30+3+1+16+1)*100/914, 2)`%) artefacts, mostly from T6 (n=86).
+
+When crossing the class data with type of raw material (@tbl-general-solutrean), we see that for local chert, the majority of the debitage is composed of blanks (~80%).
+Cores are also present (~10%), alongside a small percentage of core preparation and maintenance products (1.3%).
 In comparison to debitage products, retouched tools are scarce in the sample (\~3%).
 Non-local cherts, especially T6 and T8, show similar patterns to the local cherts, with high percentages of blanks (\>50%).
-Cores and core preparation and maintenance are also present in small percentages (\~3-6% and 3.5% only for T6, respectively), as well as retouched tools (\<10%).
+Cores and core preparation and maintenance are also present in small percentages (\~6-3% and 3.5% only for T6, respectively), as well as retouched tools (\<10%).
 However, these specific results are truncated by the lack of shatter, which were not present in the sample used for the study (see Materials and Methods section).
@@
 The analysed Solutrean sample has a total of 43 retouched tools (complete and fragments), with `r 3+13+1` corresponding to local cherts and `r 7+1+2+3` to non-local cherts, with the remaining being cherts of unidentified type.
-These results are significantly reduced from the original dataset (Cascalheira et al., 2013), possibly due to no bucket coordinates being used.
+These results are significantly reduced from the original dataset, possibly due to no bucket coordinates being used.
 From the 43 retouched tools, 11 different typologies were identified (without accounting for fragments).
@@
 Two types of points were identified: shouldered points (n=3, including fragments) and tanged and winged points (n=4).
-Other frequent retouched typologies are retouched blanks (n=6), notches (n=5), splintered pieces (n=5) and endscrappers (n=4).
+Other frequent retouched typologies are retouched blanks (n=7), notches (n=5), splintered pieces (n=5) and endscrappers (n=4).
 Local types show a higher variability retouched typologies (9 types), although this may be related to the higher number of tools.
@@
 
+# Tool-to-debitage ratios
+
+```{r}
+#| label: ratios-retouchdebitage
+
+# Solutrean
+solutrean_ratiotd <- solutrean_dataset %>% 
+  filter(RMTYPE != "INDET") %>% 
+  mutate(CLASS = case_when(
+    CLASS == "Flake" | CLASS == "ElongBlank" | CLASS == "Thinning flake" ~ "Blank",
+    CLASS == "FlakeFrag" | CLASS == "ElongBlankFrag" ~ "BlankFrag",
+    CLASS == "Retouched piece" ~ "RetouchedPiece",
+    CLASS == "Retouched piece fragment" ~ "RetouchedPieceFrag",
+    TRUE ~ CLASS)) %>% 
+  mutate(CLASS = case_when(
+    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
+    CLASS == "RetouchedPieceFrag" & PIECECOMPLETENESS == "Proximal" ~ "RetouchedPiece",
+    TRUE ~ CLASS)) %>% 
+  filter(CLASS %in% c("Blank", "RetouchedPiece")) %>% 
+  count(RMTYPE, CLASS, sort = TRUE) %>% 
+  group_by(RMTYPE, CLASS) %>%
+  mutate(RMTYPE = factor(RMTYPE, levels = c("T1", "T2", "T3", "T4", "T5", "T6","T7", 
+                                        "T8", "T9", "T10", "T11", "TL"))) |>
+  summarise(n=sum(n)) %>% 
+  spread(CLASS, n, fill=0) %>% 
+  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
+  filter(!sum <= 5) %>% 
+  mutate("Tool/Blank" = RetouchedPiece / Blank)
+
+# Proto-Solutrean
+proto_ratiotd <- phase_dataset %>% 
+  filter(Level %in% c("4E", "4E/5")) %>% 
+  filter(RMTYPE != "INDET") %>% 
+  mutate(CLASS = case_when(
+    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
+    CLASS == "RetouchedPieceFrag" & PIECECOMPLETENESS == "Proximal" ~ "RetouchedPiece",
+    TRUE ~ CLASS)) %>% 
+  filter(CLASS %in% c("Blank", "RetouchedPiece")) %>% 
+  count(RMTYPE, CLASS, sort = TRUE) %>% 
+  group_by(RMTYPE, CLASS) %>%
+  mutate(RMTYPE = factor(RMTYPE, levels = c("T1", "T2", "T3", "T4", "T5", "T6","T7", 
+                                        "T8", "T9", "T10", "TL"))) |>
+  summarise(n=sum(n)) %>% 
+  spread(CLASS, n, fill=0) %>% 
+  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
+  filter(!sum <= 5) %>% 
+  mutate("Tool/Blank" = RetouchedPiece / Blank)
+
+# Gravettian
+gravettian_ratiotd <- phase_dataset %>% 
+  filter(Level %in% c(6, 7)) %>% 
+  filter(RMTYPE != "INDET") %>% 
+  mutate(CLASS = case_when(
+    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
+    CLASS == "RetouchedPieceFrag" & PIECECOMPLETENESS == "Proximal" ~ "RetouchedPiece",
+    TRUE ~ CLASS)) %>% 
+  filter(CLASS %in% c("Blank", "RetouchedPiece")) %>% 
+  count(RMTYPE, CLASS, sort = TRUE) %>% 
+  group_by(RMTYPE, CLASS) %>%
+  mutate(RMTYPE = factor(RMTYPE, levels = c("T1", "T2", "T3", "T4", "T5", "T6","T7", 
+                                        "T8", "T9", "T10", "TL"))) |>
+  summarise(n=sum(n)) %>% 
+  spread(CLASS, n, fill=0) %>% 
+  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
+  filter(!sum <= 5) %>% 
+  mutate("Tool/Blank" = RetouchedPiece / Blank)
+
+```
+
+```{r}
+#| label: tbl-tdratio-solutrean
+#| tbl-cap: Tools to debitage ratio by raw material of the Gravettian (a), Proto-Solutrean (b) and Solutrean (c) assemblages.
+#| tbl-subcap: 
+#|  - "Tools to debitage ratio by raw material of the Gravettian assemblage (levels 6 and 7) of the Terrace area."
+#|  - "Tools to debitage ratio by raw material of the Proto-Solutrean assemblage (levels 4E and top of level 5) of the Terrace area."
+#|  - "Tools to debitage ratio by raw material of the Solutrean assemblage (levels A to C) of the Shelter area."
+#| layout-nrow: 3
+
+kable(gravettian_ratiotd)
+kable(proto_ratiotd)
+kable(solutrean_ratiotd)
+
+```
+
 # Blank-to-core ratios
+
 ```{r}
@@
 
-# Tool-to-debitage ratios
-```{r}
-#| label: ratios-retouchdebitage
-
-# Solutrean
-solutrean_ratiotd <- solutrean_dataset %>% 
-  filter(RMTYPE != "INDET") %>% 
-  mutate(CLASS = case_when(
-    CLASS == "Flake" | CLASS == "ElongBlank" | CLASS == "Thinning flake" ~ "Blank",
-    CLASS == "FlakeFrag" | CLASS == "ElongBlankFrag" ~ "BlankFrag",
-    CLASS == "Retouched piece" ~ "RetouchedPiece",
-    CLASS == "Retouched piece fragment" ~ "RetouchedPieceFrag",
-    TRUE ~ CLASS)) %>% 
-  mutate(CLASS = case_when(
-    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
-    CLASS == "RetouchedPieceFrag" & PIECECOMPLETENESS == "Proximal" ~ "RetouchedPiece",
-    TRUE ~ CLASS)) %>% 
-  filter(CLASS %in% c("Blank", "RetouchedPiece")) %>% 
-  count(RMTYPE, CLASS, sort = TRUE) %>% 
-  group_by(RMTYPE, CLASS) %>%
-  mutate(RMTYPE = factor(RMTYPE, levels = c("T1", "T2", "T3", "T4", "T5", "T6","T7", 
-                                        "T8", "T9", "T10", "T11", "TL"))) |>
-  summarise(n=sum(n)) %>% 
-  spread(CLASS, n, fill=0) %>% 
-  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
-  filter(!sum <= 5) %>% 
-  mutate("Tool/Blank" = RetouchedPiece / Blank)
-
-# Proto-Solutrean
-proto_ratiotd <- phase_dataset %>% 
-  filter(Level %in% c("4E", "4E/5")) %>% 
-  filter(RMTYPE != "INDET") %>% 
-  mutate(CLASS = case_when(
-    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
-    CLASS == "RetouchedPieceFrag" & PIECECOMPLETENESS == "Proximal" ~ "RetouchedPiece",
-    TRUE ~ CLASS)) %>% 
-  filter(CLASS %in% c("Blank", "RetouchedPiece")) %>% 
-  count(RMTYPE, CLASS, sort = TRUE) %>% 
-  group_by(RMTYPE, CLASS) %>%
-   mutate(RMTYPE = factor(RMTYPE, levels = c("T1", "T2", "T3", "T4", "T5", "T6","T7", 
-                                        "T8", "T9", "T10", "TL"))) |>
-  summarise(n=sum(n)) %>% 
-  spread(CLASS, n, fill=0) %>% 
-  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
-  filter(!sum <= 5) %>% 
-  mutate("Tool/Blank" = RetouchedPiece / Blank)
-
-# Gravettian
-gravettian_ratiotd <- phase_dataset %>% 
-  filter(Level %in% c(6, 7)) %>% 
-  filter(RMTYPE != "INDET") %>% 
-  mutate(CLASS = case_when(
-    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
-    CLASS == "RetouchedPieceFrag" & PIECECOMPLETENESS == "Proximal" ~ "RetouchedPiece",
-    TRUE ~ CLASS)) %>% 
-  filter(CLASS %in% c("Blank", "RetouchedPiece")) %>% 
-  count(RMTYPE, CLASS, sort = TRUE) %>% 
-  group_by(RMTYPE, CLASS) %>%
-  mutate(RMTYPE = factor(RMTYPE, levels = c("T1", "T2", "T3", "T4", "T5", "T6","T7", 
-                                        "T8", "T9", "T10", "TL"))) |>
-  summarise(n=sum(n)) %>% 
-  spread(CLASS, n, fill=0) %>% 
-  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
-  filter(!sum <= 5) %>% 
-  mutate("Tool/Blank" = RetouchedPiece / Blank)
-
-```
-
-```{r}
-#| label: tbl-tdratio-solutrean
-#| tbl-cap: Tools to debitage ratio by raw material of the Gravettian (a), Proto-Solutrean (b) and Solutrean (c) assemblages.
-#| tbl-subcap: 
-#|  - "Tools to debitage ratio by raw material of the Gravettian assemblage (levels 6 and 7) of the Terrace area."
-#|  - "Tools to debitage ratio by raw material of the Proto-Solutrean assemblage (levels 4E and top of level 5) of the Terrace area."
-#|  - "Tools to debitage ratio by raw material of the Solutrean assemblage (levels A to C) of the Shelter area."
-#| layout-nrow: 3
-
-kable(gravettian_ratiotd)
-kable(proto_ratiotd)
-kable(solutrean_ratiotd)
-
-```
-
 # Measurements
+
 ```{r}
@@
   mutate(CLASS = case_when(
-    CLASS == "Thinning flake" | CLASS == "ElongBlank" | CLASS == "Flake" ~ "Blank",
+    CLASS == "ElongBlank" | CLASS == "Flake" ~ "Blank",
     TRUE ~ CLASS)) |> 
@@
 
-The results indicate generally significant differences in blank measurements between local and non-local cherts in all technocomplexes; in almost all measurements and groups, local cherts show higher medians.
-Length seems to be consistently significant (p < 0.01 and < 0.001) between the two groups in all assemblages, with local cherts showing higher median lengths than non-local cherts.
-In the Gravettian and Proto-Solutrean assemblages, weight is also significantly different between local and non-local blanks (p < 0.05), with local cherts presenting higher medians.
+The results indicate generally significant differences in blank measurements between local and non-local cherts in all technocomplexes; in almost all measurements and groups, local cherts show higher medians [@tbl-pvalue-1].
+Length seems to be consistently significant (p \< 0.01 and \< 0.001) between the two groups in all assemblages, with local cherts showing higher median lengths than non-local cherts.
+In the Gravettian and Proto-Solutrean assemblages, weight is also significantly different between local and non-local blanks (p \< 0.05), with local cherts presenting higher medians.
 In comparison, width differences show no statistical significance, with exception of the Solutrean assemblage.
-In the Solutrean, blank width is significant (p < 0.05) with the non-local category showing higher medians compared local cherts.
-
-Regarding the cores, in almost all measurements and groups local cherts show higher means than non-local cherts.
-Despite these differences, the results are not significant  (p > 0.05).
-It is important to note, however, that the sample size for cores in the Gravettian assemblage and non-local cherts in the Solutrean assemblage are smaller (n < 15) than the blank sample size.
+In the Solutrean, blank width is significant (p \< 0.05) with the non-local category showing higher medians compared local cherts.
+
+Regarding the cores, in almost all measurements and groups local cherts show higher means than non-local cherts [@tbl-pvalue-2].
+Despite these differences, the results are not significant (p \> 0.05).
+It is important to note, however, that the sample size for cores in the Gravettian assemblage and non-local cherts in the Solutrean assemblage are smaller (n \< 15) than the blank sample size.
 
@@
 ```
-
-# TL cherts
-
-```{r}
-#| label: tbl-TL-classes
-#| tbl-cap: Trace lithologies cherts from the Gravettian and Proto-Solutrean assemblages of the Terrace and Solutrean assemblage from the Shelter and corresponding technological class.
-
-TL_classes_terrace <- phase_dataset |> 
-  filter(RMTYPE == "TL") |> 
-  filter(Level != "5") |>
-  select(Technocomplex, SUBTYPE, ID, CLASS, RETOUCHEDPIECETYPOLOGY)
-
-TL_classes_shelter <- solutrean_dataset |> 
-  mutate(Technocomplex = case_when(
-    LEVEL == "A" | LEVEL == "B" | LEVEL == "C" ~ "Solutrean")) |> 
-  filter(RMTYPE == "TL") |> 
-  select(Technocomplex, SUBTYPE, ID, CLASS, RETOUCHEDPIECETYPOLOGY)
-
-TL_classes <- rbind(TL_classes_terrace, TL_classes_shelter)
-TL_classes <- TL_classes |> 
-  rename(TL = SUBTYPE) |> 
-  arrange(TL)
-
-kable(TL_classes)
-
-```

### 2025-04-15 analysis/paper/paper.qmd
Cleaning up paper and som for submission 02

@@
       email: jfbelmiro@ualg.pt
-      orcid: 0000-0003-1689-0557
+      orcid: https://orcid.org/0000-0003-4951-2907
       institute:
-        - ICArEHB
+        - Interdisciplinary Center for Archaeology and the Evolution of Human Behaviour, University of Algarve, Faro, Portugal
   - Nuno Bicho:
-      institute: ICArEHB
-      orcid: 0000-0003-1689-0558
+      orcid: https://orcid.org/0000-0001-9655-0549
+      institute:
+        - Interdisciplinary Center for Archaeology and the Evolution of Human Behaviour, University of Algarve, Faro, Portugal
   - Xavier Terradas:
-      institute: CSIC
-      orcid: 0000-0003-1689-0559
+      orcid: https://orcid.org/0000-0002-8000-5607
+      institute:
+        - Institución Milá y Fontanals de Investigación en Humanidades, Consejo Superior de Investigaciones Científicas, Barcelona, Spain
   - João Cascalheira:
-      orcid: 0000-0003-1689-0551
-      institute: ICArEHB
+      orcid: https://orcid.org/0000-0003-0321-8892
       institute:
-  - ICArEHB:
-      name: Interdisciplinary Center for Archaeology and Human Evolution
-      address: 
-  - CSIC: CSIC
+        - Interdisciplinary Center for Archaeology and the Evolution of Human Behaviour, University of Algarve, Faro, Portugal
 title-block-published: "Last updated"
-date: now
-date-format: long
 format:
@@
   fig-dpi: 600
+number-sections: true
 filters:
@@
 bibliography: references.bib
-csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
+csl: "../templates/archaeological-and-anthropological-sciences.csl" # Insert path for the bib-style
 abstract: |
@@
 
-Highlights: `r rmarkdown::metadata$highlights`
-
 <!-- The actual document text starts here: -->
 
-# Acknowledgements
+# Acknowledgements {.unnumbered}
+
+We would like to thank the Fundacão para a Ciência e Tecnologia for the funding essential for our research, and ICArEHB for funding and providing access to the archaeological materials.
 
@@
 
-In Westernmost Europe, five main technocomplexes have been identified within the UP  - Aurignacian, Gravettian, Proto-Solutrean, Solutrean, and Magdalenian - each contributing new evidence on resilient behaviors and cultural continuity.
+In westernmost Europe, five main technocomplexes have been identified within the UP - Aurignacian, Gravettian, Proto-Solutrean, Solutrean, and Magdalenian - each contributing new evidence on resilient behaviors and cultural continuity.
 These technocomplexes are distinguished by regional adaptations as well as technological and subsistence innovations [@baker2024; @bicho2017; @bicho2011; @bicho2008; @cascalheira2017; @straus1991; @straus1993; @zilhão1997].
@@
 
-In particular, the Iberian Peninsula has been recognized as a key region for studying UP adaptations and cultural transitions [e.g., @aubry2016; @aubry2022; @bicho2011; @bicho2017a; @bradtmöller2012; @cascalheira2017a; @cascalheira2021; @cascalheira2013; @finlayson2006; @gamble2004; @jennings2011; @jiménez-espejo2007; @linscott2023; @rasilla2020; @schmidt2012; @zilhão1997a; @zilhão2021; @zilhão2017a].
+In particular, the Iberian Peninsula has been recognized as a key region for studying UP adaptations and cultural transitions [e.g., @aubry2016; @aubry2022; @bicho2011; @bicho2017a; @bradtmöller2012; @cascalheira2017; @cascalheira2021; @cascalheira2013; @finlayson2006; @gamble2004; @jennings2011; @jiménez-espejo2007; @linscott2023; @rasilla2020; @schmidt2012; @zilhão1997; @zilhão2021; @zilhão2017].
 This significance is largely due to its location in southwestern Europe, which enabled it to maintain a relatively temperate climate even during the coldest periods, such as the LGM.
@@
 These differences were suggested to be related to Vale Boi’s peripheral location and the distinct ecological characteristics of westernmost Algarve [@marreiros2013].
-In contrast, Proto-Solutrean evidence from the site indicates technological similarities and strong connections with central Portugal [@belmiro2021], while the Solutrean period stands out as a highly dynamic phase at Vale Boi, during which the site functioned as a key connection point between central Portugal and southern Spain, facilitating social networks and cultural exchanges [@bicho2017a; \@cascalheira2013a].
+In contrast, Proto-Solutrean evidence from the site indicates technological similarities and strong connections with central Portugal [@belmiro2021], while the Solutrean period stands out as a highly dynamic phase at Vale Boi, during which the site functioned as a key connection point between central Portugal and southern Spain, facilitating social networks and cultural exchanges [@bicho2017a; @cascalheira2013c].
 
@@
 Raw material procurement is a fundamental aspect of lithic technological organisation, as it played a crucial role in the lives of past hunter-gatherer groups by determining the possibilities for tool production and use [@bamforth1997].
-Approaches focused on the study of raw materials provide key knowledge on several topics of hunter-gatherer organisation such as procurement and mobility strategies [@ambrose1990; @binford1979; @binford1985; @gould1985; @kuhn1991; @mccall2007; @gramly1980], occupation types and their duration [@torrence1983; @grove2023; @kuhn2004a; @surovell2009], and social networks and exchanges [@gamble1999; @whallon2006].
+Approaches focused on the study of raw materials provide key knowledge on several topics of hunter-gatherer organisation such as procurement and mobility strategies [@ambrose1990; @binford1979; @binford1985; @gould1985; @kuhn1991; @mccall2007; @gramly1980], occupation types and their duration [@torrence1983; @grove2023; @kuhn2004; @surovell2009], and social networks and exchanges [@gamble1999; @whallon2006].
 However, raw material studies often focus on the characterisation of geological and archaeological raw materials to identify local and non-local raw materials within an assemblage, their possible sources, and examine their proportions within a site over different occupations and periods.
@@
 
-Vale Boi is located on the western coast of the Algarve region, southern Portugal, within a small valley extending \~2 km southward to the Atlantic Ocean (\@fig-map a).
-The site is nestled among limestone outcrops that create west- and southwest-facing rock shelters (\@fig-map b).
-Covering more than 10,000 square meters along the valley slope, Vale Boi includes three main excavated areas, Slope, Terrace, and Shelter (\@fig-map c), which were excavated between 2000 and 2019 [@bicho2007; @bicho2012; @bicho2013; @cascalheira2017; @cascalheira2013; @manne2012].
+Vale Boi is located on the western coast of the Algarve region, southern Portugal, at a latitude of N 37º5'24.1" and longitude of W 8º48'32.4" (@gibajaFUNCIONINSTRUMENTOSLITICOS2006; the original site coordinates were converted from Degree/Minute to Degree/Minute/Second), within a small valley extending \~2 km southward to the Atlantic Ocean [@fig-map a].
+The site is nestled among limestone outcrops that create west- and southwest-facing rock shelters [@fig-map b].
+Covering more than 10,000 square meters along the valley slope, Vale Boi includes three main excavated areas, Slope, Terrace, and Shelter [@fig-map c], which were excavated between 2000 and 2019 [@bicho2007; @bicho2012; @bicho2013; @cascalheira2017; @cascalheira2013; @manne2012].
 Based on the available raw material and technological evidence, particularly the macroscopic and petrographic data [@belmiro2025], the Terrace and Shelter areas were selected for the current study.
@@
 #| fig-cap: "a) Location of the archaeological site of Vale Boi (orange box) within the Iberian Peninsula; b) Photo of the limestone hill with terrace and shelter locations; c) Schematic of the site area.
-Figures b-c adapted from Cascalheira et al., 2017"
+Figures b-c) were adapted from Cascalheira et al., 2017.
+Figure a) was produced with resource to the open-source software QGIS (https://qgis.org/)"
+
+knitr::include_graphics(here::here("analysis/figures/VB_map.png"))
 
@@
 32 and 27 ka cal BP, Proto-Solutrean (levels 5 to 4E) between approximately c.
-26 and 24 ka cal BP, and Solutrean (levels 4D, 4C, 4, and Lower 3) between approximately c. 24 and 20 ka cal BP [@belmiro2021; @cascalheira2017; @cascalheira2013b].
+26 and 24 ka cal BP, and Solutrean (levels 4D, 4C, 4, and Lower 3) between approximately c.
+24 and 20 ka cal BP [@belmiro2021; @cascalheira2017; @cascalheira2013b].
 
 The Shelter area contains four main lithostratigraphic units, with occupations from the Magdalenian, Solutrean, and Gravettian periods.
-The Magdalenian and Gravettian levels (Z and D, respectively) were partially eroded, while the Solutrean levels represent the most intensive occupation, with three well-preserved archaeological horizons (layers A to C), dated to approximately c. 24-22 ka cal BP.
+The Magdalenian and Gravettian levels (Z and D, respectively) were partially eroded, while the Solutrean levels represent the most intensive occupation, with three well-preserved archaeological horizons (layers A to C), dated to approximately c.
+24-22 ka cal BP.
 These occupation levels were found beneath blocks of limestone that likely collapsed from the rock shelter ceiling, probably shortly after the LGM [@cascalheira2010; @cascalheira2017].
@@
 Technological analyses of these assemblages have provided insights into the adaptive strategies of UP hunter-gatherers at Vale Boi, revealing technological shifts across different cultural phases.
-Given the chosen sample for this study (see **Section 2.2.
-Datasets and analysis**), we summarise here the results for the technological analysis of the Solutrean assemblages (levels A-C) of the Shelter [@cascalheira2010; @cascalheira2013a] and the Proto-Solutrean assemblages (levels 4E-5) of the Terrace [@belmiro2020; @belmiro2021].
+Given the chosen sample for this study (see Section 2.2.
+Datasets and analysis), we summarise here the results for the technological analysis of the Solutrean assemblages (levels A-C) of the Shelter [@cascalheira2010; @cascalheira2013a] and the Proto-Solutrean assemblages (levels 4E-5) of the Terrace [@belmiro2020; @belmiro2021].
 The Gravettian occupations (levels 6-8) of the Terrace have been recently studied, but the results remain unpublished.
@@
 Quartz and greywacke are locally available from streambeds near the site, whereas regional chert and chalcedony sources lie further afield, within a \~20 km radius.
-Our previous studies showed, as expected, the existence of local cherts in the assemblages (\@fig-cherts a), originating from the Lower Jurassic (cherts) and Triassic (chalcedony) outcrops in proximity to the site [@belmiro2025].
+Our previous studies showed, as expected, the existence of local cherts in the assemblages [@fig-cherts a], originating from the Lower Jurassic (cherts) and Triassic (chalcedony) outcrops in proximity to the site [@belmiro2025].
 Despite petrographic differences between cherts and chalcedony, for this study any chert varieties (e.g., chalcedony or jasper) are referred to as chert [@luedtke1978].
-A significant  portion of non-local cherts was also identified (\@fig-cherts b-c), with some identified sources, such as the Cretaceous outcrops of central Portugal (Lisbon and Rio Maior areas) and the Upper Jurassic oolitic/peloidal cherts of the Betic Systems in southern Spain.
+A significant portion of non-local cherts was also identified [@fig-cherts b-c], with some identified sources, such as the Cretaceous outcrops of central Portugal (Lisbon and Rio Maior areas) and the Upper Jurassic oolitic/peloidal cherts of the Betic Systems in southern Spain.
 This indicates varying degrees of contact or long-range mobility throughout the UP, likely influenced by shifts in settlement patterns and the duration of site occupation.
@@
 Summarised from Belmiro et al., (2025)"
+#| out-width: "500"
+
+knitr::include_graphics(here::here("analysis/figures/types_short.png"))
 
@@
 A similar conclusion was suggested based on the interpretation of the chert sourcing and proportions in the several UP occupations of the Terrace and Shelter of Vale Boi and following other models relating local/non-local raw materials with mobility [@grove2023; @surovell2009].
-The high proportions of non-local chert (\~50%) in the Gravettian occupations (levels 6 and 7), and gradual reduction in the following occupations, were interpreted as a possible result of settlement changes, where the Gravettian occupations corresponded to short-term occupations, and the other occupations corresponded to residential, long-term occupations with more infrequent moves [@belmiro2025].\
+The high proportions of non-local chert (\~50%) in the Gravettian occupations (levels 6 and 7), and gradual reduction in the following occupations, were interpreted as a possible result of settlement changes, where the Gravettian occupations corresponded to short-term occupations, and the other occupations corresponded to residential, long-term occupations with more infrequent moves [@belmiro2025].
 
@@
 The Solutrean analysis in this study is limited to the Shelter area due to the inexistence of technological studies for the better preserved Solutrean levels of the Terrace (e.g., level 4).
-However, and as seen in [@belmiro2025], the raw material patterns seem to show little differences between the Solutrean in both areas; coupled with the similar radiocarbon dates between the areas (c. 24-22 and 24-20 ka cal BP), we assume that they are representative of similar, long-term, residential occupations which aptly characterise the Solutrean groups of Vale Boi.
+However, and as seen in [@belmiro2025], the raw material patterns seem to show little differences between the Solutrean in both areas; coupled with the similar radiocarbon dates between the areas (c.
+24-22 and 24-20 ka cal BP), we assume that they are representative of similar, long-term, residential occupations which aptly characterise the Solutrean groups of Vale Boi.
 
@@
 Given that the raw material analysis focused exclusively on individually piece-plotted artifacts (dimensions \>2 cm), the final dataset does not include all excavated material but retains the most diagnostically relevant pieces for analysis.
-As a result, the final dataset used in this study does not directly correspond to the full datasets analysed in the original studies (\@tbl-samples).
+As a result, the final dataset used in this study does not directly correspond to the full datasets analysed in the original studies [@tbl-samples].
 Similarly, the raw material study used only currently accessible artifacts from the assemblages.
@@
 
-Following data homogenization and integration, we conducted descriptive statistical analyses and non-parametric statistical analyses to examine correlations between raw material use and technological patterns and test expectations related to the technological organisation of local and non-local cherts (see @tbl-expectations), and between different UP occupations.
+Following data homogenization and integration, we conducted descriptive statistical analyses and non-parametric statistical analyses to examine correlations between raw material use and technological patterns and test expectations related to the technological organisation of local and non-local cherts [see @tbl-expectations], and between different UP occupations.
 To test these expectations, we evaluated 9 parameters for each dataset and assemblage; the parameters and their calculation/measurement methods are detailed in @tbl-parameters.
@@
 table_parameters <- tibble::tribble(
-  ~Parameter, ~Calculation/Measurements, ~Unit,
+  ~Parameter, ~`Calculation/Measurements`, ~Unit,
   "P1 - Frequency of retouched tools", "N retouched tools / Total assemblage * 100", "%",
@@
 
+knitr::kable(table_parameters)
+
 ```
@@
     Level == 6 | Level == 7 ~ "Gravettian")) |> 
-  mutate(RETOUCHEDPIECETYPOLOGY = case_when( # Simplifying the retouched typology
+  mutate(RETOUCHEDPIECETYPOLOGY = case_when(
     RETOUCHEDPIECETYPOLOGY == "Splintered piece with notch" | RETOUCHEDPIECETYPOLOGY == "Splintered piece with lateral retouch" | RETOUCHEDPIECETYPOLOGY == "Splintered piece fragment" ~ "Splintered piece",
@@
     RETOUCHEDPIECETYPOLOGY == "Backed bladelet" | RETOUCHEDPIECETYPOLOGY == "Backed flake" | RETOUCHEDPIECETYPOLOGY == "Double backed bladelet" | RETOUCHEDPIECETYPOLOGY == "Backed bladelet parcial" ~ "Backed tool",
+    RETOUCHEDPIECETYPOLOGY == "Carenated endscraper" ~ "Endscraper",
     TRUE ~ RETOUCHEDPIECETYPOLOGY))
@@
 ```{r}
-#| label: general-tables
-#| include: false
-
-# Tables with class by raw material type (n and %)
-##
-general_table <- phase_dataset |> 
-    dplyr::mutate(RMTYPE = case_when(
-    RMTYPE == "T1" | RMTYPE == "T2" | RMTYPE == "T3" | RMTYPE == "T4" | RMTYPE == "T5" ~ "Local",
-    TRUE ~ RMTYPE)) |> 
-  dplyr::mutate(CLASS = case_when(
-    CLASS == "BlankFrag" ~ "Blank fragment",
-    CLASS == "BurinSpall" ~ "Burin spall",
-    CLASS == "CoreFrag" ~ "Core fragment",
-    CLASS == "CorePreparProd" ~ "Core prep product",
-    CLASS == "RetouchedPiece" ~ "Retouched",
-    CLASS == "RetouchedPieceFrag" ~ "Retouched fragment",
-    TRUE ~ CLASS))
-
-general_table_solutrean <- solutrean_dataset |> 
-   dplyr::mutate(RMTYPE = case_when(
-    RMTYPE == "T1" | RMTYPE == "T2" | RMTYPE == "T3" | RMTYPE == "T4" | RMTYPE == "T5" ~ "Local",
-    TRUE ~ RMTYPE)) |> 
-  dplyr::mutate(CLASS = case_when(
-    CLASS == "ElongBlank" | CLASS == "Flake" ~ "Blank",
-    CLASS == "ElongBlankFrag" | CLASS == "FlakeFrag" ~ "Blank fragment",
-    CLASS == "Thinning flake" ~ "Blank",
-    CLASS == "CorePrepProd" ~ "Core prep product",
-    CLASS == "RetouchedPiece" | CLASS == "Retouched piece" ~ "Retouched",
-    CLASS == "Retouched piece fragment" ~ "Retouched fragment",
-    TRUE ~ CLASS))
-
-## For Gravettian
-gravettian_general <- general_table |>
-  filter(Level %in% c(6, 7)) |>
-  filter(RMTYPE != "INDET") |> 
-  mutate(RMTYPE = factor(RMTYPE, levels = c("Local", "T6","T7", 
-                                        "T8", "T9", "T10", "TL"))) |>
-  tabyl(CLASS, RMTYPE) |> 
-  adorn_totals(c("row", "col")) |> 
-  adorn_percentages("col") |> 
-  adorn_pct_formatting(2) |> 
-  adorn_ns("front") |> 
-  mutate(across(where(is.character), ~ gsub("^0   \\(0\\.00%\\)$", " ", .)))
-
-## For Proto-Solutrean (4E and top of 5)
-proto_general <- general_table |> 
-  filter(Level %in% c("4E", "4E/5")) |> 
-  filter(CLASS != "Chip") |> 
-  filter(RMTYPE != "INDET") |> 
-  mutate(RMTYPE = factor(RMTYPE, levels = c("Local", "T6","T7", 
-                                        "T8", "T9", "T10", "TL"))) |> 
-  tabyl(CLASS, RMTYPE) |> 
-  adorn_totals(c("row", "col")) |> 
-  adorn_percentages("col") |> 
-  adorn_pct_formatting(2) |> 
-  adorn_ns("front") |> 
-  mutate(across(where(is.character), ~ gsub("^0   \\(0\\.00%\\)$", " ", .)))
-
-## For Solutrean
-solutrean_general <- general_table_solutrean |> 
-  filter(RMTYPE != "INDET") |> 
-  mutate(RMTYPE = factor(RMTYPE, levels = c("Local", "T6","T7", 
-                                        "T8", "T9", "T10", "T11", "TL"))) |> 
-  tabyl(CLASS, RMTYPE) |> 
-  adorn_totals(c("row", "col")) |> 
-  adorn_percentages("col") |> 
-  adorn_pct_formatting(2) |> 
-  adorn_ns("front") |> 
-  mutate(across(where(is.character), ~ gsub("^0   \\(0\\.00%\\)$", " ", .)))
-
-```
-
-```{r}
 #| label: general-graphs
@@
   geom_bar(stat = "identity", width = 0.5) +
-  geom_text(aes(label=ifelse(n > 3, n, "")), position = position_stack(vjust = .5), size = 3, color = "white") +
+  geom_text(aes(label=ifelse(n > 7, n, "")), position = position_stack(vjust = .5), size = 3, color = "white") +
   scale_y_continuous(labels = scales::percent_format()) +
@@
    mutate(CLASS = case_when(
-    CLASS == "ElongBlank" ~ "Blank",
-    CLASS == "Flake" ~ "Blank",
+    CLASS == "ElongBlank" | CLASS == "Flake" ~ "Blank",
     CLASS == "FlakeFrag" | CLASS == "ElongBlankFrag" ~ "BlankFrag",
@@
 
-Results are structured according to the predefined analytical parameters, comparing local and non-local cherts (\@fig-cherts) across different assemblages.
-A comprehensive description of each assemblage and detailed tables for all analyses are provided in Online Resource1.
-
-Blanks and blank fragments represent the most abundant lithic class in all assemblages, accounting for over 40% of artifacts across all chert types (\@fig-general-graphs).
+Results are structured according to the predefined analytical parameters, comparing local and non-local cherts [@fig-cherts] across different assemblages.
+A comprehensive description of each assemblage and detailed tables for all analyses are provided in Online Resource 1.
+
+Blanks and blank fragments represent the most abundant lithic class in all assemblages, accounting for over 40% of artifacts across all chert types [@fig-general-graphs].
 As expected, cores are present in lower frequencies (\~3-8%) but are more numerous in local cherts than in non-local types and absent from the TL group.
@@
 Albeit in very small numbers in all assemblages (n\<10), core preparation and maintenance products are also present in local and non-local cherts, especially in the non-local types which show the larger sample number such as T6-7.
-The presence of knapping by-products such as shatter is also seen in both local and non-local cherts in the Gravettian (\@fig-general-graphs a)  and Proto-Solutrean (\@fig-general-graphs b) assemblages.
+The presence of knapping by-products such as shatter is also seen in both local and non-local cherts in the Gravettian [@fig-general-graphs a] and Proto-Solutrean [@fig-general-graphs b] assemblages.
 Retouched tools are present in all chert types across all assemblages.
@@
 #| label: fig-general-graphs
-#| fig-cap: Class by chert type in the Gravettian (a), Proto-Solutrean (b) and Solutrean (c) assemblages.
+#| fig-cap: "Class by chert type in the Gravettian (a), Proto-Solutrean (b) and Solutrean (c) assemblages.
 Labels within the plots correspond to the number of artefacts within each class.
-Only labels of n>3 are shown
+Only labels of n>7 are shown"
+#| fig-width: 5
+#| fig-asp: 1.3
+#| fig-align: center
 
@@
 
-The Gravettian assemblage exhibits the highest proportion of retouched tools (\~21%; @fig-general-graphs **a**), compared to \<10% in other assemblages.
+The Gravettian assemblage exhibits the highest proportion of retouched tools [\~21%; @fig-general-graphs a], compared to \<10% in other assemblages.
 Tool diversity is greater among non-local cherts, which contain a high frequency of generalized tools (e.g., notches, retouched pieces) alongside hunting implements such as backed bladelets.
@@
 
-Retouched tools were further explored through the tools-to-debitage ratio (\@fig-ratios a and Online Resource1 Table s8).
+Retouched tools were further explored through the tools-to-debitage ratio [@fig-ratios a, Online Resource 1 Table s8].
 In general, the results show that non-local chert types have higher retouch-to-debitage ratios than local types, albeit with differences between assemblages.
@@
 
-Blank-to-core ratios were analysed across assemblages (\@fig-ratios b, Online Resource1 Table s7).
+Blank-to-core ratios were analysed across assemblages [@fig-ratios b, Online Resource 1 Table s7].
 In all cases, local cherts generally exhibit lower blank-to-core ratios than non-local cherts, though variability in core sample sizes may influence these patterns.
@@
 The Proto-Solutrean assemblage has most blank-to-core ratios below 10, except for non-local T7.
-However, non-local cores are poorly represented within the assemblage (\@tbl-ratios b; Online Resource 1 Table s3).
+However, non-local cores are poorly represented within the assemblage [@fig-ratios b, Online Resource 1 Table s3].
 
@@
 #| label: fig-ratios
-#| fig-cap: Plotted ratios of blanks-to-core (a) and tool-to-debitage (b) by assemblage (Gravettian, Proto-Solutrean and Solutrean) and by chert type.
-All local cherts were grouped in a single category "Local", while non-local cherts maintained their individual types
+#| fig-cap: "Plotted ratios of blanks-to-core (a) and tool-to-debitage (b) by assemblage (Gravettian, Proto-Solutrean and Solutrean) and by chert type.
+All local cherts were grouped in a single category Local, while non-local cherts maintained their individual types"
+#| fig-width: 5
+#| fig-asp: 1.2
+#| fig-align: center
 
@@
 
-The sizes of blanks and cores were explored through scatterplots of width and length (\@fig-gravettian-plots**, @fig-proto-plots**, \@fig-solutrean-plots a) and boxplots for weight (\@fig-gravettian-plots, @fig-proto-plots, \@fig-solutrean-plots b) and through non-parametric statistical tests; complete tables with descriptive statistics and non-parametric statistical test results can be found in Online Resource1 Tables s9-11 and s12, respectively.
+The sizes of blanks and cores were explored through scatterplots of width and length [@fig-gravettian-plots; @fig-proto-plots; @fig-solutrean-plots a] and boxplots for weight [@fig-gravettian-plots; @fig-proto-plots; @fig-solutrean-plots b] and through non-parametric statistical tests; complete tables with descriptive statistics and non-parametric statistical test results can be found in Online Resource 1 Tables s9-11 and s12, respectively.
 
 The scatterplots and boxplots show different patterns in the several analysed classes.
-In the Gravettian assemblage (@fig-gravettian-plots), non-local chert blanks  are significantly smaller than local chert blanks in both length (\<25 mm) and weight (\<5 g), with p-values of 0.00009\*\* and 0.012, respectively.
+In the Gravettian assemblage [@fig-gravettian-plots], non-local chert blanks are significantly smaller than local chert blanks in both length (\<25 mm) and weight (\<5 g), with p-values of 0.00009 and 0.012, respectively.
 While local chert blanks also cluster below 25 mm in width and length, they display greater variability and a higher number of outliers.
-The same patterns can be seen in the Proto-Solutrean assemblage (@fig-proto-plots) , where length and weight also show statistically significant differences between chert types (0.046\* and 0.004), with non-local types showing smaller values.
-In comparison, in the Solutrean assemblage (@fig-solutrean-plots), only non-local chert blank length seems to show statistically significant smaller values (p value of 0.0005\*\*) than local chert blanks, being mostly concentrated below \~30 mm, while local chert blank lengths are more variable.
-While width seems to be more variable in both local and non-local chert blanks, the statistical analysis shows that non-local chert width is significantly higher than local chert blanks (p value of 0.02\*).
+The same patterns can be seen in the Proto-Solutrean assemblage [@fig-proto-plots], where length and weight also show statistically significant differences between chert types (0.046 and 0.004), with non-local types showing smaller values.
+In comparison, in the Solutrean assemblage [@fig-solutrean-plots], only non-local chert blank length seems to show statistically significant smaller values (p value of 0.0005) than local chert blanks, being mostly concentrated below \~30 mm, while local chert blank lengths are more variable.
+While width seems to be more variable in both local and non-local chert blanks, the statistical analysis shows that non-local chert width is significantly higher than local chert blanks (p value of 0.02).
 
@@
 b) Boxplots of weight; diamond symbol represents mean values within each group"
+#| fig-width: 6
+#| fig-asp: 1
+#| fig-align: center
 
@@
 For cores, size and weight differences between local and non-local cherts are less pronounced than those observed in blanks.
-In the Gravettian assemblage (@fig-gravettian-plots), there is a cluster of non-local chert cores below the 25 mm width and length, while in general, local chert cores seem to be more variable.
+In the Gravettian assemblage [@fig-gravettian-plots], there is a cluster of non-local chert cores below the 25 mm width and length, while in general, local chert cores seem to be more variable.
 Especially regarding weight, T6 cores (non-local) show lighter cores (\~12 gr means) compared to local cherts.
-In the Solutrean assemblage (@fig-solutrean-plots), weight is the only measurement where differences can be seen between local and non-local chert cores; similarly to the Gravettian assemblage, T6 cores show slightly smaller cores (\~18 gr means) than local cherts (\~22 gr means).
+In the Solutrean assemblage [@fig-solutrean-plots], weight is the only measurement where differences can be seen between local and non-local chert cores; similarly to the Gravettian assemblage, T6 cores show slightly smaller cores (\~18 gr means) than local cherts (\~22 gr means).
 Despite these differences, the statistical tests applied to the dimensions between local and non-local chert cores show no statistically significant differences between the groups (p \> 0.05).
-The Proto-Solutrean assemblage (@fig-proto-plots) contains a limited sample of non-local chert cores (n=4), preventing clear pattern identification.
+The Proto-Solutrean assemblage [@fig-proto-plots] contains a limited sample of non-local chert cores (n=4), preventing clear pattern identification.
 However, as in other assemblages, local chert cores display a broad size range (\~15-50 mm in width/length, \~3-85 g in weight).
@@
 Our findings confirm initial expectations regarding mobility patterns, raw material use, and technological strategies in southwestern Iberia during the UP.
-While core reduction differences between local and non-local cherts were anticipated (\@tbl-expectations, expectations a-b), these differences were largely shaped by transportation strategies used to bring non-local cherts to Vale Boi.
+While core reduction differences between local and non-local cherts were anticipated [@tbl-expectations a-b], these differences were largely shaped by transportation strategies used to bring non-local cherts to Vale Boi.
 As such, the occurrence of non-local chert cores at Vale Boi depended on whether these materials arrived as finished tools or blanks—likely reflecting an individual provisioning strategy [@kuhn2004]—or as trimmed cores and larger stone volumes, aligning with site provisioning models [@kelly1988; @roth1998; @clark1987; @kuhn2004] .
@@
 Non-local chert cores appear in all studied assemblages, though in limited numbers (e.g., in the Proto-Solutrean).
-While chips—often proxies for on-site knapping—were not analysed, the presence of knapping by-products, such as shatter (except in the Solutrean assemblage; see **Section 2.2.
-Datasets and analysis**), and core preparation and maintenance products in non-local cherts suggests active core reduction and on-site knapping.
+While chips—often proxies for on-site knapping—were not analysed, the presence of knapping by-products, such as shatter (except in the Solutrean assemblage; see Section Datasets and analysis), and core preparation and maintenance products in non-local cherts suggests active core reduction and on-site knapping.
 
 Our results indicate that most non-local cherts were introduced to Vale Boi as cores or nodules, subsequently knapped on-site.
-However, an exception is seen in TL cherts (TL2-6, TL9-10, TL14; Online Resource1 Table S13), which lack associated cores and knapping debris.
+However, an exception is seen in TL cherts (TL2-6, TL9-10, TL14; Online Resource 1 Table S13), which lack associated cores and knapping debris.
 This absence suggests that TL cherts were introduced to the site primarily as finished tools or blanks, aligning with long-distance individual provisioning models [@kuhn2004].
@@
 However, our results show that these differences are not limited to long-term occupations and low mobility.
-High blank-to-core ratios and smaller blank sizes in non-local cherts are also observed in the Gravettian occupation, previously interpreted as a short-term occupation with frequent moves [@cascalheira2017a; @belmiro2025].
+High blank-to-core ratios and smaller blank sizes in non-local cherts are also observed in the Gravettian occupation, previously interpreted as a short-term occupation with frequent moves [@cascalheira2017; @belmiro2025].
 Due to this, we may interpret that non-local raw materials were generally more intensively reduced than local cherts throughout all the UP, independent of the duration of the occupation.
@@
 
-While the high intensity of retouch in non-local cherts aligns with expectations (\@tbl-expectations, expectations 1c and 2c), their unexpectedly high tool diversity contradicts common mobility models which expect mobility to be linked with limited tool variability and an increase of easily reconfigurable tools [@odell1981; @shott1986; @siegel1984].
+While the high intensity of retouch in non-local cherts aligns with expectations [@tbl-expectations 1c and 2c], their unexpectedly high tool diversity contradicts common mobility models which expect mobility to be linked with limited tool variability and an increase of easily reconfigurable tools [@odell1981; @shott1986; @siegel1984].
 Non-local cherts at Vale Boi contain a diverse set of tool types, although less specialized tools, such as retouched blanks, remain predominant in non-local cherts across all assemblages.
@@
 
-This study was supported by
+This study was supported by the Fundacão para a Ciência e Tecnologia (https://www.fct.pt/), attributed to Joana Belmiro with the contract reference 2020.08722.BD (DOI: 10.54499/2020.08722.BD).
+This study was also supported by the Seed Fund’21 attributed to Joana Belmiro, by ICArEHB – The interdisciplinary Center for Archaeology and Evolution of Human Behaviour, funded by the Fundacão para a Ciência e Tecnologia (https://www.fct.pt/) under program UIDP/04211/2020.
+João Cascalheira is funded by the European Union (ERC-CoG-2021, FINISTERRA, 101045506).
+Views and opinions expressed are however those of the author(s) only and do not necessarily reflect those of the European Union or the European Research Council Executive Agency.
+Neither the European Union nor the granting authority can be held responsible for them.
+Open Access was funded by Fundacão para a Ciência e Tecnologia (https://www.fct.pt/) and Universidade do Algarve.
 
@@
 
-All datasets and code used to write this article and supplementary materials, including analyses and visual representations, can be accessed and downloaded in an online repository: <http://www.doi.org/10.17605/OSF.IO/UWMPK>.
+The authors confirm that all data generated and analysed during this study are available to the public in the Open Science Framework repository at: <http://www.doi.org/10.17605/OSF.IO/UWMPK>.
 

### 2025-04-15 analysis/paper/paper.qmd
Final commit before submission 03. Project is up to date with final documents.

@@
 abstract: |
-  The Upper Palaeolithic of Westernmost Europe was marked by technological and cultural transformations shaped by Late Pleistocene climatic shifts.
-The Iberian Peninsula, particularly southwestern Iberia, served as a key eco-cultural niche, sustaining hunter-gatherer populations through environmental fluctuations.
+  The Upper Palaeolithic (UP) of westernmost Europe was marked by technological and cultural transformations and abrupt climatic shifts.
+The Iberian Peninsula, particularly southwestern Iberia, served as a refugia and key eco-cultural niche, making it a key region for UP studies.
 Vale Boi, the only site in southwestern Iberia with a nearly complete UP sequence, provides critical insights into technological, economical and mobility patterns over time.
-This study examines lithic raw material procurement and technological organization at Vale Boi, focusing on the role of non-local cherts in mobility strategies and social interactions.
-Using a combined techno-typological and geoarchaeological approach, we analyzed chert assemblages from Gravettian, Proto-Solutrean, and Solutrean levels, integrating previously published macroscopic and petrographic data.
-Our results indicate that non-local cherts were intensively reduced, suggesting strategic curation.
-The presence of diverse tool types challenges the assumption that non-local materials were transported solely as finished tools, instead highlighting their role as versatile raw material sources.
-Gravettian occupations exhibited greater reliance on non-local cherts, implying frequent mobility, while later phases saw increased dependence on local materials, aligning with more long-term occupations.These findings refine our understanding of UP technological adaptations in southwestern Iberia, emphasizing the evolving role of lithic resources in mobility, settlement, and social networks.
-By integrating lithic technology and raw material sourcing, this study contributes to broader discussions on eco-cultural niches and UP hunter-gatherer adaptations.
+This study examines lithic raw material procurement and technological organisation at Vale Boi, focusing on the differences between local and non-local cherts.
+We analysed chert assemblages from Gravettian, Proto-Solutrean, and Solutrean levels, integrating previously published techno-typological and raw material macroscopic and petrographic data.
+Our results indicate that non-local cherts were intensively reduced.
+The presence of varied techno-typological classes suggests that non-local cherts were not exclusively transported as finished tools, instead highlighting their role as versatile, transportable raw material volumes.
+Gravettian occupations exhibited greater reliance on non-local cherts, suggesting short-term occupations and higher mobility, while Proto-Solutrean and Solutrean occupations showed increased dependence on local materials, suggesting long-term occupations.
+These findings expand our understanding of UP technological adaptations in southwestern Iberia, emphasizing the complex role of lithic resources in mobility, settlement, and social networks.
 
@@
 
-In westernmost Europe, five main technocomplexes have been identified within the UP - Aurignacian, Gravettian, Proto-Solutrean, Solutrean, and Magdalenian - each contributing new evidence on resilient behaviors and cultural continuity.
+In westernmost Europe, five main technocomplexes have been identified within the UP - Aurignacian, Gravettian, Proto-Solutrean, Solutrean, and Magdalenian - each contributing new evidence on resilient behaviours and cultural continuity.
 These technocomplexes are distinguished by regional adaptations as well as technological and subsistence innovations [@baker2024; @bicho2017; @bicho2011; @bicho2008; @cascalheira2017; @straus1991; @straus1993; @zilhão1997].
@@
 These innovations are believed to have resulted from human adaptations and resilience to the shifting conditions of the Late Pleistocene, which was marked by abrupt climatic events such as the Heinrich events and the Last Glacial Maximum (LGM) [@sanchezgoñi2010].
-These climatic shifts had a profound impact on flora and fauna, with exception of spatially limited refugia, likely disrupting settlement patterns and the organization of hunter-gatherer groups, ultimately leading to a process of cultural reorganization [@bradtmöller2012; @morin2008; @schmidt2012], which has frequently been a focus of UP research.
+These climatic shifts had a profound impact on flora and fauna, with exception of spatially limited refugia, likely disrupting settlement patterns and the organisation of hunter-gatherer groups, ultimately leading to a process of cultural reorganisation [@bradtmöller2012; @morin2008; @schmidt2012], which has frequently been a focus of UP research.
 
@@
 
-In this context, comparisons between Vale Boi and the Gravettian occupations in central Portugal and southern Spain, revealed marked regional differences, suggesting more limited movement of people and information in southwestern Iberia.
+In this context, comparisons between Vale Boi and the Gravettian occupations in central Portugal and southern Spain revealed marked regional differences, suggesting more limited movement of people and information in southwestern Iberia.
 These differences were suggested to be related to Vale Boi’s peripheral location and the distinct ecological characteristics of westernmost Algarve [@marreiros2013].
@@
 
-To address this gap, this paper presents a combined analysis of lithic technology and raw material sourcing, aiming to refine our understanding of technological and social adaptations throughout the UP in southwestern Iberia.
+This paper presents a combined analysis of lithic technology and raw material sourcing, aiming to refine our understanding of technological and social adaptations throughout the UP in southwestern Iberia.
 By integrating archaeological and ethnographic perspectives, we test expectations regarding the organisation of hunter-gatherer technological systems and their implications for mobility, resource management, and social interactions.
@@
 
-Lithic technological organization has long been a central focus in archaeology, offering critical insights into past human behaviour and adaptive strategies.
+Lithic technological organisation has long been a central focus in archaeology, offering critical insights into past human behaviour and adaptive strategies.
 It encompasses all stages of stone tool production and use, from raw material acquisition to discard.
 Nelson [-@nelson1991] defines it as the selection and integration of strategies for tool production, use, transport, and maintenance, while Kelly [-@kelly1988] emphasizes its spatial and temporal dimensions within cultural systems.
-These perspectives highlight the interrelationship between raw material sourcing, mobility, and settlement organization—key factors in understanding hunter-gatherer adaptations.
+These perspectives highlight the interrelationship between raw material sourcing, mobility, and settlement organisation—key factors in understanding hunter-gatherer adaptations.
 
@@
 
-Vale Boi is located on the western coast of the Algarve region, southern Portugal, at a latitude of N 37º5'24.1" and longitude of W 8º48'32.4" (@gibajaFUNCIONINSTRUMENTOSLITICOS2006; the original site coordinates were converted from Degree/Minute to Degree/Minute/Second), within a small valley extending \~2 km southward to the Atlantic Ocean [@fig-map a].
-The site is nestled among limestone outcrops that create west- and southwest-facing rock shelters [@fig-map b].
-Covering more than 10,000 square meters along the valley slope, Vale Boi includes three main excavated areas, Slope, Terrace, and Shelter [@fig-map c], which were excavated between 2000 and 2019 [@bicho2007; @bicho2012; @bicho2013; @cascalheira2017; @cascalheira2013; @manne2012].
+Vale Boi is located on the western coast of the Algarve region, southern Portugal, at a latitude of N 37º5'24.1" and longitude of W 8º48'32.4" [@gibajaFUNCIONINSTRUMENTOSLITICOS2006, the original site coordinates were converted from Degree/Minute to Degree/Minute/Second], within a small valley extending \~2 km southward to the Atlantic Ocean ([@fig-map] a).
+The site is nestled among limestone outcrops that create west- and southwest-facing rock shelters ([@fig-map] b).
+Covering more than 10,000 square meters along the valley slope, Vale Boi includes three main excavated areas, Slope, Terrace, and Shelter ([@fig-map] c), which were excavated between 2000 and 2019 [@bicho2007; @bicho2012; @bicho2013; @cascalheira2017; @cascalheira2013; @manne2012].
 Based on the available raw material and technological evidence, particularly the macroscopic and petrographic data [@belmiro2025], the Terrace and Shelter areas were selected for the current study.
@@
 Technological analyses of these assemblages have provided insights into the adaptive strategies of UP hunter-gatherers at Vale Boi, revealing technological shifts across different cultural phases.
-Given the chosen sample for this study (see Section 2.2.
-Datasets and analysis), we summarise here the results for the technological analysis of the Solutrean assemblages (levels A-C) of the Shelter [@cascalheira2010; @cascalheira2013a] and the Proto-Solutrean assemblages (levels 4E-5) of the Terrace [@belmiro2020; @belmiro2021].
+Given the chosen sample for this study (see Section Datasets and analysis), we summarise here the results for the technological analysis of the Solutrean assemblages (levels A-C) of the Shelter [@cascalheira2010; @cascalheira2013a] and the Proto-Solutrean assemblages (levels 4E-5) of the Terrace [@belmiro2020; @belmiro2021].
 The Gravettian occupations (levels 6-8) of the Terrace have been recently studied, but the results remain unpublished.
@@
 Notably, the lower Proto-Solutrean levels show high frequencies of quartz [@belmiro2021], whereas chert becomes the dominant raw material in the Solutrean occupations of the Shelter [@cascalheira2010].
-Likewise, different raw materials appear to have been used for specific purposes,with chert being particularly associated with the production of high percentage of blanks and retouched tools, and greywacke being extensively used for anvils [@horta2019].
-
-Quartz and greywacke are locally available from streambeds near the site, whereas regional chert and chalcedony sources lie further afield, within a \~20 km radius.
-Our previous studies showed, as expected, the existence of local cherts in the assemblages [@fig-cherts a], originating from the Lower Jurassic (cherts) and Triassic (chalcedony) outcrops in proximity to the site [@belmiro2025].
+Likewise, different raw materials appear to have been used for specific purposes, with chert being particularly associated with the production of high percentage of blanks and retouched tools, and greywacke being extensively used for anvils [@horta2019].
+
+Quartz and greywacke are locally available from streambeds near the site, whereas regional chert and chalcedony sources lie further afield, within a \~20 km radius [@pereira2016].
+Our previous studies showed, as expected, the existence of local cherts in the assemblages ([@fig-cherts] a), originating from the Lower Jurassic (cherts) and Triassic (chalcedony) outcrops in proximity to the site [@belmiro2025].
 Despite petrographic differences between cherts and chalcedony, for this study any chert varieties (e.g., chalcedony or jasper) are referred to as chert [@luedtke1978].
-A significant portion of non-local cherts was also identified [@fig-cherts b-c], with some identified sources, such as the Cretaceous outcrops of central Portugal (Lisbon and Rio Maior areas) and the Upper Jurassic oolitic/peloidal cherts of the Betic Systems in southern Spain.
+A significant portion of non-local cherts was also identified (@fig-cherts b-c), with some identified sources, such as the Cretaceous outcrops of central Portugal (Lisbon and Rio Maior areas) and the Upper Jurassic oolitic/peloidal cherts of the Betic Systems in southern Spain.
 This indicates varying degrees of contact or long-range mobility throughout the UP, likely influenced by shifts in settlement patterns and the duration of site occupation.
@@
 Cascalheira et al.
-(2017) applied the Whole Assemblage Behavioral Index (WABI) [@riel-salvatore2004; @riel-salvatore2010] to reconstruct and compare mobility strategies at Vale Boi.
+(2017) applied the Whole Assemblage Behavioural Index (WABI) [@riel-salvatore2004; @riel-salvatore2010] to reconstruct and compare mobility strategies at Vale Boi.
 By analysing lithic volumetric density and retouched tool frequencies from the Shelter’s Solutrean [@cascalheira2010; @cascalheira2013a] and Magdalenian [@mendonça2009] levels, as well as the Terrace’s Gravettian levels [@marreiros2015], the authors confirmed that most occupations corresponded to long-term residential base camps, composed of high-density assemblages with low degree of retouch, except for the Early Gravettian and Magdalenian occupations.
-Regarding the Terrace area, however, the results were based on previous excavations (\<2012), and the identification of new layers and reorganization of the occupations is yet to be tested through the use of lithic techno-typological methods.
+Regarding the Terrace area, however, the results were based on previous excavations (\<2012), and the identification of new layers and reorganisation of the occupations is yet to be tested through the use of lithic techno-typological methods.
 
@@
 One challenge of integrating datasets collected by different researchers is variability in methodologies and recorded variables.
-However, this study minimizes such inconsistencies by ensuring methodological alignment across datasets, as all data collection followed standardized attribute-based lithic analysis protocols.
+However, this study minimizes such inconsistencies by ensuring methodological alignment across datasets, as all data collection followed standardised attribute-based lithic analysis protocols.
 The lithic analysis datasets used in this study adopted an attribute-based methodology, which aims to describe morphological and metrical attributes of technological classes.
-The attributes analysed followed those present in specialized literature, such as Brézillon [-@brézillon1968], and [@tixier1980], paired with other Palaeolithic lithic attribute analyses [e.g., @bicho1992; @zilhão1997; @cascalheira2019; @tostevin2012].
+The attributes analysed followed those present in specialised literature, such as Brézillon [-@brézillon1968], and [@tixier1980], paired with other Palaeolithic lithic attribute analyses [e.g., @bicho1992; @zilhão1997; @cascalheira2019; @tostevin2012].
 
 After data collection, the databases were imported into R Programming Environment, where the information was processed through the creation of descriptive statistical analysis and writing of this article.
-Some differences in the collection process, especially between the study by Cascalheira [-@cascalheira2010] and the following studies, required homogenization.
+Some differences in the collection process, especially between the study by Cascalheira [-@cascalheira2010] and the following studies, required homogenisation.
 For example, where in Cascalheira (2010) retouch was an independent variable (e.g., Class as flake, Retouch as Yes), in Belmiro et al.
@@
 To integrate technological and raw material data, we merged multiple datasets using the artifact identifier (“ID”), which corresponds to total station plot records.
-This identifier links artifacts to either precise excavation coordinates (for individually plotted pieces) or generalized spatial data (for sieved assemblages).
+This identifier links artifacts to either precise excavation coordinates (for individually plotted pieces) or generalised spatial data (for sieved assemblages).
 
 Given that the raw material analysis focused exclusively on individually piece-plotted artifacts (dimensions \>2 cm), the final dataset does not include all excavated material but retains the most diagnostically relevant pieces for analysis.
-As a result, the final dataset used in this study does not directly correspond to the full datasets analysed in the original studies [@tbl-samples].
+As a result, the final dataset used in this study does not directly correspond to the full datasets analysed in the original studies ([@tbl-samples]).
 Similarly, the raw material study used only currently accessible artifacts from the assemblages.
@@
 
-Following data homogenization and integration, we conducted descriptive statistical analyses and non-parametric statistical analyses to examine correlations between raw material use and technological patterns and test expectations related to the technological organisation of local and non-local cherts [see @tbl-expectations], and between different UP occupations.
+Following data homogenization and integration, we conducted descriptive statistical analyses and non-parametric statistical analyses to examine correlations between raw material use and technological patterns and test expectations related to the technological organisation of local and non-local cherts (see [@tbl-expectations]), and between different UP occupations.
 To test these expectations, we evaluated 9 parameters for each dataset and assemblage; the parameters and their calculation/measurement methods are detailed in @tbl-parameters.
@@
 The provided files include the complete set of raw data used in the analysis, along with a custom R project [@wickham2015] containing the code required to generate all tables and figures.
-To enable maximum reuse, the code is made available under the MIT license, data under CC‐0, and figures under CC‐BY (additional details can be found in @Marwick2017).
+To enable maximum reuse, the code is made available under the MIT license, data under CC‐0, and figures under CC‐BY [additional details can be found in @Marwick2017].
 
@@
 
-Results are structured according to the predefined analytical parameters, comparing local and non-local cherts [@fig-cherts] across different assemblages.
+Results are structured according to the predefined analytical parameters, comparing local and non-local cherts ([@fig-cherts]) across different assemblages.
 A comprehensive description of each assemblage and detailed tables for all analyses are provided in Online Resource 1.
 
-Blanks and blank fragments represent the most abundant lithic class in all assemblages, accounting for over 40% of artifacts across all chert types [@fig-general-graphs].
+Blanks and blank fragments represent the most abundant lithic class in all assemblages, accounting for over 40% of artifacts across all chert types ([@fig-general-graphs]).
 As expected, cores are present in lower frequencies (\~3-8%) but are more numerous in local cherts than in non-local types and absent from the TL group.
@@
 Albeit in very small numbers in all assemblages (n\<10), core preparation and maintenance products are also present in local and non-local cherts, especially in the non-local types which show the larger sample number such as T6-7.
-The presence of knapping by-products such as shatter is also seen in both local and non-local cherts in the Gravettian [@fig-general-graphs a] and Proto-Solutrean [@fig-general-graphs b] assemblages.
+The presence of knapping by-products such as shatter is also seen in both local and non-local cherts in the Gravettian (@fig-general-graphs a) and Proto-Solutrean (@fig-general-graphs b) assemblages.
 Retouched tools are present in all chert types across all assemblages.
@@
 
-The Gravettian assemblage exhibits the highest proportion of retouched tools [\~21%; @fig-general-graphs a], compared to \<10% in other assemblages.
+The Gravettian assemblage exhibits the highest proportion of retouched tools (\~21%; [@fig-general-graphs] a), compared to \<10% in other assemblages.
 Tool diversity is greater among non-local cherts, which contain a high frequency of generalized tools (e.g., notches, retouched pieces) alongside hunting implements such as backed bladelets.
-The abundance of burins and burin spalls suggests they may have functioned as cores, indicating a technological strategy rather than solely a focus on retouch (Online Resource1 Table s2).
+The abundance of burins and burin spalls suggests they may have functioned as cores, indicating a technological strategy rather than solely a focus on retouch (Online Resource 1 Tables s1 and s3).
 In comparison to the Gravettian, other assemblages show smaller numbers of retouched tools.
-While in the Proto-Solutrean (Online Resource 1 Table s4) no significant difference was identified between local and non-local retouched tools, in the Solutrean assemblage (Online Resource 1 Table s6) both tool diversity and tool specialization are higher in local cherts.
-
-Retouched tools were further explored through the tools-to-debitage ratio [@fig-ratios a, Online Resource 1 Table s8].
-In general, the results show that non-local chert types have higher retouch-to-debitage ratios than local types, albeit with differences between assemblages.
+While in the Proto-Solutrean (Online Resource 1 Table s5) no significant difference was identified between local and non-local retouched tools, in the Solutrean assemblage (Online Resource 1 Table s7) both tool diversity and tool specialization are higher in local cherts.
+
+Retouched tools were further explored through the tools-to-debitage ratio (@fig-ratios a, Online Resource 1 Table s8).
+In general, the results show that non-local chert types have higher tool-to-debitage ratios than local types, albeit with differences between assemblages.
 The Gravettian assemblage shows the highest tool-to-debitage ratios compared to the remaining assemblages; non-local cherts show ratios above 0.9, while local cherts show values below 0.5.
@@
 Non-local cherts show ratios above 0.1 (with exception of T8) while local cherts show ratios below 0.1 (with exception of T1; Online Resource1 Table s8 c).
-
-Blank-to-core ratios were analysed across assemblages [@fig-ratios b, Online Resource 1 Table s7].
-In all cases, local cherts generally exhibit lower blank-to-core ratios than non-local cherts, though variability in core sample sizes may influence these patterns.
-For the Gravettian assemblages, the results reveal that non-local cherts show ratios above 7 (with the exception of T10), while local cherts have ratios located below 4.
-The Solutrean assemblage also shows higher ratios for non-local cherts (\>12), with local cherts showing ratios below 7.
-The Proto-Solutrean assemblage has most blank-to-core ratios below 10, except for non-local T7.
-However, non-local cores are poorly represented within the assemblage [@fig-ratios b, Online Resource 1 Table s3].
 
@@
 
-The sizes of blanks and cores were explored through scatterplots of width and length [@fig-gravettian-plots; @fig-proto-plots; @fig-solutrean-plots a] and boxplots for weight [@fig-gravettian-plots; @fig-proto-plots; @fig-solutrean-plots b] and through non-parametric statistical tests; complete tables with descriptive statistics and non-parametric statistical test results can be found in Online Resource 1 Tables s9-11 and s12, respectively.
+Blank-to-core ratios were analysed across assemblages (@fig-ratios b, Online Resource 1 Table s9).
+In all cases, local cherts generally exhibit lower blank-to-core ratios than non-local cherts, though variability in core sample sizes may influence these patterns.
+For the Gravettian assemblages, the results reveal that non-local cherts show ratios above 7 (with the exception of T10), while local cherts have ratios located below 4.
+The Solutrean assemblage also shows higher ratios for non-local cherts (\>12), with local cherts showing ratios below 7.
+The Proto-Solutrean assemblage has most blank-to-core ratios below 10, except for non-local T7.
+However, non-local cores are poorly represented within the assemblage (@fig-ratios b, Online Resource 1 Table s4).
+
+The sizes of blanks and cores were explored through scatterplots of width and length (@fig-gravettian-plots, @fig-proto-plots, @fig-solutrean-plots a) and boxplots for weight (@fig-gravettian-plots, @fig-proto-plots, @fig-solutrean-plots b) and through non-parametric statistical tests; complete tables with descriptive statistics and non-parametric statistical test results can be found in Online Resource 1 Tables s10-12 and s13, respectively.
 
 The scatterplots and boxplots show different patterns in the several analysed classes.
-In the Gravettian assemblage [@fig-gravettian-plots], non-local chert blanks are significantly smaller than local chert blanks in both length (\<25 mm) and weight (\<5 g), with p-values of 0.00009 and 0.012, respectively.
+In the Gravettian assemblage ([@fig-gravettian-plots]), non-local chert blanks are significantly smaller than local chert blanks in both length (\<25 mm) and weight (\<5 g), with p values of 0.00009 and 0.012, respectively.
 While local chert blanks also cluster below 25 mm in width and length, they display greater variability and a higher number of outliers.
-The same patterns can be seen in the Proto-Solutrean assemblage [@fig-proto-plots], where length and weight also show statistically significant differences between chert types (0.046 and 0.004), with non-local types showing smaller values.
-In comparison, in the Solutrean assemblage [@fig-solutrean-plots], only non-local chert blank length seems to show statistically significant smaller values (p value of 0.0005) than local chert blanks, being mostly concentrated below \~30 mm, while local chert blank lengths are more variable.
+The same patterns can be seen in the Proto-Solutrean assemblage ([@fig-proto-plots]), where length and weight also show statistically significant differences between chert types (0.046 and 0.004), with non-local types showing smaller values.
+In comparison, in the Solutrean assemblage ([@fig-solutrean-plots]), only non-local chert blank length seems to show statistically significant smaller values (p value of 0.0005) than local chert blanks, being mostly concentrated below \~30 mm, while local chert blank lengths are more variable.
 While width seems to be more variable in both local and non-local chert blanks, the statistical analysis shows that non-local chert width is significantly higher than local chert blanks (p value of 0.02).
@@
 For cores, size and weight differences between local and non-local cherts are less pronounced than those observed in blanks.
-In the Gravettian assemblage [@fig-gravettian-plots], there is a cluster of non-local chert cores below the 25 mm width and length, while in general, local chert cores seem to be more variable.
+In the Gravettian assemblage ([@fig-gravettian-plots]), there is a cluster of non-local chert cores below the 25 mm width and length, while in general, local chert cores seem to be more variable.
 Especially regarding weight, T6 cores (non-local) show lighter cores (\~12 gr means) compared to local cherts.
-In the Solutrean assemblage [@fig-solutrean-plots], weight is the only measurement where differences can be seen between local and non-local chert cores; similarly to the Gravettian assemblage, T6 cores show slightly smaller cores (\~18 gr means) than local cherts (\~22 gr means).
+In the Solutrean assemblage ([@fig-solutrean-plots]), weight is the only measurement where differences can be seen between local and non-local chert cores; similarly to the Gravettian assemblage, T6 cores show slightly smaller cores (\~18 gr means) than local cherts (\~22 gr means).
 Despite these differences, the statistical tests applied to the dimensions between local and non-local chert cores show no statistically significant differences between the groups (p \> 0.05).
-The Proto-Solutrean assemblage [@fig-proto-plots] contains a limited sample of non-local chert cores (n=4), preventing clear pattern identification.
+The Proto-Solutrean assemblage ([@fig-proto-plots]) contains a limited sample of non-local chert cores (n=4), preventing clear pattern identification.
 However, as in other assemblages, local chert cores display a broad size range (\~15-50 mm in width/length, \~3-85 g in weight).
@@
 Similarly to the Proto-Solutrean, tools-to-debitage ratios (P5) and blank-to-core ratios (P7) are slightly higher in non-local cherts.
-Both tool diversity (P8) and tool specialization (P9) are higher in local cherts.
+Both tool diversity (P8) and tool specialisation (P9) are higher in local cherts.
 
@@
 Our findings confirm initial expectations regarding mobility patterns, raw material use, and technological strategies in southwestern Iberia during the UP.
-While core reduction differences between local and non-local cherts were anticipated [@tbl-expectations a-b], these differences were largely shaped by transportation strategies used to bring non-local cherts to Vale Boi.
-As such, the occurrence of non-local chert cores at Vale Boi depended on whether these materials arrived as finished tools or blanks—likely reflecting an individual provisioning strategy [@kuhn2004]—or as trimmed cores and larger stone volumes, aligning with site provisioning models [@kelly1988; @roth1998; @clark1987; @kuhn2004] .
+While core reduction differences between local and non-local cherts were anticipated ([@tbl-expectations] a-b), these differences were largely shaped by transportation strategies used to bring non-local cherts to Vale Boi.
+As such, the occurrence of non-local chert cores at Vale Boi depended on whether these materials arrived as finished tools or blanks—likely reflecting an individual provisioning strategy [@kuhn2004]—or as trimmed cores and larger stone volumes, aligning with site provisioning models [@kelly1988; @roth1998; @clark1987; @kuhn2004].
 
@@
 Our results indicate that most non-local cherts were introduced to Vale Boi as cores or nodules, subsequently knapped on-site.
-However, an exception is seen in TL cherts (TL2-6, TL9-10, TL14; Online Resource 1 Table S13), which lack associated cores and knapping debris.
+However, an exception is seen in TL cherts (TL2-6, TL9-10, TL14; Online Resource 1 Table S2), which lack associated cores and knapping debris.
 This absence suggests that TL cherts were introduced to the site primarily as finished tools or blanks, aligning with long-distance individual provisioning models [@kuhn2004].
@@
 These results align with previous studies, which propose that the intensive reduction of cores—manifested in smaller core sizes, smaller debitage products, and a higher amount of debitage per core—may be linked to lower mobility.
-In this context, lower mobility results in reduced access to non-local raw materials, which are therefore more intensively reduced and eventually replaced by local raw materials [@kuhn2004; @roth1998; @grove2023; @surovell2009].
+In this context, lower mobility results in decreased access to non-local raw materials, which are therefore more intensively reduced and eventually replaced by local raw materials [@kuhn2004; @roth1998; @grove2023; @surovell2009].
 However, our results show that these differences are not limited to long-term occupations and low mobility.
@@
 [-@belmiro2023] highlight that chert nodules in the Algarve region vary between \~5-10 cm, with a maximum diameter of 20 cm.
-Although abundant and easily accessible in exposed cliffs, local chert nodules are generally small and prone to fracturing when extracted, and normally occur fractured in secondary deposits associated to primary outcrops.
+Although abundant and easily accessible in exposed cliffs, local chert nodules are generally small and prone to fracturing when extracted and normally occur fractured in secondary deposits associated to primary outcrops.
 As a result, the small size of local chert cores may not necessarily indicate intensive reduction but could instead reflect the initial size of available raw material volumes.
@@
 
-While the high intensity of retouch in non-local cherts aligns with expectations [@tbl-expectations 1c and 2c], their unexpectedly high tool diversity contradicts common mobility models which expect mobility to be linked with limited tool variability and an increase of easily reconfigurable tools [@odell1981; @shott1986; @siegel1984].
+While the high intensity of retouch in non-local cherts aligns with expectations ([@tbl-expectations] 1c and 2c), their unexpectedly high tool diversity contradicts common mobility models which expect mobility to be linked with limited tool variability and an increase of easily reconfigurable tools [@odell1981; @shott1986; @siegel1984].
 Non-local cherts at Vale Boi contain a diverse set of tool types, although less specialized tools, such as retouched blanks, remain predominant in non-local cherts across all assemblages.
@@
 Regarding Vale Boi, quality does not appear to be a defining factor in raw material choice, as typologies typically associated with high-quality lithic resources are also present in other materials.
-For example, Solutrean bifacial tools and points were produced using Lower Jurassic and Triassic local chert, which may be considered lower quality than Rio Maior Cretaceous chert due to its coarser texture and less homogeneous composition.
+For example, Solutrean bifacial tools and points were produced using Lower Jurassic and Triassic local chert, which may be considered lower quality than Rio Maior Cretaceous chert (central Portugal) due to its coarser texture and less homogeneous composition.
 This suggests that the choice of raw material was not solely dependent on quality, especially when techniques such as heat treatment—the process of heating lithic raw material to improve knapping quality [@schmidt2016]—have been identified during the Solutrean in European UP contexts [@bachellerie2022; @bordes1969; @inizan1976; @tiffagom1998].
 
-Similarly, Vale Comprido points, a fossil indicator of the Proto-Solutrean, were primarily made from other lithic resources (e.g., dolerite), with only two identified points made of chalcedony and chert [@belmiro2021].
+Similarly, Vale Comprido points, the index fossil of the Proto-Solutrean, were primarily made from other lithic resources (e.g., dolerite), with only two identified points made of chalcedony and chert [@belmiro2021].
 This further supports the idea that raw material choice was not strictly based on quality, as these points were made from a variety of materials, including those less commonly associated with formal tools at the site.
@@
 
-Thus, non-local cherts were likely not procured solely for their superior quality or due to local resource limitations.
+Thus, non-local cherts were likely not procured solely for their quality or due to local resource limitations.
 Instead, their presence reflects mobility strategies, territorial exploitation, and social networks that facilitated the circulation of long-distance raw materials within southwestern Iberian hunter-gatherer groups.
@@
 When combined with technological data, this also reveals significant differences in the frequency of retouched non-local chert tools and the tool-to-blank ratio (\~30% and \~1, respectively), compared to other technocomplexes (\~15-19% and 0.1-0.5).
+
 Several studies have shown that short-term occupations often rely heavily on non-local toolkits and raw materials [@torrence1983; @surovell2009] and are characterized by higher reduction intensity and increased tool-to-debitage ratios [@grove2023].
@@
 Alternatively, non-local raw materials were brought in as blanks, which were consistently retouched, although individual toolkits made from limited non-local raw materials—primarily blanks or tools—were also present.
-Raw material volumes were intensively exploited at the site, yielding both specialized and non-specialized tools.
+Raw material volumes were intensively exploited at the site, yielding both specialised and non-specialised tools.
 Local raw materials were also used to produce high frequencies of retouched tools, though this was done less intensively.
@@
 Over time, Vale Boi was occupied as a long-term residential site.
-At the beginning and during the Last Glacial Maximum, Proto-Solutrean and later Solutrean groups, while still having access to non-local raw materials, moved less frequently, which limited the influx of these materials.
+At the beginning and during the LGM, Proto-Solutrean and later Solutrean groups, while still having access to non-local raw materials, moved less frequently, which limited the influx of these materials.
 
@@
 
-This study demonstrates that lithic raw material procurement and technological organization at Vale Boi were shaped by a combination of raw material accessibility, mobility strategies, and social interactions:
+This study demonstrates that lithic raw material procurement and technological organisation at Vale Boi were shaped by a combination of raw material accessibility, mobility strategies, and social interactions:
 
@@
 Future work should explore whether Vale Boi’s non-local cherts were obtained through direct procurement, down-the-line exchange, or social trade networks.
-Additionally, applying use-wear analysis to retouched tools could help clarify whether non-local chert tools had specialized functions, further informing our understanding of raw material management and mobility strategies.
-
-# Supplementary Information
+Additionally, applying use-wear analysis to retouched tools could help clarify whether non-local chert tools had specialised functions, further informing our understanding of raw material management and mobility strategies.
+
+# Supplementary Information {.unnumbered}
 
@@
 
-# Statements and Declarations
-
-## Funding
+# Statements and Declarations {.unnumbered}
+
+## Funding {.unnumbered}
 
@@
 
-## Competing Interests
+## Competing Interests {.unnumbered}
 
@@
 
-## Data and code availability
+## Data and code availability {.unnumbered}
 

### 2025-04-15 analysis/supplementary-materials/ESM_1.qmd
Final commit before submission 03. Project is up to date with final documents.

--- 2025-04-15 analysis/Supplementary/ESM_1.qmd
@@
 title: "Online Resources 1: Supplementary results and tables"
-subtitle: "From Stone to Tool: How Raw Materials Influenced Upper Paleolithic Technology in Southwestern Iberia (Vale Boi). Archaeological and Anthroplogical Sciences"
+subtitle: "From Stone to Tool: How Raw Materials Influenced Upper Paleolithic Technology in Southwestern Iberia (Vale Boi). Archaeological and Anthropological Sciences"
 author:
@@
     Type == "28" ~ "Burin",
-    Type == "7" | Type == "5b" | Type == "4"  ~ "Endscrapper",
+    Type == "7" | Type == "5b" | Type == "4"  ~ "Endscraper",
     TRUE ~ Type)) |>
@@
 #| label: tbl-general-gravettian
-#| tbl-cap: Number and percentage of technotypological classes by raw material type from the Gravettian assemblage (levels 6 and 7) of the Terrace area of Vale Boi.
+#| tbl-cap: Number and percentage of techno-typological classes by cherts type from the Gravettian assemblage (levels 6 and 7) of the Terrace area of Vale Boi.
 
@@
 #| label: tbl-TL-classes
-#| tbl-cap: Trace lithologies cherts from the Gravettian and Proto-Solutrean assemblages of the Terrace and Solutrean assemblage from the Shelter and corresponding technological class.
+#| tbl-cap: Trace lithologies cherts (TL) from the Gravettian and Proto-Solutrean assemblages of the Terrace and Solutrean assemblage from the Shelter and corresponding technological class.
 
@@
 As seen in @tbl-retouch-gravettian, 170 retouched tools were identified, with 39 (`r round(39*100/170, 2)`%) having an unidentified chert type.
-From the identified chert types, `r 3+20+4+2+9` (`r round((3+20+4+2+9)*100/170, 2)`%) of retouched tools are from local types, `r 58+24+3+1+3` (`r round((58+24+3+1+3)*100/170, 2)`%) from non-local cherts and 4 are from the TL group.
+From the identified chert types, `r 3+35` (`r round((3+35)*100/170, 2)`%) of retouched tools are from local types, `r 58+24+3+1+3` (`r round((58+24+3+1+3)*100/170, 2)`%) from non-local cherts and 4 are from the TL group.
 From all chert types, T6 shows the highest numbers of retouched pieces (n=58, `r round(58*100/170, 2)`%).
-Regarding typologies, 9 types of retouched tools were identified (excluding retouched tool fragments).
+Regarding typologies, 8 types of retouched tools were identified (excluding retouched tool fragments).
 All these typologies were found in the non-local chert group, while only 5 retouched tool typologies were identified in local cherts/chalcedony.
-The most common typologies identified in the assemblage (and for all chert types) were burins (n=54), endscrappers (n=18), retouched blanks (n=38) and splintered pieces (n=26).
+The most common typologies identified in the assemblage (and for all chert types) were burins (n=54), endscrapers (n=18), retouched blanks (n=38) and splintered pieces (n=26).
 From all these, the majority were produced in non-local chert, mainly T6 and in lesser numbers T7.
-Although in lesser numbers as well, backed bladelets were also identified (n=6), but only present in non-local chert types (T6 and T7).
+Although in lesser numbers as well, backed tools were also identified (n=7), but only present in non-local chert types (T6 and T7).
 The presence of burin spalls in local chert, especially T2 (n=11) and non-local cherts (T6 with n=12 and T7 with n=4) is expected due to the prevalence of burins (@tbl-general-gravettian).
@@
 #| label: tbl-retouch-gravettian
-#| tbl-cap: Number and percentage of retouched tools by raw material of the Gravettian assemblage (levels 6 and 7) of the Terrace area.
+#| tbl-cap: Retouched tools (n) by chert type of the Gravettian assemblage (levels 6 and 7) of the Terrace area.
 
@@
 Of these, 167 (`r round(167*100/635, 2)`%) have an indeterminate chert type.
-Local cherts correspond to `r 32+62+32+158+51` artefacts (`r round((32+62+32+158+51)*100/635, 2)`%), with T4 being the most present local chert.
+Local cherts correspond to `r 32+303` artefacts (`r round((32+303)*100/635, 2)`%), with T4 being the most present local chert.
 Non-local types are present in smaller numbers (n=`r 60+47+13+9+1` or `r round((60+47+13+9+1)*100/635, 2)`%), with T6 and T7 being the most relevant (n=60 and n=47, respectively), and the TL group includes 3 artefacts, corresponding to 2 types of TL cherts [@tbl-TL-classes].
@@
 As expected, the TL artifacts include a blank and a retouched.
-However, a shatter was also identified.
+However, a shatter was also identified [@tbl-TL-classes].
 
@@
 #| label: tbl-general-proto
-#| tbl-cap: Number and percentage of technotypological classes by raw material type from the Proto-Solutrean assemblage (levels 4E and top of level 5) of the Terrace area of Vale Boi.
+#| tbl-cap: Number and percentage of techno-typological classes by chert type from the Proto-Solutrean assemblage (levels 4E and top of level 5) of the Terrace area of Vale Boi.
 
@@
 In total, 65 chert retouched tools were identified, with 23.1% (n=15) having an unidentified chert type (@tbl-retouch-proto).
-A total of `r 1+10+2+8+5` (`r round((1+10+2+8+5)*100/65, 2)`%) retouched tools correspond to local cherts, while `r 10+9+3+1+1` (`r round((10+9+3+1+1)*100/65, 2)`%) correspond to non-local and TL cherts.
+A total of `r 1+25` (`r round((1+25)*100/65, 2)`%) retouched tools correspond to local cherts, while `r 10+9+3+1+1` (`r round((10+9+3+1+1)*100/65, 2)`%) correspond to non-local and TL cherts.
 As seen in @tbl-retouch-proto, 11 different types of retouched tools were identified.
-Endscrappers (n=18), retouched blanks (n=14), splintered pieces (n=11) and burins (n=8) were the most prevalent.
+Endscrapers (n=18), retouched blanks (n=14), splintered pieces (n=11) and burins (n=8) were the most prevalent.
 Local raw materials show 8 different typologies, with Vale Comprido points, Solutrean retouch and notches being present only in the local types, even if in small numbers.
 Non-local cherts show 8 different typologies of retouched tools.
-Perforator-endscrapper, truncation and backed bladelet typologies can only be found in non-local varieties, albeit also in small numbers.
+Perforator-endscraper, truncation and backed bladelet typologies can only be found in non-local varieties, albeit also in small numbers.
 
@@
 #| label: tbl-retouch-proto
-#| tbl-cap: Number and percentage of retouched tools by raw material of the Proto-Solutrean assemblage (levels 4E and top of level 5) of the Terrace area.
+#| tbl-cap: Retouched tools (n) by chert type of the Proto-Solutrean assemblage (levels 4E and top of level 5) of the Terrace area.
 
@@
 Local cherts correspond to `r 16+548` (`r round((16+548)*100/914, 2)`%) artefacts, mostly from T2.
-Non-local cherts (including the TL group) correspond to `r 85+9+30+3+1+16+1` (`r round((85+9+30+3+1+16+1)*100/914, 2)`%) artefacts, mostly from T6 (n=86).
+Non-local cherts (including the TL group) correspond to `r 85+9+30+3+1+16+1` (`r round((85+9+30+3+1+16+1)*100/914, 2)`%) artefacts, mostly from T6 (n=85).
 
@@
 #| label: tbl-general-solutrean
-#| tbl-cap: Number and percentage of technotypological classes by raw material type from the Solutrean assemblage (levels A to C) of the Shelter area of Vale Boi.
+#| tbl-cap: Number and percentage of techno-typological classes by chert type from the Solutrean assemblage (levels A to C) of the Shelter area of Vale Boi.
 
@@
 
-The analysed Solutrean sample has a total of 43 retouched tools (complete and fragments), with `r 3+13+1` corresponding to local cherts and `r 7+1+2+3` to non-local cherts, with the remaining being cherts of unidentified type.
+The analysed Solutrean sample has a total of 43 retouched tools (complete and fragments), with `r 3+14` corresponding to local cherts and `r 7+1+2+3` to non-local cherts, with the remaining being cherts of unidentified type.
 These results are significantly reduced from the original dataset, possibly due to no bucket coordinates being used.
-From the 43 retouched tools, 11 different typologies were identified (without accounting for fragments).
-The typologies which are most frequently present in the studied sample are bifacial blanks and bifacial fragments (n=10).
+From the 43 retouched tools, 10 different typologies were identified (without accounting for fragments).
+The typologies which are most frequently present in the studied sample are bifacial blanks (n=10).
 Two types of points were identified: shouldered points (n=3, including fragments) and tanged and winged points (n=4).
-Other frequent retouched typologies are retouched blanks (n=7), notches (n=5), splintered pieces (n=5) and endscrappers (n=4).
+Other frequent retouched typologies are retouched blanks (n=7), notches (n=5), splintered pieces (n=5) and endscrapers (n=4).
 Local types show a higher variability retouched typologies (9 types), although this may be related to the higher number of tools.
@@
 #| label: tbl-retouch-solutrean
-#| tbl-cap: Number and percentage of retouched tools by raw material of the Solutrean assemblage (level A to C) of the Shelter area.
+#| tbl-cap: Retouched tools (n) by chert type of the Solutrean assemblage (level A to C) of the Shelter area.
 
@@
 #| label: tbl-tdratio-solutrean
-#| tbl-cap: Tools to debitage ratio by raw material of the Gravettian (a), Proto-Solutrean (b) and Solutrean (c) assemblages.
+#| tbl-cap: Tools-to-debitage ratio by chert type of the Gravettian (a), Proto-Solutrean (b) and Solutrean (c) assemblages.
 #| tbl-subcap: 
-#|  - "Tools to debitage ratio by raw material of the Gravettian assemblage (levels 6 and 7) of the Terrace area."
-#|  - "Tools to debitage ratio by raw material of the Proto-Solutrean assemblage (levels 4E and top of level 5) of the Terrace area."
-#|  - "Tools to debitage ratio by raw material of the Solutrean assemblage (levels A to C) of the Shelter area."
+#|  - "Tools-to-debitage ratio by chert type of the Gravettian assemblage (levels 6 and 7) of the Terrace area."
+#|  - "Tools-to-debitage ratio by chert type of the Proto-Solutrean assemblage (levels 4E and top of level 5) of the Terrace area."
+#|  - "Tools-to-debitage ratio by chert type of the Solutrean assemblage (levels A to C) of the Shelter area."
 #| layout-nrow: 3
@@
 #| label: tbl-bcratio
-#| tbl-cap: Blank to core ratio by raw material of the Gravettian (a), Proto-Solutrean (b) and Solutrean (c) assemblages.
+#| tbl-cap: Blank-to-core ratio by chert type of the Gravettian (a), Proto-Solutrean (b) and Solutrean (c) assemblages.
 #| tbl-subcap: 
-#|  - "Blank to core ratio by raw material of the Gravettian assemblage (levels 6 and 7) of the Terrace area."
-#|  - "Blank to core ratio by raw material of the Proto-Solutrean assemblage (levels 4E and top of level 5) of the Terrace area."
-#|  - "Blank to core ratio by raw material of the Solutrean assemblage (levels A to C) of the Shelter area."
+#|  - "Blank-to-core ratio by chert type of the Gravettian assemblage (levels 6 and 7) of the Terrace area."
+#|  - "Blank-to-core ratio by chert type of the Proto-Solutrean assemblage (levels 4E and top of level 5) of the Terrace area."
+#|  - "Blank-to-core ratio by chert type of the Solutrean assemblage (levels A to C) of the Shelter area."
 #| layout-nrow: 3
@@
 
-# Measurements
+# Measurements descriptive statistics
 
@@
 #| label: tbl-gravettian
-#| tbl-cap: "Stats for Gravettian occupation."
+#| tbl-cap: "Mean and standard deviation (SD) for Max width, Length and Weight measurements of blanks (a) and cores (b) by chert type of the Gravettian assemblage (levels 6-7) of the Terrace area."
 #| tbl-subcap: 
-#|  - "Stats for blanks"
-#|  - "Stats for cores"
+#|  - "Mean and SD for Max width, Length and Weight measurements of blanks."
+#|  - "Mean and SD for Max width, Length and Weight measurements of cores."
 #| layout-nrow: 2
@@
 #| label: tbl-proto
-#| tbl-cap: "Stats for Proto-Solutrean occupation."
+#| tbl-cap: "Mean and standard deviation (SD) for Max width, Length and Weight measurements of blanks (a) and cores (b) by chert type of the Proto-Solutrean assemblage (levels 4E and top of level 5) of the Terrace area."
 #| tbl-subcap: 
-#|  - "Stats for blanks"
-#|  - "Stats for cores"
+#|  - "Mean and SD for Max width, Length and Weight measurements of blanks."
+#|  - "Mean and SD for Max width, Length and Weight measurements of cores."
 #| layout-nrow: 2
@@
 #| label: tbl-solutrean
-#| tbl-cap: "Stats for Solutrean occupation."
+#| tbl-cap: "Mean and standard deviation (SD) for Max width, Length and Weight measurements of blanks (a) and cores (b) by chert type of the Solutrean assemblage (levels A to C) of the Shelter area."
 #| tbl-subcap: 
-#|  - "Stats for blanks"
-#|  - "Stats for cores"
+#|  - "Mean and SD for Max width, Length and Weight measurements of blanks."
+#|  - "Mean and SD for Max width, Length and Weight measurements of cores."
 #| layout-nrow: 2
@@
 
-# Statistical tests
+# Measurements non-parametric statistical analysis
 
@@
 #| label: tbl-pvalue
-#| tbl-cap: "Median values and p value results for the Mann-Whitney test applied to blank and core measurements (weight, length and width)."
+#| tbl-cap: "Median values and p value results for the Mann-Whitney U statistical test applied to blank (a) and core (b) measurements (Weight, Length and Max width).
+All local chert types were grouped in the L category and non-local groups in the NL category.
+(n) corresponds to the total number of samples used for the statistical tests.
+* corresponds to significant p values (<0.05), ** to very significant p values (< 0.01) and *** to highly significant p values (0.001)."
 #| tbl-subcap: 
-#|  - "Median values and p value results for the Mann-Whitney test applied to blank measurements"
-#|  - "Median values and p value results for the Mann-Whitney test applied to core measurements"
+#|  - "Median values and p value results for the Mann-Whitney U statistical test applied to blank measurements."
+#|  - "Median values and p value results for the Mann-Whitney U statistical test applied to core measurements."
 #| layout-nrow: 2

### 2025-04-15 README.qmd
Updated read_me

@@
 # Please put your title here to include it in the file below.
-Title <- "Title of your paper goes here"
+Title <- "From Stone to Tool: How Raw Materials Influenced Upper Paleolithic Technology in Southwestern Iberia (Vale Boi)"
 ```
 
-# article3
+# from-stone-to-tool
 
@@
 
-> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>
+> Belmiro, J., Bicho, N., Terradas, X., Cascalheira, J., (in prep). _`r Title`_. Archaeological and Anthropological Sciences <https://doi.org/xxx/xxx>
 
@@
 
-> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+> Belmiro, J., Bicho, N., Terradas, X., Cascalheira, J., (`r format(Sys.Date(), "%Y")`).
+_Compendium of R code and data for `r Title`_.
+Accessed `r format(Sys.Date(), "%d %b %Y")`.
+Online at <https://doi.org/10.17605/OSF.IO/UWMPK>
 
@@
     for manuscript. Includes code to reproduce the figures and tables
-    generated by the analysis. It also has a rendered version,
-    `paper.docx`, suitable for reading (the code is replaced by figures
-    and tables in this file)
+    generated by the analysis.
   - [:file\_folder: data](/analysis/data): Data used in the analysis.

### 2025-04-15 analysis/paper/paper.qmd
Update institution

@@
       institute:
-        - Institución Milá y Fontanals de Investigación en Humanidades, Consejo Superior de Investigaciones Científicas, Barcelona, Spain
+        - Spanish National Research Council - Institute Milá y Fontanals of research on Humanities, Barcelona, Spain
   - João Cascalheira:

### 2025-05-06 analysis/paper/paper.qmd
Improvements
Added ggsave codes for future graph saving

@@
 
-**Author contributions:** Study conception and design were performed by Joana Belmiro and João Cascalheira.
+**Author contributions:** Study conception and design were performed by Joana Belmiro, Nuno Bicho and João Cascalheira.
 Material preparation, data collection, database collection and treatment and analysis were performed by Joana Belmiro.
@@
 
+#ggsave("general_graphs.png", width = 6, height = 8, units = "in", dpi = 300, bg = "white")
+
 ```
@@
 ggpubr::ggarrange(td_ratio, bc_ratio, legend_b, labels = c("a", "b", ""), ncol = 1, heights = c(3, 3, 1))
+
+#ggsave("ratios.png", width = 6, height = 7, units = "in", dpi = 300, bg = "white")
 
@@
 
+#ggsave("gravettian_plots.png", width = 6, height = 6, units = "in", dpi = 300, bg = "white")
+
 ```
@@
 
+#ggsave("proto_plots.png", width = 6, height = 6, units = "in", dpi = 300, bg = "white")
+
 ```
@@
 
+#ggsave("solutrean_plots.png", width = 6, height = 6, units = "in", dpi = 300, bg = "white")
+
 ```

### 2025-06-19 analysis/paper/paper.qmd
Review corrections. Tables and graphs altered to include flakes and elongated blanks.

@@
     RETOUCHEDPIECETYPOLOGY == "Carenated endscraper" ~ "Endscraper",
-    TRUE ~ RETOUCHEDPIECETYPOLOGY))
+    TRUE ~ RETOUCHEDPIECETYPOLOGY)) |>
+  # --Article review--
+  mutate(elong = MAXWIDTH*2-LENGTH) %>%
+  mutate(CLASS_ELONG = case_when( 
+    CLASS == "Blank" & elong > 0 ~ "Flake",
+    CLASS == "Blank" & elong <= 0 ~ "ElongatedProd",
+    TRUE ~ CLASS))
+  # ---
 
@@
     TRUE ~ RMTYPE)) |> 
-  dplyr::mutate(CLASS = case_when(
-    CLASS == "BlankFrag" ~ "Blank",
-    CLASS == "BurinSpall" ~ "Burin spall",
-    CLASS == "CoreFrag" ~ "Core",
-    CLASS == "CorePreparProd" ~ "Core prep product",
-    CLASS == "RetouchedPiece" | CLASS == "RetouchedPieceFrag" ~ "Retouched",
-    TRUE ~ CLASS)) |> 
+  # --Article review--
+  dplyr::mutate(CLASS_ELONG = case_when(
+    CLASS_ELONG == "BlankFrag" ~ "Blank fragment",
+    CLASS_ELONG == "ElongatedProd" ~ "Elongated blank",
+    CLASS_ELONG == "BurinSpall" ~ "Burin spall",
+    CLASS_ELONG == "CoreFrag" ~ "Core",
+    CLASS_ELONG == "CorePreparProd" ~ "Core prep product",
+    CLASS_ELONG == "RetouchedPiece" | CLASS == "RetouchedPieceFrag" ~ "Retouched",
+    TRUE ~ CLASS_ELONG)) |> 
   filter(Level %in% c(6, 7)) |>
@@
                                         "T8", "T9", "T10", "TL"))) |>
-  dplyr::group_by(RMTYPE, CLASS) |> 
+  mutate(CLASS_ELONG = factor(CLASS_ELONG, levels = c("Flake", "Elongated blank", "Blank fragment", "Burin spall", "Core", "Core prep product", "Retouched", "Shatter"))) |> 
+  dplyr::group_by(RMTYPE, CLASS_ELONG) |> 
   dplyr::summarise(n = n()) |> 
   mutate(percent = n/sum(n)) |>
-  ggplot(aes(x = RMTYPE, y = percent, fill = CLASS)) +
+  ggplot(aes(x = RMTYPE, y = percent, fill = CLASS_ELONG)) +
   geom_bar(stat = "identity", width = 0.5) +
-  geom_text(aes(label=ifelse(n > 7, n, "")), position = position_stack(vjust = .5), size = 3, color = "white") +
+  geom_text(aes(label=ifelse(n > 4, n, "")), position = position_stack(vjust = .5), size = 3, color = "white") +
   scale_y_continuous(labels = scales::percent_format()) +
@@
   labs(y = " ") +
-  scale_fill_manual(name = "Class", values = c("#D49B54", "#8B6969", "#CDAF95", "#2B4F60",
+  scale_fill_manual(name = "Class", values = c("#D49B54", "#E66101", "#FDB863", "#8B6969", "#CDAF95", "#2B4F60",
                               "tomato4", "#798777"))
@@
     TRUE ~ RMTYPE)) |> 
-  dplyr::mutate(CLASS = case_when(
-    CLASS == "BlankFrag" ~ "Blank",
-    CLASS == "BurinSpall" ~ "Burin spall",
-    CLASS == "CoreFrag" ~ "Core",
-    CLASS == "CorePreparProd" ~ "Core prep product",
-    CLASS == "RetouchedPiece" | CLASS == "RetouchedPieceFrag" ~ "Retouched",
-    TRUE ~ CLASS)) |> 
+  dplyr::mutate(CLASS_ELONG = case_when(
+    CLASS_ELONG == "BlankFrag" ~ "Blank fragment",
+    CLASS_ELONG == "ElongatedProd" ~ "Elongated blank",
+    CLASS_ELONG == "BurinSpall" ~ "Burin spall",
+    CLASS_ELONG == "CoreFrag" ~ "Core",
+    CLASS_ELONG == "CorePreparProd" ~ "Core prep product",
+    CLASS_ELONG == "RetouchedPiece" | CLASS == "RetouchedPieceFrag" ~ "Retouched",
+    TRUE ~ CLASS_ELONG)) |>
   filter(Level %in% c("4E", "4E/5")) |>
@@
   theme(axis.title.x=element_blank()) +
-  scale_fill_manual(values = c("#D49B54", "#8B6969", "#CDAF95", "#2B4F60",
+  scale_fill_manual(values = c("#D49B54", "#E66101", "#FDB863", "#8B6969", "#CDAF95", "#2B4F60",
                               "tomato4", "#798777"))
@@
   dplyr::mutate(CLASS = case_when(
-    CLASS == "ElongBlank" | CLASS == "Flake" ~ "Blank",
-    CLASS == "ElongBlankFrag" | CLASS == "FlakeFrag" ~ "Blank",
-    CLASS == "Thinning flake" ~ "Blank",
+    CLASS == "ElongBlank" ~ "Elongated blank",
+    CLASS == "ElongBlankFrag" | CLASS == "FlakeFrag" ~ "Blank fragment",
+    #CLASS == "Thinning flake" ~ "Blank",
     CLASS == "CorePrepProd" ~ "Core prep product",
@@
                                         "T8", "T9", "T10","T11", "TL"))) |>
+  mutate(CLASS = factor(CLASS, levels = c("Flake", "Elongated blank", "Thinning flake", "Blank fragment", "Burin spall", "Core", "Core prep product", "Retouched", "Shatter"))) |> 
   dplyr::group_by(RMTYPE, CLASS) |> 
@@
   geom_bar(stat = "identity", width = 0.5) +
-  geom_text(aes(label=ifelse(n > 7, n, "")), position = position_stack(vjust = .5), size = 3, color = "white") +
+  geom_text(aes(label=ifelse(n > 4, n, "")), position = position_stack(vjust = .5), size = 3, color = "white") +
   scale_y_continuous(labels = scales::percent_format()) +
@@
   theme(axis.title.x=element_blank()) +
-  scale_fill_manual(values = c("#D49B54", "#8B6969", "#CDAF95", "#2B4F60",
+  scale_fill_manual(values = c("#D49B54","#E66101", "#D27D70", "#FDB863", "#8B6969", "#CDAF95", "#2B4F60",
                               "tomato4", "#798777"))
@@
    mutate(CLASS = case_when(
-    CLASS == "ElongBlank" | CLASS == "Flake" ~ "Blank",
-    CLASS == "FlakeFrag" | CLASS == "ElongBlankFrag" ~ "BlankFrag",
+    CLASS == "ElongBlank" ~ "Elongated blank",
     TRUE ~ CLASS)) |> 
@@
 solutrean_scatter <- solutrean_stats |>
-  filter(CLASS %in% c("Blank", "Core")) |>
+  filter(CLASS %in% c("Flake", "Elongated blank", "Core")) |>
   filter(MAXWIDTH !="0" & LENGTH != "0") |>
@@
                                         "T8", "T9", "T10", "T11", "TL"))) |>
-  mutate(CLASS = factor(CLASS, levels = c("Blank", "Core"))) |>
+  mutate(CLASS = factor(CLASS, levels = c("Flake", "Elongated blank", "Core"))) |>
   ggplot(aes(x = LENGTH, y = MAXWIDTH, fill = RMTYPE)) +
   geom_point(alpha = 0.9, size = 3, shape = 21, colour = "black") +
-  facet_wrap(~CLASS) +
+  facet_wrap(~CLASS, ncol = 1, scales = "free_y") +
   labs(x="Length (mm)", y="Max. Width (mm)") +
@@
 solutrean_stats_weight <- solutrean_stats |>
-  filter(CLASS %in% c("Blank", "Core")) |> 
+  filter(CLASS %in% c("Flake", "Elongated blank", "Core")) |> 
   filter(WEIGHT !="0") |>
@@
                                         "T8", "T9", "T10", "T11", "TL"))) |>
-  mutate(CLASS = factor(CLASS, levels = c("Blank", "Core"))) |>
+  mutate(CLASS = factor(CLASS, levels = c("Flake", "Elongated blank", "Core"))) |>
   ggplot(aes(RMTYPE, WEIGHT, fill = RMTYPE)) +
   geom_boxplot(aes(fill=factor(RMTYPE)), outlier.shape=21) +
-  facet_wrap(~CLASS, scales = "free_y") +
+  facet_wrap(~CLASS, ncol = 1, scales = "free_y") +
   stat_summary(fun=mean, geom="point", shape=23) +
-  labs(y="Weight (gr)") +
+  labs(x= " ", y="Weight (gr)") +
   #coord_cartesian(ylim =  c(0, 50)) +
@@
   theme(axis.text.x=element_blank(),
-        axis.ticks.x=element_blank(), axis.title.x = element_blank())
+        axis.ticks.x=element_blank())
 
@@
     RMTYPE == "T1" | RMTYPE == "T2" | RMTYPE == "T3" | RMTYPE == "T4" | RMTYPE == "T5" ~ "Local",
-    TRUE ~ RMTYPE))
+    TRUE ~ RMTYPE)) |> 
+  dplyr::mutate(CLASS_ELONG = case_when(
+    CLASS_ELONG == "ElongatedProd" ~ "Elongated blank",
+    TRUE ~ CLASS_ELONG))
 
@@
 proto_scatter <- proto_stats |>
-  filter(CLASS %in% c("Blank", "Core")) |>
+  filter(CLASS_ELONG %in% c("Flake", "Elongated blank", "Core")) |>
   filter(MAXWIDTH !="0" & LENGTH != "0") |>
@@
                                         "T8", "T9", "T10", "TL"))) |>
-  mutate(CLASS = factor(CLASS, levels = c("Blank", "Core"))) |>
+  mutate(CLASS_ELONG = factor(CLASS_ELONG, levels = c("Flake", "Elongated blank", "Core"))) |>
   ggplot(aes(x = LENGTH, y = MAXWIDTH, fill = RMTYPE)) +
   geom_point(alpha = 0.9, size = 3, shape = 21, colour = "black") +
-  facet_wrap(~CLASS) +
+  facet_wrap(~CLASS_ELONG, ncol = 1, scales = "free_y") +
   labs(x="Length (mm)", y="Max. Width (mm)") +
@@
 proto_stats_weight <- proto_stats |>
-  filter(CLASS %in% c("Blank", "Core")) |> 
+  filter(CLASS_ELONG %in% c("Flake", "Elongated blank", "Core")) |> 
   filter(WEIGHT !="0") |>
   mutate(RMTYPE = factor(RMTYPE, levels = c("Local", "T6", "T7", "T8", "T9", "T10", "TL"))) |>
-  mutate(CLASS = factor(CLASS, levels = c("Blank", "Core"))) |>
+  mutate(CLASS_ELONG = factor(CLASS_ELONG, levels = c("Flake", "Elongated blank", "Core"))) |>
   ggplot(aes(RMTYPE, WEIGHT, fill = RMTYPE)) +
   geom_boxplot(aes(fill=factor(RMTYPE)), outlier.shape=21) +
-  facet_wrap(~CLASS, scales = "free_y") +
+  facet_wrap(~CLASS_ELONG, ncol = 1, scales = "free_y") +
   stat_summary(fun.y=mean, geom="point", shape=23) +
-  labs(y="Weight (gr)") +
+  labs(x= " ", y="Weight (gr)") +
   #coord_cartesian(ylim =  c(0, 50)) +
@@
 gravettian_scatter <- gravettian_stats |>
-  filter(CLASS %in% c("Blank", "Core")) |>
+  filter(CLASS_ELONG %in% c("Flake", "Elongated blank", "Core")) |>
   filter(MAXWIDTH !="0" & LENGTH != "0") |>
@@
 gravettian_stats_weight <- gravettian_stats |>
-  filter(CLASS %in% c("Blank", "Core")) |> 
+  filter(CLASS_ELONG %in% c("Flake", "Elongated blank", "Core")) |> 
   filter(WEIGHT !="0") |>
@@
   mutate(RMTYPE = factor(RMTYPE, levels = c("Local", "T6", "T7", "T8", "T9", "T10", "TL"))) |>
-  mutate(CLASS = factor(CLASS, levels = c("Blank", "Core"))) |>
+  mutate(CLASS_ELONG = factor(CLASS_ELONG, levels = c("Flake", "Elongated blank", "Core"))) |>
   ggplot(aes(RMTYPE, WEIGHT, fill = RMTYPE)) +
   geom_boxplot(aes(fill=factor(RMTYPE)), outlier.shape=21) +
-  facet_wrap(~CLASS, scales = "free_y") +
+  facet_wrap(~CLASS_ELONG, ncol = 1, scales = "free_y") +
   stat_summary(fun.y=mean, geom="point", shape=23) +
-  labs(y="Weight (gr)") +
+  labs(x= " ", y="Weight (gr)") +
   #coord_cartesian(ylim =  c(0, 60)) +
@@
 #| label: fig-gravettian-plots
-#| fig-cap: "Gravettian assemblage blank and core measurements and weight descriptive statistics in local and non-local chert types.
+#| fig-cap: "Gravettian assemblage flake, elongated blank and core measurements and weight descriptive statistics in local and non-local chert types.
 a) Scatterplots of width and length.
@@
 
-ggpubr::ggarrange(gravettian_scatter, gravettian_stats_weight, labels = c("a", "b"), nrow = 2, common.legend = TRUE, legend = "bottom")
+ggpubr::ggarrange(gravettian_scatter, gravettian_stats_weight, labels = c("a", "b"), ncol = 2, common.legend = TRUE, legend = "bottom")
 
@@
 #| label: fig-proto-plots
-#| fig-cap: "Proto-Solutrean assemblage blank and core measurements and weight descriptive statistics in local and non-local chert types.
+#| fig-cap: "Proto-Solutrean assemblage flake, elongated blank and core measurements and weight descriptive statistics in local and non-local chert types.
 a) Scatterplots of width and length.
@@
 
-ggpubr::ggarrange(proto_scatter, proto_stats_weight, labels = c("a", "b"), nrow = 2, common.legend = TRUE, legend = "bottom")
+ggpubr::ggarrange(proto_scatter, proto_stats_weight, labels = c("a", "b"), ncol = 2, common.legend = TRUE, legend = "bottom")
 
@@
 #| label: fig-solutrean-plots
-#| fig-cap: "Solutrean assemblage blank and core measurements and weight descriptive statistics in local and non-local chert types.
+#| fig-cap: "Solutrean assemblage flake, elongated blank and core measurements and weight descriptive statistics in local and non-local chert types.
 a) Scatterplots of width and length.
@@
 
-ggpubr::ggarrange(solutrean_scatter, solutrean_stats_weight, labels = c("a", "b"), nrow = 2, common.legend = TRUE, legend = "bottom")
+ggpubr::ggarrange(solutrean_scatter, solutrean_stats_weight, labels = c("a", "b"), ncol = 2, common.legend = TRUE, legend = "bottom")
 

### 2025-06-19 analysis/supplementary-materials/ESM_1.qmd
Review corrections. Tables and graphs altered to include flakes and elongated blanks.

@@
     RETOUCHEDPIECETYPOLOGY == "Carenated endscraper" ~ "Endscraper",
-    TRUE ~ RETOUCHEDPIECETYPOLOGY))
+    TRUE ~ RETOUCHEDPIECETYPOLOGY)) |> 
+  # --Article review--
+  mutate(elong = MAXWIDTH*2-LENGTH) %>%
+  mutate(CLASS_ELONG = case_when( 
+    CLASS == "Blank" & elong > 0 ~ "Flake",
+    CLASS == "Blank" & elong <= 0 ~ "ElongatedProd",
+    TRUE ~ CLASS))
 
@@
 general_table_som <- phase_dataset |> 
-  dplyr::mutate(CLASS = case_when(
-    CLASS == "BlankFrag" ~ "Blank fragment",
-    CLASS == "BurinSpall" ~ "Burin spall",
-    CLASS == "CoreFrag" ~ "Core fragment",
-    CLASS == "CorePreparProd" ~ "Core prep product",
-    CLASS == "RetouchedPiece" ~ "Retouched",
-    CLASS == "RetouchedPieceFrag" ~ "Retouched fragment",
-    TRUE ~ CLASS)) |> 
+  dplyr::mutate(CLASS_ELONG = case_when(
+    CLASS_ELONG == "BlankFrag" ~ "Blank fragment",
+    CLASS_ELONG == "ElongatedProd" ~ "Elongated blank",
+    CLASS_ELONG == "BurinSpall" ~ "Burin spall",
+    CLASS_ELONG == "CoreFrag" ~ "Core fragment",
+    CLASS_ELONG == "CorePreparProd" ~ "Core prep product",
+    CLASS_ELONG == "RetouchedPiece" ~ "Retouched",
+    CLASS_ELONG == "RetouchedPieceFrag" ~ "Retouched fragment",
+    TRUE ~ CLASS_ELONG)) |> 
   dplyr::mutate(RMTYPE = case_when(
@@
                                         "T8", "T9", "T10", "TL", "INDET"))) |>
-  tabyl(CLASS, RMTYPE) |> 
+  mutate(CLASS_ELONG = factor(CLASS_ELONG, levels = c("Flake", "Elongated blank", "Blank fragment", "Burin spall", "Core", "Core fragment", "Core prep product", "Retouched", "Retouched fragment", "Shatter"))) |>
+  tabyl(CLASS_ELONG, RMTYPE) |> 
   adorn_totals(c("row", "col")) |> 
@@
   dplyr::mutate(CLASS = case_when(
-    CLASS == "ElongBlank" | CLASS == "Flake" ~ "Blank",
+    CLASS == "ElongBlank" ~ "Elongated blank",
     CLASS == "ElongBlankFrag" | CLASS == "FlakeFrag" ~ "Blank fragment",
-    CLASS == "Thinning flake" ~ "Blank",
     CLASS == "CorePrepProd" ~ "Core prep product",
@@
                                         "T8", "T9", "T10", "T11", "TL", "INDET"))) |>
+    mutate(CLASS = factor(CLASS, levels = c("Flake", "Elongated blank", "Thinning flake", "Blank fragment", "Burin spall", "Core", "Core fragment", "Core prep product", "Retouched", "Retouched fragment", "Shatter"))) |>
   tabyl(CLASS, RMTYPE) |> 
@@
   mutate(CLASS = case_when(
-    CLASS == "ElongBlank" | CLASS == "Flake" ~ "Blank",
+    CLASS == "ElongBlank" ~ "Elongated blank",
     TRUE ~ CLASS)) |> 
@@
 solutrean_stats_blank <- solutrean_stats |>  
-  filter(CLASS == "Blank") |> 
+  filter(CLASS == "Flake") |> 
+  select(RMTYPE, CLASS, MAXWIDTH, LENGTH, WEIGHT) |> 
+  drop_na() |> 
+  group_by(RMTYPE, CLASS) |> 
+  mutate(RMTYPE = factor(RMTYPE, levels = c("T1", "T2-5", "T6","T7", 
+                                        "T8", "T9", "T10", "T11", "TL"))) |>
+  summarise(across(c(MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd)))
+# For elongated blanks (complete) --Article review--
+solutrean_stats_elong <- solutrean_stats |>  
+  filter(CLASS == "Elongated blank") |> 
   select(RMTYPE, CLASS, MAXWIDTH, LENGTH, WEIGHT) |> 
@@
 proto_stats_blank <- proto_stats |> 
-  filter(CLASS == "Blank") |> 
-  select(RMTYPE, CLASS, MAXWIDTH, LENGTH, WEIGHT) |> 
-  group_by(RMTYPE, CLASS) |> 
+  filter(CLASS_ELONG == "Flake") |> 
+  select(RMTYPE, CLASS_ELONG, MAXWIDTH, LENGTH, WEIGHT) |> 
+  group_by(RMTYPE, CLASS_ELONG) |> 
+  rename(CLASS = CLASS_ELONG) |> 
+  mutate(RMTYPE = factor(RMTYPE, levels = c("T1", "T2-5", "T6","T7", 
+                                        "T8", "T9", "T10", "TL"))) |>
+  summarise(across(c(MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd)))
+# For elongated blanks (complete) --Article review--
+proto_stats_elong <- proto_stats |> 
+  dplyr::mutate(CLASS_ELONG = case_when(
+    CLASS_ELONG == "ElongatedProd" ~ "Elongated blank",
+    TRUE ~ CLASS_ELONG)) |> 
+  filter(CLASS_ELONG == "Elongated blank") |> 
+  select(RMTYPE, CLASS_ELONG, MAXWIDTH, LENGTH, WEIGHT) |> 
+  group_by(RMTYPE, CLASS_ELONG) |> 
+  rename(CLASS = CLASS_ELONG) |> 
   mutate(RMTYPE = factor(RMTYPE, levels = c("T1", "T2-5", "T6","T7", 
@@
 gravettian_stats_blank <- gravettian_stats |> 
-  filter(CLASS == "Blank") |> 
-  select(RMTYPE, CLASS, MAXWIDTH, LENGTH, WEIGHT) |> 
-  group_by(RMTYPE, CLASS) |> 
+  filter(CLASS_ELONG == "Flake") |> 
+  select(RMTYPE, CLASS_ELONG, MAXWIDTH, LENGTH, WEIGHT) |> 
+  group_by(RMTYPE, CLASS_ELONG) |> 
+  rename(CLASS = CLASS_ELONG) |> 
+  mutate(RMTYPE = factor(RMTYPE, levels = c("T1", "T2-5", "T6","T7", 
+                                        "T8", "T9", "T10", "TL"))) |>
+  summarise(across(c(MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd)))
+# For elongated blanks (complete) --Article review--
+gravettian_stats_elong <- gravettian_stats |> 
+  dplyr::mutate(CLASS_ELONG = case_when(
+    CLASS_ELONG == "ElongatedProd" ~ "Elongated blank",
+    TRUE ~ CLASS_ELONG)) |> 
+  filter(CLASS_ELONG == "Elongated blank") |> 
+  select(RMTYPE, CLASS_ELONG, MAXWIDTH, LENGTH, WEIGHT) |> 
+  group_by(RMTYPE, CLASS_ELONG) |> 
+  rename(CLASS = CLASS_ELONG) |> 
   mutate(RMTYPE = factor(RMTYPE, levels = c("T1", "T2-5", "T6","T7", 
@@
 #| label: tbl-gravettian
-#| tbl-cap: "Mean and standard deviation (SD) for Max width, Length and Weight measurements of blanks (a) and cores (b) by chert type of the Gravettian assemblage (levels 6-7) of the Terrace area."
+#| tbl-cap: "Mean and standard deviation (SD) for Max width, Length and Weight measurements of flakes (a), elongated blanks (b) and cores (c) by chert type of the Gravettian assemblage (levels 6-7) of the Terrace area."
 #| tbl-subcap: 
-#|  - "Mean and SD for Max width, Length and Weight measurements of blanks."
+#|  - "Mean and SD for Max width, Length and Weight measurements of flakes."
+#|  - "Mean and SD for Max width, Length and Weight measurements of elongated blanks."
 #|  - "Mean and SD for Max width, Length and Weight measurements of cores."
-#| layout-nrow: 2
+#| layout-nrow: 3
 
 kable(gravettian_stats_blank)
+kable(gravettian_stats_elong)
 kable(gravettian_stats_cores)
@@
 #| label: tbl-proto
-#| tbl-cap: "Mean and standard deviation (SD) for Max width, Length and Weight measurements of blanks (a) and cores (b) by chert type of the Proto-Solutrean assemblage (levels 4E and top of level 5) of the Terrace area."
+#| tbl-cap: "Mean and standard deviation (SD) for Max width, Length and Weight measurements of flakes (a), elongated blanks (b) and cores (c) by chert type of the Proto-Solutrean assemblage (levels 4E and top of level 5) of the Terrace area."
 #| tbl-subcap: 
-#|  - "Mean and SD for Max width, Length and Weight measurements of blanks."
+#|  - "Mean and SD for Max width, Length and Weight measurements of flakes."
+#|  - "Mean and SD for Max width, Length and Weight measurements of elongated blanks."
 #|  - "Mean and SD for Max width, Length and Weight measurements of cores."
-#| layout-nrow: 2
+#| layout-nrow: 3
 
 kable(proto_stats_blank)
+kable(proto_stats_elong)
 kable(proto_stats_cores)
@@
 #| label: tbl-solutrean
-#| tbl-cap: "Mean and standard deviation (SD) for Max width, Length and Weight measurements of blanks (a) and cores (b) by chert type of the Solutrean assemblage (levels A to C) of the Shelter area."
+#| tbl-cap: "Mean and standard deviation (SD) for Max width, Length and Weight measurements of flakes (a), elongated blanks (b) and cores (c) by chert type of the Solutrean assemblage (levels A to C) of the Shelter area."
 #| tbl-subcap: 
-#|  - "Mean and SD for Max width, Length and Weight measurements of blanks."
+#|  - "Mean and SD for Max width, Length and Weight measurements of flakes."
+#|  - "Mean and SD for Max width, Length and Weight measurements of elongated blanks."
 #|  - "Mean and SD for Max width, Length and Weight measurements of cores."
-#| layout-nrow: 2
+#| layout-nrow: 3
 
 kable(solutrean_stats_blank)
+kable(solutrean_stats_elong)
 kable(solutrean_stats_cores)

### 2025-06-19 analysis/paper/paper.qmd
Review corrections. Tweeks to the ggsave code.

@@
 
-#ggsave("gravettian_plots.png", width = 6, height = 6, units = "in", dpi = 300, bg = "white")
+#ggsave("gravettian_plots.png", width = 6, height = 7, units = "in", dpi = 300, bg = "white")
 
@@
 
-#ggsave("proto_plots.png", width = 6, height = 6, units = "in", dpi = 300, bg = "white")
+#ggsave("proto_plots.png", width = 6, height = 7, units = "in", dpi = 300, bg = "white")
 
@@
 
-#ggsave("solutrean_plots.png", width = 6, height = 6, units = "in", dpi = 300, bg = "white")
+#ggsave("solutrean_plots.png", width = 6, height = 7, units = "in", dpi = 300, bg = "white")
 

### 2025-06-20 analysis/supplementary-materials/ESM_1.qmd
Article review. Statistical analysis and stats changed to separate flakes from elongated blanks.

@@
 
+```{r}
+#| label: stats-wilcox-review
+#| include: false
+
+# Terrace wilcox statistics
+## DB prep
+wilcox_t <- phase_dataset |> 
+  dplyr::mutate(Prov = case_when(
+    RMTYPE == "T1" | RMTYPE == "T2" | RMTYPE == "T3" | RMTYPE == "T4" | RMTYPE == "T5" ~ "Local",
+    RMTYPE == "T6" | RMTYPE == "T7" | RMTYPE == "T8" | RMTYPE == "T9" ~ "Non-local",
+    TRUE ~ RMTYPE)) |> 
+  filter(Prov %in% c("Local", "Non-local")) |> 
+  select(ID, CLASS_ELONG, THICKNESS, MAXWIDTH, LENGTH, WEIGHT, Prov, Level)
+
+# Proto-Solutrean flakes statistical test
+## NL cores were not represented enough (n) for statistics
+wt_flake_proto <- wilcox_t |> 
+  filter(Level %in% c("4E", "4E/5")) |> 
+  filter(CLASS_ELONG == "Flake")
+
+## Flake weight (fw)
+dplyr::group_by(wt_flake_proto, Prov) |> 
+  summarise(count = n(), median = median(WEIGHT, na.rm = TRUE))
+res_fw_proto <- wilcox.test(WEIGHT ~ Prov, data = wt_flake_proto, exact = FALSE)
+## Flake length (fl)
+dplyr::group_by(wt_flake_proto, Prov) |> 
+  summarise(count = n(), median = median(LENGTH, na.rm = TRUE))
+res_fl_proto <- wilcox.test(LENGTH ~ Prov, data = wt_flake_proto, exact = FALSE)
+## Flake width (fwi)
+dplyr::group_by(wt_flake_proto, Prov) |> 
+  summarise(count = n(), median = median(MAXWIDTH, na.rm = TRUE))
+res_fwi_proto <- wilcox.test(MAXWIDTH ~ Prov, data = wt_flake_proto, exact = FALSE)
+
+# Proto-Solutrean elongated blanks statistical test
+wt_elong_proto <- wilcox_t |> 
+  filter(Level %in% c("4E", "4E/5")) |> 
+  filter(CLASS_ELONG == "ElongatedProd")
+
+## Elongated blank weight (ew)
+dplyr::group_by(wt_elong_proto, Prov) |> 
+  summarise(count = n(), median = median(WEIGHT, na.rm = TRUE))
+res_ew_proto <- wilcox.test(WEIGHT ~ Prov, data = wt_elong_proto, exact = FALSE)
+## Elongated blank length (el)
+dplyr::group_by(wt_elong_proto, Prov) |> 
+  summarise(count = n(), median = median(LENGTH, na.rm = TRUE))
+res_el_proto <- wilcox.test(LENGTH ~ Prov, data = wt_elong_proto, exact = FALSE)
+## Elongated blank width (ewi)
+dplyr::group_by(wt_elong_proto, Prov) |> 
+  summarise(count = n(), median = median(MAXWIDTH, na.rm = TRUE))
+res_ewi_proto <- wilcox.test(MAXWIDTH ~ Prov, data = wt_elong_proto, exact = FALSE)
+
+# Gravettian flakes statistical test
+wt_flake_grav <- wilcox_t |> 
+  filter(Level %in% c("6", "7")) |> 
+  filter(CLASS_ELONG == "Flake")
+
+## Flake weight
+dplyr::group_by(wt_flake_grav, Prov) |> 
+  summarise(count = n(), median = median(WEIGHT, na.rm = TRUE))
+res_fw_grav <- wilcox.test(WEIGHT ~ Prov, data = wt_flake_grav, exact = FALSE)
+## Flake length
+dplyr::group_by(wt_flake_grav, Prov) |> 
+  summarise(count = n(), median = median(LENGTH, na.rm = TRUE))
+res_fl_grav <- wilcox.test(LENGTH ~ Prov, data = wt_flake_grav, exact = FALSE)
+## Flake width
+dplyr::group_by(wt_flake_grav, Prov) |> 
+  summarise(count = n(), median = median(MAXWIDTH, na.rm = TRUE))
+res_fwi_grav <- wilcox.test(MAXWIDTH ~ Prov, data = wt_flake_grav, exact = FALSE)
+
+# Gravettian elongated blanks statistical test
+wt_elong_grav <- wilcox_t |> 
+  filter(Level %in% c("6", "7")) |> 
+  filter(CLASS_ELONG == "ElongatedProd")
+
+## Elongated blanks weight
+dplyr::group_by(wt_elong_grav, Prov) |> 
+  summarise(count = n(), median = median(WEIGHT, na.rm = TRUE))
+res_ew_grav <- wilcox.test(WEIGHT ~ Prov, data = wt_elong_grav, exact = FALSE)
+## Elongated blanks length
+dplyr::group_by(wt_elong_grav, Prov) |> 
+  summarise(count = n(), median = median(LENGTH, na.rm = TRUE))
+res_el_grav <- wilcox.test(LENGTH ~ Prov, data = wt_elong_grav, exact = FALSE)
+## Elongated blanks width
+dplyr::group_by(wt_elong_grav, Prov) |> 
+  summarise(count = n(), median = median(MAXWIDTH, na.rm = TRUE))
+res_ewi_grav <- wilcox.test(MAXWIDTH ~ Prov, data = wt_elong_grav, exact = FALSE)
+
+## Gravettian cores statistical test
+wt_core_grav <- wilcox_t |> 
+  filter(Level %in% c("6", "7")) |> 
+  filter(CLASS == "Core")
+
+## Cores weight
+dplyr::group_by(wt_core_grav, Prov) |> 
+  summarise(count = n(), median = median(WEIGHT, na.rm = TRUE))
+res_cw_grav <- wilcox.test(WEIGHT ~ Prov, data = wt_core_grav, exact = FALSE)
+## Cores length
+dplyr::group_by(wt_core_grav, Prov) |> 
+  summarise(count = n(), median = median(LENGTH, na.rm = TRUE))
+res_cl_grav <- wilcox.test(LENGTH ~ Prov, data = wt_core_grav, exact = FALSE)
+## Cores width
+dplyr::group_by(wt_core_grav, Prov) |> 
+  summarise(count = n(), median = median(MAXWIDTH, na.rm = TRUE))
+res_cwi_grav <- wilcox.test(MAXWIDTH ~ Prov, data = wt_core_grav, exact = FALSE)
+
+# Shelter wilcox statistics
+## DB prep
+wilcox_t_shelter <- solutrean_dataset |> 
+  dplyr::mutate(Prov = case_when(
+    RMTYPE == "T1" | RMTYPE == "T2" | RMTYPE == "T3" | RMTYPE == "T4" | RMTYPE == "T5" ~ "Local",
+    RMTYPE == "T6" | RMTYPE == "T7" | RMTYPE == "T8" | RMTYPE == "T9" | RMTYPE == "T11" ~ "Non-local",
+    TRUE ~ RMTYPE)) |> 
+   mutate(CLASS = case_when(
+    CLASS == "ElongBlank" ~ "ElongatedProd",
+    TRUE ~ CLASS)) |>
+  filter(Prov %in% c("Local", "Non-local")) |> 
+  select(ID, CLASS, THICKNESS, MAXWIDTH, LENGTH, WEIGHT, Prov, LEVEL)
+
+# Solutrean flakes statistical test
+wt_flake_s <- wilcox_t_shelter |>
+  filter(CLASS == "Flake")
+
+## Flake weight
+dplyr::group_by(wt_flake_s, Prov) |> 
+  summarise(count = n(), median = median(WEIGHT, na.rm = TRUE))
+res_fw_s <- wilcox.test(WEIGHT ~ Prov, data = wt_flake_s, exact = FALSE)
+## Flake length
+dplyr::group_by(wt_flake_s, Prov) |> 
+  summarise(count = n(), median = median(LENGTH, na.rm = TRUE))
+res_fl_s <- wilcox.test(LENGTH ~ Prov, data = wt_flake_s, exact = FALSE)
+## Flake width
+dplyr::group_by(wt_flake_s, Prov) |> 
+  summarise(count = n(), median = median(MAXWIDTH, na.rm = TRUE))
+res_fwi_s <- wilcox.test(MAXWIDTH ~ Prov, data = wt_flake_s, exact = FALSE)
+
+# Solutrean elongated blanks statistical test
+wt_elong_s <- wilcox_t_shelter |>
+  filter(CLASS == "ElongatedProd")
+
+## Elongated blanks weight
+dplyr::group_by(wt_elong_s, Prov) |> 
+  summarise(count = n(), median = median(WEIGHT, na.rm = TRUE))
+res_ew_s <- wilcox.test(WEIGHT ~ Prov, data = wt_elong_s, exact = FALSE)
+## Elongated blanks length
+dplyr::group_by(wt_elong_s, Prov) |> 
+  summarise(count = n(), median = median(LENGTH, na.rm = TRUE))
+res_el_s <- wilcox.test(LENGTH ~ Prov, data = wt_elong_s, exact = FALSE)
+## Elongated blanks width
+dplyr::group_by(wt_elong_s, Prov) |> 
+  summarise(count = n(), median = median(MAXWIDTH, na.rm = TRUE))
+res_ewi_s <- wilcox.test(MAXWIDTH ~ Prov, data = wt_elong_s, exact = FALSE)
+
+wt_core_s <- wilcox_t_shelter |>
+  filter(CLASS == "Core")
+
+## Cores weight
+dplyr::group_by(wt_core_s, Prov) |> 
+  summarise(count = n(), median = median(WEIGHT, na.rm = TRUE))
+res_cw_s <- wilcox.test(WEIGHT ~ Prov, data = wt_core_s, exact = FALSE)
+## Cores length
+dplyr::group_by(wt_core_s, Prov) |> 
+  summarise(count = n(), median = median(LENGTH, na.rm = TRUE))
+res_cl_s <- wilcox.test(LENGTH ~ Prov, data = wt_core_s, exact = FALSE)
+## Cores width
+dplyr::group_by(wt_core_s, Prov) |> 
+  summarise(count = n(), median = median(MAXWIDTH, na.rm = TRUE))
+res_cwi_s <- wilcox.test(MAXWIDTH ~ Prov, data = wt_core_s, exact = FALSE)
+
+# Flake stats table
+p.value_flake <- tibble(
+  Technocomplex = c("Gravettian", " "," ", " ", "Proto-Solutrean", " ", " ", " ", "Solutrean", " ", " ", " "),
+  Variable = c("Weight", "Length", "Width", "(n)","Weight", "Length", "Width","(n)", "Weight", "Length", "Width", "(n)"),
+  L_median = c("3.2", "25.3", "21.4", "65", "2.5", "22.1", "18.8", "123", "1.9", "24", "18.2", "240"),
+  NL_median = c("2", "20.9", "19.3", "60", "2.2", "19.1", "17.7", "36", "1.55", "20.7", "20.1", "65"),
+  P_Value = round(c(res_fw_grav$p.value, res_fl_grav$p.value, res_fwi_grav$p.value, 0, res_fw_proto$p.value, res_fl_proto$p.value, res_fwi_proto$p.value, 0, res_fw_s$p.value, res_fl_s$p.value, res_fwi_s$p.value, 0), 5)) |>   mutate(Significance = case_when(
+      P_Value <= 0.001 ~ "***",
+      P_Value <= 0.01  ~ "**",
+      P_Value <= 0.05  ~ "*",
+      TRUE             ~ ""),
+    P_value = paste0(format(P_Value, nsmall = 4), Significance)  # Keep decimal places
+  ) |> 
+  select(-P_Value, -Significance) |> 
+  dplyr::mutate(P_value = case_when(
+    P_value == "0.00000***" ~ "-",
+    TRUE ~ P_value))
+
+# Elongated blanks stats table
+p.value_elong <- tibble(
+  Technocomplex = c("Gravettian", " "," ", " ", "Proto-Solutrean", " ", " ", " ", "Solutrean", " ", " ", " "),
+  Variable = c("Weight", "Length", "Width", "(n)","Weight", "Length", "Width","(n)", "Weight", "Length", "Width", "(n)"),
+  L_median = c("1.4", "27.4", "10.7", "17", "1.2", "29", "11", "31", "1.25", "28", "11.4", "98"),
+  NL_median = c("0.6", "23.7", "9.01", "9", "0.7", "21.5", "10.4", "13", "0.9", "28", "10.2", "15"),
+  P_Value = round(c(res_ew_grav$p.value, res_el_grav$p.value, res_ewi_grav$p.value, 0, res_ew_proto$p.value, res_el_proto$p.value, res_ewi_proto$p.value, 0, res_ew_s$p.value, res_el_s$p.value, res_ewi_s$p.value, 0), 5)) |>   mutate(Significance = case_when(
+      P_Value <= 0.001 ~ "***",
+      P_Value <= 0.01  ~ "**",
+      P_Value <= 0.05  ~ "*",
+      TRUE             ~ ""),
+    P_value = paste0(format(P_Value, nsmall = 4), Significance)  # Keep decimal places
+  ) |> 
+  select(-P_Value, -Significance) |> 
+  dplyr::mutate(P_value = case_when(
+    P_value == "0.00000***" ~ "-",
+    TRUE ~ P_value))
+
+p.value_cores <- tibble(
+  Technocomplex = c("Gravettian", "", "", "", "Solutrean", "", "", ""),
+  Variable = c("Weight", "Length", "Width", "(n)", "Weight", "Length", "Width", "(n)"),
+  L_median = c("20", "29.6", "27.6", "14", "16.5", "28.3", "30.7", "59"),
+  NL_median = c("11.8", "21.9", "29.7", "10", "11.7", "25.6", "26.0", "7"),
+  P_Value = round(c(res_cw_grav$p.value, res_cl_grav$p.value, res_cwi_grav$p.value, 0, res_cw_s$p.value, res_cl_s$p.value, res_cwi_s$p.value, 0), 3)) |> 
+  mutate(Significance = case_when(
+      P_Value <= 0.001 ~ "***",
+      P_Value <= 0.01  ~ "**",
+      P_Value <= 0.05  ~ "*",
+      TRUE             ~ ""),
+    P_value = paste0(format(P_Value, nsmall = 3), Significance)  # Keep decimal places
+  ) |> 
+  select(-P_Value, -Significance) |> 
+  dplyr::mutate(P_value = case_when(
+    P_value == "0.000***" ~ "-",
+    TRUE ~ P_value))
+
+```
+
 A Mann-Whitney U statistical test was conducted to determine whether there are differences in the measurements (weight, length and width) of blanks and cores between local and non-local cherts.
